### 2020-06-05 submission/manuscript.Rmd
Initial draft

---
title: ''
csl: msphere.csl
fontsize: 11pt
output:
  word_document: default
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  html_document:
    df_print: paged
geometry: margin=1.0in
bibliography: references.bib
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
options(tidyverse.quiet = TRUE)

library(tidyverse)
library(knitr)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)

	if(is.list(x)){
		x <- unlist(x)
	}

	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
		}
	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```

\vspace{35mm}

# Initial gut microbiota and response to antibiotic perturbation influence *Clostridioides difficile* colonization in mice

\vspace{35mm}

Sarah Tomkovich${^1}$, Joshua M.A. Stough${^1}$, Lucas Bishop${^1}$, Patrick D. Schloss^1$\dagger$^

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109

\newpage
\linenumbers

## Abstract
The microbiota plays a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs). However, much of the mechanistic work examining CDIs in mouse models use a single university colony or vendor, which have lower inter-individual microbiota variation compared to humans. We treated mice from 6 different colony sources (2 University of Michigan colonies and 4 vendors) with a single clindamycin dose, followed by *C. difficile* challenge 1 day later and measured *C. difficile* colonization levels through 9 days post-infection. The microbiota was profiled throughout the experiment via 16S rRNA gene sequencing analysis to examine variation across colony sources and alterations due to clindamycin treatment and *C. difficile* challenge. While all sources of mice were colonized 1-day post-infection, variation in *C. difficile* colonization levels emerged from days 3-7 post-infection with 3 sources colonized with *C. difficile* for longer and at higher levels. We identified bacterial taxa with different relative abundances across colony sources throughout the experiment, as well as taxa that were consistently impacted by clindamycin treatment. We created bacterial community-based logistic regression models that successfully classified mice based on their day 7 *C. difficile* colonization status. Importantly, after examining the taxa that were most important to the classification models, we identified a subset of key taxa that varied across colony sources (*Bacteroides*, *Deferribacteraceae*), were altered by clindamycin (*Porphyromonadaceae*, *Ruminococcaceae*), or both (*Enterobacteriaceae*, *Enterococcus*, *Bifidobacteriaceae*, *Coriobacteriaceae*, *Lachnospiraceae*, and *Verrucomicrobiaceae*). These results suggest the response of the initial gut microbiota to clindamycin treatment influences *C. difficile* 630 colonization dynamics.

## Importance
*Clostridioides difficile* is a leading nosocomial infection. Although the microbiota has been established as a key risk factor, there is variation in who becomes asymptomatically colonized, develops an infection, or has an infection with adverse outcomes. *C. difficile* infection (CDI) mouse models are widely used to answer a variety of *C. difficile* pathogenesis questions. However, the inter-individual variation between mice is less than what is observed in humans, particularly if just one source of mice is used. In this study, we administered clindamycin to mice from 6 different colony sources and challenged them with *C. difficile*. Interestingly, only a subset of the taxa that vary across sources were associated with how long *C. difficile* was able to colonize. Future studies examining the interplay between the microbiota and *C. difficile* should consider using mice from multiple sources to narrow down the microbes driving the observed phenotypes and reflect human interindividual variation.

\newpage

## Introduction

Antibiotics are a clear risk factor for *Clostridioides difficile* infections (CDIs), but there is variation in who goes on to develop severe or recurrent CDIs after exposure [@Teng2019; @Kelly2012]. Additionally, asymptomatic colonization, where *C. difficile* is detectable, but symptoms are absent has been documented in infants and adults [@Zacharioudakis2015; @Crobach2018]. The intestinal microbiome has been implicated in asymptomatic colonization [@Zhang2015; @VanInsberghe2020], susceptibility to CDIs [@Mancabelli2017, @Duvallet2017], and adverse CDI outcomes [@Seekatz2016; @Khanna2016; @Pakpour2017; @Lee2020].

Mouse models of CDIs have been a great tool for understanding *C. difficile* pathogenesis [@Hutton2014]. The number of CDI mouse model studies has grown substantially since Chen et al. published their C57BL/6 model in 2008, which disrupted the gut microbiota with antibiotics to enable *C. difficile* colonization, diarrhea, and weight loss [@Chen2008]. CDI mouse models have been used to examine translationally relevant questions regarding *C. difficile*, including the role of the microbiota and efficacy of potential therapeutics for treating CDIs [@Best2012]. However, microbiome variation between lab mice is much less than the variation observed between humans [@Baxter2014; @Nagpal2018]. Additionally, studying the contribution of the microbiota to a particular disease phenotype in one set of lab mice after the same perturbation could yield a number of findings of which only a fraction may be driving the phenotype.

In the past, our group has attempted to introduce more microbiome variation into the CDI mouse model by using a variety of antibiotic treatments [@Reeves2011; @Schubert2015; @Jenior2017; @Jenior2018]. An alternative approach to maximize microbiome variation is to use mice from multiple sources [@Velazquez2019; @Osbelt2020]. Microbiome differences between different mouse vendors have been well documented and shown to influence susceptibility to a variety of diseases [@Stough2016; @Alegre2019], including enteric infections [@EtienneMesmin2017; @Velazquez2019; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018]. Additionally, different research groups have observed different CDI outcomes in mice despite using similar models and the microbiome has been proposed as one factor potentially mediating susceptibility [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014]. Here we examined how variations in the baseline microbiome and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C. difficile* colonization and the time needed to clear the infection.

```{r, cfu}
cfu_stats_all <- read_tsv("../data/process/cfu_stats_all_days.tsv")
padj_d1_cfu <- cfu_stats_all %>% filter(day == 1) %>% pull(p.value.adj) 
mean_d1 <- cfu_stats_all %>% filter(day == 1) %>% select(-day, -statistic, -p.value, -parameter, -method, -p.value.adj) %>% pivot_longer(everything(), values_to = "mean_cfu", names_to = "Source") %>% arrange(mean_cfu)
min_mean_d1 <- mean_d1 %>%  select(mean_cfu) %>% min
max_mean_d1 <- mean_d1 %>% select(mean_cfu) %>% max

cfu_stats_sig <- cfu_stats_all %>% filter(p.value.adj < 0.05)
max_padj_d3to7 <- cfu_stats_sig %>% select(p.value.adj) %>% max
```
## Results

**Clindamycin treatment renders all mice susceptible to *C. difficile* 630 colonization regardless of colony source.** To test how the microbiotas of mice from different colony sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan (the Young and Schloss lab colonies), the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan). These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019]. After a 13-day acclimation period for the mice ordered from vendors, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C. difficile* 630 spores (Fig. 1A). Clindamycin was chosen because we have previously demonstrated mice are rendered susceptible, but consistently cleared the CDI within 9 days [@Jenior2018; @Tomkovich2019], clindamycin is frequently implicated with human CDIs [@Guh2018], and is also part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008]. The day after infection, *C. difficile* was detectable in all mice at a similar level (Mean CFU range: `r format(min_mean_d1, digits = 2, scientific=TRUE)`-`r format(max_mean_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of colony source (Fig. 1B). Interestingly, variation in *C. difficile* CFU levels across sources of mice emerged from days 3-7 post-infection (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig. 1B and Table S1), suggesting mouse colony source is associated with *C. difficile* clearance. We conducted two experiments approximately 3 months apart and combined the data because the colonization dynamics across sources of mice varied similarily in both experiments (Fig. S1). Although *C. difficile* 630 causes mild symptoms in mice comparied to other *C. difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig. 1C and Table S2). Interestingly, mice ordered from Jackson, Taconic, and Envigo tended to lose the most weight, have higher *C. difficile* CFU levels and take longer to clear the infection compared to the other sources of mice, which was particularly evident 7 days post-infection (Fig. 1C and D). Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection. Thus, clindamycin rendered all mice susceptible to *C. difficile* 630 colonization, regardless of colony source, but variation across sources emerged with 3 out of 6 sources taking longer to clear *C. difficile*.

```{r, alpha}
richness_stats_all <- read_tsv("../data/process/richness_stats_all_days.tsv")
padj_dn1_richness <- richness_stats_all %>% filter(day == -1) %>% pull(p.value.adj)

shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
```
**Bacterial communities vary across mouse colony sources.** Given the well known variation in mouse microbiomes across vendors and university colonies [@Alegre2019], we hypothesized that the variation in *C. difficile* clearance could be explained by microbiota variation across the 6 sources. We used 16S rRNA gene sequencing to characterize the fecal bacterial communities from the mice over the course of the experiment. Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity measures across the 6 sources of mice. Examining the bacterial communities at baseline, prior to clindamycin treatment there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig. 2A-B and Table S3). As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and communities started to recover 1 day post-infection (Fig. 2C-D). Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after both clindamycin and *C. difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other groups (Fig 2C-F and Table S4). While Charles River mice had more diverse microbiotas, Young and Schloss lab mice were also able to clear *C. difficile* faster, suggesting microbiota diversity alone does not explain the observed variation in *C. difficile* colonization across vendors.
```{r, adonis}
adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max

adonis_stats_dn101 <- read_tsv("../data/process/adonis_dn1-1.tsv")
dn1_rsq_sum <- adonis_stats_dn101 %>% filter(day == -1) %>% select(r_sq) %>% sum
d0_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 0) %>% select(r_sq) %>% sum
d1_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 1) %>% select(r_sq) %>% sum
p_adonis_intial <- adonis_stats_dn101 %>% select(p) %>% max 
p_adonis_min <- adonis_stats_dn101 %>% select(p) %>% min #All p values are same
```
Next, we compared the bacterial communities from the 6 colonies over the course of the experiment using principal coordinate analysis (PCoA) of the Theta YC distances. Permutational multivariate analysis of variance (PERMANOVA) analysis revealed colony source was the major factor explaining the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment (Movie S1 and Table S5). Since, the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the bacterial communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1). For all 3 timepoints, source and the interaction with cage significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig. 3 and Table S6). We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig. S2 and Table S7), suggesting vendor communities were relatively stable between the 2 experiments. Thus, mouse colony source was the factor that explained the most variation observed in the bacterial communities. Importantly with the exception of the 2 University colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C. difficile* challenge.
```{r, source_taxa}
#Number of different otus and families across sources at day -1, 0, and 1 timepoints
source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
source_families <- read_tsv("../data/process/family_stats_dn1to1_combined.tsv")
source_families_dn1 <- source_families %>% filter(day == -1) %>% nrow()
source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
source_families_d0 <- source_families %>% filter(day == 0) %>% nrow()
source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()
source_families_d1 <- source_families %>% filter(day == 1) %>% nrow()

#Function to find which significant otus/genera/families are shared 
intersect_all <- function(a,b,...){
  Reduce(intersect, list(a,b,...))
}

#OTUs shared across baseline, post-clindamycin, and post-infection timepoints
source_o_dn1 <- source_otus %>% filter(day == -1) %>% pull(otu)
source_o_d0 <- source_otus %>% filter(day == 0) %>% pull(otu)
source_o_d1 <- source_otus %>% filter(day == 1) %>% pull(otu)
shared_source_otus_dn1to1 <- length(intersect_all(source_o_dn1, source_o_d0, source_o_d1))

#Families shared across baseline, post-clindamycin, and post-infection timepoints
source_f_dn1 <- source_families %>% filter(day == -1) %>% pull(family)
source_f_d0 <- source_families %>% filter(day == 0) %>% pull(family)
source_f_d1 <- source_families %>% filter(day == 1) %>% pull(family)
shared_source_families_dn1to1 <- length(intersect_all(source_f_dn1, source_f_d0, source_f_d1))
```
After finding differences at the community level, we next identified the taxa that varied across sources of mice over the initial days of the experiment. We examined bacterial relative abundances at the operational taxonomic unit (OTU) and family levels, expecting the number of differences to be reduced at the family level due to the nature of bacterial taxonomy [@Nguyen2015]. Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs and only `r source_families_dn1` families (Table S8-9) with relative abundances that varied across colony sources. Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs and `r source_families_d0` families (Table S8-9). After *C. difficile* challenge, there were `r source_otus_d1` OTUs and `r source_families_d1` families (Table S8-9) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment. In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 4A-C) and `r shared_source_families_dn1to1` families with relative abundances that consistently varied across colony sources (Fig. 4D-F). Importantly, some of the taxa that consistently varied across sources also shifted with clindamycin treatment. For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice. *Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment. In summary, mouse bacterial communities significantly varied according to colony source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C. difficile* challenge.

```{r, source_clindamycin}
clindamycin_otus <- read_tsv("../data/process/otu_stats_dn1to0.tsv")
clindamycin_otus_sig <- clindamycin_otus %>% filter(p.value.adj < 0.05) %>% nrow()

clindamycin_families <- read_tsv("../data/process/family_stats_dn1to0.tsv")
clindamycin_families_sig <- clindamycin_families %>% filter(p.value.adj < 0.05) %>% nrow()
```
**Clindamycin treatment alters a subset of taxa that were found in all colony sources.** Although there were bacteria that consistently varied across colony sources, we also wanted to identify the bacteria that shifted after clindamycin treatment, regardless of colony source. By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs and `r clindamycin_families_sig` families that were altered after clindamycin treatment (Fig. 5 and Table S10-11). Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) and 3 families (*Porphyromonadaceae*, *Enterococcaceae*, *Lachnospiraceae*) overlapped (Fig. 4, Fig. 5C-D). These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all colony sources and only a subset of the taxa also varied across colony sources.

```{r, classification}
models_vs_random <- read_tsv("../data/process/classification_to_random.tsv")
max_p_vs_random <- models_vs_random %>% select(p.value) %>% max
median_otu_day0_AUROC <- models_vs_random %>% filter(model_name == "otu_day0") %>% pull(median)

padj_pairwise_comp <- read_tsv("../data/process/classification_model_pairwise_stats.tsv")
max_padj_otu_day0_comp <- padj_pairwise_comp %>% filter(str_detect(comparison, "otu_day0")) %>% select(p.value.adj) %>% max

#Function to find which significant otus/genera/families are shared 
intersect_all <- function(a,b,...){
  Reduce(intersect, list(a,b,...))
}

#Read in components of venn diagram script to be able to calculate comparisons:----
#OTUs identified in logistic regression classification models (20 OTUs with the highest ranking for each model)
interp_otus <- read_tsv("../data/process/combined_top20_otus_all_models.tsv")
interp_otus_dn1 <- interp_otus  %>% filter(model_input_day == -1) %>% pull(OTU)
interp_otus_d0 <- interp_otus  %>% filter(model_input_day == 0) %>% pull(OTU)
interp_otus_d1 <- interp_otus  %>% filter(model_input_day == 1) %>% pull(OTU)
interp_combined <- c(interp_otus_dn1, interp_otus_d0, interp_otus_d1)

#OTUs that vary across sources of mice on day -1, 0, or 1
sig_otu_source <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
`sig_otu_day-1` <- sig_otu_source %>% filter(day == -1) %>% pull(otu)
`sig_otu_day0` <- sig_otu_source %>% filter(day == 0) %>% pull(otu)
`sig_otu_day1` <- sig_otu_source %>% filter(day == 1) %>% pull(otu)
#OTUs that consistently vary across sources of mice on day-1, 0, or 1
shared_sig_otus_Dn1toD1 <- intersect_all(`sig_otu_day-1`, sig_otu_day0, sig_otu_day1)

#OTUs that were impacted by clindamycin treatment
`sig_otu_pairs` <- read_tsv("../data/process/otu_stats_dn1to0.tsv") %>% 
  filter(p.value.adj < 0.05) %>% pull(otu)

# Day -1 overlapping OTUs----
# Overlap between OTUs that varied across sources of mice on day -1 and top 20 taxa from logistic regression model based on day -1 community:
dayn1_and_interp_dn1_o <- intersect_all(`sig_otu_day-1`, `interp_otus_dn1`)
# Overlap between OTUs that were altered by clindamycin treatment and top 20 taxa from logistic regression model based on day -1 community:
paired_and_interp_dn1_o <- intersect_all(`sig_otu_pairs`, `interp_otus_dn1`)

# Day 0 overlapping OTUs----
# Overlap between OTUs that varied across sources of mice on day 0 and top 20 taxa from logistic regression model based on day 0 community:
day0_and_interp_d0_o <- intersect_all(`sig_otu_day0`, `interp_otus_d0`)
# Overlap between OTUs that were altered by clindamycin treatment and top 20 taxa from logistic regression model based on day 0 community:
paired_and_interp_d0_o <- intersect_all(`sig_otu_pairs`, `interp_otus_d0`)

# Day 1 overlapping OTUs----
# Overlap between OTUs that varied across sources of mice on day 1 and top 20 taxa from logistic regression model based on day 1 community:
day1_and_interp_d1_o <- intersect_all(`sig_otu_day1`, `interp_otus_d1`)
# Overlap between OTUs that were altered by clindamycin treatment and top 20 taxa from logistic regression model based on day 1 community:
paired_and_interp_d1_o <- intersect_all(`sig_otu_pairs`, `interp_otus_d1`)

#Combined overlapping OTUs based on day -1, 0, and 1 comparisons----
#Combined overlapping OTUs that varied by source and were in the top 20 taxa from at least one logistic regression model & remove any duplicates
dayn1to1_and_interp_combined <- unique(c(`dayn1_and_interp_dn1_o`, `day0_and_interp_d0_o`, `day1_and_interp_d1_o`))
#See which source OTUs show up as significant over multiple days
source_otus_w_duplicates <- c(`dayn1_and_interp_dn1_o`, `day0_and_interp_d0_o`, `day1_and_interp_d1_o`)
key_source_otus <- sort(source_otus_w_duplicates[duplicated(source_otus_w_duplicates)])
#Key source OTUs: "Bacteroides (OTU 2)", "Enterococcus (OTU 23)"
#Combined overlapping OTUs that were altered by clindamycin treatment and were in the top 20 taxa from at least one logistic regression model
paired_and_interp_combined <- unique(c(`paired_and_interp_dn1_o`, `paired_and_interp_d0_o`, `paired_and_interp_d1_o`))
#See which clindamycin-associated OTUs show up as significant over multiple days
clind_otus_w_duplicates <- c(`paired_and_interp_dn1_o`, `paired_and_interp_d0_o`, `paired_and_interp_d1_o`)
key_clind_otus <- sort(clind_otus_w_duplicates[duplicated(clind_otus_w_duplicates)])
#List of key OTUs from source and clindamycin varying OTUs:
key_otus <- unique(c(key_source_otus, key_clind_otus))
list_key_otus <- paste(key_otus, collapse= ", ") #Format to output in paper

#Families identified in logistic regression classification models (20 families with the highest ranking for each model)
intrep_families <- read_tsv("../data/process/combined_top20_families_all_models.tsv")
interp_families_dn1 <- intrep_families %>% filter(model_input_day == -1) %>% pull(family)
interp_families_d0 <- intrep_families %>% filter(model_input_day == 0) %>% pull(family)
interp_families_d1 <- intrep_families %>% filter(model_input_day == 1) %>% pull(family)
interp_families_combined <- c(interp_families_dn1, interp_families_d0, interp_families_d1)
#Families identified in logistic regression classification models (20 families with the highest ranking for each model)
intrep_families <- read_tsv("../data/process/combined_top20_families_all_models.tsv")
interp_families_dn1 <- intrep_families %>% filter(model_input_day == -1) %>% pull(family)
interp_families_d0 <- intrep_families %>% filter(model_input_day == 0) %>% pull(family)
interp_families_d1 <- intrep_families %>% filter(model_input_day == 1) %>% pull(family)
interp_families_combined <- c(interp_families_dn1, interp_families_d0, interp_families_d1)

#Families that vary across sources of mice on day -1, 0, or 1.
sig_family_source <- read_tsv("../data/process/family_stats_dn1to1_combined.tsv")
`sig_family_day-1` <- sig_family_source %>% filter(day == -1) %>% pull(family)
`sig_family_day0` <- sig_family_source %>% filter(day == 0) %>% pull(family)
`sig_family_day1` <- sig_family_source %>% filter(day == 1) %>% pull(family)
#Families that consistently vary across sources of mice on day-1, 0, or 1
shared_sig_families_Dn1toD1 <- intersect_all(`sig_family_day-1`, sig_family_day0, sig_family_day1)

#Families that were impacted by clindamycin treatment
`sig_family_pairs` <- read_tsv("../data/process/family_stats_dn1to0.tsv") %>% 
  filter(p.value.adj < 0.05) %>% pull(family)

# Day -1 overlapping families----
# Overlap between families that varied across sources of mice on day -1 and top 20 taxa from logistic regression model based on day -1 community:
dayn1_and_interp_dn1_f <- intersect_all(`sig_family_day-1`, `interp_families_dn1`)
# Overlap between families that were altered by clindamycin treatment and top 20 taxa from logistic regression model based on day -1 community:
paired_and_interp_dn1_f <- intersect_all(`sig_family_pairs`, `interp_families_dn1`)

# Day 0 overlapping families----
# Overlap between families that varied across sources of mice on day 0 and top 20 taxa from logistic regression model based on day 0 community:
day0_and_interp_d0_f <- intersect_all(`sig_family_day0`, `interp_families_d0`)
# Overlap between families that were altered by clindamycin treatment and top 20 taxa from logistic regression model based on day 0 community:
paired_and_interp_d0_f <- intersect_all(`sig_family_pairs`, `interp_families_d0`)

# Day 1 overlapping families----
# Overlap between families that varied across sources of mice on day 1 and top 20 taxa from logistic regression model based on day 1 community:
day1_and_interp_d1_f <- intersect_all(`sig_family_day1`, `interp_families_d1`)
# Overlap between families that were altered by clindamycin treatment and top 20 taxa from logistic regression model based on day 1 community:
paired_and_interp_d1_f <- intersect_all(`sig_family_pairs`, `interp_families_d1`)

#Combined overlapping families based on day -1, 0, and 1 comparisons----
#Combined overlapping families that varied by source and were in the top 20 taxa from at least one logistic regression model & remove any duplicates
dayn1to1_and_interp_combined_f <- unique(c(`dayn1_and_interp_dn1_f`, `day0_and_interp_d0_f`, day1_and_interp_d1_f))
#See which source families show up as significant over multiple days
source_families_w_duplicates <- c(`dayn1_and_interp_dn1_f`, `day0_and_interp_d0_f`, day1_and_interp_d1_f)
key_source_families <- sort(unique(source_families_w_duplicates[duplicated(source_families_w_duplicates)]))
#Key source families: "Bacteroidaceae" -seen across 3 days, "Deferribacteraceae", "Enterococcaceae", "Lachnospiraceae"
#Combined overlapping families that were altered by clindamycin treatment and were in the top 20 taxa from at least one logistic regression model
paired_and_interp_combined_f <- unique(c(`paired_and_interp_dn1_f`,`paired_and_interp_d0_f`, paired_and_interp_d1_f))
#See which clindamycin-associated families show up as significant over multiple days
clind_familes_w_duplicates <- c(`paired_and_interp_dn1_f`,`paired_and_interp_d0_f`, paired_and_interp_d1_f)
key_clind_families <- sort(unique(clind_familes_w_duplicates[duplicated(clind_familes_w_duplicates)]))
#List of key families from source and clindamycin varying families:
key_families <- unique(c(key_source_families, key_clind_families))
list_key_families <- paste(key_families, collapse= ", ") #Format to output in paper

#Venn diagram calculations for OTUs----
venn_otus_source <- length(dayn1to1_and_interp_combined)
venn_otus_intersect <- length(intersect(dayn1to1_and_interp_combined, paired_and_interp_combined))
venn_otus_clind <-  length(paired_and_interp_combined)
#Venn diagram calculations for families
venn_families_source <- length(dayn1to1_and_interp_combined_f)
venn_families_intersect <- length(intersect(dayn1to1_and_interp_combined_f, paired_and_interp_combined_f))
venn_families_clind <-  length(paired_and_interp_combined_f)
```
**Source-specific and clindamycin impacted bacteria distinguish *C. difficile* colonization status in mice.** After identifying taxa that varied by colony source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C. difficile* colonization status at day 7 (Fig. 1D). We trained L2-regularized logistic regression models with input bacterial community data from the baseline, post-clindamycin, and post-infection timepoints of the experiment to predict *C. difficile* colonization status on day 7 (Fig. S3). All models were better at predicting *C. difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12), however the models trained with OTU level data generally performed better than those trained with family level data with the expception of the models based on the post-infection (day 1) communities (Fig. 6). Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than all other models with an AUROC of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13). Thus, we were able to use community bacterial relative abundance data alone to differentiate mice that were still colonized with *C. difficile* on day 7 from those mice that had cleared the *C. difficile*. Interestingly, the model built with OTU relative abundance data post-clindamycin treatment had the best performance, suggesting how the bacterial community responds to clindamycin treatment has the greatest influence on subsequent *C. difficile* colonization dynamics.

Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 taxa that had the highest absolute feature weights in each of the 6 models (Table S14-15). The top 20 taxa from each model were compared to the list of taxa that varied across colony source (Fig. 4 and Table S8-9) at the same timepoint and the taxa that were altered by clindamycin treatment (Fig. 5 and Table S10-11). We found a subset of OTUs and families that were important to the model and overlapped with bacteria that varied by either source, clindamycin treatment, or both (Fig. S4, S5, S6). Combining the overall results for the 3 OTU models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig. 7A). Combining the overall results for the 3 family models identified `r venn_families_source` families associated with source, `r venn_families_clind` families associated with clindamycin treatment and `r venn_families_intersect` families associated with both (Fig. 7B). Several OTUs (*`r list_key_otus`*) and families (*`r list_key_families`*) appeared across at least 2 models, so we examined how the relative abundances of these key taxa varied over the course of the experiment (Fig. 8 and Fig. S7). Throughout the experiment, there was at least 1 timepoint where relative abundances of these taxa significantly varied across sources (Table S16-17). Interestingly, there were no taxa that emerged as consistently enriched or depleted in Jackson, Taconic, and Envigo mice that were colonized for longer with *C. difficile* 630, suggesting multiple bacteria influence the time needed to clear the infection. Together, these results suggest the initial bacterial communities and their responses to clindamycin have a large influence on the time needed to clear *C. difficile*.

## Discussion

By examining the *C. difficile* colonization dynamics within mice from 6 different colony sources after perturbing the microbiota with clindamycin treatment, we were able to identify bacterial taxa that were unique to sources throughout the experiment as well as taxa that were universally impacted by clindamycin. We built L2 logistic regression models with baseline, post-clindamycin treatment, and post-infection fecal community data that successfully predicted *C. difficile* colonization status 7 days after infection better than random chance. We identified *`r list_key_otus`*, *`r list_key_families`* (Fig. 8, Fig. S7) as candidate bacteria within these communities that were influencing variation in *C. difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by colony source, were impacted by clindamycin treatment, or both. Overall, our results demonstrate clindamycin is sufficient to render mice from multiple sources susceptible to CDI and only a subset of the interindividual microbiota variation across mice from different sources was associated with the time needed to clear *C. difficile*.

Other groups have taken similar approaches by using mice from multiple colony sources to identify bacteria that either promote colonization resistance or increase susceptibility to enteric infections [@Velazquez2019; @EtienneMesmin2017; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018]. For example, in the context of *Salmonella* infections, *Enterobacteriaceae* and segmented filamentous bacteria have emerged as protective [@Velazquez2019; @Lai2020]. A previous study with *C. difficile* identified an endogenous protective *C. difficile* strain LEM1 that bloomed after antibiotic treatment in mice from Jackson or Charles River Laboratories, but not Taconic that protected mice against the more toxigenic *C. difficile* VPI10463 [@EtienneMesmin2017]. Given that we ordered mice from the same vendors, we checked all mice for endogenous *C. difficile* by plating stool samples that were collected after clindamycin treatment. However, we did not identify any endogenous *C. difficile* strains prior to challenge, suggesting there were no endogenous protective strains in the mice we received and other bacterial taxa were mediating the variation in *C. difficile* colonization across sources. Although all mice were susceptible to *C. difficile* colonization, by following colonization over time we found Jackson, Taconic, and Envigo mice remained colonized for longer. We identified a subset of bacteria that were important in predicting whether a mouse was still colonized with *C. difficile* status 7 days post-infection. These results suggest a subset of the bacterial community is responsible for determining the length of time needed to clear *C. difficile* colonization.

In the past variation between different CDI mouse model studies have been attributed to intestinal microbiome differences in mice across different institutional environments. For example, groups using the same clindamycin treatment and C57BL/6 mice had different *C. difficile* outcomes, one having sustained colonization [@Buffie2011], while the other had transient [@Reeves2011]. Baseline differences in the microbiota composition have been hypothesized to partially explain the differences in colonization outcomes and overall susceptibility to *C. difficile* after treatment with the same antibiotic [@Lawley2013; @Hutton2014]. We have shown that mice from 6 different sources were all susceptible to *C. difficile* 630, suggesting the microbiota influences *C. difficile* clearance more than susceptibility. Fortunately, the bacterial perturbations induced by clindamycin treatment have been well characterized and our findings agree with previous CDI mouse model work demonstrating *Enterococcus* and *Enterobacteriaceae* were associated with *C. difficile* susceptibility and *Porpyhromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, and *Turicibacter* were associated with resistance [@Schubert2015; @Jenior2018; @Tomkovich2019; @Buffie2011; @Buffie2014; @Lawley2009; @Lawley2012; @Jump2014]. While we have demonstrated that susceptibility is uniform across vendors after clindamycin treatment, there could be different outcomes for either susceptibility or clearance in the case of other antibiotic treatments. The *C. difficile* strain used could also be contributing to the variation in *C. difficile* outcomes seen across different groups [@Lawley2012]. We found the time needed to naturally clear *C. difficile* varied across sources of mice implying that at least in the context of the same perturbation, microbiota differences seemed to influence infection outcome more than susceptibility. More importantly, we were able to narrow down from all the variation observed across colony sources to a subset of bacterial taxa that were also important for predicting *C. difficile* colonization status 7 days post-infection.

Our approach successfully increased the diversity of murine bacterial communities tested in our clindamycin *C. difficile* model. One alternative approach that has been used in some CDI studies [@NagaoKitamoto2020; @Battaglioli2018; @Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018] is to use mice associated with human microbiotas. However, a major caveat to this method is the substantial loss of human microbiota community members upon transfer to mice [@Fouladi2020; @Walter2020]. Additionally with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of the CDI mouse model studies to date associated mice with just 1 types of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018], which does not aid in the goal of figuring out how a variety of unique microbiotas influence susceptiblity to CDIs and adverse outcomes. Encouragingly, decreased *Bifidobacterium*, *Porphyromonas*, *Ruminococcaceae* and *Lachnospiraceae* and increased *Enterobacteriaceae*, *Enterococcus*, *Lactobacillus*, and *Proteus* have all been associated with human CDIs [@Mancabelli2017, @Duvallet2017] and were well represented in our study, suggesting most of the vendors are suitable for gaining insights into microbiota associated factors influencing *C. difficile* colonization and infections in humans. An important exception was *Enterococcus*, which was primarily absent from the mice from University of Michigan colonies and *Proteus*, which was only found in Taconic mice. Importantly, the fact that some CDI associated bacteria were only found in a subset of mice has important implications for future CDI mouse model studies.

There are several limitations to our work. The microbiome is composed of viruses, fungi, and parasites in addition to bacteria, and these non-bacterial members can also vary across mouse vendors [@Rasmussen2019; @Mims2020]. While our study focused solely on the bacterial portion, viruses and fungi have also begun to be implicated in the context of CDIs or FMT treatments for recurrent CDIs [@Markey2018; @Stewart2019; @Ott2017; @Zuo2017; @Zuo2018]. Beyond community composition, the metabolic function of the microbiota also has a CDI signature [@Robinson2019; @Jenior2017; @Fletcher2018; @Jump2014] and can vary across mice from different sources [@Xiao2015]. For example, microbial metabolites, particularly secondary bile acids and butyrate production, have been implicated as important contributors to *C. difficile* resistance [@Buffie2014; @Lawley2012]. Although, we only looked at composition, *Ruminococcaceae* and *Lachnospiraceae* both emerged as important taxa for classifying day 7 *C. difficile* colonization status and metagenomes from these bacteria have been shown to contain the bile acid-inducible gene cluster necessary for secondary bile acid formation and ability to produce butyrate [@Vital2019; @Collins2015]. Interestingly, butyrate has previously been shown to vary across vendors and mediates resistance to *Citrobacter rodentium* infection, a model of enterohemorrhagic and enteropathogenic *Escherichia coli* infections [@Osbelt2020]. Evidence for immunological toning differences in IgA and Th17 cells across mice from different vendors have also been documented and [@Fransen2015; @Ivanov2009], may also influence response to CDI, particularly in the context of severe CDIs [@Azrad2018; @Saleh2019]. The outcome after *C. difficile* exposure depends on a multitude of factors, including age, diet, and immunity; all of which are also influenced by the microbiota. We have demonstrated that the ways different baseline microbiotas from different mouse colony sources respond to clindamycin treatment influences the length of time mice remained colonized with *C. difficile* 630. For those interested in dissecting the contribution of the microbiome to *C. difficile* pathogenesis and treatments, using multiple sources of mice may yield more insights than a single model alone. Furthermore, for studies wanting to examine the interplay between a particular bacterial taxon such as *Enterococcus* and *C. difficile*, these results could serve as a resource for selecting which mice to order to address the question.

\newpage

## Acknowledgements
This work was supported by the National Institutes of Health (U01AI124255). ST was supported by the Michigan Institute for Clincial and Health Research Postdoctoral Translation Scholars Program (UL1TR002240 and KL2TR002241). We thank members of the Schloss lab for feedback on planning the experiments and data presentation, as well as code tutorials and feedback through Code club. In particular, we want to thank Begum Topçuoğlu for help with implementing logistic regression models using her [pipeline](https://github.com/SchlossLab/ML_pipeline_microbiome) and Ana Taylor for help with some media preparation and sample collection. We also thank members of Vincent Young’s lab, particularly  Kimberly Vendrov, for guidance with the *C. difficile* infection mouse model and donating the mice. We also want to thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the institutional support for our mouse experiments. Finally, we thank Kwi Kim, Austin Campbell, and Kimberly Vendrov for their help in maintaining the Schloss lab’s anaerobic chamber.

\newpage

## Materials and Methods

**(i) Animals.** All experiments were approved by the University of Michigan Animal Care and Use Committee (IACUC) under protocol number PRO00006983. Female C57BL/7 mice were obtained from 6 different colony sources: The Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, Envigo, and two colonies at the University of Michigan (the Schloss lab colony and the Young lab colony). The Young lab colony was originally established with mice purchased from Jackson, and the Schloss lab colony was later founded with mice donated from the Young lab. The 4 groups of mice purchased from vendors were allowed to acclimate to the University of Michigan mouse facility for 13 days prior to starting the experiment. At least 4 female mice (age 5-10 weeks) were obtained per source and mice from the same source were primarily housed at a density of 2 mice per cage. The experiment was repeated once, approximately 3 months after the start of the first experiment.

**(ii) Antibiotic treatment.** After the 13-day acclimation period and 1 day prior to challenge (Fig. 1A), all mice received 10 mg/kg clindamycin (filter sterilized through a 0.22 micron syringe filter prior to administration) via intraperitoneal injection.

**(iii) *C. difficile* infection model.** Mice were challenged with 10^3^ spores of *C. difficile* strain 630 via oral gavage post-infection 1 day after clindamycin treatment as described previously [@Jenior2018]. Mice weights and stool samples were taken daily through 9 days post-challenge. Collected stool was split for *C. difficile* CFU quantification and 16S rRNA sequencing analysis. *C. difficile* quantification stool samples were transferred to the anaerobic chamber, serially diluted in PBS, plated on taurocholate-cycloserine-cefoxitin-fructose agar (TCCFA) plates, and counted after 24 hours of incubation at 37&deg;C under anaerobic conditions. A sample from the day 0 timepoint (post-clindamycin and prior to *C. difficile* challenge) was also plated on TCCFA to ensure mice were not already colonized with *C. difficile* prior to infection. There were 3 deaths recorded over the course of the experiment, 1 Taconic mouse died prior to *C. difficile* challenge and 1 Jackson and 1 Envigo mouse died between 1- and 3-days post-infection. Mice were categorized as cleared when no *C. difficile* was detected in the first serial dilution (limit of detection: 100 CFU). Stool samples for 16S rRNA sequencing were snap frozen in liquid nitrogen and stored at -80&deg;C until DNA extraction.

**(iv) 16S rRNA sequencing.** DNA was extracted from -80&deg;C stored stool samples using the DNeasy Powersoil HTP 96 kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf). The V4 region was amplified for 16S rRNA with the AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom barcoded primers, as previously described [@Kozich2013]. The ZymoBIOMICS microbial community DNA standards was used as a mock community control [@Sze2019] and water was used as a negative control per 96-well extraction plate. The PCR amplicons were cleaned up and normalized with the SequalPrep normalization plate kit (Thermo Fisher Scientific). Amplicons were pooled and quantified with the KAPA library quantification kit (KAPA biosystems), prior to sequencing using the MiSeq system (Illumina).

**(v) 16S rRNA gene sequence analysis.** mothur (v. 1.43) was used to process all sequences [@Schloss2009] with a previously published protocol [@Kozich2013]. Reads were combined and aligned with the SILVA reference database [@Quast2012]. Chimeras were removed with the VSEARCH algorithm and taxonomic assignment was completed with a modified version (v16) of the Ribosomal Database Project reference database (v11.5) [@Cole2013] with an 80% cutoff. Operational taxonomic units (OTUs) were assigned with a 97% similarity threshold using the opticlust algorithm [@Westcott2017]. To account for uneven sequencing across samples, samples were rarefied to 5,437 sequences 1,000 times for alpha and beta diversity analyses. PCoAs were generated based on Theta YC distances. Permutational multivariate analysis of variance (PERMANOVA) was performed on mothur-generated Theta YC distance matrices with the adonis function in the vegan package [@Vegan2018] in R [@r_citation_2018].

**(vi) Classification model training and evaluation.** Models were generated based on mice that were categorized as either cleared or colonized 7 days post-infection and had sequencing data from the baseline (day -1), post-clindamycin (day 0), and post-infection (day 1) timepoints of the experiment. Input bacterial community relative abundance data at either the OTU or family level from the baseline, post-clindamycin, and post-infection timepoints was used to generate 6 classification models that predicted *C. difficile* colonization status 7 days post-infection. The L2-regularized logistic regression models were trained and tested using the caret package [@Kuhn2008] in R as previously described [@Topcuoglu2019] with the exception that we used 60% training and 40% testing data splits for the cross-validation of the training data to select the best cost hyperparameter and the testing of the held out test data to measure model performance. The modified training/testing ratio was selected to accommodate the small number of samples in the dataset. Code was modified from https://github.com/SchlossLab/ML_pipeline_microbiome to update the classification outcomes and change the data split ratios. The modified repository to regenerate this analysis is available at https://github.com/tomkoset/ML_pipeline_microbiome.

**(vii) Statistical analysis.** All statistical tests were performed in R (v `r paste(version$major, ".", version$minor, sep = "")`) [@r_citation_2018]. The Kruskal-Wallis test was used to analyze differences in *C. difficile* CFU, mouse weight change, and alpha diversity across vendors with a Benjamini-Hochberg correction for testing multiple timepoints, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction. For taxonomic analysis and generation of logistic regression model input data, *C. difficile* (OTU 20) was removed. Bacterial relative abundances that varied across sources at the OTU and family taxonomic levels were identified with the Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified taxa at each level, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction. Taxa impacted by clindamycin treatment were identified using the Wilcoxon signed rank test with matched pairs of mice samples for day -1 and day 0. To determine whether classification models had better performance (test AUROCs) than random chance (0.5), we used the one-sample Wilcoxon signed rank test. To examine whether there was an overall difference in predictive performance across the 6 classification models we used the Kruskal-Wallis test followed by pairwise Wilcoxan comparisons with Benjamini-Hochberg correction for multiple hypothesis testing. The tidyverse package was used to wrangle and graph data (v `r paste(packageVersion("tidyverse")[1])`) [@Tidyverse2019].

**(viii) Code availability.** Code for all data analysis and generating this manuscript is available at https://github.com/SchlossLab/Tomkovich_vendor_difs_XXXX_2020.

**(ix) Data availability.** The 16S rRNA sequencing data have been deposited in the National Center for Biotechnology Information Sequence Read Archive (BioProject Accession no. PRJNA608529).

\newpage

## References

<div id="refs"></div>

\newpage

## Figures
```{r, sample_size}
#figure out in functions.R and export results as a tsv to read in
#analysis_n <- read_tsv("../data/process/____.tsv")

```
\includegraphics{figure_1.pdf}
**Figure 1. Clindamycin is sufficient to promote *C. difficile* colonization regardless of colony source, but clearance time varies across sources of C57BL/6 mice.** A. Setup of the experimental timeline. Mice for the experiments were obtained from 6 different sources: the Schloss (N =8) and Young lab (N = 9) colonies at the University of Michigan, the Jackson Laboratory (N = 8), Charles River Laboratory (N = 8), Taconic Biosciences (N = 8), and Envigo (N = 8). All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with *C. difficile* 630 spores on day 0. Mice were weighed and feces was collected daily through the end of the experiment (9 days post-infection). Note: 3 mice died during course of experiment. 1 Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse between 1- and 3-days post-infection. B. *C. difficile* CFU/gram stool measured over time (N = 20-49 mice per timepoint) via serial dilutions. The black line represents the limit of detection for the first serial dilution. CFU quantification data was not available for each mouse due to early deaths, stool sampling difficulties, and not plating all of the serial dilutions. C. Mouse weight change measured in grams over time (N = 45-49 mice per timepoint), all mice were normalized to the weight recorded 1 day before infection. For B-C: timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S1 and Table S2) are reflected by the asterisk(s) above each timepoint (\*\, P < 0.05, \*\*\, P < 0.01, \*\*\*\, P < 0.001). Lines represent the mean for each source and circles represent individual mice. D. *C. difficile* CFU status on Day 7 within each mouse colony source. Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU). Schloss (N = 6), Young (N = 9), Jackson (N = 5), Charles River (N = 8), Taconic (N = 5), Envigo (N = 7). Alternative: N = 5-9 mice per group.

\newpage
\includegraphics{figure_2.pdf}
**Figure 2. Differences in microbial richness and diversity across mouse colony sources emerge after clindamycin treatment and infection.** A-B. Boxplots of the number of observed OTUs and Shannon diversity index values at baseline: day -1 (A-B), after clindamycin: day 0 (C-D) and post-infection: day 1 (E-F) timepoints of the experiment. Each circle represents the value for a stool sample from an individual mouse. Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 for all panels except for B (Table S3). Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S4).

\newpage
\includegraphics{figure_3.pdf}
**Figure 3. Mouse colony source is the variable that explains most of the variation observed in the baseline, post-clindamycin, and post-infection bacterial communities.** A-C. Principal Coordinates Analysis of Theta YC distances from stools collected at baseline (A), post-clindamycin (B), and post-infection (C) timepoints of the experiment. Each circle represents a stool sample from an individual mouse. PERMANOVA analysis demonstrated that source and the interaction between source and cage explained most of the variation observed in the baseline (combined R^2^ = `r dn1_rsq_sum`), post-clindamycin (combined R^2^ = `r d0_rsq_sum`), and post-infection (combined R^2^ = `r d1_rsq_sum`) communities (all *P* = `r p_adonis_intial`, see Table S6).

\newpage
\includegraphics{figure_4.pdf}
**Figure 4. A subset of bacterial consistently vary across mouse colony sources despite clindamycin perturbation and *C. difficile* challenge.** A-C. Boxplots of the relative abundances for the `r shared_source_otus_dn1to1` OTUs that consistently varied across sources of mice at the baseline (A), post-clindamycin (B), and post-infection (C) timepoints of the experiment. D-F. Boxplots of the relative abundances for the `r shared_source_families_dn1to1` families that consistently varied across sources of mice at the baseline (D), post-clindamycin (E), and post-infection (F) timepoints of the experiment. For each timepoint bacteria with differential relative abundances across sources of mice were identified by Kruskal-Wallis test at the family and OTU level with Benjamini-Hochberg correction for testing all identified taxa at the respective level (Table S8-9). The grey vertical line indicates the limit of detection for A-F.

\newpage
\includegraphics{figure_5.pdf}
**Figure 5. Clindamycin treatment has the same effects on a subset of taxa regardless of colony source.** A-B. Boxplots of the top 10 most significant (adjusted P value < 0.05) OTUs with relative abundances that changed post clindamycin treatment. C-D. Boxplots of the top 10 most significant families with relative abundances that changed post clindamycin treatment. Data were analyzed by Wilcoxon signed rank test limited to mice that had paired sequence data for day -1 and 0 (N = 31). Tests were performed at the OTU and family level with Benjamini-Hochberg correction for testing all identified OTUs and families. See Table S10-11 for complete list of OTUs and families significantly impacted by clindamycin treatment. The grey vertical line indicates the limit of detection for A-D.

\newpage
\includegraphics{figure_6.pdf}
**Figure 6. Bacterial community composition before, after clindamycin perturbation, and post-infection can predict *C. difficile* colonization status 7 days post-challenge.** A-B. Logistic regression classification model AUROCs to predict *C. difficile* CFU on D7 (Fig. 1D, Fig. S3) based on the community relative abundances at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1) at either the OTU (A) or family (B) level. All models performed better than random chance (AUROC = 0.5), see (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13). A List of the 20 taxa that were ranked as most important to each model are listed in Table S14-15.

\newpage
\includegraphics{figure_7.pdf}
**Figure 7. Key taxa that influence *C. difficile* colonization status vary by mouse colony source, change after clindamycin treatment, or both.** A-B. Venn diagrams that combine Fig. S4-6 summaries of taxa that were important to the day -1, 0, and 1 classification models (Table S14-15) and either overlapped with taxa that varied across vendors at the same timepoint, were impacted by clindamycin treatment, or both. See Fig. S4-6 for separate comparisons of taxa from the day -1, 0, and 1 classification models. Bold taxa signify bacteria that were important to more than 1 classification model.

\newpage
\includegraphics{figure_8.pdf}
**Figure 8: Key OTUs vary across sources throughout the experiment.** A-C. Relative abundances of bold OTUs from Fig. 7A  that were important for at least two classification models are shown over time. A. *Enterococcus* (OTU 23), which significantly varied across sources and was impacted by clindamycin treatment. B. *Bacteroides* (OTU 2), which varied across sources throughout the experiment. C. *Enterobacteriaceae* (OTU 1) and *Porphyromonadaceae* (OTU 7) were significantly impacted by clindamycin treatment and examining relative abundance dynamics over the course of the experiment indicated timepoints where relative abundances also significantly varied across sources of mice. Each point represents the relative abundance of an individual mouse stool sample and colored lines represent the mean relative abundances for each source of mice. The gray horizontal line represents the limit of detection. Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S16) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05, \*\*\, P < 0.01, \*\*\*\, P < 0.001).

\newpage
\includegraphics{figure_S1.pdf}
**Figure S1. *C. difficile* CFU variation across vendors is similar between the 2 experiments.** (A-B). *C. difficile* CFU/gram of stool quantification over time for experiment 1 (A) and 2 (B). Experiments were conducted approximately 3 months apart. Lines represent the mean CFU for each source and circles represent individual mice and the black line represents the limit of detection.

\newpage
\includegraphics{figure_S2.pdf}
**Figure S2. Only bacterial communities from University of Michigan mice significantly vary between experiments.** A-F. PCoA of Theta YC distances for the baseline fecal bacterial communities within each source of mice. Each symbol represents a stool sample from an individual mouse with color corresponding to experiment and shape representing cage mates. PERMANOVA was performed within each group to examine the contributions of experiment and cage to observed variation. Experiment number and cage only significantly explained observed variation for mice from the Schloss (add r2 and P values) and Young (add r2 and P values) lab colonies (Table S5).

\newpage
\includegraphics{figure_S3.pdf}
**Figure S3. *C. difficile* colonization status 7 days post-infection varies across mice.** A. Bar graph visualizations of overall day 7 classification outcomes used to build logistic regression models. Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU) based on CFU g/stool data from 7 days post-infection. B-D. Classification input data used for models based on day -1 (B), day 0 (C), and day 1 (D) community relative abundance data. Only mice with day 7 *C. difficile* status and community relative abundance data for each respective timepoint were used as input data for the classification models.

\newpage
\includegraphics{figure_S4.pdf}
**Figure S4. Key taxa from classification model based on baseline community data vary by mouse colony source, clindamycin treatment, or both.** A-B. Venn diagrams of top 20 important OTUs (A) and families (B) from day -1 classification models (Table S14-15) that overlapped with taxa that varied across vendors at baseline, were impacted by clindamycin treatment, or both. Bold taxa signify bacteria that were important to more than 1 classification model.

\newpage
\includegraphics{figure_S5.pdf}
**Figure S5. Key taxa from classification model based on post-clindamycin treatment community data vary by mouse colony source, clindamycin treatment, or both.** A-B. Venn diagrams of top 20 important OTUs (A) and families (B) from day 0 classification models (Table S14-15) that overlapped with taxa that still varied across vendors after clindamycin, were impacted by clindamycin treatment, or both. Bold taxa signify bacteria that were important to more than 1 classification model.

\newpage
\includegraphics{figure_S6.pdf}
**Figure S6. Key taxa from classification model based on post-infection community data vary by mouse colony source, clindamycin treatment, or both.** A-B. Venn diagrams of top 20 important OTUs (A) and families (B) from day 1 classification models (Table S14-15) that overlapped with taxa that varied across vendors post-infection, were impacted by clindamycin treatment, or both. Bold taxa signify bacteria that were important to more than 1 classification model.

\newpage
\includegraphics{figure_S7.pdf}
**Figure S7. Key families vary across sources throughout experiment.** Relative abundances of bold families from Fig. 7B that were important for at least two classification models are shown over time. A. *Enterococcaceae* and *Lachnospiraceae*, which significantly varied across sources and were impacted by clindamycin treatment. B. *Bacteroidaceae* and *Deferribacteraceae*, which varied across sources throughout the experiment. C. *Bifidobacteriaceae*, *Coriobacteriaceae*, *Ruminococcaceae*, and *Verrucomicrobiaceae* were significantly impacted by clindamycin treatment. Examining the relative abundance dynamics throughout the experiment, identified timepoints where relative abundances also significantly varied across sources of mice. Each point represents the relative abundance of an individual mouse stool sample and colored lines represent the mean relative abundances for each source of mice. The gray horizontal line represents the limit of detection. Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S17) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05, \*\*\, P < 0.01, \*\*\*\, P < 0.001).

\newpage
## Supplementary Tables and Movie
All supplemental material is available at: https://github.com/SchlossLab/Tomkovich_vendor_difs_XXXX_2020/submission.

**Movie S1. Large shifts in bacterial community structure occurred after clindamycin and *C. difficile* infection.** PCoA of Theta YC distances animated from 0 through 9 days post-infection. Transparency of the circle corresponds to the day of the experiment, each circle represents a sample from an individual mouse at a specific timepoint. See Table S5 for PERMANOVA results).

**Table S1. *C. difficile* CFU statistical results.**

**Table S2. Mouse weight change statistical results.**

**Table S3. Diversity metrics Kruskal-Wallis statistical results.**

**Table S4. Diversity metrics pairwise Wilcoxon statistical results.**

**Table S5. PERMANOVA results for all mice, all timepoints.**

**Table S6. PERMANOVA results for all mice, all timepoints.**

**Table S7. PERMANOVA results of baseline communities within each source.**

**Table S8. OTUs with relative abudances that significantly vary across sources at baseline, post-clindamycin, or post-infection timepoints.**

**Table S9. Families with relative abudances that significantly vary across sources at baseline, post-clindamycin, or post-infection timepoints.**

**Table S10. OTUs with relative abudances that significantly changed after clindamycin treatment.**

**Table S11. Families with relative abudances that significantly changed after clindamycin treatment.**
**Table S12. Statistical results of logistic regression model performances compared to random chance.**

**Table S13. Pairwise Wilcoxan results of comparing all 6 logistic regression model performances.**

**Table S14. Top 20 most important OTUs for each of the 3 logistic regression models based on OTU relative abundance data.**

**Table S15. Top 20 most important families for each of the 3  logistic regression models based on OTU relative abundance data.**

**Table S16. OTUs with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment.**

**Table S17. Families with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment.**

### 2020-06-05 submission/manuscript.tex
Initial draft

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

\begin{document}

\vspace{35mm}

\section{\texorpdfstring{Initial gut microbiota and response to
antibiotic perturbation influence \emph{Clostridioides difficile}
colonization in
mice}{Initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice}}\label{initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}

\vspace{35mm}

Sarah Tomkovich\({^1}\), Joshua M.A.~Stough\({^1}\), Lucas
Bishop\({^1}\), Patrick D. Schloss\textsuperscript{1\(\dagger\)}

\vspace{40mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

\(1\) Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI 48109

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

The microbiota plays a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use a single university
colony or vendor, which have lower inter-individual microbiota variation
compared to humans. We treated mice from 6 different colony sources (2
University of Michigan colonies and 4 vendors) with a single clindamycin
dose, followed by \emph{C. difficile} challenge 1 day later and measured
\emph{C. difficile} colonization levels through 9 days post-infection.
The microbiota was profiled throughout the experiment via 16S rRNA gene
sequencing analysis to examine variation across colony sources and
alterations due to clindamycin treatment and \emph{C. difficile}
challenge. While all sources of mice were colonized 1-day
post-infection, variation in \emph{C. difficile} colonization levels
emerged from days 3-7 post-infection with 3 sources colonized with
\emph{C. difficile} for longer and at higher levels. We identified
bacterial taxa with different relative abundances across colony sources
throughout the experiment, as well as taxa that were consistently
impacted by clindamycin treatment. We created bacterial community-based
logistic regression models that successfully classified mice based on
their day 7 \emph{C. difficile} colonization status. Importantly, after
examining the taxa that were most important to the classification
models, we identified a subset of key taxa that varied across colony
sources (\emph{Bacteroides}, \emph{Deferribacteraceae}), were altered by
clindamycin (\emph{Porphyromonadaceae}, \emph{Ruminococcaceae}), or both
(\emph{Enterobacteriaceae}, \emph{Enterococcus},
\emph{Bifidobacteriaceae}, \emph{Coriobacteriaceae},
\emph{Lachnospiraceae}, and \emph{Verrucomicrobiaceae}). These results
suggest the response of the initial gut microbiota to clindamycin
treatment influences \emph{C. difficile} 630 colonization dynamics.

\subsection{Importance}\label{importance}

\emph{Clostridioides difficile} is a leading nosocomial infection.
Although the microbiota has been established as a key risk factor, there
is variation in who becomes asymptomatically colonized, develops an
infection, or has an infection with adverse outcomes. \emph{C.
difficile} infection (CDI) mouse models are widely used to answer a
variety of \emph{C. difficile} pathogenesis questions. However, the
inter-individual variation between mice is less than what is observed in
humans, particularly if just one source of mice is used. In this study,
we administered clindamycin to mice from 6 different colony sources and
challenged them with \emph{C. difficile}. Interestingly, only a subset
of the taxa that vary across sources were associated with how long
\emph{C. difficile} was able to colonize. Future studies examining the
interplay between the microbiota and \emph{C. difficile} should consider
using mice from multiple sources to narrow down the microbes driving the
observed phenotypes and reflect human interindividual variation.

\newpage

\subsection{Introduction}\label{introduction}

Antibiotics are a clear risk factor for \emph{Clostridioides difficile}
infections (CDIs), but there is variation in who goes on to develop
severe or recurrent CDIs after exposure (1, 2). Additionally,
asymptomatic colonization, where \emph{C. difficile} is detectable, but
symptoms are absent has been documented in infants and adults (3, 4).
The intestinal microbiome has been implicated in asymptomatic
colonization (5, 6), susceptibility to CDIs (7), and adverse CDI
outcomes (9--12).

Mouse models of CDIs have been a great tool for understanding \emph{C.
difficile} pathogenesis (13). The number of CDI mouse model studies has
grown substantially since Chen et al. published their C57BL/6 model in
2008, which disrupted the gut microbiota with antibiotics to enable
\emph{C. difficile} colonization, diarrhea, and weight loss (14). CDI
mouse models have been used to examine translationally relevant
questions regarding \emph{C. difficile}, including the role of the
microbiota and efficacy of potential therapeutics for treating CDIs
(15). However, microbiome variation between lab mice is much less than
the variation observed between humans (16, 17). Additionally, studying
the contribution of the microbiota to a particular disease phenotype in
one set of lab mice after the same perturbation could yield a number of
findings of which only a fraction may be driving the phenotype.

In the past, our group has attempted to introduce more microbiome
variation into the CDI mouse model by using a variety of antibiotic
treatments (18--21). An alternative approach to maximize microbiome
variation is to use mice from multiple sources (22, 23). Microbiome
differences between different mouse vendors have been well documented
and shown to influence susceptibility to a variety of diseases (24, 25),
including enteric infections (22, 23, 26--30). Additionally, different
research groups have observed different CDI outcomes in mice despite
using similar models and the microbiome has been proposed as one factor
potentially mediating susceptibility (13, 18, 21, 31--33). Here we
examined how variations in the baseline microbiome and responses to
clindamycin treatment in C57BL/6 mice from six different sources
influenced susceptibility to \emph{C. difficile} colonization and the
time needed to clear the infection.

\subsection{Results}\label{results}

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization regardless of colony source.} To test how
the microbiotas of mice from different colony sources impact
colonization dynamics after clindamycin exposure, we utilized C57BL/6
mice from 6 different sources: two colonies from the University of
Michigan (the Young and Schloss lab colonies), the Jackson Laboratory,
Charles River Laboratories, Taconic Biosciences, and Envigo (which was
formerly Harlan). These 4 vendors were chosen because they represent
commonly used vendors for CDI studies in mice (26, 34--40). After a
13-day acclimation period for the mice ordered from vendors, all mice
were treated with 10 mg/kg clindamycin via intraperitoneal injection and
one day later challenged with 10\textsuperscript{3} \emph{C. difficile}
630 spores (Fig. 1A). Clindamycin was chosen because we have previously
demonstrated mice are rendered susceptible, but consistently cleared the
CDI within 9 days (21, 41), clindamycin is frequently implicated with
human CDIs (42), and is also part of the antibiotic treatment for the
frequently cited 2008 CDI mouse model (14). The day after infection,
\emph{C. difficile} was detectable in all mice at a similar level (Mean
CFU range: 3e+07-1.5e+08; \emph{P}\textsubscript{FDR} = 0.15),
indicating clindamycin rendered all mice susceptible regardless of
colony source (Fig. 1B). Interestingly, variation in \emph{C. difficile}
CFU levels across sources of mice emerged from days 3-7 post-infection
(all \emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 1B and Table S1),
suggesting mouse colony source is associated with \emph{C. difficile}
clearance. We conducted two experiments approximately 3 months apart and
combined the data because the colonization dynamics across sources of
mice varied similarily in both experiments (Fig. S1). Although \emph{C.
difficile} 630 causes mild symptoms in mice comparied to other \emph{C.
difficile} strains (43), we also saw that weight change significantly
varied across sources of mice with the most weight lost two days
post-infection (Fig. 1C and Table S2). Interestingly, mice ordered from
Jackson, Taconic, and Envigo tended to lose the most weight, have higher
\emph{C. difficile} CFU levels and take longer to clear the infection
compared to the other sources of mice, which was particularly evident 7
days post-infection (Fig. 1C and D). Importantly, there was also one
Jackson and one Envigo mouse that died between 1- and 3-days
post-infection. Thus, clindamycin rendered all mice susceptible to
\emph{C. difficile} 630 colonization, regardless of colony source, but
variation across sources emerged with 3 out of 6 sources taking longer
to clear \emph{C. difficile}.

\textbf{Bacterial communities vary across mouse colony sources.} Given
the well known variation in mouse microbiomes across vendors and
university colonies (25), we hypothesized that the variation in \emph{C.
difficile} clearance could be explained by microbiota variation across
the 6 sources. We used 16S rRNA gene sequencing to characterize the
fecal bacterial communities from the mice over the course of the
experiment. Since antibiotics and other risk factors of CDIs are
associated with decreased microbiota diversity (44), we first examined
alpha diversity measures across the 6 sources of mice. Examining the
bacterial communities at baseline, prior to clindamycin treatment there
was a significant difference in the number of observed OTUs
(\emph{P}\textsubscript{FDR} = 0.03), but not Shannon diversity index
(\emph{P}\textsubscript{FDR} = 0.052) across sources of mice (Fig. 2A-B
and Table S3). As expected, clindamycin treatment decreased richness and
Shannon diversity across all sources of mice, and communities started to
recover 1 day post-infection (Fig. 2C-D). Interestingly, significant
differences in diversity metrics across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after both
clindamycin and \emph{C. difficile} infection, with Charles River mice
having higher richness and Shannon Diversity than most of the other
groups (Fig 2C-F and Table S4). While Charles River mice had more
diverse microbiotas, Young and Schloss lab mice were also able to clear
\emph{C. difficile} faster, suggesting microbiota diversity alone does
not explain the observed variation in \emph{C. difficile} colonization
across vendors.

Next, we compared the bacterial communities from the 6 colonies over the
course of the experiment using principal coordinate analysis (PCoA) of
the Theta YC distances. Permutational multivariate analysis of variance
(PERMANOVA) analysis revealed colony source was the major factor
explaining the observed variation across fecal communities
(R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
interactions between cage and day of the experiment (Movie S1 and Table
S5). Since, the majority of the perturbations happened over the initial
days of the experiment, we decided to focus on the bacterial communities
at baseline (day -1), after clindamycin treatment (day 0), and
post-infection (day 1). For all 3 timepoints, source and the interaction
with cage significantly explained most of the observed community
variation (combined R\textsuperscript{2} = 0.90, 0.99, 0.88,
respectively; \emph{P} = 0.0001; Fig. 3 and Table S6). We also compared
baseline communities across the 2 experiments, and found experiment and
cage significantly explained the observed variation only for the Schloss
and Young lab mouse colonies (Fig. S2 and Table S7), suggesting vendor
communities were relatively stable between the 2 experiments. Thus,
mouse colony source was the factor that explained the most variation
observed in the bacterial communities. Importantly with the exception of
the 2 University colonies, the community of each source clustered apart
from one another suggesting each community had a unique response to
clindamycin treatment and \emph{C. difficile} challenge.

After finding differences at the community level, we next identified the
taxa that varied across sources of mice over the initial days of the
experiment. We examined bacterial relative abundances at the operational
taxonomic unit (OTU) and family levels, expecting the number of
differences to be reduced at the family level due to the nature of
bacterial taxonomy (45). Focusing on the baseline communities first,
there were 268 OTUs and only 20 families (Table S8-9) with relative
abundances that varied across colony sources. Clindamycin treatment
reduced the number of taxa with relative abundances that varied across
sources to 18 OTUs and 10 families (Table S8-9). After \emph{C.
difficile} challenge, there were 44 OTUs and 18 families (Table S8-9)
with significantly different relative abundances across sources, as the
communities started to recover from antibiotic treatment. In spite of
the experimental perturbations that occurred during these 3 timepoints,
there were 12 OTUs (Fig 4A-C) and 8 families with relative abundances
that consistently varied across colony sources (Fig. 4D-F). Importantly,
some of the taxa that consistently varied across sources also shifted
with clindamycin treatment. For example, \emph{Proteus} increased after
clindamycin treatment, but only in Taconic mice. \emph{Enterococcus} was
primarily found only in mice purchased from commercial vendors and also
increased after clindamycin treatment. In summary, mouse bacterial
communities significantly varied according to colony source throughout
the course of the experiment and a consistent subset of bacterial taxa
remained different across sources regardless of clindamycin and \emph{C.
difficile} challenge.

\textbf{Clindamycin treatment alters a subset of taxa that were found in
all colony sources.} Although there were bacteria that consistently
varied across colony sources, we also wanted to identify the bacteria
that shifted after clindamycin treatment, regardless of colony source.
By analyzing all mice that had sequence data from fecal samples
collected at baseline and after clindamycin treatment, we identified 153
OTUs and 18 families that were altered after clindamycin treatment (Fig.
5 and Table S10-11). Interestingly, when we compared the list of
significant clindamycin impacted bacteria with the bacteria that
consistently varied across groups over the initial 3 timepoints of our
experiment, we found 3 OTUs (\emph{Lachnospiraceae} (OTU 130),
\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) and 3
families (\emph{Porphyromonadaceae}, \emph{Enterococcaceae},
\emph{Lachnospiraceae}) overlapped (Fig. 4, Fig. 5C-D). These findings
demonstrate that clindamycin has a consistent impact on the fecal
bacterial communities of mice from all colony sources and only a subset
of the taxa also varied across colony sources.

\textbf{Source-specific and clindamycin impacted bacteria distinguish
\emph{C. difficile} colonization status in mice.} After identifying taxa
that varied by colony source, changed after clindamycin treatment, or
both, we next wanted to determine which taxa were influencing the
variation in \emph{C. difficile} colonization status at day 7 (Fig. 1D).
We trained L2-regularized logistic regression models with input
bacterial community data from the baseline, post-clindamycin, and
post-infection timepoints of the experiment to predict \emph{C.
difficile} colonization status on day 7 (Fig. S3). All models were
better at predicting \emph{C. difficile} colonization status on day 7
than random chance (all \emph{P} \(\le\) 5e-15; Table S12), however the
models trained with OTU level data generally performed better than those
trained with family level data with the expception of the models based
on the post-infection (day 1) communities (Fig. 6). Interestingly, the
model based on the post-clindamycin (day 0) community OTU data performed
significantly better than all other models with an AUROC of 0.75
(\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise comparisons;
Table S13). Thus, we were able to use community bacterial relative
abundance data alone to differentiate mice that were still colonized
with \emph{C. difficile} on day 7 from those mice that had cleared the
\emph{C. difficile}. Interestingly, the model built with OTU relative
abundance data post-clindamycin treatment had the best performance,
suggesting how the bacterial community responds to clindamycin treatment
has the greatest influence on subsequent \emph{C. difficile}
colonization dynamics.

Next, to examine the bacteria that were driving each model's
performance, we pulled out the top 20 taxa that had the highest absolute
feature weights in each of the 6 models (Table S14-15). The top 20 taxa
from each model were compared to the list of taxa that varied across
colony source (Fig. 4 and Table S8-9) at the same timepoint and the taxa
that were altered by clindamycin treatment (Fig. 5 and Table S10-11). We
found a subset of OTUs and families that were important to the model and
overlapped with bacteria that varied by either source, clindamycin
treatment, or both (Fig. S4, S5, S6). Combining the overall results for
the 3 OTU models identified 14 OTUs associated with source, 21 OTUs
associated with clindamycin treatment, and 6 OTUs associated with both
(Fig. 7A). Combining the overall results for the 3 family models
identified 18 families associated with source, 14 families associated
with clindamycin treatment and 8 families associated with both (Fig.
7B). Several OTUs (\emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)}) and families
(\emph{Bacteroidaceae, Deferribacteraceae, Enterococcaceae,
Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
Verrucomicrobiaceae}) appeared across at least 2 models, so we examined
how the relative abundances of these key taxa varied over the course of
the experiment (Fig. 8 and Fig. S7). Throughout the experiment, there
was at least 1 timepoint where relative abundances of these taxa
significantly varied across sources (Table S16-17). Interestingly, there
were no taxa that emerged as consistently enriched or depleted in
Jackson, Taconic, and Envigo mice that were colonized for longer with
\emph{C. difficile} 630, suggesting multiple bacteria influence the time
needed to clear the infection. Together, these results suggest the
initial bacterial communities and their responses to clindamycin have a
large influence on the time needed to clear \emph{C. difficile}.

\subsection{Discussion}\label{discussion}

By examining the \emph{C. difficile} colonization dynamics within mice
from 6 different colony sources after perturbing the microbiota with
clindamycin treatment, we were able to identify bacterial taxa that were
unique to sources throughout the experiment as well as taxa that were
universally impacted by clindamycin. We built L2 logistic regression
models with baseline, post-clindamycin treatment, and post-infection
fecal community data that successfully predicted \emph{C. difficile}
colonization status 7 days after infection better than random chance. We
identified \emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)},
\emph{Bacteroidaceae, Deferribacteraceae, Enterococcaceae,
Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
Verrucomicrobiaceae} (Fig. 8, Fig. S7) as candidate bacteria within
these communities that were influencing variation in \emph{C. difficile}
colonization dynamics since these bacteria were all important in the
logistic regression models and varied by colony source, were impacted by
clindamycin treatment, or both. Overall, our results demonstrate
clindamycin is sufficient to render mice from multiple sources
susceptible to CDI and only a subset of the interindividual microbiota
variation across mice from different sources was associated with the
time needed to clear \emph{C. difficile}.

Other groups have taken similar approaches by using mice from multiple
colony sources to identify bacteria that either promote colonization
resistance or increase susceptibility to enteric infections (22, 23,
26--30). For example, in the context of \emph{Salmonella} infections,
\emph{Enterobacteriaceae} and segmented filamentous bacteria have
emerged as protective (22, 27). A previous study with \emph{C.
difficile} identified an endogenous protective \emph{C. difficile}
strain LEM1 that bloomed after antibiotic treatment in mice from Jackson
or Charles River Laboratories, but not Taconic that protected mice
against the more toxigenic \emph{C. difficile} VPI10463 (26). Given that
we ordered mice from the same vendors, we checked all mice for
endogenous \emph{C. difficile} by plating stool samples that were
collected after clindamycin treatment. However, we did not identify any
endogenous \emph{C. difficile} strains prior to challenge, suggesting
there were no endogenous protective strains in the mice we received and
other bacterial taxa were mediating the variation in \emph{C. difficile}
colonization across sources. Although all mice were susceptible to
\emph{C. difficile} colonization, by following colonization over time we
found Jackson, Taconic, and Envigo mice remained colonized for longer.
We identified a subset of bacteria that were important in predicting
whether a mouse was still colonized with \emph{C. difficile} status 7
days post-infection. These results suggest a subset of the bacterial
community is responsible for determining the length of time needed to
clear \emph{C. difficile} colonization.

In the past variation between different CDI mouse model studies have
been attributed to intestinal microbiome differences in mice across
different institutional environments. For example, groups using the same
clindamycin treatment and C57BL/6 mice had different \emph{C. difficile}
outcomes, one having sustained colonization (32), while the other had
transient (18). Baseline differences in the microbiota composition have
been hypothesized to partially explain the differences in colonization
outcomes and overall susceptibility to \emph{C. difficile} after
treatment with the same antibiotic (13, 31). We have shown that mice
from 6 different sources were all susceptible to \emph{C. difficile}
630, suggesting the microbiota influences \emph{C. difficile} clearance
more than susceptibility. Fortunately, the bacterial perturbations
induced by clindamycin treatment have been well characterized and our
findings agree with previous CDI mouse model work demonstrating
\emph{Enterococcus} and \emph{Enterobacteriaceae} were associated with
\emph{C. difficile} susceptibility and \emph{Porpyhromonadaceae},
\emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and \emph{Turicibacter}
were associated with resistance (19, 21, 32, 33, 41, 46--48). While we
have demonstrated that susceptibility is uniform across vendors after
clindamycin treatment, there could be different outcomes for either
susceptibility or clearance in the case of other antibiotic treatments.
The \emph{C. difficile} strain used could also be contributing to the
variation in \emph{C. difficile} outcomes seen across different groups
(47). We found the time needed to naturally clear \emph{C. difficile}
varied across sources of mice implying that at least in the context of
the same perturbation, microbiota differences seemed to influence
infection outcome more than susceptibility. More importantly, we were
able to narrow down from all the variation observed across colony
sources to a subset of bacterial taxa that were also important for
predicting \emph{C. difficile} colonization status 7 days
post-infection.

Our approach successfully increased the diversity of murine bacterial
communities tested in our clindamycin \emph{C. difficile} model. One
alternative approach that has been used in some CDI studies (49--54) is
to use mice associated with human microbiotas. However, a major caveat
to this method is the substantial loss of human microbiota community
members upon transfer to mice (55, 56). Additionally with the exception
of 2 recent studies (49, 50), most of the CDI mouse model studies to
date associated mice with just 1 types of human microbiota either from a
single donor or a single pool from multiple donors (51--54), which does
not aid in the goal of figuring out how a variety of unique microbiotas
influence susceptiblity to CDIs and adverse outcomes. Encouragingly,
decreased \emph{Bifidobacterium}, \emph{Porphyromonas},
\emph{Ruminococcaceae} and \emph{Lachnospiraceae} and increased
\emph{Enterobacteriaceae}, \emph{Enterococcus}, \emph{Lactobacillus},
and \emph{Proteus} have all been associated with human CDIs (7) and were
well represented in our study, suggesting most of the vendors are
suitable for gaining insights into microbiota associated factors
influencing \emph{C. difficile} colonization and infections in humans.
An important exception was \emph{Enterococcus}, which was primarily
absent from the mice from University of Michigan colonies and
\emph{Proteus}, which was only found in Taconic mice. Importantly, the
fact that some CDI associated bacteria were only found in a subset of
mice has important implications for future CDI mouse model studies.

There are several limitations to our work. The microbiome is composed of
viruses, fungi, and parasites in addition to bacteria, and these
non-bacterial members can also vary across mouse vendors (57, 58). While
our study focused solely on the bacterial portion, viruses and fungi
have also begun to be implicated in the context of CDIs or FMT
treatments for recurrent CDIs (35, 59--62). Beyond community
composition, the metabolic function of the microbiota also has a CDI
signature (20, 48, 63, 64) and can vary across mice from different
sources (65). For example, microbial metabolites, particularly secondary
bile acids and butyrate production, have been implicated as important
contributors to \emph{C. difficile} resistance (33, 47). Although, we
only looked at composition, \emph{Ruminococcaceae} and
\emph{Lachnospiraceae} both emerged as important taxa for classifying
day 7 \emph{C. difficile} colonization status and metagenomes from these
bacteria have been shown to contain the bile acid-inducible gene cluster
necessary for secondary bile acid formation and ability to produce
butyrate (52, 66). Interestingly, butyrate has previously been shown to
vary across vendors and mediates resistance to \emph{Citrobacter
rodentium} infection, a model of enterohemorrhagic and enteropathogenic
\emph{Escherichia coli} infections (23). Evidence for immunological
toning differences in IgA and Th17 cells across mice from different
vendors have also been documented and (67, 68), may also influence
response to CDI, particularly in the context of severe CDIs (69, 70).
The outcome after \emph{C. difficile} exposure depends on a multitude of
factors, including age, diet, and immunity; all of which are also
influenced by the microbiota. We have demonstrated that the ways
different baseline microbiotas from different mouse colony sources
respond to clindamycin treatment influences the length of time mice
remained colonized with \emph{C. difficile} 630. For those interested in
dissecting the contribution of the microbiome to \emph{C. difficile}
pathogenesis and treatments, using multiple sources of mice may yield
more insights than a single model alone. Furthermore, for studies
wanting to examine the interplay between a particular bacterial taxon
such as \emph{Enterococcus} and \emph{C. difficile}, these results could
serve as a resource for selecting which mice to order to address the
question.

\newpage

\subsection{Acknowledgements}\label{acknowledgements}

This work was supported by the National Institutes of Health
(U01AI124255). ST was supported by the Michigan Institute for Clincial
and Health Research Postdoctoral Translation Scholars Program
(UL1TR002240 and KL2TR002241). We thank members of the Schloss lab for
feedback on planning the experiments and data presentation, as well as
code tutorials and feedback through Code club. In particular, we want to
thank Begum Topçuoğlu for help with implementing logistic regression
models using her
\href{https://github.com/SchlossLab/ML_pipeline_microbiome}{pipeline}
and Ana Taylor for help with some media preparation and sample
collection. We also thank members of Vincent Young's lab, particularly
Kimberly Vendrov, for guidance with the \emph{C. difficile} infection
mouse model and donating the mice. We also want to thank the Unit for
Laboratory Animal Medicine at the University of Michigan for maintaining
our mouse colony and providing the institutional support for our mouse
experiments. Finally, we thank Kwi Kim, Austin Campbell, and Kimberly
Vendrov for their help in maintaining the Schloss lab's anaerobic
chamber.

\newpage

\subsection{Materials and Methods}\label{materials-and-methods}

\textbf{(i) Animals.} All experiments were approved by the University of
Michigan Animal Care and Use Committee (IACUC) under protocol number
PRO00006983. Female C57BL/7 mice were obtained from 6 different colony
sources: The Jackson Laboratory, Charles River Laboratories, Taconic
Biosciences, Envigo, and two colonies at the University of Michigan (the
Schloss lab colony and the Young lab colony). The Young lab colony was
originally established with mice purchased from Jackson, and the Schloss
lab colony was later founded with mice donated from the Young lab. The 4
groups of mice purchased from vendors were allowed to acclimate to the
University of Michigan mouse facility for 13 days prior to starting the
experiment. At least 4 female mice (age 5-10 weeks) were obtained per
source and mice from the same source were primarily housed at a density
of 2 mice per cage. The experiment was repeated once, approximately 3
months after the start of the first experiment.

\textbf{(ii) Antibiotic treatment.} After the 13-day acclimation period
and 1 day prior to challenge (Fig. 1A), all mice received 10 mg/kg
clindamycin (filter sterilized through a 0.22 micron syringe filter
prior to administration) via intraperitoneal injection.

\textbf{(iii) \emph{C. difficile} infection model.} Mice were challenged
with 10\textsuperscript{3} spores of \emph{C. difficile} strain 630 via
oral gavage post-infection 1 day after clindamycin treatment as
described previously (21). Mice weights and stool samples were taken
daily through 9 days post-challenge. Collected stool was split for
\emph{C. difficile} CFU quantification and 16S rRNA sequencing analysis.
\emph{C. difficile} quantification stool samples were transferred to the
anaerobic chamber, serially diluted in PBS, plated on
taurocholate-cycloserine-cefoxitin-fructose agar (TCCFA) plates, and
counted after 24 hours of incubation at 37°C under anaerobic conditions.
A sample from the day 0 timepoint (post-clindamycin and prior to
\emph{C. difficile} challenge) was also plated on TCCFA to ensure mice
were not already colonized with \emph{C. difficile} prior to infection.
There were 3 deaths recorded over the course of the experiment, 1
Taconic mouse died prior to \emph{C. difficile} challenge and 1 Jackson
and 1 Envigo mouse died between 1- and 3-days post-infection. Mice were
categorized as cleared when no \emph{C. difficile} was detected in the
first serial dilution (limit of detection: 100 CFU). Stool samples for
16S rRNA sequencing were snap frozen in liquid nitrogen and stored at
-80°C until DNA extraction.

\textbf{(iv) 16S rRNA sequencing.} DNA was extracted from -80°C stored
stool samples using the DNeasy Powersoil HTP 96 kit (Qiagen) and an
EpMotion 5075 automated pipetting system (Eppendorf). The V4 region was
amplified for 16S rRNA with the AccuPrime Pfx DNA polymerase (Thermo
Fisher Scientific) using custom barcoded primers, as previously
described (71). The ZymoBIOMICS microbial community DNA standards was
used as a mock community control (72) and water was used as a negative
control per 96-well extraction plate. The PCR amplicons were cleaned up
and normalized with the SequalPrep normalization plate kit (Thermo
Fisher Scientific). Amplicons were pooled and quantified with the KAPA
library quantification kit (KAPA biosystems), prior to sequencing using
the MiSeq system (Illumina).

\textbf{(v) 16S rRNA gene sequence analysis.} mothur (v. 1.43) was used
to process all sequences (73) with a previously published protocol (71).
Reads were combined and aligned with the SILVA reference database (74).
Chimeras were removed with the VSEARCH algorithm and taxonomic
assignment was completed with a modified version (v16) of the Ribosomal
Database Project reference database (v11.5) (75) with an 80\% cutoff.
Operational taxonomic units (OTUs) were assigned with a 97\% similarity
threshold using the opticlust algorithm (76). To account for uneven
sequencing across samples, samples were rarefied to 5,437 sequences
1,000 times for alpha and beta diversity analyses. PCoAs were generated
based on Theta YC distances. Permutational multivariate analysis of
variance (PERMANOVA) was performed on mothur-generated Theta YC distance
matrices with the adonis function in the vegan package (77) in R (78).

\textbf{(vi) Classification model training and evaluation.} Models were
generated based on mice that were categorized as either cleared or
colonized 7 days post-infection and had sequencing data from the
baseline (day -1), post-clindamycin (day 0), and post-infection (day 1)
timepoints of the experiment. Input bacterial community relative
abundance data at either the OTU or family level from the baseline,
post-clindamycin, and post-infection timepoints was used to generate 6
classification models that predicted \emph{C. difficile} colonization
status 7 days post-infection. The L2-regularized logistic regression
models were trained and tested using the caret package (79) in R as
previously described (80) with the exception that we used 60\% training
and 40\% testing data splits for the cross-validation of the training
data to select the best cost hyperparameter and the testing of the held
out test data to measure model performance. The modified
training/testing ratio was selected to accommodate the small number of
samples in the dataset. Code was modified from
\url{https://github.com/SchlossLab/ML_pipeline_microbiome} to update the
classification outcomes and change the data split ratios. The modified
repository to regenerate this analysis is available at
\url{https://github.com/tomkoset/ML_pipeline_microbiome}.

\textbf{(vii) Statistical analysis.} All statistical tests were
performed in R (v 3.5.2) (78). The Kruskal-Wallis test was used to
analyze differences in \emph{C. difficile} CFU, mouse weight change, and
alpha diversity across vendors with a Benjamini-Hochberg correction for
testing multiple timepoints, followed by pairwise Wilcoxon comparisons
with Benjamini-Hochberg correction. For taxonomic analysis and
generation of logistic regression model input data, \emph{C. difficile}
(OTU 20) was removed. Bacterial relative abundances that varied across
sources at the OTU and family taxonomic levels were identified with the
Kruskal-Wallis test with Benjamini-Hochberg correction for testing all
identified taxa at each level, followed by pairwise Wilcoxon comparisons
with Benjamini-Hochberg correction. Taxa impacted by clindamycin
treatment were identified using the Wilcoxon signed rank test with
matched pairs of mice samples for day -1 and day 0. To determine whether
classification models had better performance (test AUROCs) than random
chance (0.5), we used the one-sample Wilcoxon signed rank test. To
examine whether there was an overall difference in predictive
performance across the 6 classification models we used the
Kruskal-Wallis test followed by pairwise Wilcoxan comparisons with
Benjamini-Hochberg correction for multiple hypothesis testing. The
tidyverse package was used to wrangle and graph data (v 1.3.0) (81).

\textbf{(viii) Code availability.} Code for all data analysis and
generating this manuscript is available at
\url{https://github.com/SchlossLab/Tomkovich_vendor_difs_XXXX_2020}.

\textbf{(ix) Data availability.} The 16S rRNA sequencing data have been
deposited in the National Center for Biotechnology Information Sequence
Read Archive (BioProject Accession no. PRJNA608529).

\newpage

\subsection{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-Teng2019}{}
1. Teng C, Reveles KR, Obodozie-Ofoegbu OO, Frei CR. 2019. Clostridium
difficile infection risk with important antibiotic classes: An analysis
of the FDA adverse event reporting system. International Journal of
Medical Sciences 16:630--635.

\hypertarget{ref-Kelly2012}{}
2. Kelly C. 2012. Can we identify patients at high risk of recurrent
clostridium difficile infection? Clinical Microbiology and Infection
18:21--27.

\hypertarget{ref-Zacharioudakis2015}{}
3. Zacharioudakis IM, Zervou FN, Pliakos EE, Ziakas PD, Mylonakis E.
2015. Colonization with toxinogenic c. difficile upon hospital
admission, and risk of infection: A systematic review and meta-analysis.
American Journal of Gastroenterology 110:381--390.

\hypertarget{ref-Crobach2018}{}
4. Crobach MJT, Vernon JJ, Loo VG, Kong LY, Péchiné S, Wilcox MH,
Kuijper EJ. 2018. Understanding clostridium difficile colonization.
Clinical Microbiology Reviews 31.

\hypertarget{ref-Zhang2015}{}
5. Zhang L, Dong D, Jiang C, Li Z, Wang X, Peng Y. 2015. Insight into
alteration of gut microbiota in clostridium difficile infection and
asymptomatic c. difficile colonization. Anaerobe 34:1--7.

\hypertarget{ref-VanInsberghe2020}{}
6. VanInsberghe D, Elsherbini JA, Varian B, Poutahidis T, Erdman S, Polz
MF. 2020. Diarrhoeal events can trigger long-term clostridium difficile
colonization with recurrent blooms. Nature Microbiology 5:642--650.

\hypertarget{ref-Mancabelli2017}{}
7. Mancabelli L, Milani C, Lugli GA, Turroni F, Cocconi D, Sinderen D
van, Ventura M. 2017. Identification of universal gut microbial
biomarkers of common human intestinal diseases by meta-analysis. FEMS
Microbiology Ecology 93.

\hypertarget{ref-Duvallet2017}{}
8. Duvallet C, Gibbons SM, Gurry T, Irizarry RA, Alm EJ. 2017.
Meta-analysis of gut microbiome studies identifies disease-specific and
shared responses. Nature Communications 8.

\hypertarget{ref-Seekatz2016}{}
9. Seekatz AM, Rao K, Santhosh K, Young VB. 2016. Dynamics of the fecal
microbiome in patients with recurrent and nonrecurrent clostridium
difficile infection. Genome Medicine 8.

\hypertarget{ref-Khanna2016}{}
10. Khanna S, Montassier E, Schmidt B, Patel R, Knights D, Pardi DS,
Kashyap PC. 2016. Gut microbiome predictors of treatment response and
recurrence in primaryClostridium difficileinfection. Alimentary
Pharmacology \& Therapeutics 44:715--727.

\hypertarget{ref-Pakpour2017}{}
11. Pakpour S, Bhanvadia A, Zhu R, Amarnani A, Gibbons SM, Gurry T, Alm
EJ, Martello LA. 2017. Identifying predictive features of clostridium
difficile infection recurrence before, during, and after primary
antibiotic treatment. Microbiome 5.

\hypertarget{ref-Lee2020}{}
12. Lee AA, Rao K, Limsrivilai J, Gillilland M, Malamet B, Briggs E,
Young VB, Higgins PDR. 2020. Temporal gut microbial changes predict
recurrent clostridiodes difficile infection in patients with and without
ulcerative colitis. Inflammatory Bowel Diseases.

\hypertarget{ref-Hutton2014}{}
13. Hutton ML, Mackin KE, Chakravorty A, Lyras D. 2014. Small animal
models for the study ofClostridium difficiledisease pathogenesis. FEMS
Microbiology Letters 352:140--149.

\hypertarget{ref-Chen2008}{}
14. Chen X, Katchar K, Goldsmith JD, Nanthakumar N, Cheknis A, Gerding
DN, Kelly CP. 2008. A mouse model of clostridium difficileAssociated
disease. Gastroenterology 135:1984--1992.

\hypertarget{ref-Best2012}{}
15. Best EL, Freeman J, Wilcox MH. 2012. Models for the study of
clostridium difficile infection. Gut Microbes 3:145--167.

\hypertarget{ref-Baxter2014}{}
16. Baxter NT, Wan JJ, Schubert AM, Jenior ML, Myers P, Schloss PD.
2014. Intra- and interindividual variations mask interspecies variation
in the microbiota of sympatric peromyscus populations. Applied and
Environmental Microbiology 81:396--404.

\hypertarget{ref-Nagpal2018}{}
17. Nagpal R, Wang S, Woods LCS, Seshie O, Chung ST, Shively CA,
Register TC, Craft S, McClain DA, Yadav H. 2018. Comparative microbiome
signatures and short-chain fatty acids in mouse, rat, non-human primate,
and human feces. Frontiers in Microbiology 9.

\hypertarget{ref-Reeves2011}{}
18. Reeves AE, Theriot CM, Bergin IL, Huffnagle GB, Schloss PD, Young
VB. 2011. The interplay between microbiome dynamics and pathogen
dynamics in a murine model ofClostridium difficileInfection. Gut
Microbes 2:145--158.

\hypertarget{ref-Schubert2015}{}
19. Schubert AM, Sinani H, Schloss PD. 2015. Antibiotic-induced
alterations of the murine gut microbiota and subsequent effects on
colonization resistance against \emph{Clostridium difficile}. mBio 6.

\hypertarget{ref-Jenior2017}{}
20. Jenior ML, Leslie JL, Young VB, Schloss PD. 2017. \emph{Clostridium
difficile} colonizes alternative nutrient niches during infection across
distinct murine gut microbiomes. mSystems 2.

\hypertarget{ref-Jenior2018}{}
21. Jenior ML, Leslie JL, Young VB, Schloss PD. 2018. \emph{Clostridium
difficile} alters the structure and metabolism of distinct cecal
microbiomes during initial infection to promote sustained colonization.
mSphere 3.

\hypertarget{ref-Velazquez2019}{}
22. Velazquez EM, Nguyen H, Heasley KT, Saechao CH, Gil LM, Rogers AWL,
Miller BM, Rolston MR, Lopez CA, Litvak Y, Liou MJ, Faber F, Bronner DN,
Tiffany CR, Byndloss MX, Byndloss AJ, Bäumler AJ. 2019. Endogenous
enterobacteriaceae underlie variation in susceptibility to salmonella
infection. Nature Microbiology 4:1057--1064.

\hypertarget{ref-Osbelt2020}{}
23. Osbelt L, Thiemann S, Smit N, Lesker TR, Schröter M, Gálvez EJC,
Schmidt-Hohagen K, Pils MC, Mühlen S, Dersch P, Hiller K, Schlüter D,
Neumann-Schaal M, Strowig T. 2020. Variations in microbiota composition
of laboratory mice influence citrobacter rodentium infection via
variable short-chain fatty acid production. PLOS Pathogens 16:e1008448.

\hypertarget{ref-Stough2016}{}
24. Stough JMA, Dearth SP, Denny JE, LeCleir GR, Schmidt NW, Campagna
SR, Wilhelm SW. 2016. Functional characteristics of the gut microbiome
in c57bl/6 mice differentially susceptible to plasmodium yoelii.
Frontiers in Microbiology 7.

\hypertarget{ref-Alegre2019}{}
25. Alegre M-L. 2019. Mouse microbiomes: Overlooked culprits of
experimental variability. Genome Biology 20.

\hypertarget{ref-EtienneMesmin2017}{}
26. Etienne-Mesmin L, Chassaing B, Adekunle O, Mattei LM, Bushman FD,
Gewirtz AT. 2017. Toxin-positive clostridium difficile latently infect
mouse colonies and protect against highly pathogenic c. difficile. Gut
67:860--871.

\hypertarget{ref-Lai2020}{}
27. Lai NY, Musser MA, Pinho-Ribeiro FA, Baral P, Jacobson A, Ma P,
Potts DE, Chen Z, Paik D, Soualhi S, Yan Y, Misra A, Goldstein K,
Lagomarsino VN, Nordstrom A, Sivanathan KN, Wallrapp A, Kuchroo VK,
Nowarski R, Starnbach MN, Shi H, Surana NK, An D, Wu C, Huh JR, Rao M,
Chiu IM. 2020. Gut-innervating nociceptor neurons regulate peyer's patch
microfold cells and SFB levels to mediate salmonella host defense. Cell
180:33--49.e22.

\hypertarget{ref-Thiemann2017}{}
28. Thiemann S, Smit N, Roy U, Lesker TR, Gálvez EJ, Helmecke J, Basic
M, Bleich A, Goodman AL, Kalinke U, Flavell RA, Erhardt M, Strowig T.
2017. Enhancement of IFNgamma production by distinct commensals
ameliorates salmonella-induced disease. Cell Host \& Microbe
21:682--694.e5.

\hypertarget{ref-Rolig2013}{}
29. Rolig AS, Cech C, Ahler E, Carter JE, Ottemann KM. 2013. The degree
of helicobacter pylori-triggered inflammation is manipulated by
preinfection host microbiota. Infection and Immunity 81:1382--1389.

\hypertarget{ref-Ge2018}{}
30. Ge Z, Sheh A, Feng Y, Muthupalani S, Ge L, Wang C, Kurnick S,
Mannion A, Whary MT, Fox JG. 2018. Helicobacter pylori-infected c57bl/6
mice with different gastrointestinal microbiota have contrasting gastric
pathology, microbial and host immune responses. Scientific Reports 8.

\hypertarget{ref-Lawley2013}{}
31. Lawley TD, Young VB. 2013. Murine models to study clostridium
difficile infection and transmission. Anaerobe 24:94--97.

\hypertarget{ref-Buffie2011}{}
32. Buffie CG, Jarchum I, Equinda M, Lipuma L, Gobourne A, Viale A,
Ubeda C, Xavier J, Pamer EG. 2011. Profound alterations of intestinal
microbiota following a single dose of clindamycin results in sustained
susceptibility to clostridium difficile-induced colitis. Infection and
Immunity 80:62--73.

\hypertarget{ref-Buffie2014}{}
33. Buffie CG, Bucci V, Stein RR, McKenney PT, Ling L, Gobourne A, No D,
Liu H, Kinnebrew M, Viale A, Littmann E, Brink MRM van den, Jenq RR,
Taur Y, Sander C, Cross JR, Toussaint NC, Xavier JB, Pamer EG. 2014.
Precision microbiome reconstitution restores bile acid mediated
resistance to clostridium difficile. Nature 517:205--208.

\hypertarget{ref-Spinler2016}{}
34. Spinler JK, Brown A, Ross CL, Boonma P, Conner ME, Savidge TC. 2016.
Administration of probiotic kefir to mice with clostridium difficile
infection exacerbates disease. Anaerobe 40:54--57.

\hypertarget{ref-Markey2018}{}
35. Markey L, Shaban L, Green ER, Lemon KP, Mecsas J, Kumamoto CA. 2018.
Pre-colonization with the commensal fungus candida albicans reduces
murine susceptibility to clostridium difficile infection. Gut Microbes
1--13.

\hypertarget{ref-McKee2018}{}
36. McKee RW, Aleksanyan N, Garrett EM, Tamayo R. 2018. Type IV pili
promote clostridium difficile adherence and persistence in a mouse model
of infection. Infection and Immunity 86.

\hypertarget{ref-Yamaguchi2020}{}
37. Yamaguchi T, Konishi H, Aoki K, Ishii Y, Chono K, Tateda K. 2020.
The gut microbiome diversity of clostridioides difficile-inoculated mice
treated with vancomycin and fidaxomicin. Journal of Infection and
Chemotherapy 26:483--491.

\hypertarget{ref-Stroke2018}{}
38. Stroke IL, Letourneau JJ, Miller TE, Xu Y, Pechik I, Savoly DR, Ma
L, Sturzenbecker LJ, Sabalski J, Stein PD, Webb ML, Hilbert DW. 2018.
Treatment of clostridium difficile infection with a small-molecule
inhibitor of toxin UDP-glucose hydrolysis activity. Antimicrobial Agents
and Chemotherapy 62.

\hypertarget{ref-Quigley2019}{}
39. Quigley L, Coakley M, Alemayehu D, Rea MC, Casey PG, O'Sullivan,
Murphy E, Kiely B, Cotter PD, Hill C, Ross RP. 2019. Lactobacillus
gasseri APC 678 reduces shedding of the pathogen clostridium difficile
in a murine model. Frontiers in Microbiology 10.

\hypertarget{ref-Mullish2019}{}
40. Mullish BH, McDonald JAK, Pechlivanis A, Allegretti JR, Kao D,
Barker GF, Kapila D, Petrof EO, Joyce SA, Gahan CGM, Glegola-Madejska I,
Williams HRT, Holmes E, Clarke TB, Thursz MR, Marchesi JR. 2019.
Microbial bile salt hydrolases mediate the efficacy of faecal microbiota
transplant in the treatment of recurrent clostridioides difficile
infection. Gut 68:1791--1800.

\hypertarget{ref-Tomkovich2019}{}
41. Tomkovich S, Lesniak NA, Li Y, Bishop L, Fitzgerald MJ, Schloss PD.
2019. The proton pump inhibitor omeprazole does not promote
clostridioides difficile colonization in a murine model. mSphere 4.

\hypertarget{ref-Guh2018}{}
42. Guh AY, Kutty PK. 2018. Clostridioides difficileInfection. Annals of
Internal Medicine 169:ITC49.

\hypertarget{ref-Theriot2011}{}
43. Theriot CM, Koumpouras CC, Carlson PE, Bergin II, Aronoff DM, Young
VB. 2011. Cefoperazone-treated mice as an experimental platform to
assess differential virulence ofClostridium difficilestrains. Gut
Microbes 2:326--334.

\hypertarget{ref-Ross2016}{}
44. Ross CL, Spinler JK, Savidge TC. 2016. Structural and functional
changes within the gut microbiota and susceptibility to clostridium
difficile infection. Anaerobe 41:37--43.

\hypertarget{ref-Nguyen2015}{}
45. Nguyen TLA, Vieira-Silva S, Liston A, Raes J. 2015. How informative
is the mouse for human gut microbiota research? Disease Models \&
Mechanisms 8:1--16.

\hypertarget{ref-Lawley2009}{}
46. Lawley TD, Clare S, Walker AW, Goulding D, Stabler RA, Croucher N,
Mastroeni P, Scott P, Raisen C, Mottram L, Fairweather NF, Wren BW,
Parkhill J, Dougan G. 2009. Antibiotic treatment of clostridium
difficile carrier mice triggers a supershedder state, spore-mediated
transmission, and severe disease in immunocompromised hosts. Infection
and Immunity 77:3661--3669.

\hypertarget{ref-Lawley2012}{}
47. Lawley TD, Clare S, Walker AW, Stares MD, Connor TR, Raisen C,
Goulding D, Rad R, Schreiber F, Brandt C, Deakin LJ, Pickard DJ, Duncan
SH, Flint HJ, Clark TG, Parkhill J, Dougan G. 2012. Targeted restoration
of the intestinal microbiota with a simple, defined bacteriotherapy
resolves relapsing clostridium difficile disease in mice. PLoS Pathogens
8:e1002995.

\hypertarget{ref-Jump2014}{}
48. Jump RLP, Polinkovsky A, Hurless K, Sitzlar B, Eckart K, Tomas M,
Deshpande A, Nerandzic MM, Donskey CJ. 2014. Metabolomics analysis
identifies intestinal microbiota-derived biomarkers of colonization
resistance in clindamycin-treated mice. PLoS ONE 9:e101267.

\hypertarget{ref-NagaoKitamoto2020}{}
49. Nagao-Kitamoto H, Leslie JL, Kitamoto S, Jin C, Thomsson KA,
Gillilland MG, Kuffa P, Goto Y, Jenq RR, Ishii C, Hirayama A, Seekatz
AM, Martens EC, Eaton KA, Kao JY, Fukuda S, Higgins PDR, Karlsson NG,
Young VB, Kamada N. 2020. Interleukin-22-mediated host glycosylation
prevents clostridioides difficile infection by modulating the metabolic
activity of the gut microbiota. Nature Medicine 26:608--617.

\hypertarget{ref-Battaglioli2018}{}
50. Battaglioli EJ, Hale VL, Chen J, Jeraldo P, Ruiz-Mojica C, Schmidt
BA, Rekdal VM, Till LM, Huq L, Smits SA, Moor WJ, Jones-Hall Y, Smyrk T,
Khanna S, Pardi DS, Grover M, Patel R, Chia N, Nelson H, Sonnenburg JL,
Farrugia G, Kashyap PC. 2018. Clostridioides difficile uses amino acids
associated with gut microbial dysbiosis in a subset of patients with
diarrhea. Science Translational Medicine 10:eaam7019.

\hypertarget{ref-Robinson2014}{}
51. Robinson CD, Auchtung JM, Collins J, Britton RA. 2014. Epidemic
clostridium difficile strains demonstrate increased competitive fitness
compared to nonepidemic isolates. Infection and Immunity 82:2815--2825.

\hypertarget{ref-Collins2015}{}
52. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. 2015.
Humanized microbiota mice as a model of recurrent clostridium difficile
disease. Microbiome 3.

\hypertarget{ref-Collins2018}{}
53. Collins J, Robinson C, Danhof H, Knetsch CW, Leeuwen HC van, Lawley
TD, Auchtung JM, Britton RA. 2018. Dietary trehalose enhances virulence
of epidemic clostridium difficile. Nature 553:291--294.

\hypertarget{ref-Hryckowian2018}{}
54. Hryckowian AJ, Treuren WV, Smits SA, Davis NM, Gardner JO, Bouley
DM, Sonnenburg JL. 2018. Microbiota-accessible carbohydrates suppress
clostridium difficile infection in a murine model. Nature Microbiology
3:662--669.

\hypertarget{ref-Fouladi2020}{}
55. Fouladi F, Glenny EM, Bulik-Sullivan EC, Tsilimigras MCB, Sioda M,
Thomas SA, Wang Y, Djukic Z, Tang Q, Tarantino LM, Bulik CM, Fodor AA,
Carroll IM. 2020. Sequence variant analysis reveals poor correlations in
microbial taxonomic abundance between humans and mice after gnotobiotic
transfer. The ISME Journal.

\hypertarget{ref-Walter2020}{}
56. Walter J, Armet AM, Finlay BB, Shanahan F. 2020. Establishing or
exaggerating causality for the gut microbiome: Lessons from human
microbiota-associated rodents. Cell 180:221--232.

\hypertarget{ref-Rasmussen2019}{}
57. Rasmussen TS, Vries L de, Kot W, Hansen LH, Castro-Mejía JL,
Vogensen FK, Hansen AK, Nielsen DS. 2019. Mouse vendor influence on the
bacterial and viral gut composition exceeds the effect of diet. Viruses
11:435.

\hypertarget{ref-Mims2020}{}
58. Mims TS, Abdallah QA, Watts S, White C, Han J, Willis KA, Pierre JF.
2020. Variability in interkingdom gut microbiomes between different
commercial vendors shapes fat gain in response to diet. The FASEB
Journal 34:1--1.

\hypertarget{ref-Stewart2019}{}
59. Stewart DB, Wright JR, Fowler M, McLimans CJ, Tokarev V, Amaniera I,
Baker O, Wong H-T, Brabec J, Drucker R, Lamendella R. 2019. Integrated
meta-omics reveals a fungus-associated bacteriome and distinct
functional pathways in clostridioides difficile infection. mSphere 4.

\hypertarget{ref-Ott2017}{}
60. Ott SJ, Waetzig GH, Rehman A, Moltzau-Anderson J, Bharti R, Grasis
JA, Cassidy L, Tholey A, Fickenscher H, Seegert D, Rosenstiel P,
Schreiber S. 2017. Efficacy of sterile fecal filtrate transfer for
treating patients with clostridium difficile infection. Gastroenterology
152:799--811.e7.

\hypertarget{ref-Zuo2017}{}
61. Zuo T, Wong SH, Lam K, Lui R, Cheung K, Tang W, Ching JYL, Chan PKS,
Chan MCW, Wu JCY, Chan FKL, Yu J, Sung JJY, Ng SC. 2017. Bacteriophage
transfer during faecal microbiota transplantation inClostridium
difficile infection is associated with treatment outcome. Gut
gutjnl--2017--313952.

\hypertarget{ref-Zuo2018}{}
62. Zuo T, Wong SH, Cheung CP, Lam K, Lui R, Cheung K, Zhang F, Tang W,
Ching JYL, Wu JCY, Chan PKS, Sung JJY, Yu J, Chan FKL, Ng SC. 2018. Gut
fungal dysbiosis correlates with reduced efficacy of fecal microbiota
transplantation in clostridium difficile infection. Nature
Communications 9.

\hypertarget{ref-Robinson2019}{}
63. Robinson JI, Weir WH, Crowley JR, Hink T, Reske KA, Kwon JH, Burnham
C-AD, Dubberke ER, Mucha PJ, Henderson JP. 2019. Metabolomic networks
connect host-microbiome processes to human clostridioides difficile
infections. Journal of Clinical Investigation 129:3792--3806.

\hypertarget{ref-Fletcher2018}{}
64. Fletcher JR, Erwin S, Lanzas C, Theriot CM. 2018. Shifts in the gut
metabolome and clostridium difficile transcriptome throughout
colonization and infection in a mouse model. mSphere 3.

\hypertarget{ref-Xiao2015}{}
65. Xiao L, Feng Q, Liang S, Sonne SB, Xia Z, Qiu X, Li X, Long H, Zhang
J, Zhang D, Liu C, Fang Z, Chou J, Glanville J, Hao Q, Kotowska D,
Colding C, Licht TR, Wu D, Yu J, Sung JJY, Liang Q, Li J, Jia H, Lan Z,
Tremaroli V, Dworzynski P, Nielsen HB, Bäckhed F, Doré J, Chatelier EL,
Ehrlich SD, Lin JC, Arumugam M, Wang J, Madsen L, Kristiansen K. 2015. A
catalog of the mouse gut metagenome. Nature Biotechnology 33:1103--1108.

\hypertarget{ref-Vital2019}{}
66. Vital M, Rud T, Rath S, Pieper DH, Schlüter D. 2019. Diversity of
bacteria exhibiting bile acid-inducible 7alpha-dehydroxylation genes in
the human gut. Computational and Structural Biotechnology Journal
17:1016--1019.

\hypertarget{ref-Fransen2015}{}
67. Fransen F, Zagato E, Mazzini E, Fosso B, Manzari C, Aidy SE,
Chiavelli A, D'Erchia AM, Sethi MK, Pabst O, Marzano M, Moretti S,
Romani L, Penna G, Pesole G, Rescigno M. 2015. BALB/c and c57bl/6 mice
differ in polyreactive IgA abundance, which impacts the generation of
antigen-specific IgA and microbiota diversity. Immunity 43:527--540.

\hypertarget{ref-Ivanov2009}{}
68. Ivanov II, Atarashi K, Manel N, Brodie EL, Shima T, Karaoz U, Wei D,
Goldfarb KC, Santee CA, Lynch SV, Tanoue T, Imaoka A, Itoh K, Takeda K,
Umesaki Y, Honda K, Littman DR. 2009. Induction of intestinal th17 cells
by segmented filamentous bacteria. Cell 139:485--498.

\hypertarget{ref-Azrad2018}{}
69. Azrad M, Hamo Z, Tkhawkho L, Peretz A. 2018. Elevated serum
immunoglobulin a levels in patients with clostridium difficile infection
are associated with mortality. Pathogens and Disease 76.

\hypertarget{ref-Saleh2019}{}
70. Saleh MM, Frisbee AL, Leslie JL, Buonomo EL, Cowardin CA, Ma JZ,
Simpson ME, Scully KW, Abhyankar MM, Petri WA. 2019. Colitis-induced
th17 cells increase the risk for severe subsequent clostridium difficile
infection. Cell Host \& Microbe 25:756--765.e5.

\hypertarget{ref-Kozich2013}{}
71. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq illumina sequencing
platform. Applied and Environmental Microbiology 79:5112--5120.

\hypertarget{ref-Sze2019}{}
72. Sze MA, Schloss PD. 2019. The impact of DNA polymerase and number of
rounds of amplification in PCR on 16S rRNA gene sequence data. mSphere
4.

\hypertarget{ref-Schloss2009}{}
73. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
Thallinger GG, Horn DJV, Weber CF. 2009. Introducing mothur:
Open-source, platform-independent, community-supported software for
describing and comparing microbial communities. Applied and
Environmental Microbiology 75:7537--7541.

\hypertarget{ref-Quast2012}{}
74. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
J, Glöckner FO. 2012. The SILVA ribosomal RNA gene database project:
Improved data processing and web-based tools. Nucleic Acids Research
41:D590--D596.

\hypertarget{ref-Cole2013}{}
75. Cole JR, Wang Q, Fish JA, Chai B, McGarrell DM, Sun Y, Brown CT,
Porras-Alfaro A, Kuske CR, Tiedje JM. 2013. Ribosomal database project:
Data and tools for high throughput rRNA analysis. Nucleic Acids Research
42:D633--D642.

\hypertarget{ref-Westcott2017}{}
76. Westcott SL, Schloss PD. 2017. OptiClust, an improved method for
assigning amplicon-based sequence data to operational taxonomic units.
mSphere 2.

\hypertarget{ref-Vegan2018}{}
77. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
Minchin PR, O'Hara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
Wagner H. 2018. Vegan: Community ecology package.

\hypertarget{ref-r_citation_2018}{}
78. R Core Team. 2018. R: A language and environment for statistical
computing. R Foundation for Statistical Computing, Vienna, Austria.

\hypertarget{ref-Kuhn2008}{}
79. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
Journal of Statistical Software 28.

\hypertarget{ref-Topcuoglu2019}{}
80. Topçuoğlu BD, Lesniak NA, Ruffin M, Wiens J, Schloss PD. 2019. A
framework for effective application of machine learning to
microbiome-based classification problems.

\hypertarget{ref-Tidyverse2019}{}
81. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K,
Vaughan D, Wilke C, Woo K, Yutani H. 2019. Welcome to the tidyverse.
Journal of Open Source Software 4:1686.

\newpage

\subsection{Figures}\label{figures}

\includegraphics{figure_1.pdf} \textbf{Figure 1. Clindamycin is
sufficient to promote \emph{C. difficile} colonization regardless of
colony source, but clearance time varies across sources of C57BL/6
mice.} A. Setup of the experimental timeline. Mice for the experiments
were obtained from 6 different sources: the Schloss (N =8) and Young lab
(N = 9) colonies at the University of Michigan, the Jackson Laboratory
(N = 8), Charles River Laboratory (N = 8), Taconic Biosciences (N = 8),
and Envigo (N = 8). All mice were administered 10 mg/kg clindamycin
intraperitoneally (IP) 1 day before challenge with \emph{C. difficile}
630 spores on day 0. Mice were weighed and feces was collected daily
through the end of the experiment (9 days post-infection). Note: 3 mice
died during course of experiment. 1 Taconic mouse prior to infection and
1 Jackson and 1 Envigo mouse between 1- and 3-days post-infection. B.
\emph{C. difficile} CFU/gram stool measured over time (N = 20-49 mice
per timepoint) via serial dilutions. The black line represents the limit
of detection for the first serial dilution. CFU quantification data was
not available for each mouse due to early deaths, stool sampling
difficulties, and not plating all of the serial dilutions. C. Mouse
weight change measured in grams over time (N = 45-49 mice per
timepoint), all mice were normalized to the weight recorded 1 day before
infection. For B-C: timepoints where differences across sources of mice
were statistically significant by Kruskal-Wallis test with
Benjamini-Hochberg correction for testing across multiple days (Table S1
and Table S2) are reflected by the asterisk(s) above each timepoint (*,
P \textless{} 0.05, **, P \textless{} 0.01, ***, P \textless{} 0.001).
Lines represent the mean for each source and circles represent
individual mice. D. \emph{C. difficile} CFU status on Day 7 within each
mouse colony source. Mice were classified as colonized or cleared (not
detectable at the limit of detection of 100 CFU). Schloss (N = 6), Young
(N = 9), Jackson (N = 5), Charles River (N = 8), Taconic (N = 5), Envigo
(N = 7). Alternative: N = 5-9 mice per group.

\newpage

\includegraphics{figure_2.pdf} \textbf{Figure 2. Differences in
microbial richness and diversity across mouse colony sources emerge
after clindamycin treatment and infection.} A-B. Boxplots of the number
of observed OTUs and Shannon diversity index values at baseline: day -1
(A-B), after clindamycin: day 0 (C-D) and post-infection: day 1 (E-F)
timepoints of the experiment. Each circle represents the value for a
stool sample from an individual mouse. Data were analyzed by
Kruskal-Wallis test with Benjamini-Hochberg correction for testing each
day of the experiment and the adjusted \emph{P} value was \textless{}
0.05 for all panels except for B (Table S3). Significant \emph{P} values
from the pairwise Wilcoxon comparisons between sources with
Benjamini-Hochberg correction are shown (Table S4).

\newpage

\includegraphics{figure_3.pdf} \textbf{Figure 3. Mouse colony source is
the variable that explains most of the variation observed in the
baseline, post-clindamycin, and post-infection bacterial communities.}
A-C. Principal Coordinates Analysis of Theta YC distances from stools
collected at baseline (A), post-clindamycin (B), and post-infection (C)
timepoints of the experiment. Each circle represents a stool sample from
an individual mouse. PERMANOVA analysis demonstrated that source and the
interaction between source and cage explained most of the variation
observed in the baseline (combined R\textsuperscript{2} = 0.90),
post-clindamycin (combined R\textsuperscript{2} = 0.99), and
post-infection (combined R\textsuperscript{2} = 0.88) communities (all
\emph{P} = 0.0001, see Table S6).

\newpage

\includegraphics{figure_4.pdf} \textbf{Figure 4. A subset of bacterial
consistently vary across mouse colony sources despite clindamycin
perturbation and \emph{C. difficile} challenge.} A-C. Boxplots of the
relative abundances for the 12 OTUs that consistently varied across
sources of mice at the baseline (A), post-clindamycin (B), and
post-infection (C) timepoints of the experiment. D-F. Boxplots of the
relative abundances for the 8 families that consistently varied across
sources of mice at the baseline (D), post-clindamycin (E), and
post-infection (F) timepoints of the experiment. For each timepoint
bacteria with differential relative abundances across sources of mice
were identified by Kruskal-Wallis test at the family and OTU level with
Benjamini-Hochberg correction for testing all identified taxa at the
respective level (Table S8-9). The grey vertical line indicates the
limit of detection for A-F.

\newpage

\includegraphics{figure_5.pdf} \textbf{Figure 5. Clindamycin treatment
has the same effects on a subset of taxa regardless of colony source.}
A-B. Boxplots of the top 10 most significant (adjusted P value
\textless{} 0.05) OTUs with relative abundances that changed post
clindamycin treatment. C-D. Boxplots of the top 10 most significant
families with relative abundances that changed post clindamycin
treatment. Data were analyzed by Wilcoxon signed rank test limited to
mice that had paired sequence data for day -1 and 0 (N = 31). Tests were
performed at the OTU and family level with Benjamini-Hochberg correction
for testing all identified OTUs and families. See Table S10-11 for
complete list of OTUs and families significantly impacted by clindamycin
treatment. The grey vertical line indicates the limit of detection for
A-D.

\newpage

\includegraphics{figure_6.pdf} \textbf{Figure 6. Bacterial community
composition before, after clindamycin perturbation, and post-infection
can predict \emph{C. difficile} colonization status 7 days
post-challenge.} A-B. Logistic regression classification model AUROCs to
predict \emph{C. difficile} CFU on D7 (Fig. 1D, Fig. S3) based on the
community relative abundances at baseline (day -1), post-clindamycin
(day 0), and post-infection (day 1) at either the OTU (A) or family (B)
level. All models performed better than random chance (AUROC = 0.5), see
(all \emph{P} \(\le\) 5e-15; Table S12) and the model built with
post-clindamycin treated bacterial OTU relative abundances had the best
performance ((\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise
comparisons; Table S13). A List of the 20 taxa that were ranked as most
important to each model are listed in Table S14-15.

\newpage

\includegraphics{figure_7.pdf} \textbf{Figure 7. Key taxa that influence
\emph{C. difficile} colonization status vary by mouse colony source,
change after clindamycin treatment, or both.} A-B. Venn diagrams that
combine Fig. S4-6 summaries of taxa that were important to the day -1,
0, and 1 classification models (Table S14-15) and either overlapped with
taxa that varied across vendors at the same timepoint, were impacted by
clindamycin treatment, or both. See Fig. S4-6 for separate comparisons
of taxa from the day -1, 0, and 1 classification models. Bold taxa
signify bacteria that were important to more than 1 classification
model.

\newpage

\includegraphics{figure_8.pdf} \textbf{Figure 8: Key OTUs vary across
sources throughout the experiment.} A-C. Relative abundances of bold
OTUs from Fig. 7A that were important for at least two classification
models are shown over time. A. \emph{Enterococcus} (OTU 23), which
significantly varied across sources and was impacted by clindamycin
treatment. B. \emph{Bacteroides} (OTU 2), which varied across sources
throughout the experiment. C. \emph{Enterobacteriaceae} (OTU 1) and
\emph{Porphyromonadaceae} (OTU 7) were significantly impacted by
clindamycin treatment and examining relative abundance dynamics over the
course of the experiment indicated timepoints where relative abundances
also significantly varied across sources of mice. Each point represents
the relative abundance of an individual mouse stool sample and colored
lines represent the mean relative abundances for each source of mice.
The gray horizontal line represents the limit of detection. Timepoints
where differences across sources of mice were statistically significant
by Kruskal-Wallis test with Benjamini-Hochberg correction for testing
across multiple days (Table S16) are identified by the asterisk(s) above
each timepoint (*, P \textless{} 0.05, **, P \textless{} 0.01, ***, P
\textless{} 0.001).

\newpage

\includegraphics{figure_S1.pdf} \textbf{Figure S1. \emph{C. difficile}
CFU variation across vendors is similar between the 2 experiments.}
(A-B). \emph{C. difficile} CFU/gram of stool quantification over time
for experiment 1 (A) and 2 (B). Experiments were conducted approximately
3 months apart. Lines represent the mean CFU for each source and circles
represent individual mice and the black line represents the limit of
detection.

\newpage

\includegraphics{figure_S2.pdf} \textbf{Figure S2. Only bacterial
communities from University of Michigan mice significantly vary between
experiments.} A-F. PCoA of Theta YC distances for the baseline fecal
bacterial communities within each source of mice. Each symbol represents
a stool sample from an individual mouse with color corresponding to
experiment and shape representing cage mates. PERMANOVA was performed
within each group to examine the contributions of experiment and cage to
observed variation. Experiment number and cage only significantly
explained observed variation for mice from the Schloss (add r2 and P
values) and Young (add r2 and P values) lab colonies (Table S5).

\newpage

\includegraphics{figure_S3.pdf} \textbf{Figure S3. \emph{C. difficile}
colonization status 7 days post-infection varies across mice.} A. Bar
graph visualizations of overall day 7 classification outcomes used to
build logistic regression models. Mice were classified as colonized or
cleared (not detectable at the limit of detection of 100 CFU) based on
CFU g/stool data from 7 days post-infection. B-D. Classification input
data used for models based on day -1 (B), day 0 (C), and day 1 (D)
community relative abundance data. Only mice with day 7 \emph{C.
difficile} status and community relative abundance data for each
respective timepoint were used as input data for the classification
models.

\newpage

\includegraphics{figure_S4.pdf} \textbf{Figure S4. Key taxa from
classification model based on baseline community data vary by mouse
colony source, clindamycin treatment, or both.} A-B. Venn diagrams of
top 20 important OTUs (A) and families (B) from day -1 classification
models (Table S14-15) that overlapped with taxa that varied across
vendors at baseline, were impacted by clindamycin treatment, or both.
Bold taxa signify bacteria that were important to more than 1
classification model.

\newpage

\includegraphics{figure_S5.pdf} \textbf{Figure S5. Key taxa from
classification model based on post-clindamycin treatment community data
vary by mouse colony source, clindamycin treatment, or both.} A-B. Venn
diagrams of top 20 important OTUs (A) and families (B) from day 0
classification models (Table S14-15) that overlapped with taxa that
still varied across vendors after clindamycin, were impacted by
clindamycin treatment, or both. Bold taxa signify bacteria that were
important to more than 1 classification model.

\newpage

\includegraphics{figure_S6.pdf} \textbf{Figure S6. Key taxa from
classification model based on post-infection community data vary by
mouse colony source, clindamycin treatment, or both.} A-B. Venn diagrams
of top 20 important OTUs (A) and families (B) from day 1 classification
models (Table S14-15) that overlapped with taxa that varied across
vendors post-infection, were impacted by clindamycin treatment, or both.
Bold taxa signify bacteria that were important to more than 1
classification model.

\newpage

\includegraphics{figure_S7.pdf} \textbf{Figure S7. Key families vary
across sources throughout experiment.} Relative abundances of bold
families from Fig. 7B that were important for at least two
classification models are shown over time. A. \emph{Enterococcaceae} and
\emph{Lachnospiraceae}, which significantly varied across sources and
were impacted by clindamycin treatment. B. \emph{Bacteroidaceae} and
\emph{Deferribacteraceae}, which varied across sources throughout the
experiment. C. \emph{Bifidobacteriaceae}, \emph{Coriobacteriaceae},
\emph{Ruminococcaceae}, and \emph{Verrucomicrobiaceae} were
significantly impacted by clindamycin treatment. Examining the relative
abundance dynamics throughout the experiment, identified timepoints
where relative abundances also significantly varied across sources of
mice. Each point represents the relative abundance of an individual
mouse stool sample and colored lines represent the mean relative
abundances for each source of mice. The gray horizontal line represents
the limit of detection. Timepoints where differences across sources of
mice were statistically significant by Kruskal-Wallis test with
Benjamini-Hochberg correction for testing across multiple days (Table
S17) are identified by the asterisk(s) above each timepoint (*, P
\textless{} 0.05, **, P \textless{} 0.01, ***, P \textless{} 0.001).

\newpage

\subsection{Supplementary Tables and
Movie}\label{supplementary-tables-and-movie}

All supplemental material is available at:
\url{https://github.com/SchlossLab/Tomkovich_vendor_difs_XXXX_2020/submission}.

\textbf{Movie S1. Large shifts in bacterial community structure occurred
after clindamycin and \emph{C. difficile} infection.} PCoA of Theta YC
distances animated from 0 through 9 days post-infection. Transparency of
the circle corresponds to the day of the experiment, each circle
represents a sample from an individual mouse at a specific timepoint.
See Table S5 for PERMANOVA results).

\textbf{Table S1. \emph{C. difficile} CFU statistical results.}

\textbf{Table S2. Mouse weight change statistical results.}

\textbf{Table S3. Diversity metrics Kruskal-Wallis statistical results.}

\textbf{Table S4. Diversity metrics pairwise Wilcoxon statistical
results.}

\textbf{Table S5. PERMANOVA results for all mice, all timepoints.}

\textbf{Table S6. PERMANOVA results for all mice, all timepoints.}

\textbf{Table S7. PERMANOVA results of baseline communities within each
source.}

\textbf{Table S8. OTUs with relative abudances that significantly vary
across sources at baseline, post-clindamycin, or post-infection
timepoints.}

\textbf{Table S9. Families with relative abudances that significantly
vary across sources at baseline, post-clindamycin, or post-infection
timepoints.}

\textbf{Table S10. OTUs with relative abudances that significantly
changed after clindamycin treatment.}

\textbf{Table S11. Families with relative abudances that significantly
changed after clindamycin treatment.} \textbf{Table S12. Statistical
results of logistic regression model performances compared to random
chance.}

\textbf{Table S13. Pairwise Wilcoxan results of comparing all 6 logistic
regression model performances.}

\textbf{Table S14. Top 20 most important OTUs for each of the 3 logistic
regression models based on OTU relative abundance data.}

\textbf{Table S15. Top 20 most important families for each of the 3
logistic regression models based on OTU relative abundance data.}

\textbf{Table S16. OTUs with relative abudances that significantly
varied across sources of mice on at least 1 day of the experiment.}

\textbf{Table S17. Families with relative abudances that significantly
varied across sources of mice on at least 1 day of the experiment.}

\end{document}

### 2020-06-08 code/functions.R
Edit and remove old code, output processed_metadata to use for quantifying n for different analyses in the manuscript

@@ -73,8 +73,7 @@ metadata <- full_join(metadata, cfu_d7, by = "mouse_id") %>%
                                          cfu_d7 == NA ~ "NA"))
 
 #Create a column to denote baseline_weight and weight_change from baseline_weight for each mouse----
-#Percent baseline weight for each mouse will be calculated 
-#based on the weight recorded on D-1 of the experiment
+#Calculated based on the weight recorded on D-1 of the experiment
 baseline_weight <- metadata %>% select(mouse_id, weight, day) %>% 
   filter(day == -1) %>% 
   mutate(baseline_weight = weight) %>% 
@@ -122,7 +121,11 @@ dropped_by_miseq_ids <- anti_join(sequenced, matching, by = "id") %>% pull(id)
 color_scheme <- c("#1f78b4", "#e6ab02", "#d95f02", "#e7298a", "#7570b3", "#1b9e77") #Adapted from http://colorbrewer2.org/#type=qualitative&scheme=Dark2&n=6
 color_vendors <- levels(metadata$vendor)
 
-### Caluclate Standard Error----
+#Export processed metadata as a .tsv to read in to manuscript.Rmd for pulling out
+#sample sizes to add to figure legends.
+write_tsv(metadata, "data/process/processed_metadata.tsv")
+
+### Calculate Standard Error----
 calc_se <- function(stdev, n){
   stdev/sqrt(n)
 }

### 2020-06-08 code/taxa_stats.R

@@ -26,9 +26,8 @@ agg_taxa_data <- inner_join(otu_data, taxonomy)
 
 # Function to summarize relative abundance level for a given taxonomic level (ex. genus, family, phlyum, etc.)
 agg_taxonomic_data <- function(taxonomic_level) {
-  taxonomic_level <- enquo(taxonomic_level) #Part of transformation of taxonomic_level argument into a column name
   agg_taxa_data %>% 
-    group_by(id, !!taxonomic_level) %>% #!!Completes the transformation of taxonomic_level argument into a column name
+    group_by(id, {{ taxonomic_level }}) %>% #Embracing treats the taxonomic_level argument as a column name
     summarize(agg_rel_abund=sum(rel_abund)) %>% 
     # Merge relative abundance data to specifci taxonomic_level data
     inner_join(., metadata, by = "id") %>% 
@@ -62,8 +61,6 @@ agg_otu_data <- inner_join(agg_otu, taxa_info, by="key") %>%
   rename(otu=key) %>% 
   mutate(otu=paste0(gsub('TU0*', 'TU ', otu))) 
 
-# Relative abundance data at the genus level:
-agg_genus_data <- agg_taxonomic_data(genus)
 # Relative abundance data at the family level:
 agg_family_data <- agg_taxonomic_data(family)
 
@@ -156,13 +153,6 @@ Dn1toD1_families_d1 <- plot_families_dx(shared_sig_families_Dn1toD1, 1) +
   theme(plot.title = element_text(hjust = 0.5)) #Center plot titile
 save_plot("results/figures/Dn1toD1_families_d1.png", Dn1toD1_families_d1, base_height = 6, base_width = 8)
 
-#Plots of the relative abundances of families that significantly varied across sources of mice on day 7----
-#and were important in the 3 logistic regression models
-sig_families_d7_model_overlap <- intersect_all(`sig_family_day7`, `interp_families_combined`)
-sig_families_d7_overlap <- plot_families_dx(sig_families_d7_model_overlap, 7) +
-  ggtitle("Day 7 Model Overlap")+ #Title plot
-  theme(plot.title = element_text(hjust = 0.5)) #Center plot title
-
[…]

### 2020-06-08 code/weight_plot.R

@@ -1,6 +1,8 @@
 source("code/functions.R")
 
 #Graph weight data as weight change (g) from baseline weight.----
+weight_data <- metadata %>%
+  filter(!is.na(weight)) #512 observations that are not NAs
 
 #Range of N mice per day
 weight_data %>% group_by(day) %>% count() %>% arrange(n) 

### 2020-06-08 submission/manuscript.Rmd
Fix code chunk labels, add code to figure legends to indicate N for analyses

@@
 
-```{r, cfu}
+```{r cfu}
 cfu_stats_all <- read_tsv("../data/process/cfu_stats_all_days.tsv")
@@
 
-```{r, alpha}
+```{r alpha}
 richness_stats_all <- read_tsv("../data/process/richness_stats_all_days.tsv")
@@
 difficile* colonization across vendors.
-```{r, adonis}
+```{r adonis}
 adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
@@
 p_adonis_min <- adonis_stats_dn101 %>% select(p) %>% min #All p values are same
+
+source_adonis_dn1 <- read_tsv("../data/process/adonis_dn1_source.tsv")
+adonis_schloss_rsq_sum <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(r_sq) %>% sum
+p_adonis_schloss_max <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(p) %>% max
+adonis_young_rsq_sum <- source_adonis_dn1 %>% filter(source == "Young") %>% select(r_sq) %>% sum
+p_adonis_young_max <- source_adonis_dn1 %>% filter(source == "Young") %>% select(p) %>% max
 ```
@@
 difficile* challenge.
-```{r, source_taxa}
+```{r source_taxa}
 #Number of different otus and families across sources at day -1, 0, and 1 timepoints
@@
 
-```{r, source_clindamycin}
+```{r clindamycin_taxa}
 clindamycin_otus <- read_tsv("../data/process/otu_stats_dn1to0.tsv")
@@
 
-```{r, classification}
+```{r classification}
 models_vs_random <- read_tsv("../data/process/classification_to_random.tsv")
@@
 ## Figures
-```{r, sample_size}
+```{r sample_size}
 #figure out in functions.R and export results as a tsv to read in
-#analysis_n <- read_tsv("../data/process/____.tsv")
+analysis_n <- read_tsv("../data/process/processed_metadata.tsv")
 
+#Initial N for each source from the 2 experiments combined:
+schloss_i <- analysis_n %>% filter(day == -1 & vendor == "Schloss") %>% nrow()
+young_i <- analysis_n %>% filter(day == -1 & vendor == "Young") %>% nrow()
+jackson_i <- analysis_n %>% filter(day == -1 & vendor == "Jackson") %>% nrow()
+cr_i <- analysis_n %>% filter(day == -1 & vendor == "Charles River") %>% nrow()
+taconic_i <- analysis_n %>% filter(day == -1 & vendor == "Taconic") %>% nrow()
+envigo_i <- analysis_n %>% filter(day == -1 & vendor == "Envigo") %>% nrow()
+
+#Range in N for C. difficile CFU quantification:
+cfu_n_by_day <- analysis_n %>% 
+  filter(!is.na(cfu)) %>% #remove rows with NAs
+  group_by(day) %>% count() %>% arrange(n) %>% ungroup()
+min_cfu_n <- cfu_n_by_day %>% select(n) %>% min
+max_cfu_n <- cfu_n_by_day %>% select(n) %>% max
+
+#Range in N for mouse weight change data over the course of the experiment:
+weight_n_by_day <- analysis_n %>% 
+  filter(!is.na(weight)) %>% #remove rows with NAs
+  group_by(day) %>% count() %>% arrange(n) %>% ungroup()
+min_weight_n <- weight_n_by_day %>% select(n) %>% min
+max_weight_n <- weight_n_by_day %>% select(n) %>% max
+
+#C. difficile day 7 colonization status N range across sources:
+day7_status_n <- analysis_n %>% 
+  filter(!is.na(clearance_status_d7)) %>% #remove rows with NAs
+  filter(day == 7) %>% 
+  group_by(vendor) %>% count() %>% arrange(n) %>% ungroup
+day7_status_n_min <- day7_status_n %>% select(n) %>% min
+day7_status_n_max <-day7_status_n %>% select(n) %>% max
+
+#N for mice with sequencing data on both day -1 and day 0 (used to determine taxa that were altered by clindamycin treatment)
+n_paired_for_clind <- read_tsv("../data/process/vendors.diversity.summary") %>% #read in sequencing data & join to metadata
+  filter(method == "ave") %>% 
+  select(group, sobs, shannon, invsimpson, coverage) %>% 
+  inner_join(metadata, by = c("group" = "id")) %>% 
+  filter(day == -1 | day == 0) %>% 
+  filter(duplicated(mouse_id)) %>% 
+  nrow()
 ```
@@
 Setup of the experimental timeline.
-Mice for the experiments were obtained from 6 different sources: the Schloss (N =8) and Young lab (N = 9) colonies at the University of Michigan, the Jackson Laboratory (N = 8), Charles River Laboratory (N = 8), Taconic Biosciences (N = 8), and Envigo (N = 8).
+Mice for the experiments were obtained from 6 different sources: the Schloss (N = `r schloss_i`) and Young lab (N = `r young_i`) colonies at the University of Michigan, the Jackson Laboratory (N = `r jackson_i`), Charles River Laboratory (N = `r cr_i`), Taconic Biosciences (N = `r taconic_i`), and Envigo (N = `r envigo_i`).
 All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with *C.
@@
 *C.
-difficile* CFU/gram stool measured over time (N = 20-49 mice per timepoint) via serial dilutions.
+difficile* CFU/gram stool measured over time (N = `r min_cfu_n`-`r max_cfu_n` mice per timepoint) via serial dilutions.
 The black line represents the limit of detection for the first serial dilution.
-CFU quantification data was not available for each mouse due to early deaths, stool sampling difficulties, and not plating all of the serial dilutions.
+CFU quantification data was not available for each mouse due to early deaths, stool sampling difficulties, and not plating all of the seriral dilutions.
 C.
-Mouse weight change measured in grams over time (N = 45-49 mice per timepoint), all mice were normalized to the weight recorded 1 day before infection.
+Mouse weight change measured in grams over time (N = `r min_weight_n`-`r max_weight_n` mice per timepoint), all mice were normalized to the weight recorded 1 day before infection.
 For B-C: timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S1 and Table S2) are reflected by the asterisk(s) above each timepoint (\*\, P < 0.05, \*\*\, P < 0.01, \*\*\*\, P < 0.001).
@@
 Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU).
-Schloss (N = 6), Young (N = 9), Jackson (N = 5), Charles River (N = 8), Taconic (N = 5), Envigo (N = 7).
-Alternative: N = 5-9 mice per group.
+N = `r day7_status_n_min`-`r day7_status_n_max` mice per group.
 
@@
 Clindamycin treatment has the same effects on a subset of taxa regardless of colony source.** A-B.
-Boxplots of the top 10 most significant (adjusted P value < 0.05) OTUs with relative abundances that changed post clindamycin treatment.
+Boxplots of the top 10 most significant (adjusted *P* value < 0.05) OTUs with relative abundances that changed post clindamycin treatment.
 C-D.
 Boxplots of the top 10 most significant families with relative abundances that changed post clindamycin treatment.
-Data were analyzed by Wilcoxon signed rank test limited to mice that had paired sequence data for day -1 and 0 (N = 31).
-Tests were performed at the OTU and family level with Benjamini-Hochberg correction for testing all identified OTUs and families.
+Data were analyzed by Wilcoxon signed rank test limited to mice that had paired sequence data for day -1 and 0 (N = `r n_paired_for_clind`).
+Tests were performed at the OTU and family levels with Benjamini-Hochberg correction for testing all identified OTUs and families.
 See Table S10-11 for complete list of OTUs and families significantly impacted by clindamycin treatment.
@@
 PERMANOVA was performed within each group to examine the contributions of experiment and cage to observed variation.
-Experiment number and cage only significantly explained observed variation for mice from the Schloss (add r2 and P values) and Young (add r2 and P values) lab colonies (Table S5).
+Experiment number and cage only significantly explained observed variation for mice from the Schloss (combined R^2^ = `r adonis_schloss_rsq_sum`; *P* $\le$ `r p_adonis_schloss_max`) and Young (combined R^2^ = `r adonis_young_rsq_sum`; *P* $\le$ `r p_adonis_young_max`) lab colonies (Table S7).
 
@@
 difficile* colonization status 7 days post-infection varies across mice.** A.
-Bar graph visualizations of overall day 7 classification outcomes used to build logistic regression models.
+Bar graph visualizations of overall day 7 C.
+difficile colonization status that were used as classification outcomes to build logistic regression models.
 Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU) based on CFU g/stool data from 7 days post-infection.
@@
 difficile* infection.** PCoA of Theta YC distances animated from 0 through 9 days post-infection.
+PERMANOVA analysis indicated colony source was the variable that explained the most observed variation across fecal communities (source R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment.
 Transparency of the circle corresponds to the day of the experiment, each circle represents a sample from an individual mouse at a specific timepoint.

### 2020-06-08 code/cfu_plot.R
Remove unused code and figures

@@ -89,99 +89,6 @@ exp2_cfu <- summarize_plot(cfu_data_final %>% filter(experiment == 2))+
   theme(plot.title = element_text(hjust = 0.5)) #Center plot titile
 save_plot(filename = paste0("results/figures/exp2_cfu_time.png"), plot = exp2_cfu, base_aspect_ratio = 2)
 
-# Boxplots of C. diff CFU data at timepoints where there are significant differences in CFU levels across the different sources of mice:
-#Function to plot all significant genus relative abundances across vendors at a specific timepoint----
-#Arguments:
-# timepoint = timepoint to be analyzed
-plot_C.diff_timepoint <- function(timepoint){
-  plot_CFU_DX <- cfu_data_final %>% 
-    filter(day == timepoint) %>% 
-    ggplot(aes(x= vendor, y=cfu, color=vendor))+
-    scale_colour_manual(name=NULL,
-                        values=color_scheme,
-                        breaks=color_vendors,
-                        labels=color_vendors)+
-    geom_hline(yintercept = 100, linetype=2) +
-    geom_text(x = 11, y = 104, color = "black", label = "LOD")+
-    geom_boxplot(outlier.shape = NA, size = 1.2)+
-    geom_jitter(shape=19, size=2, alpha=0.6, position=position_jitterdodge(dodge.width=0.5, jitter.width=0.2)) +
-    labs(title=NULL, 
-         x=NULL,
-         y="CFU/g Feces")+
-    scale_y_log10(labels=fancy_scientific, breaks = c(10, 100, 10^3, 10^4, 10^5, 10^6, 10^7, 10^8, 10^9), limits=c(NA, 10^9))+
-    theme_classic()+
-    theme(plot.title=element_text(hjust=0.5))+
-    theme(legend.position = "none") + #Get rid of legend title & move legend position
-    theme(text = element_text(size = 16))  #Remove legend
-  save_plot(filename = paste0("results/figures/C.diff_CFU_D", timepoint,".png"), plot_CFU_DX, base_height = 11, base_width = 8.5, base_aspect_ratio = 2)
-}
-#Plot all the timepoints where C. diff CFUs were significantly different across sources of mice
-for(d in sig_cfu_days){
-  plot_C.diff_timepoint(d)
-}
-
[…]

### 2020-06-08 code/diversity.R

@@ -240,8 +240,6 @@ shannon_d1 <- shannon_dx_plot(1) +
   stat_pvalue_manual(data = pairwise_shannon_day1_plot, label = "p.adj", y.position = "y.position") 
 save_plot("results/figures/shannon_d1.png", shannon_d1)
 
-shannon_d7 <- shannon_dx_plot(7)
-
 #Function to make boxplots of sobs (richness) for all sources of mice on a specific experimental day----
 #Arguments: 
 #  timepoint = day of the experiment
@@ -306,7 +304,4 @@ sobs_d1 <- sobs_dx_plot(1) +
   ggtitle("Post-infection")+ #Title plot
   theme(plot.title = element_text(hjust = 0.5)) #Center plot titile
 save_plot("results/figures/richness_d1.png", sobs_d1) #Use save_plot instead of ggsave because it works better with cowplot
-
-  
-sobs_d7 <- sobs_dx_plot(7)
 

### 2020-06-08 code/pcoa_stats.R

@@ -120,7 +120,7 @@ all_axis2 <- all_axis_labels %>% filter(axis == 2) %>% pull(loading) %>% round(d
 
 #PCoA plot that combines the 2 experiments and save the plot----  
 pcoa_plot_combined <- plot_pcoa(pcoa_data)+
-  labs(x = paste("PCoA 1 (", all_axis1, "%)", sep = ""), #Anotations for each axis from loadings file
+  labs(x = paste("PCoA 1 (", all_axis1, "%)", sep = ""), #Annotations for each axis from loadings file
        y = paste("PCoA 2 (", all_axis2,"%)", sep = ""))
 save_plot(filename = paste0("results/figures/pcoa.png"), pcoa_plot_combined)
 

### 2020-06-08 code/weight_plot.R

@@ -36,9 +36,6 @@ exp1_weight <- summarize_plot(weight_data %>% filter(experiment == 1))
 #Percent weight change plot for the 2nd experiment----
 exp2_weight <- summarize_plot(weight_data %>% filter(experiment == 2))
 
-plot_grid(combined_exp_weight, exp1_weight, exp2_weight, labels = c("Combined Experiments", "Experiment 1", "Experiment 2"), ncol = 1, label_x = 0, label_y = 1)+
-  ggsave("exploratory/notebook/weight_changes.pdf", width = 8.5, height = 11)
-
 #Statistical analysis----
 #Kruskal_wallis test for differences across groups at different timepoints with Benjamini-Hochburg correction----
 set.seed(19881117) #Same seed used for mothur analysis
@@ -81,95 +78,6 @@ weight_stats_pairwise <- weight_kruskal_stats %>%
   unnest(model) %>% 
   select(-data, -parameter, -statistic) %>% 
   write_tsv("data/process/weight_stats_sig_days.tsv")
-
-#Format pairwise stats to use with ggpubr package
-plot_format_stats <- weight_stats_pairwise %>%
-  #Remove all columns except pairwise comparisons and day
-  select(-p.value, -method,-Schloss, -Young, -Jackson, -`Charles River`, -Taconic, -Envigo) %>% 
-  group_split() %>% #Keeps a attr(,"ptype") to track prototype of the splits
-  lapply(tidy_pairwise) %>% 
-  bind_rows()
-
-plot_format_stats %>% filter(day == 8 & p.adj <= 0.05) #7 significant
-plot_format_stats %>% filter(day == 7 & p.adj <= 0.05) #4 significant
-plot_format_stats %>% filter(day == 9 & p.adj <= 0.05) # 6 significant
-plot_format_stats %>% filter(day == 5 & p.adj <= 0.05) #5 significant
-plot_format_stats %>% filter(day == 1 & p.adj <= 0.05) #0 significant
-plot_format_stats %>% filter(day == 2 & p.adj <= 0.05) #0 significant
-plot_format_stats %>% filter(day == 3 & p.adj <= 0.05) #0 significant
-plot_format_stats %>% filter(day == 6 & p.adj <= 0.05) # 0 significant
-
-# Boxplots of weight_change data at timepoints where there were significant differences in CFU levels across the different sources of mice:
[…]

### 2020-06-08 submission/manuscript.Rmd
Fix broken dataframe reference

@@
 #N for mice with sequencing data on both day -1 and day 0 (used to determine taxa that were altered by clindamycin treatment)
-n_paired_for_clind <- read_tsv("../data/process/vendors.diversity.summary") %>% #read in sequencing data & join to metadata
+n_paired_for_clind <- read_tsv("../data/process/vendors.diversity.summary") %>% #read in sequencing data & join to analysis_n
   filter(method == "ave") %>% 
   select(group, sobs, shannon, invsimpson, coverage) %>% 
-  inner_join(metadata, by = c("group" = "id")) %>% 
+  inner_join(analysis_n, by = c("group" = "id")) %>% 
   filter(day == -1 | day == 0) %>% 

### 2020-06-08 submission/manuscript.tex
Updated version after modifying figure ledends

@@
 mice.} A. Setup of the experimental timeline. Mice for the experiments
-were obtained from 6 different sources: the Schloss (N =8) and Young lab
-(N = 9) colonies at the University of Michigan, the Jackson Laboratory
-(N = 8), Charles River Laboratory (N = 8), Taconic Biosciences (N = 8),
-and Envigo (N = 8). All mice were administered 10 mg/kg clindamycin
-intraperitoneally (IP) 1 day before challenge with \emph{C. difficile}
-630 spores on day 0. Mice were weighed and feces was collected daily
-through the end of the experiment (9 days post-infection). Note: 3 mice
-died during course of experiment. 1 Taconic mouse prior to infection and
-1 Jackson and 1 Envigo mouse between 1- and 3-days post-infection. B.
-\emph{C. difficile} CFU/gram stool measured over time (N = 20-49 mice
-per timepoint) via serial dilutions. The black line represents the limit
-of detection for the first serial dilution. CFU quantification data was
-not available for each mouse due to early deaths, stool sampling
-difficulties, and not plating all of the serial dilutions. C. Mouse
-weight change measured in grams over time (N = 45-49 mice per
-timepoint), all mice were normalized to the weight recorded 1 day before
-infection. For B-C: timepoints where differences across sources of mice
-were statistically significant by Kruskal-Wallis test with
-Benjamini-Hochberg correction for testing across multiple days (Table S1
-and Table S2) are reflected by the asterisk(s) above each timepoint (*,
-P \textless{} 0.05, **, P \textless{} 0.01, ***, P \textless{} 0.001).
-Lines represent the mean for each source and circles represent
-individual mice. D. \emph{C. difficile} CFU status on Day 7 within each
-mouse colony source. Mice were classified as colonized or cleared (not
-detectable at the limit of detection of 100 CFU). Schloss (N = 6), Young
-(N = 9), Jackson (N = 5), Charles River (N = 8), Taconic (N = 5), Envigo
-(N = 7). Alternative: N = 5-9 mice per group.
+were obtained from 6 different sources: the Schloss (N = 8) and Young
+lab (N = 9) colonies at the University of Michigan, the Jackson
+Laboratory (N = 8), Charles River Laboratory (N = 8), Taconic
+Biosciences (N = 8), and Envigo (N = 8). All mice were administered 10
+mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with
+\emph{C. difficile} 630 spores on day 0. Mice were weighed and feces was
+collected daily through the end of the experiment (9 days
+post-infection). Note: 3 mice died during course of experiment. 1
+Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse
+between 1- and 3-days post-infection. B. \emph{C. difficile} CFU/gram
+stool measured over time (N = 20-49 mice per timepoint) via serial
+dilutions. The black line represents the limit of detection for the
+first serial dilution. CFU quantification data was not available for
+each mouse due to early deaths, stool sampling difficulties, and not
+plating all of the seriral dilutions. C. Mouse weight change measured in
+grams over time (N = 45-49 mice per timepoint), all mice were normalized
+to the weight recorded 1 day before infection. For B-C: timepoints where
+differences across sources of mice were statistically significant by
+Kruskal-Wallis test with Benjamini-Hochberg correction for testing
+across multiple days (Table S1 and Table S2) are reflected by the
+asterisk(s) above each timepoint (*, P \textless{} 0.05, **, P
+\textless{} 0.01, ***, P \textless{} 0.001). Lines represent the mean
+for each source and circles represent individual mice. D. \emph{C.
+difficile} CFU status on Day 7 within each mouse colony source. Mice
+were classified as colonized or cleared (not detectable at the limit of
+detection of 100 CFU). N = 5-9 mice per group.
 
@@
 has the same effects on a subset of taxa regardless of colony source.}
-A-B. Boxplots of the top 10 most significant (adjusted P value
+A-B. Boxplots of the top 10 most significant (adjusted \emph{P} value
 \textless{} 0.05) OTUs with relative abundances that changed post
@@
 mice that had paired sequence data for day -1 and 0 (N = 31). Tests were
-performed at the OTU and family level with Benjamini-Hochberg correction
-for testing all identified OTUs and families. See Table S10-11 for
-complete list of OTUs and families significantly impacted by clindamycin
-treatment. The grey vertical line indicates the limit of detection for
-A-D.
+performed at the OTU and family levels with Benjamini-Hochberg
+correction for testing all identified OTUs and families. See Table
+S10-11 for complete list of OTUs and families significantly impacted by
+clindamycin treatment. The grey vertical line indicates the limit of
+detection for A-D.
 
@@
 observed variation. Experiment number and cage only significantly
-explained observed variation for mice from the Schloss (add r2 and P
-values) and Young (add r2 and P values) lab colonies (Table S5).
+explained observed variation for mice from the Schloss (combined
+R\textsuperscript{2} = 0.99; \emph{P} \(\le\) 0.033) and Young (combined
+R\textsuperscript{2} = 0.95; \emph{P} \(\le\) 0.029) lab colonies (Table
+S7).
 
@@
 colonization status 7 days post-infection varies across mice.} A. Bar
-graph visualizations of overall day 7 classification outcomes used to
-build logistic regression models. Mice were classified as colonized or
-cleared (not detectable at the limit of detection of 100 CFU) based on
-CFU g/stool data from 7 days post-infection. B-D. Classification input
-data used for models based on day -1 (B), day 0 (C), and day 1 (D)
-community relative abundance data. Only mice with day 7 \emph{C.
-difficile} status and community relative abundance data for each
-respective timepoint were used as input data for the classification
-models.
+graph visualizations of overall day 7 C. difficile colonization status
+that were used as classification outcomes to build logistic regression
+models. Mice were classified as colonized or cleared (not detectable at
+the limit of detection of 100 CFU) based on CFU g/stool data from 7 days
+post-infection. B-D. Classification input data used for models based on
+day -1 (B), day 0 (C), and day 1 (D) community relative abundance data.
+Only mice with day 7 \emph{C. difficile} status and community relative
+abundance data for each respective timepoint were used as input data for
+the classification models.
 
@@
 after clindamycin and \emph{C. difficile} infection.} PCoA of Theta YC
-distances animated from 0 through 9 days post-infection. Transparency of
-the circle corresponds to the day of the experiment, each circle
-represents a sample from an individual mouse at a specific timepoint.
-See Table S5 for PERMANOVA results).
+distances animated from 0 through 9 days post-infection. PERMANOVA
+analysis indicated colony source was the variable that explained the
+most observed variation across fecal communities (source
+R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by interactions
+between cage and day of the experiment. Transparency of the circle
+corresponds to the day of the experiment, each circle represents a
+sample from an individual mouse at a specific timepoint. See Table S5
+for PERMANOVA results).
 

### 2020-06-21 code/cfu_plot.R
Figure edits based on Nick's feedback, plot medians instead of median, use shapes to denote experiment or whether mice cleared by day 7, examine intra- and inter-group variation, update Readme to include new anaylsis script.

@@ -15,8 +15,8 @@ cfu_kruskal_stats <- cfu_data_final %>%
   group_by(day) %>% 
   nest() %>% 
   mutate(model=map(data, ~kruskal.test(x=.x$cfu, g=as.factor(.x$vendor)) %>%  tidy())) %>% 
-  mutate(mean = map(data, get_cfu_mean_vendor)) %>% 
-  unnest(c(model, mean)) %>% 
+  mutate(median = map(data, get_cfu_median_vendor)) %>% 
+  unnest(c(model, median)) %>% 
   ungroup() #Ungroup before adjusting p values
 
 #Adjust p-values for testing multiple days and write results to table
@@ -56,18 +56,22 @@ plot_format_stats <- cfu_stats_pairwise %>%
   lapply(tidy_pairwise) %>% 
   bind_rows()
 
-#Function to summarize data (calculate the mean for each group) and plot the data
+#Function to summarize data (calculate the median for each group) and plot the data
 summarize_plot <- function(df){
-  mean_summary <- df %>% 
+  median_summary <- df %>% 
     group_by(vendor, day) %>% 
-    summarize(mean_cfu = mean(cfu, na.rm = TRUE))
+    summarize(median_cfu = median(cfu, na.rm = TRUE))
   ggplot(NULL) +
-    geom_point(df, mapping = aes(x = day, y = cfu, color= vendor, fill = vendor), alpha = .2, size = .5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
-    geom_line(mean_summary, mapping = aes(x = day, y = mean_cfu, color = vendor), size = 1) +
+    geom_point(df, mapping = aes(x = day, y = cfu, color= vendor, fill = vendor, shape = experiment), alpha = .2, size = 1.5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
+    geom_line(median_summary, mapping = aes(x = day, y = median_cfu, color = vendor), size = 1.5) +
     scale_colour_manual(name=NULL,
                         values=color_scheme,
                         breaks=color_vendors,
                         labels=color_vendors)+
+    scale_shape_manual(name=NULL,
+                       values=shape_scheme,
+                       breaks=shape_experiment,
+                       labels=shape_experiment) +
     labs(x = "Days Post-Infection", y = "CFU/g Feces") +
     scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9),
                        limits = c(-0.5, 9.5)) +
[…]

### 2020-06-21 code/diversity.R

@@ -6,7 +6,7 @@ diversity_data <- read_tsv("data/process/vendors.diversity.summary") %>%
   inner_join(metadata, by = c("group" = "id")) %>% #Match only the samples we have sequence data for
   mutate(day=factor(day, levels=c("-1", "0", "1", "2", "3", "4", "5", "6", "7", "8", "9"))) # Treat day as a factor
 
-#Boxplots of diversity metrics (shannon and richness) for all sources of mice at all timepoints:
+#Plots of diversity metrics (shannon and richness) for all sources of mice at all timepoints:
 shannon_all <- diversity_data %>% 
   ggplot(aes(x=vendor, y = shannon, colour= vendor))+
   scale_colour_manual(name=NULL,
@@ -74,8 +74,8 @@ shannon_stats <- diversity_data %>%
     group_by(day) %>% 
     nest() %>% 
     mutate(model=map(data, ~kruskal.test(x=.x$shannon, g=as.factor(.x$vendor)) %>% tidy())) %>% 
-    mutate(mean = map(data, get_shannon_mean_vendor)) %>% 
-    unnest(c(model, mean)) %>% 
+    mutate(median = map(data, get_shannon_median_vendor)) %>% 
+    unnest(c(model, median)) %>% 
     ungroup() 
 
 #Adjust p-values for testing multiple days and write results to table
@@ -123,8 +123,8 @@ richness_stats <- diversity_data %>%
   group_by(day) %>% 
   nest() %>% 
   mutate(model=map(data, ~kruskal.test(x=.x$sobs, g=as.factor(.x$vendor)) %>% tidy())) %>% 
-  mutate(mean = map(data, get_sobs_mean_vendor)) %>% 
-  unnest(c(model, mean)) %>% 
+  mutate(median = map(data, get_sobs_median_vendor)) %>% 
+  unnest(c(model, median)) %>% 
   ungroup() 
 
 #Adjust p-values for testing multiple days and write results to table
@@ -182,20 +182,27 @@ plot_format_richness <- richness_stats_pairwise %>%
   lapply(tidy_pairwise) %>% 
   bind_rows()
 
-#Function to make boxplots of shannon for all sources of mice on a specific experimental day----
+#Function to make plots of shannon values for all sources of mice on a specific experimental day----
 #Arguments: 
 #  timepoint = day of the experiment
 shannon_dx_plot <- function(timepoint){
   diversity_data %>% 
     filter(day == timepoint) %>% 
+    group_by(vendor) %>% 
[…]

### 2020-06-21 code/figure_1.R

@@ -3,9 +3,11 @@ source("code/functions.R")
 a <- ggdraw() + draw_image("results/figures/Vendor_exp_scheme.png")
 b <- ggdraw() + draw_image("results/figures/cfu_over_time.png")
 c <- ggdraw() + draw_image("results/figures/weight_over_time.png")
-d <- ggdraw() + draw_image("results/figures/c.diff_status_d7.png")
+d <- ggdraw() + draw_image("results/figures/C.diff_CFU_D7_stats.png")
+e <- ggdraw() + draw_image("results/figures/weight_D2.png")
+bottom_row <- plot_grid(d, e, labels = c("D", "E"), label_size = 12)
 
-plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=1)+
+plot_grid(a, b, c, bottom_row, labels = c("A", "B", "C", ""), label_size = 12, ncol=1)+
   ggsave("results/figures/figure_1.pdf", width=3.875, height=8.5)+
   ggsave("submission/figure_1.pdf", width=3.875, height=8.5)
 

### 2020-06-21 code/figure_3.R

@@ -5,7 +5,7 @@ b <- ggdraw() + draw_image("results/figures/pcoa_day0.png")
 c <- ggdraw() + draw_image("results/figures/pcoa_day1.png")
 
 plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_3.pdf", width=3.4, height=8.5)+
-  ggsave("submission/figure_3.pdf", width=3.4, height=8.5)
+  ggsave("results/figures/figure_3.pdf", width=5, height=10)+
+  ggsave("submission/figure_3.pdf", width=5, height=10)
 
 

### 2020-06-21 code/figure_4.R

@@ -1,13 +1,21 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/Dn1toD1_otus_dn1.png")
-b <- ggdraw() + draw_image("results/figures/Dn1toD1_otus_d0.png")
-c <- ggdraw() + draw_image("results/figures/Dn1toD1_otus_d1.png")
-d <- ggdraw() + draw_image("results/figures/Dn1toD1_families_dn1.png")
-e <- ggdraw() + draw_image("results/figures/Dn1toD1_families_d0.png")
-f <- ggdraw() + draw_image("results/figures/Dn1toD1_families_d1.png")
+a1 <- ggdraw() + draw_image("results/figures/within_exp_dn1.png")
+a2 <- ggdraw() + draw_image("results/figures/between_exp_dn1.png")
+a3 <- ggdraw() + draw_image("results/figures/within_vendors_dn1.png")
+a4 <- ggdraw() + draw_image("results/figures/between_vendors_dn1.png")
+a <- plot_grid(a1, a2, a3, a4, labels = NULL, nrow = 1)
+b1 <- ggdraw() + draw_image("results/figures/within_exp_d0.png")
+b2 <- ggdraw() + draw_image("results/figures/between_exp_d0.png")
+b3 <- ggdraw() + draw_image("results/figures/within_vendors_d0.png")
+b4 <- ggdraw() + draw_image("results/figures/between_vendors_d0.png")
+b <- plot_grid(b1, b2, b3, b4, labels = NULL, nrow = 1)
+c1 <- ggdraw() + draw_image("results/figures/within_exp_d1.png")
+c2 <- ggdraw() + draw_image("results/figures/between_exp_d1.png")
+c3 <- ggdraw() + draw_image("results/figures/within_vendors_d1.png")
+c4 <- ggdraw() + draw_image("results/figures/between_vendors_d1.png")
+c <- plot_grid(c1, c2, c3, c4, labels = NULL, nrow = 1)
 
-
-plot_grid(a, b, c, d, e, f, labels = "AUTO", label_size = 12, ncol=3)+
-  ggsave("results/figures/figure_4.pdf", width=8, height=6)+
-  ggsave("submission/figure_4.pdf", width=8, height=6)
+plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_4.pdf", width=6, height=5)+
+  ggsave("submission/figure_4.pdf", width=6, height=5)

### 2020-06-21 code/figure_5.R

@@ -1,11 +1,14 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/clind_impacted_otus_plot_dn1.png")
-b <- ggdraw() + draw_image("results/figures/clind_impacted_otus_plot_d0.png")
-c <- ggdraw() + draw_image("results/figures/clind_impacted_families_plot_dn1.png")
-d <- ggdraw() + draw_image("results/figures/clind_impacted_families_plot_d0.png")
+a <- ggdraw() + draw_image("results/figures/Dn1toD1_otus_dn1.png")
+b <- ggdraw() + draw_image("results/figures/Dn1toD1_otus_d0.png")
+c <- ggdraw() + draw_image("results/figures/Dn1toD1_otus_d1.png")
+d <- ggdraw() + draw_image("results/figures/Dn1toD1_families_dn1.png")
+e <- ggdraw() + draw_image("results/figures/Dn1toD1_families_d0.png")
+f <- ggdraw() + draw_image("results/figures/Dn1toD1_families_d1.png")
 
-plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_5.pdf", width=8, height=11)+
-  ggsave("submission/figure_5.pdf", width=8, height=11)
+
+plot_grid(a, b, c, d, e, f, labels = "AUTO", label_size = 12, ncol=3)+
+  ggsave("results/figures/figure_5.pdf", width=8, height=6)+
+  ggsave("submission/figure_5.pdf", width=8, height=6)
 

### 2020-06-21 code/figure_6.R

@@ -1,8 +1,10 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/class._logistic_regression_60.png")
-b <- ggdraw() + draw_image("results/figures/class._logistic_regression_60_family.png")
+a <- ggdraw() + draw_image("results/figures/clind_impacted_otus_plot_dn1.png")
+b <- ggdraw() + draw_image("results/figures/clind_impacted_otus_plot_d0.png")
+c <- ggdraw() + draw_image("results/figures/clind_impacted_families_plot_dn1.png")
+d <- ggdraw() + draw_image("results/figures/clind_impacted_families_plot_d0.png")
 
-plot_grid(a, b, labels = "AUTO", label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_6.pdf", width=6, height=8)+
-  ggsave("submission/figure_6.pdf", width=6, height=8)
+plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_6.pdf", width=8, height=11)+
+  ggsave("submission/figure_6.pdf", width=8, height=11)

### 2020-06-21 code/figure_7.R

@@ -1,9 +1,17 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/venn_overall_otus.png")
-b <- ggdraw() + draw_image("results/figures/venn_overall_families.png")
+a1 <- ggdraw() + draw_image("results/figures/d0_model_otu_Enterobacteriaceae (OTU 1).png")
+a2 <- ggdraw() + draw_image("results/figures/d0_model_otu_Bacteroides (OTU 2).png")
+a3 <- ggdraw() + draw_image("results/figures/d0_model_otu_Proteus (OTU 16).png")
+a <- plot_grid(a1, a2, a3, labels = NULL, ncol = 1)
+b1 <- ggdraw() + draw_image("results/figures/d0_model_otu_Lactobacillus (OTU 18).png")
+b2 <- ggdraw() + draw_image("results/figures/d0_model_otu_Lachnospiraceae (OTU 9).png")
+b3 <- ggdraw() + draw_image("results/figures/d0_model_otu_Clostridium (OTU 99).png")
+b <- plot_grid(b1, b2, b3, labels = NULL, ncol = 1)
+top_half <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, ncol=2)
+c <- ggdraw() + draw_image("results/figures/venn_overall_otus.png")
 
-plot_grid(a, b, labels = "AUTO", label_size = 12, ncol=1)+
+plot_grid(top_half, c, labels = c("", "C"), label_size = 12, ncol=1)+
   ggsave("results/figures/figure_7.pdf", width=6, height=8)+
   ggsave("submission/figure_7.pdf", width=6, height=8)
 

### 2020-06-21 code/figure_S1.R

@@ -2,9 +2,10 @@ source("code/functions.R")
 
 a <- ggdraw() + draw_image("results/figures/exp1_cfu_time.png")
 b <- ggdraw() + draw_image("results/figures/exp2_cfu_time.png")
+c <- ggdraw() + draw_image("results/figures/C.diff_percent_colonized.png")
 
-plot_grid(a, b, labels = "AUTO", label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_S1.pdf", width=3.875, height=4.5)+
-  ggsave("submission/figure_S1.pdf", width=3.875, height=4.5)
+plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_S1.pdf", width=8, height=5.5)+
+  ggsave("submission/figure_S1.pdf", width=8, height=5.5)
 
 

### 2020-06-21 code/figure_S3.R

@@ -1,12 +1,11 @@
 source("code/functions.R")
 
 a <- ggdraw() + draw_image("results/figures/c.diff_status_d7_binary.png")
-b <- ggdraw() + draw_image("results/figures/d7_status_dn1_input.png")
-c <- ggdraw() + draw_image("results/figures/d7_status_d0_input.png")
-d <- ggdraw() + draw_image("results/figures/d7_status_d1_input.png")
-
+b <- ggdraw() + draw_image("results/figures/c.diff_status_d7.png")
+c <- ggdraw() + draw_image("results/figures/class._logistic_regression_60.png")
+d <- ggdraw() + draw_image("results/figures/class._logistic_regression_60_family.png")
 
 plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_S3.pdf", width=5, height=3.5)+
-  ggsave("submission/figure_S3.pdf", width=5, height=3.5)
+  ggsave("results/figures/figure_S3.pdf", width=7.5, height=5.5)+
+  ggsave("submission/figure_S3.pdf", width=7.5, height=5.5)
  

### 2020-06-21 code/figure_S4.R

@@ -1,8 +1,10 @@
 source("code/functions.R")
 
 a <- ggdraw() + draw_image("results/figures/venn_dn1_otus.png")
-b <- ggdraw() + draw_image("results/figures/venn_dn1_families.png")
+b <- ggdraw() + draw_image("results/figures/venn_d0_otus.png")
+c <- ggdraw() + draw_image("results/figures/venn_d1_otus.png")
 
-plot_grid(a, b, labels = "AUTO", label_size = 12, ncol=1)+
+
+plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=1)+
   ggsave("results/figures/figure_S4.pdf", width=6, height=8)+
   ggsave("submission/figure_S4.pdf", width=6, height=8)

### 2020-06-21 code/figure_S5.R

@@ -1,8 +1,10 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/venn_d0_otus.png")
+a <- ggdraw() + draw_image("results/figures/venn_dn1_families.png")
 b <- ggdraw() + draw_image("results/figures/venn_d0_families.png")
+c <- ggdraw() + draw_image("results/figures/venn_d1_families.png")
+d <- ggdraw() + draw_image("results/figures/venn_overall_families.png")
 
-plot_grid(a, b, labels = "AUTO", label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_S5.pdf", width=6, height=8)+
-  ggsave("submission/figure_S5.pdf", width=6, height=8)
+plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_S5.pdf", width=3, height=9.5)+
+  ggsave("submission/figure_S5.pdf", width=3, height=9.5)

### 2020-06-21 code/figure_S6.R

@@ -1,9 +1,25 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/venn_d1_otus.png")
-b <- ggdraw() + draw_image("results/figures/venn_d1_families.png")
+#Key families that vary with source and clindamycin treatment
+a <- ggdraw() + draw_image("results/figures/Enterococcaceae_time.png")
+b <- ggdraw() + draw_image("results/figures/Lachnospiraceae_time.png")
+A <- plot_grid(a, b, labels = c("", ""), label_size = 12, nrow=1)
 
-plot_grid(a, b, labels = "AUTO", label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_S6.pdf", width=6, height=8)+
-  ggsave("submission/figure_S6.pdf", width=6, height=8)
+#Key families that vary with source
+c <- ggdraw() + draw_image("results/figures/Bacteroidaceae_time.png")
+d <- ggdraw() + draw_image("results/figures/Deferribacteraceae_time.png")
+B <- plot_grid(c, d, labels = c("", ""), label_size = 12, nrow=1)
 
+#Key families that vary with clindamycin treatment
+e <- ggdraw() + draw_image("results/figures/Bifidobacteriaceae_time.png")
+f <- ggdraw() + draw_image("results/figures/Coriobacteriaceae_time.png")
+g <- ggdraw() + draw_image("results/figures/Ruminococcaceae_time.png")
+h <- ggdraw() + draw_image("results/figures/Verrucomicrobiaceae_time.png")
+C <- plot_grid(e, f, labels = c("", ""), label_size = 12, nrow=1)
+D <- plot_grid(g, h, labels = c("", ""), label_size = 12, nrow=1)
+
+#Combine all plots into one figure
+plot_grid(A, B, C, D, labels = c("A", "B", "C", ""), label_size = 12, nrow=4, align = "t")+
+  ggsave("results/figures/figure_S6.pdf", width=8, height=7)+
+  ggsave("submission/figure_S6.pdf", width=8, height=7)
+ 

### 2020-06-21 code/figure_S7.R

@@ -1,25 +1,0 @@
-source("code/functions.R")
-
-#Key families that vary with source and clindamycin treatment
-a <- ggdraw() + draw_image("results/figures/Enterococcaceae_time.png")
-b <- ggdraw() + draw_image("results/figures/Lachnospiraceae_time.png")
-A <- plot_grid(a, b, labels = c("", ""), label_size = 12, nrow=1)
-
-#Key families that vary with source
-c <- ggdraw() + draw_image("results/figures/Bacteroidaceae_time.png")
-d <- ggdraw() + draw_image("results/figures/Deferribacteraceae_time.png")
-B <- plot_grid(c, d, labels = c("", ""), label_size = 12, nrow=1)
-
-#Key families that vary with clindamycin treatment
-e <- ggdraw() + draw_image("results/figures/Bifidobacteriaceae_time.png")
-f <- ggdraw() + draw_image("results/figures/Coriobacteriaceae_time.png")
-g <- ggdraw() + draw_image("results/figures/Ruminococcaceae_time.png")
-h <- ggdraw() + draw_image("results/figures/Verrucomicrobiaceae_time.png")
-C <- plot_grid(e, f, labels = c("", ""), label_size = 12, nrow=1)
-D <- plot_grid(g, h, labels = c("", ""), label_size = 12, nrow=1)
-
-#Combine all plots into one figure
-plot_grid(A, B, C, D, labels = c("A", "B", "C", ""), label_size = 12, nrow=4, align = "t")+
-  ggsave("results/figures/figure_S7.pdf", width=8, height=7)+
-  ggsave("submission/figure_S7.pdf", width=8, height=7)
- 

### 2020-06-21 code/functions.R

@@ -49,21 +49,37 @@ cfu1_0s <- metadata %>% filter(cfu1 == 0) #Now only 184 instances, which is what
 cfu2_0s <- metadata %>% filter(cfu2 == 0) #0 instances of 0.
 cfu_nas_final <- map(metadata, ~sum(is.na(.))) #196 for cfu1, 367 instances for cfu2. #161 for cfu
 
-# Create columns for C. difficile CFU at Day 5 and Day 6 since these were the 2 timepoints where there were significant differences across vendors and the most significant pairwise.wilcox values (6 pairs)
-cfu_d5 <- metadata %>% 
-  filter(day == 5) %>% 
-  mutate(cfu_d5 = cfu) %>% 
-  select(mouse_id, cfu_d5) #43 values with 6 NAs
-cfu_d6 <- metadata %>% 
-  filter(day == 6) %>% 
-  mutate(cfu_d6 = cfu) %>% 
-  select(mouse_id, cfu_d6) #34 values with 15 NAs
-cfu_d7 <- metadata %>% 
-  filter(day == 7) %>% 
-  mutate(cfu_d7 = cfu) %>% 
-  select(mouse_id, cfu_d7) #40 values with 5 NAs. 17 mice with 0, 23 still colonized. Close to half have cleared. By D8 33 mice with 0s, 3 still colonized, 13 NAs
+# Function to create columns for C. difficile CFU at Days 1-9
+cfu_dx <- function(timepoint){
+  metadata %>% 
+    filter(day == timepoint) %>% 
+    mutate("col_name" = cfu) %>% 
+    select(mouse_id, "col_name")
+}
+cfu_d1 <- cfu_dx(1) %>% 
+  rename(cfu_d1 = col_name) 
+cfu_d2 <- cfu_dx(2) %>% 
+  rename(cfu_d2 = col_name) 
+cfu_d3 <- cfu_dx(3) %>% 
+  rename(cfu_d3 = col_name) 
+cfu_d4 <- cfu_dx(4) %>% 
+  rename(cfu_d4 = col_name) 
+cfu_d5 <- cfu_dx(5) %>% 
+  rename(cfu_d5 = col_name) #43 values with 6 NAs
+cfu_d6 <- cfu_dx(6) %>% 
+  rename(cfu_d6 = col_name) #34 values with 15 NAs
+cfu_d7 <- cfu_dx(7) %>% 
[…]

### 2020-06-21 code/pcoa_stats.R

@@ -94,7 +94,7 @@ tibble(effects = c("source", "day", "source:unique_cage", "source:run", "source:
 
 #Function to plot pcoa data for all vendors----
 plot_pcoa <- function(df){
-  ggplot(df, aes(x=axis1, y=axis2, color = vendor, alpha = day)) +
+  ggplot(df, aes(x=axis1, y=axis2, color = vendor, alpha = day, shape = experiment)) +
     geom_point(size=2) +
     scale_colour_manual(name=NULL,
                         values=color_scheme,
@@ -103,6 +103,10 @@ plot_pcoa <- function(df){
     scale_alpha_continuous(range = c(.3, 1),
                            breaks= c(2, 4, 6, 8, 10),
                            labels=c(2, 4, 6, 8, 10))+
+    scale_shape_manual(name="Experiment",
+                       values=shape_scheme,
+                       breaks=shape_experiment,
+                       labels=shape_experiment) +
     coord_fixed() + 
     xlim(-0.4, 0.65)+
     ylim(-0.45, 0.6)+
@@ -122,7 +126,7 @@ all_axis2 <- all_axis_labels %>% filter(axis == 2) %>% pull(loading) %>% round(d
 pcoa_plot_combined <- plot_pcoa(pcoa_data)+
   labs(x = paste("PCoA 1 (", all_axis1, "%)", sep = ""), #Annotations for each axis from loadings file
        y = paste("PCoA 2 (", all_axis2,"%)", sep = ""))
-save_plot(filename = paste0("results/figures/pcoa.png"), pcoa_plot_combined)
+save_plot(filename = paste0("results/figures/pcoa.png"), pcoa_plot_combined, base_height = 5, base_width = 4.5)
 
 #Animation of PCoA plot over time for all sequenced samples ----
 #Source: Will Close's Code Club from 4/12/2020 on plot animation
@@ -142,20 +146,23 @@ anim_save(animation = pcoa_gif, filename = 'results/pcoa_over_time.gif')
 
 #Function to plot pcoa data for all sources of mice at a specific timepoint----
 plot_pcoa <- function(df, timepoint){
-  plot <- ggplot(df, aes(x=axis1, y=axis2, color = vendor)) +
+  plot <- ggplot(df, aes(x=axis1, y=axis2, color = vendor, shape = experiment)) +
     geom_point(size=4, alpha = 0.4) +
     scale_colour_manual(name=NULL,
                         values=color_scheme,
[…]

### 2020-06-21 code/relative_distance_comparisons.R

@@ -1,0 +1,368 @@
+source("code/functions.R")
+
+read_dist_df <- function(dist_file_name){
+  linear_data <- scan(dist_file_name, what="character", quiet=TRUE)[-1]
+  
+  samples <- str_subset(linear_data, "D") #Pull out sample ids with E (denotes experiment number
+  n_samples <- length(samples)
+  distance_strings <- str_subset(linear_data, "\\.")
+  
+  distance_matrix <- matrix(0, nrow=n_samples, ncol=n_samples)
+  colnames(distance_matrix) <- samples
+  as.tibble(cbind(rows=samples, distance_matrix)) %>%
+    gather(columns, distances, -rows) %>%
+    filter(rows < columns) %>%
+    arrange(columns, rows) %>%
+    mutate(distances = as.numeric(distance_strings))
+}
+
+#Day -1 Inter and Intra-group variation----
+dn1_dist <- read_dist_df("data/mothur/d-1/vendors.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.opti_mcc.0.03.subsample.thetayc.0.03.lt.ave.dist") %>% 
+  #Relabel rows and columns by separating sample id into mouse and exp labels
+  separate(rows, into=c("row_mouse", "row_day", "row_exp"), sep = c(3, -2)) %>% #separate mouse ID from D and experiment number
+  separate(columns, into=c("col_mouse", "col_day", "col_exp"), sep = c(3, -2)) %>% #separate mouse ID from D and experiment number
+  select(-row_day, -col_day) %>% #get rid of day column, since all samples are from day -1 timepoint
+  mutate(experiment = recode(row_exp, E1 = 1, E2 = 2)) %>% #Match experiment column variable names from metadata
+  mutate(col_exp = recode(col_exp, E1 = 1, E2 = 2)) %>% #Match experiment column variable names from metadata
+  #Make 2 columns to designate row and column vendors
+  mutate(vendor = case_when(str_detect(row_mouse, "S") ~ "Schloss",
+                                str_detect(row_mouse, "Y") ~ "Young",
+                                str_detect(row_mouse, "J") ~ "Jackson",
+                                str_detect(row_mouse, "C") ~ "Charles River",
+                                str_detect(row_mouse, "T") ~ "Taconic",
+                                str_detect(row_mouse, "E") ~ "Envigo")
+         ) %>% 
[…]

### 2020-06-21 code/taxa_stats.R

@@ -87,8 +87,8 @@ kruskal_wallis_f <- function(timepoint){
   group_by(family) %>% 
   nest() %>% 
   mutate(model=map(data, ~kruskal.test(x=.x$agg_rel_abund, g=as.factor(.x$vendor)) %>% tidy())) %>% 
-  mutate(mean = map(data, get_rel_abund_mean_vendor)) %>% 
-  unnest(c(model, mean)) %>% 
+  mutate(median = map(data, get_rel_abund_median_vendor)) %>% 
+  unnest(c(model, median)) %>% 
   ungroup() 
   #Adjust p-values for testing multiple families
   family_stats_adjust <- family_stats %>% 
@@ -119,14 +119,22 @@ plot_families_dx <- function(families, timepoint){
   filter(day == timepoint) %>% 
   mutate(family=factor(family, levels=families)) %>% 
   mutate(agg_rel_abund = agg_rel_abund + 1/10874) %>% # 10,874 is 2 times the subsampling parameter of 5437
-  ggplot(aes(x= family, y=agg_rel_abund, color=vendor))+
+#  group_by(vendor, family) %>%   
+#  mutate(median = median(agg_rel_abund)) %>% #create a column of median_cfu
+#  ungroup() %>%      
+  ggplot(aes(x = family, y=agg_rel_abund, color=vendor))+
   scale_colour_manual(name=NULL,
                       values=color_scheme,
                       breaks=color_vendors,
                       labels=color_vendors)+
   geom_hline(yintercept=1/5437, color="gray")+
-  geom_boxplot(outlier.shape = NA, size = 1.2)+
-#  geom_jitter(shape=19, size=2, alpha=0.6, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.2)) +
+  geom_boxplot(outlier.shape = NA, size = 1.2)+  
+#  geom_errorbar(aes(ymax = median, ymin = median))+ #Add lines to indicate the median for each group to the plot. Median calculated before y axis transformation
+#  geom_jitter(aes(shape = experiment), size=2, alpha=0.6, position=position_dodge(1.6)) +
+#  scale_shape_manual(name=NULL,
+#                     values=shape_scheme,
+#                     breaks=shape_experiment,
+#                     labels=shape_experiment) +
   labs(title=NULL, 
        x=NULL,
[…]

### 2020-06-21 code/venn_diagram_comparisons.R

@@ -128,19 +128,19 @@ venn_families <- function(source_comp, clind_comp, title){
 }
 
 # Venn diagram of Day -1 overlapping families----
-dn1_families_venn_plot <- venn_families(dayn1_and_interp_dn1_f, paired_and_interp_dn1_f, "Day -1 model key taxa comparisons")
+dn1_families_venn_plot <- venn_families(dayn1_and_interp_dn1_f, paired_and_interp_dn1_f, "Day -1 model key family comparisons")
 save_plot("results/figures/venn_dn1_families.png", dn1_families_venn_plot, base_aspect_ratio = 1.8)
 
 # Venn diagram of Day 0 overlapping families----
-d0_families_venn_plot <- venn_families(day0_and_interp_d0_f, paired_and_interp_d0_f, "Day 0 model key taxa comparisons")
+d0_families_venn_plot <- venn_families(day0_and_interp_d0_f, paired_and_interp_d0_f, "Day 0 model key family comparisons")
 save_plot("results/figures/venn_d0_families.png", d0_families_venn_plot, base_aspect_ratio = 1.8)
 
 # Venn diagram of Day 1 overlapping families----
-d1_families_venn_plot <- venn_families(day1_and_interp_d1_f, paired_and_interp_d1_f, "Day 1 model key taxa comparisons")
+d1_families_venn_plot <- venn_families(day1_and_interp_d1_f, paired_and_interp_d1_f, "Day 1 model key family comparisons")
 save_plot("results/figures/venn_d1_families.png", d1_families_venn_plot, base_aspect_ratio = 1.8)
 
 # Venn diagram of overlapping families for all days combined----
-combined_families_venn_plot <- venn_families(dayn1to1_and_interp_combined_f, paired_and_interp_combined_f, "Key taxa comparisons for day -1, 0, and 1 models")
+combined_families_venn_plot <- venn_families(dayn1to1_and_interp_combined_f, paired_and_interp_combined_f, "Key family comparisons for day -1, 0, and 1 models")
 save_plot("results/figures/venn_overall_families.png", combined_families_venn_plot, base_aspect_ratio = 1.8)
 
 #OTUs identified in logistic regression classification models (20 OTUs with the highest ranking for each model)
@@ -201,7 +201,7 @@ key_otus <- unique(c(key_source_otus, key_clind_otus))
 #Arguments:
[…]

### 2020-06-21 code/weight_plot.R

@@ -9,18 +9,22 @@ weight_data %>% group_by(day) %>% count() %>% arrange(n)
 
 #Plot weight data as Weight change (g)----
 
-#Function to summarize weight change data (calculate the mean for each group) and plot the data
+#Function to summarize weight change data (calculate the median for each group) and plot the data
 summarize_plot <- function(df){
-  mean_summary <- df %>% 
+  median_summary <- df %>% 
     group_by(vendor, day) %>% 
-    summarize(mean_weight_change = mean(weight_change, na.rm = TRUE))
+    summarize(median_weight_change = median(weight_change, na.rm = TRUE))
   ggplot(NULL) +
-    geom_point(df, mapping = aes(x = day, y = weight_change, color= vendor, fill = vendor), alpha = .2, size = .5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
-    geom_line(mean_summary, mapping = aes(x = day, y = mean_weight_change, color = vendor), size = 1) +
+    geom_point(df, mapping = aes(x = day, y = weight_change, color= vendor, fill = vendor, shape = experiment), alpha = .2, size = .5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
+    geom_line(median_summary, mapping = aes(x = day, y = median_weight_change, color = vendor), size = 1) +
     scale_colour_manual(name=NULL,
                         values=color_scheme,
                         breaks=color_vendors,
                         labels=color_vendors)+
+    scale_shape_manual(name=NULL,
+                       values=shape_scheme,
+                       breaks=shape_experiment,
+                       labels=shape_experiment) +
     labs(x = "Days Post-Infection", y = "Weight Change (g)") +
     scale_x_continuous(breaks = c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9),
                        limits = c(-1.5, 9.5)) +
@@ -79,6 +83,14 @@ weight_stats_pairwise <- weight_kruskal_stats %>%
   select(-data, -parameter, -statistic) %>% 
   write_tsv("data/process/weight_stats_sig_days.tsv")
 
+#Format pairwise stats to use with ggpubr package
+plot_format_stats <- weight_stats_pairwise %>%
+  #Remove all columns except pairwise comparisons and day
+  select(-p.value, -method,-Schloss, -Young, -Jackson, -`Charles River`, -Taconic, -Envigo) %>% 
[…]

### 2020-06-22 code/figure_7.R
Additional figure edits

@@ -4,14 +4,10 @@ a1 <- ggdraw() + draw_image("results/figures/d0_model_otu_Enterobacteriaceae (OT
 a2 <- ggdraw() + draw_image("results/figures/d0_model_otu_Bacteroides (OTU 2).png")
 a3 <- ggdraw() + draw_image("results/figures/d0_model_otu_Proteus (OTU 16).png")
 a <- plot_grid(a1, a2, a3, labels = NULL, ncol = 1)
-b1 <- ggdraw() + draw_image("results/figures/d0_model_otu_Lactobacillus (OTU 18).png")
-b2 <- ggdraw() + draw_image("results/figures/d0_model_otu_Lachnospiraceae (OTU 9).png")
-b3 <- ggdraw() + draw_image("results/figures/d0_model_otu_Clostridium (OTU 99).png")
-b <- plot_grid(b1, b2, b3, labels = NULL, ncol = 1)
-top_half <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, ncol=2)
-c <- ggdraw() + draw_image("results/figures/venn_overall_otus.png")
+top_half <- plot_grid(a, labels = c("A"), label_size = 12, ncol=1)
+b <- ggdraw() + draw_image("results/figures/venn_overall_otus.png")
 
-plot_grid(top_half, c, labels = c("", "C"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_7.pdf", width=6, height=8)+
-  ggsave("submission/figure_7.pdf", width=6, height=8)
+plot_grid(top_half, b, labels = c("", "B"), label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_7.pdf", width=4, height=8)+
+  ggsave("submission/figure_7.pdf", width=4, height=8)
 

### 2020-06-22 code/relative_distance_comparisons.R

@@ -52,6 +52,7 @@ within_exp <- dn1_dist %>%
   geom_boxplot(outlier.shape = NA, size = 1.2)+
   geom_jitter(aes(shape = experiment), size=2, alpha=0.6, show.legend = FALSE) +
   scale_x_discrete(guide = guide_axis(n.dodge = 2))+
+  ylim(0, 1.25)+
   scale_colour_manual(name=NULL,
                       values=color_scheme,
                       breaks=color_vendors,
@@ -71,6 +72,7 @@ between_exp <- dn1_dist %>%
   geom_boxplot(outlier.shape = NA, size = 1.2)+
   geom_jitter(aes(shape = experiment), size=2, alpha=0.6, show.legend = FALSE) +
   scale_x_discrete(guide = guide_axis(n.dodge = 2))+
+  ylim(0, 1.25)+
   scale_colour_manual(name=NULL,
                       values=color_scheme,
                       breaks=color_vendors,
@@ -89,6 +91,7 @@ intra_vendors <- dn1_dist %>%
   geom_boxplot(outlier.shape = NA, size = 1.2)+
   geom_jitter(aes(shape = experiment), size=2, alpha=0.6, show.legend = FALSE) +
   scale_x_discrete(guide = guide_axis(n.dodge = 2))+
+  ylim(0, 1.25)+
   scale_colour_manual(name=NULL,
                       values=color_scheme,
                       breaks=color_vendors,
@@ -124,6 +127,7 @@ inter_all <- rbind(inter_schloss, inter_young, inter_jackson, inter_charles,
   ggplot(aes(x = vendor, y = distances, color = vendor)) +
   geom_boxplot(outlier.shape = NA, size = 1.2)+
   scale_x_discrete(guide = guide_axis(n.dodge = 2))+
+  ylim(0, 1.25)+
   scale_colour_manual(name=NULL,
                       values=color_scheme,
                       breaks=color_vendors,
@@ -169,6 +173,7 @@ within_exp_d0 <- d0_dist %>%
   geom_boxplot(outlier.shape = NA, size = 1.2)+
   geom_jitter(aes(shape = experiment), size=2, alpha=0.6, show.legend = FALSE) +
   scale_x_discrete(guide = guide_axis(n.dodge = 2))+
+  ylim(0, 1.25)+
   scale_colour_manual(name=NULL,
                       values=color_scheme,
                       breaks=color_vendors,
[…]

### 2020-06-22 submission/manuscript.Rmd
Edits after updating figures, Nick's review, and my review of manuscript

@@
 The microbiota plays a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs).
-However, much of the mechanistic work examining CDIs in mouse models use a single university colony or vendor, which have lower inter-individual microbiota variation compared to humans.
+However, much of the mechanistic work examining CDIs in mouse models use a single university colony or vendor.
 We treated mice from 6 different colony sources (2 University of Michigan colonies and 4 vendors) with a single clindamycin dose, followed by *C.
@@
 difficile* colonization levels through 9 days post-infection.
-The microbiota was profiled throughout the experiment via 16S rRNA gene sequencing analysis to examine variation across colony sources and alterations due to clindamycin treatment and *C.
+The microbiota was profiled via 16S rRNA gene sequencing analysis to examine variation across colony sources and alterations due to clindamycin treatment and *C.
 difficile* challenge.
@@
 difficile* colonization levels emerged from days 3-7 post-infection with 3 sources colonized with *C.
-difficile* for longer and at higher levels.
-We identified bacterial taxa with different relative abundances across colony sources throughout the experiment, as well as taxa that were consistently impacted by clindamycin treatment.
-We created bacterial community-based logistic regression models that successfully classified mice based on their day 7 *C.
-difficile* colonization status.
-Importantly, after examining the taxa that were most important to the classification models, we identified a subset of key taxa that varied across colony sources (*Bacteroides*, *Deferribacteraceae*), were altered by clindamycin (*Porphyromonadaceae*, *Ruminococcaceae*), or both (*Enterobacteriaceae*, *Enterococcus*, *Bifidobacteriaceae*, *Coriobacteriaceae*, *Lachnospiraceae*, and *Verrucomicrobiaceae*).
+difficile* for slightly longer and at higher levels.
+We identified bacterial taxa with different relative abundances across colony sources throughout the experiment, as well as taxa that were consistently impacted by clindamycin treatment in all sources of mice.
+We created logistic regression models that successfully classified mice based on whether they cleared *C.
+difficile* by 7 days post-infection using baseline, post-clindamycin, and post-infection community composition data.
+After examining the taxa that were most important to the classification models, we identified a subset of key taxa that varied across colony sources (*Bacteroides*, *Deferribacteraceae*), were altered by clindamycin (*Porphyromonadaceae*, *Ruminococcaceae*), or both (*Enterobacteriaceae*, *Enterococcus*, *Bifidobacteriaceae*, *Coriobacteriaceae*, *Lachnospiraceae*, and *Verrucomicrobiaceae*).
 These results suggest the response of the initial gut microbiota to clindamycin treatment influences *C.
@@
 published their C57BL/6 model in 2008, which disrupted the gut microbiota with antibiotics to enable *C.
-difficile* colonization, diarrhea, and weight loss [@Chen2008].
+difficile* colonization and symptoms such as diarrhea and weight loss [@Chen2008].
 CDI mouse models have been used to examine translationally relevant questions regarding *C.
@@
 padj_d1_cfu <- cfu_stats_all %>% filter(day == 1) %>% pull(p.value.adj) 
-mean_d1 <- cfu_stats_all %>% filter(day == 1) %>% select(-day, -statistic, -p.value, -parameter, -method, -p.value.adj) %>% pivot_longer(everything(), values_to = "mean_cfu", names_to = "Source") %>% arrange(mean_cfu)
-min_mean_d1 <- mean_d1 %>%  select(mean_cfu) %>% min
-max_mean_d1 <- mean_d1 %>% select(mean_cfu) %>% max
+median_d1 <- cfu_stats_all %>% filter(day == 1) %>% select(-day, -statistic, -p.value, -parameter, -method, -p.value.adj) %>% pivot_longer(everything(), values_to = "median_cfu", names_to = "Source") %>% arrange(median_cfu)
+min_median_d1 <- median_d1 %>%  select(median_cfu) %>% min
+max_median_d1 <- median_d1 %>% select(median_cfu) %>% max
 
@@
 max_padj_d3to7 <- cfu_stats_sig %>% select(p.value.adj) %>% max
+
+#Percent of mice colonized with C. difficile each day
+metadata <- read_tsv("../data/process/processed_metadata.tsv")
+#Add clearance_status_dx columns for the rest of the days in addition to d7
+metadata_cfu_status <- metadata %>% 
+  mutate(clearance_status_d7 = case_when(cfu_d7 > 0 ~ "colonized",
+                                         cfu_d7 == 0 ~ "cleared", #changed to match plot labeling used above
+                                         cfu_d7 == NA ~ "NA"),
+         clearance_status_d1 = case_when(cfu_d1 > 0 ~ "colonized",
+                                         cfu_d1 == 0 ~ "cleared", 
+                                         cfu_d1 == NA ~ "NA"),
+         clearance_status_d2 = case_when(cfu_d2 > 0 ~ "colonized",
+                                         cfu_d2 == 0 ~ "cleared", 
+                                         cfu_d2 == NA ~ "NA"),
+         clearance_status_d3 = case_when(cfu_d3 > 0 ~ "colonized",
+                                         cfu_d3 == 0 ~ "cleared", 
+                                         cfu_d3 == NA ~ "NA"),
+         clearance_status_d4 = case_when(cfu_d4 > 0 ~ "colonized",
+                                         cfu_d4 == 0 ~ "cleared", 
+                                         cfu_d4 == NA ~ "NA"),
+         clearance_status_d5 = case_when(cfu_d5 > 0 ~ "colonized",
+                                         cfu_d5 == 0 ~ "cleared", 
+                                         cfu_d5 == NA ~ "NA"),
+         clearance_status_d6 = case_when(cfu_d6 > 0 ~ "colonized",
+                                         cfu_d6 == 0 ~ "cleared", 
+                                         cfu_d6 == NA ~ "NA"),
+         clearance_status_d8 = case_when(cfu_d8 > 0 ~ "colonized",
+                                         cfu_d8 == 0 ~ "cleared", 
+                                         cfu_d8 == NA ~ "NA"),
+         clearance_status_d9 = case_when(cfu_d9 > 0 ~ "colonized",
+                                         cfu_d9 == 0 ~ "cleared", 
+                                         cfu_d9 == NA ~ "NA"))
+
+#Percent mice colonized each day, 2 experiments combined----
+count_colonized <- function(timepoint, clearance_status_dx){
+  metadata_cfu_status %>% 
+    filter(day == timepoint, !is.na(cfu)) %>% #get rid of NA values
+    count({{ clearance_status_dx }}) %>% 
+    mutate(day = timepoint) %>% 
+    rename(cfu_status = {{ clearance_status_dx }})
+}
+
+#Figure out number of mice that either cleared C. difficile or were still colonized with C. difficile on each day
+day_1 <- count_colonized(1, clearance_status_d1)
+day_2 <- count_colonized(2, clearance_status_d2)
+day_3 <- count_colonized(3, clearance_status_d3)
+day_4 <- count_colonized(4, clearance_status_d4)
+day_5 <- count_colonized(5, clearance_status_d5)
+day_6 <- count_colonized(6, clearance_status_d6)
+day_7 <- count_colonized(7, clearance_status_d7)
+day_8 <- count_colonized(8, clearance_status_d8)
+day_9 <- count_colonized(9, clearance_status_d9)
+
+#Combine into a data frame and figure out percent colonized per day
+percent_colonized_day <- rbind(day_1, day_2, day_3, day_4, day_5, day_6, day_7, day_8, day_9) %>% 
+  group_by(day) %>% 
+  mutate(total_n = sum(n)) %>% #Figure out total n per day (sum of cleared and colonized counts)
+  filter(cfu_status == "colonized") %>% #Only need colonized rows to figure out percent colonized
+  mutate(percent_colonized = (n/total_n) * 100)#Add new column for percent_colonized
+percent_colonized_d7 <- percent_colonized_day %>% filter(day == 7) %>% 
+  pull(percent_colonized)
+n_day1 <- percent_colonized_day %>% filter(day == 1) %>% 
+  pull(total_n)
+n_day2 <- percent_colonized_day %>% filter(day == 2) %>% 
+  pull(total_n)
+n_day3 <- percent_colonized_day %>% filter(day == 3) %>% 
+  pull(total_n)
+n_day4 <- percent_colonized_day %>% filter(day == 4) %>% 
+  pull(total_n)
+n_day5 <- percent_colonized_day %>% filter(day == 5) %>% 
+  pull(total_n)
+n_day6 <- percent_colonized_day %>% filter(day == 6) %>% 
+  pull(total_n)
+n_day7 <- percent_colonized_day %>% filter(day == 7) %>% 
+  pull(total_n)
+n_day8 <- percent_colonized_day %>% filter(day == 8) %>% 
+  pull(total_n)
+n_day9 <- percent_colonized_day %>% filter(day == 9) %>% 
+  pull(total_n)
 ```
@@
 The day after infection, *C.
-difficile* was detectable in all mice at a similar level (Mean CFU range: `r format(min_mean_d1, digits = 2, scientific=TRUE)`-`r format(max_mean_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of colony source (Fig.
+difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of colony source (Fig.
 1B).
@@
 difficile* clearance.
-We conducted two experiments approximately 3 months apart and combined the data because the colonization dynamics across sources of mice varied similarily in both experiments (Fig.
-S1).
+We conducted two experiments approximately 3 months apart and while the colonization dynamics were similar across most sources of mice, there was some variation between the 2 experiments, particularly for the Schloss and Envigo mice (Fig.
+S1A-B, Fig.
+1D).
 Although *C.
@@
 difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig.
-1C and Table S2).
-Interestingly, mice ordered from Jackson, Taconic, and Envigo tended to lose the most weight, have higher *C.
+1C, E and Table S2).
+Interestingly, mice ordered from Jackson, Taconic, and Envigo tended to lose more weight (although there was variation between experiments with Schloss and Envigo mice), have higher *C.
 difficile* CFU levels and take longer to clear the infection compared to the other sources of mice, which was particularly evident 7 days post-infection (Fig.
-1C and D).
-Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection.
+1B-E), when `r percent_colonized_d7` of the mice were still colonized with *C.
+difficile* (Fig.
+S1C).
+By 9 days post-infection the majority of the mice from all sources had cleared *C.
+difficile* (Fig.
+S1C) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment.
+Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment.
 Thus, clindamycin rendered all mice susceptible to *C.
@@
 ```
-**Bacterial communities vary across mouse colony sources.** Given the well known variation in mouse microbiomes across vendors and university colonies [@Alegre2019], we hypothesized that the variation in *C.
+**Bacterial communities consistently vary across mouse colony sources despite antibiotic and infection perturbations.** Given the well known variation in mouse microbiomes across vendors and university colonies [@Alegre2019], we hypothesized that the variation in *C.
 difficile* clearance could be explained by microbiota variation across the 6 sources.
@@
 We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig.
-S2 and Table S7), suggesting vendor communities were relatively stable between the 2 experiments.
+S2 and Table S7), although most of the vendors also clustered by experiment, suggesting there was some community variation between the 2 experiments within each vendor.
 Thus, mouse colony source was the factor that explained the most variation observed in the bacterial communities.
@@
 difficile* challenge.
+
+Since there was some variation in microbiota communities between experiments at baseline, we next looked at how similar the communities were within the same source and between sources in response to clindamycin treatment and *C.
+difficile* challenge (Fig.
+4) .
+The baseline communities varied most between experiments for Schloss, Young, and Envigo mice and variation between sources of mice was high (Fig.
+4A).
+Clindamycin treatment reduced variation between experiments within Schloss, Young and Jackson mice and some of the variation between sources diminished (Fig.
+4B).
+Post-infection, the community variation started to increase within sources of mice and variation between sources of mice started to return (Fig.
+4C).
+By using mice from multiple sources we were able to increase the number of microbiota communities we tested with the clindamycin *C.
+difficile* colonization mouse model.
 ```{r source_taxa}
@@
 We examined bacterial relative abundances at the operational taxonomic unit (OTU) and family levels, expecting the number of differences to be reduced at the family level due to the nature of bacterial taxonomy [@Nguyen2015].
-Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs and only `r source_families_dn1` families (Table S8-9) with relative abundances that varied across colony sources.
+Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs and `r source_families_dn1` families (Table S8-9) with relative abundances that varied across colony sources.
 Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs and `r source_families_d0` families (Table S8-9).
@@
 difficile* challenge, there were `r source_otus_d1` OTUs and `r source_families_d1` families (Table S8-9) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment.
-In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 4A-C) and `r shared_source_families_dn1to1` families with relative abundances that consistently varied across colony sources (Fig.
-4D-F).
+In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 5A-C) and `r shared_source_families_dn1to1` families with relative abundances that consistently varied across colony sources (Fig.
+5D-F).
 Importantly, some of the taxa that consistently varied across sources also shifted with clindamycin treatment.
@@
 By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs and `r clindamycin_families_sig` families that were altered after clindamycin treatment (Fig.
-5 and Table S10-11).
+6 and Table S10-11).
 Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) and 3 families (*Porphyromonadaceae*, *Enterococcaceae*, *Lachnospiraceae*) overlapped (Fig.
-4, Fig.
-5C-D).
+5, Fig.
+6C-D).
 These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all colony sources and only a subset of the taxa also varied across colony sources.
@@
 difficile* colonization status in mice.** After identifying taxa that varied by colony source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C.
-difficile* colonization status at day 7 (Fig.
-1D).
+difficile* colonization at day 7 (Fig.
+1D, Fig.
+S1C).
 We trained L2-regularized logistic regression models with input bacterial community data from the baseline, post-clindamycin, and post-infection timepoints of the experiment to predict *C.
 difficile* colonization status on day 7 (Fig.
-S3).
+S3A-B).
 All models were better at predicting *C.
 difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12), however the models trained with OTU level data generally performed better than those trained with family level data with the expception of the models based on the post-infection (day 1) communities (Fig.
-6).
+S3C-D).
 Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than all other models with an AUROC of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13).
-Thus, we were able to use community bacterial relative abundance data alone to differentiate mice that were still colonized with *C.
-difficile* on day 7 from those mice that had cleared the *C.
+Thus, we were able to use community bacterial relative abundance data alone to differentiate mice that had cleared *C.
+difficile* before day 7 from the mice still colonized with *C.
 difficile*.
@@
 Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 taxa that had the highest absolute feature weights in each of the 6 models (Table S14-15).
-The top 20 taxa from each model were compared to the list of taxa that varied across colony source (Fig.
-4 and Table S8-9) at the same timepoint and the taxa that were altered by clindamycin treatment (Fig.
-5 and Table S10-11).
+First, we looked at OTUs from the model with the best performance that was based on the post-clindamycin treatment bacterial community data.
+While most of the 20 OTUs had low relative abundances on day 0, *Enterobacteriace*, *Bacteroides* and *Proteus* had high relative abundances in at least one source of mice and significantly varied across sources (Fig.
+7A).
+Next, the top 20 taxa from each model were compared to the list of taxa that varied across colony source (Fig.
+5 and Table S8-9) at the same timepoint and the taxa that were altered by clindamycin treatment (Fig.
+6 and Table S10-11).
 We found a subset of OTUs and families that were important to the model and overlapped with bacteria that varied by either source, clindamycin treatment, or both (Fig.
-S4, S5, S6).
+S4, S5A-C).
 Combining the overall results for the 3 OTU models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig.
-7A).
+7B).
 Combining the overall results for the 3 family models identified `r venn_families_source` families associated with source, `r venn_families_clind` families associated with clindamycin treatment and `r venn_families_intersect` families associated with both (Fig.
-7B).
+S5D).
 Several OTUs (*`r list_key_otus`*) and families (*`r list_key_families`*) appeared across at least 2 models, so we examined how the relative abundances of these key taxa varied over the course of the experiment (Fig.
 8 and Fig.
-S7).
+S6).
 Throughout the experiment, there was at least 1 timepoint where relative abundances of these taxa significantly varied across sources (Table S16-17).
-Interestingly, there were no taxa that emerged as consistently enriched or depleted in Jackson, Taconic, and Envigo mice that were colonized for longer with *C.
+Interestingly, there were no taxa that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection with *C.
 difficile* 630, suggesting multiple bacteria influence the time needed to clear the infection.
@@
 8, Fig.
-S7) as candidate bacteria within these communities that were influencing variation in *C.
+S6) as candidate bacteria within these communities that were influencing variation in *C.
 difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by colony source, were impacted by clindamycin treatment, or both.
@@
 However, we did not identify any endogenous *C.
-difficile* strains prior to challenge, suggesting there were no endogenous protective strains in the mice we received and other bacterial taxa were mediating the variation in *C.
+difficile* strains prior to challenge, suggesting there were no endogenous protective strains in the mice we received and other bacterial taxa mediated the variation in *C.
 difficile* colonization across sources.
 Although all mice were susceptible to *C.
-difficile* colonization, by following colonization over time we found Jackson, Taconic, and Envigo mice remained colonized for longer.
+difficile* colonization, by following colonization over time we found Jackson, Taconic, and Envigo mice remained colonized beyond 7 days post-infection.
 We identified a subset of bacteria that were important in predicting whether a mouse was still colonized with *C.
-difficile* status 7 days post-infection.
+difficile* 7 days post-infection.
 These results suggest a subset of the bacterial community is responsible for determining the length of time needed to clear *C.
@@
 difficile* colonization status 7 days post-infection.
+Since all but 3 mice eventually cleared *C.
+difficile* 630 by 9 days post-infection and the model built with the post-clindamycin OTU relative abundance data had the best performance, our results suggest clindamycin treatment had a large role in determining *C.
+difficile* susceptibility and clearance in the mice.
 
@@
 difficile* model.
-One alternative approach that has been used in some CDI studies [@NagaoKitamoto2020; @Battaglioli2018; @Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018] is to use mice associated with human microbiotas.
+One alternative approach that has been used in some CDI studies [@NagaoKitamoto2020; @Battaglioli2018; @Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018] is to associate mice with human microbiotas.
 However, a major caveat to this method is the substantial loss of human microbiota community members upon transfer to mice [@Fouladi2020; @Walter2020].
 Additionally with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of the CDI mouse model studies to date associated mice with just 1 types of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018], which does not aid in the goal of figuring out how a variety of unique microbiotas influence susceptiblity to CDIs and adverse outcomes.
-Encouragingly, decreased *Bifidobacterium*, *Porphyromonas*, *Ruminococcaceae* and *Lachnospiraceae* and increased *Enterobacteriaceae*, *Enterococcus*, *Lactobacillus*, and *Proteus* have all been associated with human CDIs [@Mancabelli2017, @Duvallet2017] and were well represented in our study, suggesting most of the vendors are suitable for gaining insights into microbiota associated factors influencing *C.
+Encouragingly, decreased *Bifidobacterium*, *Porphyromonas*, *Ruminococcaceae* and *Lachnospiraceae* and increased *Enterobacteriaceae*, *Enterococcus*, *Lactobacillus*, and *Proteus* have all been associated with human CDIs [@Mancabelli2017, @Duvallet2017] and were well represented in our study, suggesting most of the mouse sources are suitable for gaining insights into microbiota associated factors influencing *C.
 difficile* colonization and infections in humans.
@@
 Interestingly, butyrate has previously been shown to vary across vendors and mediates resistance to *Citrobacter rodentium* infection, a model of enterohemorrhagic and enteropathogenic *Escherichia coli* infections [@Osbelt2020].
-Evidence for immunological toning differences in IgA and Th17 cells across mice from different vendors have also been documented and [@Fransen2015; @Ivanov2009], may also influence response to CDI, particularly in the context of severe CDIs [@Azrad2018; @Saleh2019].
+Evidence for immunological toning differences in IgA and Th17 cells across mice from different vendors have also been documented and [@Fransen2015; @Ivanov2009] may also influence response to CDI, particularly in the context of severe CDIs [@Azrad2018; @Saleh2019].
 The outcome after *C.
@@
 This work was supported by the National Institutes of Health (U01AI124255).
-ST was supported by the Michigan Institute for Clincial and Health Research Postdoctoral Translation Scholars Program (UL1TR002240 and KL2TR002241).
+ST was supported by the Michigan Institute for Clincial and Health Research Postdoctoral Translation Scholars Program (UL1TR002240).
 We thank members of the Schloss lab for feedback on planning the experiments and data presentation, as well as code tutorials and feedback through Code club.
-In particular, we want to thank Begum Topçuoğlu for help with implementing logistic regression models using her [pipeline](https://github.com/SchlossLab/ML_pipeline_microbiome) and Ana Taylor for help with some media preparation and sample collection.
+In particular, we want to thank Begüm Topçuoğlu for help with implementing L2 logistic regression models using her [pipeline](https://github.com/SchlossLab/ML_pipeline_microbiome), Ana Taylor for help with some media preparation and sample collection, and Nicholas Lesniak for his critical feedback on the manuscript.
 We also thank members of Vincent Young’s lab, particularly  Kimberly Vendrov, for guidance with the *C.
@@
 We also want to thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the institutional support for our mouse experiments.
-Finally, we thank Kwi Kim, Austin Campbell, and Kimberly Vendrov for their help in maintaining the Schloss lab’s anaerobic chamber.
+Finally, we thank Kwi Kim, Austin Campbell, and Kimberly Vendrov for their help in maintaining the Schloss lab's anaerobic chamber.
 
@@
 difficile* colonization status 7 days post-infection.
-The L2-regularized logistic regression models were trained and tested using the caret package [@Kuhn2008] in R as previously described [@Topcuoglu2019] with the exception that we used 60% training and 40% testing data splits for the cross-validation of the training data to select the best cost hyperparameter and the testing of the held out test data to measure model performance.
+The L2-regularized logistic regression models were trained and tested using the caret package [@Kuhn2008] in R as previously described [@Topcuoglu2020] with the exception that we used 60% training and 40% testing data splits for the cross-validation of the training data to select the best cost hyperparameter and the testing of the held out test data to measure model performance.
 The modified training/testing ratio was selected to accommodate the small number of samples in the dataset.
@@
 Clindamycin is sufficient to promote *C.
-difficile* colonization regardless of colony source, but clearance time varies across sources of C57BL/6 mice.** A.
+difficile* colonization in all mice, but clearance time varies across sources of C57BL/6 mice.** A.
 Setup of the experimental timeline.
@@
 The black line represents the limit of detection for the first serial dilution.
-CFU quantification data was not available for each mouse due to early deaths, stool sampling difficulties, and not plating all of the seriral dilutions.
+CFU quantification data was not available for each mouse due to early deaths, stool sampling difficulties, and not plating all of the serial dilutions.
 C.
 Mouse weight change measured in grams over time (N = `r min_weight_n`-`r max_weight_n` mice per timepoint), all mice were normalized to the weight recorded 1 day before infection.
-For B-C: timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S1 and Table S2) are reflected by the asterisk(s) above each timepoint (\*\, P < 0.05, \*\*\, P < 0.01, \*\*\*\, P < 0.001).
-Lines represent the mean for each source and circles represent individual mice.
+For B-C: timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S1 and Table S2) are reflected by the asterisk(s) above each timepoint (\*\, *P* < 0.05).
+Lines represent the median for each source and circles represent individual mice from experiment 1 while triangles represent mice from experiment 2.
 D.
 *C.
-difficile* CFU status on Day 7 within each mouse colony source.
-Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU).
-N = `r day7_status_n_min`-`r day7_status_n_max` mice per group.
+difficile* CFU/gram stool on day 7 post-infection across sources of mice with asterisks for pairwise Wilcoxon comparisons with Benjamini-Hochberg correction where *P* < 0.05.
+E.
+Mouse weight change 2 days post-infection across sources of mice, no pairwise Wilcoxon comparisons were significant after Benjamini-Hochberg correction.
+For D-E.
+Circles represent experiment 1 mice, triangles represent experiment 2 mice and gray lines indicate the median values for each group.
 
@@
 **Figure 2.
-Differences in microbial richness and diversity across mouse colony sources emerge after clindamycin treatment and infection.** A-B.
-Boxplots of the number of observed OTUs and Shannon diversity index values at baseline: day -1 (A-B), after clindamycin: day 0 (C-D) and post-infection: day 1 (E-F) timepoints of the experiment.
-Each circle represents the value for a stool sample from an individual mouse.
+Differences in microbial richness and diversity across mouse colony sources emerge after clindamycin treatment and infection.** A-F.
+Number of observed OTUs and Shannon diversity index values at baseline: day -1 (A-B), after clindamycin: day 0 (C-D) and post-infection: day 1 (E-F) timepoints of the experiment.
 Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 for all panels except for B (Table S3).
 Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S4).
+For A-F: circles represent experiment 1 mice, while triangles represent experiment 2 mice with each symbol representing the value for a stool sample from an individual mouse.
+Gray lines represent the median values for each source of mice.
 
@@
 Principal Coordinates Analysis of Theta YC distances from stools collected at baseline (A), post-clindamycin (B), and post-infection (C) timepoints of the experiment.
-Each circle represents a stool sample from an individual mouse.
+Each symbol represents a stool sample from an individual mouse, with circles representing experiment 1 mice and triangles representing experiment 2 mice.
 PERMANOVA analysis demonstrated that source and the interaction between source and cage explained most of the variation observed in the baseline (combined R^2^ = `r dn1_rsq_sum`), post-clindamycin (combined R^2^ = `r d0_rsq_sum`), and post-infection (combined R^2^ = `r d1_rsq_sum`) communities (all *P* = `r p_adonis_intial`, see Table S6).
@@
 **Figure 4.
-A subset of bacterial consistently vary across mouse colony sources despite clindamycin perturbation and *C.
+High inter-group variation across mouse sources is diminished by clindamycin treatment** A-C.
+Boxplots of the Theta YC distances of the 6 sources of mice relative to mice within the same source and experiment, mice within the same source and between experiments, mice within the same source, and mice from other groups at the baseline (A), after clindamycin treatment (B), and post-infection (C) timepoints.
+For comparisons within mice from the same source, symbols represent individual mouse samples: circles for experiment 1 and triangles for experiment 2.
+
+\newpage
+\includegraphics{figure_5.pdf}
+**Figure 5.
+A subset of bacteria consistently vary across mouse colony sources despite clindamycin perturbation and *C.
 difficile* challenge.** A-C.
@@
 \newpage
-\includegraphics{figure_5.pdf}
-**Figure 5.
+\includegraphics{figure_6.pdf}
+**Figure 6.
 Clindamycin treatment has the same effects on a subset of taxa regardless of colony source.** A-B.
@@
 \newpage
-\includegraphics{figure_6.pdf}
-**Figure 6.
-Bacterial community composition before, after clindamycin perturbation, and post-infection can predict *C.
-difficile* colonization status 7 days post-challenge.** A-B.
-Logistic regression classification model AUROCs to predict *C.
-difficile* CFU on D7 (Fig.
-1D, Fig.
-S3) based on the community relative abundances at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1) at either the OTU (A) or family (B) level.
-All models performed better than random chance (AUROC = 0.5), see (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13).
-A List of the 20 taxa that were ranked as most important to each model are listed in Table S14-15.
-
-\newpage
 \includegraphics{figure_7.pdf}
 **Figure 7.
-Key taxa that influence *C.
-difficile* colonization status vary by mouse colony source, change after clindamycin treatment, or both.** A-B.
-Venn diagrams that combine Fig.
-S4-6 summaries of taxa that were important to the day -1, 0, and 1 classification models (Table S14-15) and either overlapped with taxa that varied across vendors at the same timepoint, were impacted by clindamycin treatment, or both.
+Key OTUs that influence whether mice cleared *C.
+difficile* by day 7.** A.
+Baseline relative abundance data for 3 of the OTUs from the classification model based on day 0 OTU relative abundances that significantly varied across sources of mice and had high relative abundances.
+Symbols represent the relative abundance data for an individual mouse, circles represent mice that cleared *C.
+difficile* by day 7, X-shapes represent mice that were still colonized with *C.
+difficile*, and open circles represent mice that did not have *C.
+difficile* CFU counts for day 7 post-infection.
+Gray lines indicate the median relative abundances for each source.
+Asterisks are shown for pairwise Wilcoxon comparisons with Benjamini-Hochberg correction where *P* < 0.05.
+B.
+Venn diagram that combines Fig.
+S4 summaries of OTUs that were important to the day -1, 0, and 1 classification models (Table S14) and either overlapped with taxa that varied across vendors at the same timepoint, were impacted by clindamycin treatment, or both.
 See Fig.
-S4-6 for separate comparisons of taxa from the day -1, 0, and 1 classification models.
-Bold taxa signify bacteria that were important to more than 1 classification model.
+S4 for separate comparisons of taxa from the day -1, 0, and 1 classification models.
+Bold OTUs signify OTUs that were important to more than 1 classification model.
 
@@
 *Enterobacteriaceae* (OTU 1) and *Porphyromonadaceae* (OTU 7) were significantly impacted by clindamycin treatment and examining relative abundance dynamics over the course of the experiment indicated timepoints where relative abundances also significantly varied across sources of mice.
-Each point represents the relative abundance of an individual mouse stool sample and colored lines represent the mean relative abundances for each source of mice.
+Symbols represent the relative abundance data for an individual mouse, circles represent mice that cleared *C.
+difficile* by day 7, X-shapes represent mice that were still colonized with *C.
+difficile*, and open circles represent mice that did not have *C.
+difficile* CFU counts for day 7 post-infection.
+Colored lines indicate the median relative abundances for each source.
 The gray horizontal line represents the limit of detection.
-Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S16) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05, \*\*\, P < 0.01, \*\*\*\, P < 0.001).
+Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S16) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05).
 
@@
 *C.
-difficile* CFU variation across vendors is similar between the 2 experiments.** (A-B).
+difficile* CFU variation across vendors varies slightly across the 2 experiments.** A-B.
 *C.
@@
 Experiments were conducted approximately 3 months apart.
-Lines represent the mean CFU for each source and circles represent individual mice and the black line represents the limit of detection.
+Lines represent the median CFU for each source, symbols represent individual mice and the black line represents the limit of detection.
+C.
+Percent of mice that were colonized with *C.
+difficile* over the course of the experiment.
+Each day the percent is calculated based on the mice where *C.
+difficile* CFU was quantified for that particular day.
+Total N for each day: day 1 (N = `r n_day1`), day 2 (N = `r n_day2`), day 3 (N = `r n_day3`), day 4 (N = `r n_day4`), day 5 (N = `r n_day5`), day 6 (N = `r n_day6`), day 7 (N = `r n_day7`), day 8 (N = `r n_day8`), and day 9 (N = `r n_day9`).
 
@@
 **Figure S3.
-*C.
-difficile* colonization status 7 days post-infection varies across mice.** A.
+Bacterial community composition before, after clindamycin perturbation, and post-infection can predict *C.
+difficile* colonization status 7 days post-challenge.** A.
 Bar graph visualizations of overall day 7 C.
@@
 Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU) based on CFU g/stool data from 7 days post-infection.
-B-D.
-Classification input data used for models based on day -1 (B), day 0 (C), and day 1 (D) community relative abundance data.
-Only mice with day 7 *C.
-difficile* status and community relative abundance data for each respective timepoint were used as input data for the classification models.
+B.
+*C.
+difficile* CFU status on Day 7 within each mouse colony source.
+N = `r day7_status_n_min`-`r day7_status_n_max` mice per group.
+C-D.
+L2-regularized logistic regression classification model AUROCs to predict *C.
+difficile* CFU on D7 (Fig.
+1D, Fig.
+S3) based on the community relative abundances at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1) at either the OTU (C) or family (D) level.
+All models performed better than random chance (AUROC = 0.5), see (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13).
+A List of the 20 taxa that were ranked as most important to each model are listed in Table S14-15.
 
@@
 **Figure S4.
-Key taxa from classification model based on baseline community data vary by mouse colony source, clindamycin treatment, or both.** A-B.
-Venn diagrams of top 20 important OTUs (A) and families (B) from day -1 classification models (Table S14-15) that overlapped with taxa that varied across vendors at baseline, were impacted by clindamycin treatment, or both.
-Bold taxa signify bacteria that were important to more than 1 classification model.
+Key OTUs from classification models based on baseline, post-clindamycin treatment, or post-infection community data vary by mouse colony source, clindamycin treatment, or both.** A-C.
+Venn diagrams of top 20 important OTUs from baseline (A), post-clindamycin treatment (B), and post-infection (C) classification models (Table S14) that overlapped with OTUs that varied across vendors at baseline, were impacted by clindamycin treatment, or both.
+Bold OTUs signify OTUs that were important to more than 1 classification model.
 
@@
 **Figure S5.
-Key taxa from classification model based on post-clindamycin treatment community data vary by mouse colony source, clindamycin treatment, or both.** A-B.
-Venn diagrams of top 20 important OTUs (A) and families (B) from day 0 classification models (Table S14-15) that overlapped with taxa that still varied across vendors after clindamycin, were impacted by clindamycin treatment, or both.
-Bold taxa signify bacteria that were important to more than 1 classification model.
+Key families from classification models based on baseline, post-clindamycin treatment, or post-infection community data vary by mouse colony source, clindamycin treatment, or both.** A-C.
+Venn diagrams of top 20 important families from baseline (A), post-clindamycin treatment (B), and post-infection (C) classification models (Table S15) that overlapped with families that varied across vendors after clindamycin, were impacted by clindamycin treatment, or both.
+D.
+Venn diagrams that combines A-C summaries of familes that were important to the day -1, 0, and 1 classification models (Table S15) and either overlapped with familes that varied across vendors at the same timepoint, were impacted by clindamycin treatment, or both.
+Bold families signify families that were important to more than 1 classification model.
 
@@
 **Figure S6.
-Key taxa from classification model based on post-infection community data vary by mouse colony source, clindamycin treatment, or both.** A-B.
-Venn diagrams of top 20 important OTUs (A) and families (B) from day 1 classification models (Table S14-15) that overlapped with taxa that varied across vendors post-infection, were impacted by clindamycin treatment, or both.
-Bold taxa signify bacteria that were important to more than 1 classification model.
-
-\newpage
-\includegraphics{figure_S7.pdf}
-**Figure S7.
 Key families vary across sources throughout experiment.** Relative abundances of bold families from Fig.
-7B that were important for at least two classification models are shown over time.
+S5D that were important for at least two classification models are shown over time.
 A.
@@
 Examining the relative abundance dynamics throughout the experiment, identified timepoints where relative abundances also significantly varied across sources of mice.
-Each point represents the relative abundance of an individual mouse stool sample and colored lines represent the mean relative abundances for each source of mice.
+Symbols represent the relative abundance data for an individual mouse, circles represent mice that cleared *C.
+difficile* by day 7, X-shapes represent mice that were still colonized with *C.
+difficile*, and open circles represent mice that did not have *C.
+difficile* CFU counts for day 7 post-infection.
+Colored lines indicate the median relative abundances for each source.
 The gray horizontal line represents the limit of detection.
-Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S17) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05, \*\*\, P < 0.01, \*\*\*\, P < 0.001).
+Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S17) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05).
 
@@
 See Table S5 for PERMANOVA results).
+Circles represent mice from experiment 1 and triangles represent mice from expeirment 2.
 
@@
 **Table S11. Families with relative abudances that significantly changed after clindamycin treatment.**
-**Table S12. Statistical results of logistic regression model performances compared to random chance.**
+**Table S12. Statistical results of L2-regularized logistic regression model performances compared to random chance.**
 
-**Table S13. Pairwise Wilcoxan results of comparing all 6 logistic regression model performances.**
+**Table S13. Pairwise Wilcoxan results of comparing all 6 L2-regularized logistic regression model performances.**
 
-**Table S14. Top 20 most important OTUs for each of the 3 logistic regression models based on OTU relative abundance data.**
+**Table S14. Top 20 most important OTUs for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
 
-**Table S15. Top 20 most important families for each of the 3  logistic regression models based on OTU relative abundance data.**
+**Table S15. Top 20 most important families for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
 

### 2020-06-24 submission/manuscript.tex

@@
 mechanistic work examining CDIs in mouse models use a single university
-colony or vendor, which have lower inter-individual microbiota variation
-compared to humans. We treated mice from 6 different colony sources (2
+colony or vendor. We treated mice from 6 different colony sources (2
 University of Michigan colonies and 4 vendors) with a single clindamycin
@@
 \emph{C. difficile} colonization levels through 9 days post-infection.
-The microbiota was profiled throughout the experiment via 16S rRNA gene
-sequencing analysis to examine variation across colony sources and
-alterations due to clindamycin treatment and \emph{C. difficile}
-challenge. While all sources of mice were colonized 1-day
-post-infection, variation in \emph{C. difficile} colonization levels
-emerged from days 3-7 post-infection with 3 sources colonized with
-\emph{C. difficile} for longer and at higher levels. We identified
-bacterial taxa with different relative abundances across colony sources
-throughout the experiment, as well as taxa that were consistently
-impacted by clindamycin treatment. We created bacterial community-based
-logistic regression models that successfully classified mice based on
-their day 7 \emph{C. difficile} colonization status. Importantly, after
-examining the taxa that were most important to the classification
-models, we identified a subset of key taxa that varied across colony
-sources (\emph{Bacteroides}, \emph{Deferribacteraceae}), were altered by
-clindamycin (\emph{Porphyromonadaceae}, \emph{Ruminococcaceae}), or both
-(\emph{Enterobacteriaceae}, \emph{Enterococcus},
-\emph{Bifidobacteriaceae}, \emph{Coriobacteriaceae},
-\emph{Lachnospiraceae}, and \emph{Verrucomicrobiaceae}). These results
-suggest the response of the initial gut microbiota to clindamycin
-treatment influences \emph{C. difficile} 630 colonization dynamics.
+The microbiota was profiled via 16S rRNA gene sequencing analysis to
+examine variation across colony sources and alterations due to
+clindamycin treatment and \emph{C. difficile} challenge. While all
+sources of mice were colonized 1-day post-infection, variation in
+\emph{C. difficile} colonization levels emerged from days 3-7
+post-infection with 3 sources colonized with \emph{C. difficile} for
+slightly longer and at higher levels. We identified bacterial taxa with
+different relative abundances across colony sources throughout the
+experiment, as well as taxa that were consistently impacted by
+clindamycin treatment in all sources of mice. We created logistic
+regression models that successfully classified mice based on whether
+they cleared \emph{C. difficile} by 7 days post-infection using
+baseline, post-clindamycin, and post-infection community composition
+data. After examining the taxa that were most important to the
+classification models, we identified a subset of key taxa that varied
+across colony sources (\emph{Bacteroides}, \emph{Deferribacteraceae}),
+were altered by clindamycin (\emph{Porphyromonadaceae},
+\emph{Ruminococcaceae}), or both (\emph{Enterobacteriaceae},
+\emph{Enterococcus}, \emph{Bifidobacteriaceae},
+\emph{Coriobacteriaceae}, \emph{Lachnospiraceae}, and
+\emph{Verrucomicrobiaceae}). These results suggest the response of the
+initial gut microbiota to clindamycin treatment influences \emph{C.
+difficile} 630 colonization dynamics.
 
@@
 2008, which disrupted the gut microbiota with antibiotics to enable
-\emph{C. difficile} colonization, diarrhea, and weight loss (14). CDI
-mouse models have been used to examine translationally relevant
-questions regarding \emph{C. difficile}, including the role of the
-microbiota and efficacy of potential therapeutics for treating CDIs
-(15). However, microbiome variation between lab mice is much less than
-the variation observed between humans (16, 17). Additionally, studying
-the contribution of the microbiota to a particular disease phenotype in
-one set of lab mice after the same perturbation could yield a number of
-findings of which only a fraction may be driving the phenotype.
+\emph{C. difficile} colonization and symptoms such as diarrhea and
+weight loss (14). CDI mouse models have been used to examine
+translationally relevant questions regarding \emph{C. difficile},
+including the role of the microbiota and efficacy of potential
+therapeutics for treating CDIs (15). However, microbiome variation
+between lab mice is much less than the variation observed between humans
+(16, 17). Additionally, studying the contribution of the microbiota to a
+particular disease phenotype in one set of lab mice after the same
+perturbation could yield a number of findings of which only a fraction
+may be driving the phenotype.
 
@@
 frequently cited 2008 CDI mouse model (14). The day after infection,
-\emph{C. difficile} was detectable in all mice at a similar level (Mean
-CFU range: 3e+07-1.5e+08; \emph{P}\textsubscript{FDR} = 0.15),
+\emph{C. difficile} was detectable in all mice at a similar level
+(median CFU range: 2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15),
 indicating clindamycin rendered all mice susceptible regardless of
@@
 clearance. We conducted two experiments approximately 3 months apart and
-combined the data because the colonization dynamics across sources of
-mice varied similarily in both experiments (Fig. S1). Although \emph{C.
+while the colonization dynamics were similar across most sources of
+mice, there was some variation between the 2 experiments, particularly
+for the Schloss and Envigo mice (Fig. S1A-B, Fig. 1D). Although \emph{C.
 difficile} 630 causes mild symptoms in mice comparied to other \emph{C.
@@
 varied across sources of mice with the most weight lost two days
-post-infection (Fig. 1C and Table S2). Interestingly, mice ordered from
-Jackson, Taconic, and Envigo tended to lose the most weight, have higher
-\emph{C. difficile} CFU levels and take longer to clear the infection
-compared to the other sources of mice, which was particularly evident 7
-days post-infection (Fig. 1C and D). Importantly, there was also one
-Jackson and one Envigo mouse that died between 1- and 3-days
-post-infection. Thus, clindamycin rendered all mice susceptible to
-\emph{C. difficile} 630 colonization, regardless of colony source, but
-variation across sources emerged with 3 out of 6 sources taking longer
-to clear \emph{C. difficile}.
-
-\textbf{Bacterial communities vary across mouse colony sources.} Given
-the well known variation in mouse microbiomes across vendors and
-university colonies (25), we hypothesized that the variation in \emph{C.
-difficile} clearance could be explained by microbiota variation across
-the 6 sources. We used 16S rRNA gene sequencing to characterize the
-fecal bacterial communities from the mice over the course of the
-experiment. Since antibiotics and other risk factors of CDIs are
-associated with decreased microbiota diversity (44), we first examined
-alpha diversity measures across the 6 sources of mice. Examining the
-bacterial communities at baseline, prior to clindamycin treatment there
-was a significant difference in the number of observed OTUs
+post-infection (Fig. 1C, E and Table S2). Interestingly, mice ordered
+from Jackson, Taconic, and Envigo tended to lose more weight (although
+there was variation between experiments with Schloss and Envigo mice),
+have higher \emph{C. difficile} CFU levels and take longer to clear the
+infection compared to the other sources of mice, which was particularly
+evident 7 days post-infection (Fig. 1B-E), when 57.50 of the mice were
+still colonized with \emph{C. difficile} (Fig. S1C). By 9 days
+post-infection the majority of the mice from all sources had cleared
+\emph{C. difficile} (Fig. S1C) with the exception of 1 Taconic mouse
+from the first experiment and 2 Envigo mice from the second experiment.
+Importantly, there was also one Jackson and one Envigo mouse that died
+between 1- and 3-days post-infection during the second experiment. Thus,
+clindamycin rendered all mice susceptible to \emph{C. difficile} 630
+colonization, regardless of colony source, but variation across sources
+emerged with 3 out of 6 sources taking longer to clear \emph{C.
+difficile}.
+
+\textbf{Bacterial communities consistently vary across mouse colony
+sources despite antibiotic and infection perturbations.} Given the well
+known variation in mouse microbiomes across vendors and university
+colonies (25), we hypothesized that the variation in \emph{C. difficile}
+clearance could be explained by microbiota variation across the 6
+sources. We used 16S rRNA gene sequencing to characterize the fecal
+bacterial communities from the mice over the course of the experiment.
+Since antibiotics and other risk factors of CDIs are associated with
+decreased microbiota diversity (44), we first examined alpha diversity
+measures across the 6 sources of mice. Examining the bacterial
+communities at baseline, prior to clindamycin treatment there was a
+significant difference in the number of observed OTUs
 (\emph{P}\textsubscript{FDR} = 0.03), but not Shannon diversity index
@@
 cage significantly explained the observed variation only for the Schloss
-and Young lab mouse colonies (Fig. S2 and Table S7), suggesting vendor
-communities were relatively stable between the 2 experiments. Thus,
+and Young lab mouse colonies (Fig. S2 and Table S7), although most of
+the vendors also clustered by experiment, suggesting there was some
+community variation between the 2 experiments within each vendor. Thus,
 mouse colony source was the factor that explained the most variation
@@
 clindamycin treatment and \emph{C. difficile} challenge.
+
+Since there was some variation in microbiota communities between
+experiments at baseline, we next looked at how similar the communities
+were within the same source and between sources in response to
+clindamycin treatment and \emph{C. difficile} challenge (Fig. 4) . The
+baseline communities varied most between experiments for Schloss, Young,
+and Envigo mice and variation between sources of mice was high (Fig.
+4A). Clindamycin treatment reduced variation between experiments within
+Schloss, Young and Jackson mice and some of the variation between
+sources diminished (Fig. 4B). Post-infection, the community variation
+started to increase within sources of mice and variation between sources
+of mice started to return (Fig. 4C). By using mice from multiple sources
+we were able to increase the number of microbiota communities we tested
+with the clindamycin \emph{C. difficile} colonization mouse model.
 
@@
 bacterial taxonomy (45). Focusing on the baseline communities first,
-there were 268 OTUs and only 20 families (Table S8-9) with relative
+there were 268 OTUs and 20 families (Table S8-9) with relative
 abundances that varied across colony sources. Clindamycin treatment
@@
 the experimental perturbations that occurred during these 3 timepoints,
-there were 12 OTUs (Fig 4A-C) and 8 families with relative abundances
-that consistently varied across colony sources (Fig. 4D-F). Importantly,
+there were 12 OTUs (Fig 5A-C) and 8 families with relative abundances
+that consistently varied across colony sources (Fig. 5D-F). Importantly,
 some of the taxa that consistently varied across sources also shifted
@@
 OTUs and 18 families that were altered after clindamycin treatment (Fig.
-5 and Table S10-11). Interestingly, when we compared the list of
+6 and Table S10-11). Interestingly, when we compared the list of
 significant clindamycin impacted bacteria with the bacteria that
@@
 families (\emph{Porphyromonadaceae}, \emph{Enterococcaceae},
-\emph{Lachnospiraceae}) overlapped (Fig. 4, Fig. 5C-D). These findings
+\emph{Lachnospiraceae}) overlapped (Fig. 5, Fig. 6C-D). These findings
 demonstrate that clindamycin has a consistent impact on the fecal
@@
 both, we next wanted to determine which taxa were influencing the
-variation in \emph{C. difficile} colonization status at day 7 (Fig. 1D).
-We trained L2-regularized logistic regression models with input
+variation in \emph{C. difficile} colonization at day 7 (Fig. 1D, Fig.
+S1C). We trained L2-regularized logistic regression models with input
 bacterial community data from the baseline, post-clindamycin, and
 post-infection timepoints of the experiment to predict \emph{C.
-difficile} colonization status on day 7 (Fig. S3). All models were
+difficile} colonization status on day 7 (Fig. S3A-B). All models were
 better at predicting \emph{C. difficile} colonization status on day 7
@@
 trained with family level data with the expception of the models based
-on the post-infection (day 1) communities (Fig. 6). Interestingly, the
-model based on the post-clindamycin (day 0) community OTU data performed
-significantly better than all other models with an AUROC of 0.75
-(\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise comparisons;
-Table S13). Thus, we were able to use community bacterial relative
-abundance data alone to differentiate mice that were still colonized
-with \emph{C. difficile} on day 7 from those mice that had cleared the
+on the post-infection (day 1) communities (Fig. S3C-D). Interestingly,
+the model based on the post-clindamycin (day 0) community OTU data
+performed significantly better than all other models with an AUROC of
+0.75 (\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise
+comparisons; Table S13). Thus, we were able to use community bacterial
+relative abundance data alone to differentiate mice that had cleared
+\emph{C. difficile} before day 7 from the mice still colonized with
 \emph{C. difficile}. Interestingly, the model built with OTU relative
@@
 performance, we pulled out the top 20 taxa that had the highest absolute
-feature weights in each of the 6 models (Table S14-15). The top 20 taxa
-from each model were compared to the list of taxa that varied across
-colony source (Fig. 4 and Table S8-9) at the same timepoint and the taxa
-that were altered by clindamycin treatment (Fig. 5 and Table S10-11). We
-found a subset of OTUs and families that were important to the model and
-overlapped with bacteria that varied by either source, clindamycin
-treatment, or both (Fig. S4, S5, S6). Combining the overall results for
-the 3 OTU models identified 14 OTUs associated with source, 21 OTUs
-associated with clindamycin treatment, and 6 OTUs associated with both
-(Fig. 7A). Combining the overall results for the 3 family models
-identified 18 families associated with source, 14 families associated
-with clindamycin treatment and 8 families associated with both (Fig.
-7B). Several OTUs (\emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
-Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)}) and families
-(\emph{Bacteroidaceae, Deferribacteraceae, Enterococcaceae,
-Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
+feature weights in each of the 6 models (Table S14-15). First, we looked
+at OTUs from the model with the best performance that was based on the
+post-clindamycin treatment bacterial community data. While most of the
+20 OTUs had low relative abundances on day 0, \emph{Enterobacteriace},
+\emph{Bacteroides} and \emph{Proteus} had high relative abundances in at
+least one source of mice and significantly varied across sources (Fig.
+7A). Next, the top 20 taxa from each model were compared to the list of
+taxa that varied across colony source (Fig. 5 and Table S8-9) at the
+same timepoint and the taxa that were altered by clindamycin treatment
+(Fig. 6 and Table S10-11). We found a subset of OTUs and families that
+were important to the model and overlapped with bacteria that varied by
+either source, clindamycin treatment, or both (Fig. S4, S5A-C).
+Combining the overall results for the 3 OTU models identified 14 OTUs
+associated with source, 21 OTUs associated with clindamycin treatment,
+and 6 OTUs associated with both (Fig. 7B). Combining the overall results
+for the 3 family models identified 18 families associated with source,
+14 families associated with clindamycin treatment and 8 families
+associated with both (Fig. S5D). Several OTUs (\emph{Bacteroides (OTU
+2), Enterococcus (OTU 23), Enterobacteriaceae (OTU 1),
+Porphyromonadaceae (OTU 7)}) and families (\emph{Bacteroidaceae,
+Deferribacteraceae, Enterococcaceae, Lachnospiraceae,
+Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
 Verrucomicrobiaceae}) appeared across at least 2 models, so we examined
 how the relative abundances of these key taxa varied over the course of
-the experiment (Fig. 8 and Fig. S7). Throughout the experiment, there
+the experiment (Fig. 8 and Fig. S6). Throughout the experiment, there
 was at least 1 timepoint where relative abundances of these taxa
 significantly varied across sources (Table S16-17). Interestingly, there
-were no taxa that emerged as consistently enriched or depleted in
-Jackson, Taconic, and Envigo mice that were colonized for longer with
-\emph{C. difficile} 630, suggesting multiple bacteria influence the time
-needed to clear the infection. Together, these results suggest the
-initial bacterial communities and their responses to clindamycin have a
-large influence on the time needed to clear \emph{C. difficile}.
+were no taxa that emerged as consistently enriched or depleted in mice
+that were colonized past 7 days post-infection with \emph{C. difficile}
+630, suggesting multiple bacteria influence the time needed to clear the
+infection. Together, these results suggest the initial bacterial
+communities and their responses to clindamycin have a large influence on
+the time needed to clear \emph{C. difficile}.
 
@@
 Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
-Verrucomicrobiaceae} (Fig. 8, Fig. S7) as candidate bacteria within
+Verrucomicrobiaceae} (Fig. 8, Fig. S6) as candidate bacteria within
 these communities that were influencing variation in \emph{C. difficile}
@@
 there were no endogenous protective strains in the mice we received and
-other bacterial taxa were mediating the variation in \emph{C. difficile}
+other bacterial taxa mediated the variation in \emph{C. difficile}
 colonization across sources. Although all mice were susceptible to
 \emph{C. difficile} colonization, by following colonization over time we
-found Jackson, Taconic, and Envigo mice remained colonized for longer.
-We identified a subset of bacteria that were important in predicting
-whether a mouse was still colonized with \emph{C. difficile} status 7
-days post-infection. These results suggest a subset of the bacterial
-community is responsible for determining the length of time needed to
-clear \emph{C. difficile} colonization.
+found Jackson, Taconic, and Envigo mice remained colonized beyond 7 days
+post-infection. We identified a subset of bacteria that were important
+in predicting whether a mouse was still colonized with \emph{C.
+difficile} 7 days post-infection. These results suggest a subset of the
+bacterial community is responsible for determining the length of time
+needed to clear \emph{C. difficile} colonization.
 
@@
 predicting \emph{C. difficile} colonization status 7 days
-post-infection.
+post-infection. Since all but 3 mice eventually cleared \emph{C.
+difficile} 630 by 9 days post-infection and the model built with the
+post-clindamycin OTU relative abundance data had the best performance,
+our results suggest clindamycin treatment had a large role in
+determining \emph{C. difficile} susceptibility and clearance in the
+mice.
 
@@
 alternative approach that has been used in some CDI studies (49--54) is
-to use mice associated with human microbiotas. However, a major caveat
-to this method is the substantial loss of human microbiota community
+to associate mice with human microbiotas. However, a major caveat to
+this method is the substantial loss of human microbiota community
 members upon transfer to mice (55, 56). Additionally with the exception
@@
 and \emph{Proteus} have all been associated with human CDIs (7) and were
-well represented in our study, suggesting most of the vendors are
+well represented in our study, suggesting most of the mouse sources are
 suitable for gaining insights into microbiota associated factors
@@
 toning differences in IgA and Th17 cells across mice from different
-vendors have also been documented and (67, 68), may also influence
+vendors have also been documented and (67, 68) may also influence
 response to CDI, particularly in the context of severe CDIs (69, 70).
@@
 and Health Research Postdoctoral Translation Scholars Program
-(UL1TR002240 and KL2TR002241). We thank members of the Schloss lab for
-feedback on planning the experiments and data presentation, as well as
-code tutorials and feedback through Code club. In particular, we want to
-thank Begum Topçuoğlu for help with implementing logistic regression
+(UL1TR002240). We thank members of the Schloss lab for feedback on
+planning the experiments and data presentation, as well as code
+tutorials and feedback through Code club. In particular, we want to
+thank Begüm Topçuoğlu for help with implementing L2 logistic regression
 models using her
-\href{https://github.com/SchlossLab/ML_pipeline_microbiome}{pipeline}
-and Ana Taylor for help with some media preparation and sample
-collection. We also thank members of Vincent Young's lab, particularly
-Kimberly Vendrov, for guidance with the \emph{C. difficile} infection
-mouse model and donating the mice. We also want to thank the Unit for
-Laboratory Animal Medicine at the University of Michigan for maintaining
-our mouse colony and providing the institutional support for our mouse
+\href{https://github.com/SchlossLab/ML_pipeline_microbiome}{pipeline},
+Ana Taylor for help with some media preparation and sample collection,
+and Nicholas Lesniak for his critical feedback on the manuscript. We
+also thank members of Vincent Young's lab, particularly Kimberly
+Vendrov, for guidance with the \emph{C. difficile} infection mouse model
+and donating the mice. We also want to thank the Unit for Laboratory
+Animal Medicine at the University of Michigan for maintaining our mouse
+colony and providing the institutional support for our mouse
 experiments. Finally, we thank Kwi Kim, Austin Campbell, and Kimberly
@@
 \hypertarget{ref-Teng2019}{}
-1. Teng C, Reveles KR, Obodozie-Ofoegbu OO, Frei CR. 2019. Clostridium
-difficile infection risk with important antibiotic classes: An analysis
-of the FDA adverse event reporting system. International Journal of
-Medical Sciences 16:630--635.
+1. Teng C, Reveles KR, Obodozie-Ofoegbu OO, Frei CR. 2019.
+\emph{Clostridium difficile} infection risk with important antibiotic
+classes: An analysis of the FDA adverse event reporting system.
+International Journal of Medical Sciences 16:630--635.
 
@@
 2. Kelly C. 2012. Can we identify patients at high risk of recurrent
-clostridium difficile infection? Clinical Microbiology and Infection
-18:21--27.
+\emph{Clostridium difficile} infection? Clinical Microbiology and
+Infection 18:21--27.
 
@@
 3. Zacharioudakis IM, Zervou FN, Pliakos EE, Ziakas PD, Mylonakis E.
-2015. Colonization with toxinogenic c. difficile upon hospital
+2015. Colonization with toxinogenic \emph{C. difficile} upon hospital
 admission, and risk of infection: A systematic review and meta-analysis.
@@
 4. Crobach MJT, Vernon JJ, Loo VG, Kong LY, Péchiné S, Wilcox MH,
-Kuijper EJ. 2018. Understanding clostridium difficile colonization.
-Clinical Microbiology Reviews 31.
+Kuijper EJ. 2018. Understanding \emph{Clostridium difficile}
+colonization. Clinical Microbiology Reviews 31.
 
@@
 5. Zhang L, Dong D, Jiang C, Li Z, Wang X, Peng Y. 2015. Insight into
-alteration of gut microbiota in clostridium difficile infection and
-asymptomatic c. difficile colonization. Anaerobe 34:1--7.
+alteration of gut microbiota in \emph{Clostridium difficile} infection
+and asymptomatic c. difficile colonization. Anaerobe 34:1--7.
 
@@
 6. VanInsberghe D, Elsherbini JA, Varian B, Poutahidis T, Erdman S, Polz
-MF. 2020. Diarrhoeal events can trigger long-term clostridium difficile
-colonization with recurrent blooms. Nature Microbiology 5:642--650.
+MF. 2020. Diarrhoeal events can trigger long-term \emph{Clostridium
+difficile} colonization with recurrent blooms. Nature Microbiology
+5:642--650.
 
@@
 9. Seekatz AM, Rao K, Santhosh K, Young VB. 2016. Dynamics of the fecal
-microbiome in patients with recurrent and nonrecurrent clostridium
-difficile infection. Genome Medicine 8.
+microbiome in patients with recurrent and nonrecurrent \emph{Clostridium
+difficile} infection. Genome Medicine 8.
 
@@
 Kashyap PC. 2016. Gut microbiome predictors of treatment response and
-recurrence in primaryClostridium difficileinfection. Alimentary
+recurrence in primary \emph{Clostridium difficile} infection. Alimentary
 Pharmacology \& Therapeutics 44:715--727.
@@
 11. Pakpour S, Bhanvadia A, Zhu R, Amarnani A, Gibbons SM, Gurry T, Alm
-EJ, Martello LA. 2017. Identifying predictive features of clostridium
-difficile infection recurrence before, during, and after primary
-antibiotic treatment. Microbiome 5.
+EJ, Martello LA. 2017. Identifying predictive features of
+\emph{Clostridium difficile} infection recurrence before, during, and
+after primary antibiotic treatment. Microbiome 5.
 
@@
 Young VB, Higgins PDR. 2020. Temporal gut microbial changes predict
-recurrent clostridiodes difficile infection in patients with and without
-ulcerative colitis. Inflammatory Bowel Diseases.
+recurrent \emph{Clostridioides difficile} infection in patients with and
+without ulcerative colitis. Inflammatory Bowel Diseases.
 
@@
 13. Hutton ML, Mackin KE, Chakravorty A, Lyras D. 2014. Small animal
-models for the study ofClostridium difficiledisease pathogenesis. FEMS
-Microbiology Letters 352:140--149.
+models for the study of \emph{Clostridium difficile} disease
+pathogenesis. FEMS Microbiology Letters 352:140--149.
 
@@
 14. Chen X, Katchar K, Goldsmith JD, Nanthakumar N, Cheknis A, Gerding
-DN, Kelly CP. 2008. A mouse model of clostridium difficileAssociated
-disease. Gastroenterology 135:1984--1992.
+DN, Kelly CP. 2008. A mouse model of \emph{Clostridium
+difficile}-associated disease. Gastroenterology 135:1984--1992.
 
@@
 15. Best EL, Freeman J, Wilcox MH. 2012. Models for the study of
-clostridium difficile infection. Gut Microbes 3:145--167.
+\emph{Clostridium difficile} infection. Gut Microbes 3:145--167.
 
@@
 VB. 2011. The interplay between microbiome dynamics and pathogen
-dynamics in a murine model ofClostridium difficileInfection. Gut
-Microbes 2:145--158.
+dynamics in a murine model of \emph{Clostridium difficile} infection
+2:145--158.
 
@@
 Tiffany CR, Byndloss MX, Byndloss AJ, Bäumler AJ. 2019. Endogenous
-enterobacteriaceae underlie variation in susceptibility to salmonella
-infection. Nature Microbiology 4:1057--1064.
+Enterobacteriaceae underlie variation in susceptibility to
+\emph{Salmonella} infection. Nature Microbiology 4:1057--1064.
 
@@
 Neumann-Schaal M, Strowig T. 2020. Variations in microbiota composition
-of laboratory mice influence citrobacter rodentium infection via
+of laboratory mice influence \emph{Citrobacter rodentium} infection via
 variable short-chain fatty acid production. PLOS Pathogens 16:e1008448.
@@
 SR, Wilhelm SW. 2016. Functional characteristics of the gut microbiome
-in c57bl/6 mice differentially susceptible to plasmodium yoelii.
+in C57BL/6 mice differentially susceptible to \emph{Plasmodium yoelii}.
 Frontiers in Microbiology 7.
@@
 26. Etienne-Mesmin L, Chassaing B, Adekunle O, Mattei LM, Bushman FD,
-Gewirtz AT. 2017. Toxin-positive clostridium difficile latently infect
-mouse colonies and protect against highly pathogenic c. difficile. Gut
-67:860--871.
+Gewirtz AT. 2017. Toxin-positive \emph{Clostridium difficile} latently
+infect mouse colonies and protect against highly pathogenic \emph{C.
+difficile}. Gut 67:860--871.
 
@@
 Chiu IM. 2020. Gut-innervating nociceptor neurons regulate peyer's patch
-microfold cells and SFB levels to mediate salmonella host defense. Cell
-180:33--49.e22.
+microfold cells and SFB levels to mediate \emph{Salmonella} host
+defense. Cell 180:33--49.e22.
 
@@
 2017. Enhancement of IFNgamma production by distinct commensals
-ameliorates salmonella-induced disease. Cell Host \& Microbe
+ameliorates \emph{Salmonella}-induced disease. Cell Host \& Microbe
 21:682--694.e5.
@@
 29. Rolig AS, Cech C, Ahler E, Carter JE, Ottemann KM. 2013. The degree
-of helicobacter pylori-triggered inflammation is manipulated by
+of \emph{Helicobacter pylori}-triggered inflammation is manipulated by
 preinfection host microbiota. Infection and Immunity 81:1382--1389.
@@
 30. Ge Z, Sheh A, Feng Y, Muthupalani S, Ge L, Wang C, Kurnick S,
-Mannion A, Whary MT, Fox JG. 2018. Helicobacter pylori-infected c57bl/6
-mice with different gastrointestinal microbiota have contrasting gastric
-pathology, microbial and host immune responses. Scientific Reports 8.
+Mannion A, Whary MT, Fox JG. 2018. \emph{Helicobacter pylori}-infected
+C57BL/6 mice with different gastrointestinal microbiota have contrasting
+gastric pathology, microbial and host immune responses. Scientific
+Reports 8.
 
 \hypertarget{ref-Lawley2013}{}
-31. Lawley TD, Young VB. 2013. Murine models to study clostridium
-difficile infection and transmission. Anaerobe 24:94--97.
+31. Lawley TD, Young VB. 2013. Murine models to study \emph{Clostridium
+difficile} infection and transmission. Anaerobe 24:94--97.
 
@@
 microbiota following a single dose of clindamycin results in sustained
-susceptibility to clostridium difficile-induced colitis. Infection and
-Immunity 80:62--73.
+susceptibility to \emph{Clostridium difficile}-induced colitis.
+Infection and Immunity 80:62--73.
 
@@
 Precision microbiome reconstitution restores bile acid mediated
-resistance to clostridium difficile. Nature 517:205--208.
+resistance to \emph{Clostridium difficile}. Nature 517:205--208.
 
@@
 34. Spinler JK, Brown A, Ross CL, Boonma P, Conner ME, Savidge TC. 2016.
-Administration of probiotic kefir to mice with clostridium difficile
-infection exacerbates disease. Anaerobe 40:54--57.
+Administration of probiotic kefir to mice with \emph{Clostridium
+difficile} infection exacerbates disease. Anaerobe 40:54--57.
 
@@
 Pre-colonization with the commensal fungus candida albicans reduces
-murine susceptibility to clostridium difficile infection. Gut Microbes
-1--13.
+murine susceptibility to \emph{Clostridium difficile} infection. Gut
+Microbes 1--13.
 
@@
 36. McKee RW, Aleksanyan N, Garrett EM, Tamayo R. 2018. Type IV pili
-promote clostridium difficile adherence and persistence in a mouse model
-of infection. Infection and Immunity 86.
+promote \emph{Clostridium difficile} adherence and persistence in a
+mouse model of infection. Infection and Immunity 86.
 
@@
 37. Yamaguchi T, Konishi H, Aoki K, Ishii Y, Chono K, Tateda K. 2020.
-The gut microbiome diversity of clostridioides difficile-inoculated mice
-treated with vancomycin and fidaxomicin. Journal of Infection and
-Chemotherapy 26:483--491.
+The gut microbiome diversity of \emph{Clostridioides
+difficile}-inoculated mice treated with vancomycin and fidaxomicin.
+Journal of Infection and Chemotherapy 26:483--491.
 
@@
 L, Sturzenbecker LJ, Sabalski J, Stein PD, Webb ML, Hilbert DW. 2018.
-Treatment of clostridium difficile infection with a small-molecule
-inhibitor of toxin UDP-glucose hydrolysis activity. Antimicrobial Agents
-and Chemotherapy 62.
+Treatment of \emph{Clostridium difficile} infection with a
+small-molecule inhibitor of toxin UDP-glucose hydrolysis activity.
+Antimicrobial Agents and Chemotherapy 62.
 
@@
 39. Quigley L, Coakley M, Alemayehu D, Rea MC, Casey PG, O'Sullivan,
-Murphy E, Kiely B, Cotter PD, Hill C, Ross RP. 2019. Lactobacillus
-gasseri APC 678 reduces shedding of the pathogen clostridium difficile
-in a murine model. Frontiers in Microbiology 10.
+Murphy E, Kiely B, Cotter PD, Hill C, Ross RP. 2019. \emph{Lactobacillus
+gasseri} APC 678 reduces shedding of the pathogen \emph{Clostridium
+difficile} in a murine model. Frontiers in Microbiology 10.
 
@@
 Microbial bile salt hydrolases mediate the efficacy of faecal microbiota
-transplant in the treatment of recurrent clostridioides difficile
+transplant in the treatment of recurrent \emph{Clostridioides difficile}
 infection. Gut 68:1791--1800.
@@
 2019. The proton pump inhibitor omeprazole does not promote
-clostridioides difficile colonization in a murine model. mSphere 4.
+\emph{Clostridioides difficile} colonization in a murine model. mSphere
+4.
 
 \hypertarget{ref-Guh2018}{}
-42. Guh AY, Kutty PK. 2018. Clostridioides difficileInfection. Annals of
-Internal Medicine 169:ITC49.
+42. Guh AY, Kutty PK. 2018. \emph{Clostridioides difficile} infection
+169:ITC49.
 
@@
 VB. 2011. Cefoperazone-treated mice as an experimental platform to
-assess differential virulence ofClostridium difficilestrains. Gut
-Microbes 2:326--334.
+assess differential virulence of \emph{Clostridium difficile} strains.
+Gut Microbes 2:326--334.
 
@@
 44. Ross CL, Spinler JK, Savidge TC. 2016. Structural and functional
-changes within the gut microbiota and susceptibility to clostridium
-difficile infection. Anaerobe 41:37--43.
+changes within the gut microbiota and susceptibility to
+\emph{Clostridium difficile} infection. Anaerobe 41:37--43.
 
@@
 Mastroeni P, Scott P, Raisen C, Mottram L, Fairweather NF, Wren BW,
-Parkhill J, Dougan G. 2009. Antibiotic treatment of clostridium
-difficile carrier mice triggers a supershedder state, spore-mediated
+Parkhill J, Dougan G. 2009. Antibiotic treatment of \emph{Clostridium
+difficile} carrier mice triggers a supershedder state, spore-mediated
 transmission, and severe disease in immunocompromised hosts. Infection
@@
 of the intestinal microbiota with a simple, defined bacteriotherapy
-resolves relapsing clostridium difficile disease in mice. PLoS Pathogens
-8:e1002995.
+resolves relapsing \emph{Clostridium difficile} disease in mice. PLoS
+Pathogens 8:e1002995.
 
@@
 Young VB, Kamada N. 2020. Interleukin-22-mediated host glycosylation
-prevents clostridioides difficile infection by modulating the metabolic
-activity of the gut microbiota. Nature Medicine 26:608--617.
+prevents \emph{Clostridioides difficile} infection by modulating the
+metabolic activity of the gut microbiota. Nature Medicine 26:608--617.
 
@@
 Khanna S, Pardi DS, Grover M, Patel R, Chia N, Nelson H, Sonnenburg JL,
-Farrugia G, Kashyap PC. 2018. Clostridioides difficile uses amino acids
-associated with gut microbial dysbiosis in a subset of patients with
-diarrhea. Science Translational Medicine 10:eaam7019.
+Farrugia G, Kashyap PC. 2018. \emph{Clostridioides difficile} uses amino
+acids associated with gut microbial dysbiosis in a subset of patients
+with diarrhea. Science Translational Medicine 10:eaam7019.
 
@@
 51. Robinson CD, Auchtung JM, Collins J, Britton RA. 2014. Epidemic
-clostridium difficile strains demonstrate increased competitive fitness
-compared to nonepidemic isolates. Infection and Immunity 82:2815--2825.
+\emph{Clostridium difficile} strains demonstrate increased competitive
+fitness compared to nonepidemic isolates. Infection and Immunity
+82:2815--2825.
 
@@
 52. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. 2015.
-Humanized microbiota mice as a model of recurrent clostridium difficile
-disease. Microbiome 3.
+Humanized microbiota mice as a model of recurrent \emph{Clostridium
+difficile} disease. Microbiome 3.
 
@@
 TD, Auchtung JM, Britton RA. 2018. Dietary trehalose enhances virulence
-of epidemic clostridium difficile. Nature 553:291--294.
+of epidemic \emph{Clostridium difficile}. Nature 553:291--294.
 
@@
 DM, Sonnenburg JL. 2018. Microbiota-accessible carbohydrates suppress
-clostridium difficile infection in a murine model. Nature Microbiology
-3:662--669.
+\emph{Clostridium difficile} infection in a murine model. Nature
+Microbiology 3:662--669.
 
@@
 meta-omics reveals a fungus-associated bacteriome and distinct
-functional pathways in clostridioides difficile infection. mSphere 4.
+functional pathways in \emph{Clostridioides difficile} infection.
+mSphere 4.
 
@@
 Schreiber S. 2017. Efficacy of sterile fecal filtrate transfer for
-treating patients with clostridium difficile infection. Gastroenterology
-152:799--811.e7.
+treating patients with \emph{Clostridium difficile} infection.
+Gastroenterology 152:799--811.e7.
 
@@
 Chan MCW, Wu JCY, Chan FKL, Yu J, Sung JJY, Ng SC. 2017. Bacteriophage
-transfer during faecal microbiota transplantation inClostridium
-difficile infection is associated with treatment outcome. Gut
+transfer during faecal microbiota transplantation in \emph{Clostridium
+difficile} infection is associated with treatment outcome. Gut
 gutjnl--2017--313952.
@@
 fungal dysbiosis correlates with reduced efficacy of fecal microbiota
-transplantation in clostridium difficile infection. Nature
+transplantation in \emph{Clostridium difficile} infection. Nature
 Communications 9.
@@
 C-AD, Dubberke ER, Mucha PJ, Henderson JP. 2019. Metabolomic networks
-connect host-microbiome processes to human clostridioides difficile
-infections. Journal of Clinical Investigation 129:3792--3806.
+connect host-microbiome processes to human \emph{Clostridioides
+difficile} infections. Journal of Clinical Investigation 129:3792--3806.
 
@@
 64. Fletcher JR, Erwin S, Lanzas C, Theriot CM. 2018. Shifts in the gut
-metabolome and clostridium difficile transcriptome throughout
+metabolome and \emph{Clostridium difficile} transcriptome throughout
 colonization and infection in a mouse model. mSphere 3.
@@
 Chiavelli A, D'Erchia AM, Sethi MK, Pabst O, Marzano M, Moretti S,
-Romani L, Penna G, Pesole G, Rescigno M. 2015. BALB/c and c57bl/6 mice
+Romani L, Penna G, Pesole G, Rescigno M. 2015. BALB/c and C57BL/6 mice
 differ in polyreactive IgA abundance, which impacts the generation of
@@
 69. Azrad M, Hamo Z, Tkhawkho L, Peretz A. 2018. Elevated serum
-immunoglobulin a levels in patients with clostridium difficile infection
-are associated with mortality. Pathogens and Disease 76.
+immunoglobulin a levels in patients with \emph{Clostridium difficile}
+infection are associated with mortality. Pathogens and Disease 76.
 
@@
 Simpson ME, Scully KW, Abhyankar MM, Petri WA. 2019. Colitis-induced
-th17 cells increase the risk for severe subsequent clostridium difficile
-infection. Cell Host \& Microbe 25:756--765.e5.
+th17 cells increase the risk for severe subsequent \emph{Clostridium
+difficile} infection. Cell Host \& Microbe 25:756--765.e5.
 
@@
 
-\hypertarget{ref-Topcuoglu2019}{}
-80. Topçuoğlu BD, Lesniak NA, Ruffin M, Wiens J, Schloss PD. 2019. A
+\hypertarget{ref-Topcuoglu2020}{}
+80. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
 framework for effective application of machine learning to
-microbiome-based classification problems.
+microbiome-based classification problems. mBio 11.
 
@@
 \includegraphics{figure_1.pdf} \textbf{Figure 1. Clindamycin is
-sufficient to promote \emph{C. difficile} colonization regardless of
-colony source, but clearance time varies across sources of C57BL/6
-mice.} A. Setup of the experimental timeline. Mice for the experiments
-were obtained from 6 different sources: the Schloss (N = 8) and Young
-lab (N = 9) colonies at the University of Michigan, the Jackson
-Laboratory (N = 8), Charles River Laboratory (N = 8), Taconic
-Biosciences (N = 8), and Envigo (N = 8). All mice were administered 10
-mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with
-\emph{C. difficile} 630 spores on day 0. Mice were weighed and feces was
-collected daily through the end of the experiment (9 days
-post-infection). Note: 3 mice died during course of experiment. 1
-Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse
-between 1- and 3-days post-infection. B. \emph{C. difficile} CFU/gram
-stool measured over time (N = 20-49 mice per timepoint) via serial
-dilutions. The black line represents the limit of detection for the
-first serial dilution. CFU quantification data was not available for
-each mouse due to early deaths, stool sampling difficulties, and not
-plating all of the seriral dilutions. C. Mouse weight change measured in
-grams over time (N = 45-49 mice per timepoint), all mice were normalized
-to the weight recorded 1 day before infection. For B-C: timepoints where
-differences across sources of mice were statistically significant by
-Kruskal-Wallis test with Benjamini-Hochberg correction for testing
-across multiple days (Table S1 and Table S2) are reflected by the
-asterisk(s) above each timepoint (*, P \textless{} 0.05, **, P
-\textless{} 0.01, ***, P \textless{} 0.001). Lines represent the mean
-for each source and circles represent individual mice. D. \emph{C.
-difficile} CFU status on Day 7 within each mouse colony source. Mice
-were classified as colonized or cleared (not detectable at the limit of
-detection of 100 CFU). N = 5-9 mice per group.
+sufficient to promote \emph{C. difficile} colonization in all mice, but
+clearance time varies across sources of C57BL/6 mice.} A. Setup of the
+experimental timeline. Mice for the experiments were obtained from 6
+different sources: the Schloss (N = 8) and Young lab (N = 9) colonies at
+the University of Michigan, the Jackson Laboratory (N = 8), Charles
+River Laboratory (N = 8), Taconic Biosciences (N = 8), and Envigo (N =
+8). All mice were administered 10 mg/kg clindamycin intraperitoneally
+(IP) 1 day before challenge with \emph{C. difficile} 630 spores on day
+0. Mice were weighed and feces was collected daily through the end of
+the experiment (9 days post-infection). Note: 3 mice died during course
+of experiment. 1 Taconic mouse prior to infection and 1 Jackson and 1
+Envigo mouse between 1- and 3-days post-infection. B. \emph{C.
+difficile} CFU/gram stool measured over time (N = 20-49 mice per
+timepoint) via serial dilutions. The black line represents the limit of
+detection for the first serial dilution. CFU quantification data was not
+available for each mouse due to early deaths, stool sampling
+difficulties, and not plating all of the serial dilutions. C. Mouse
+weight change measured in grams over time (N = 45-49 mice per
+timepoint), all mice were normalized to the weight recorded 1 day before
+infection. For B-C: timepoints where differences across sources of mice
+were statistically significant by Kruskal-Wallis test with
+Benjamini-Hochberg correction for testing across multiple days (Table S1
+and Table S2) are reflected by the asterisk(s) above each timepoint (*,
+\emph{P} \textless{} 0.05). Lines represent the median for each source
+and circles represent individual mice from experiment 1 while triangles
+represent mice from experiment 2. D. \emph{C. difficile} CFU/gram stool
+on day 7 post-infection across sources of mice with asterisks for
+pairwise Wilcoxon comparisons with Benjamini-Hochberg correction where
+\emph{P} \textless{} 0.05. E. Mouse weight change 2 days post-infection
+across sources of mice, no pairwise Wilcoxon comparisons were
+significant after Benjamini-Hochberg correction. For D-E. Circles
+represent experiment 1 mice, triangles represent experiment 2 mice and
+gray lines indicate the median values for each group.
 
@@
 microbial richness and diversity across mouse colony sources emerge
-after clindamycin treatment and infection.} A-B. Boxplots of the number
-of observed OTUs and Shannon diversity index values at baseline: day -1
-(A-B), after clindamycin: day 0 (C-D) and post-infection: day 1 (E-F)
-timepoints of the experiment. Each circle represents the value for a
-stool sample from an individual mouse. Data were analyzed by
-Kruskal-Wallis test with Benjamini-Hochberg correction for testing each
-day of the experiment and the adjusted \emph{P} value was \textless{}
-0.05 for all panels except for B (Table S3). Significant \emph{P} values
-from the pairwise Wilcoxon comparisons between sources with
-Benjamini-Hochberg correction are shown (Table S4).
+after clindamycin treatment and infection.} A-F. Number of observed OTUs
+and Shannon diversity index values at baseline: day -1 (A-B), after
+clindamycin: day 0 (C-D) and post-infection: day 1 (E-F) timepoints of
+the experiment. Data were analyzed by Kruskal-Wallis test with
+Benjamini-Hochberg correction for testing each day of the experiment and
+the adjusted \emph{P} value was \textless{} 0.05 for all panels except
+for B (Table S3). Significant \emph{P} values from the pairwise Wilcoxon
+comparisons between sources with Benjamini-Hochberg correction are shown
+(Table S4). For A-F: circles represent experiment 1 mice, while
+triangles represent experiment 2 mice with each symbol representing the
+value for a stool sample from an individual mouse. Gray lines represent
+the median values for each source of mice.
 
@@
 collected at baseline (A), post-clindamycin (B), and post-infection (C)
-timepoints of the experiment. Each circle represents a stool sample from
-an individual mouse. PERMANOVA analysis demonstrated that source and the
-interaction between source and cage explained most of the variation
-observed in the baseline (combined R\textsuperscript{2} = 0.90),
-post-clindamycin (combined R\textsuperscript{2} = 0.99), and
-post-infection (combined R\textsuperscript{2} = 0.88) communities (all
-\emph{P} = 0.0001, see Table S6).
-
-\newpage
-
-\includegraphics{figure_4.pdf} \textbf{Figure 4. A subset of bacterial
+timepoints of the experiment. Each symbol represents a stool sample from
+an individual mouse, with circles representing experiment 1 mice and
+triangles representing experiment 2 mice. PERMANOVA analysis
+demonstrated that source and the interaction between source and cage
+explained most of the variation observed in the baseline (combined
+R\textsuperscript{2} = 0.90), post-clindamycin (combined
+R\textsuperscript{2} = 0.99), and post-infection (combined
+R\textsuperscript{2} = 0.88) communities (all \emph{P} = 0.0001, see
+Table S6).
+
+\newpage
+
+\includegraphics{figure_4.pdf} \textbf{Figure 4. High inter-group
+variation across mouse sources is diminished by clindamycin treatment}
+A-C. Boxplots of the Theta YC distances of the 6 sources of mice
+relative to mice within the same source and experiment, mice within the
+same source and between experiments, mice within the same source, and
+mice from other groups at the baseline (A), after clindamycin treatment
+(B), and post-infection (C) timepoints. For comparisons within mice from
+the same source, symbols represent individual mouse samples: circles for
+experiment 1 and triangles for experiment 2.
+
+\newpage
+
+\includegraphics{figure_5.pdf} \textbf{Figure 5. A subset of bacteria
 consistently vary across mouse colony sources despite clindamycin
@@
 
-\includegraphics{figure_5.pdf} \textbf{Figure 5. Clindamycin treatment
+\includegraphics{figure_6.pdf} \textbf{Figure 6. Clindamycin treatment
 has the same effects on a subset of taxa regardless of colony source.}
@@
 
-\includegraphics{figure_6.pdf} \textbf{Figure 6. Bacterial community
-composition before, after clindamycin perturbation, and post-infection
-can predict \emph{C. difficile} colonization status 7 days
-post-challenge.} A-B. Logistic regression classification model AUROCs to
-predict \emph{C. difficile} CFU on D7 (Fig. 1D, Fig. S3) based on the
-community relative abundances at baseline (day -1), post-clindamycin
-(day 0), and post-infection (day 1) at either the OTU (A) or family (B)
-level. All models performed better than random chance (AUROC = 0.5), see
-(all \emph{P} \(\le\) 5e-15; Table S12) and the model built with
-post-clindamycin treated bacterial OTU relative abundances had the best
-performance ((\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise
-comparisons; Table S13). A List of the 20 taxa that were ranked as most
-important to each model are listed in Table S14-15.
-
-\newpage
-
-\includegraphics{figure_7.pdf} \textbf{Figure 7. Key taxa that influence
-\emph{C. difficile} colonization status vary by mouse colony source,
-change after clindamycin treatment, or both.} A-B. Venn diagrams that
-combine Fig. S4-6 summaries of taxa that were important to the day -1,
-0, and 1 classification models (Table S14-15) and either overlapped with
-taxa that varied across vendors at the same timepoint, were impacted by
-clindamycin treatment, or both. See Fig. S4-6 for separate comparisons
-of taxa from the day -1, 0, and 1 classification models. Bold taxa
-signify bacteria that were important to more than 1 classification
-model.
+\includegraphics{figure_7.pdf} \textbf{Figure 7. Key OTUs that influence
+whether mice cleared \emph{C. difficile} by day 7.} A. Baseline relative
+abundance data for 3 of the OTUs from the classification model based on
+day 0 OTU relative abundances that significantly varied across sources
+of mice and had high relative abundances. Symbols represent the relative
+abundance data for an individual mouse, circles represent mice that
+cleared \emph{C. difficile} by day 7, X-shapes represent mice that were
+still colonized with \emph{C. difficile}, and open circles represent
+mice that did not have \emph{C. difficile} CFU counts for day 7
+post-infection. Gray lines indicate the median relative abundances for
+each source. Asterisks are shown for pairwise Wilcoxon comparisons with
+Benjamini-Hochberg correction where \emph{P} \textless{} 0.05. B. Venn
+diagram that combines Fig. S4 summaries of OTUs that were important to
+the day -1, 0, and 1 classification models (Table S14) and either
+overlapped with taxa that varied across vendors at the same timepoint,
+were impacted by clindamycin treatment, or both. See Fig. S4 for
+separate comparisons of taxa from the day -1, 0, and 1 classification
+models. Bold OTUs signify OTUs that were important to more than 1
+classification model.
 
@@
 course of the experiment indicated timepoints where relative abundances
-also significantly varied across sources of mice. Each point represents
-the relative abundance of an individual mouse stool sample and colored
-lines represent the mean relative abundances for each source of mice.
-The gray horizontal line represents the limit of detection. Timepoints
-where differences across sources of mice were statistically significant
-by Kruskal-Wallis test with Benjamini-Hochberg correction for testing
-across multiple days (Table S16) are identified by the asterisk(s) above
-each timepoint (*, P \textless{} 0.05, **, P \textless{} 0.01, ***, P
-\textless{} 0.001).
+also significantly varied across sources of mice. Symbols represent the
+relative abundance data for an individual mouse, circles represent mice
+that cleared \emph{C. difficile} by day 7, X-shapes represent mice that
+were still colonized with \emph{C. difficile}, and open circles
+represent mice that did not have \emph{C. difficile} CFU counts for day
+7 post-infection. Colored lines indicate the median relative abundances
+for each source. The gray horizontal line represents the limit of
+detection. Timepoints where differences across sources of mice were
+statistically significant by Kruskal-Wallis test with Benjamini-Hochberg
+correction for testing across multiple days (Table S16) are identified
+by the asterisk(s) above each timepoint (*, P \textless{} 0.05).
 
@@
 \includegraphics{figure_S1.pdf} \textbf{Figure S1. \emph{C. difficile}
-CFU variation across vendors is similar between the 2 experiments.}
-(A-B). \emph{C. difficile} CFU/gram of stool quantification over time
-for experiment 1 (A) and 2 (B). Experiments were conducted approximately
-3 months apart. Lines represent the mean CFU for each source and circles
+CFU variation across vendors varies slightly across the 2 experiments.}
+A-B. \emph{C. difficile} CFU/gram of stool quantification over time for
+experiment 1 (A) and 2 (B). Experiments were conducted approximately 3
+months apart. Lines represent the median CFU for each source, symbols
 represent individual mice and the black line represents the limit of
-detection.
+detection. C. Percent of mice that were colonized with \emph{C.
+difficile} over the course of the experiment. Each day the percent is
+calculated based on the mice where \emph{C. difficile} CFU was
+quantified for that particular day. Total N for each day: day 1 (N =
+42), day 2 (N = 20), day 3 (N = 39), day 4 (N = 29), day 5 (N = 43), day
+6 (N = 34), day 7 (N = 40), day 8 (N = 36), and day 9 (N = 46).
 
@@
 R\textsuperscript{2} = 0.99; \emph{P} \(\le\) 0.033) and Young (combined
-R\textsuperscript{2} = 0.95; \emph{P} \(\le\) 0.029) lab colonies (Table
+R\textsuperscript{2} = 0.95; \emph{P} \(\le\) 0.027) lab colonies (Table
 S7).
@@
 
-\includegraphics{figure_S3.pdf} \textbf{Figure S3. \emph{C. difficile}
-colonization status 7 days post-infection varies across mice.} A. Bar
-graph visualizations of overall day 7 C. difficile colonization status
-that were used as classification outcomes to build logistic regression
-models. Mice were classified as colonized or cleared (not detectable at
-the limit of detection of 100 CFU) based on CFU g/stool data from 7 days
-post-infection. B-D. Classification input data used for models based on
-day -1 (B), day 0 (C), and day 1 (D) community relative abundance data.
-Only mice with day 7 \emph{C. difficile} status and community relative
-abundance data for each respective timepoint were used as input data for
-the classification models.
-
-\newpage
-
-\includegraphics{figure_S4.pdf} \textbf{Figure S4. Key taxa from
-classification model based on baseline community data vary by mouse
-colony source, clindamycin treatment, or both.} A-B. Venn diagrams of
-top 20 important OTUs (A) and families (B) from day -1 classification
-models (Table S14-15) that overlapped with taxa that varied across
-vendors at baseline, were impacted by clindamycin treatment, or both.
-Bold taxa signify bacteria that were important to more than 1
+\includegraphics{figure_S3.pdf} \textbf{Figure S3. Bacterial community
+composition before, after clindamycin perturbation, and post-infection
+can predict \emph{C. difficile} colonization status 7 days
+post-challenge.} A. Bar graph visualizations of overall day 7 C.
+difficile colonization status that were used as classification outcomes
+to build logistic regression models. Mice were classified as colonized
+or cleared (not detectable at the limit of detection of 100 CFU) based
+on CFU g/stool data from 7 days post-infection. B. \emph{C. difficile}
+CFU status on Day 7 within each mouse colony source. N = 5-9 mice per
+group. C-D. L2-regularized logistic regression classification model
+AUROCs to predict \emph{C. difficile} CFU on D7 (Fig. 1D, Fig. S3) based
+on the community relative abundances at baseline (day -1),
+post-clindamycin (day 0), and post-infection (day 1) at either the OTU
+(C) or family (D) level. All models performed better than random chance
+(AUROC = 0.5), see (all \emph{P} \(\le\) 5e-15; Table S12) and the model
+built with post-clindamycin treated bacterial OTU relative abundances
+had the best performance ((\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10
+for pairwise comparisons; Table S13). A List of the 20 taxa that were
+ranked as most important to each model are listed in Table S14-15.
+
+\newpage
+
+\includegraphics{figure_S4.pdf} \textbf{Figure S4. Key OTUs from
+classification models based on baseline, post-clindamycin treatment, or
+post-infection community data vary by mouse colony source, clindamycin
+treatment, or both.} A-C. Venn diagrams of top 20 important OTUs from
+baseline (A), post-clindamycin treatment (B), and post-infection (C)
+classification models (Table S14) that overlapped with OTUs that varied
+across vendors at baseline, were impacted by clindamycin treatment, or
+both. Bold OTUs signify OTUs that were important to more than 1
 classification model.
@@
 
-\includegraphics{figure_S5.pdf} \textbf{Figure S5. Key taxa from
-classification model based on post-clindamycin treatment community data
-vary by mouse colony source, clindamycin treatment, or both.} A-B. Venn
-diagrams of top 20 important OTUs (A) and families (B) from day 0
-classification models (Table S14-15) that overlapped with taxa that
-still varied across vendors after clindamycin, were impacted by
-clindamycin treatment, or both. Bold taxa signify bacteria that were
-important to more than 1 classification model.
-
-\newpage
-
-\includegraphics{figure_S6.pdf} \textbf{Figure S6. Key taxa from
-classification model based on post-infection community data vary by
-mouse colony source, clindamycin treatment, or both.} A-B. Venn diagrams
-of top 20 important OTUs (A) and families (B) from day 1 classification
-models (Table S14-15) that overlapped with taxa that varied across
-vendors post-infection, were impacted by clindamycin treatment, or both.
-Bold taxa signify bacteria that were important to more than 1
-classification model.
-
-\newpage
-
-\includegraphics{figure_S7.pdf} \textbf{Figure S7. Key families vary
+\includegraphics{figure_S5.pdf} \textbf{Figure S5. Key families from
+classification models based on baseline, post-clindamycin treatment, or
+post-infection community data vary by mouse colony source, clindamycin
+treatment, or both.} A-C. Venn diagrams of top 20 important families
+from baseline (A), post-clindamycin treatment (B), and post-infection
+(C) classification models (Table S15) that overlapped with families that
+varied across vendors after clindamycin, were impacted by clindamycin
+treatment, or both. D. Venn diagrams that combines A-C summaries of
+familes that were important to the day -1, 0, and 1 classification
+models (Table S15) and either overlapped with familes that varied across
+vendors at the same timepoint, were impacted by clindamycin treatment,
+or both. Bold families signify families that were important to more than
+1 classification model.
+
+\newpage
+
+\includegraphics{figure_S6.pdf} \textbf{Figure S6. Key families vary
 across sources throughout experiment.} Relative abundances of bold
-families from Fig. 7B that were important for at least two
+families from Fig. S5D that were important for at least two
 classification models are shown over time. A. \emph{Enterococcaceae} and
@@
 where relative abundances also significantly varied across sources of
-mice. Each point represents the relative abundance of an individual
-mouse stool sample and colored lines represent the mean relative
-abundances for each source of mice. The gray horizontal line represents
-the limit of detection. Timepoints where differences across sources of
-mice were statistically significant by Kruskal-Wallis test with
-Benjamini-Hochberg correction for testing across multiple days (Table
-S17) are identified by the asterisk(s) above each timepoint (*, P
-\textless{} 0.05, **, P \textless{} 0.01, ***, P \textless{} 0.001).
+mice. Symbols represent the relative abundance data for an individual
+mouse, circles represent mice that cleared \emph{C. difficile} by day 7,
+X-shapes represent mice that were still colonized with \emph{C.
+difficile}, and open circles represent mice that did not have \emph{C.
+difficile} CFU counts for day 7 post-infection. Colored lines indicate
+the median relative abundances for each source. The gray horizontal line
+represents the limit of detection. Timepoints where differences across
+sources of mice were statistically significant by Kruskal-Wallis test
+with Benjamini-Hochberg correction for testing across multiple days
+(Table S17) are identified by the asterisk(s) above each timepoint (*, P
+\textless{} 0.05).
 
@@
 sample from an individual mouse at a specific timepoint. See Table S5
-for PERMANOVA results).
+for PERMANOVA results). Circles represent mice from experiment 1 and
+triangles represent mice from expeirment 2.
 
@@
 changed after clindamycin treatment.} \textbf{Table S12. Statistical
-results of logistic regression model performances compared to random
-chance.}
-
-\textbf{Table S13. Pairwise Wilcoxan results of comparing all 6 logistic
-regression model performances.}
-
-\textbf{Table S14. Top 20 most important OTUs for each of the 3 logistic
-regression models based on OTU relative abundance data.}
+results of L2-regularized logistic regression model performances
+compared to random chance.}
+
+\textbf{Table S13. Pairwise Wilcoxan results of comparing all 6
+L2-regularized logistic regression model performances.}
+
+\textbf{Table S14. Top 20 most important OTUs for each of the 3
+L2-regularized logistic regression models based on OTU relative
+abundance data.}
 
 \textbf{Table S15. Top 20 most important families for each of the 3
-logistic regression models based on OTU relative abundance data.}
+L2-regularized logistic regression models based on OTU relative
+abundance data.}
 

### 2020-06-26 code/taxa_stats.R
For Fig. 5 and 6: change boxplots to stat summary plots of median and interquartile range

@@ -128,13 +128,10 @@ plot_families_dx <- function(families, timepoint){
                       breaks=color_vendors,
                       labels=color_vendors)+
   geom_hline(yintercept=1/5437, color="gray")+
-  geom_boxplot(outlier.shape = NA, size = 1.2)+  
-#  geom_errorbar(aes(ymax = median, ymin = median))+ #Add lines to indicate the median for each group to the plot. Median calculated before y axis transformation
-#  geom_jitter(aes(shape = experiment), size=2, alpha=0.6, position=position_dodge(1.6)) +
-#  scale_shape_manual(name=NULL,
-#                     values=shape_scheme,
-#                     breaks=shape_experiment,
-#                     labels=shape_experiment) +
+  stat_summary(fun = 'median', 
+               fun.max = function(x) quantile(x, 0.75), 
+               fun.min = function(x) quantile(x, 0.25),
+               position = position_dodge(width = 1)) +  
   labs(title=NULL, 
        x=NULL,
        y="Relative abundance (%)")+
@@ -337,8 +334,10 @@ plot_otus_dx <- function(otus, timepoint){
                         breaks=color_vendors,
                         labels=color_vendors)+
     geom_hline(yintercept=1/5437, color="gray")+
-    geom_boxplot(outlier.shape = NA, size = 1.2)+
-    #  geom_jitter(shape=19, size=2, alpha=0.6, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.2)) + #Too busy when indiv. mice are shown
+    stat_summary(fun = 'median', 
+                 fun.max = function(x) quantile(x, 0.75), 
+                 fun.min = function(x) quantile(x, 0.25),
+                 position = position_dodge(width = 1)) +
     labs(title=NULL, 
          x=NULL,
          y="Relative abundance (%)")+
@@ -543,9 +542,11 @@ clind_impacted_families_plot_dx <- function(timepoint){
                         breaks=color_vendors,
                         labels=color_vendors)+
     geom_hline(yintercept=1/5437, color="gray")+
-    geom_boxplot(outlier.shape = NA, size = 1.2)+
-    #  geom_jitter(shape=19, size=2, alpha=0.6, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.2)) +
-    labs(title=NULL, 
+    stat_summary(fun = 'median', 
[…]

### 2020-06-26 code/figure_1.R
Adjust figure arrangement and sizing

@@ -3,12 +3,10 @@ source("code/functions.R")
 a <- ggdraw() + draw_image("results/figures/Vendor_exp_scheme.png")
 b <- ggdraw() + draw_image("results/figures/cfu_over_time.png")
 c <- ggdraw() + draw_image("results/figures/weight_over_time.png")
-d <- ggdraw() + draw_image("results/figures/C.diff_CFU_D7_stats.png")
-e <- ggdraw() + draw_image("results/figures/weight_D2.png")
-bottom_row <- plot_grid(d, e, labels = c("D", "E"), label_size = 12)
 
-plot_grid(a, b, c, bottom_row, labels = c("A", "B", "C", ""), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_1.pdf", width=3.875, height=8.5)+
-  ggsave("submission/figure_1.pdf", width=3.875, height=8.5)
+
+plot_grid(a, b, c, labels = c("A", "B", "C", ""), label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_1.pdf", width=5, height=7.5)+
+  ggsave("submission/figure_1.pdf", width=5, height=7.5)
 
 

### 2020-06-26 code/figure_3.R

@@ -4,8 +4,8 @@ a <- ggdraw() + draw_image("results/figures/pcoa_day-1.png")
 b <- ggdraw() + draw_image("results/figures/pcoa_day0.png")
 c <- ggdraw() + draw_image("results/figures/pcoa_day1.png")
 
-plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_3.pdf", width=5, height=10)+
-  ggsave("submission/figure_3.pdf", width=5, height=10)
+plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_3.pdf", width=8, height=7)+
+  ggsave("submission/figure_3.pdf", width=8, height=7)
 
 

### 2020-06-26 code/figure_7.R

@@ -8,6 +8,6 @@ top_half <- plot_grid(a, labels = c("A"), label_size = 12, ncol=1)
 b <- ggdraw() + draw_image("results/figures/venn_overall_otus.png")
 
 plot_grid(top_half, b, labels = c("", "B"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_7.pdf", width=4, height=8)+
-  ggsave("submission/figure_7.pdf", width=4, height=8)
+  ggsave("results/figures/figure_7.pdf", width=6, height=8)+
+  ggsave("submission/figure_7.pdf", width=6, height=8)
 

### 2020-06-26 code/figure_8.R

@@ -6,6 +6,6 @@ c <- ggdraw() + draw_image("results/figures/Enterobacteriaceae (OTU 1)_time.png"
 d <- ggdraw() + draw_image("results/figures/Porphyromonadaceae (OTU 7)_time.png")
 
 
-plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_8.pdf", width=3, height=7)+
-  ggsave("submission/figure_8.pdf", width=3, height=7)
+plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_8.pdf", width=6, height=4)+
+  ggsave("submission/figure_8.pdf", width=6, height=4)

### 2020-06-26 code/figure_S1.R

@@ -2,10 +2,14 @@ source("code/functions.R")
 
 a <- ggdraw() + draw_image("results/figures/exp1_cfu_time.png")
 b <- ggdraw() + draw_image("results/figures/exp2_cfu_time.png")
-c <- ggdraw() + draw_image("results/figures/C.diff_percent_colonized.png")
+c <- ggdraw() + draw_image("results/figures/C.diff_CFU_D7_stats.png")
+d <- ggdraw() + draw_image("results/figures/weight_D2.png")
+e <- ggdraw() + draw_image("results/figures/C.diff_percent_colonized.png")
+right_panel <- plot_grid(c, d, e, labels = c("C", "D", "E"), label_size = 12, nrow = 1)
 
-plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_S1.pdf", width=8, height=5.5)+
-  ggsave("submission/figure_S1.pdf", width=8, height=5.5)
+
+plot_grid(a, right_panel, b,  labels = c("A", "", "B"), label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_S1.pdf", width=8, height=4.5)+
+  ggsave("submission/figure_S1.pdf", width=8, height=4.5)
 
 

### 2020-06-26 code/figure_S2.R

@@ -8,7 +8,7 @@ e <- ggdraw() + draw_image("results/figures/pcoa_dn1_taconic.png")
 f <- ggdraw() + draw_image("results/figures/pcoa_dn1_envigo.png")
 
 plot_grid(a, b, c, d, e, f, labels = "AUTO", label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_S2.pdf", width=3, height=4)+
-  ggsave("submission/figure_S2.pdf", width=3, height=4)
+  ggsave("results/figures/figure_S2.pdf", width=6, height=8)+
+  ggsave("submission/figure_S2.pdf", width=6, height=8)
 
 

### 2020-06-26 code/figure_S5.R

@@ -5,6 +5,6 @@ b <- ggdraw() + draw_image("results/figures/venn_d0_families.png")
 c <- ggdraw() + draw_image("results/figures/venn_d1_families.png")
 d <- ggdraw() + draw_image("results/figures/venn_overall_families.png")
 
-plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_S5.pdf", width=3, height=9.5)+
-  ggsave("submission/figure_S5.pdf", width=3, height=9.5)
+plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_S5.pdf", width=7.5, height=4.5)+
+  ggsave("submission/figure_S5.pdf", width=7.5, height=4.5)

### 2020-06-26 submission/manuscript.Rmd
Additional edits and incorporate Josh's suggestion

@@
 While all sources of mice were colonized 1-day post-infection, variation in *C.
-difficile* colonization levels emerged from days 3-7 post-infection with 3 sources colonized with *C.
+difficile* colonization levels emerged from days 3-7 post-infection with some sources colonized with *C.
 difficile* for slightly longer and at higher levels.
@@
 Microbiome differences between different mouse vendors have been well documented and shown to influence susceptibility to a variety of diseases [@Stough2016; @Alegre2019], including enteric infections [@EtienneMesmin2017; @Velazquez2019; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018].
-Additionally, different research groups have observed different CDI outcomes in mice despite using similar models and the microbiome has been proposed as one factor potentially mediating susceptibility [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014].
+Additionally, different research groups have observed different CDI outcomes in mice despite using similar models and the microbiome has been proposed as one factor potentially mediating CDI susceptibility and outcomes [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014].
 Here we examined how variations in the baseline microbiome and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C.
@@
 We conducted two experiments approximately 3 months apart and while the colonization dynamics were similar across most sources of mice, there was some variation between the 2 experiments, particularly for the Schloss and Envigo mice (Fig.
-S1A-B, Fig.
-1D).
+S1A-B).
 Although *C.
@@
 difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig.
-1C, E and Table S2).
+1C and Table S2).
+Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment.
 Interestingly, mice ordered from Jackson, Taconic, and Envigo tended to lose more weight (although there was variation between experiments with Schloss and Envigo mice), have higher *C.
 difficile* CFU levels and take longer to clear the infection compared to the other sources of mice, which was particularly evident 7 days post-infection (Fig.
-1B-E), when `r percent_colonized_d7` of the mice were still colonized with *C.
+1B-C, Fig.
+S2C-D), when `r format(percent_colonized_d7, digits = 1)`% of the mice were still colonized with *C.
 difficile* (Fig.
-S1C).
+S1E).
 By 9 days post-infection the majority of the mice from all sources had cleared *C.
@@
 S1C) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment.
-Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment.
 Thus, clindamycin rendered all mice susceptible to *C.
@@
 Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after both clindamycin and *C.
-difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other groups (Fig 2C-F and Table S4).
+difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other sources (Fig 2C-F and Table S4).
 While Charles River mice had more diverse microbiotas, Young and Schloss lab mice were also able to clear *C.
@@
 We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig.
-S2 and Table S7), although most of the vendors also clustered by experiment, suggesting there was some community variation between the 2 experiments within each vendor.
+S2A-B and Table S7).
+However, most of the vendors also clustered by experiment (Fig.
+S2C-D, F), suggesting there was some community variation between the 2 experiments within each source.
 Thus, mouse colony source was the factor that explained the most variation observed in the bacterial communities.
@@
 4A).
-Clindamycin treatment reduced variation between experiments within Schloss, Young and Jackson mice and some of the variation between sources diminished (Fig.
+Clindamycin treatment reduced variation between experiments within Schloss, Young and Jackson mice and some of the variation between sources diminished, particularly for the University of Michigan and Charles River mice (Fig.
 4B).
-Post-infection, the community variation started to increase within sources of mice and variation between sources of mice started to return (Fig.
+Post-infection, the community variation started to increase within sources of mice and variation between sources of mice started to increase towards previous levels (Fig.
 4C).
@@
 We examined bacterial relative abundances at the operational taxonomic unit (OTU) and family levels, expecting the number of differences to be reduced at the family level due to the nature of bacterial taxonomy [@Nguyen2015].
-Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs and `r source_families_dn1` families (Table S8-9) with relative abundances that varied across colony sources.
-Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs and `r source_families_d0` families (Table S8-9).
+Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs and `r source_families_dn1` families (Tables S8-9) with relative abundances that varied across colony sources.
+Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs and `r source_families_d0` families (Tables S8-9).
 After *C.
-difficile* challenge, there were `r source_otus_d1` OTUs and `r source_families_d1` families (Table S8-9) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment.
+difficile* challenge, there were `r source_otus_d1` OTUs and `r source_families_d1` families (Tables S8-9) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment.
 In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 5A-C) and `r shared_source_families_dn1to1` families with relative abundances that consistently varied across colony sources (Fig.
@@
 By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs and `r clindamycin_families_sig` families that were altered after clindamycin treatment (Fig.
-6 and Table S10-11).
+6 and Tables S10-11).
 Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) and 3 families (*Porphyromonadaceae*, *Enterococcaceae*, *Lachnospiraceae*) overlapped (Fig.
@@
 All models were better at predicting *C.
-difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12), however the models trained with OTU level data generally performed better than those trained with family level data with the expception of the models based on the post-infection (day 1) communities (Fig.
+difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12).
+However, the models trained with OTU level data generally performed better than those trained with family level data with the expception of the models based on the post-infection (day 1) communities (Fig.
 S3C-D).
-Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than all other models with an AUROC of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13).
+Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than all other models with an area under the receiving operator characteristic curve (AUROC) of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13).
 Thus, we were able to use community bacterial relative abundance data alone to differentiate mice that had cleared *C.
 difficile* before day 7 from the mice still colonized with *C.
-difficile*.
+difficile* at that timepoint.
 Interestingly, the model built with OTU relative abundance data post-clindamycin treatment had the best performance, suggesting how the bacterial community responds to clindamycin treatment has the greatest influence on subsequent *C.
@@
 
-Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 taxa that had the highest absolute feature weights in each of the 6 models (Table S14-15).
+Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 taxa that had the highest absolute feature weights in each of the 6 models (Tables S14-15).
 First, we looked at OTUs from the model with the best performance that was based on the post-clindamycin treatment bacterial community data.
@@
 7A).
-Next, the top 20 taxa from each model were compared to the list of taxa that varied across colony source (Fig.
-5 and Table S8-9) at the same timepoint and the taxa that were altered by clindamycin treatment (Fig.
+Next, the top 20 taxa from each model were compared to the list of taxa that varied across colony source at the same timepoint (Fig.
+5 and Tables S8-9) and the taxa that were altered by clindamycin treatment (Fig.
 6 and Table S10-11).
@@
 difficile* 630, suggesting multiple bacteria influence the time needed to clear the infection.
-Together, these results suggest the initial bacterial communities and their responses to clindamycin have a large influence on the time needed to clear *C.
+Together, these results suggest the initial bacterial communities and their responses to clindamycin influence on the time needed to clear *C.
 difficile*.
@@
 difficile* colonization dynamics within mice from 6 different colony sources after perturbing the microbiota with clindamycin treatment, we were able to identify bacterial taxa that were unique to sources throughout the experiment as well as taxa that were universally impacted by clindamycin.
-We built L2 logistic regression models with baseline, post-clindamycin treatment, and post-infection fecal community data that successfully predicted *C.
-difficile* colonization status 7 days after infection better than random chance.
+We built L2 logistic regression models with baseline, post-clindamycin treatment, and post-infection fecal community data that successfully predicted whether mice cleared *C.
+difficile* by 7 days post-infection better than random chance.
 We identified *`r list_key_otus`*, *`r list_key_families`* (Fig.
@@
 Although all mice were susceptible to *C.
-difficile* colonization, by following colonization over time we found Jackson, Taconic, and Envigo mice remained colonized beyond 7 days post-infection.
+difficile* colonization, by following colonization over time we found Jackson, Taconic, and Envigo mice tended to remain colonized through at least 7 days post-infection.
 We identified a subset of bacteria that were important in predicting whether a mouse was still colonized with *C.
@@
 difficile* strain used could also be contributing to the variation in *C.
-difficile* outcomes seen across different groups [@Lawley2012].
+difficile* outcomes seen across different research groups [@Lawley2012].
 We found the time needed to naturally clear *C.
@@
 Additionally with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of the CDI mouse model studies to date associated mice with just 1 types of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018], which does not aid in the goal of figuring out how a variety of unique microbiotas influence susceptiblity to CDIs and adverse outcomes.
+Importantly, our study using mice from 6 different sources increased the variation between groups of mice compared to using 1 source alone, to better reflect the inter-individual microbiota variation observed in humans.
 Encouragingly, decreased *Bifidobacterium*, *Porphyromonas*, *Ruminococcaceae* and *Lachnospiraceae* and increased *Enterobacteriaceae*, *Enterococcus*, *Lactobacillus*, and *Proteus* have all been associated with human CDIs [@Mancabelli2017, @Duvallet2017] and were well represented in our study, suggesting most of the mouse sources are suitable for gaining insights into microbiota associated factors influencing *C.
@@
 An important exception was *Enterococcus*, which was primarily absent from the mice from University of Michigan colonies and *Proteus*, which was only found in Taconic mice.
-Importantly, the fact that some CDI associated bacteria were only found in a subset of mice has important implications for future CDI mouse model studies.
+Importantly, the fact that some CDI-associated bacteria were only found in a subset of mice has important implications for future CDI mouse model studies.
 
@@
 difficile* colonization status and metagenomes from these bacteria have been shown to contain the bile acid-inducible gene cluster necessary for secondary bile acid formation and ability to produce butyrate [@Vital2019; @Collins2015].
-Interestingly, butyrate has previously been shown to vary across vendors and mediates resistance to *Citrobacter rodentium* infection, a model of enterohemorrhagic and enteropathogenic *Escherichia coli* infections [@Osbelt2020].
+Interestingly, butyrate has previously been shown to vary across vendors and mediated resistance to *Citrobacter rodentium* infection, a model of enterohemorrhagic and enteropathogenic *Escherichia coli* infections [@Osbelt2020].
 Evidence for immunological toning differences in IgA and Th17 cells across mice from different vendors have also been documented and [@Fransen2015; @Ivanov2009] may also influence response to CDI, particularly in the context of severe CDIs [@Azrad2018; @Saleh2019].
@@
 difficile* exposure depends on a multitude of factors, including age, diet, and immunity; all of which are also influenced by the microbiota.
-We have demonstrated that the ways different baseline microbiotas from different mouse colony sources respond to clindamycin treatment influences the length of time mice remained colonized with *C.
+We have demonstrated that the ways baseline microbiotas from different mouse colony sources respond to clindamycin treatment influences the length of time mice remained colonized with *C.
 difficile* 630.
@@
 We thank members of the Schloss lab for feedback on planning the experiments and data presentation, as well as code tutorials and feedback through Code club.
-In particular, we want to thank Begüm Topçuoğlu for help with implementing L2 logistic regression models using her [pipeline](https://github.com/SchlossLab/ML_pipeline_microbiome), Ana Taylor for help with some media preparation and sample collection, and Nicholas Lesniak for his critical feedback on the manuscript.
+In particular, we want to thank Begüm Topçuoğlu for help with implementing L2 logistic regression models using her [pipeline](https://github.com/SchlossLab/ML_pipeline_microbiome), Ana Taylor for help with some of the media preparation and sample collection, and Nicholas Lesniak for his critical feedback on the manuscript.
 We also thank members of Vincent Young’s lab, particularly  Kimberly Vendrov, for guidance with the *C.
@@
 The L2-regularized logistic regression models were trained and tested using the caret package [@Kuhn2008] in R as previously described [@Topcuoglu2020] with the exception that we used 60% training and 40% testing data splits for the cross-validation of the training data to select the best cost hyperparameter and the testing of the held out test data to measure model performance.
-The modified training/testing ratio was selected to accommodate the small number of samples in the dataset.
+The modified training to testing ratio was selected to accommodate the small number of samples in the dataset.
 Code was modified from https://github.com/SchlossLab/ML_pipeline_microbiome to update the classification outcomes and change the data split ratios.
@@
 Lines represent the median for each source and circles represent individual mice from experiment 1 while triangles represent mice from experiment 2.
-D.
-*C.
-difficile* CFU/gram stool on day 7 post-infection across sources of mice with asterisks for pairwise Wilcoxon comparisons with Benjamini-Hochberg correction where *P* < 0.05.
-E.
-Mouse weight change 2 days post-infection across sources of mice, no pairwise Wilcoxon comparisons were significant after Benjamini-Hochberg correction.
-For D-E.
-Circles represent experiment 1 mice, triangles represent experiment 2 mice and gray lines indicate the median values for each group.
 
@@
 A subset of bacteria consistently vary across mouse colony sources despite clindamycin perturbation and *C.
-difficile* challenge.** A-C.
-Boxplots of the relative abundances for the `r shared_source_otus_dn1to1` OTUs that consistently varied across sources of mice at the baseline (A), post-clindamycin (B), and post-infection (C) timepoints of the experiment.
-D-F.
-Boxplots of the relative abundances for the `r shared_source_families_dn1to1` families that consistently varied across sources of mice at the baseline (D), post-clindamycin (E), and post-infection (F) timepoints of the experiment.
+difficile* challenge.** A-C: plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the `r shared_source_otus_dn1to1` OTUs that consistently varied across sources of mice at the baseline (A), post-clindamycin (B), and post-infection (C) timepoints of the experiment.
+D-F: plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the `r shared_source_families_dn1to1` families that consistently varied across sources of mice at the baseline (D), post-clindamycin (E), and post-infection (F) timepoints of the experiment.
 For each timepoint bacteria with differential relative abundances across sources of mice were identified by Kruskal-Wallis test at the family and OTU level with Benjamini-Hochberg correction for testing all identified taxa at the respective level (Table S8-9).
@@
 **Figure 6.
-Clindamycin treatment has the same effects on a subset of taxa regardless of colony source.** A-B.
-Boxplots of the top 10 most significant (adjusted *P* value < 0.05) OTUs with relative abundances that changed post clindamycin treatment.
-C-D.
-Boxplots of the top 10 most significant families with relative abundances that changed post clindamycin treatment.
+Clindamycin treatment has the same effects on a subset of taxa regardless of colony source.** A-B: plots highlighting the median (point) and interquantile range (colored lines) of the top 10 most significant (adjusted *P* value < 0.05) OTUs with relative abundances that changed after clindamycin treatment.
+C-D: plots highlighting the median (point) and interquantile range (colored lines) of the top 10 most significant families with relative abundances that changed after clindamycin treatment.
 Data were analyzed by Wilcoxon signed rank test limited to mice that had paired sequence data for day -1 and 0 (N = `r n_paired_for_clind`).
@@
 difficile* by day 7.** A.
-Baseline relative abundance data for 3 of the OTUs from the classification model based on day 0 OTU relative abundances that significantly varied across sources of mice and had high relative abundances.
+Baseline relative abundance data for 3 of the OTUs from the classification model based on day 0 OTU relative abundances that significantly varied across sources of mice and had high relative abundances in the community.
 Symbols represent the relative abundance data for an individual mouse, circles represent mice that cleared *C.
@@
 C.
+*C.
+difficile* CFU/gram stool on day 7 post-infection across sources of mice with asterisks for pairwise Wilcoxon comparisons with Benjamini-Hochberg correction where *P* < 0.05.
+D.
+Mouse weight change 2 days post-infection across sources of mice, no pairwise Wilcoxon comparisons were significant after Benjamini-Hochberg correction.
+For C-D: circles represent experiment 1 mice, triangles represent experiment 2 mice and gray lines indicate the median values for each group.
+E.
 Percent of mice that were colonized with *C.
@@
 Bar graph visualizations of overall day 7 C.
-difficile colonization status that were used as classification outcomes to build logistic regression models.
+difficile colonization status that were used as classification outcomes to build L2-regularized logistic regression models.
 Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU) based on CFU g/stool data from 7 days post-infection.
@@
 C-D.
-L2-regularized logistic regression classification model AUROCs to predict *C.
-difficile* CFU on D7 (Fig.
+L2-regularized logistic regression classification model area under the receiving operator characteristic curve (AUROCs) to predict *C.
+difficile* CFU on day 7 post-infectoin (Fig.
 1D, Fig.
 S3) based on the community relative abundances at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1) at either the OTU (C) or family (D) level.
-All models performed better than random chance (AUROC = 0.5), see (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13).
-A List of the 20 taxa that were ranked as most important to each model are listed in Table S14-15.
+All models performed better than random chance (AUROC = 0.5; all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for all pairwise comparisons; Table S13).
+For lists of the 20 taxa that were ranked as most important to each model, see Tables S14-15.
 
@@
 
-**Table S6. PERMANOVA results for all mice, all timepoints.**
+**Table S6. PERMANOVA results for all mice at baseline, post clindamycin, and post-infection timepoints.**
 
@@
 **Table S11. Families with relative abudances that significantly changed after clindamycin treatment.**
+
 **Table S12. Statistical results of L2-regularized logistic regression model performances compared to random chance.**
 
-**Table S13. Pairwise Wilcoxan results of comparing all 6 L2-regularized logistic regression model performances.**
+**Table S13. Pairwise Wilcoxan results comparing all 6 L2-regularized logistic regression model performances.**
 
@@
 
-**Table S16. OTUs with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment.**
+**Table S16. OTUs with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**
 
-**Table S17. Families with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment.**
+**Table S17. Families with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**

### 2020-06-26 submission/manuscript.tex
Additional edits and incorporate Josh's suggestion

@@
 \emph{C. difficile} colonization levels emerged from days 3-7
-post-infection with 3 sources colonized with \emph{C. difficile} for
+post-infection with some sources colonized with \emph{C. difficile} for
 slightly longer and at higher levels. We identified bacterial taxa with
@@
 using similar models and the microbiome has been proposed as one factor
-potentially mediating susceptibility (13, 18, 21, 31--33). Here we
-examined how variations in the baseline microbiome and responses to
-clindamycin treatment in C57BL/6 mice from six different sources
-influenced susceptibility to \emph{C. difficile} colonization and the
-time needed to clear the infection.
+potentially mediating CDI susceptibility and outcomes (13, 18, 21,
+31--33). Here we examined how variations in the baseline microbiome and
+responses to clindamycin treatment in C57BL/6 mice from six different
+sources influenced susceptibility to \emph{C. difficile} colonization
+and the time needed to clear the infection.
 
@@
 mice, there was some variation between the 2 experiments, particularly
-for the Schloss and Envigo mice (Fig. S1A-B, Fig. 1D). Although \emph{C.
+for the Schloss and Envigo mice (Fig. S1A-B). Although \emph{C.
 difficile} 630 causes mild symptoms in mice comparied to other \emph{C.
@@
 varied across sources of mice with the most weight lost two days
-post-infection (Fig. 1C, E and Table S2). Interestingly, mice ordered
+post-infection (Fig. 1C and Table S2). Importantly, there was also one
+Jackson and one Envigo mouse that died between 1- and 3-days
+post-infection during the second experiment. Interestingly, mice ordered
 from Jackson, Taconic, and Envigo tended to lose more weight (although
@@
 infection compared to the other sources of mice, which was particularly
-evident 7 days post-infection (Fig. 1B-E), when 57.50 of the mice were
-still colonized with \emph{C. difficile} (Fig. S1C). By 9 days
+evident 7 days post-infection (Fig. 1B-C, Fig. S2C-D), when 57\% of the
+mice were still colonized with \emph{C. difficile} (Fig. S1E). By 9 days
 post-infection the majority of the mice from all sources had cleared
@@
 from the first experiment and 2 Envigo mice from the second experiment.
-Importantly, there was also one Jackson and one Envigo mouse that died
-between 1- and 3-days post-infection during the second experiment. Thus,
-clindamycin rendered all mice susceptible to \emph{C. difficile} 630
-colonization, regardless of colony source, but variation across sources
-emerged with 3 out of 6 sources taking longer to clear \emph{C.
+Thus, clindamycin rendered all mice susceptible to \emph{C. difficile}
+630 colonization, regardless of colony source, but variation across
+sources emerged with 3 out of 6 sources taking longer to clear \emph{C.
 difficile}.
@@
 having higher richness and Shannon Diversity than most of the other
-groups (Fig 2C-F and Table S4). While Charles River mice had more
+sources (Fig 2C-F and Table S4). While Charles River mice had more
 diverse microbiotas, Young and Schloss lab mice were also able to clear
@@
 cage significantly explained the observed variation only for the Schloss
-and Young lab mouse colonies (Fig. S2 and Table S7), although most of
-the vendors also clustered by experiment, suggesting there was some
-community variation between the 2 experiments within each vendor. Thus,
-mouse colony source was the factor that explained the most variation
-observed in the bacterial communities. Importantly with the exception of
-the 2 University colonies, the community of each source clustered apart
-from one another suggesting each community had a unique response to
-clindamycin treatment and \emph{C. difficile} challenge.
+and Young lab mouse colonies (Fig. S2A-B and Table S7). However, most of
+the vendors also clustered by experiment (Fig. S2C-D, F), suggesting
+there was some community variation between the 2 experiments within each
+source. Thus, mouse colony source was the factor that explained the most
+variation observed in the bacterial communities. Importantly with the
+exception of the 2 University colonies, the community of each source
+clustered apart from one another suggesting each community had a unique
+response to clindamycin treatment and \emph{C. difficile} challenge.
 
@@
 Schloss, Young and Jackson mice and some of the variation between
-sources diminished (Fig. 4B). Post-infection, the community variation
+sources diminished, particularly for the University of Michigan and
+Charles River mice (Fig. 4B). Post-infection, the community variation
 started to increase within sources of mice and variation between sources
-of mice started to return (Fig. 4C). By using mice from multiple sources
-we were able to increase the number of microbiota communities we tested
-with the clindamycin \emph{C. difficile} colonization mouse model.
+of mice started to increase towards previous levels (Fig. 4C). By using
+mice from multiple sources we were able to increase the number of
+microbiota communities we tested with the clindamycin \emph{C.
+difficile} colonization mouse model.
 
@@
 bacterial taxonomy (45). Focusing on the baseline communities first,
-there were 268 OTUs and 20 families (Table S8-9) with relative
+there were 268 OTUs and 20 families (Tables S8-9) with relative
 abundances that varied across colony sources. Clindamycin treatment
 reduced the number of taxa with relative abundances that varied across
-sources to 18 OTUs and 10 families (Table S8-9). After \emph{C.
-difficile} challenge, there were 44 OTUs and 18 families (Table S8-9)
+sources to 18 OTUs and 10 families (Tables S8-9). After \emph{C.
+difficile} challenge, there were 44 OTUs and 18 families (Tables S8-9)
 with significantly different relative abundances across sources, as the
@@
 OTUs and 18 families that were altered after clindamycin treatment (Fig.
-6 and Table S10-11). Interestingly, when we compared the list of
+6 and Tables S10-11). Interestingly, when we compared the list of
 significant clindamycin impacted bacteria with the bacteria that
@@
 better at predicting \emph{C. difficile} colonization status on day 7
-than random chance (all \emph{P} \(\le\) 5e-15; Table S12), however the
+than random chance (all \emph{P} \(\le\) 5e-15; Table S12). However, the
 models trained with OTU level data generally performed better than those
@@
 the model based on the post-clindamycin (day 0) community OTU data
-performed significantly better than all other models with an AUROC of
-0.75 (\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise
-comparisons; Table S13). Thus, we were able to use community bacterial
-relative abundance data alone to differentiate mice that had cleared
-\emph{C. difficile} before day 7 from the mice still colonized with
-\emph{C. difficile}. Interestingly, the model built with OTU relative
-abundance data post-clindamycin treatment had the best performance,
-suggesting how the bacterial community responds to clindamycin treatment
-has the greatest influence on subsequent \emph{C. difficile}
-colonization dynamics.
+performed significantly better than all other models with an area under
+the receiving operator characteristic curve (AUROC) of 0.75
+(\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise comparisons;
+Table S13). Thus, we were able to use community bacterial relative
+abundance data alone to differentiate mice that had cleared \emph{C.
+difficile} before day 7 from the mice still colonized with \emph{C.
+difficile} at that timepoint. Interestingly, the model built with OTU
+relative abundance data post-clindamycin treatment had the best
+performance, suggesting how the bacterial community responds to
+clindamycin treatment has the greatest influence on subsequent \emph{C.
+difficile} colonization dynamics.
 
@@
 performance, we pulled out the top 20 taxa that had the highest absolute
-feature weights in each of the 6 models (Table S14-15). First, we looked
-at OTUs from the model with the best performance that was based on the
-post-clindamycin treatment bacterial community data. While most of the
-20 OTUs had low relative abundances on day 0, \emph{Enterobacteriace},
-\emph{Bacteroides} and \emph{Proteus} had high relative abundances in at
-least one source of mice and significantly varied across sources (Fig.
-7A). Next, the top 20 taxa from each model were compared to the list of
-taxa that varied across colony source (Fig. 5 and Table S8-9) at the
-same timepoint and the taxa that were altered by clindamycin treatment
-(Fig. 6 and Table S10-11). We found a subset of OTUs and families that
-were important to the model and overlapped with bacteria that varied by
-either source, clindamycin treatment, or both (Fig. S4, S5A-C).
-Combining the overall results for the 3 OTU models identified 14 OTUs
-associated with source, 21 OTUs associated with clindamycin treatment,
-and 6 OTUs associated with both (Fig. 7B). Combining the overall results
-for the 3 family models identified 18 families associated with source,
-14 families associated with clindamycin treatment and 8 families
-associated with both (Fig. S5D). Several OTUs (\emph{Bacteroides (OTU
-2), Enterococcus (OTU 23), Enterobacteriaceae (OTU 1),
-Porphyromonadaceae (OTU 7)}) and families (\emph{Bacteroidaceae,
-Deferribacteraceae, Enterococcaceae, Lachnospiraceae,
-Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
+feature weights in each of the 6 models (Tables S14-15). First, we
+looked at OTUs from the model with the best performance that was based
+on the post-clindamycin treatment bacterial community data. While most
+of the 20 OTUs had low relative abundances on day 0,
+\emph{Enterobacteriace}, \emph{Bacteroides} and \emph{Proteus} had high
+relative abundances in at least one source of mice and significantly
+varied across sources (Fig. 7A). Next, the top 20 taxa from each model
+were compared to the list of taxa that varied across colony source at
+the same timepoint (Fig. 5 and Tables S8-9) and the taxa that were
+altered by clindamycin treatment (Fig. 6 and Table S10-11). We found a
+subset of OTUs and families that were important to the model and
+overlapped with bacteria that varied by either source, clindamycin
+treatment, or both (Fig. S4, S5A-C). Combining the overall results for
+the 3 OTU models identified 14 OTUs associated with source, 21 OTUs
+associated with clindamycin treatment, and 6 OTUs associated with both
+(Fig. 7B). Combining the overall results for the 3 family models
+identified 18 families associated with source, 14 families associated
+with clindamycin treatment and 8 families associated with both (Fig.
+S5D). Several OTUs (\emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
+Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)}) and families
+(\emph{Bacteroidaceae, Deferribacteraceae, Enterococcaceae,
+Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
 Verrucomicrobiaceae}) appeared across at least 2 models, so we examined
@@
 infection. Together, these results suggest the initial bacterial
-communities and their responses to clindamycin have a large influence on
-the time needed to clear \emph{C. difficile}.
+communities and their responses to clindamycin influence on the time
+needed to clear \emph{C. difficile}.
 
@@
 models with baseline, post-clindamycin treatment, and post-infection
-fecal community data that successfully predicted \emph{C. difficile}
-colonization status 7 days after infection better than random chance. We
-identified \emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
+fecal community data that successfully predicted whether mice cleared
+\emph{C. difficile} by 7 days post-infection better than random chance.
+We identified \emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
 Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)},
@@
 \emph{C. difficile} colonization, by following colonization over time we
-found Jackson, Taconic, and Envigo mice remained colonized beyond 7 days
-post-infection. We identified a subset of bacteria that were important
-in predicting whether a mouse was still colonized with \emph{C.
-difficile} 7 days post-infection. These results suggest a subset of the
-bacterial community is responsible for determining the length of time
-needed to clear \emph{C. difficile} colonization.
+found Jackson, Taconic, and Envigo mice tended to remain colonized
+through at least 7 days post-infection. We identified a subset of
+bacteria that were important in predicting whether a mouse was still
+colonized with \emph{C. difficile} 7 days post-infection. These results
+suggest a subset of the bacterial community is responsible for
+determining the length of time needed to clear \emph{C. difficile}
+colonization.
 
@@
 The \emph{C. difficile} strain used could also be contributing to the
-variation in \emph{C. difficile} outcomes seen across different groups
-(47). We found the time needed to naturally clear \emph{C. difficile}
-varied across sources of mice implying that at least in the context of
-the same perturbation, microbiota differences seemed to influence
-infection outcome more than susceptibility. More importantly, we were
-able to narrow down from all the variation observed across colony
-sources to a subset of bacterial taxa that were also important for
-predicting \emph{C. difficile} colonization status 7 days
+variation in \emph{C. difficile} outcomes seen across different research
+groups (47). We found the time needed to naturally clear \emph{C.
+difficile} varied across sources of mice implying that at least in the
+context of the same perturbation, microbiota differences seemed to
+influence infection outcome more than susceptibility. More importantly,
+we were able to narrow down from all the variation observed across
+colony sources to a subset of bacterial taxa that were also important
+for predicting \emph{C. difficile} colonization status 7 days
 post-infection. Since all but 3 mice eventually cleared \emph{C.
@@
 not aid in the goal of figuring out how a variety of unique microbiotas
-influence susceptiblity to CDIs and adverse outcomes. Encouragingly,
-decreased \emph{Bifidobacterium}, \emph{Porphyromonas},
+influence susceptiblity to CDIs and adverse outcomes. Importantly, our
+study using mice from 6 different sources increased the variation
+between groups of mice compared to using 1 source alone, to better
+reflect the inter-individual microbiota variation observed in humans.
+Encouragingly, decreased \emph{Bifidobacterium}, \emph{Porphyromonas},
 \emph{Ruminococcaceae} and \emph{Lachnospiraceae} and increased
@@
 \emph{Proteus}, which was only found in Taconic mice. Importantly, the
-fact that some CDI associated bacteria were only found in a subset of
+fact that some CDI-associated bacteria were only found in a subset of
 mice has important implications for future CDI mouse model studies.
@@
 butyrate (52, 66). Interestingly, butyrate has previously been shown to
-vary across vendors and mediates resistance to \emph{Citrobacter
+vary across vendors and mediated resistance to \emph{Citrobacter
 rodentium} infection, a model of enterohemorrhagic and enteropathogenic
@@
 influenced by the microbiota. We have demonstrated that the ways
-different baseline microbiotas from different mouse colony sources
-respond to clindamycin treatment influences the length of time mice
-remained colonized with \emph{C. difficile} 630. For those interested in
+baseline microbiotas from different mouse colony sources respond to
+clindamycin treatment influences the length of time mice remained
+colonized with \emph{C. difficile} 630. For those interested in
 dissecting the contribution of the microbiome to \emph{C. difficile}
@@
 \href{https://github.com/SchlossLab/ML_pipeline_microbiome}{pipeline},
-Ana Taylor for help with some media preparation and sample collection,
-and Nicholas Lesniak for his critical feedback on the manuscript. We
-also thank members of Vincent Young's lab, particularly Kimberly
-Vendrov, for guidance with the \emph{C. difficile} infection mouse model
-and donating the mice. We also want to thank the Unit for Laboratory
-Animal Medicine at the University of Michigan for maintaining our mouse
-colony and providing the institutional support for our mouse
+Ana Taylor for help with some of the media preparation and sample
+collection, and Nicholas Lesniak for his critical feedback on the
+manuscript. We also thank members of Vincent Young's lab, particularly
+Kimberly Vendrov, for guidance with the \emph{C. difficile} infection
+mouse model and donating the mice. We also want to thank the Unit for
+Laboratory Animal Medicine at the University of Michigan for maintaining
+our mouse colony and providing the institutional support for our mouse
 experiments. Finally, we thank Kwi Kim, Austin Campbell, and Kimberly
@@
 data to select the best cost hyperparameter and the testing of the held
-out test data to measure model performance. The modified
-training/testing ratio was selected to accommodate the small number of
-samples in the dataset. Code was modified from
+out test data to measure model performance. The modified training to
+testing ratio was selected to accommodate the small number of samples in
+the dataset. Code was modified from
 \url{https://github.com/SchlossLab/ML_pipeline_microbiome} to update the
@@
 and circles represent individual mice from experiment 1 while triangles
-represent mice from experiment 2. D. \emph{C. difficile} CFU/gram stool
-on day 7 post-infection across sources of mice with asterisks for
-pairwise Wilcoxon comparisons with Benjamini-Hochberg correction where
-\emph{P} \textless{} 0.05. E. Mouse weight change 2 days post-infection
-across sources of mice, no pairwise Wilcoxon comparisons were
-significant after Benjamini-Hochberg correction. For D-E. Circles
-represent experiment 1 mice, triangles represent experiment 2 mice and
-gray lines indicate the median values for each group.
+represent mice from experiment 2.
 
@@
 consistently vary across mouse colony sources despite clindamycin
-perturbation and \emph{C. difficile} challenge.} A-C. Boxplots of the
+perturbation and \emph{C. difficile} challenge.} A-C: plots highlighting
+the median (point) and interquantile range (colored lines) of the
 relative abundances for the 12 OTUs that consistently varied across
 sources of mice at the baseline (A), post-clindamycin (B), and
-post-infection (C) timepoints of the experiment. D-F. Boxplots of the
+post-infection (C) timepoints of the experiment. D-F: plots highlighting
+the median (point) and interquantile range (colored lines) of the
 relative abundances for the 8 families that consistently varied across
@@
 has the same effects on a subset of taxa regardless of colony source.}
-A-B. Boxplots of the top 10 most significant (adjusted \emph{P} value
-\textless{} 0.05) OTUs with relative abundances that changed post
-clindamycin treatment. C-D. Boxplots of the top 10 most significant
-families with relative abundances that changed post clindamycin
+A-B: plots highlighting the median (point) and interquantile range
+(colored lines) of the top 10 most significant (adjusted \emph{P} value
+\textless{} 0.05) OTUs with relative abundances that changed after
+clindamycin treatment. C-D: plots highlighting the median (point) and
+interquantile range (colored lines) of the top 10 most significant
+families with relative abundances that changed after clindamycin
 treatment. Data were analyzed by Wilcoxon signed rank test limited to
@@
 day 0 OTU relative abundances that significantly varied across sources
-of mice and had high relative abundances. Symbols represent the relative
-abundance data for an individual mouse, circles represent mice that
-cleared \emph{C. difficile} by day 7, X-shapes represent mice that were
-still colonized with \emph{C. difficile}, and open circles represent
-mice that did not have \emph{C. difficile} CFU counts for day 7
-post-infection. Gray lines indicate the median relative abundances for
-each source. Asterisks are shown for pairwise Wilcoxon comparisons with
-Benjamini-Hochberg correction where \emph{P} \textless{} 0.05. B. Venn
-diagram that combines Fig. S4 summaries of OTUs that were important to
-the day -1, 0, and 1 classification models (Table S14) and either
-overlapped with taxa that varied across vendors at the same timepoint,
-were impacted by clindamycin treatment, or both. See Fig. S4 for
-separate comparisons of taxa from the day -1, 0, and 1 classification
-models. Bold OTUs signify OTUs that were important to more than 1
-classification model.
+of mice and had high relative abundances in the community. Symbols
+represent the relative abundance data for an individual mouse, circles
+represent mice that cleared \emph{C. difficile} by day 7, X-shapes
+represent mice that were still colonized with \emph{C. difficile}, and
+open circles represent mice that did not have \emph{C. difficile} CFU
+counts for day 7 post-infection. Gray lines indicate the median relative
+abundances for each source. Asterisks are shown for pairwise Wilcoxon
+comparisons with Benjamini-Hochberg correction where \emph{P}
+\textless{} 0.05. B. Venn diagram that combines Fig. S4 summaries of
+OTUs that were important to the day -1, 0, and 1 classification models
+(Table S14) and either overlapped with taxa that varied across vendors
+at the same timepoint, were impacted by clindamycin treatment, or both.
+See Fig. S4 for separate comparisons of taxa from the day -1, 0, and 1
+classification models. Bold OTUs signify OTUs that were important to
+more than 1 classification model.
 
@@
 represent individual mice and the black line represents the limit of
-detection. C. Percent of mice that were colonized with \emph{C.
+detection. C. \emph{C. difficile} CFU/gram stool on day 7 post-infection
+across sources of mice with asterisks for pairwise Wilcoxon comparisons
+with Benjamini-Hochberg correction where \emph{P} \textless{} 0.05. D.
+Mouse weight change 2 days post-infection across sources of mice, no
+pairwise Wilcoxon comparisons were significant after Benjamini-Hochberg
+correction. For C-D: circles represent experiment 1 mice, triangles
+represent experiment 2 mice and gray lines indicate the median values
+for each group. E. Percent of mice that were colonized with \emph{C.
 difficile} over the course of the experiment. Each day the percent is
@@
 difficile colonization status that were used as classification outcomes
-to build logistic regression models. Mice were classified as colonized
-or cleared (not detectable at the limit of detection of 100 CFU) based
-on CFU g/stool data from 7 days post-infection. B. \emph{C. difficile}
-CFU status on Day 7 within each mouse colony source. N = 5-9 mice per
-group. C-D. L2-regularized logistic regression classification model
-AUROCs to predict \emph{C. difficile} CFU on D7 (Fig. 1D, Fig. S3) based
-on the community relative abundances at baseline (day -1),
+to build L2-regularized logistic regression models. Mice were classified
+as colonized or cleared (not detectable at the limit of detection of 100
+CFU) based on CFU g/stool data from 7 days post-infection. B. \emph{C.
+difficile} CFU status on Day 7 within each mouse colony source. N = 5-9
+mice per group. C-D. L2-regularized logistic regression classification
+model area under the receiving operator characteristic curve (AUROCs) to
+predict \emph{C. difficile} CFU on day 7 post-infectoin (Fig. 1D, Fig.
+S3) based on the community relative abundances at baseline (day -1),
 post-clindamycin (day 0), and post-infection (day 1) at either the OTU
 (C) or family (D) level. All models performed better than random chance
-(AUROC = 0.5), see (all \emph{P} \(\le\) 5e-15; Table S12) and the model
-built with post-clindamycin treated bacterial OTU relative abundances
-had the best performance ((\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10
-for pairwise comparisons; Table S13). A List of the 20 taxa that were
-ranked as most important to each model are listed in Table S14-15.
+(AUROC = 0.5; all \emph{P} \(\le\) 5e-15; Table S12) and the model built
+with post-clindamycin treated bacterial OTU relative abundances had the
+best performance ((\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for all
+pairwise comparisons; Table S13). For lists of the 20 taxa that were
+ranked as most important to each model, see Tables S14-15.
 
@@
 
-\textbf{Table S6. PERMANOVA results for all mice, all timepoints.}
+\textbf{Table S6. PERMANOVA results for all mice at baseline, post
+clindamycin, and post-infection timepoints.}
 
@@
 \textbf{Table S11. Families with relative abudances that significantly
-changed after clindamycin treatment.} \textbf{Table S12. Statistical
-results of L2-regularized logistic regression model performances
-compared to random chance.}
-
-\textbf{Table S13. Pairwise Wilcoxan results of comparing all 6
+changed after clindamycin treatment.}
+
+\textbf{Table S12. Statistical results of L2-regularized logistic
+regression model performances compared to random chance.}
+
+\textbf{Table S13. Pairwise Wilcoxan results comparing all 6
 L2-regularized logistic regression model performances.}
@@
 \textbf{Table S16. OTUs with relative abudances that significantly
-varied across sources of mice on at least 1 day of the experiment.}
+varied across sources of mice on at least 1 day of the experiment by
+Kruskal-Wallis test.}
 
 \textbf{Table S17. Families with relative abudances that significantly
-varied across sources of mice on at least 1 day of the experiment.}
+varied across sources of mice on at least 1 day of the experiment by
+Kruskal-Wallis test.}
 

### 2020-07-08 submission/manuscript.Rmd
Text edits before reorganization

@@
 output:
-  word_document: default
   pdf_document:
@@
     keep_tex: yes
+  word_document: default
   html_document:
@@
 
-# Initial gut microbiota and response to antibiotic perturbation influence *Clostridioides difficile* colonization in mice
+# The initial gut microbiota and response to antibiotic perturbation influence *Clostridioides difficile* colonization in mice
 
@@
 ## Abstract
-The microbiota plays a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs).
-However, much of the mechanistic work examining CDIs in mouse models use a single university colony or vendor.
-We treated mice from 6 different colony sources (2 University of Michigan colonies and 4 vendors) with a single clindamycin dose, followed by *C.
-difficile* challenge 1 day later and measured *C.
+The gut microbiota has a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs).
+However, much of the mechanistic work examining CDIs in mouse models use animals obtained from a single university colony or vendor.
+We treated mice from 6 different sources (2 University of Michigan colonies and 4 vendors) with a single clindamycin dose, followed by a *C.
+difficile* challenge 1 day later and then measured *C.
 difficile* colonization levels through 9 days post-infection.
-The microbiota was profiled via 16S rRNA gene sequencing analysis to examine variation across colony sources and alterations due to clindamycin treatment and *C.
+The microbiota was profiled via 16S rRNA gene sequencing to examine the variation across sources and alterations due to clindamycin treatment and *C.
 difficile* challenge.
-While all sources of mice were colonized 1-day post-infection, variation in *C.
-difficile* colonization levels emerged from days 3-7 post-infection with some sources colonized with *C.
-difficile* for slightly longer and at higher levels.
-We identified bacterial taxa with different relative abundances across colony sources throughout the experiment, as well as taxa that were consistently impacted by clindamycin treatment in all sources of mice.
-We created logistic regression models that successfully classified mice based on whether they cleared *C.
+While all sources of mice were colonized 1-day post-infection, variation emerged from days 3-7 post-infection with animals from some sources colonized with *C.
+difficile* for longer and at higher levels.
+We identified bacteria that varied in relative abundance across sources and throughout the experiment.
+Some bacteria were consistently impacted by clindamycin treatment in all sources of mice including *Lachnospiraceae*, *Ruminococcaceae*, and *Enterobacteriaceae*.
+To identify bacteria that were most important to colonization regardless of the source, we created logistic regression models that successfully classified mice based on whether they cleared *C.
 difficile* by 7 days post-infection using baseline, post-clindamycin, and post-infection community composition data.
-After examining the taxa that were most important to the classification models, we identified a subset of key taxa that varied across colony sources (*Bacteroides*, *Deferribacteraceae*), were altered by clindamycin (*Porphyromonadaceae*, *Ruminococcaceae*), or both (*Enterobacteriaceae*, *Enterococcus*, *Bifidobacteriaceae*, *Coriobacteriaceae*, *Lachnospiraceae*, and *Verrucomicrobiaceae*).
-These results suggest the response of the initial gut microbiota to clindamycin treatment influences *C.
-difficile* 630 colonization dynamics.
+With these models, we identified 4 bacteria that varied across sources (*Bacteroides*), were altered by clindamycin (*Porphyromonadaceae*), or both (*Enterobacteriaceae* and *Enterococcus*).
+Microbiota variation across sources better emulates human interindividual variation and can help identify bacterial drivers of phenotypic variation in the context of CDIs.
 
@@
 *Clostridioides difficile* is a leading nosocomial infection.
-Although the microbiota has been established as a key risk factor, there is variation in who becomes asymptomatically colonized, develops an infection, or has an infection with adverse outcomes.
+Although perturbation to the gut microbiota has been established as a key risk factor, there is variation in who becomes asymptomatically colonized, develops an infection, or has an infection with adverse outcomes.
 *C.
@@
 However, the inter-individual variation between mice is less than what is observed in humans, particularly if just one source of mice is used.
-In this study, we administered clindamycin to mice from 6 different colony sources and challenged them with *C.
+In this study, we administered clindamycin to mice from 6 different breeding colonies and challenged them with *C.
 difficile*.
-Interestingly, only a subset of the taxa that vary across sources were associated with how long *C.
+Interestingly, only a subset of the bacteria that vary across sources were associated with how long *C.
 difficile* was able to colonize.
 Future studies examining the interplay between the microbiota and *C.
-difficile* should consider using mice from multiple sources to narrow down the microbes driving the observed phenotypes and reflect human interindividual variation.
+difficile* should consider using mice from multiple sources to better reflect human interindividual variation.
 
@@
 
-Antibiotics are a clear risk factor for *Clostridioides difficile* infections (CDIs), but there is variation in who goes on to develop severe or recurrent CDIs after exposure [@Teng2019; @Kelly2012].
+Antibiotics are a common risk factor for *Clostridioides difficile* infections (CDIs), but there is variation in who goes on to develop severe or recurrent CDIs after exposure [@Teng2019; @Kelly2012].
 Additionally, asymptomatic colonization, where *C.
@@
 difficile*, including the role of the microbiota and efficacy of potential therapeutics for treating CDIs [@Best2012].
-However, microbiome variation between lab mice is much less than the variation observed between humans [@Baxter2014; @Nagpal2018].
+However, variation in the microbiome between mice from the same breeding colony is much less than the variation observed between humans [@Baxter2014; @Nagpal2018].
 Additionally, studying the contribution of the microbiota to a particular disease phenotype in one set of lab mice after the same perturbation could yield a number of findings of which only a fraction may be driving the phenotype.
@@
 Microbiome differences between different mouse vendors have been well documented and shown to influence susceptibility to a variety of diseases [@Stough2016; @Alegre2019], including enteric infections [@EtienneMesmin2017; @Velazquez2019; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018].
-Additionally, different research groups have observed different CDI outcomes in mice despite using similar models and the microbiome has been proposed as one factor potentially mediating CDI susceptibility and outcomes [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014].
+Additionally, research groups have observed different CDI outcomes in mice despite using similar models and the microbiome has been proposed as one factor potentially mediating CDI susceptibility and outcomes [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014].
 Here we examined how variations in the baseline microbiome and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C.
@@
 **Clindamycin treatment renders all mice susceptible to *C.
-difficile* 630 colonization regardless of colony source.** To test how the microbiotas of mice from different colony sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan (the Young and Schloss lab colonies), the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan).
+difficile* 630 colonization regardless of source.** To test how the microbiotas of mice from different sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other 10 years ago (the Young and Schloss lab colonies) and four commercial source: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan).
 These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019].
+
 After a 13-day acclimation period for the mice ordered from vendors, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C.
@@
 The day after infection, *C.
-difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of colony source (Fig.
+difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of source (Fig.
 1B).
@@
 difficile* CFU levels across sources of mice emerged from days 3-7 post-infection (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig.
-1B and Table S1), suggesting mouse colony source is associated with *C.
+1B and Table S1), suggesting mouse source is associated with *C.
 difficile* clearance.
@@
 Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment.
-Interestingly, mice ordered from Jackson, Taconic, and Envigo tended to lose more weight (although there was variation between experiments with Schloss and Envigo mice), have higher *C.
+Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to lose more weight (although there was variation between experiments with Schloss and Envigo mice), have higher *C.
 difficile* CFU levels and take longer to clear the infection compared to the other sources of mice, which was particularly evident 7 days post-infection (Fig.
@@
 Thus, clindamycin rendered all mice susceptible to *C.
-difficile* 630 colonization, regardless of colony source, but variation across sources emerged with 3 out of 6 sources taking longer to clear *C.
-difficile*.
+difficile* 630 colonization, regardless of source, but there was significant variation in disease phenotype across the sources of mice.
 
@@
 ```
-**Bacterial communities consistently vary across mouse colony sources despite antibiotic and infection perturbations.** Given the well known variation in mouse microbiomes across vendors and university colonies [@Alegre2019], we hypothesized that the variation in *C.
-difficile* clearance could be explained by microbiota variation across the 6 sources.
+**Bacterial communities consistently vary across sources despite antibiotic and infection perturbations.** Given the well known variation in mouse microbiomes across breeding colonies [@Alegre2019], we hypothesized that the variation in *C.
+difficile* clearance could be explained by microbiota variation across the 6 sources of mice.
 We used 16S rRNA gene sequencing to characterize the fecal bacterial communities from the mice over the course of the experiment.
 Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity measures across the 6 sources of mice.
-Examining the bacterial communities at baseline, prior to clindamycin treatment there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig.
+Examining the bacterial communities at baseline, prior to clindamycin treatment, there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig.
 2A-B and Table S3).
-As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and communities started to recover 1 day post-infection (Fig.
+As expected, the clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and richness and Shannon diversity started to increase 1 day post-infection (Fig.
 2C-D).
@@
 difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other sources (Fig 2C-F and Table S4).
-While Charles River mice had more diverse microbiotas, Young and Schloss lab mice were also able to clear *C.
-difficile* faster, suggesting microbiota diversity alone does not explain the observed variation in *C.
+Although the Charles River mice had more diverse microbiotas, the Young and Schloss lab mice were also able to clear *C.
+difficile* faster than the other sources, suggesting microbiota diversity alone does not explain the observed variation in *C.
 difficile* colonization across vendors.
@@
 ```
-Next, we compared the bacterial communities from the 6 colonies over the course of the experiment using principal coordinate analysis (PCoA) of the Theta YC distances.
-Permutational multivariate analysis of variance (PERMANOVA) analysis revealed colony source was the major factor explaining the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment (Movie S1 and Table S5).
-Since, the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the bacterial communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1).
-For all 3 timepoints, source and the interaction with cage significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig.
+Next, we compared the community structure of mice from the 6 sources over the course of the experiment using principal coordinate analysis (PCoA) of the $\theta_{YC}$ distances.
+Permutational multivariate analysis of variance (PERMANOVA) analysis revealed source was the dominant factor that explained the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage (mice from the same sources were housed together, primarily at a density of 2 mice per cage) and day of the experiment (Movie S1 and Table S5).
+Mice that are co-housed have been shown to have similar gut microbiotas due to coprophagy [@Nguyen2015] and since mice within the same source were housed together, it is not surprising that cage also contributes to the observed microbiota variation.
+Since the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1).
+For all 3 timepoints, the source and cage interaction significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig.
 3 and Table S6).
@@
 S2C-D, F), suggesting there was some community variation between the 2 experiments within each source.
-Thus, mouse colony source was the factor that explained the most variation observed in the bacterial communities.
-Importantly with the exception of the 2 University colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C.
+Thus, source was the factor that explained the most variation observed in the bacterial communities.
+Importantly with the exception of the Schloss and Young colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C.
 difficile* challenge.
 
-Since there was some variation in microbiota communities between experiments at baseline, we next looked at how similar the communities were within the same source and between sources in response to clindamycin treatment and *C.
+Since there was some variation in the microbiota structure of mice from the same source between experiments at baseline, we next looked at how similar the communities were within the same source and between sources in response to clindamycin treatment and *C.
 difficile* challenge (Fig.
@@
 4A).
-Clindamycin treatment reduced variation between experiments within Schloss, Young and Jackson mice and some of the variation between sources diminished, particularly for the University of Michigan and Charles River mice (Fig.
+Clindamycin treatment reduced the variation between experiments within Schloss, Young and Jackson mice and some of the variation between sources diminished, particularly for the Schloss, Young, and Charles River mice (Fig.
 4B).
@@
 ```{r source_taxa}
-#Number of different otus and families across sources at day -1, 0, and 1 timepoints
+#Number of different otus across sources at day -1, 0, and 1 timepoints
 source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
 source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
-source_families <- read_tsv("../data/process/family_stats_dn1to1_combined.tsv")
-source_families_dn1 <- source_families %>% filter(day == -1) %>% nrow()
 source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
-source_families_d0 <- source_families %>% filter(day == 0) %>% nrow()
 source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()
-source_families_d1 <- source_families %>% filter(day == 1) %>% nrow()
 
-#Function to find which significant otus/genera/families are shared 
+#Function to find which significant otus are shared 
 intersect_all <- function(a,b,...){
@@
 
-#Families shared across baseline, post-clindamycin, and post-infection timepoints
-source_f_dn1 <- source_families %>% filter(day == -1) %>% pull(family)
-source_f_d0 <- source_families %>% filter(day == 0) %>% pull(family)
-source_f_d1 <- source_families %>% filter(day == 1) %>% pull(family)
-shared_source_families_dn1to1 <- length(intersect_all(source_f_dn1, source_f_d0, source_f_d1))
 ```
-After finding differences at the community level, we next identified the taxa that varied across sources of mice over the initial days of the experiment.
-We examined bacterial relative abundances at the operational taxonomic unit (OTU) and family levels, expecting the number of differences to be reduced at the family level due to the nature of bacterial taxonomy [@Nguyen2015].
-Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs and `r source_families_dn1` families (Tables S8-9) with relative abundances that varied across colony sources.
-Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs and `r source_families_d0` families (Tables S8-9).
+After finding differences at the community level, we next identified the bacteria that varied across sources of mice over the initial days of the experiment.
+We examined bacterial relative abundances at the operational taxonomic unit (OTU) level.
+Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs (Table S8) with relative abundances that varied across sources.
+Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs (Table S8).
 After *C.
-difficile* challenge, there were `r source_otus_d1` OTUs and `r source_families_d1` families (Tables S8-9) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment.
-In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 5A-C) and `r shared_source_families_dn1to1` families with relative abundances that consistently varied across colony sources (Fig.
-5D-F).
-Importantly, some of the taxa that consistently varied across sources also shifted with clindamycin treatment.
+difficile* challenge, there were `r source_otus_d1` OTUs (Table S8) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment.
+In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 5A-C) with relative abundances that consistently varied across sources.
+Importantly, some of the OTus that consistently varied across sources also shifted with clindamycin treatment.
 For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice.
 *Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment.
-In summary, mouse bacterial communities significantly varied according to colony source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C.
+In summary, mouse bacterial communities varied significantly according to source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C.
 difficile* challenge.
@@
 clindamycin_otus_sig <- clindamycin_otus %>% filter(p.value.adj < 0.05) %>% nrow()
-
-clindamycin_families <- read_tsv("../data/process/family_stats_dn1to0.tsv")
-clindamycin_families_sig <- clindamycin_families %>% filter(p.value.adj < 0.05) %>% nrow()
 ```
-**Clindamycin treatment alters a subset of taxa that were found in all colony sources.** Although there were bacteria that consistently varied across colony sources, we also wanted to identify the bacteria that shifted after clindamycin treatment, regardless of colony source.
-By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs and `r clindamycin_families_sig` families that were altered after clindamycin treatment (Fig.
-6 and Tables S10-11).
-Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) and 3 families (*Porphyromonadaceae*, *Enterococcaceae*, *Lachnospiraceae*) overlapped (Fig.
+**Clindamycin treatment alters a subset of taxa that were found in all sources.** Although there were bacteria that consistently varied across sources, we also wanted to identify the bacteria that shifted after clindamycin treatment, regardless of source.
+By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment (Fig.
+6 and Tables S9).
+Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) overlapped (Fig.
 5, Fig.
 6C-D).
-These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all colony sources and only a subset of the taxa also varied across colony sources.
+These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all sources and only a subset of the OTUs also varied across sources.
 
@@
 
-#Function to find which significant otus/genera/families are shared 
+#Function to find which significant otus are shared 
 intersect_all <- function(a,b,...){
@@
 
-#Families identified in logistic regression classification models (20 families with the highest ranking for each model)
-intrep_families <- read_tsv("../data/process/combined_top20_families_all_models.tsv")
-interp_families_dn1 <- intrep_families %>% filter(model_input_day == -1) %>% pull(family)
-interp_families_d0 <- intrep_families %>% filter(model_input_day == 0) %>% pull(family)
-interp_families_d1 <- intrep_families %>% filter(model_input_day == 1) %>% pull(family)
-interp_families_combined <- c(interp_families_dn1, interp_families_d0, interp_families_d1)
-#Families identified in logistic regression classification models (20 families with the highest ranking for each model)
-intrep_families <- read_tsv("../data/process/combined_top20_families_all_models.tsv")
-interp_families_dn1 <- intrep_families %>% filter(model_input_day == -1) %>% pull(family)
-interp_families_d0 <- intrep_families %>% filter(model_input_day == 0) %>% pull(family)
-interp_families_d1 <- intrep_families %>% filter(model_input_day == 1) %>% pull(family)
-interp_families_combined <- c(interp_families_dn1, interp_families_d0, interp_families_d1)
-
-#Families that vary across sources of mice on day -1, 0, or 1.
-sig_family_source <- read_tsv("../data/process/family_stats_dn1to1_combined.tsv")
-`sig_family_day-1` <- sig_family_source %>% filter(day == -1) %>% pull(family)
-`sig_family_day0` <- sig_family_source %>% filter(day == 0) %>% pull(family)
-`sig_family_day1` <- sig_family_source %>% filter(day == 1) %>% pull(family)
-#Families that consistently vary across sources of mice on day-1, 0, or 1
-shared_sig_families_Dn1toD1 <- intersect_all(`sig_family_day-1`, sig_family_day0, sig_family_day1)
-
-#Families that were impacted by clindamycin treatment
-`sig_family_pairs` <- read_tsv("../data/process/family_stats_dn1to0.tsv") %>% 
-  filter(p.value.adj < 0.05) %>% pull(family)
-
-# Day -1 overlapping families----
-# Overlap between families that varied across sources of mice on day -1 and top 20 taxa from logistic regression model based on day -1 community:
-dayn1_and_interp_dn1_f <- intersect_all(`sig_family_day-1`, `interp_families_dn1`)
-# Overlap between families that were altered by clindamycin treatment and top 20 taxa from logistic regression model based on day -1 community:
-paired_and_interp_dn1_f <- intersect_all(`sig_family_pairs`, `interp_families_dn1`)
-
-# Day 0 overlapping families----
-# Overlap between families that varied across sources of mice on day 0 and top 20 taxa from logistic regression model based on day 0 community:
-day0_and_interp_d0_f <- intersect_all(`sig_family_day0`, `interp_families_d0`)
-# Overlap between families that were altered by clindamycin treatment and top 20 taxa from logistic regression model based on day 0 community:
-paired_and_interp_d0_f <- intersect_all(`sig_family_pairs`, `interp_families_d0`)
-
-# Day 1 overlapping families----
-# Overlap between families that varied across sources of mice on day 1 and top 20 taxa from logistic regression model based on day 1 community:
-day1_and_interp_d1_f <- intersect_all(`sig_family_day1`, `interp_families_d1`)
-# Overlap between families that were altered by clindamycin treatment and top 20 taxa from logistic regression model based on day 1 community:
-paired_and_interp_d1_f <- intersect_all(`sig_family_pairs`, `interp_families_d1`)
-
-#Combined overlapping families based on day -1, 0, and 1 comparisons----
-#Combined overlapping families that varied by source and were in the top 20 taxa from at least one logistic regression model & remove any duplicates
-dayn1to1_and_interp_combined_f <- unique(c(`dayn1_and_interp_dn1_f`, `day0_and_interp_d0_f`, day1_and_interp_d1_f))
-#See which source families show up as significant over multiple days
-source_families_w_duplicates <- c(`dayn1_and_interp_dn1_f`, `day0_and_interp_d0_f`, day1_and_interp_d1_f)
-key_source_families <- sort(unique(source_families_w_duplicates[duplicated(source_families_w_duplicates)]))
-#Key source families: "Bacteroidaceae" -seen across 3 days, "Deferribacteraceae", "Enterococcaceae", "Lachnospiraceae"
-#Combined overlapping families that were altered by clindamycin treatment and were in the top 20 taxa from at least one logistic regression model
-paired_and_interp_combined_f <- unique(c(`paired_and_interp_dn1_f`,`paired_and_interp_d0_f`, paired_and_interp_d1_f))
-#See which clindamycin-associated families show up as significant over multiple days
-clind_familes_w_duplicates <- c(`paired_and_interp_dn1_f`,`paired_and_interp_d0_f`, paired_and_interp_d1_f)
-key_clind_families <- sort(unique(clind_familes_w_duplicates[duplicated(clind_familes_w_duplicates)]))
-#List of key families from source and clindamycin varying families:
-key_families <- unique(c(key_source_families, key_clind_families))
-list_key_families <- paste(key_families, collapse= ", ") #Format to output in paper
-
 #Venn diagram calculations for OTUs----
@@
 venn_otus_clind <-  length(paired_and_interp_combined)
-#Venn diagram calculations for families
-venn_families_source <- length(dayn1to1_and_interp_combined_f)
-venn_families_intersect <- length(intersect(dayn1to1_and_interp_combined_f, paired_and_interp_combined_f))
-venn_families_clind <-  length(paired_and_interp_combined_f)
 ```
 **Source-specific and clindamycin impacted bacteria distinguish *C.
-difficile* colonization status in mice.** After identifying taxa that varied by colony source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C.
+difficile* colonization status in mice.** After identifying taxa that varied by source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C.
 difficile* colonization at day 7 (Fig.
@@
 S1C).
-We trained L2-regularized logistic regression models with input bacterial community data from the baseline, post-clindamycin, and post-infection timepoints of the experiment to predict *C.
+We trained three L2-regularized logistic regression models with input bacterial community data from the baseline, post-clindamycin, and post-infection timepoints of the experiment to predict *C.
 difficile* colonization status on day 7 (Fig.
@@
 difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12).
-However, the models trained with OTU level data generally performed better than those trained with family level data with the expception of the models based on the post-infection (day 1) communities (Fig.
-S3C-D).
 Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than all other models with an area under the receiving operator characteristic curve (AUROC) of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13).
-Thus, we were able to use community bacterial relative abundance data alone to differentiate mice that had cleared *C.
+Thus, we were able to use bacterial relative abundance data alone to differentiate mice that had cleared *C.
 difficile* before day 7 from the mice still colonized with *C.
@@
 7A).
-Next, the top 20 taxa from each model were compared to the list of taxa that varied across colony source at the same timepoint (Fig.
+Next, the top 20 taxa from each model were compared to the list of taxa that varied across source at the same timepoint (Fig.
 5 and Tables S8-9) and the taxa that were altered by clindamycin treatment (Fig.
 6 and Table S10-11).
-We found a subset of OTUs and families that were important to the model and overlapped with bacteria that varied by either source, clindamycin treatment, or both (Fig.
+We found a subset of OTUs that were important to the model and overlapped with bacteria that varied by either source, clindamycin treatment, or both (Fig.
 S4, S5A-C).
@@
 7B).
-Combining the overall results for the 3 family models identified `r venn_families_source` families associated with source, `r venn_families_clind` families associated with clindamycin treatment and `r venn_families_intersect` families associated with both (Fig.
-S5D).
-Several OTUs (*`r list_key_otus`*) and families (*`r list_key_families`*) appeared across at least 2 models, so we examined how the relative abundances of these key taxa varied over the course of the experiment (Fig.
-8 and Fig.
-S6).
-Throughout the experiment, there was at least 1 timepoint where relative abundances of these taxa significantly varied across sources (Table S16-17).
-Interestingly, there were no taxa that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection with *C.
+Several OTUs (*`r list_key_otus`*) appeared across at least 2 models, so we examined how the relative abundances of these key taxa varied over the course of the experiment (Fig.
+8).
+Throughout the experiment, there was at least 1 timepoint where relative abundances of these OTUs significantly varied across sources (Table S13).
+Interestingly, there were no OTUs that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection with *C.
 difficile* 630, suggesting multiple bacteria influence the time needed to clear the infection.
-Together, these results suggest the initial bacterial communities and their responses to clindamycin influence on the time needed to clear *C.
+Together, these results suggest the initial bacterial communities and their responses to clindamycin influence the clearance of *C.
 difficile*.
@@
 
-By examining the *C.
-difficile* colonization dynamics within mice from 6 different colony sources after perturbing the microbiota with clindamycin treatment, we were able to identify bacterial taxa that were unique to sources throughout the experiment as well as taxa that were universally impacted by clindamycin.
-We built L2 logistic regression models with baseline, post-clindamycin treatment, and post-infection fecal community data that successfully predicted whether mice cleared *C.
+By running our CDI model with mice from 6 different sources, we were able to identify bacterial taxa that were unique to sources throughout the experiment as well as taxa that were universally impacted by clindamycin.
+We trained L2 logistic regression models with baseline, post-clindamycin treatment, and post-infection fecal community data that could predict whether mice cleared *C.
 difficile* by 7 days post-infection better than random chance.
-We identified *`r list_key_otus`*, *`r list_key_families`* (Fig.
-8, Fig.
-S6) as candidate bacteria within these communities that were influencing variation in *C.
-difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by colony source, were impacted by clindamycin treatment, or both.
+We identified *`r list_key_otus`* (Fig.
+8) as candidate bacteria within these communities that were influencing variation in *C.
+difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by source, were impacted by clindamycin treatment, or both.
 Overall, our results demonstrate clindamycin is sufficient to render mice from multiple sources susceptible to CDI and only a subset of the interindividual microbiota variation across mice from different sources was associated with the time needed to clear *C.
@@
 
-Other groups have taken similar approaches by using mice from multiple colony sources to identify bacteria that either promote colonization resistance or increase susceptibility to enteric infections [@Velazquez2019; @EtienneMesmin2017; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018].
+Other studies have taken similar approaches by using mice from multiple sources to identify bacteria that either promote colonization resistance or increase susceptibility to enteric infections [@Velazquez2019; @EtienneMesmin2017; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018].
 For example, in the context of *Salmonella* infections, *Enterobacteriaceae* and segmented filamentous bacteria have emerged as protective [@Velazquez2019; @Lai2020].
@@
 difficile* VPI10463 [@EtienneMesmin2017].
-Given that we ordered mice from the same vendors, we checked all mice for endogenous *C.
+Given that we obtained mice from the same vendors, we checked all mice for endogenous *C.
 difficile* by plating stool samples that were collected after clindamycin treatment.
@@
 difficile* colonization across sources.
+
 Although all mice were susceptible to *C.
-difficile* colonization, by following colonization over time we found Jackson, Taconic, and Envigo mice tended to remain colonized through at least 7 days post-infection.
+difficile* colonization, mice from Jackson, Taconic, and Envigo mice tended to remain colonized through at least 7 days post-infection.
 We identified a subset of bacteria that were important in predicting whether a mouse was still colonized with *C.
@@
 
-In the past variation between different CDI mouse model studies have been attributed to intestinal microbiome differences in mice across different institutional environments.
+Differences in CDI mouse model studies have been attributed to different intestinal microbiotas of mice from different sources.
 For example, groups using the same clindamycin treatment and C57BL/6 mice had different *C.
-difficile* outcomes, one having sustained colonization [@Buffie2011], while the other had transient [@Reeves2011].
+difficile* outcomes, one having sustained colonization [@Buffie2011], while the other had transient [@Reeves2011], despite both using *C.
+difficile* VPI 10643.
 Baseline differences in the microbiota composition have been hypothesized to partially explain the differences in colonization outcomes and overall susceptibility to *C.
 difficile* after treatment with the same antibiotic [@Lawley2013; @Hutton2014].
-We have shown that mice from 6 different sources were all susceptible to *C.
-difficile* 630, suggesting the microbiota influences *C.
-difficile* clearance more than susceptibility.
-Fortunately, the bacterial perturbations induced by clindamycin treatment have been well characterized and our findings agree with previous CDI mouse model work demonstrating *Enterococcus* and *Enterobacteriaceae* were associated with *C.
+The bacterial perturbations induced by clindamycin treatment have been well characterized and our findings agree with previous CDI mouse model work demonstrating *Enterococcus* and *Enterobacteriaceae* were associated with *C.
 difficile* susceptibility and *Porpyhromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, and *Turicibacter* were associated with resistance [@Schubert2015; @Jenior2018; @Tomkovich2019; @Buffie2011; @Buffie2014; @Lawley2009; @Lawley2012; @Jump2014].
-While we have demonstrated that susceptibility is uniform across vendors after clindamycin treatment, there could be different outcomes for either susceptibility or clearance in the case of other antibiotic treatments.
+While we have demonstrated that susceptibility is uniform across sources of mice after clindamycin treatment, there could be different outcomes for either susceptibility or clearance in the case of other antibiotic treatments.
 The *C.
@@
 difficile* varied across sources of mice implying that at least in the context of the same perturbation, microbiota differences seemed to influence infection outcome more than susceptibility.
-More importantly, we were able to narrow down from all the variation observed across colony sources to a subset of bacterial taxa that were also important for predicting *C.
+More importantly, we were able to reduce the variation observed across sources to identify a subset of OTUs that were also important for predicting *C.
 difficile* colonization status 7 days post-infection.
@@
 However, a major caveat to this method is the substantial loss of human microbiota community members upon transfer to mice [@Fouladi2020; @Walter2020].
-Additionally with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of the CDI mouse model studies to date associated mice with just 1 types of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018], which does not aid in the goal of figuring out how a variety of unique microbiotas influence susceptiblity to CDIs and adverse outcomes.
+Additionally with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of the CDI mouse model studies to date associated mice with just 1 types of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018], which does not aid in the goal of modeling the interpersonal variation seen in humans to understand how the microbiota influences susceptiblity to CDIs and adverse outcomes.
 Importantly, our study using mice from 6 different sources increased the variation between groups of mice compared to using 1 source alone, to better reflect the inter-individual microbiota variation observed in humans.
@@
 difficile* exposure depends on a multitude of factors, including age, diet, and immunity; all of which are also influenced by the microbiota.
-We have demonstrated that the ways baseline microbiotas from different mouse colony sources respond to clindamycin treatment influences the length of time mice remained colonized with *C.
+
+We have demonstrated that the ways baseline microbiotas from different mouse sources respond to clindamycin treatment influences the length of time mice remained colonized with *C.
 difficile* 630.
@@
 difficile*, these results could serve as a resource for selecting which mice to order to address the question.
+Using mice from multiple sources helps model the interpersonal microbiota variation among humans to aid our understanding of how the gut microbiota contributes to CDIs.
 
@@
 We thank members of the Schloss lab for feedback on planning the experiments and data presentation, as well as code tutorials and feedback through Code club.
-In particular, we want to thank Begüm Topçuoğlu for help with implementing L2 logistic regression models using her [pipeline](https://github.com/SchlossLab/ML_pipeline_microbiome), Ana Taylor for help with some of the media preparation and sample collection, and Nicholas Lesniak for his critical feedback on the manuscript.
+In particular, we want to thank Begüm Topçuoğlu for help with implementing L2 logistic regression models using her [pipeline](https://github.com/SchlossLab/ML_pipeline_microbiome), Ana Taylor for help with media preparation and sample collection, and Nicholas Lesniak for his critical feedback on the manuscript.
 We also thank members of Vincent Young’s lab, particularly  Kimberly Vendrov, for guidance with the *C.
@@
 **(i) Animals.** All experiments were approved by the University of Michigan Animal Care and Use Committee (IACUC) under protocol number PRO00006983.
-Female C57BL/7 mice were obtained from 6 different colony sources: The Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, Envigo, and two colonies at the University of Michigan (the Schloss lab colony and the Young lab colony).
-The Young lab colony was originally established with mice purchased from Jackson, and the Schloss lab colony was later founded with mice donated from the Young lab.
+Female C57BL/7 mice were obtained from 6 different sources: The Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, Envigo, and two colonies at the University of Michigan (the Schloss lab colony and the Young lab colony).
+The Young lab colony was originally established with mice purchased from Jackson, and the Schloss lab colony was established 10 years ago with mice donated from the Young lab.
 The 4 groups of mice purchased from vendors were allowed to acclimate to the University of Michigan mouse facility for 13 days prior to starting the experiment.
@@
 To account for uneven sequencing across samples, samples were rarefied to 5,437 sequences 1,000 times for alpha and beta diversity analyses.
-PCoAs were generated based on Theta YC distances.
-Permutational multivariate analysis of variance (PERMANOVA) was performed on mothur-generated Theta YC distance matrices with the adonis function in the vegan package [@Vegan2018] in R [@r_citation_2018].
+PCoAs were generated based on $\theta_{YC}$ distances.
+Permutational multivariate analysis of variance (PERMANOVA) was performed on mothur-generated $\theta_{YC}$ distance matrices with the adonis function in the vegan package [@Vegan2018] in R [@r_citation_2018].
 
 **(vi) Classification model training and evaluation.** Models were generated based on mice that were categorized as either cleared or colonized 7 days post-infection and had sequencing data from the baseline (day -1), post-clindamycin (day 0), and post-infection (day 1) timepoints of the experiment.
-Input bacterial community relative abundance data at either the OTU or family level from the baseline, post-clindamycin, and post-infection timepoints was used to generate 6 classification models that predicted *C.
+Input bacterial community relative abundance data at the OTU level from the baseline, post-clindamycin, and post-infection timepoints was used to generate 6 classification models that predicted *C.
 difficile* colonization status 7 days post-infection.
@@
 difficile* (OTU 20) was removed.
-Bacterial relative abundances that varied across sources at the OTU and family taxonomic levels were identified with the Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified taxa at each level, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
-Taxa impacted by clindamycin treatment were identified using the Wilcoxon signed rank test with matched pairs of mice samples for day -1 and day 0.
+Bacterial relative abundances that varied across sources at the OTU level were identified with the Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
+OTUs impacted by clindamycin treatment were identified using the Wilcoxon signed rank test with matched pairs of mice samples for day -1 and day 0.
 To determine whether classification models had better performance (test AUROCs) than random chance (0.5), we used the one-sample Wilcoxon signed rank test.
@@
 
-**(viii) Code availability.** Code for all data analysis and generating this manuscript is available at https://github.com/SchlossLab/Tomkovich_vendor_difs_XXXX_2020.
+**(viii) Code availability.** Code for all data analysis and generating this manuscript is available at https://github.com/SchlossLab/Tomkovich_Vendor_XXXX_2020.
 
@@
 **Figure 2.
-Differences in microbial richness and diversity across mouse colony sources emerge after clindamycin treatment and infection.** A-F.
+Differences in microbial richness and diversity across mouse sources emerge after clindamycin treatment and infection.** A-F.
 Number of observed OTUs and Shannon diversity index values at baseline: day -1 (A-B), after clindamycin: day 0 (C-D) and post-infection: day 1 (E-F) timepoints of the experiment.
@@
 **Figure 3.
-Mouse colony source is the variable that explains most of the variation observed in the baseline, post-clindamycin, and post-infection bacterial communities.** A-C.
-Principal Coordinates Analysis of Theta YC distances from stools collected at baseline (A), post-clindamycin (B), and post-infection (C) timepoints of the experiment.
+Mouse source is the variable that explains most of the variation observed in the baseline, post-clindamycin, and post-infection bacterial communities.** A-C.
+Principal Coordinates Analysis of $\theta_{YC}$ distances from stools collected at baseline (A), post-clindamycin (B), and post-infection (C) timepoints of the experiment.
 Each symbol represents a stool sample from an individual mouse, with circles representing experiment 1 mice and triangles representing experiment 2 mice.
@@
 High inter-group variation across mouse sources is diminished by clindamycin treatment** A-C.
-Boxplots of the Theta YC distances of the 6 sources of mice relative to mice within the same source and experiment, mice within the same source and between experiments, mice within the same source, and mice from other groups at the baseline (A), after clindamycin treatment (B), and post-infection (C) timepoints.
+Boxplots of the $\theta_{YC}$ distances of the 6 sources of mice relative to mice within the same source and experiment, mice within the same source and between experiments, mice within the same source, and mice from other groups at the baseline (A), after clindamycin treatment (B), and post-infection (C) timepoints.
 For comparisons within mice from the same source, symbols represent individual mouse samples: circles for experiment 1 and triangles for experiment 2.
@@
 **Figure 5.
-A subset of bacteria consistently vary across mouse colony sources despite clindamycin perturbation and *C.
+A subset of bacteria consistently vary across sources despite clindamycin perturbation and *C.
 difficile* challenge.** A-C: plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the `r shared_source_otus_dn1to1` OTUs that consistently varied across sources of mice at the baseline (A), post-clindamycin (B), and post-infection (C) timepoints of the experiment.
-D-F: plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the `r shared_source_families_dn1to1` families that consistently varied across sources of mice at the baseline (D), post-clindamycin (E), and post-infection (F) timepoints of the experiment.
-For each timepoint bacteria with differential relative abundances across sources of mice were identified by Kruskal-Wallis test at the family and OTU level with Benjamini-Hochberg correction for testing all identified taxa at the respective level (Table S8-9).
-The grey vertical line indicates the limit of detection for A-F.
+For each timepoint OTus with differential relative abundances across sources of mice were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs (Table S8).
+The grey vertical line indicates the limit of detection.
 
@@
 **Figure 6.
-Clindamycin treatment has the same effects on a subset of taxa regardless of colony source.** A-B: plots highlighting the median (point) and interquantile range (colored lines) of the top 10 most significant (adjusted *P* value < 0.05) OTUs with relative abundances that changed after clindamycin treatment.
-C-D: plots highlighting the median (point) and interquantile range (colored lines) of the top 10 most significant families with relative abundances that changed after clindamycin treatment.
+Clindamycin treatment has the same effects on a subset of taxa regardless of source.** A-B: plots highlighting the median (point) and interquantile range (colored lines) of the top 10 most significant (adjusted *P* value < 0.05) OTUs with relative abundances that changed after clindamycin treatment.
 Data were analyzed by Wilcoxon signed rank test limited to mice that had paired sequence data for day -1 and 0 (N = `r n_paired_for_clind`).
-Tests were performed at the OTU and family levels with Benjamini-Hochberg correction for testing all identified OTUs and families.
-See Table S10-11 for complete list of OTUs and families significantly impacted by clindamycin treatment.
-The grey vertical line indicates the limit of detection for A-D.
+Tests were performed at the OTU level with Benjamini-Hochberg correction for testing all identified OTUs.
+See Table S9 for complete list of OTUs significantly impacted by clindamycin treatment.
+The grey vertical line indicates the limit of detection.
 
@@
 \includegraphics{figure_8.pdf}
-**Figure 8: Key OTUs vary across sources throughout the experiment.** A-C.
+**Figure 8: Key OTUs vary across sources throughout the experiment.** A-D.
 Relative abundances of bold OTUs from Fig.
@@
 A.
-*Enterococcus* (OTU 23), which significantly varied across sources and was impacted by clindamycin treatment.
-B.
 *Bacteroides* (OTU 2), which varied across sources throughout the experiment.
-C.
-*Enterobacteriaceae* (OTU 1) and *Porphyromonadaceae* (OTU 7) were significantly impacted by clindamycin treatment and examining relative abundance dynamics over the course of the experiment indicated timepoints where relative abundances also significantly varied across sources of mice.
+B-C.
+*Enterobacteriaceae* (B) and *Enterococcus* (C), which significantly varied across sources and were impacted by clindamycin treatment.
+D.
+*Porphyromonadaceae* (OTU 7), which was significantly impacted by clindamycin treatment and examining relative abundance dynamics over the course of the experiment, revealed timepoints where relative abundances also significantly varied across sources of mice.
 Symbols represent the relative abundance data for an individual mouse, circles represent mice that cleared *C.
@@
 The gray horizontal line represents the limit of detection.
-Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S16) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05).
+Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S13) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05).
 
@@
 Only bacterial communities from University of Michigan mice significantly vary between experiments.** A-F.
-PCoA of Theta YC distances for the baseline fecal bacterial communities within each source of mice.
+PCoA of $\theta_{YC}$ distances for the baseline fecal bacterial communities within each source of mice.
 Each symbol represents a stool sample from an individual mouse with color corresponding to experiment and shape representing cage mates.
@@
 *C.
-difficile* CFU status on Day 7 within each mouse colony source.
+difficile* CFU status on Day 7 within each mouse source.
 N = `r day7_status_n_min`-`r day7_status_n_max` mice per group.
-C-D.
+C.
 L2-regularized logistic regression classification model area under the receiving operator characteristic curve (AUROCs) to predict *C.
@@
 1D, Fig.
-S3) based on the community relative abundances at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1) at either the OTU (C) or family (D) level.
-All models performed better than random chance (AUROC = 0.5; all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for all pairwise comparisons; Table S13).
-For lists of the 20 taxa that were ranked as most important to each model, see Tables S14-15.
+S3) based on the OTU community relative abundances at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1).
+All models performed better than random chance (AUROC = 0.5; all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for all pairwise comparisons; Table S11).
+For list of the 20 OTUs that were ranked as most important to each model, see Table S12.
 
@@
 **Figure S4.
-Key OTUs from classification models based on baseline, post-clindamycin treatment, or post-infection community data vary by mouse colony source, clindamycin treatment, or both.** A-C.
-Venn diagrams of top 20 important OTUs from baseline (A), post-clindamycin treatment (B), and post-infection (C) classification models (Table S14) that overlapped with OTUs that varied across vendors at baseline, were impacted by clindamycin treatment, or both.
+Key OTUs from classification models based on baseline, post-clindamycin treatment, or post-infection community data vary by source, clindamycin treatment, or both.** A-C.
+Venn diagrams of top 20 important OTUs from baseline (A), post-clindamycin treatment (B), and post-infection (C) classification models (Table S12) that overlapped with OTUs that varied across vendors at the corresponding timepoint, were impacted by clindamycin treatment, or both.
 Bold OTUs signify OTUs that were important to more than 1 classification model.
@@
 \newpage
-\includegraphics{figure_S5.pdf}
-**Figure S5.
-Key families from classification models based on baseline, post-clindamycin treatment, or post-infection community data vary by mouse colony source, clindamycin treatment, or both.** A-C.
-Venn diagrams of top 20 important families from baseline (A), post-clindamycin treatment (B), and post-infection (C) classification models (Table S15) that overlapped with families that varied across vendors after clindamycin, were impacted by clindamycin treatment, or both.
-D.
-Venn diagrams that combines A-C summaries of familes that were important to the day -1, 0, and 1 classification models (Table S15) and either overlapped with familes that varied across vendors at the same timepoint, were impacted by clindamycin treatment, or both.
-Bold families signify families that were important to more than 1 classification model.
-
-\newpage
-\includegraphics{figure_S6.pdf}
-**Figure S6.
-Key families vary across sources throughout experiment.** Relative abundances of bold families from Fig.
-S5D that were important for at least two classification models are shown over time.
-A.
-*Enterococcaceae* and *Lachnospiraceae*, which significantly varied across sources and were impacted by clindamycin treatment.
-B.
-*Bacteroidaceae* and *Deferribacteraceae*, which varied across sources throughout the experiment.
-C.
-*Bifidobacteriaceae*, *Coriobacteriaceae*, *Ruminococcaceae*, and *Verrucomicrobiaceae* were significantly impacted by clindamycin treatment.
-Examining the relative abundance dynamics throughout the experiment, identified timepoints where relative abundances also significantly varied across sources of mice.
-Symbols represent the relative abundance data for an individual mouse, circles represent mice that cleared *C.
-difficile* by day 7, X-shapes represent mice that were still colonized with *C.
-difficile*, and open circles represent mice that did not have *C.
-difficile* CFU counts for day 7 post-infection.
-Colored lines indicate the median relative abundances for each source.
-The gray horizontal line represents the limit of detection.
-Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S17) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05).
-
-\newpage
 ## Supplementary Tables and Movie
-All supplemental material is available at: https://github.com/SchlossLab/Tomkovich_vendor_difs_XXXX_2020/submission.
+All supplemental material is available at: https://github.com/SchlossLab/Tomkovich_Vendor_XXXX_2020/submission.
 
@@
 Large shifts in bacterial community structure occurred after clindamycin and *C.
-difficile* infection.** PCoA of Theta YC distances animated from 0 through 9 days post-infection.
-PERMANOVA analysis indicated colony source was the variable that explained the most observed variation across fecal communities (source R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment.
+difficile* infection.** PCoA of $\theta_{YC}$ distances animated from 0 through 9 days post-infection.
+PERMANOVA analysis indicated source was the variable that explained the most observed variation across fecal communities (source R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment.
 Transparency of the circle corresponds to the day of the experiment, each circle represents a sample from an individual mouse at a specific timepoint.
@@
 
-**Table S9. Families with relative abudances that significantly vary across sources at baseline, post-clindamycin, or post-infection timepoints.**
+**Table S9. OTUs with relative abudances that significantly changed after clindamycin treatment.**
 
-**Table S10. OTUs with relative abudances that significantly changed after clindamycin treatment.**
+**Table S10. Statistical results of L2-regularized logistic regression model performances compared to random chance.**
 
-**Table S11. Families with relative abudances that significantly changed after clindamycin treatment.**
+**Table S11. Pairwise Wilcoxan results comparing all 3 L2-regularized logistic regression model performances.**
 
-**Table S12. Statistical results of L2-regularized logistic regression model performances compared to random chance.**
+**Table S12. Top 20 most important OTUs for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
 
-**Table S13. Pairwise Wilcoxan results comparing all 6 L2-regularized logistic regression model performances.**
-
-**Table S14. Top 20 most important OTUs for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
-
-**Table S15. Top 20 most important families for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
-
-**Table S16. OTUs with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**
-
-**Table S17. Families with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**
+**Table S13. OTUs with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**

### 2020-07-08 submission/manuscript.tex
Text edits before reorganization

@@
 
-\section{\texorpdfstring{Initial gut microbiota and response to
+\section{\texorpdfstring{The initial gut microbiota and response to
 antibiotic perturbation influence \emph{Clostridioides difficile}
 colonization in
-mice}{Initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice}}\label{initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}
+mice}{The initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice}}\label{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}
 
@@
 
-The microbiota plays a key role in determining susceptibility to
+The gut microbiota has a key role in determining susceptibility to
 \emph{Clostridioides difficile} infections (CDIs). However, much of the
-mechanistic work examining CDIs in mouse models use a single university
-colony or vendor. We treated mice from 6 different colony sources (2
-University of Michigan colonies and 4 vendors) with a single clindamycin
-dose, followed by \emph{C. difficile} challenge 1 day later and measured
-\emph{C. difficile} colonization levels through 9 days post-infection.
-The microbiota was profiled via 16S rRNA gene sequencing analysis to
-examine variation across colony sources and alterations due to
-clindamycin treatment and \emph{C. difficile} challenge. While all
-sources of mice were colonized 1-day post-infection, variation in
-\emph{C. difficile} colonization levels emerged from days 3-7
-post-infection with some sources colonized with \emph{C. difficile} for
-slightly longer and at higher levels. We identified bacterial taxa with
-different relative abundances across colony sources throughout the
-experiment, as well as taxa that were consistently impacted by
-clindamycin treatment in all sources of mice. We created logistic
-regression models that successfully classified mice based on whether
-they cleared \emph{C. difficile} by 7 days post-infection using
-baseline, post-clindamycin, and post-infection community composition
-data. After examining the taxa that were most important to the
-classification models, we identified a subset of key taxa that varied
-across colony sources (\emph{Bacteroides}, \emph{Deferribacteraceae}),
-were altered by clindamycin (\emph{Porphyromonadaceae},
-\emph{Ruminococcaceae}), or both (\emph{Enterobacteriaceae},
-\emph{Enterococcus}, \emph{Bifidobacteriaceae},
-\emph{Coriobacteriaceae}, \emph{Lachnospiraceae}, and
-\emph{Verrucomicrobiaceae}). These results suggest the response of the
-initial gut microbiota to clindamycin treatment influences \emph{C.
-difficile} 630 colonization dynamics.
+mechanistic work examining CDIs in mouse models use animals obtained
+from a single university colony or vendor. We treated mice from 6
+different sources (2 University of Michigan colonies and 4 vendors) with
+a single clindamycin dose, followed by a \emph{C. difficile} challenge 1
+day later and then measured \emph{C. difficile} colonization levels
+through 9 days post-infection. The microbiota was profiled via 16S rRNA
+gene sequencing to examine the variation across sources and alterations
+due to clindamycin treatment and \emph{C. difficile} challenge. While
+all sources of mice were colonized 1-day post-infection, variation
+emerged from days 3-7 post-infection with animals from some sources
+colonized with \emph{C. difficile} for longer and at higher levels. We
+identified bacteria that varied in relative abundance across sources and
+throughout the experiment. Some bacteria were consistently impacted by
+clindamycin treatment in all sources of mice including
+\emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and
+\emph{Enterobacteriaceae}. To identify bacteria that were most important
+to colonization regardless of the source, we created logistic regression
+models that successfully classified mice based on whether they cleared
+\emph{C. difficile} by 7 days post-infection using baseline,
+post-clindamycin, and post-infection community composition data. With
+these models, we identified 4 bacteria that varied across sources
+(\emph{Bacteroides}), were altered by clindamycin
+(\emph{Porphyromonadaceae}), or both (\emph{Enterobacteriaceae} and
+\emph{Enterococcus}). Microbiota variation across sources better
+emulates human interindividual variation and can help identify bacterial
+drivers of phenotypic variation in the context of CDIs.
 
@@
 \emph{Clostridioides difficile} is a leading nosocomial infection.
-Although the microbiota has been established as a key risk factor, there
-is variation in who becomes asymptomatically colonized, develops an
-infection, or has an infection with adverse outcomes. \emph{C.
-difficile} infection (CDI) mouse models are widely used to answer a
-variety of \emph{C. difficile} pathogenesis questions. However, the
-inter-individual variation between mice is less than what is observed in
-humans, particularly if just one source of mice is used. In this study,
-we administered clindamycin to mice from 6 different colony sources and
-challenged them with \emph{C. difficile}. Interestingly, only a subset
-of the taxa that vary across sources were associated with how long
-\emph{C. difficile} was able to colonize. Future studies examining the
-interplay between the microbiota and \emph{C. difficile} should consider
-using mice from multiple sources to narrow down the microbes driving the
-observed phenotypes and reflect human interindividual variation.
+Although perturbation to the gut microbiota has been established as a
+key risk factor, there is variation in who becomes asymptomatically
+colonized, develops an infection, or has an infection with adverse
+outcomes. \emph{C. difficile} infection (CDI) mouse models are widely
+used to answer a variety of \emph{C. difficile} pathogenesis questions.
+However, the inter-individual variation between mice is less than what
+is observed in humans, particularly if just one source of mice is used.
+In this study, we administered clindamycin to mice from 6 different
+breeding colonies and challenged them with \emph{C. difficile}.
+Interestingly, only a subset of the bacteria that vary across sources
+were associated with how long \emph{C. difficile} was able to colonize.
+Future studies examining the interplay between the microbiota and
+\emph{C. difficile} should consider using mice from multiple sources to
+better reflect human interindividual variation.
 
@@
 
-Antibiotics are a clear risk factor for \emph{Clostridioides difficile}
+Antibiotics are a common risk factor for \emph{Clostridioides difficile}
 infections (CDIs), but there is variation in who goes on to develop
@@
 including the role of the microbiota and efficacy of potential
-therapeutics for treating CDIs (15). However, microbiome variation
-between lab mice is much less than the variation observed between humans
-(16, 17). Additionally, studying the contribution of the microbiota to a
-particular disease phenotype in one set of lab mice after the same
-perturbation could yield a number of findings of which only a fraction
-may be driving the phenotype.
+therapeutics for treating CDIs (15). However, variation in the
+microbiome between mice from the same breeding colony is much less than
+the variation observed between humans (16, 17). Additionally, studying
+the contribution of the microbiota to a particular disease phenotype in
+one set of lab mice after the same perturbation could yield a number of
+findings of which only a fraction may be driving the phenotype.
 
@@
 and shown to influence susceptibility to a variety of diseases (24, 25),
-including enteric infections (22, 23, 26--30). Additionally, different
-research groups have observed different CDI outcomes in mice despite
-using similar models and the microbiome has been proposed as one factor
+including enteric infections (22, 23, 26--30). Additionally, research
+groups have observed different CDI outcomes in mice despite using
+similar models and the microbiome has been proposed as one factor
 potentially mediating CDI susceptibility and outcomes (13, 18, 21,
@@
 \textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
-difficile} 630 colonization regardless of colony source.} To test how
-the microbiotas of mice from different colony sources impact
-colonization dynamics after clindamycin exposure, we utilized C57BL/6
-mice from 6 different sources: two colonies from the University of
-Michigan (the Young and Schloss lab colonies), the Jackson Laboratory,
-Charles River Laboratories, Taconic Biosciences, and Envigo (which was
-formerly Harlan). These 4 vendors were chosen because they represent
-commonly used vendors for CDI studies in mice (26, 34--40). After a
-13-day acclimation period for the mice ordered from vendors, all mice
-were treated with 10 mg/kg clindamycin via intraperitoneal injection and
-one day later challenged with 10\textsuperscript{3} \emph{C. difficile}
-630 spores (Fig. 1A). Clindamycin was chosen because we have previously
-demonstrated mice are rendered susceptible, but consistently cleared the
-CDI within 9 days (21, 41), clindamycin is frequently implicated with
-human CDIs (42), and is also part of the antibiotic treatment for the
-frequently cited 2008 CDI mouse model (14). The day after infection,
-\emph{C. difficile} was detectable in all mice at a similar level
-(median CFU range: 2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15),
-indicating clindamycin rendered all mice susceptible regardless of
-colony source (Fig. 1B). Interestingly, variation in \emph{C. difficile}
-CFU levels across sources of mice emerged from days 3-7 post-infection
-(all \emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 1B and Table S1),
-suggesting mouse colony source is associated with \emph{C. difficile}
-clearance. We conducted two experiments approximately 3 months apart and
-while the colonization dynamics were similar across most sources of
-mice, there was some variation between the 2 experiments, particularly
-for the Schloss and Envigo mice (Fig. S1A-B). Although \emph{C.
-difficile} 630 causes mild symptoms in mice comparied to other \emph{C.
-difficile} strains (43), we also saw that weight change significantly
-varied across sources of mice with the most weight lost two days
-post-infection (Fig. 1C and Table S2). Importantly, there was also one
-Jackson and one Envigo mouse that died between 1- and 3-days
-post-infection during the second experiment. Interestingly, mice ordered
-from Jackson, Taconic, and Envigo tended to lose more weight (although
-there was variation between experiments with Schloss and Envigo mice),
-have higher \emph{C. difficile} CFU levels and take longer to clear the
-infection compared to the other sources of mice, which was particularly
-evident 7 days post-infection (Fig. 1B-C, Fig. S2C-D), when 57\% of the
-mice were still colonized with \emph{C. difficile} (Fig. S1E). By 9 days
-post-infection the majority of the mice from all sources had cleared
-\emph{C. difficile} (Fig. S1C) with the exception of 1 Taconic mouse
-from the first experiment and 2 Envigo mice from the second experiment.
-Thus, clindamycin rendered all mice susceptible to \emph{C. difficile}
-630 colonization, regardless of colony source, but variation across
-sources emerged with 3 out of 6 sources taking longer to clear \emph{C.
-difficile}.
-
-\textbf{Bacterial communities consistently vary across mouse colony
-sources despite antibiotic and infection perturbations.} Given the well
-known variation in mouse microbiomes across vendors and university
-colonies (25), we hypothesized that the variation in \emph{C. difficile}
-clearance could be explained by microbiota variation across the 6
-sources. We used 16S rRNA gene sequencing to characterize the fecal
-bacterial communities from the mice over the course of the experiment.
-Since antibiotics and other risk factors of CDIs are associated with
-decreased microbiota diversity (44), we first examined alpha diversity
-measures across the 6 sources of mice. Examining the bacterial
-communities at baseline, prior to clindamycin treatment there was a
-significant difference in the number of observed OTUs
-(\emph{P}\textsubscript{FDR} = 0.03), but not Shannon diversity index
-(\emph{P}\textsubscript{FDR} = 0.052) across sources of mice (Fig. 2A-B
-and Table S3). As expected, clindamycin treatment decreased richness and
-Shannon diversity across all sources of mice, and communities started to
-recover 1 day post-infection (Fig. 2C-D). Interestingly, significant
-differences in diversity metrics across sources
-(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after both
-clindamycin and \emph{C. difficile} infection, with Charles River mice
-having higher richness and Shannon Diversity than most of the other
-sources (Fig 2C-F and Table S4). While Charles River mice had more
-diverse microbiotas, Young and Schloss lab mice were also able to clear
-\emph{C. difficile} faster, suggesting microbiota diversity alone does
-not explain the observed variation in \emph{C. difficile} colonization
-across vendors.
-
-Next, we compared the bacterial communities from the 6 colonies over the
-course of the experiment using principal coordinate analysis (PCoA) of
-the Theta YC distances. Permutational multivariate analysis of variance
-(PERMANOVA) analysis revealed colony source was the major factor
-explaining the observed variation across fecal communities
-(R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
-interactions between cage and day of the experiment (Movie S1 and Table
-S5). Since, the majority of the perturbations happened over the initial
-days of the experiment, we decided to focus on the bacterial communities
-at baseline (day -1), after clindamycin treatment (day 0), and
-post-infection (day 1). For all 3 timepoints, source and the interaction
-with cage significantly explained most of the observed community
-variation (combined R\textsuperscript{2} = 0.90, 0.99, 0.88,
-respectively; \emph{P} = 0.0001; Fig. 3 and Table S6). We also compared
-baseline communities across the 2 experiments, and found experiment and
-cage significantly explained the observed variation only for the Schloss
-and Young lab mouse colonies (Fig. S2A-B and Table S7). However, most of
-the vendors also clustered by experiment (Fig. S2C-D, F), suggesting
-there was some community variation between the 2 experiments within each
-source. Thus, mouse colony source was the factor that explained the most
-variation observed in the bacterial communities. Importantly with the
-exception of the 2 University colonies, the community of each source
-clustered apart from one another suggesting each community had a unique
-response to clindamycin treatment and \emph{C. difficile} challenge.
-
-Since there was some variation in microbiota communities between
-experiments at baseline, we next looked at how similar the communities
-were within the same source and between sources in response to
-clindamycin treatment and \emph{C. difficile} challenge (Fig. 4) . The
-baseline communities varied most between experiments for Schloss, Young,
-and Envigo mice and variation between sources of mice was high (Fig.
-4A). Clindamycin treatment reduced variation between experiments within
-Schloss, Young and Jackson mice and some of the variation between
-sources diminished, particularly for the University of Michigan and
-Charles River mice (Fig. 4B). Post-infection, the community variation
-started to increase within sources of mice and variation between sources
-of mice started to increase towards previous levels (Fig. 4C). By using
-mice from multiple sources we were able to increase the number of
-microbiota communities we tested with the clindamycin \emph{C.
-difficile} colonization mouse model.
+difficile} 630 colonization regardless of source.} To test how the
+microbiotas of mice from different sources impact colonization dynamics
+after clindamycin exposure, we utilized C57BL/6 mice from 6 different
+sources: two colonies from the University of Michigan that were split
+from each other 10 years ago (the Young and Schloss lab colonies) and
+four commercial source: the Jackson Laboratory, Charles River
+Laboratories, Taconic Biosciences, and Envigo (which was formerly
+Harlan). These 4 vendors were chosen because they represent commonly
+used vendors for CDI studies in mice (26, 34--40).
+
+After a 13-day acclimation period for the mice ordered from vendors, all
+mice were treated with 10 mg/kg clindamycin via intraperitoneal
+injection and one day later challenged with 10\textsuperscript{3}
+\emph{C. difficile} 630 spores (Fig. 1A). Clindamycin was chosen because
+we have previously demonstrated mice are rendered susceptible, but
+consistently cleared the CDI within 9 days (21, 41), clindamycin is
+frequently implicated with human CDIs (42), and is also part of the
+antibiotic treatment for the frequently cited 2008 CDI mouse model (14).
+The day after infection, \emph{C. difficile} was detectable in all mice
+at a similar level (median CFU range: 2.2e+07-1.3e+08;
+\emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
+mice susceptible regardless of source (Fig. 1B). Interestingly,
+variation in \emph{C. difficile} CFU levels across sources of mice
+emerged from days 3-7 post-infection (all \emph{P}\textsubscript{FDR}
+\(\le\) 0.019; Fig. 1B and Table S1), suggesting mouse source is
+associated with \emph{C. difficile} clearance. We conducted two
+experiments approximately 3 months apart and while the colonization
+dynamics were similar across most sources of mice, there was some
+variation between the 2 experiments, particularly for the Schloss and
+Envigo mice (Fig. S1A-B). Although \emph{C. difficile} 630 causes mild
+symptoms in mice comparied to other \emph{C. difficile} strains (43), we
+also saw that weight change significantly varied across sources of mice
+with the most weight lost two days post-infection (Fig. 1C and Table
+S2). Importantly, there was also one Jackson and one Envigo mouse that
+died between 1- and 3-days post-infection during the second experiment.
+Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to
+lose more weight (although there was variation between experiments with
+Schloss and Envigo mice), have higher \emph{C. difficile} CFU levels and
+take longer to clear the infection compared to the other sources of
+mice, which was particularly evident 7 days post-infection (Fig. 1B-C,
+Fig. S2C-D), when 57\% of the mice were still colonized with \emph{C.
+difficile} (Fig. S1E). By 9 days post-infection the majority of the mice
+from all sources had cleared \emph{C. difficile} (Fig. S1C) with the
+exception of 1 Taconic mouse from the first experiment and 2 Envigo mice
+from the second experiment. Thus, clindamycin rendered all mice
+susceptible to \emph{C. difficile} 630 colonization, regardless of
+source, but there was significant variation in disease phenotype across
+the sources of mice.
+
+\textbf{Bacterial communities consistently vary across sources despite
+antibiotic and infection perturbations.} Given the well known variation
+in mouse microbiomes across breeding colonies (25), we hypothesized that
+the variation in \emph{C. difficile} clearance could be explained by
+microbiota variation across the 6 sources of mice. We used 16S rRNA gene
+sequencing to characterize the fecal bacterial communities from the mice
+over the course of the experiment. Since antibiotics and other risk
+factors of CDIs are associated with decreased microbiota diversity (44),
+we first examined alpha diversity measures across the 6 sources of mice.
+Examining the bacterial communities at baseline, prior to clindamycin
+treatment, there was a significant difference in the number of observed
+OTUs (\emph{P}\textsubscript{FDR} = 0.03), but not Shannon diversity
+index (\emph{P}\textsubscript{FDR} = 0.052) across sources of mice (Fig.
+2A-B and Table S3). As expected, the clindamycin treatment decreased
+richness and Shannon diversity across all sources of mice, and richness
+and Shannon diversity started to increase 1 day post-infection (Fig.
+2C-D). Interestingly, significant differences in diversity metrics
+across sources (\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged
+after both clindamycin and \emph{C. difficile} infection, with Charles
+River mice having higher richness and Shannon Diversity than most of the
+other sources (Fig 2C-F and Table S4). Although the Charles River mice
+had more diverse microbiotas, the Young and Schloss lab mice were also
+able to clear \emph{C. difficile} faster than the other sources,
+suggesting microbiota diversity alone does not explain the observed
+variation in \emph{C. difficile} colonization across vendors.
+
+Next, we compared the community structure of mice from the 6 sources
+over the course of the experiment using principal coordinate analysis
+(PCoA) of the \(\theta_{YC}\) distances. Permutational multivariate
+analysis of variance (PERMANOVA) analysis revealed source was the
+dominant factor that explained the observed variation across fecal
+communities (R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
+interactions between cage (mice from the same sources were housed
+together, primarily at a density of 2 mice per cage) and day of the
+experiment (Movie S1 and Table S5). Mice that are co-housed have been
+shown to have similar gut microbiotas due to coprophagy (45) and since
+mice within the same source were housed together, it is not surprising
+that cage also contributes to the observed microbiota variation. Since
+the majority of the perturbations happened over the initial days of the
+experiment, we decided to focus on the communities at baseline (day -1),
+after clindamycin treatment (day 0), and post-infection (day 1). For all
+3 timepoints, the source and cage interaction significantly explained
+most of the observed community variation (combined R\textsuperscript{2}
+= 0.90, 0.99, 0.88, respectively; \emph{P} = 0.0001; Fig. 3 and Table
+S6). We also compared baseline communities across the 2 experiments, and
+found experiment and cage significantly explained the observed variation
+only for the Schloss and Young lab mouse colonies (Fig. S2A-B and Table
+S7). However, most of the vendors also clustered by experiment (Fig.
+S2C-D, F), suggesting there was some community variation between the 2
+experiments within each source. Thus, source was the factor that
+explained the most variation observed in the bacterial communities.
+Importantly with the exception of the Schloss and Young colonies, the
+community of each source clustered apart from one another suggesting
+each community had a unique response to clindamycin treatment and
+\emph{C. difficile} challenge.
+
+Since there was some variation in the microbiota structure of mice from
+the same source between experiments at baseline, we next looked at how
+similar the communities were within the same source and between sources
+in response to clindamycin treatment and \emph{C. difficile} challenge
+(Fig. 4) . The baseline communities varied most between experiments for
+Schloss, Young, and Envigo mice and variation between sources of mice
+was high (Fig. 4A). Clindamycin treatment reduced the variation between
+experiments within Schloss, Young and Jackson mice and some of the
+variation between sources diminished, particularly for the Schloss,
+Young, and Charles River mice (Fig. 4B). Post-infection, the community
+variation started to increase within sources of mice and variation
+between sources of mice started to increase towards previous levels
+(Fig. 4C). By using mice from multiple sources we were able to increase
+the number of microbiota communities we tested with the clindamycin
+\emph{C. difficile} colonization mouse model.
 
 After finding differences at the community level, we next identified the
-taxa that varied across sources of mice over the initial days of the
+bacteria that varied across sources of mice over the initial days of the
 experiment. We examined bacterial relative abundances at the operational
-taxonomic unit (OTU) and family levels, expecting the number of
-differences to be reduced at the family level due to the nature of
-bacterial taxonomy (45). Focusing on the baseline communities first,
-there were 268 OTUs and 20 families (Tables S8-9) with relative
-abundances that varied across colony sources. Clindamycin treatment
-reduced the number of taxa with relative abundances that varied across
-sources to 18 OTUs and 10 families (Tables S8-9). After \emph{C.
-difficile} challenge, there were 44 OTUs and 18 families (Tables S8-9)
-with significantly different relative abundances across sources, as the
+taxonomic unit (OTU) level. Focusing on the baseline communities first,
+there were 268 OTUs (Table S8) with relative abundances that varied
+across sources. Clindamycin treatment reduced the number of taxa with
+relative abundances that varied across sources to 18 OTUs (Table S8).
+After \emph{C. difficile} challenge, there were 44 OTUs (Table S8) with
+significantly different relative abundances across sources, as the
 communities started to recover from antibiotic treatment. In spite of
 the experimental perturbations that occurred during these 3 timepoints,
-there were 12 OTUs (Fig 5A-C) and 8 families with relative abundances
-that consistently varied across colony sources (Fig. 5D-F). Importantly,
-some of the taxa that consistently varied across sources also shifted
-with clindamycin treatment. For example, \emph{Proteus} increased after
-clindamycin treatment, but only in Taconic mice. \emph{Enterococcus} was
-primarily found only in mice purchased from commercial vendors and also
-increased after clindamycin treatment. In summary, mouse bacterial
-communities significantly varied according to colony source throughout
-the course of the experiment and a consistent subset of bacterial taxa
-remained different across sources regardless of clindamycin and \emph{C.
-difficile} challenge.
+there were 12 OTUs (Fig 5A-C) with relative abundances that consistently
+varied across sources. Importantly, some of the OTus that consistently
+varied across sources also shifted with clindamycin treatment. For
+example, \emph{Proteus} increased after clindamycin treatment, but only
+in Taconic mice. \emph{Enterococcus} was primarily found only in mice
+purchased from commercial vendors and also increased after clindamycin
+treatment. In summary, mouse bacterial communities varied significantly
+according to source throughout the course of the experiment and a
+consistent subset of bacterial taxa remained different across sources
+regardless of clindamycin and \emph{C. difficile} challenge.
 
 \textbf{Clindamycin treatment alters a subset of taxa that were found in
-all colony sources.} Although there were bacteria that consistently
-varied across colony sources, we also wanted to identify the bacteria
-that shifted after clindamycin treatment, regardless of colony source.
-By analyzing all mice that had sequence data from fecal samples
-collected at baseline and after clindamycin treatment, we identified 153
-OTUs and 18 families that were altered after clindamycin treatment (Fig.
-6 and Tables S10-11). Interestingly, when we compared the list of
-significant clindamycin impacted bacteria with the bacteria that
-consistently varied across groups over the initial 3 timepoints of our
-experiment, we found 3 OTUs (\emph{Lachnospiraceae} (OTU 130),
-\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) and 3
-families (\emph{Porphyromonadaceae}, \emph{Enterococcaceae},
-\emph{Lachnospiraceae}) overlapped (Fig. 5, Fig. 6C-D). These findings
-demonstrate that clindamycin has a consistent impact on the fecal
-bacterial communities of mice from all colony sources and only a subset
-of the taxa also varied across colony sources.
+all sources.} Although there were bacteria that consistently varied
+across sources, we also wanted to identify the bacteria that shifted
+after clindamycin treatment, regardless of source. By analyzing all mice
+that had sequence data from fecal samples collected at baseline and
+after clindamycin treatment, we identified 153 OTUs that were altered
+after clindamycin treatment (Fig. 6 and Tables S9). Interestingly, when
+we compared the list of significant clindamycin impacted bacteria with
+the bacteria that consistently varied across groups over the initial 3
+timepoints of our experiment, we found 3 OTUs (\emph{Lachnospiraceae}
+(OTU 130), \emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23))
+overlapped (Fig. 5, Fig. 6C-D). These findings demonstrate that
+clindamycin has a consistent impact on the fecal bacterial communities
+of mice from all sources and only a subset of the OTUs also varied
+across sources.
 
@@
 \emph{C. difficile} colonization status in mice.} After identifying taxa
-that varied by colony source, changed after clindamycin treatment, or
-both, we next wanted to determine which taxa were influencing the
-variation in \emph{C. difficile} colonization at day 7 (Fig. 1D, Fig.
-S1C). We trained L2-regularized logistic regression models with input
+that varied by source, changed after clindamycin treatment, or both, we
+next wanted to determine which taxa were influencing the variation in
+\emph{C. difficile} colonization at day 7 (Fig. 1D, Fig. S1C). We
+trained three L2-regularized logistic regression models with input
 bacterial community data from the baseline, post-clindamycin, and
@@
 better at predicting \emph{C. difficile} colonization status on day 7
-than random chance (all \emph{P} \(\le\) 5e-15; Table S12). However, the
-models trained with OTU level data generally performed better than those
-trained with family level data with the expception of the models based
-on the post-infection (day 1) communities (Fig. S3C-D). Interestingly,
-the model based on the post-clindamycin (day 0) community OTU data
-performed significantly better than all other models with an area under
-the receiving operator characteristic curve (AUROC) of 0.75
-(\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise comparisons;
-Table S13). Thus, we were able to use community bacterial relative
-abundance data alone to differentiate mice that had cleared \emph{C.
-difficile} before day 7 from the mice still colonized with \emph{C.
-difficile} at that timepoint. Interestingly, the model built with OTU
-relative abundance data post-clindamycin treatment had the best
-performance, suggesting how the bacterial community responds to
-clindamycin treatment has the greatest influence on subsequent \emph{C.
-difficile} colonization dynamics.
+than random chance (all \emph{P} \(\le\) 5.2e-31; Table S12).
+Interestingly, the model based on the post-clindamycin (day 0) community
+OTU data performed significantly better than all other models with an
+area under the receiving operator characteristic curve (AUROC) of 0.75
+(\emph{P}\textsubscript{FDR} \(\le\) 3.1e-11 for pairwise comparisons;
+Table S13). Thus, we were able to use bacterial relative abundance data
+alone to differentiate mice that had cleared \emph{C. difficile} before
+day 7 from the mice still colonized with \emph{C. difficile} at that
+timepoint. Interestingly, the model built with OTU relative abundance
+data post-clindamycin treatment had the best performance, suggesting how
+the bacterial community responds to clindamycin treatment has the
+greatest influence on subsequent \emph{C. difficile} colonization
+dynamics.
 
@@
 varied across sources (Fig. 7A). Next, the top 20 taxa from each model
-were compared to the list of taxa that varied across colony source at
-the same timepoint (Fig. 5 and Tables S8-9) and the taxa that were
-altered by clindamycin treatment (Fig. 6 and Table S10-11). We found a
-subset of OTUs and families that were important to the model and
-overlapped with bacteria that varied by either source, clindamycin
-treatment, or both (Fig. S4, S5A-C). Combining the overall results for
-the 3 OTU models identified 14 OTUs associated with source, 21 OTUs
-associated with clindamycin treatment, and 6 OTUs associated with both
-(Fig. 7B). Combining the overall results for the 3 family models
-identified 18 families associated with source, 14 families associated
-with clindamycin treatment and 8 families associated with both (Fig.
-S5D). Several OTUs (\emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
-Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)}) and families
-(\emph{Bacteroidaceae, Deferribacteraceae, Enterococcaceae,
-Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
-Verrucomicrobiaceae}) appeared across at least 2 models, so we examined
-how the relative abundances of these key taxa varied over the course of
-the experiment (Fig. 8 and Fig. S6). Throughout the experiment, there
-was at least 1 timepoint where relative abundances of these taxa
-significantly varied across sources (Table S16-17). Interestingly, there
-were no taxa that emerged as consistently enriched or depleted in mice
+were compared to the list of taxa that varied across source at the same
+timepoint (Fig. 5 and Tables S8-9) and the taxa that were altered by
+clindamycin treatment (Fig. 6 and Table S10-11). We found a subset of
+OTUs that were important to the model and overlapped with bacteria that
+varied by either source, clindamycin treatment, or both (Fig. S4,
+S5A-C). Combining the overall results for the 3 OTU models identified 14
+OTUs associated with source, 21 OTUs associated with clindamycin
+treatment, and 6 OTUs associated with both (Fig. 7B). Several OTUs
+(\emph{Bacteroides (OTU 2), Enterococcus (OTU 23), Enterobacteriaceae
+(OTU 1), Porphyromonadaceae (OTU 7)}) appeared across at least 2 models,
+so we examined how the relative abundances of these key taxa varied over
+the course of the experiment (Fig. 8). Throughout the experiment, there
+was at least 1 timepoint where relative abundances of these OTUs
+significantly varied across sources (Table S13). Interestingly, there
+were no OTUs that emerged as consistently enriched or depleted in mice
 that were colonized past 7 days post-infection with \emph{C. difficile}
@@
 infection. Together, these results suggest the initial bacterial
-communities and their responses to clindamycin influence on the time
-needed to clear \emph{C. difficile}.
+communities and their responses to clindamycin influence the clearance
+of \emph{C. difficile}.
 
@@
 
-By examining the \emph{C. difficile} colonization dynamics within mice
-from 6 different colony sources after perturbing the microbiota with
-clindamycin treatment, we were able to identify bacterial taxa that were
-unique to sources throughout the experiment as well as taxa that were
-universally impacted by clindamycin. We built L2 logistic regression
-models with baseline, post-clindamycin treatment, and post-infection
-fecal community data that successfully predicted whether mice cleared
-\emph{C. difficile} by 7 days post-infection better than random chance.
-We identified \emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
-Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)},
-\emph{Bacteroidaceae, Deferribacteraceae, Enterococcaceae,
-Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
-Verrucomicrobiaceae} (Fig. 8, Fig. S6) as candidate bacteria within
-these communities that were influencing variation in \emph{C. difficile}
-colonization dynamics since these bacteria were all important in the
-logistic regression models and varied by colony source, were impacted by
+By running our CDI model with mice from 6 different sources, we were
+able to identify bacterial taxa that were unique to sources throughout
+the experiment as well as taxa that were universally impacted by
+clindamycin. We trained L2 logistic regression models with baseline,
+post-clindamycin treatment, and post-infection fecal community data that
+could predict whether mice cleared \emph{C. difficile} by 7 days
+post-infection better than random chance. We identified
+\emph{Bacteroides (OTU 2), Enterococcus (OTU 23), Enterobacteriaceae
+(OTU 1), Porphyromonadaceae (OTU 7)} (Fig. 8) as candidate bacteria
+within these communities that were influencing variation in \emph{C.
+difficile} colonization dynamics since these bacteria were all important
+in the logistic regression models and varied by source, were impacted by
 clindamycin treatment, or both. Overall, our results demonstrate
@@
 
-Other groups have taken similar approaches by using mice from multiple
-colony sources to identify bacteria that either promote colonization
-resistance or increase susceptibility to enteric infections (22, 23,
-26--30). For example, in the context of \emph{Salmonella} infections,
+Other studies have taken similar approaches by using mice from multiple
+sources to identify bacteria that either promote colonization resistance
+or increase susceptibility to enteric infections (22, 23, 26--30). For
+example, in the context of \emph{Salmonella} infections,
 \emph{Enterobacteriaceae} and segmented filamentous bacteria have
@@
 against the more toxigenic \emph{C. difficile} VPI10463 (26). Given that
-we ordered mice from the same vendors, we checked all mice for
+we obtained mice from the same vendors, we checked all mice for
 endogenous \emph{C. difficile} by plating stool samples that were
@@
 other bacterial taxa mediated the variation in \emph{C. difficile}
-colonization across sources. Although all mice were susceptible to
-\emph{C. difficile} colonization, by following colonization over time we
-found Jackson, Taconic, and Envigo mice tended to remain colonized
+colonization across sources.
+
+Although all mice were susceptible to \emph{C. difficile} colonization,
+mice from Jackson, Taconic, and Envigo mice tended to remain colonized
 through at least 7 days post-infection. We identified a subset of
@@
 
-In the past variation between different CDI mouse model studies have
-been attributed to intestinal microbiome differences in mice across
-different institutional environments. For example, groups using the same
-clindamycin treatment and C57BL/6 mice had different \emph{C. difficile}
-outcomes, one having sustained colonization (32), while the other had
-transient (18). Baseline differences in the microbiota composition have
-been hypothesized to partially explain the differences in colonization
-outcomes and overall susceptibility to \emph{C. difficile} after
-treatment with the same antibiotic (13, 31). We have shown that mice
-from 6 different sources were all susceptible to \emph{C. difficile}
-630, suggesting the microbiota influences \emph{C. difficile} clearance
-more than susceptibility. Fortunately, the bacterial perturbations
-induced by clindamycin treatment have been well characterized and our
-findings agree with previous CDI mouse model work demonstrating
-\emph{Enterococcus} and \emph{Enterobacteriaceae} were associated with
-\emph{C. difficile} susceptibility and \emph{Porpyhromonadaceae},
-\emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and \emph{Turicibacter}
-were associated with resistance (19, 21, 32, 33, 41, 46--48). While we
-have demonstrated that susceptibility is uniform across vendors after
-clindamycin treatment, there could be different outcomes for either
-susceptibility or clearance in the case of other antibiotic treatments.
-The \emph{C. difficile} strain used could also be contributing to the
-variation in \emph{C. difficile} outcomes seen across different research
-groups (47). We found the time needed to naturally clear \emph{C.
-difficile} varied across sources of mice implying that at least in the
-context of the same perturbation, microbiota differences seemed to
-influence infection outcome more than susceptibility. More importantly,
-we were able to narrow down from all the variation observed across
-colony sources to a subset of bacterial taxa that were also important
-for predicting \emph{C. difficile} colonization status 7 days
-post-infection. Since all but 3 mice eventually cleared \emph{C.
-difficile} 630 by 9 days post-infection and the model built with the
-post-clindamycin OTU relative abundance data had the best performance,
-our results suggest clindamycin treatment had a large role in
-determining \emph{C. difficile} susceptibility and clearance in the
-mice.
+Differences in CDI mouse model studies have been attributed to different
+intestinal microbiotas of mice from different sources. For example,
+groups using the same clindamycin treatment and C57BL/6 mice had
+different \emph{C. difficile} outcomes, one having sustained
+colonization (32), while the other had transient (18), despite both
+using \emph{C. difficile} VPI 10643. Baseline differences in the
+microbiota composition have been hypothesized to partially explain the
+differences in colonization outcomes and overall susceptibility to
+\emph{C. difficile} after treatment with the same antibiotic (13, 31).
+The bacterial perturbations induced by clindamycin treatment have been
+well characterized and our findings agree with previous CDI mouse model
+work demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae}
+were associated with \emph{C. difficile} susceptibility and
+\emph{Porpyhromonadaceae}, \emph{Lachnospiraceae},
+\emph{Ruminococcaceae}, and \emph{Turicibacter} were associated with
+resistance (19, 21, 32, 33, 41, 46--48). While we have demonstrated that
+susceptibility is uniform across sources of mice after clindamycin
+treatment, there could be different outcomes for either susceptibility
+or clearance in the case of other antibiotic treatments. The \emph{C.
+difficile} strain used could also be contributing to the variation in
+\emph{C. difficile} outcomes seen across different research groups (47).
+We found the time needed to naturally clear \emph{C. difficile} varied
+across sources of mice implying that at least in the context of the same
+perturbation, microbiota differences seemed to influence infection
+outcome more than susceptibility. More importantly, we were able to
+reduce the variation observed across sources to identify a subset of
+OTUs that were also important for predicting \emph{C. difficile}
+colonization status 7 days post-infection. Since all but 3 mice
+eventually cleared \emph{C. difficile} 630 by 9 days post-infection and
+the model built with the post-clindamycin OTU relative abundance data
+had the best performance, our results suggest clindamycin treatment had
+a large role in determining \emph{C. difficile} susceptibility and
+clearance in the mice.
 
@@
 single donor or a single pool from multiple donors (51--54), which does
-not aid in the goal of figuring out how a variety of unique microbiotas
-influence susceptiblity to CDIs and adverse outcomes. Importantly, our
-study using mice from 6 different sources increased the variation
-between groups of mice compared to using 1 source alone, to better
-reflect the inter-individual microbiota variation observed in humans.
-Encouragingly, decreased \emph{Bifidobacterium}, \emph{Porphyromonas},
-\emph{Ruminococcaceae} and \emph{Lachnospiraceae} and increased
-\emph{Enterobacteriaceae}, \emph{Enterococcus}, \emph{Lactobacillus},
-and \emph{Proteus} have all been associated with human CDIs (7) and were
-well represented in our study, suggesting most of the mouse sources are
-suitable for gaining insights into microbiota associated factors
-influencing \emph{C. difficile} colonization and infections in humans.
-An important exception was \emph{Enterococcus}, which was primarily
-absent from the mice from University of Michigan colonies and
-\emph{Proteus}, which was only found in Taconic mice. Importantly, the
-fact that some CDI-associated bacteria were only found in a subset of
-mice has important implications for future CDI mouse model studies.
+not aid in the goal of modeling the interpersonal variation seen in
+humans to understand how the microbiota influences susceptiblity to CDIs
+and adverse outcomes. Importantly, our study using mice from 6 different
+sources increased the variation between groups of mice compared to using
+1 source alone, to better reflect the inter-individual microbiota
+variation observed in humans. Encouragingly, decreased
+\emph{Bifidobacterium}, \emph{Porphyromonas}, \emph{Ruminococcaceae} and
+\emph{Lachnospiraceae} and increased \emph{Enterobacteriaceae},
+\emph{Enterococcus}, \emph{Lactobacillus}, and \emph{Proteus} have all
+been associated with human CDIs (7) and were well represented in our
+study, suggesting most of the mouse sources are suitable for gaining
+insights into microbiota associated factors influencing \emph{C.
+difficile} colonization and infections in humans. An important exception
+was \emph{Enterococcus}, which was primarily absent from the mice from
+University of Michigan colonies and \emph{Proteus}, which was only found
+in Taconic mice. Importantly, the fact that some CDI-associated bacteria
+were only found in a subset of mice has important implications for
+future CDI mouse model studies.
 
@@
 factors, including age, diet, and immunity; all of which are also
-influenced by the microbiota. We have demonstrated that the ways
-baseline microbiotas from different mouse colony sources respond to
-clindamycin treatment influences the length of time mice remained
-colonized with \emph{C. difficile} 630. For those interested in
-dissecting the contribution of the microbiome to \emph{C. difficile}
-pathogenesis and treatments, using multiple sources of mice may yield
-more insights than a single model alone. Furthermore, for studies
-wanting to examine the interplay between a particular bacterial taxon
-such as \emph{Enterococcus} and \emph{C. difficile}, these results could
-serve as a resource for selecting which mice to order to address the
-question.
+influenced by the microbiota.
+
+We have demonstrated that the ways baseline microbiotas from different
+mouse sources respond to clindamycin treatment influences the length of
+time mice remained colonized with \emph{C. difficile} 630. For those
+interested in dissecting the contribution of the microbiome to \emph{C.
+difficile} pathogenesis and treatments, using multiple sources of mice
+may yield more insights than a single model alone. Furthermore, for
+studies wanting to examine the interplay between a particular bacterial
+taxon such as \emph{Enterococcus} and \emph{C. difficile}, these results
+could serve as a resource for selecting which mice to order to address
+the question. Using mice from multiple sources helps model the
+interpersonal microbiota variation among humans to aid our understanding
+of how the gut microbiota contributes to CDIs.
 
@@
 \href{https://github.com/SchlossLab/ML_pipeline_microbiome}{pipeline},
-Ana Taylor for help with some of the media preparation and sample
-collection, and Nicholas Lesniak for his critical feedback on the
-manuscript. We also thank members of Vincent Young's lab, particularly
-Kimberly Vendrov, for guidance with the \emph{C. difficile} infection
-mouse model and donating the mice. We also want to thank the Unit for
-Laboratory Animal Medicine at the University of Michigan for maintaining
-our mouse colony and providing the institutional support for our mouse
-experiments. Finally, we thank Kwi Kim, Austin Campbell, and Kimberly
-Vendrov for their help in maintaining the Schloss lab's anaerobic
-chamber.
+Ana Taylor for help with media preparation and sample collection, and
+Nicholas Lesniak for his critical feedback on the manuscript. We also
+thank members of Vincent Young's lab, particularly Kimberly Vendrov, for
+guidance with the \emph{C. difficile} infection mouse model and donating
+the mice. We also want to thank the Unit for Laboratory Animal Medicine
+at the University of Michigan for maintaining our mouse colony and
+providing the institutional support for our mouse experiments. Finally,
+we thank Kwi Kim, Austin Campbell, and Kimberly Vendrov for their help
+in maintaining the Schloss lab's anaerobic chamber.
 
@@
 Michigan Animal Care and Use Committee (IACUC) under protocol number
-PRO00006983. Female C57BL/7 mice were obtained from 6 different colony
-sources: The Jackson Laboratory, Charles River Laboratories, Taconic
-Biosciences, Envigo, and two colonies at the University of Michigan (the
-Schloss lab colony and the Young lab colony). The Young lab colony was
-originally established with mice purchased from Jackson, and the Schloss
-lab colony was later founded with mice donated from the Young lab. The 4
+PRO00006983. Female C57BL/7 mice were obtained from 6 different sources:
+The Jackson Laboratory, Charles River Laboratories, Taconic Biosciences,
+Envigo, and two colonies at the University of Michigan (the Schloss lab
+colony and the Young lab colony). The Young lab colony was originally
+established with mice purchased from Jackson, and the Schloss lab colony
+was established 10 years ago with mice donated from the Young lab. The 4
 groups of mice purchased from vendors were allowed to acclimate to the
@@
 1,000 times for alpha and beta diversity analyses. PCoAs were generated
-based on Theta YC distances. Permutational multivariate analysis of
-variance (PERMANOVA) was performed on mothur-generated Theta YC distance
-matrices with the adonis function in the vegan package (77) in R (78).
+based on \(\theta_{YC}\) distances. Permutational multivariate analysis
+of variance (PERMANOVA) was performed on mothur-generated
+\(\theta_{YC}\) distance matrices with the adonis function in the vegan
+package (77) in R (78).
 
@@
 timepoints of the experiment. Input bacterial community relative
-abundance data at either the OTU or family level from the baseline,
-post-clindamycin, and post-infection timepoints was used to generate 6
-classification models that predicted \emph{C. difficile} colonization
-status 7 days post-infection. The L2-regularized logistic regression
-models were trained and tested using the caret package (79) in R as
-previously described (80) with the exception that we used 60\% training
-and 40\% testing data splits for the cross-validation of the training
-data to select the best cost hyperparameter and the testing of the held
-out test data to measure model performance. The modified training to
-testing ratio was selected to accommodate the small number of samples in
-the dataset. Code was modified from
+abundance data at the OTU level from the baseline, post-clindamycin, and
+post-infection timepoints was used to generate 6 classification models
+that predicted \emph{C. difficile} colonization status 7 days
+post-infection. The L2-regularized logistic regression models were
+trained and tested using the caret package (79) in R as previously
+described (80) with the exception that we used 60\% training and 40\%
+testing data splits for the cross-validation of the training data to
+select the best cost hyperparameter and the testing of the held out test
+data to measure model performance. The modified training to testing
+ratio was selected to accommodate the small number of samples in the
+dataset. Code was modified from
 \url{https://github.com/SchlossLab/ML_pipeline_microbiome} to update the
@@
 (OTU 20) was removed. Bacterial relative abundances that varied across
-sources at the OTU and family taxonomic levels were identified with the
-Kruskal-Wallis test with Benjamini-Hochberg correction for testing all
-identified taxa at each level, followed by pairwise Wilcoxon comparisons
-with Benjamini-Hochberg correction. Taxa impacted by clindamycin
-treatment were identified using the Wilcoxon signed rank test with
-matched pairs of mice samples for day -1 and day 0. To determine whether
-classification models had better performance (test AUROCs) than random
-chance (0.5), we used the one-sample Wilcoxon signed rank test. To
-examine whether there was an overall difference in predictive
-performance across the 6 classification models we used the
-Kruskal-Wallis test followed by pairwise Wilcoxan comparisons with
-Benjamini-Hochberg correction for multiple hypothesis testing. The
-tidyverse package was used to wrangle and graph data (v 1.3.0) (81).
+sources at the OTU level were identified with the Kruskal-Wallis test
+with Benjamini-Hochberg correction for testing all identified OTUs,
+followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg
+correction. OTUs impacted by clindamycin treatment were identified using
+the Wilcoxon signed rank test with matched pairs of mice samples for day
+-1 and day 0. To determine whether classification models had better
+performance (test AUROCs) than random chance (0.5), we used the
+one-sample Wilcoxon signed rank test. To examine whether there was an
+overall difference in predictive performance across the 6 classification
+models we used the Kruskal-Wallis test followed by pairwise Wilcoxan
+comparisons with Benjamini-Hochberg correction for multiple hypothesis
+testing. The tidyverse package was used to wrangle and graph data (v
+1.3.0) (81).
 
@@
 generating this manuscript is available at
-\url{https://github.com/SchlossLab/Tomkovich_vendor_difs_XXXX_2020}.
+\url{https://github.com/SchlossLab/Tomkovich_Vendor_XXXX_2020}.
 
@@
 \includegraphics{figure_2.pdf} \textbf{Figure 2. Differences in
-microbial richness and diversity across mouse colony sources emerge
-after clindamycin treatment and infection.} A-F. Number of observed OTUs
-and Shannon diversity index values at baseline: day -1 (A-B), after
+microbial richness and diversity across mouse sources emerge after
+clindamycin treatment and infection.} A-F. Number of observed OTUs and
+Shannon diversity index values at baseline: day -1 (A-B), after
 clindamycin: day 0 (C-D) and post-infection: day 1 (E-F) timepoints of
@@
 
-\includegraphics{figure_3.pdf} \textbf{Figure 3. Mouse colony source is
-the variable that explains most of the variation observed in the
-baseline, post-clindamycin, and post-infection bacterial communities.}
-A-C. Principal Coordinates Analysis of Theta YC distances from stools
+\includegraphics{figure_3.pdf} \textbf{Figure 3. Mouse source is the
+variable that explains most of the variation observed in the baseline,
+post-clindamycin, and post-infection bacterial communities.} A-C.
+Principal Coordinates Analysis of \(\theta_{YC}\) distances from stools
 collected at baseline (A), post-clindamycin (B), and post-infection (C)
@@
 variation across mouse sources is diminished by clindamycin treatment}
-A-C. Boxplots of the Theta YC distances of the 6 sources of mice
+A-C. Boxplots of the \(\theta_{YC}\) distances of the 6 sources of mice
 relative to mice within the same source and experiment, mice within the
@@
 \includegraphics{figure_5.pdf} \textbf{Figure 5. A subset of bacteria
-consistently vary across mouse colony sources despite clindamycin
-perturbation and \emph{C. difficile} challenge.} A-C: plots highlighting
-the median (point) and interquantile range (colored lines) of the
-relative abundances for the 12 OTUs that consistently varied across
-sources of mice at the baseline (A), post-clindamycin (B), and
-post-infection (C) timepoints of the experiment. D-F: plots highlighting
-the median (point) and interquantile range (colored lines) of the
-relative abundances for the 8 families that consistently varied across
-sources of mice at the baseline (D), post-clindamycin (E), and
-post-infection (F) timepoints of the experiment. For each timepoint
-bacteria with differential relative abundances across sources of mice
-were identified by Kruskal-Wallis test at the family and OTU level with
-Benjamini-Hochberg correction for testing all identified taxa at the
-respective level (Table S8-9). The grey vertical line indicates the
-limit of detection for A-F.
+consistently vary across sources despite clindamycin perturbation and
+\emph{C. difficile} challenge.} A-C: plots highlighting the median
+(point) and interquantile range (colored lines) of the relative
+abundances for the 12 OTUs that consistently varied across sources of
+mice at the baseline (A), post-clindamycin (B), and post-infection (C)
+timepoints of the experiment. For each timepoint OTus with differential
+relative abundances across sources of mice were identified by
+Kruskal-Wallis test with Benjamini-Hochberg correction for testing all
+identified OTUs (Table S8). The grey vertical line indicates the limit
+of detection.
 
@@
 \includegraphics{figure_6.pdf} \textbf{Figure 6. Clindamycin treatment
-has the same effects on a subset of taxa regardless of colony source.}
-A-B: plots highlighting the median (point) and interquantile range
-(colored lines) of the top 10 most significant (adjusted \emph{P} value
+has the same effects on a subset of taxa regardless of source.} A-B:
+plots highlighting the median (point) and interquantile range (colored
+lines) of the top 10 most significant (adjusted \emph{P} value
 \textless{} 0.05) OTUs with relative abundances that changed after
-clindamycin treatment. C-D: plots highlighting the median (point) and
-interquantile range (colored lines) of the top 10 most significant
-families with relative abundances that changed after clindamycin
-treatment. Data were analyzed by Wilcoxon signed rank test limited to
-mice that had paired sequence data for day -1 and 0 (N = 31). Tests were
-performed at the OTU and family levels with Benjamini-Hochberg
-correction for testing all identified OTUs and families. See Table
-S10-11 for complete list of OTUs and families significantly impacted by
-clindamycin treatment. The grey vertical line indicates the limit of
-detection for A-D.
+clindamycin treatment. Data were analyzed by Wilcoxon signed rank test
+limited to mice that had paired sequence data for day -1 and 0 (N = 31).
+Tests were performed at the OTU level with Benjamini-Hochberg correction
+for testing all identified OTUs. See Table S9 for complete list of OTUs
+significantly impacted by clindamycin treatment. The grey vertical line
+indicates the limit of detection.
 
@@
 \includegraphics{figure_8.pdf} \textbf{Figure 8: Key OTUs vary across
-sources throughout the experiment.} A-C. Relative abundances of bold
+sources throughout the experiment.} A-D. Relative abundances of bold
 OTUs from Fig. 7A that were important for at least two classification
-models are shown over time. A. \emph{Enterococcus} (OTU 23), which
-significantly varied across sources and was impacted by clindamycin
-treatment. B. \emph{Bacteroides} (OTU 2), which varied across sources
-throughout the experiment. C. \emph{Enterobacteriaceae} (OTU 1) and
-\emph{Porphyromonadaceae} (OTU 7) were significantly impacted by
+models are shown over time. A. \emph{Bacteroides} (OTU 2), which varied
+across sources throughout the experiment. B-C. \emph{Enterobacteriaceae}
+(B) and \emph{Enterococcus} (C), which significantly varied across
+sources and were impacted by clindamycin treatment. D.
+\emph{Porphyromonadaceae} (OTU 7), which was significantly impacted by
 clindamycin treatment and examining relative abundance dynamics over the
-course of the experiment indicated timepoints where relative abundances
+course of the experiment, revealed timepoints where relative abundances
 also significantly varied across sources of mice. Symbols represent the
@@
 statistically significant by Kruskal-Wallis test with Benjamini-Hochberg
-correction for testing across multiple days (Table S16) are identified
+correction for testing across multiple days (Table S13) are identified
 by the asterisk(s) above each timepoint (*, P \textless{} 0.05).
@@
 communities from University of Michigan mice significantly vary between
-experiments.} A-F. PCoA of Theta YC distances for the baseline fecal
-bacterial communities within each source of mice. Each symbol represents
-a stool sample from an individual mouse with color corresponding to
-experiment and shape representing cage mates. PERMANOVA was performed
-within each group to examine the contributions of experiment and cage to
-observed variation. Experiment number and cage only significantly
-explained observed variation for mice from the Schloss (combined
-R\textsuperscript{2} = 0.99; \emph{P} \(\le\) 0.033) and Young (combined
-R\textsuperscript{2} = 0.95; \emph{P} \(\le\) 0.027) lab colonies (Table
-S7).
+experiments.} A-F. PCoA of \(\theta_{YC}\) distances for the baseline
+fecal bacterial communities within each source of mice. Each symbol
+represents a stool sample from an individual mouse with color
+corresponding to experiment and shape representing cage mates. PERMANOVA
+was performed within each group to examine the contributions of
+experiment and cage to observed variation. Experiment number and cage
+only significantly explained observed variation for mice from the
+Schloss (combined R\textsuperscript{2} = 0.99; \emph{P} \(\le\) 0.033)
+and Young (combined R\textsuperscript{2} = 0.95; \emph{P} \(\le\) 0.027)
+lab colonies (Table S7).
 
@@
 CFU) based on CFU g/stool data from 7 days post-infection. B. \emph{C.
-difficile} CFU status on Day 7 within each mouse colony source. N = 5-9
-mice per group. C-D. L2-regularized logistic regression classification
-model area under the receiving operator characteristic curve (AUROCs) to
+difficile} CFU status on Day 7 within each mouse source. N = 5-9 mice
+per group. C. L2-regularized logistic regression classification model
+area under the receiving operator characteristic curve (AUROCs) to
 predict \emph{C. difficile} CFU on day 7 post-infectoin (Fig. 1D, Fig.
-S3) based on the community relative abundances at baseline (day -1),
-post-clindamycin (day 0), and post-infection (day 1) at either the OTU
-(C) or family (D) level. All models performed better than random chance
-(AUROC = 0.5; all \emph{P} \(\le\) 5e-15; Table S12) and the model built
-with post-clindamycin treated bacterial OTU relative abundances had the
-best performance ((\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for all
-pairwise comparisons; Table S13). For lists of the 20 taxa that were
-ranked as most important to each model, see Tables S14-15.
+S3) based on the OTU community relative abundances at baseline (day -1),
+post-clindamycin (day 0), and post-infection (day 1). All models
+performed better than random chance (AUROC = 0.5; all \emph{P} \(\le\)
+5.2e-31; Table S12) and the model built with post-clindamycin treated
+bacterial OTU relative abundances had the best performance
+((\emph{P}\textsubscript{FDR} \(\le\) 3.1e-11 for all pairwise
+comparisons; Table S11). For list of the 20 OTUs that were ranked as
+most important to each model, see Table S12.
 
@@
 classification models based on baseline, post-clindamycin treatment, or
-post-infection community data vary by mouse colony source, clindamycin
-treatment, or both.} A-C. Venn diagrams of top 20 important OTUs from
-baseline (A), post-clindamycin treatment (B), and post-infection (C)
-classification models (Table S14) that overlapped with OTUs that varied
-across vendors at baseline, were impacted by clindamycin treatment, or
-both. Bold OTUs signify OTUs that were important to more than 1
+post-infection community data vary by source, clindamycin treatment, or
+both.} A-C. Venn diagrams of top 20 important OTUs from baseline (A),
+post-clindamycin treatment (B), and post-infection (C) classification
+models (Table S12) that overlapped with OTUs that varied across vendors
+at the corresponding timepoint, were impacted by clindamycin treatment,
+or both. Bold OTUs signify OTUs that were important to more than 1
 classification model.
-
-\newpage
-
-\includegraphics{figure_S5.pdf} \textbf{Figure S5. Key families from
-classification models based on baseline, post-clindamycin treatment, or
-post-infection community data vary by mouse colony source, clindamycin
-treatment, or both.} A-C. Venn diagrams of top 20 important families
-from baseline (A), post-clindamycin treatment (B), and post-infection
-(C) classification models (Table S15) that overlapped with families that
-varied across vendors after clindamycin, were impacted by clindamycin
-treatment, or both. D. Venn diagrams that combines A-C summaries of
-familes that were important to the day -1, 0, and 1 classification
-models (Table S15) and either overlapped with familes that varied across
-vendors at the same timepoint, were impacted by clindamycin treatment,
-or both. Bold families signify families that were important to more than
-1 classification model.
-
-\newpage
-
-\includegraphics{figure_S6.pdf} \textbf{Figure S6. Key families vary
-across sources throughout experiment.} Relative abundances of bold
-families from Fig. S5D that were important for at least two
-classification models are shown over time. A. \emph{Enterococcaceae} and
-\emph{Lachnospiraceae}, which significantly varied across sources and
-were impacted by clindamycin treatment. B. \emph{Bacteroidaceae} and
-\emph{Deferribacteraceae}, which varied across sources throughout the
-experiment. C. \emph{Bifidobacteriaceae}, \emph{Coriobacteriaceae},
-\emph{Ruminococcaceae}, and \emph{Verrucomicrobiaceae} were
-significantly impacted by clindamycin treatment. Examining the relative
-abundance dynamics throughout the experiment, identified timepoints
-where relative abundances also significantly varied across sources of
-mice. Symbols represent the relative abundance data for an individual
-mouse, circles represent mice that cleared \emph{C. difficile} by day 7,
-X-shapes represent mice that were still colonized with \emph{C.
-difficile}, and open circles represent mice that did not have \emph{C.
-difficile} CFU counts for day 7 post-infection. Colored lines indicate
-the median relative abundances for each source. The gray horizontal line
-represents the limit of detection. Timepoints where differences across
-sources of mice were statistically significant by Kruskal-Wallis test
-with Benjamini-Hochberg correction for testing across multiple days
-(Table S17) are identified by the asterisk(s) above each timepoint (*, P
-\textless{} 0.05).
 
@@
 All supplemental material is available at:
-\url{https://github.com/SchlossLab/Tomkovich_vendor_difs_XXXX_2020/submission}.
+\url{https://github.com/SchlossLab/Tomkovich_Vendor_XXXX_2020/submission}.
 
 \textbf{Movie S1. Large shifts in bacterial community structure occurred
-after clindamycin and \emph{C. difficile} infection.} PCoA of Theta YC
-distances animated from 0 through 9 days post-infection. PERMANOVA
-analysis indicated colony source was the variable that explained the
+after clindamycin and \emph{C. difficile} infection.} PCoA of
+\(\theta_{YC}\) distances animated from 0 through 9 days post-infection.
+PERMANOVA analysis indicated source was the variable that explained the
 most observed variation across fecal communities (source
@@
 
-\textbf{Table S9. Families with relative abudances that significantly
-vary across sources at baseline, post-clindamycin, or post-infection
-timepoints.}
-
-\textbf{Table S10. OTUs with relative abudances that significantly
+\textbf{Table S9. OTUs with relative abudances that significantly
 changed after clindamycin treatment.}
 
-\textbf{Table S11. Families with relative abudances that significantly
-changed after clindamycin treatment.}
-
-\textbf{Table S12. Statistical results of L2-regularized logistic
+\textbf{Table S10. Statistical results of L2-regularized logistic
 regression model performances compared to random chance.}
 
-\textbf{Table S13. Pairwise Wilcoxan results comparing all 6
+\textbf{Table S11. Pairwise Wilcoxan results comparing all 3
 L2-regularized logistic regression model performances.}
 
-\textbf{Table S14. Top 20 most important OTUs for each of the 3
+\textbf{Table S12. Top 20 most important OTUs for each of the 3
 L2-regularized logistic regression models based on OTU relative
@@
 
-\textbf{Table S15. Top 20 most important families for each of the 3
-L2-regularized logistic regression models based on OTU relative
-abundance data.}
-
-\textbf{Table S16. OTUs with relative abudances that significantly
+\textbf{Table S13. OTUs with relative abudances that significantly
 varied across sources of mice on at least 1 day of the experiment by
@@
 
-\textbf{Table S17. Families with relative abudances that significantly
-varied across sources of mice on at least 1 day of the experiment by
-Kruskal-Wallis test.}
-
 \end{document}

### 2020-07-08 code/figure_8.R

@@ -1,8 +1,8 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/Enterococcus (OTU 23)_time.png")
-b <- ggdraw() + draw_image("results/figures/Bacteroides (OTU 2)_time.png")
-c <- ggdraw() + draw_image("results/figures/Enterobacteriaceae (OTU 1)_time.png")
+a <- ggdraw() + draw_image("results/figures/Bacteroides (OTU 2)_time.png")
+b <- ggdraw() + draw_image("results/figures/Enterobacteriaceae (OTU 1)_time.png")
+c <- ggdraw() + draw_image("results/figures/Enterococcus (OTU 23)_time.png")
 d <- ggdraw() + draw_image("results/figures/Porphyromonadaceae (OTU 7)_time.png")
 
 

### 2020-07-08 code/functions.R
Add libraries for OTU label formatting on plots

@@ -11,6 +11,8 @@ library(ggpubr)
 library(ggforce) #Use geom_circle to make venn diagrams
 library(gganimate) #Use to create PCoA animation
 library(writexl) #For writing supplemental tables
+library(ggtext)
+library(glue)
 
 ### Load in metadata and make sure experiment and vendor columns are treated as factors
 metadata <- read_csv("data/process/vendor_metadata.csv") %>%

### 2020-07-08 code/functions.R
Change color scheme for Jackson mice

@@ -136,9 +136,8 @@ dropped_by_miseq <- nrow(sequenced)-nrow(matching)
 #IDs of the 19 samples that were dropped after subsampling to 5437 sequences
 dropped_by_miseq_ids <- anti_join(sequenced, matching, by = "id") %>% pull(id)
 
-
 #Define color scheme----
-color_scheme <- c("#1f78b4", "#e6ab02", "#d95f02", "#e7298a", "#7570b3", "#1b9e77") #Adapted from http://colorbrewer2.org/#type=qualitative&scheme=Dark2&n=6
+color_scheme <- c("#1f78b4", "#e6ab02", "#323232", "#e7298a", "#7570b3", "#1b9e77") #Adapted from http://colorbrewer2.org/#type=qualitative&scheme=Dark2&n=6. Previous color for Jackson mice: #d95f02
 color_vendors <- levels(metadata$vendor)
 
 #Define shape scheme to differentiate experiment 1 and 2----

### 2020-07-09 submission/manuscript.Rmd
Figure edits before reorganization

@@
 difficile* strain used could also be contributing to the variation in *C.
-difficile* outcomes seen across different research groups [@Lawley2012].
+difficile* outcomes seen across different research groups.
+For example, a group found differential colonization outcomes after clindamycin treatment, with *C.
+difficile* 630 and M68 eventually becoming undectable while BI-7 remained detectable up to 70 days post-treatment [@Lawley2012].
 We found the time needed to naturally clear *C.

### 2020-07-09 code/cfu_plot.R

@@ -62,13 +62,13 @@ summarize_plot <- function(df){
     group_by(vendor, day) %>% 
     summarize(median_cfu = median(cfu, na.rm = TRUE))
   ggplot(NULL) +
-    geom_point(df, mapping = aes(x = day, y = cfu, color= vendor, fill = vendor, shape = experiment), alpha = .2, size = 1.5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
-    geom_line(median_summary, mapping = aes(x = day, y = median_cfu, color = vendor), size = 1.5) +
+    geom_point(df, mapping = aes(x = day, y = cfu, color= vendor, fill = vendor, shape = experiment), alpha = .2, size = 1.5, position = position_dodge(width = 0.6)) +
+    geom_line(median_summary, mapping = aes(x = day, y = median_cfu, color = vendor), show.legend = FALSE, size = 1.5) +
     scale_colour_manual(name=NULL,
                         values=color_scheme,
                         breaks=color_vendors,
                         labels=color_vendors)+
-    scale_shape_manual(name=NULL,
+    scale_shape_manual(name="Experiment",
                        values=shape_scheme,
                        breaks=shape_experiment,
                        labels=shape_experiment) +
@@ -92,7 +92,7 @@ save_plot(filename = paste0("results/figures/exp1_cfu_time.png"), plot = exp1_cf
 exp2_cfu <- summarize_plot(cfu_data_final %>% filter(experiment == 2))+
   ggtitle("2nd experiment")+ #Title plot
   theme(plot.title = element_text(hjust = 0.5),
-        legend.position = "bottom") #Center plot titile
+        legend.position = "none") #Center plot titile
 save_plot(filename = paste0("results/figures/exp2_cfu_time.png"), plot = exp2_cfu, base_aspect_ratio = 2)
 
 #CFU over time plot with astericks on days where cfu varied significantly across sources of mice using annotate()----
@@ -114,13 +114,13 @@ median_summary <- cfu_data_final %>%
   group_by(vendor, day) %>% 
   summarize(median_cfu = median(cfu, na.rm = TRUE))
 cfu_stats <- ggplot(NULL) +
-  geom_point(cfu_data_final, mapping = aes(x = day, y = cfu, color= vendor, fill = vendor, shape = experiment), alpha = .2, size = 1.5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
[…]

### 2020-07-09 code/figure_6.R

@@ -1,8 +1,7 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/clind_impacted_otus_plot_dn1.png")
-b <- ggdraw() + draw_image("results/figures/clind_impacted_otus_plot_d0.png")
+a <- ggdraw() + draw_image("results/figures/clind_impacted_otus_plot.png")
 
-plot_grid(a, b, labels = "AUTO", label_size = 12, ncol=2)+
+plot_grid(a, labels = "", label_size = 12, ncol=1)+
   ggsave("results/figures/figure_6.pdf", width=8, height=5.5)+
   ggsave("submission/figure_6.pdf", width=8, height=5.5)

### 2020-07-09 code/figure_7.R

@@ -8,6 +8,6 @@ top_half <- plot_grid(a, labels = c("A"), label_size = 12, ncol=1)
 b <- ggdraw() + draw_image("results/figures/venn_overall_otus.png")
 
 plot_grid(top_half, b, labels = c("", "B"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_7.pdf", width=6, height=8)+
-  ggsave("submission/figure_7.pdf", width=6, height=8)
+  ggsave("results/figures/figure_7.pdf", width=6, height=9)+
+  ggsave("submission/figure_7.pdf", width=6, height=9)
 

### 2020-07-09 code/figure_8.R

@@ -7,5 +7,5 @@ d <- ggdraw() + draw_image("results/figures/Porphyromonadaceae (OTU 7)_time.png"
 
 
 plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_8.pdf", width=6, height=4)+
-  ggsave("submission/figure_8.pdf", width=6, height=4)
+  ggsave("results/figures/figure_8.pdf", width=7.5, height=4)+
+  ggsave("submission/figure_8.pdf", width=7.5, height=4)

### 2020-07-09 code/pcoa_stats.R

@@ -156,7 +156,9 @@ plot_pcoa <- function(df, timepoint){
                        values=shape_scheme,
                        breaks=shape_experiment,
                        labels=shape_experiment) +
-    coord_fixed() + 
+    coord_fixed() +
+    xlim(-0.7, 0.5)+
+    ylim(-0.6, 0.7)+
     labs(x="PCoA 1",
          y="PCoA 2",
          color= "Source") +
@@ -279,8 +281,14 @@ write_tsv("data/process/adonis_dn1-1.tsv") %>%
 #Function to plot pcoa data for each source of mice at day -1, colored according to experiment with shapes to differentiate cage----
 pcoa_vendor <- function(df, label){
   plot <- ggplot(df, aes(x=axis1, y=axis2, color = experiment, shape = cage)) +
-    geom_point(size=4, alpha = 0.4) +
+    geom_point(size=4, alpha = 0.4, fill = "white", stroke = 1.5) +
     coord_fixed() + 
+    xlim(-0.55, 0.5)+
+    ylim(-0.45, 0.52)+
+    scale_shape_manual(name="Cage",
+                       values=c(21, 24),
+                       breaks=c(1, 2),
+                       labels=c(1, 2)) +
     labs(title = label,
          x="PCoA 1",
          y="PCoA 2",
@@ -320,7 +328,8 @@ s_dn1_axis2 <- s_dn1_axis_labels %>% filter(axis == 2) %>% pull(loading) %>% rou
 
 s_dn1 <- pcoa_vendor(s_dn1_pcoa, "Schloss")+
   labs(x = paste("PCoA 1 (", s_dn1_axis1, "%)", sep = ""), #Anotations for each axis from loadings file
-       y = paste("PCoA 2 (", s_dn1_axis2,"%)", sep = ""))
+       y = paste("PCoA 2 (", s_dn1_axis2,"%)", sep = ""))+
+  theme(legend.position = "none")
 save_plot(filename = paste0("results/figures/pcoa_dn1_schloss.png"), s_dn1)
 
 # Read in thetayc distance matrix that represents Day -1 Young samples----
@@ -352,7 +361,8 @@ y_dn1_axis2 <- y_dn1_axis_labels %>% filter(axis == 2) %>% pull(loading) %>% rou
 
 y_dn1 <- pcoa_vendor(y_dn1_pcoa, "Young")+
[…]

### 2020-07-09 code/taxa_stats.R

@@ -59,16 +59,17 @@ agg_otu_data <- inner_join(agg_otu, taxa_info, by="key") %>%
   mutate(key = paste(key,")", sep="")) %>% 
   select(-otu, -Taxonomy) %>% 
   rename(otu=key) %>% 
-  mutate(otu=paste0(gsub('TU0*', 'TU ', otu))) 
+  mutate(otu=paste0(gsub('TU0*', 'TU ', otu))) %>% 
+  separate(otu, into = c("bactname", "OTUnumber"), sep = "\\ [(]", remove = FALSE) %>% #Add columns to separate bacteria name from OTU number to utilize ggtext so that only bacteria name is italicized
+  mutate(otu_name = glue("*{bactname}* ({OTUnumber}")) #Markdown notation so that only bacteria name is italicized
 
 #Hypothesis testing & plotting----
 
 #Function to pull significant taxa (adjusted p value < 0.05) after statistical analysis
 pull_significant_taxa <- function(dataframe, taxonomic_level){
-  taxonomic_level <- enquo(taxonomic_level) #Part of transformation of taxonomic_level argument into a column name
   dataframe %>% 
     filter(p.value.adj <= 0.05) %>% 
-    pull(!!taxonomic_level) #!!Completes the transformation of taxonomic_level argument into a column name
+    pull({{ taxonomic_level }}) #Embracing transforms taxonomic_level argument into a column name
 }
 
 #List of days with sequence data----
@@ -117,9 +118,8 @@ plot_otus_dx <- function(otus, timepoint){
   agg_otu_data %>% 
     filter(otu %in% otus) %>% 
     filter(day == timepoint) %>% 
-    mutate(otu=factor(otu, levels=otus)) %>% 
     mutate(agg_rel_abund = agg_rel_abund + 1/10874) %>% # 10,874 is 2 times the subsampling parameter of 5437
-    ggplot(aes(x= otu, y=agg_rel_abund, color=vendor))+
+    ggplot(aes(x= otu_name, y=agg_rel_abund, color=vendor))+
     scale_colour_manual(name=NULL,
                         values=color_scheme,
                         breaks=color_vendors,
@@ -137,7 +137,7 @@ plot_otus_dx <- function(otus, timepoint){
     theme_classic()+
     theme(plot.title=element_text(hjust=0.5),
           legend.position = "bottom",
[…]

### 2020-07-09 code/venn_diagram_comparisons.R

@@ -12,23 +12,44 @@ ggplot(df.venn, aes(x0 = x, y0 = y, r = 19, fill = labels)) +
 
 #OTUs identified in logistic regression classification models (20 OTUs with the highest ranking for each model)
 interp_otus <- read_tsv("data/process/combined_top20_otus_all_models.tsv")
-interp_otus_dn1 <- interp_otus  %>% filter(model_input_day == -1) %>% pull(OTU)
-interp_otus_d0 <- interp_otus  %>% filter(model_input_day == 0) %>% pull(OTU)
-interp_otus_d1 <- interp_otus  %>% filter(model_input_day == 1) %>% pull(OTU)
+
+#Function to get top 20 OTUs from each model:
+get_interp_otus <- function(timepoint){
+  otus <- interp_otus  %>% filter(model_input_day == timepoint) %>% 
+    separate(OTU, into = c("bactname", "OTUnumber"), sep = "\\ [(]", remove = FALSE) %>% #Add columns to separate bacteria name from OTU number to utilize ggtext so that only bacteria name is italicized
+    mutate(otu_name = glue("*{bactname}* ({OTUnumber}")) %>% #Markdown notation so that only bacteria name is italicized
+    pull(otu_name)
+}
+#format_otus <- (paste(otus, collapse = "<br>")) #Format so that there's line breaks separating each OTU
+#return(format_otus)
+
+interp_otus_dn1 <- get_interp_otus(-1)
+interp_otus_d0 <- get_interp_otus(0)
+interp_otus_d1 <- get_interp_otus(1)
 interp_combined <- c(interp_otus_dn1, interp_otus_d0, interp_otus_d1)
 
 #OTUs that vary across sources of mice on day -1, 0, or 1
 sig_otu_source <- read_tsv("data/process/otu_stats_dn1to1_combined.tsv")
-`sig_otu_day-1` <- sig_otu_source %>% filter(day == -1) %>% pull(otu)
-`sig_otu_day0` <- sig_otu_source %>% filter(day == 0) %>% pull(otu)
-`sig_otu_day1` <- sig_otu_source %>% filter(day == 1) %>% pull(otu)
+#Function to get OTUs that vary across sources for each timepoint:
+get_source_otus <- function(timepoint){
+  otus <- sig_otu_source  %>% filter(day == timepoint) %>% 
[…]

### 2020-07-09 code/weight_plot.R

@@ -124,7 +124,8 @@ weight_stats <-  ggplot(NULL) +
                        limits = c(-1.5, 9.5)) +
     theme(text = element_text(size = 16))+  # Change font size for entire plot
     annotate("text", y = y_position, x = x_annotation, label = label, size =8)+
-    theme_classic()
+    theme_classic()+
+    theme(legend.position = "none") #Get rid of legend 
 save_plot(filename = paste0("results/figures/weight_over_time.png"), plot = weight_stats, base_aspect_ratio = 2)
 
 #Add p.value manually to timepoints of interest with ggpubr stat_pvalue_manual() function: 2 ----

### 2020-07-10 code/cfu_plot.R
Reorganize figures and tables

@@ -26,7 +26,7 @@ cfu_kruskal_stats_adjust <- cfu_kruskal_stats %>%
   arrange(p.value.adj) %>% 
   write_tsv("data/process/cfu_stats_all_days.tsv")
 #Also write to supplemental table excel file
-cfu_kruskal_stats_adjust %>% write_xlsx("submission/table_S1_cfu_kruskal-wallis.xlsx", format_headers = FALSE)
+cfu_kruskal_stats_adjust %>% write_xlsx("submission/table_S4_cfu_kruskal-wallis.xlsx", format_headers = FALSE)
 
 #List significant days after BH adjustment of p-values:
 sig_cfu_days <- cfu_kruskal_stats_adjust %>% 

### 2020-07-10 code/class_interpretation.R

@@ -297,7 +297,7 @@ day1_model_top20 <- get_table_of_top_20(interp_D1_60) %>%
 combined_top20_all_models <- rbind(dayn1_model_top20, day0_model_top20, day1_model_top20) 
 write_tsv(combined_top20_all_models, path = "data/process/combined_top20_otus_all_models.tsv") 
 #Also write results to supplemental table excel file
-combined_top20_all_models %>% write_xlsx("submission/table_S12_combined_top20_otus_each_model.xlsx", format_headers = FALSE)
+combined_top20_all_models %>% write_xlsx("submission/table_S8_combined_top20_otus_each_model.xlsx", format_headers = FALSE)
 
 #OTUs important across models (found within at least 2 models):
 overlapping_across_models <- combined_top20_all_models %>% 

### 2020-07-10 code/compare_models.R

@@ -18,14 +18,12 @@ all <- rbind(otu_dayn1, otu_day0, otu_day1) %>%
 #Statistical analysis:
 set.seed(19881117) #Same seed used for mothur analysis
 #Compare test AUCs for each model to 0.5. Wilcoxan signed rank test. Null hypothesis is the distribution of test AUCs for each model is symmetric about mu (0.5)
-all %>% 
+all_vs_random <- all %>% 
   group_by(model_name) %>% 
   summarize(test=list(tidy(wilcox.test(test_aucs, mu=0.5, paired = FALSE, alternative="greater"))), 
             summary=list(tidy(summary(test_aucs)))) %>% 
   unnest() %>% 
   write_tsv("data/process/classification_to_random.tsv") 
-  #Also write results to supplemental table excel file
-write_xlsx(all, "submission/table_S10_classification_to_random.xlsx", format_headers = FALSE)
 #All models perform significantly better than random (AUC of 0.5)
 
 #Kruskal-wallis test comparing test AUCs from all 3 models: 
@@ -41,9 +39,7 @@ model_stats_pairwise <-
   select(compare, p.value.adj) %>% 
   rename(comparison = compare) %>% 
   arrange(p.value.adj) %>% 
-  write_tsv("data/process/classification_model_pairwise_stats.tsv") %>% 
-  #Also write results to supplemental table excel file
-  write_xlsx("submission/table_S11_classification_model_pairwise_stats.xlsx", format_headers = FALSE)
+  write_tsv("data/process/classification_model_pairwise_stats.tsv") 
 
 #Compare each model's cv and test AUCs  
 all %>% 
@@ -67,3 +63,9 @@ all %>%
   ) %>%
   unnest() %>%
   write_tsv("data/process/classification_cv_test_compare.tsv")
+
+
+#Write statistical analysis results to excel file with each data frame saved as a separate sheet:
+table_S7_sheets <- (list("vs_random" = all_vs_random, "pairwise" = model_stats_pairwise))
+write_xlsx(table_S7_sheets, "submission/table_S7_classification_model_stats.xlsx", format_headers = FALSE)
+

### 2020-07-10 code/diversity.R

@@ -139,8 +139,8 @@ shannon_results <- shannon_kruskal_stats_adjust %>%
   mutate(diversity_metric = "Shannon Diversity")
 richness_results <- richness_kruskal_stats_adjust %>% 
   mutate(diversity_metric = "Richness")
-diversity_kruskal_stats_adjust <- rbind(shannon_results, richness_results) %>% 
-  write_xlsx("submission/table_S3_diversity_kruskal-wallis.xlsx", format_headers = FALSE)
+#Kruskal-Wallis test results with adjusted P values----
+diversity_kruskal_stats_adjust <- rbind(shannon_results, richness_results)
 
 #List significant days after BH adjustment of p-values:
 sig_richness_days <- richness_kruskal_stats_adjust %>% 
@@ -171,8 +171,11 @@ shannon_pairwise_results <- shannon_stats_pairwise %>%
   mutate(diversity_metric = "Shannon Diversity")
 richness_pairwise_results <- richness_stats_pairwise %>% 
   mutate(diversity_metric = "Richness")
-diversity_pairwise_results <- rbind(shannon_pairwise_results, richness_pairwise_results) %>% 
-  write_xlsx("submission/table_S4_diversity_pairwise.xlsx", format_headers = FALSE)
+diversity_pairwise_results <- rbind(shannon_pairwise_results, richness_pairwise_results)
+
+#Combine Kruskal-Wallis and pairwise Wilcoxon results into one excel file----
+table_S1_sheets <- list("Kruskal_Wallis" = diversity_kruskal_stats_adjust, "pairwise_Wilcoxon" = diversity_pairwise_results)
+write_xlsx(table_S1_sheets, "submission/table_S1_diversity_stats.xlsx", format_headers = FALSE)
 
 #Format richness pairwise stats to use with ggpubr package
 plot_format_richness <- richness_stats_pairwise %>%
@@ -213,9 +216,7 @@ shannon_dx_plot <- function(timepoint){
 }
 
 #Plots of shannon for all sources of mice on day -1, the baseline microbiota community for each mouse----
-shannon_dn1 <- shannon_dx_plot(-1) +
-  ggtitle("Baseline")+ #Title plot
-  theme(plot.title = element_text(hjust = 0.5)) #Center plot title
+shannon_dn1 <- shannon_dx_plot(-1)
[…]

### 2020-07-10 code/figure_1.R

@@ -1,12 +1,13 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/Vendor_exp_scheme.png")
-b <- ggdraw() + draw_image("results/figures/cfu_over_time.png")
-c <- ggdraw() + draw_image("results/figures/weight_over_time.png")
+a <- ggdraw() + draw_image("results/figures/richness_dn1.png")
+b <- ggdraw() + draw_image("results/figures/shannon_dn1.png")
+c <- ggdraw() + draw_image("results/figures/pcoa_day-1.png")
+left_panel <- plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol = 1)
+d <- ggdraw() + draw_image("results/figures/Dn1top20_otus.png")
 
-
-plot_grid(a, b, c, labels = c("A", "B", "C", ""), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_1.pdf", width=5, height=7.5)+
-  ggsave("submission/figure_1.pdf", width=5, height=7.5)
+plot_grid(left_panel, d, labels = c("", "D"), label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_1.pdf", width=8, height=5.5)+
+  ggsave("submission/figure_1.pdf", width=8, height=5.5)
 
 

### 2020-07-10 code/figure_2.R

@@ -1,14 +1,13 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/richness_dn1.png")
-c <- ggdraw() + draw_image("results/figures/richness_d0.png")
-e <- ggdraw() + draw_image("results/figures/richness_d1.png")
-b <- ggdraw() + draw_image("results/figures/shannon_dn1.png")
-d <- ggdraw() + draw_image("results/figures/shannon_d0.png")
-f <- ggdraw() + draw_image("results/figures/shannon_d1.png")
+a <- ggdraw() + draw_image("results/figures/Vendor_exp_scheme.png")
+b <- ggdraw() + draw_image("results/figures/cfu_over_time.png")
+c <- ggdraw() + draw_image("results/figures/weight_over_time.png")
 
 
-plot_grid(a, c, e, b, d, f, labels = c("A", "C", "E", "B", "D", "F"), label_size = 12, ncol=3)+
-  ggsave("results/figures/figure_2.pdf", width=6, height=4)+
-  ggsave("submission/figure_2.pdf", width=6, height=4)
+plot_grid(a, b, c, labels = c("A", "B", "C", ""), label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_2.pdf", width=5, height=7.5)+
+  ggsave("submission/figure_2.pdf", width=5, height=7.5)
+
+
 

### 2020-07-10 code/figure_3.R

@@ -1,11 +1,15 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/pcoa_day-1.png")
-b <- ggdraw() + draw_image("results/figures/pcoa_day0.png")
-c <- ggdraw() + draw_image("results/figures/pcoa_day1.png")
+a <- ggdraw() + draw_image("results/figures/richness_d0.png")
+b <- ggdraw() + draw_image("results/figures/shannon_d0.png")
+c <- ggdraw() + draw_image("results/figures/pcoa_day0.png")
+top_row <- plot_grid(a, b, c, labels = "AUTO", label_size = 12, nrow=1)
+d <- ggdraw() + draw_image("results/figures/clind_impacted_otus_plot.png")
+e <- ggdraw() + draw_image("results/figures/D0top18_otus.png")
+bottom_row <- plot_grid(d, e, labels = c("D", "E"), label_size = 12, nrow=1)
 
-plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_3.pdf", width=8, height=7)+
-  ggsave("submission/figure_3.pdf", width=8, height=7)
+plot_grid(top_row, bottom_row, labels = c("", ""), label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_3.pdf", width=7.5, height=5.75)+
+  ggsave("submission/figure_3.pdf", width=7.5, height=5.75)
 
 

### 2020-07-10 code/figure_4.R

@@ -1,21 +1,11 @@
 source("code/functions.R")
 
-a1 <- ggdraw() + draw_image("results/figures/within_exp_dn1.png")
-a2 <- ggdraw() + draw_image("results/figures/between_exp_dn1.png")
-a3 <- ggdraw() + draw_image("results/figures/within_vendors_dn1.png")
-a4 <- ggdraw() + draw_image("results/figures/between_vendors_dn1.png")
-a <- plot_grid(a1, a2, a3, a4, labels = NULL, nrow = 1)
-b1 <- ggdraw() + draw_image("results/figures/within_exp_d0.png")
-b2 <- ggdraw() + draw_image("results/figures/between_exp_d0.png")
-b3 <- ggdraw() + draw_image("results/figures/within_vendors_d0.png")
-b4 <- ggdraw() + draw_image("results/figures/between_vendors_d0.png")
-b <- plot_grid(b1, b2, b3, b4, labels = NULL, nrow = 1)
-c1 <- ggdraw() + draw_image("results/figures/within_exp_d1.png")
-c2 <- ggdraw() + draw_image("results/figures/between_exp_d1.png")
-c3 <- ggdraw() + draw_image("results/figures/within_vendors_d1.png")
-c4 <- ggdraw() + draw_image("results/figures/between_vendors_d1.png")
-c <- plot_grid(c1, c2, c3, c4, labels = NULL, nrow = 1)
+a <- ggdraw() + draw_image("results/figures/richness_d1.png")
+b <- ggdraw() + draw_image("results/figures/shannon_d1.png")
+c <- ggdraw() + draw_image("results/figures/pcoa_day1.png")
+left_panel <- plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol = 1)
+d <- ggdraw() + draw_image("results/figures/D1top20_otus.png")
 
-plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_4.pdf", width=6, height=5)+
-  ggsave("submission/figure_4.pdf", width=6, height=5)
+plot_grid(left_panel, d, labels = c("", "D"), label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_4.pdf", width=8, height=5.5)+
+  ggsave("submission/figure_4.pdf", width=8, height=5.5)

### 2020-07-10 code/figure_5.R

@@ -1,10 +1,13 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/Dn1toD1_otus_dn1.png")
-b <- ggdraw() + draw_image("results/figures/Dn1toD1_otus_d0.png")
-c <- ggdraw() + draw_image("results/figures/Dn1toD1_otus_d1.png")
+a1 <- ggdraw() + draw_image("results/figures/d0_model_otu_Enterobacteriaceae (OTU 1).png")
+a2 <- ggdraw() + draw_image("results/figures/d0_model_otu_Bacteroides (OTU 2).png")
+a3 <- ggdraw() + draw_image("results/figures/d0_model_otu_Proteus (OTU 16).png")
+a <- plot_grid(a1, a2, a3, labels = NULL, ncol = 1)
+top_half <- plot_grid(a, labels = c("A"), label_size = 12, ncol=1)
+b <- ggdraw() + draw_image("results/figures/venn_overall_otus.png")
 
-plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=3)+
-  ggsave("results/figures/figure_5.pdf", width=8, height=3)+
-  ggsave("submission/figure_5.pdf", width=8, height=3)
+plot_grid(top_half, b, labels = c("", "B"), label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_5.pdf", width=6, height=9)+
+  ggsave("submission/figure_5.pdf", width=6, height=9)
 

### 2020-07-10 code/figure_6.R

@@ -1,7 +1,11 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/clind_impacted_otus_plot.png")
+a <- ggdraw() + draw_image("results/figures/Bacteroides (OTU 2)_time.png")
+b <- ggdraw() + draw_image("results/figures/Enterobacteriaceae (OTU 1)_time.png")
+c <- ggdraw() + draw_image("results/figures/Enterococcus (OTU 23)_time.png")
+d <- ggdraw() + draw_image("results/figures/Porphyromonadaceae (OTU 7)_time.png")
 
-plot_grid(a, labels = "", label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_6.pdf", width=8, height=5.5)+
-  ggsave("submission/figure_6.pdf", width=8, height=5.5)
+
+plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_6.pdf", width=7.5, height=4)+
+  ggsave("submission/figure_6.pdf", width=7.5, height=4)

### 2020-07-10 code/figure_7.R

@@ -1,13 +1,0 @@
-source("code/functions.R")
-
-a1 <- ggdraw() + draw_image("results/figures/d0_model_otu_Enterobacteriaceae (OTU 1).png")
-a2 <- ggdraw() + draw_image("results/figures/d0_model_otu_Bacteroides (OTU 2).png")
-a3 <- ggdraw() + draw_image("results/figures/d0_model_otu_Proteus (OTU 16).png")
-a <- plot_grid(a1, a2, a3, labels = NULL, ncol = 1)
-top_half <- plot_grid(a, labels = c("A"), label_size = 12, ncol=1)
-b <- ggdraw() + draw_image("results/figures/venn_overall_otus.png")
-
-plot_grid(top_half, b, labels = c("", "B"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_7.pdf", width=6, height=9)+
-  ggsave("submission/figure_7.pdf", width=6, height=9)
-

### 2020-07-10 code/figure_8.R

@@ -1,11 +1,0 @@
-source("code/functions.R")
-
-a <- ggdraw() + draw_image("results/figures/Bacteroides (OTU 2)_time.png")
-b <- ggdraw() + draw_image("results/figures/Enterobacteriaceae (OTU 1)_time.png")
-c <- ggdraw() + draw_image("results/figures/Enterococcus (OTU 23)_time.png")
-d <- ggdraw() + draw_image("results/figures/Porphyromonadaceae (OTU 7)_time.png")
-
-
-plot_grid(a, b, c, d, labels = "AUTO", label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_8.pdf", width=7.5, height=4)+
-  ggsave("submission/figure_8.pdf", width=7.5, height=4)

### 2020-07-10 code/figure_S1.R

@@ -1,15 +1,17 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/exp1_cfu_time.png")
-b <- ggdraw() + draw_image("results/figures/exp2_cfu_time.png")
-c <- ggdraw() + draw_image("results/figures/C.diff_CFU_D7_stats.png")
-d <- ggdraw() + draw_image("results/figures/weight_D2.png")
-e <- ggdraw() + draw_image("results/figures/C.diff_percent_colonized.png")
-right_panel <- plot_grid(c, d, e, labels = c("C", "D", "E"), label_size = 12, nrow = 1)
+a <- ggdraw() + draw_image("results/figures/pcoa_dn1_schloss.png")
+b <- ggdraw() + draw_image("results/figures/pcoa_dn1_young.png")
+c <- ggdraw() + draw_image("results/figures/pcoa_dn1_jackson.png")
+d <- ggdraw() + draw_image("results/figures/pcoa_dn1_charles_river.png")
+e <- ggdraw() + draw_image("results/figures/pcoa_dn1_taconic.png")
+f <- ggdraw() + draw_image("results/figures/pcoa_dn1_envigo.png")
+g <- ggdraw() + draw_image("results/figures/within_exp_dn1.png")
+h <- ggdraw() + draw_image("results/figures/between_exp_dn1.png")
 
+plot_grid(a, b, c, d, e, f, g, h, labels = "AUTO", label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_S1.pdf", width=6, height=10)+
+  ggsave("submission/figure_S1.pdf", width=6, height=10)
 
-plot_grid(a, right_panel, b,  labels = c("A", "", "B"), label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_S1.pdf", width=8, height=4.5)+
-  ggsave("submission/figure_S1.pdf", width=8, height=4.5)
 
 

### 2020-07-10 code/figure_S2.R

@@ -1,14 +1,15 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/pcoa_dn1_schloss.png")
-b <- ggdraw() + draw_image("results/figures/pcoa_dn1_young.png")
-c <- ggdraw() + draw_image("results/figures/pcoa_dn1_jackson.png")
-d <- ggdraw() + draw_image("results/figures/pcoa_dn1_charles_river.png")
-e <- ggdraw() + draw_image("results/figures/pcoa_dn1_taconic.png")
-f <- ggdraw() + draw_image("results/figures/pcoa_dn1_envigo.png")
+a <- ggdraw() + draw_image("results/figures/exp1_cfu_time.png")
+b <- ggdraw() + draw_image("results/figures/exp2_cfu_time.png")
+c <- ggdraw() + draw_image("results/figures/C.diff_CFU_D7_stats.png")
+d <- ggdraw() + draw_image("results/figures/weight_D2.png")
+e <- ggdraw() + draw_image("results/figures/C.diff_percent_colonized.png")
+right_panel <- plot_grid(c, d, e, labels = c("C", "D", "E"), label_size = 12, nrow = 1)
 
-plot_grid(a, b, c, d, e, f, labels = "AUTO", label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_S2.pdf", width=6, height=8)+
-  ggsave("submission/figure_S2.pdf", width=6, height=8)
+
+plot_grid(a, right_panel, b,  labels = c("A", "", "B"), label_size = 12, ncol=2)+
+  ggsave("results/figures/figure_S2.pdf", width=8, height=4.5)+
+  ggsave("submission/figure_S2.pdf", width=8, height=4.5)
 
 

### 2020-07-10 code/pcoa_stats.R

@@ -85,12 +85,10 @@ day_adonis <- adonis(all_dist~ day, strata = variables$mouse_id, data = variable
 
 all_adonis <- adonis(all_dist~(source/(unique_cage*experiment*run))*day, strata = variables$mouse_id, data = variables, permutations = 9999)
 
-tibble(effects = c("source", "day", "source:unique_cage", "source:run", "source:day", "source:unique_cage:run", "source:unique_cage:day"),
+adonis_all_samples <- tibble(effects = c("source", "day", "source:unique_cage", "source:run", "source:day", "source:unique_cage:run", "source:unique_cage:day"),
         r_sq = all_adonis$aov.tab$R2[1:7],
         p = all_adonis$aov.tab$Pr[1:7]) %>% 
-  write_tsv("data/process/adonis_all.tsv") %>% 
-  #Also write results to supplemental table excel file
-  write_xlsx("submission/table_S5_PERMANOVA_all.xlsx", format_headers = FALSE)
+  write_tsv("data/process/adonis_all.tsv") 
 
 #Function to plot pcoa data for all vendors----
 plot_pcoa <- function(df){
@@ -195,7 +193,6 @@ dn1_axis1 <- dn1_axis_labels %>% filter(axis == 1) %>% pull(loading) %>% round(d
 dn1_axis2 <- dn1_axis_labels %>% filter(axis == 2) %>% pull(loading) %>% round(digits = 1) #Pull value & round to 1 decimal
 
 dn1 <- plot_pcoa(dn1_pcoa, -1)+
-  ggtitle("Baseline")+ #Title plot
   labs(x = paste("PCoA 1 (", dn1_axis1, "%)", sep = ""), #Anotations for each axis from loadings file
        y = paste("PCoA 2 (", dn1_axis2,"%)", sep = ""))+
   theme(plot.title = element_text(hjust = 0.5)) #Center plot title
@@ -229,7 +226,6 @@ d0_axis1 <- d0_axis_labels %>% filter(axis == 1) %>% pull(loading) %>% round(dig
 d0_axis2 <- d0_axis_labels %>% filter(axis == 2) %>% pull(loading) %>% round(digits = 1) #Pull value & round to 1 decimal
 
 d0 <- plot_pcoa(d0_pcoa, 0)+
-  ggtitle("Clindamycin")+ #Title plot
   labs(x = paste("PCoA 1 (", d0_axis1, "%)", sep = ""), #Anotations for each axis from loadings file
[…]

### 2020-07-10 code/relative_distance_comparisons.R

@@ -49,7 +49,7 @@ within_exp <- dn1_dist %>%
   group_by(vendor) %>% 
 #  summarize(median_dist = median(distances), n = n()) %>% 
   ggplot(aes(x = vendor, y = distances, color = vendor)) +
-  geom_boxplot(outlier.shape = NA, size = 1.2)+
+  geom_boxplot(outlier.shape = NA, size = 1.2, show.legend = FALSE)+
   geom_jitter(aes(shape = experiment), size=2, alpha=0.6, show.legend = FALSE) +
   scale_x_discrete(guide = guide_axis(n.dodge = 2))+
   ylim(0, 1.25)+
@@ -57,8 +57,8 @@ within_exp <- dn1_dist %>%
                       values=color_scheme,
                       breaks=color_vendors,
                       labels=color_vendors)+
-  labs(title = "Baseline: within experiment", x = NULL, y = "Theta YC Distance")+
-  theme(plot.title = element_text(hjust = 0.5))+ #Center plot title
+  labs(title = "Within experiment", x = NULL, y = "Theta YC Distance")+
+  theme(plot.title = element_text(hjust = 0.5)) +#Center plot title
   theme_classic()
 save_plot("results/figures/within_exp_dn1.png", within_exp)
 
@@ -77,7 +77,7 @@ between_exp <- dn1_dist %>%
                       values=color_scheme,
                       breaks=color_vendors,
                       labels=color_vendors)+
-  labs(title = "Baseline: between experiments", x = NULL, y = "Theta YC Distance")+
+  labs(title = "Between experiments", x = NULL, y = "Theta YC Distance")+
   theme(plot.title = element_text(hjust = 0.5))+ #Center plot title
   theme_classic()
 save_plot("results/figures/between_exp_dn1.png", between_exp)

### 2020-07-10 code/taxa_stats.R

@@ -155,6 +155,21 @@ Dn1toD1_otus_d1 <- plot_otus_dx(shared_sig_otus_Dn1toD1, 1) +
   theme(plot.title = element_text(hjust = 0.5)) #Center plot title
 save_plot("results/figures/Dn1toD1_otus_d1.png", Dn1toD1_otus_d1, base_height = 7, base_width = 8)
 
+#Plots of the top 18-20 OTUs that varied across sources at each timepoint (only 18 OTUs varied significantly across sources post-clindamycin treatment)
+#Baseline: top 20 OTUs that vary across sources
+Dn1top20_otus <- plot_otus_dx(`sig_otu_day-1`[1:20], -1) #Pick top 20 significant OTUs
+save_plot("results/figures/Dn1top20_otus.png", Dn1top20_otus, base_height = 9, base_width = 7)
+#Post-clindamycin: top 18 OTUs that vary across sources
+D0top18_otus <- plot_otus_dx(`sig_otu_day0`[1:18], 0) #Pick top 20 significant OTUs
+save_plot("results/figures/D0top18_otus.png", D0top18_otus, base_height = 9, base_width = 7)
+#Post-infection: top 20 OTUs that vary across sources
+D1top20_otus <- plot_otus_dx(`sig_otu_day1`[1:20], 1) #Pick top 20 significant OTUs
+save_plot("results/figures/D1top20_otus.png", D1top20_otus, base_height = 9, base_width = 7)
+
+#Overlap between the top 18-20 OTUs that varied across sources at each timepoints
+shared_top_otus_Dn1toD1 <- intersect_all(`sig_otu_day-1`[1:20], sig_otu_day0[1:18], sig_otu_day1[1:20])
+#5 OTUs: 1] "Parasutterella (OTU 26)", "Burkholderiales (OTU 34)", "Betaproteobacteria (OTU 58)","Lactobacillus (OTU 49)","Parabacteroides (OTU 5)"  
+
 # Perform pairwise Wilcoxan rank sum tests for otus that were significantly different across sources of mice on a series of days----
 pairwise_day_otu <- function(timepoint, sig_otu_dayX){
   otu_stats <- agg_otu_data %>% 
@@ -223,11 +238,12 @@ kw_w_sig_d1 <- read_tsv(file="data/process/otu_stats_day_1.tsv") %>%
   mutate(day = 1) %>% #Add experiment day for statistics
   arrange(p.value.adj) #Arrange by adjusted p.values
 
[…]

### 2020-07-10 code/weight_plot.R

@@ -49,8 +49,8 @@ weight_kruskal_stats <- weight_data %>%
   group_by(day) %>% 
   nest() %>% 
   mutate(model=map(data, ~kruskal.test(x=.x$weight_change, g=as.factor(.x$vendor)) %>% tidy())) %>% 
-  mutate(mean = map(data, get_weight_mean_vendor)) %>% 
-  unnest(c(model, mean)) %>% 
+  mutate(median = map(data, get_weight_median_vendor)) %>% 
+  unnest(c(model, median)) %>% 
   ungroup() #Ungroup before adjusting p values
 
 #Adjust p-values for testing multiple days and write results to table
@@ -60,7 +60,7 @@ weight_kruskal_stats_adjust <- weight_kruskal_stats %>%
   arrange(p.value.adj) %>% 
   write_tsv("data/process/weight_stats_all_days.tsv")
 #Also write to supplemental table excel file
-weight_kruskal_stats_adjust %>% write_xlsx("submission/table_S2_weight_kruskal-wallis.xlsx", format_headers = FALSE)
+weight_kruskal_stats_adjust %>% write_xlsx("submission/table_S5_weight_kruskal-wallis.xlsx", format_headers = FALSE)
 
 #Timepoints where weight_change is significantly different across the sources of mice
 sig_weight_days <- weight_kruskal_stats_adjust %>% 

### 2020-07-10 code/taxa_stats.R
Examine overlap between OTUs that vary across source on day 0 and OTUs impacted by clindamycin

@@ -335,6 +335,12 @@ sig_otu_pairs <- pull_significant_taxa(o_dn1to0_pairs_stats_adjust, otu)
 # 153 OTUs
 sig_otu_pairs_top10 <- sig_otu_pairs[1:10]
 
+#OTUs that overlap between Fig. 3D and E----
+Fig3D_vs_E_overlap <- intersect_all(sig_otu_day0[1:18], sig_otu_pairs[1:10])
+#"Enterobacteriaceae (OTU 1)"
+Clind_vs_source_D0 <- intersect_all(sig_otu_day0, sig_otu_pairs)
+#4 OTUs: "Enterobacteriaceae (OTU 1)","Enterococcus (OTU 23)","Lactobacillus (OTU 6)","Lachnospiraceae (OTU 130)"
+
 #Plot of the otus with significantly different relative abundances post clindamycin treatment across all sources of mice:----
 #Facet plot by day with the following label names:
 facet_labels <- c("Baseline", "Clindamycin")

### 2020-07-10 submission/manuscript.Rmd
Reorganization edits

@@
 output:
+  word_document: default
   pdf_document:
@@
     keep_tex: yes
-  word_document: default
   html_document:
@@
 
+## Results
+```{r alpha}
+richness_stats_all <- read_tsv("../data/process/richness_stats_all_days.tsv")
+padj_dn1_richness <- richness_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
+
+shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
+padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
+```
+
+```{r adonis}
+adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
+r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
+rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
+p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max
+
+adonis_stats_dn101 <- read_tsv("../data/process/adonis_dn1-1.tsv")
+dn1_rsq_sum <- adonis_stats_dn101 %>% filter(day == -1) %>% select(r_sq) %>% sum
+d0_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 0) %>% select(r_sq) %>% sum
+d1_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 1) %>% select(r_sq) %>% sum
+p_adonis_intial <- adonis_stats_dn101 %>% select(p) %>% max 
+p_adonis_min <- adonis_stats_dn101 %>% select(p) %>% min #All p values are same
+
+source_adonis_dn1 <- read_tsv("../data/process/adonis_dn1_source.tsv")
+adonis_schloss_rsq_sum <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(r_sq) %>% sum
+p_adonis_schloss_max <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(p) %>% max
+adonis_young_rsq_sum <- source_adonis_dn1 %>% filter(source == "Young") %>% select(r_sq) %>% sum
+p_adonis_young_max <- source_adonis_dn1 %>% filter(source == "Young") %>% select(p) %>% max
+```
+
+```{r source_taxa}
+#Number of different otus across sources at day -1, 0, and 1 timepoints
+source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
+source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
+source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
+source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()
+
+#Function to find which significant otus are shared 
+intersect_all <- function(a,b,...){
+  Reduce(intersect, list(a,b,...))
+}
+
+#OTUs shared across baseline, post-clindamycin, and post-infection timepoints
+source_o_dn1 <- source_otus %>% filter(day == -1) %>% pull(otu)
+source_o_d0 <- source_otus %>% filter(day == 0) %>% pull(otu)
+source_o_d1 <- source_otus %>% filter(day == 1) %>% pull(otu)
+shared_source_otus_dn1to1 <- length(intersect_all(source_o_dn1, source_o_d0, source_o_d1))
+
+```
+
+**Microbiota variation is high between mice from different sources.** To test how the microbiotas of mice from different sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other 10 years ago (the Young and Schloss lab colonies) and four commercial source: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan).
+These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019].
+We used 16S rRNA gene sequencing to characterize the fecal bacterial communities across sources at baseline.
+
+Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity across the 6 sources of mice.
+Intially, there was a significant difference in the number of observed operational taxonomic units (OTUs), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_richness` and *P*~FDR~ = `r padj_dn1_shannon`, respectively) across sources of mice (Fig.
+1A-B and Table S1).
+Next, we compared the community structure of mice from the 6 sources over the course of the experiment using principal coordinate analysis (PCoA) of the $\theta_{YC}$ distances.
+Permutational multivariate analysis of variance (PERMANOVA) analysis revealed source and cage interactions explained most of the observed variation across fecal communities explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, *P* = `r p_adonis_intial`, Fig.
+3 and Table S2).
+Mice that are co-housed have been shown to have similar gut microbiotas due to coprophagy [@Nguyen2015] and since mice within the same source were housed together, it is not surprising that cage also contributed to the observed microbiota variation.
+Since we conducted two experiments approximately 3 months apart, we also compared baseline communities across the 2 experiments.
+Experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig.
+S2A-B and Table S2).
+However, most of the vendors also clustered by experiment (Fig.
+S2C-D, F), suggesting there was some community variation between the 2 experiments within each source, particularly for Schloss, Young, and Envigo mice (Fig.
+S2G-H).
+After finding differences at the community level, we next identified the bacteria that initially varied across sources of mice.
+There were `r source_otus_dn1` OTUs with relative abundances that varied across sources (Fig.
+1D and Table S3).
+By using mice from six sources we were able to increase the number of microbiota communities to test with the clindamycin *C.
+difficile* colonization mouse model.
+
 ```{r cfu}
@@
 ```
-## Results
-
 **Clindamycin treatment renders all mice susceptible to *C.
-difficile* 630 colonization regardless of source.** To test how the microbiotas of mice from different sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other 10 years ago (the Young and Schloss lab colonies) and four commercial source: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan).
-These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019].
-
-After a 13-day acclimation period for the mice ordered from vendors, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C.
+difficile* 630 colonization, but clearance time varies across sources.** After the acclimation period, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C.
 difficile* 630 spores (Fig.
-1A).
-Clindamycin was chosen because we have previously demonstrated mice are rendered susceptible, but consistently cleared the CDI within 9 days [@Jenior2018; @Tomkovich2019], clindamycin is frequently implicated with human CDIs [@Guh2018], and is also part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008].
+2A).
+Clindamycin was chosen because we have previously demonstrated mice are rendered susceptible, but consistently cleared *C.
+difficile* within 9 days [@Jenior2018; @Tomkovich2019], clindamycin is frequently implicated with human CDIs [@Guh2018], and is also part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008].
 The day after infection, *C.
 difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of source (Fig.
-1B).
+2B).
 Interestingly, variation in *C.
 difficile* CFU levels across sources of mice emerged from days 3-7 post-infection (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig.
-1B and Table S1), suggesting mouse source is associated with *C.
+2B and Table S4), suggesting mouse source is associated with *C.
 difficile* clearance.
-We conducted two experiments approximately 3 months apart and while the colonization dynamics were similar across most sources of mice, there was some variation between the 2 experiments, particularly for the Schloss and Envigo mice (Fig.
-S1A-B).
+We conducted two experiments and while the colonization dynamics were similar across most sources of mice, there was some variation between the 2 experiments, particularly for the Schloss and Envigo mice (Fig.
+S2A-B).
 Although *C.
@@
 difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig.
-1C and Table S2).
+2C and Table S5).
 Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment.
-Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to lose more weight (although there was variation between experiments with Schloss and Envigo mice), have higher *C.
-difficile* CFU levels and take longer to clear the infection compared to the other sources of mice, which was particularly evident 7 days post-infection (Fig.
-1B-C, Fig.
+Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to lose more weight, have higher *C.
+difficile* CFU levels and take longer to clear the infection compared to the other sources of mice (although there was variation between experiments with Schloss and Envigo mice), which was particularly evident 7 days post-infection (Fig.
+2B-C, Fig.
 S2C-D), when `r format(percent_colonized_d7, digits = 1)`% of the mice were still colonized with *C.
 difficile* (Fig.
-S1E).
+S2E).
 By 9 days post-infection the majority of the mice from all sources had cleared *C.
 difficile* (Fig.
-S1C) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment.
+2C) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment.
 Thus, clindamycin rendered all mice susceptible to *C.
 difficile* 630 colonization, regardless of source, but there was significant variation in disease phenotype across the sources of mice.
-
-```{r alpha}
-richness_stats_all <- read_tsv("../data/process/richness_stats_all_days.tsv")
-padj_dn1_richness <- richness_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
-
-shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
-padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
-```
-**Bacterial communities consistently vary across sources despite antibiotic and infection perturbations.** Given the well known variation in mouse microbiomes across breeding colonies [@Alegre2019], we hypothesized that the variation in *C.
-difficile* clearance could be explained by microbiota variation across the 6 sources of mice.
-We used 16S rRNA gene sequencing to characterize the fecal bacterial communities from the mice over the course of the experiment.
-Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity measures across the 6 sources of mice.
-Examining the bacterial communities at baseline, prior to clindamycin treatment, there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig.
-2A-B and Table S3).
-As expected, the clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and richness and Shannon diversity started to increase 1 day post-infection (Fig.
-2C-D).
-Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after both clindamycin and *C.
-difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other sources (Fig 2C-F and Table S4).
-Although the Charles River mice had more diverse microbiotas, the Young and Schloss lab mice were also able to clear *C.
-difficile* faster than the other sources, suggesting microbiota diversity alone does not explain the observed variation in *C.
-difficile* colonization across vendors.
-```{r adonis}
-adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
-r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
-rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
-p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max
-
-adonis_stats_dn101 <- read_tsv("../data/process/adonis_dn1-1.tsv")
-dn1_rsq_sum <- adonis_stats_dn101 %>% filter(day == -1) %>% select(r_sq) %>% sum
-d0_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 0) %>% select(r_sq) %>% sum
-d1_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 1) %>% select(r_sq) %>% sum
-p_adonis_intial <- adonis_stats_dn101 %>% select(p) %>% max 
-p_adonis_min <- adonis_stats_dn101 %>% select(p) %>% min #All p values are same
-
-source_adonis_dn1 <- read_tsv("../data/process/adonis_dn1_source.tsv")
-adonis_schloss_rsq_sum <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(r_sq) %>% sum
-p_adonis_schloss_max <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(p) %>% max
-adonis_young_rsq_sum <- source_adonis_dn1 %>% filter(source == "Young") %>% select(r_sq) %>% sum
-p_adonis_young_max <- source_adonis_dn1 %>% filter(source == "Young") %>% select(p) %>% max
-```
-Next, we compared the community structure of mice from the 6 sources over the course of the experiment using principal coordinate analysis (PCoA) of the $\theta_{YC}$ distances.
-Permutational multivariate analysis of variance (PERMANOVA) analysis revealed source was the dominant factor that explained the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage (mice from the same sources were housed together, primarily at a density of 2 mice per cage) and day of the experiment (Movie S1 and Table S5).
-Mice that are co-housed have been shown to have similar gut microbiotas due to coprophagy [@Nguyen2015] and since mice within the same source were housed together, it is not surprising that cage also contributes to the observed microbiota variation.
-Since the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1).
-For all 3 timepoints, the source and cage interaction significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig.
-3 and Table S6).
-We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig.
-S2A-B and Table S7).
-However, most of the vendors also clustered by experiment (Fig.
-S2C-D, F), suggesting there was some community variation between the 2 experiments within each source.
-Thus, source was the factor that explained the most variation observed in the bacterial communities.
-Importantly with the exception of the Schloss and Young colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C.
-difficile* challenge.
-
-Since there was some variation in the microbiota structure of mice from the same source between experiments at baseline, we next looked at how similar the communities were within the same source and between sources in response to clindamycin treatment and *C.
-difficile* challenge (Fig.
-4) .
-The baseline communities varied most between experiments for Schloss, Young, and Envigo mice and variation between sources of mice was high (Fig.
-4A).
-Clindamycin treatment reduced the variation between experiments within Schloss, Young and Jackson mice and some of the variation between sources diminished, particularly for the Schloss, Young, and Charles River mice (Fig.
-4B).
-Post-infection, the community variation started to increase within sources of mice and variation between sources of mice started to increase towards previous levels (Fig.
-4C).
-By using mice from multiple sources we were able to increase the number of microbiota communities we tested with the clindamycin *C.
-difficile* colonization mouse model.
-```{r source_taxa}
-#Number of different otus across sources at day -1, 0, and 1 timepoints
-source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
-source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
-source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
-source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()
-
-#Function to find which significant otus are shared 
-intersect_all <- function(a,b,...){
-  Reduce(intersect, list(a,b,...))
-}
-
-#OTUs shared across baseline, post-clindamycin, and post-infection timepoints
-source_o_dn1 <- source_otus %>% filter(day == -1) %>% pull(otu)
-source_o_d0 <- source_otus %>% filter(day == 0) %>% pull(otu)
-source_o_d1 <- source_otus %>% filter(day == 1) %>% pull(otu)
-shared_source_otus_dn1to1 <- length(intersect_all(source_o_dn1, source_o_d0, source_o_d1))
-
-```
-After finding differences at the community level, we next identified the bacteria that varied across sources of mice over the initial days of the experiment.
-We examined bacterial relative abundances at the operational taxonomic unit (OTU) level.
-Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs (Table S8) with relative abundances that varied across sources.
-Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs (Table S8).
-After *C.
-difficile* challenge, there were `r source_otus_d1` OTUs (Table S8) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment.
-In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 5A-C) with relative abundances that consistently varied across sources.
-Importantly, some of the OTus that consistently varied across sources also shifted with clindamycin treatment.
-For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice.
-*Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment.
-In summary, mouse bacterial communities varied significantly according to source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C.
-difficile* challenge.
 
@@
 ```
-**Clindamycin treatment alters a subset of taxa that were found in all sources.** Although there were bacteria that consistently varied across sources, we also wanted to identify the bacteria that shifted after clindamycin treatment, regardless of source.
-By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment (Fig.
-6 and Tables S9).
-Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) overlapped (Fig.
-5, Fig.
-6C-D).
-These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all sources and only a subset of the OTUs also varied across sources.
+**Clindamycin treatment alters bacteria in all sources, but a subset of bacterial differences across sources persists.** Given the well known variation in mouse microbiomes across breeding colonies [@Alegre2019], we hypothesized that variation in *C.
+difficile* clearance would be explained by microbiota variation across the 6 sources of mice.
+We used 16S rRNA sequencing to characterize microbiota variation across sources after clindamycin treatment and throughout the rest of the experiment.
+PERMANOVA analysis revealed source was the dominant factor that explained the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage (mice from the same sources were housed together, primarily at a density of 2 mice per cage) and day of the experiment (Movie S1 and Table S2).
+Since the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the communities after clindamycin treatment (day 0) and post-infection (day 1).
+
+As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice (Fig.
+3A-B).
+Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after clindamycin treatment, with Charles River mice having higher richness and Shannon diversity than most of the other sources (Fig 3A-B and Table S1).
+The variation in community structures was also diminished by clindamycin treatment (Fig.
+3C and Table S2).
+Next we identified the bacteria that shifted after clindamycin treatment, regardless of source by analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment.
+We identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment (Fig.
+3D and Table S6) in most mice.
+Despite the significant impact of clindamycin, there were still `r source_otus_d0` OTUs (Fig.
+3E and Table S3) with relative abundances that significantly varied across sources.
+Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that varied across sources post-clindamycin, we found 4 OTUs (*Enterobacteriaceae (OTU 1)*, *Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) overlapped (Fig.
+3D-E and Tables S3, S6).
+These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all sources and only a subset of the OTUs continued to vary across sources.
+
+**Microbiota variation across sources is maintained after *C.
+difficile* challenge.** Post-infection, significant differences in diversity metrics  were maintained across sources (*P*~FDR~ < 0.05, Fig 4A-B and Table S1).
+Although the Charles River mice had more diverse microbiotas, the Young and Schloss lab mice were also able to clear *C.
+difficile* faster than some of the other sources, suggesting microbiota diversity alone does not explain the observed variation in *C.
+difficile* colonization across sources.
+Source and cage interaction continued to explain most of the observed community variation (combined R^2^ = `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig.
+4D and Table S2).
+After *C.
+difficile* challenge, there were `r source_otus_d1` OTUs (Fig.
+4D and Table S3) with significantly different relative abundances across sources.
+In spite of the experimental perturbations that occurred, there were `r shared_source_otus_dn1to1` OTUs with relative abundances that consistently varied across sources at baseline, post-clindamycin, and post-infection.
+Importantly, some of the OTUs that consistently varied across sources also shifted with clindamycin treatment.
+For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice.
+*Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment.
+In summary, mouse bacterial communities varied significantly according to source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C.
+difficile* challenge.
 
@@
 ```
-**Source-specific and clindamycin impacted bacteria distinguish *C.
-difficile* colonization status in mice.** After identifying taxa that varied by source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C.
+**Baseline, post-clindamycin, and post-infection community data can predict mice that will clear *C.
+difficile* by 7 days post-infection.** After identifying taxa that varied by source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C.
 difficile* colonization at day 7 (Fig.
-1D, Fig.
-S1C).
+2D, Fig.
+S2C).
 We trained three L2-regularized logistic regression models with input bacterial community data from the baseline, post-clindamycin, and post-infection timepoints of the experiment to predict *C.
@@
 All models were better at predicting *C.
-difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12).
-Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than all other models with an area under the receiving operator characteristic curve (AUROC) of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13).
+difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S7).
+Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than the other models with an area under the receiving operator characteristic curve (AUROC) of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S7).
 Thus, we were able to use bacterial relative abundance data alone to differentiate mice that had cleared *C.
@@
 
-Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 taxa that had the highest absolute feature weights in each of the 6 models (Tables S14-15).
+Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 OTUs that had the highest absolute feature weights in each of the 6 models (Table S8).
 First, we looked at OTUs from the model with the best performance that was based on the post-clindamycin treatment bacterial community data.
 While most of the 20 OTUs had low relative abundances on day 0, *Enterobacteriace*, *Bacteroides* and *Proteus* had high relative abundances in at least one source of mice and significantly varied across sources (Fig.
-7A).
+5A).
 Next, the top 20 taxa from each model were compared to the list of taxa that varied across source at the same timepoint (Fig.
-5 and Tables S8-9) and the taxa that were altered by clindamycin treatment (Fig.
-6 and Table S10-11).
+1D, 3E, 4D and Table S3) and the taxa that were altered by clindamycin treatment (Fig.
+3D and Table S6).
 We found a subset of OTUs that were important to the model and overlapped with bacteria that varied by either source, clindamycin treatment, or both (Fig.
-S4, S5A-C).
-Combining the overall results for the 3 OTU models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig.
-7B).
-Several OTUs (*`r list_key_otus`*) appeared across at least 2 models, so we examined how the relative abundances of these key taxa varied over the course of the experiment (Fig.
-8).
-Throughout the experiment, there was at least 1 timepoint where relative abundances of these OTUs significantly varied across sources (Table S13).
+S4).
+Combining the overall results for the 3 models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig.
+5B).
+Several OTUs (*`r list_key_otus`*) appeared across at least 2 models, so we examined how the relative abundances of these OTUs varied over the course of the experiment (Fig.
+6).
+Throughout the experiment, there was at least 1 timepoint where relative abundances of these OTUs significantly varied across sources (Table S9).
 Interestingly, there were no OTUs that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection with *C.
@@
 We identified *`r list_key_otus`* (Fig.
-8) as candidate bacteria within these communities that were influencing variation in *C.
+6) as candidate bacteria within these communities that were influencing variation in *C.
 difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by source, were impacted by clindamycin treatment, or both.
@@
 difficile* 630 by 9 days post-infection and the model built with the post-clindamycin OTU relative abundance data had the best performance, our results suggest clindamycin treatment had a large role in determining *C.
-difficile* susceptibility and clearance in the mice.
+difficile* susceptibility and clearance than the source of the mice.
 
@@
 There are several limitations to our work.
-The microbiome is composed of viruses, fungi, and parasites in addition to bacteria, and these non-bacterial members can also vary across mouse vendors [@Rasmussen2019; @Mims2020].
+The microbiome is composed of viruses, fungi, and parasites in addition to bacteria, and these non-bacterial members can also vary across mouse sources [@Rasmussen2019; @Mims2020].
 While our study focused solely on the bacterial portion, viruses and fungi have also begun to be implicated in the context of CDIs or FMT treatments for recurrent CDIs [@Markey2018; @Stewart2019; @Ott2017; @Zuo2017; @Zuo2018].
@@
 The Kruskal-Wallis test was used to analyze differences in *C.
-difficile* CFU, mouse weight change, and alpha diversity across vendors with a Benjamini-Hochberg correction for testing multiple timepoints, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
+difficile* CFU, mouse weight change, and alpha diversity across sources with a Benjamini-Hochberg correction for testing multiple timepoints, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
 For taxonomic analysis and generation of logistic regression model input data, *C.
@@
 **Figure 1.
+Microbiota variation is high between mice from different sources.** A-B.
+Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice at baseline (day -1) of the experiment.
+Differences across sources were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 for panel A (Table S1).
+None of the *P* values from pairwise Wilcoxon comparisons between sources were significant after Benjamini-Hochberg correction (Table S1).
+Gray lines represent the median values for each source of mice.
+C.
+Principal Coordinates Analysis (PCoA) of $\theta_{YC}$ distances of baseline stool samples.
+Permutational multivariate analysis of variance (PERMANOVA) analysis demonstrated that source and the interaction between source and cage explained most of the variation (combined R^2^ = `r dn1_rsq_sum`, *P* = `r p_adonis_intial`,  see Table S2).
+For A-C:  each symbol representing the value for a stool sample from an individual mouse, circles represent experiment 1 mice and triangles represent experiment 2 mice.
+D.
+Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the top 20 bacteria out of the `r source_otus_dn1` OTUs that varied across sources at baseline (Table S3).
+
+\newpage
+\includegraphics{figure_2.pdf}
+
+**Figure 2.
 Clindamycin is sufficient to promote *C.
@@
 Mouse weight change measured in grams over time (N = `r min_weight_n`-`r max_weight_n` mice per timepoint), all mice were normalized to the weight recorded 1 day before infection.
-For B-C: timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S1 and Table S2) are reflected by the asterisk(s) above each timepoint (\*\, *P* < 0.05).
+For B-C: timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S4 and Table S5) are reflected by the asterisk(s) above each timepoint (\*\, *P* < 0.05).
 Lines represent the median for each source and circles represent individual mice from experiment 1 while triangles represent mice from experiment 2.
-
-\newpage
-\includegraphics{figure_2.pdf}
-**Figure 2.
-Differences in microbial richness and diversity across mouse sources emerge after clindamycin treatment and infection.** A-F.
-Number of observed OTUs and Shannon diversity index values at baseline: day -1 (A-B), after clindamycin: day 0 (C-D) and post-infection: day 1 (E-F) timepoints of the experiment.
-Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 for all panels except for B (Table S3).
-Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S4).
-For A-F: circles represent experiment 1 mice, while triangles represent experiment 2 mice with each symbol representing the value for a stool sample from an individual mouse.
-Gray lines represent the median values for each source of mice.
 
@@
 **Figure 3.
-Mouse source is the variable that explains most of the variation observed in the baseline, post-clindamycin, and post-infection bacterial communities.** A-C.
-Principal Coordinates Analysis of $\theta_{YC}$ distances from stools collected at baseline (A), post-clindamycin (B), and post-infection (C) timepoints of the experiment.
-Each symbol represents a stool sample from an individual mouse, with circles representing experiment 1 mice and triangles representing experiment 2 mice.
-PERMANOVA analysis demonstrated that source and the interaction between source and cage explained most of the variation observed in the baseline (combined R^2^ = `r dn1_rsq_sum`), post-clindamycin (combined R^2^ = `r d0_rsq_sum`), and post-infection (combined R^2^ = `r d1_rsq_sum`) communities (all *P* = `r p_adonis_intial`, see Table S6).
+Clindamycin treatment alters bacteria in all sources, but a subset of bacterial differences across sources persists.** A-B.
+Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice after clindamycin treatment (day 0).
+Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05.
+Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S1).
+C.
+PCoA of $\theta_{YC}$ distances from stools collected post-clindamycin.
+PERMANOVA analysis demonstrated that source and the interaction between source and cage explained most of the variation observed post-clindamycin (combined R^2^ = `r d0_rsq_sum`, *P* = `r p_adonis_intial`, see Table S2).For A-C, each symbol represents a stool sample from an individual mouse, with circles representing experiment 1 mice and triangles representing experiment 2 mice.
+D.
+Plots highlighting the median (point) and interquantile range (colored lines) of the top 10 OTUs out of `r clindamycin_otus_sig` with relative abundances that changed after clindamycin treatment (adjusted *P* value < 0.05).
+Data were analyzed by Wilcoxon signed rank test limited to mice that had paired sequence data for day -1 and 0 (N = `r n_paired_for_clind`), with Benjamini-Hochberg correction for testing all identified OTUs (Table S6).
+The gray vertical line indicates the limit of detection.
+E.
+Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the `r source_otus_d0` OTUs that varied across sources after clindamycin treatment (Table S3).
 
@@
 **Figure 4.
-High inter-group variation across mouse sources is diminished by clindamycin treatment** A-C.
-Boxplots of the $\theta_{YC}$ distances of the 6 sources of mice relative to mice within the same source and experiment, mice within the same source and between experiments, mice within the same source, and mice from other groups at the baseline (A), after clindamycin treatment (B), and post-infection (C) timepoints.
-For comparisons within mice from the same source, symbols represent individual mouse samples: circles for experiment 1 and triangles for experiment 2.
+Microbiota variation across sources is maintained after *C.
+difficile* challenge.** A-B.
+Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice post-infection (day 1).
+Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05.
+Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S1).
+PCoA of $\theta_{YC}$ distances of post-infection stool samples.
+PERMANOVA analysis demonstrated that source and the interaction between source and cage explained most of the variation (combined R^2^ = `r d1_rsq_sum`, *P* = `r p_adonis_intial`, Table S2).
+For A-C:  each symbol representing the value for a stool sample from an individual mouse, circles represent experiment 1 mice and triangles represent experiment 2 mice.
+D.
+Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the top 20 bacteria out of the `r source_otus_d1` OTUs that varied across sources post-infection.
+The gray vertical line indicates the limit of detection.
+For each timepoint OTUs with differential relative abundances across sources of mice were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs (Table S3).
 
@@
 **Figure 5.
-A subset of bacteria consistently vary across sources despite clindamycin perturbation and *C.
-difficile* challenge.** A-C: plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the `r shared_source_otus_dn1to1` OTUs that consistently varied across sources of mice at the baseline (A), post-clindamycin (B), and post-infection (C) timepoints of the experiment.
-For each timepoint OTus with differential relative abundances across sources of mice were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs (Table S8).
-The grey vertical line indicates the limit of detection.
-
-\newpage
-\includegraphics{figure_6.pdf}
-**Figure 6.
-Clindamycin treatment has the same effects on a subset of taxa regardless of source.** A-B: plots highlighting the median (point) and interquantile range (colored lines) of the top 10 most significant (adjusted *P* value < 0.05) OTUs with relative abundances that changed after clindamycin treatment.
-Data were analyzed by Wilcoxon signed rank test limited to mice that had paired sequence data for day -1 and 0 (N = `r n_paired_for_clind`).
-Tests were performed at the OTU level with Benjamini-Hochberg correction for testing all identified OTUs.
-See Table S9 for complete list of OTUs significantly impacted by clindamycin treatment.
-The grey vertical line indicates the limit of detection.
-
-\newpage
-\includegraphics{figure_7.pdf}
-**Figure 7.
-Key OTUs that influence whether mice cleared *C.
+Bacteria that influence whether mice cleared *C.
 difficile* by day 7.** A.
@@
 Venn diagram that combines Fig.
-S4 summaries of OTUs that were important to the day -1, 0, and 1 classification models (Table S14) and either overlapped with taxa that varied across vendors at the same timepoint, were impacted by clindamycin treatment, or both.
+S4 summaries of OTUs that were important to the day -1, 0, and 1 classification models (Table S8) and either overlapped with taxa that varied across sources at the same timepoint, were impacted by clindamycin treatment, or both.
 See Fig.
 S4 for separate comparisons of taxa from the day -1, 0, and 1 classification models.
-Bold OTUs signify OTUs that were important to more than 1 classification model.
+Red OTUs were important to more than 1 classification model.
 
 \newpage
-\includegraphics{figure_8.pdf}
-**Figure 8: Key OTUs vary across sources throughout the experiment.** A-D.
-Relative abundances of bold OTUs from Fig.
-7A  that were important for at least two classification models are shown over time.
+\includegraphics{figure_6.pdf}
+**Figure 6: Key OTUs vary across sources throughout the experiment.** A-D.
+Relative abundances of red OTUs from Fig.
+5A  that were important for at least two classification models are shown over time.
 A.
@@
 The gray horizontal line represents the limit of detection.
-Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S13) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05).
+Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S9) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05).
 
@@
 **Figure S1.
+Bacterial communities vary between experiments for some sources.** A-F.
+PCoA of $\theta_{YC}$ distances for the baseline fecal bacterial communities within each source of mice.
+Each symbol represents a stool sample from an individual mouse with color corresponding to experiment and shape representing cage mates.
+PERMANOVA was performed within each group to examine the contributions of experiment and cage to observed variation.
+Experiment number and cage only significantly explained observed variation for mice from the Schloss (combined R^2^ = `r adonis_schloss_rsq_sum`; *P* $\le$ `r p_adonis_schloss_max`) and Young (combined R^2^ = `r adonis_young_rsq_sum`; *P* $\le$ `r p_adonis_young_max`) lab colonies (Table S2).
+G-H: Boxplots of the $\theta_{YC}$ distances of the 6 sources of mice relative to mice within the same source and experiment (G) or mice within the same source and between experiments (H) at baseline.
+Symbols represent individual mouse samples: circles for experiment 1 and triangles for experiment 2.
+
+\newpage
+\includegraphics{figure_S2.pdf}
+**Figure S2.
 *C.
-difficile* CFU variation across vendors varies slightly across the 2 experiments.** A-B.
+difficile* CFU variation across sources varies slightly between the 2 experiments.** A-B.
 *C.
@@
 \newpage
-\includegraphics{figure_S2.pdf}
-**Figure S2.
-Only bacterial communities from University of Michigan mice significantly vary between experiments.** A-F.
-PCoA of $\theta_{YC}$ distances for the baseline fecal bacterial communities within each source of mice.
-Each symbol represents a stool sample from an individual mouse with color corresponding to experiment and shape representing cage mates.
-PERMANOVA was performed within each group to examine the contributions of experiment and cage to observed variation.
-Experiment number and cage only significantly explained observed variation for mice from the Schloss (combined R^2^ = `r adonis_schloss_rsq_sum`; *P* $\le$ `r p_adonis_schloss_max`) and Young (combined R^2^ = `r adonis_young_rsq_sum`; *P* $\le$ `r p_adonis_young_max`) lab colonies (Table S7).
-
-\newpage
 \includegraphics{figure_S3.pdf}
@@
 S3) based on the OTU community relative abundances at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1).
-All models performed better than random chance (AUROC = 0.5; all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for all pairwise comparisons; Table S11).
-For list of the 20 OTUs that were ranked as most important to each model, see Table S12.
+All models performed better than random chance (AUROC = 0.5; all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S7) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for all pairwise comparisons; Table S7).
+For list of the 20 OTUs that were ranked as most important to each model, see Table S8.
 
@@
 **Figure S4.
-Key OTUs from classification models based on baseline, post-clindamycin treatment, or post-infection community data vary by source, clindamycin treatment, or both.** A-C.
-Venn diagrams of top 20 important OTUs from baseline (A), post-clindamycin treatment (B), and post-infection (C) classification models (Table S12) that overlapped with OTUs that varied across vendors at the corresponding timepoint, were impacted by clindamycin treatment, or both.
+OTUs from classification models based on baseline, post-clindamycin treatment, or post-infection community data vary by source, clindamycin treatment, or both.** A-C.
+Venn diagrams of top 20 important OTUs from baseline (A), post-clindamycin treatment (B), and post-infection (C) classification models (Table S8) that overlapped with OTUs that varied across sources at the corresponding timepoint, were impacted by clindamycin treatment, or both.
 Bold OTUs signify OTUs that were important to more than 1 classification model.
@@
 Transparency of the circle corresponds to the day of the experiment, each circle represents a sample from an individual mouse at a specific timepoint.
-See Table S5 for PERMANOVA results).
+See Table S2 for PERMANOVA results).
 Circles represent mice from experiment 1 and triangles represent mice from expeirment 2.
 
-**Table S1. *C. difficile* CFU statistical results.**
+**Table S1. Diversity metrics Kruskal-Wallis and pairwise Wilcoxon statistical results.**
 
-**Table S2. Mouse weight change statistical results.**
+**Table S2. PERMANOVA results for mice at specific timepoints and across all timepoints.**
 
-**Table S3. Diversity metrics Kruskal-Wallis statistical results.**
+**Table S3. OTUs with relative abudances that significantly vary across sources at baseline, post-clindamycin, or post-infection timepoints.**
 
-**Table S4. Diversity metrics pairwise Wilcoxon statistical results.**
+**Table S4. *C. difficile* CFU statistical results.**
 
-**Table S5. PERMANOVA results for all mice, all timepoints.**
+**Table S5. Mouse weight change statistical results.**
 
-**Table S6. PERMANOVA results for all mice at baseline, post clindamycin, and post-infection timepoints.**
+**Table S6. OTUs with relative abudances that significantly changed after clindamycin treatment.**
 
-**Table S7. PERMANOVA results of baseline communities within each source.**
+**Table S7. Statistical results of L2-regularized logistic regression model performances.**
 
-**Table S8. OTUs with relative abudances that significantly vary across sources at baseline, post-clindamycin, or post-infection timepoints.**
+**Table S8. Top 20 most important OTUs for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
 
-**Table S9. OTUs with relative abudances that significantly changed after clindamycin treatment.**
-
-**Table S10. Statistical results of L2-regularized logistic regression model performances compared to random chance.**
-
-**Table S11. Pairwise Wilcoxan results comparing all 3 L2-regularized logistic regression model performances.**
-
-**Table S12. Top 20 most important OTUs for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
-
-**Table S13. OTUs with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**
+**Table S9. OTUs with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**

### 2020-07-10 submission/manuscript.tex
Reorganization edits

@@
 
+\textbf{Microbiota variation is high between mice from different
+sources.} To test how the microbiotas of mice from different sources
+impact colonization dynamics after clindamycin exposure, we utilized
+C57BL/6 mice from 6 different sources: two colonies from the University
+of Michigan that were split from each other 10 years ago (the Young and
+Schloss lab colonies) and four commercial source: the Jackson
+Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo
+(which was formerly Harlan). These 4 vendors were chosen because they
+represent commonly used vendors for CDI studies in mice (26, 34--40). We
+used 16S rRNA gene sequencing to characterize the fecal bacterial
+communities across sources at baseline.
+
+Since antibiotics and other risk factors of CDIs are associated with
+decreased microbiota diversity (41), we first examined alpha diversity
+across the 6 sources of mice. Intially, there was a significant
+difference in the number of observed operational taxonomic units (OTUs),
+but not Shannon diversity index (\emph{P}\textsubscript{FDR} = 0.03 and
+\emph{P}\textsubscript{FDR} = 0.052, respectively) across sources of
+mice (Fig. 1A-B and Table S1). Next, we compared the community structure
+of mice from the 6 sources over the course of the experiment using
+principal coordinate analysis (PCoA) of the \(\theta_{YC}\) distances.
+Permutational multivariate analysis of variance (PERMANOVA) analysis
+revealed source and cage interactions explained most of the observed
+variation across fecal communities explained most of the observed
+community variation (combined R\textsuperscript{2} = 0.90, \emph{P} =
+0.0001, Fig. 3 and Table S2). Mice that are co-housed have been shown to
+have similar gut microbiotas due to coprophagy (42) and since mice
+within the same source were housed together, it is not surprising that
+cage also contributed to the observed microbiota variation. Since we
+conducted two experiments approximately 3 months apart, we also compared
+baseline communities across the 2 experiments. Experiment and cage
+significantly explained the observed variation only for the Schloss and
+Young lab mouse colonies (Fig. S2A-B and Table S2). However, most of the
+vendors also clustered by experiment (Fig. S2C-D, F), suggesting there
+was some community variation between the 2 experiments within each
+source, particularly for Schloss, Young, and Envigo mice (Fig. S2G-H).
+After finding differences at the community level, we next identified the
+bacteria that initially varied across sources of mice. There were 268
+OTUs with relative abundances that varied across sources (Fig. 1D and
+Table S3). By using mice from six sources we were able to increase the
+number of microbiota communities to test with the clindamycin \emph{C.
+difficile} colonization mouse model.
+
 \textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
-difficile} 630 colonization regardless of source.} To test how the
-microbiotas of mice from different sources impact colonization dynamics
-after clindamycin exposure, we utilized C57BL/6 mice from 6 different
-sources: two colonies from the University of Michigan that were split
-from each other 10 years ago (the Young and Schloss lab colonies) and
-four commercial source: the Jackson Laboratory, Charles River
-Laboratories, Taconic Biosciences, and Envigo (which was formerly
-Harlan). These 4 vendors were chosen because they represent commonly
-used vendors for CDI studies in mice (26, 34--40).
-
-After a 13-day acclimation period for the mice ordered from vendors, all
-mice were treated with 10 mg/kg clindamycin via intraperitoneal
-injection and one day later challenged with 10\textsuperscript{3}
-\emph{C. difficile} 630 spores (Fig. 1A). Clindamycin was chosen because
-we have previously demonstrated mice are rendered susceptible, but
-consistently cleared the CDI within 9 days (21, 41), clindamycin is
-frequently implicated with human CDIs (42), and is also part of the
-antibiotic treatment for the frequently cited 2008 CDI mouse model (14).
-The day after infection, \emph{C. difficile} was detectable in all mice
-at a similar level (median CFU range: 2.2e+07-1.3e+08;
-\emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
-mice susceptible regardless of source (Fig. 1B). Interestingly,
-variation in \emph{C. difficile} CFU levels across sources of mice
-emerged from days 3-7 post-infection (all \emph{P}\textsubscript{FDR}
-\(\le\) 0.019; Fig. 1B and Table S1), suggesting mouse source is
-associated with \emph{C. difficile} clearance. We conducted two
-experiments approximately 3 months apart and while the colonization
+difficile} 630 colonization, but clearance time varies across sources.}
+After the acclimation period, all mice were treated with 10 mg/kg
+clindamycin via intraperitoneal injection and one day later challenged
+with 10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 2A).
+Clindamycin was chosen because we have previously demonstrated mice are
+rendered susceptible, but consistently cleared \emph{C. difficile}
+within 9 days (21, 43), clindamycin is frequently implicated with human
+CDIs (44), and is also part of the antibiotic treatment for the
+frequently cited 2008 CDI mouse model (14). The day after infection,
+\emph{C. difficile} was detectable in all mice at a similar level
+(median CFU range: 2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15),
+indicating clindamycin rendered all mice susceptible regardless of
+source (Fig. 2B). Interestingly, variation in \emph{C. difficile} CFU
+levels across sources of mice emerged from days 3-7 post-infection (all
+\emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 2B and Table S4),
+suggesting mouse source is associated with \emph{C. difficile}
+clearance. We conducted two experiments and while the colonization
 dynamics were similar across most sources of mice, there was some
 variation between the 2 experiments, particularly for the Schloss and
-Envigo mice (Fig. S1A-B). Although \emph{C. difficile} 630 causes mild
-symptoms in mice comparied to other \emph{C. difficile} strains (43), we
+Envigo mice (Fig. S2A-B). Although \emph{C. difficile} 630 causes mild
+symptoms in mice comparied to other \emph{C. difficile} strains (45), we
 also saw that weight change significantly varied across sources of mice
-with the most weight lost two days post-infection (Fig. 1C and Table
-S2). Importantly, there was also one Jackson and one Envigo mouse that
+with the most weight lost two days post-infection (Fig. 2C and Table
+S5). Importantly, there was also one Jackson and one Envigo mouse that
 died between 1- and 3-days post-infection during the second experiment.
 Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to
-lose more weight (although there was variation between experiments with
-Schloss and Envigo mice), have higher \emph{C. difficile} CFU levels and
-take longer to clear the infection compared to the other sources of
-mice, which was particularly evident 7 days post-infection (Fig. 1B-C,
-Fig. S2C-D), when 57\% of the mice were still colonized with \emph{C.
-difficile} (Fig. S1E). By 9 days post-infection the majority of the mice
-from all sources had cleared \emph{C. difficile} (Fig. S1C) with the
-exception of 1 Taconic mouse from the first experiment and 2 Envigo mice
-from the second experiment. Thus, clindamycin rendered all mice
+lose more weight, have higher \emph{C. difficile} CFU levels and take
+longer to clear the infection compared to the other sources of mice
+(although there was variation between experiments with Schloss and
+Envigo mice), which was particularly evident 7 days post-infection (Fig.
+2B-C, Fig. S2C-D), when 57\% of the mice were still colonized with
+\emph{C. difficile} (Fig. S2E). By 9 days post-infection the majority of
+the mice from all sources had cleared \emph{C. difficile} (Fig. 2C) with
+the exception of 1 Taconic mouse from the first experiment and 2 Envigo
+mice from the second experiment. Thus, clindamycin rendered all mice
 susceptible to \emph{C. difficile} 630 colonization, regardless of
@@
 
-\textbf{Bacterial communities consistently vary across sources despite
-antibiotic and infection perturbations.} Given the well known variation
-in mouse microbiomes across breeding colonies (25), we hypothesized that
-the variation in \emph{C. difficile} clearance could be explained by
-microbiota variation across the 6 sources of mice. We used 16S rRNA gene
-sequencing to characterize the fecal bacterial communities from the mice
-over the course of the experiment. Since antibiotics and other risk
-factors of CDIs are associated with decreased microbiota diversity (44),
-we first examined alpha diversity measures across the 6 sources of mice.
-Examining the bacterial communities at baseline, prior to clindamycin
-treatment, there was a significant difference in the number of observed
-OTUs (\emph{P}\textsubscript{FDR} = 0.03), but not Shannon diversity
-index (\emph{P}\textsubscript{FDR} = 0.052) across sources of mice (Fig.
-2A-B and Table S3). As expected, the clindamycin treatment decreased
-richness and Shannon diversity across all sources of mice, and richness
-and Shannon diversity started to increase 1 day post-infection (Fig.
-2C-D). Interestingly, significant differences in diversity metrics
-across sources (\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged
-after both clindamycin and \emph{C. difficile} infection, with Charles
-River mice having higher richness and Shannon Diversity than most of the
-other sources (Fig 2C-F and Table S4). Although the Charles River mice
-had more diverse microbiotas, the Young and Schloss lab mice were also
-able to clear \emph{C. difficile} faster than the other sources,
-suggesting microbiota diversity alone does not explain the observed
-variation in \emph{C. difficile} colonization across vendors.
-
-Next, we compared the community structure of mice from the 6 sources
-over the course of the experiment using principal coordinate analysis
-(PCoA) of the \(\theta_{YC}\) distances. Permutational multivariate
-analysis of variance (PERMANOVA) analysis revealed source was the
-dominant factor that explained the observed variation across fecal
-communities (R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
+\textbf{Clindamycin treatment alters bacteria in all sources, but a
+subset of bacterial differences across sources persists.} Given the well
+known variation in mouse microbiomes across breeding colonies (25), we
+hypothesized that variation in \emph{C. difficile} clearance would be
+explained by microbiota variation across the 6 sources of mice. We used
+16S rRNA sequencing to characterize microbiota variation across sources
+after clindamycin treatment and throughout the rest of the experiment.
+PERMANOVA analysis revealed source was the dominant factor that
+explained the observed variation across fecal communities
+(R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
 interactions between cage (mice from the same sources were housed
 together, primarily at a density of 2 mice per cage) and day of the
-experiment (Movie S1 and Table S5). Mice that are co-housed have been
-shown to have similar gut microbiotas due to coprophagy (45) and since
-mice within the same source were housed together, it is not surprising
-that cage also contributes to the observed microbiota variation. Since
-the majority of the perturbations happened over the initial days of the
-experiment, we decided to focus on the communities at baseline (day -1),
-after clindamycin treatment (day 0), and post-infection (day 1). For all
-3 timepoints, the source and cage interaction significantly explained
-most of the observed community variation (combined R\textsuperscript{2}
-= 0.90, 0.99, 0.88, respectively; \emph{P} = 0.0001; Fig. 3 and Table
-S6). We also compared baseline communities across the 2 experiments, and
-found experiment and cage significantly explained the observed variation
-only for the Schloss and Young lab mouse colonies (Fig. S2A-B and Table
-S7). However, most of the vendors also clustered by experiment (Fig.
-S2C-D, F), suggesting there was some community variation between the 2
-experiments within each source. Thus, source was the factor that
-explained the most variation observed in the bacterial communities.
-Importantly with the exception of the Schloss and Young colonies, the
-community of each source clustered apart from one another suggesting
-each community had a unique response to clindamycin treatment and
-\emph{C. difficile} challenge.
-
-Since there was some variation in the microbiota structure of mice from
-the same source between experiments at baseline, we next looked at how
-similar the communities were within the same source and between sources
-in response to clindamycin treatment and \emph{C. difficile} challenge
-(Fig. 4) . The baseline communities varied most between experiments for
-Schloss, Young, and Envigo mice and variation between sources of mice
-was high (Fig. 4A). Clindamycin treatment reduced the variation between
-experiments within Schloss, Young and Jackson mice and some of the
-variation between sources diminished, particularly for the Schloss,
-Young, and Charles River mice (Fig. 4B). Post-infection, the community
-variation started to increase within sources of mice and variation
-between sources of mice started to increase towards previous levels
-(Fig. 4C). By using mice from multiple sources we were able to increase
-the number of microbiota communities we tested with the clindamycin
-\emph{C. difficile} colonization mouse model.
-
-After finding differences at the community level, we next identified the
-bacteria that varied across sources of mice over the initial days of the
-experiment. We examined bacterial relative abundances at the operational
-taxonomic unit (OTU) level. Focusing on the baseline communities first,
-there were 268 OTUs (Table S8) with relative abundances that varied
-across sources. Clindamycin treatment reduced the number of taxa with
-relative abundances that varied across sources to 18 OTUs (Table S8).
-After \emph{C. difficile} challenge, there were 44 OTUs (Table S8) with
-significantly different relative abundances across sources, as the
-communities started to recover from antibiotic treatment. In spite of
-the experimental perturbations that occurred during these 3 timepoints,
-there were 12 OTUs (Fig 5A-C) with relative abundances that consistently
-varied across sources. Importantly, some of the OTus that consistently
-varied across sources also shifted with clindamycin treatment. For
-example, \emph{Proteus} increased after clindamycin treatment, but only
-in Taconic mice. \emph{Enterococcus} was primarily found only in mice
-purchased from commercial vendors and also increased after clindamycin
-treatment. In summary, mouse bacterial communities varied significantly
-according to source throughout the course of the experiment and a
-consistent subset of bacterial taxa remained different across sources
-regardless of clindamycin and \emph{C. difficile} challenge.
-
-\textbf{Clindamycin treatment alters a subset of taxa that were found in
-all sources.} Although there were bacteria that consistently varied
-across sources, we also wanted to identify the bacteria that shifted
-after clindamycin treatment, regardless of source. By analyzing all mice
-that had sequence data from fecal samples collected at baseline and
-after clindamycin treatment, we identified 153 OTUs that were altered
-after clindamycin treatment (Fig. 6 and Tables S9). Interestingly, when
-we compared the list of significant clindamycin impacted bacteria with
-the bacteria that consistently varied across groups over the initial 3
-timepoints of our experiment, we found 3 OTUs (\emph{Lachnospiraceae}
-(OTU 130), \emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23))
-overlapped (Fig. 5, Fig. 6C-D). These findings demonstrate that
+experiment (Movie S1 and Table S2). Since the majority of the
+perturbations happened over the initial days of the experiment, we
+decided to focus on the communities after clindamycin treatment (day 0)
+and post-infection (day 1).
+
+As expected, clindamycin treatment decreased richness and Shannon
+diversity across all sources of mice (Fig. 3A-B). Interestingly,
+significant differences in diversity metrics across sources
+(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after clindamycin
+treatment, with Charles River mice having higher richness and Shannon
+diversity than most of the other sources (Fig 3A-B and Table S1). The
+variation in community structures was also diminished by clindamycin
+treatment (Fig. 3C and Table S2). Next we identified the bacteria that
+shifted after clindamycin treatment, regardless of source by analyzing
+all mice that had sequence data from fecal samples collected at baseline
+and after clindamycin treatment. We identified 153 OTUs that were
+altered after clindamycin treatment (Fig. 3D and Table S6) in most mice.
+Despite the significant impact of clindamycin, there were still 18 OTUs
+(Fig. 3E and Table S3) with relative abundances that significantly
+varied across sources. Interestingly, when we compared the list of
+significant clindamycin impacted bacteria with the bacteria that varied
+across sources post-clindamycin, we found 4 OTUs
+(\emph{Enterobacteriaceae (OTU 1)}, \emph{Lachnospiraceae} (OTU 130),
+\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) overlapped
+(Fig. 3D-E and Tables S3, S6). These findings demonstrate that
 clindamycin has a consistent impact on the fecal bacterial communities
-of mice from all sources and only a subset of the OTUs also varied
+of mice from all sources and only a subset of the OTUs continued to vary
 across sources.
 
-\textbf{Source-specific and clindamycin impacted bacteria distinguish
-\emph{C. difficile} colonization status in mice.} After identifying taxa
-that varied by source, changed after clindamycin treatment, or both, we
-next wanted to determine which taxa were influencing the variation in
-\emph{C. difficile} colonization at day 7 (Fig. 1D, Fig. S1C). We
-trained three L2-regularized logistic regression models with input
-bacterial community data from the baseline, post-clindamycin, and
-post-infection timepoints of the experiment to predict \emph{C.
-difficile} colonization status on day 7 (Fig. S3A-B). All models were
-better at predicting \emph{C. difficile} colonization status on day 7
-than random chance (all \emph{P} \(\le\) 5.2e-31; Table S12).
-Interestingly, the model based on the post-clindamycin (day 0) community
-OTU data performed significantly better than all other models with an
-area under the receiving operator characteristic curve (AUROC) of 0.75
-(\emph{P}\textsubscript{FDR} \(\le\) 3.1e-11 for pairwise comparisons;
-Table S13). Thus, we were able to use bacterial relative abundance data
-alone to differentiate mice that had cleared \emph{C. difficile} before
-day 7 from the mice still colonized with \emph{C. difficile} at that
-timepoint. Interestingly, the model built with OTU relative abundance
-data post-clindamycin treatment had the best performance, suggesting how
-the bacterial community responds to clindamycin treatment has the
-greatest influence on subsequent \emph{C. difficile} colonization
-dynamics.
+\textbf{Microbiota variation across sources is maintained after \emph{C.
+difficile} challenge.} Post-infection, significant differences in
+diversity metrics were maintained across sources
+(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Table S1).
+Although the Charles River mice had more diverse microbiotas, the Young
+and Schloss lab mice were also able to clear \emph{C. difficile} faster
+than some of the other sources, suggesting microbiota diversity alone
+does not explain the observed variation in \emph{C. difficile}
+colonization across sources. Source and cage interaction continued to
+explain most of the observed community variation (combined
+R\textsuperscript{2} = 0.88, respectively; \emph{P} = 0.0001; Fig. 4D
+and Table S2). After \emph{C. difficile} challenge, there were 44 OTUs
+(Fig. 4D and Table S3) with significantly different relative abundances
+across sources. In spite of the experimental perturbations that
+occurred, there were 12 OTUs with relative abundances that consistently
+varied across sources at baseline, post-clindamycin, and post-infection.
+Importantly, some of the OTUs that consistently varied across sources
+also shifted with clindamycin treatment. For example, \emph{Proteus}
+increased after clindamycin treatment, but only in Taconic mice.
+\emph{Enterococcus} was primarily found only in mice purchased from
+commercial vendors and also increased after clindamycin treatment. In
+summary, mouse bacterial communities varied significantly according to
+source throughout the course of the experiment and a consistent subset
+of bacterial taxa remained different across sources regardless of
+clindamycin and \emph{C. difficile} challenge.
+
+\textbf{Baseline, post-clindamycin, and post-infection community data
+can predict mice that will clear \emph{C. difficile} by 7 days
+post-infection.} After identifying taxa that varied by source, changed
+after clindamycin treatment, or both, we next wanted to determine which
+taxa were influencing the variation in \emph{C. difficile} colonization
+at day 7 (Fig. 2D, Fig. S2C). We trained three L2-regularized logistic
+regression models with input bacterial community data from the baseline,
+post-clindamycin, and post-infection timepoints of the experiment to
+predict \emph{C. difficile} colonization status on day 7 (Fig. S3A-B).
+All models were better at predicting \emph{C. difficile} colonization
+status on day 7 than random chance (all \emph{P} \(\le\) 1.6e-17; Table
+S7). Interestingly, the model based on the post-clindamycin (day 0)
+community OTU data performed significantly better than the other models
+with an area under the receiving operator characteristic curve (AUROC)
+of 0.78 (\emph{P}\textsubscript{FDR} \(\le\) 3.3e-06 for pairwise
+comparisons; Table S7). Thus, we were able to use bacterial relative
+abundance data alone to differentiate mice that had cleared \emph{C.
+difficile} before day 7 from the mice still colonized with \emph{C.
+difficile} at that timepoint. Interestingly, the model built with OTU
+relative abundance data post-clindamycin treatment had the best
+performance, suggesting how the bacterial community responds to
+clindamycin treatment has the greatest influence on subsequent \emph{C.
+difficile} colonization dynamics.
 
 Next, to examine the bacteria that were driving each model's
-performance, we pulled out the top 20 taxa that had the highest absolute
-feature weights in each of the 6 models (Tables S14-15). First, we
-looked at OTUs from the model with the best performance that was based
-on the post-clindamycin treatment bacterial community data. While most
-of the 20 OTUs had low relative abundances on day 0,
-\emph{Enterobacteriace}, \emph{Bacteroides} and \emph{Proteus} had high
-relative abundances in at least one source of mice and significantly
-varied across sources (Fig. 7A). Next, the top 20 taxa from each model
-were compared to the list of taxa that varied across source at the same
-timepoint (Fig. 5 and Tables S8-9) and the taxa that were altered by
-clindamycin treatment (Fig. 6 and Table S10-11). We found a subset of
-OTUs that were important to the model and overlapped with bacteria that
-varied by either source, clindamycin treatment, or both (Fig. S4,
-S5A-C). Combining the overall results for the 3 OTU models identified 14
-OTUs associated with source, 21 OTUs associated with clindamycin
-treatment, and 6 OTUs associated with both (Fig. 7B). Several OTUs
-(\emph{Bacteroides (OTU 2), Enterococcus (OTU 23), Enterobacteriaceae
-(OTU 1), Porphyromonadaceae (OTU 7)}) appeared across at least 2 models,
-so we examined how the relative abundances of these key taxa varied over
-the course of the experiment (Fig. 8). Throughout the experiment, there
-was at least 1 timepoint where relative abundances of these OTUs
-significantly varied across sources (Table S13). Interestingly, there
-were no OTUs that emerged as consistently enriched or depleted in mice
-that were colonized past 7 days post-infection with \emph{C. difficile}
-630, suggesting multiple bacteria influence the time needed to clear the
-infection. Together, these results suggest the initial bacterial
-communities and their responses to clindamycin influence the clearance
-of \emph{C. difficile}.
+performance, we pulled out the top 20 OTUs that had the highest absolute
+feature weights in each of the 6 models (Table S8). First, we looked at
+OTUs from the model with the best performance that was based on the
+post-clindamycin treatment bacterial community data. While most of the
+20 OTUs had low relative abundances on day 0, \emph{Enterobacteriace},
+\emph{Bacteroides} and \emph{Proteus} had high relative abundances in at
+least one source of mice and significantly varied across sources (Fig.
+5A). Next, the top 20 taxa from each model were compared to the list of
+taxa that varied across source at the same timepoint (Fig. 1D, 3E, 4D
+and Table S3) and the taxa that were altered by clindamycin treatment
+(Fig. 3D and Table S6). We found a subset of OTUs that were important to
+the model and overlapped with bacteria that varied by either source,
+clindamycin treatment, or both (Fig. S4). Combining the overall results
+for the 3 models identified 14 OTUs associated with source, 21 OTUs
+associated with clindamycin treatment, and 6 OTUs associated with both
+(Fig. 5B). Several OTUs (\emph{Bacteroides (OTU 2), Enterococcus (OTU
+23), Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)}) appeared
+across at least 2 models, so we examined how the relative abundances of
+these OTUs varied over the course of the experiment (Fig. 6). Throughout
+the experiment, there was at least 1 timepoint where relative abundances
+of these OTUs significantly varied across sources (Table S9).
+Interestingly, there were no OTUs that emerged as consistently enriched
+or depleted in mice that were colonized past 7 days post-infection with
+\emph{C. difficile} 630, suggesting multiple bacteria influence the time
+needed to clear the infection. Together, these results suggest the
+initial bacterial communities and their responses to clindamycin
+influence the clearance of \emph{C. difficile}.
 
@@
 \emph{Bacteroides (OTU 2), Enterococcus (OTU 23), Enterobacteriaceae
-(OTU 1), Porphyromonadaceae (OTU 7)} (Fig. 8) as candidate bacteria
+(OTU 1), Porphyromonadaceae (OTU 7)} (Fig. 6) as candidate bacteria
 within these communities that were influencing variation in \emph{C.
@@
 \emph{Ruminococcaceae}, and \emph{Turicibacter} were associated with
-resistance (19, 21, 32, 33, 41, 46--48). While we have demonstrated that
+resistance (19, 21, 32, 33, 43, 46--48). While we have demonstrated that
 susceptibility is uniform across sources of mice after clindamycin
@@
 difficile} strain used could also be contributing to the variation in
-\emph{C. difficile} outcomes seen across different research groups (47).
-We found the time needed to naturally clear \emph{C. difficile} varied
-across sources of mice implying that at least in the context of the same
-perturbation, microbiota differences seemed to influence infection
-outcome more than susceptibility. More importantly, we were able to
-reduce the variation observed across sources to identify a subset of
-OTUs that were also important for predicting \emph{C. difficile}
-colonization status 7 days post-infection. Since all but 3 mice
-eventually cleared \emph{C. difficile} 630 by 9 days post-infection and
-the model built with the post-clindamycin OTU relative abundance data
-had the best performance, our results suggest clindamycin treatment had
-a large role in determining \emph{C. difficile} susceptibility and
-clearance in the mice.
+\emph{C. difficile} outcomes seen across different research groups. For
+example, a group found differential colonization outcomes after
+clindamycin treatment, with \emph{C. difficile} 630 and M68 eventually
+becoming undectable while BI-7 remained detectable up to 70 days
+post-treatment (47). We found the time needed to naturally clear
+\emph{C. difficile} varied across sources of mice implying that at least
+in the context of the same perturbation, microbiota differences seemed
+to influence infection outcome more than susceptibility. More
+importantly, we were able to reduce the variation observed across
+sources to identify a subset of OTUs that were also important for
+predicting \emph{C. difficile} colonization status 7 days
+post-infection. Since all but 3 mice eventually cleared \emph{C.
+difficile} 630 by 9 days post-infection and the model built with the
+post-clindamycin OTU relative abundance data had the best performance,
+our results suggest clindamycin treatment had a large role in
+determining \emph{C. difficile} susceptibility and clearance than the
+source of the mice.
 
@@
 viruses, fungi, and parasites in addition to bacteria, and these
-non-bacterial members can also vary across mouse vendors (57, 58). While
+non-bacterial members can also vary across mouse sources (57, 58). While
 our study focused solely on the bacterial portion, viruses and fungi
@@
 analyze differences in \emph{C. difficile} CFU, mouse weight change, and
-alpha diversity across vendors with a Benjamini-Hochberg correction for
+alpha diversity across sources with a Benjamini-Hochberg correction for
 testing multiple timepoints, followed by pairwise Wilcoxon comparisons
@@
 
+\hypertarget{ref-Ross2016}{}
+41. Ross CL, Spinler JK, Savidge TC. 2016. Structural and functional
+changes within the gut microbiota and susceptibility to
+\emph{Clostridium difficile} infection. Anaerobe 41:37--43.
+
+\hypertarget{ref-Nguyen2015}{}
+42. Nguyen TLA, Vieira-Silva S, Liston A, Raes J. 2015. How informative
+is the mouse for human gut microbiota research? Disease Models \&
+Mechanisms 8:1--16.
+
 \hypertarget{ref-Tomkovich2019}{}
-41. Tomkovich S, Lesniak NA, Li Y, Bishop L, Fitzgerald MJ, Schloss PD.
+43. Tomkovich S, Lesniak NA, Li Y, Bishop L, Fitzgerald MJ, Schloss PD.
 2019. The proton pump inhibitor omeprazole does not promote
@@
 \hypertarget{ref-Guh2018}{}
-42. Guh AY, Kutty PK. 2018. \emph{Clostridioides difficile} infection
+44. Guh AY, Kutty PK. 2018. \emph{Clostridioides difficile} infection
 169:ITC49.
@@
 \hypertarget{ref-Theriot2011}{}
-43. Theriot CM, Koumpouras CC, Carlson PE, Bergin II, Aronoff DM, Young
+45. Theriot CM, Koumpouras CC, Carlson PE, Bergin II, Aronoff DM, Young
 VB. 2011. Cefoperazone-treated mice as an experimental platform to
@@
 Gut Microbes 2:326--334.
-
-\hypertarget{ref-Ross2016}{}
-44. Ross CL, Spinler JK, Savidge TC. 2016. Structural and functional
-changes within the gut microbiota and susceptibility to
-\emph{Clostridium difficile} infection. Anaerobe 41:37--43.
-
-\hypertarget{ref-Nguyen2015}{}
-45. Nguyen TLA, Vieira-Silva S, Liston A, Raes J. 2015. How informative
-is the mouse for human gut microbiota research? Disease Models \&
-Mechanisms 8:1--16.
 
@@
 
-\includegraphics{figure_1.pdf} \textbf{Figure 1. Clindamycin is
-sufficient to promote \emph{C. difficile} colonization in all mice, but
-clearance time varies across sources of C57BL/6 mice.} A. Setup of the
-experimental timeline. Mice for the experiments were obtained from 6
-different sources: the Schloss (N = 8) and Young lab (N = 9) colonies at
-the University of Michigan, the Jackson Laboratory (N = 8), Charles
-River Laboratory (N = 8), Taconic Biosciences (N = 8), and Envigo (N =
-8). All mice were administered 10 mg/kg clindamycin intraperitoneally
-(IP) 1 day before challenge with \emph{C. difficile} 630 spores on day
-0. Mice were weighed and feces was collected daily through the end of
-the experiment (9 days post-infection). Note: 3 mice died during course
-of experiment. 1 Taconic mouse prior to infection and 1 Jackson and 1
-Envigo mouse between 1- and 3-days post-infection. B. \emph{C.
-difficile} CFU/gram stool measured over time (N = 20-49 mice per
-timepoint) via serial dilutions. The black line represents the limit of
-detection for the first serial dilution. CFU quantification data was not
-available for each mouse due to early deaths, stool sampling
-difficulties, and not plating all of the serial dilutions. C. Mouse
-weight change measured in grams over time (N = 45-49 mice per
-timepoint), all mice were normalized to the weight recorded 1 day before
-infection. For B-C: timepoints where differences across sources of mice
-were statistically significant by Kruskal-Wallis test with
-Benjamini-Hochberg correction for testing across multiple days (Table S1
-and Table S2) are reflected by the asterisk(s) above each timepoint (*,
-\emph{P} \textless{} 0.05). Lines represent the median for each source
-and circles represent individual mice from experiment 1 while triangles
-represent mice from experiment 2.
-
-\newpage
-
-\includegraphics{figure_2.pdf} \textbf{Figure 2. Differences in
-microbial richness and diversity across mouse sources emerge after
-clindamycin treatment and infection.} A-F. Number of observed OTUs and
-Shannon diversity index values at baseline: day -1 (A-B), after
-clindamycin: day 0 (C-D) and post-infection: day 1 (E-F) timepoints of
-the experiment. Data were analyzed by Kruskal-Wallis test with
+\includegraphics{figure_1.pdf} \textbf{Figure 1. Microbiota variation is
+high between mice from different sources.} A-B. Number of observed OTUs
+(A) and Shannon diversity index values (B) across sources of mice at
+baseline (day -1) of the experiment. Differences across sources were
+analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for
+testing each day of the experiment and the adjusted \emph{P} value was
+\textless{} 0.05 for panel A (Table S1). None of the \emph{P} values
+from pairwise Wilcoxon comparisons between sources were significant
+after Benjamini-Hochberg correction (Table S1). Gray lines represent the
+median values for each source of mice. C. Principal Coordinates Analysis
+(PCoA) of \(\theta_{YC}\) distances of baseline stool samples.
+Permutational multivariate analysis of variance (PERMANOVA) analysis
+demonstrated that source and the interaction between source and cage
+explained most of the variation (combined R\textsuperscript{2} = 0.90,
+\emph{P} = 0.0001, see Table S2). For A-C: each symbol representing the
+value for a stool sample from an individual mouse, circles represent
+experiment 1 mice and triangles represent experiment 2 mice. D. Plots
+highlighting the median (point) and interquantile range (colored lines)
+of the relative abundances for the top 20 bacteria out of the 268 OTUs
+that varied across sources at baseline (Table S3).
+
+\newpage
+
+\includegraphics{figure_2.pdf}
+
+\textbf{Figure 2. Clindamycin is sufficient to promote \emph{C.
+difficile} colonization in all mice, but clearance time varies across
+sources of C57BL/6 mice.} A. Setup of the experimental timeline. Mice
+for the experiments were obtained from 6 different sources: the Schloss
+(N = 8) and Young lab (N = 9) colonies at the University of Michigan,
+the Jackson Laboratory (N = 8), Charles River Laboratory (N = 8),
+Taconic Biosciences (N = 8), and Envigo (N = 8). All mice were
+administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before
+challenge with \emph{C. difficile} 630 spores on day 0. Mice were
+weighed and feces was collected daily through the end of the experiment
+(9 days post-infection). Note: 3 mice died during course of experiment.
+1 Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse
+between 1- and 3-days post-infection. B. \emph{C. difficile} CFU/gram
+stool measured over time (N = 20-49 mice per timepoint) via serial
+dilutions. The black line represents the limit of detection for the
+first serial dilution. CFU quantification data was not available for
+each mouse due to early deaths, stool sampling difficulties, and not
+plating all of the serial dilutions. C. Mouse weight change measured in
+grams over time (N = 45-49 mice per timepoint), all mice were normalized
+to the weight recorded 1 day before infection. For B-C: timepoints where
+differences across sources of mice were statistically significant by
+Kruskal-Wallis test with Benjamini-Hochberg correction for testing
+across multiple days (Table S4 and Table S5) are reflected by the
+asterisk(s) above each timepoint (*, \emph{P} \textless{} 0.05). Lines
+represent the median for each source and circles represent individual
+mice from experiment 1 while triangles represent mice from experiment 2.
+
+\newpage
+
+\includegraphics{figure_3.pdf} \textbf{Figure 3. Clindamycin treatment
+alters bacteria in all sources, but a subset of bacterial differences
+across sources persists.} A-B. Number of observed OTUs (A) and Shannon
+diversity index values (B) across sources of mice after clindamycin
+treatment (day 0). Data were analyzed by Kruskal-Wallis test with
 Benjamini-Hochberg correction for testing each day of the experiment and
-the adjusted \emph{P} value was \textless{} 0.05 for all panels except
-for B (Table S3). Significant \emph{P} values from the pairwise Wilcoxon
-comparisons between sources with Benjamini-Hochberg correction are shown
-(Table S4). For A-F: circles represent experiment 1 mice, while
-triangles represent experiment 2 mice with each symbol representing the
-value for a stool sample from an individual mouse. Gray lines represent
-the median values for each source of mice.
-
-\newpage
-
-\includegraphics{figure_3.pdf} \textbf{Figure 3. Mouse source is the
-variable that explains most of the variation observed in the baseline,
-post-clindamycin, and post-infection bacterial communities.} A-C.
-Principal Coordinates Analysis of \(\theta_{YC}\) distances from stools
-collected at baseline (A), post-clindamycin (B), and post-infection (C)
-timepoints of the experiment. Each symbol represents a stool sample from
-an individual mouse, with circles representing experiment 1 mice and
-triangles representing experiment 2 mice. PERMANOVA analysis
-demonstrated that source and the interaction between source and cage
-explained most of the variation observed in the baseline (combined
-R\textsuperscript{2} = 0.90), post-clindamycin (combined
-R\textsuperscript{2} = 0.99), and post-infection (combined
-R\textsuperscript{2} = 0.88) communities (all \emph{P} = 0.0001, see
-Table S6).
-
-\newpage
-
-\includegraphics{figure_4.pdf} \textbf{Figure 4. High inter-group
-variation across mouse sources is diminished by clindamycin treatment}
-A-C. Boxplots of the \(\theta_{YC}\) distances of the 6 sources of mice
-relative to mice within the same source and experiment, mice within the
-same source and between experiments, mice within the same source, and
-mice from other groups at the baseline (A), after clindamycin treatment
-(B), and post-infection (C) timepoints. For comparisons within mice from
-the same source, symbols represent individual mouse samples: circles for
-experiment 1 and triangles for experiment 2.
-
-\newpage
-
-\includegraphics{figure_5.pdf} \textbf{Figure 5. A subset of bacteria
-consistently vary across sources despite clindamycin perturbation and
-\emph{C. difficile} challenge.} A-C: plots highlighting the median
-(point) and interquantile range (colored lines) of the relative
-abundances for the 12 OTUs that consistently varied across sources of
-mice at the baseline (A), post-clindamycin (B), and post-infection (C)
-timepoints of the experiment. For each timepoint OTus with differential
-relative abundances across sources of mice were identified by
-Kruskal-Wallis test with Benjamini-Hochberg correction for testing all
-identified OTUs (Table S8). The grey vertical line indicates the limit
-of detection.
-
-\newpage
-
-\includegraphics{figure_6.pdf} \textbf{Figure 6. Clindamycin treatment
-has the same effects on a subset of taxa regardless of source.} A-B:
-plots highlighting the median (point) and interquantile range (colored
-lines) of the top 10 most significant (adjusted \emph{P} value
-\textless{} 0.05) OTUs with relative abundances that changed after
-clindamycin treatment. Data were analyzed by Wilcoxon signed rank test
-limited to mice that had paired sequence data for day -1 and 0 (N = 31).
-Tests were performed at the OTU level with Benjamini-Hochberg correction
-for testing all identified OTUs. See Table S9 for complete list of OTUs
-significantly impacted by clindamycin treatment. The grey vertical line
-indicates the limit of detection.
-
-\newpage
-
-\includegraphics{figure_7.pdf} \textbf{Figure 7. Key OTUs that influence
+the adjusted \emph{P} value was \textless{} 0.05. Significant \emph{P}
+values from the pairwise Wilcoxon comparisons between sources with
+Benjamini-Hochberg correction are shown (Table S1). C. PCoA of
+\(\theta_{YC}\) distances from stools collected post-clindamycin.
+PERMANOVA analysis demonstrated that source and the interaction between
+source and cage explained most of the variation observed
+post-clindamycin (combined R\textsuperscript{2} = 0.99, \emph{P} =
+0.0001, see Table S2).For A-C, each symbol represents a stool sample
+from an individual mouse, with circles representing experiment 1 mice
+and triangles representing experiment 2 mice. D. Plots highlighting the
+median (point) and interquantile range (colored lines) of the top 10
+OTUs out of 153 with relative abundances that changed after clindamycin
+treatment (adjusted \emph{P} value \textless{} 0.05). Data were analyzed
+by Wilcoxon signed rank test limited to mice that had paired sequence
+data for day -1 and 0 (N = 31), with Benjamini-Hochberg correction for
+testing all identified OTUs (Table S6). The gray vertical line indicates
+the limit of detection. E. Plots highlighting the median (point) and
+interquantile range (colored lines) of the relative abundances for the
+18 OTUs that varied across sources after clindamycin treatment (Table
+S3).
+
+\newpage
+
+\includegraphics{figure_4.pdf} \textbf{Figure 4. Microbiota variation
+across sources is maintained after \emph{C. difficile} challenge.} A-B.
+Number of observed OTUs (A) and Shannon diversity index values (B)
+across sources of mice post-infection (day 1). Data were analyzed by
+Kruskal-Wallis test with Benjamini-Hochberg correction for testing each
+day of the experiment and the adjusted \emph{P} value was \textless{}
+0.05. Significant \emph{P} values from the pairwise Wilcoxon comparisons
+between sources with Benjamini-Hochberg correction are shown (Table S1).
+PCoA of \(\theta_{YC}\) distances of post-infection stool samples.
+PERMANOVA analysis demonstrated that source and the interaction between
+source and cage explained most of the variation (combined
+R\textsuperscript{2} = 0.88, \emph{P} = 0.0001, Table S2). For A-C: each
+symbol representing the value for a stool sample from an individual
+mouse, circles represent experiment 1 mice and triangles represent
+experiment 2 mice. D. Plots highlighting the median (point) and
+interquantile range (colored lines) of the relative abundances for the
+top 20 bacteria out of the 44 OTUs that varied across sources
+post-infection. The gray vertical line indicates the limit of detection.
+For each timepoint OTUs with differential relative abundances across
+sources of mice were identified by Kruskal-Wallis test with
+Benjamini-Hochberg correction for testing all identified OTUs (Table
+S3).
+
+\newpage
+
+\includegraphics{figure_5.pdf} \textbf{Figure 5. Bacteria that influence
 whether mice cleared \emph{C. difficile} by day 7.} A. Baseline relative
@@
 OTUs that were important to the day -1, 0, and 1 classification models
-(Table S14) and either overlapped with taxa that varied across vendors
-at the same timepoint, were impacted by clindamycin treatment, or both.
-See Fig. S4 for separate comparisons of taxa from the day -1, 0, and 1
-classification models. Bold OTUs signify OTUs that were important to
-more than 1 classification model.
-
-\newpage
-
-\includegraphics{figure_8.pdf} \textbf{Figure 8: Key OTUs vary across
-sources throughout the experiment.} A-D. Relative abundances of bold
-OTUs from Fig. 7A that were important for at least two classification
-models are shown over time. A. \emph{Bacteroides} (OTU 2), which varied
-across sources throughout the experiment. B-C. \emph{Enterobacteriaceae}
-(B) and \emph{Enterococcus} (C), which significantly varied across
-sources and were impacted by clindamycin treatment. D.
-\emph{Porphyromonadaceae} (OTU 7), which was significantly impacted by
-clindamycin treatment and examining relative abundance dynamics over the
-course of the experiment, revealed timepoints where relative abundances
-also significantly varied across sources of mice. Symbols represent the
-relative abundance data for an individual mouse, circles represent mice
-that cleared \emph{C. difficile} by day 7, X-shapes represent mice that
-were still colonized with \emph{C. difficile}, and open circles
-represent mice that did not have \emph{C. difficile} CFU counts for day
-7 post-infection. Colored lines indicate the median relative abundances
-for each source. The gray horizontal line represents the limit of
-detection. Timepoints where differences across sources of mice were
-statistically significant by Kruskal-Wallis test with Benjamini-Hochberg
-correction for testing across multiple days (Table S13) are identified
-by the asterisk(s) above each timepoint (*, P \textless{} 0.05).
-
-\newpage
-
-\includegraphics{figure_S1.pdf} \textbf{Figure S1. \emph{C. difficile}
-CFU variation across vendors varies slightly across the 2 experiments.}
+(Table S8) and either overlapped with taxa that varied across sources at
+the same timepoint, were impacted by clindamycin treatment, or both. See
+Fig. S4 for separate comparisons of taxa from the day -1, 0, and 1
+classification models. Red OTUs were important to more than 1
+classification model.
+
+\newpage
+
+\includegraphics{figure_6.pdf} \textbf{Figure 6: Key OTUs vary across
+sources throughout the experiment.} A-D. Relative abundances of red OTUs
+from Fig. 5A that were important for at least two classification models
+are shown over time. A. \emph{Bacteroides} (OTU 2), which varied across
+sources throughout the experiment. B-C. \emph{Enterobacteriaceae} (B)
+and \emph{Enterococcus} (C), which significantly varied across sources
+and were impacted by clindamycin treatment. D. \emph{Porphyromonadaceae}
+(OTU 7), which was significantly impacted by clindamycin treatment and
+examining relative abundance dynamics over the course of the experiment,
+revealed timepoints where relative abundances also significantly varied
+across sources of mice. Symbols represent the relative abundance data
+for an individual mouse, circles represent mice that cleared \emph{C.
+difficile} by day 7, X-shapes represent mice that were still colonized
+with \emph{C. difficile}, and open circles represent mice that did not
+have \emph{C. difficile} CFU counts for day 7 post-infection. Colored
+lines indicate the median relative abundances for each source. The gray
+horizontal line represents the limit of detection. Timepoints where
+differences across sources of mice were statistically significant by
+Kruskal-Wallis test with Benjamini-Hochberg correction for testing
+across multiple days (Table S9) are identified by the asterisk(s) above
+each timepoint (*, P \textless{} 0.05).
+
+\newpage
+
+\includegraphics{figure_S1.pdf} \textbf{Figure S1. Bacterial communities
+vary between experiments for some sources.} A-F. PCoA of \(\theta_{YC}\)
+distances for the baseline fecal bacterial communities within each
+source of mice. Each symbol represents a stool sample from an individual
+mouse with color corresponding to experiment and shape representing cage
+mates. PERMANOVA was performed within each group to examine the
+contributions of experiment and cage to observed variation. Experiment
+number and cage only significantly explained observed variation for mice
+from the Schloss (combined R\textsuperscript{2} = 0.99; \emph{P} \(\le\)
+0.033) and Young (combined R\textsuperscript{2} = 0.95; \emph{P} \(\le\)
+0.029) lab colonies (Table S2). G-H: Boxplots of the \(\theta_{YC}\)
+distances of the 6 sources of mice relative to mice within the same
+source and experiment (G) or mice within the same source and between
+experiments (H) at baseline. Symbols represent individual mouse samples:
+circles for experiment 1 and triangles for experiment 2.
+
+\newpage
+
+\includegraphics{figure_S2.pdf} \textbf{Figure S2. \emph{C. difficile}
+CFU variation across sources varies slightly between the 2 experiments.}
 A-B. \emph{C. difficile} CFU/gram of stool quantification over time for
@@
 
-\includegraphics{figure_S2.pdf} \textbf{Figure S2. Only bacterial
-communities from University of Michigan mice significantly vary between
-experiments.} A-F. PCoA of \(\theta_{YC}\) distances for the baseline
-fecal bacterial communities within each source of mice. Each symbol
-represents a stool sample from an individual mouse with color
-corresponding to experiment and shape representing cage mates. PERMANOVA
-was performed within each group to examine the contributions of
-experiment and cage to observed variation. Experiment number and cage
-only significantly explained observed variation for mice from the
-Schloss (combined R\textsuperscript{2} = 0.99; \emph{P} \(\le\) 0.033)
-and Young (combined R\textsuperscript{2} = 0.95; \emph{P} \(\le\) 0.027)
-lab colonies (Table S7).
-
-\newpage
-
 \includegraphics{figure_S3.pdf} \textbf{Figure S3. Bacterial community
@@
 performed better than random chance (AUROC = 0.5; all \emph{P} \(\le\)
-5.2e-31; Table S12) and the model built with post-clindamycin treated
+1.6e-17; Table S7) and the model built with post-clindamycin treated
 bacterial OTU relative abundances had the best performance
-((\emph{P}\textsubscript{FDR} \(\le\) 3.1e-11 for all pairwise
-comparisons; Table S11). For list of the 20 OTUs that were ranked as
-most important to each model, see Table S12.
-
-\newpage
-
-\includegraphics{figure_S4.pdf} \textbf{Figure S4. Key OTUs from
+((\emph{P}\textsubscript{FDR} \(\le\) 3.3e-06 for all pairwise
+comparisons; Table S7). For list of the 20 OTUs that were ranked as most
+important to each model, see Table S8.
+
+\newpage
+
+\includegraphics{figure_S4.pdf} \textbf{Figure S4. OTUs from
 classification models based on baseline, post-clindamycin treatment, or
@@
 post-clindamycin treatment (B), and post-infection (C) classification
-models (Table S12) that overlapped with OTUs that varied across vendors
+models (Table S8) that overlapped with OTUs that varied across sources
 at the corresponding timepoint, were impacted by clindamycin treatment,
@@
 corresponds to the day of the experiment, each circle represents a
-sample from an individual mouse at a specific timepoint. See Table S5
+sample from an individual mouse at a specific timepoint. See Table S2
 for PERMANOVA results). Circles represent mice from experiment 1 and
@@
 
-\textbf{Table S1. \emph{C. difficile} CFU statistical results.}
-
-\textbf{Table S2. Mouse weight change statistical results.}
-
-\textbf{Table S3. Diversity metrics Kruskal-Wallis statistical results.}
-
-\textbf{Table S4. Diversity metrics pairwise Wilcoxon statistical
-results.}
-
-\textbf{Table S5. PERMANOVA results for all mice, all timepoints.}
-
-\textbf{Table S6. PERMANOVA results for all mice at baseline, post
-clindamycin, and post-infection timepoints.}
-
-\textbf{Table S7. PERMANOVA results of baseline communities within each
-source.}
-
-\textbf{Table S8. OTUs with relative abudances that significantly vary
+\textbf{Table S1. Diversity metrics Kruskal-Wallis and pairwise Wilcoxon
+statistical results.}
+
+\textbf{Table S2. PERMANOVA results for mice at specific timepoints and
+across all timepoints.}
+
+\textbf{Table S3. OTUs with relative abudances that significantly vary
 across sources at baseline, post-clindamycin, or post-infection
@@
 
-\textbf{Table S9. OTUs with relative abudances that significantly
+\textbf{Table S4. \emph{C. difficile} CFU statistical results.}
+
+\textbf{Table S5. Mouse weight change statistical results.}
+
+\textbf{Table S6. OTUs with relative abudances that significantly
 changed after clindamycin treatment.}
 
-\textbf{Table S10. Statistical results of L2-regularized logistic
-regression model performances compared to random chance.}
-
-\textbf{Table S11. Pairwise Wilcoxan results comparing all 3
-L2-regularized logistic regression model performances.}
-
-\textbf{Table S12. Top 20 most important OTUs for each of the 3
+\textbf{Table S7. Statistical results of L2-regularized logistic
+regression model performances.}
+
+\textbf{Table S8. Top 20 most important OTUs for each of the 3
 L2-regularized logistic regression models based on OTU relative
@@
 
-\textbf{Table S13. OTUs with relative abudances that significantly
-varied across sources of mice on at least 1 day of the experiment by
+\textbf{Table S9. OTUs with relative abudances that significantly varied
+across sources of mice on at least 1 day of the experiment by
 Kruskal-Wallis test.}

### 2020-08-03 code/compare_models.R
Figure edits and making sure all code works with R and R package updates

@@ -61,5 +61,6 @@ all %>%
          cv_lci, cv_q1, cv_median, cv_q3, cv_uci,
          test_lci, test_q1, test_median, test_q3, test_uci
   ) %>%
-  unnest() %>%
+  unnest(cols = c(test, cv_lci, cv_q1, cv_median, cv_q3, cv_uci, test_lci, test_q1, 
+                  test_median, test_q3, test_uci)) %>%
   write_tsv("data/process/classification_cv_test_compare.tsv")

### 2020-08-03 code/diversity.R

@@ -14,7 +14,7 @@ shannon_all <- diversity_data %>%
                       breaks=color_vendors,
                       labels=color_vendors)+
   geom_boxplot(outlier.shape = NA, size = 1.2)+
-  geom_jitter(shape=19, size=2, alpha=0.6, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.1), show.legend = FALSE) +
+  geom_jitter(shape=19, size=2, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.1), show.legend = FALSE) +
   labs(title=NULL, 
        x="Source",
        y="Shannon Diversity Index")+
@@ -30,7 +30,7 @@ richness_all <- diversity_data %>%
                       breaks=color_vendors,
                       labels=color_vendors)+
   geom_boxplot(outlier.shape = NA, size = 1.2)+
-  geom_jitter(shape=19, size=2, alpha=0.6, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.1), show.legend = FALSE) +
+  geom_jitter(shape=19, size=2, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.1), show.legend = FALSE) +
   labs(title=NULL, 
        x="Source",
        y="Number of Observed OTUs")+
@@ -186,7 +186,7 @@ shannon_dx_plot <- function(timepoint){
                        labels=shape_experiment) +
     scale_x_discrete(guide = guide_axis(n.dodge = 2))+
     geom_errorbar(aes(ymax = median_shannon, ymin = median_shannon), color = "gray50", size = 1)+ #Add lines to indicate the median for each group to the plot
-    geom_jitter(aes(shape = experiment), size=2, alpha=0.6, show.legend = FALSE) +
+    geom_jitter(aes(shape = experiment), size=2, show.legend = FALSE) +
     labs(title=NULL, 
          x="Source",
          y="Shannon Diversity Index")+
@@ -245,7 +245,7 @@ sobs_dx_plot <- function(timepoint){
                      labels=shape_experiment) +  
   scale_x_discrete(guide = guide_axis(n.dodge = 2))+
   geom_errorbar(aes(ymax = median_sobs, ymin = median_sobs), color = "gray50", size = 1)+ #Add lines to indicate the median for each group to the plot
-  geom_jitter(aes(shape = experiment), size=2, alpha=0.6, show.legend = FALSE) +
[…]

### 2020-08-03 code/figure_5.R

@@ -1,13 +1,9 @@
 source("code/functions.R")
 
-a1 <- ggdraw() + draw_image("results/figures/d0_model_otu_Enterobacteriaceae (OTU 1).png")
-a2 <- ggdraw() + draw_image("results/figures/d0_model_otu_Bacteroides (OTU 2).png")
-a3 <- ggdraw() + draw_image("results/figures/d0_model_otu_Proteus (OTU 16).png")
-a <- plot_grid(a1, a2, a3, labels = NULL, ncol = 1)
-top_half <- plot_grid(a, labels = c("A"), label_size = 12, ncol=1)
+a <- ggdraw() + draw_image("results/figures/d0_top10_otus.png")
 b <- ggdraw() + draw_image("results/figures/venn_overall_otus.png")
 
-plot_grid(top_half, b, labels = c("", "B"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_5.pdf", width=6, height=9)+
-  ggsave("submission/figure_5.pdf", width=6, height=9)
+plot_grid(a, b, labels = "AUTO", label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_5.pdf", width=7, height=9)+
+  ggsave("submission/figure_5.pdf", width=7, height=9)
 

### 2020-08-03 code/taxa_stats.R

@@ -530,7 +530,7 @@ baseline_exp_o <- function(vendor_name){
     write_tsv(path = paste0("data/process/otu_stats_experiment_", vendor_name, ".tsv"))
 }
 
-#Test only for days where input communities were used to create classification models (Day -1, 0, 1):
+#Test only Schloss, Young, and Envigo mice
 test_sources <- c("Schloss", "Young", "Envigo")
 for (v in test_sources){
   baseline_exp_o(v)
@@ -563,6 +563,14 @@ diff_otu_experiment_Schloss #0
 diff_otu_experiment_Young #58
 diff_otu_experiment_Envigo #3
   #Porphyromonadaceae OTU 52 and 139, Deltaproteobacteria OTU 174
+
+#Check if any overlap with taxa that varied between sources at baseline
+Young_exp_baseline <- intersect_all(diff_otu_experiment_Young, `sig_otu_day-1`)
+#21 overlap
+Young_exp_baseline <- intersect_all(diff_otu_experiment_Young, `sig_otu_day-1`[1:20])
+#O overlap with top 20
+Envigo_exp_baseline <- intersect_all(diff_otu_experiment_Envigo, `sig_otu_day-1`)
+#Deltaproteobacteria (OTU 174) 
 
 #Examine top 20 taxa from logistic regression model that performed the best: Communities after clindamycin treatment at the OTU level----
 interp_otus <- read_tsv("data/process/combined_top20_otus_all_models.tsv")

### 2020-08-03 code/venn_diagram_comparisons.R

@@ -149,19 +149,19 @@ venn_otus <- function(source_comp, clind_comp, title){
 }
 
 # Venn diagram of Day -1 overlapping OTUs----
-dn1_otus_venn_plot <- venn_otus(dayn1_and_interp_dn1_o, paired_and_interp_dn1_o, "Day -1 model key OTU comparisons")
+dn1_otus_venn_plot <- venn_otus(dayn1_and_interp_dn1_o, paired_and_interp_dn1_o, "Day -1 model OTU comparisons")
 save_plot("results/figures/venn_dn1_otus.png", dn1_otus_venn_plot, base_aspect_ratio = 1.8)
 
 # Venn diagram of Day 0 overlapping OTUs----
-d0_otus_venn_plot <- venn_otus(day0_and_interp_d0_o, paired_and_interp_d0_o, "Day 0 model key OTU comparisons")
+d0_otus_venn_plot <- venn_otus(day0_and_interp_d0_o, paired_and_interp_d0_o, "Day 0 model OTU comparisons")
 save_plot("results/figures/venn_d0_otus.png", d0_otus_venn_plot, base_aspect_ratio = 1.8)
 
 # Venn diagram of Day -1 overlapping OTUs----
-d1_otus_venn_plot <- venn_otus(day1_and_interp_d1_o, paired_and_interp_d1_o, "Day 1 model key OTU comparisons")
+d1_otus_venn_plot <- venn_otus(day1_and_interp_d1_o, paired_and_interp_d1_o, "Day 1 model OTU comparisons")
 save_plot("results/figures/venn_d1_otus.png", d1_otus_venn_plot, base_aspect_ratio = 1.8)
 
 # Venn diagram of combined overlapping OTUs----
-combined_otus_venn_plot <- venn_otus(dayn1to1_and_interp_combined, paired_and_interp_combined, "Key taxa comparisons for day -1, 0, and 1 models")
+combined_otus_venn_plot <- venn_otus(dayn1to1_and_interp_combined, paired_and_interp_combined, "OTU comparisons for day -1, 0, and 1 models")
 save_plot("results/figures/venn_overall_otus.png", combined_otus_venn_plot, base_aspect_ratio = 1.8)
 
 #Comparison of combined Venn diagram OTUs to significant OTUs that varied by source on day -1, 0, and 1:

### 2020-08-03 code/tables_supplemental.R
Reodering supplemental tables in combined excel workbook

@@ -20,38 +20,39 @@ richness_pairwise_results <- read_tsv("data/process/richness_stats_sig_days.tsv"
 #Table S2: Diversity pairwise wilcoxon results
 diversity_pairwise_results <- rbind(shannon_pairwise_results, richness_pairwise_results)
 
-#Table S3-S5: PERMANOVA Results----
 #Write PERMANOVA results to supplemental table excel file.
-#Table S3: PERMANOVA results for day -1, day 0, and day 1
+#Table S3: PERMANOVA results for day -1, day 0, and day 1----
 adonis_dn1_to_0 <- read_tsv("data/process/adonis_dn1-1.tsv")
-#Table S4: PERMANOVA results within sources at day -1
+#Table S4: PERMANOVA results within sources at day -1----
 dn1_source <- read_tsv("data/process/adonis_dn1_source.tsv")
-#Table S5: PERMANOVA results all sources over time
-adonis_all_samples <- read_tsv("data/process/adonis_all.tsv") 
 
-
-#Table S6-S8: OTU statistics for differences across sources of mice on day -1, 0 and 1----
+#OTU statistics for differences across sources of mice on day -1, 0 and 1----
 otu_stats_dn1to1_combined <- read_tsv("data/process/otu_stats_dn1to1_combined.tsv")
-#Table S6: OTUs that vary across sources on day -1
+#Table S5: OTUs that vary between sources on day -1----
 kw_w_sig_dn1 <- otu_stats_dn1to1_combined %>% filter(day == -1)
-#Table S7: OTUs that vary across sources on day 0
-kw_w_sig_d0 <- otu_stats_dn1to1_combined %>% filter(day == 0)
-#Table S8: OTUs that vary across sources on day 1
-kw_w_sig_d1 <- otu_stats_dn1to1_combined %>% filter(day == 1)
 
-#Table S9: C. difficile CFU statistical results----
+#Table S6: C. difficile CFU statistical results----
 cfu_kruskal_stats_adjust <- read_tsv("data/process/cfu_stats_all_days.tsv")
 
-#Table S10: Mouse weight change statistical results----
+#Table S7: Mouse weight change statistical results----
 weight_kruskal_stats_adjust <- read_tsv("data/process/weight_stats_all_days.tsv")
 
-#Table S11: OTUs that changed after clindamycin treatment----
+#Table S8: OTUs that vary between sources on day 0----
[…]

### 2020-08-03 submission/manuscript.Rmd
Incorporate Pat's and my own edits

@@
 difficile* colonization levels through 9 days post-infection.
-The microbiota was profiled via 16S rRNA gene sequencing to examine the variation across sources and alterations due to clindamycin treatment and *C.
+The microbiota were profiled via 16S rRNA gene sequencing to examine the variation across sources and alterations due to clindamycin treatment and *C.
 difficile* challenge.
@@
 To identify bacteria that were most important to colonization regardless of the source, we created logistic regression models that successfully classified mice based on whether they cleared *C.
-difficile* by 7 days post-infection using baseline, post-clindamycin, and post-infection community composition data.
+difficile* by 7 days post-infection using baseline, post-clindamycin, and 1-day post-infection community composition data.
 With these models, we identified 4 bacteria that varied across sources (*Bacteroides*), were altered by clindamycin (*Porphyromonadaceae*), or both (*Enterobacteriaceae* and *Enterococcus*).
-Microbiota variation across sources better emulates human interindividual variation and can help identify bacterial drivers of phenotypic variation in the context of CDIs.
+Microbiota variation across sources better emulates human inter-individual variation and can help identify bacterial drivers of phenotypic variation in the context of CDIs.
 
@@
 *Clostridioides difficile* is a leading nosocomial infection.
-Although perturbation to the gut microbiota has been established as a key risk factor, there is variation in who becomes asymptomatically colonized, develops an infection, or has an infection with adverse outcomes.
+Although perturbation to the gut microbiota is an established risk, there is variation in who becomes asymptomatically colonized, develops an infection, or has an infection with adverse outcomes.
 *C.
@@
 difficile* pathogenesis questions.
-However, the inter-individual variation between mice is less than what is observed in humans, particularly if just one source of mice is used.
-In this study, we administered clindamycin to mice from 6 different breeding colonies and challenged them with *C.
+However, the inter-individual variation between mice from the same breeding facility is less than what is observed in humans.
+Therefore, we challenged mice from 6 different breeding colonies with *C.
 difficile*.
-Interestingly, only a subset of the bacteria that vary across sources were associated with how long *C.
+We found that the starting microbial community structures and *C.
+difficile* persistence varied by the source of mice.
+Interestingly, a subset of the bacteria that varied across sources were associated with how long *C.
 difficile* was able to colonize.
-Future studies examining the interplay between the microbiota and *C.
-difficile* should consider using mice from multiple sources to better reflect human interindividual variation.
+By increasing the inter-individual diversity of the starting communities, we were able to better model human diversity.
+This provided a more nuanced perspective of *C.
+difficile* pathogenesis.
 
@@
 
-Antibiotics are a common risk factor for *Clostridioides difficile* infections (CDIs), but there is variation in who goes on to develop severe or recurrent CDIs after exposure [@Teng2019; @Kelly2012].
+Antibiotics are a common risk factor for *Clostridioides difficile* infections (CDIs) due to there effect on the intestinal microbiota, but there is variation in who goes on to develop severe or recurrent CDIs after exposure [@Teng2019; @Kelly2012].
 Additionally, asymptomatic colonization, where *C.
 difficile* is detectable, but symptoms are absent has been documented in infants and adults [@Zacharioudakis2015; @Crobach2018].
-The intestinal microbiome has been implicated in asymptomatic colonization [@Zhang2015; @VanInsberghe2020], susceptibility to CDIs [@Mancabelli2017, @Duvallet2017], and adverse CDI outcomes [@Seekatz2016; @Khanna2016; @Pakpour2017; @Lee2020].
+The intestinal microbiota has been implicated in asymptomatic colonization [@Zhang2015; @VanInsberghe2020], susceptibility to CDIs [@Mancabelli2017, @Duvallet2017], and adverse CDI outcomes [@Seekatz2016; @Khanna2016; @Pakpour2017; @Lee2020].
+However, it is not clear how much inter-individual microbiota variation contributes to the range of outcomes observed after *C.
+difficile* exposure relative to other risk factors.
 
@@
 difficile*, including the role of the microbiota and efficacy of potential therapeutics for treating CDIs [@Best2012].
-However, variation in the microbiome between mice from the same breeding colony is much less than the variation observed between humans [@Baxter2014; @Nagpal2018].
-Additionally, studying the contribution of the microbiota to a particular disease phenotype in one set of lab mice after the same perturbation could yield a number of findings of which only a fraction may be driving the phenotype.
+However, variation in the microbiota between mice from the same breeding colony is much less than the inter-individual variation observed between humans [@Baxter2014; @Nagpal2018].
+Studying CDIs in mice with a homogenous microbiota is likely to overstate the importance of individual mechanisms.
+Using mice that have a more heterogenous microbiota would allow researchers to identify and validate more generalizable mechanisms responsible for CDI.
 
-In the past, our group has attempted to introduce more microbiome variation into the CDI mouse model by using a variety of antibiotic treatments [@Reeves2011; @Schubert2015; @Jenior2017; @Jenior2018].
-An alternative approach to maximize microbiome variation is to use mice from multiple sources [@Velazquez2019; @Osbelt2020].
-Microbiome differences between different mouse vendors have been well documented and shown to influence susceptibility to a variety of diseases [@Stough2016; @Alegre2019], including enteric infections [@EtienneMesmin2017; @Velazquez2019; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018].
-Additionally, research groups have observed different CDI outcomes in mice despite using similar models and the microbiome has been proposed as one factor potentially mediating CDI susceptibility and outcomes [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014].
-Here we examined how variations in the baseline microbiome and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C.
+In the past, our group has attempted to introduce more variation into the mouse microbiota by using a variety of antibiotic treatments [@Reeves2011; @Schubert2015; @Jenior2017; @Jenior2018].
+An alternative approach to maximize microbiota variation is to use mice from multiple sources [@Velazquez2019; @Osbelt2020].
+The differences between the microbiota of mice from vendors have been well documented and shown to influence susceptibility to a variety of diseases [@Stough2016; @Alegre2019], including enteric infections [@EtienneMesmin2017; @Velazquez2019; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018].
+Different research groups have also observed variation in CDI outcomes despite using similar murine models [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014].
+Here we examined how variations in the baseline microbiota and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C.
 difficile* colonization and the time needed to clear the infection.
@@
 d1_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 1) %>% select(r_sq) %>% sum
-p_adonis_intial <- adonis_stats_dn101 %>% select(p) %>% max 
+p_adonis_all <- adonis_stats_dn101 %>% select(p) %>% max 
 p_adonis_min <- adonis_stats_dn101 %>% select(p) %>% min #All p values are same
@@
 
-**Microbiota variation is high between mice from different sources.** To test how the microbiotas of mice from different sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other 10 years ago (the Young and Schloss lab colonies) and four commercial source: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan).
-These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019].
-We used 16S rRNA gene sequencing to characterize the fecal bacterial communities across sources at baseline.
+**The variation in the microbiota is high between mice from different sources.** We obtained C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other in 2010 (the Young and Schloss lab colonies) and four commercial sources: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan).
+These 4 vendors were chosen because they are commonly used for murine CDI studies [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019].
+Two experiments were conducted, approximately 3 months apart.
 
-Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity across the 6 sources of mice.
-Intially, there was a significant difference in the number of observed operational taxonomic units (OTUs), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_richness` and *P*~FDR~ = `r padj_dn1_shannon`, respectively) across sources of mice (Fig.
-1A-B and Table S1).
-Next, we compared the community structure of mice from the 6 sources over the course of the experiment using principal coordinate analysis (PCoA) of the $\theta_{YC}$ distances.
-Permutational multivariate analysis of variance (PERMANOVA) analysis revealed source and cage interactions explained most of the observed variation across fecal communities explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, *P* = `r p_adonis_intial`, Fig.
-3 and Table S2).
-Mice that are co-housed have been shown to have similar gut microbiotas due to coprophagy [@Nguyen2015] and since mice within the same source were housed together, it is not surprising that cage also contributed to the observed microbiota variation.
-Since we conducted two experiments approximately 3 months apart, we also compared baseline communities across the 2 experiments.
-Experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig.
-S2A-B and Table S2).
+We sequenced the 16S rRNA gene from fecal samples collected from these mice after they acclimated to the University of Michigan animal housing environment.
+We first examined the alpha diversity across the 6 sources of mice.
+There was a significant difference in the richness (i.e.
+number of observed operational taxonomic units (OTUs)), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_richness` and *P*~FDR~ = `r padj_dn1_shannon`, respectively) across the sources of mice (Fig.
+1A-B and Tables S1-2).
+Next, we compared the community structure of mice (Fig.
+1C).
+Interactions between the source and cage effects explained most of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, Fig.
+1C and Table S3).
+Mice that are co-housed tend to have similar gut microbiotas due to coprophagy [@Nguyen2015].
+Since mice within the same source were housed together, it is not surprising that the cage effect also contributed to the observed microbiota variation.
+There were some differences between the 2 experiments we conducted, as the experiment and cage effects significantly explained the observed community variation for the Schloss and Young lab mouse colonies (Fig.
+S2A-B and Table S4).
 However, most of the vendors also clustered by experiment (Fig.
@@
 S2G-H).
-After finding differences at the community level, we next identified the bacteria that initially varied across sources of mice.
-There were `r source_otus_dn1` OTUs with relative abundances that varied across sources (Fig.
-1D and Table S3).
-By using mice from six sources we were able to increase the number of microbiota communities to test with the clindamycin *C.
-difficile* colonization mouse model.
+After finding differences at the community level, we next identified the bacteria that varied between sources of mice.
+There were `r source_otus_dn1` OTUs with relative abundances that were significantly different between the sources (Fig.
+1D and Table S5).
+Though we saw differences between experiments at the community level, there were no OTUs that were significantly different between experiments within Schloss, Young, and Envigo mice (Data not shown).
+By using mice from six sources we were able to increase the number of microbiota communities to treat with clindamycin and challenge with *C.
+difficile*.
 
@@
 **Clindamycin treatment renders all mice susceptible to *C.
-difficile* 630 colonization, but clearance time varies across sources.** After the acclimation period, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C.
+difficile* 630 colonization, but clearance time varies across sources.** Clindamycin is frequently implicated with human CDIs [@Guh2018] and was part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008].
+We have previously demonstrated mice are rendered susceptible to *C.
+difficile*, but cleared the pathogen within 9 days when treated with clindamycin alone [@Jenior2018; @Tomkovich2019].
+All mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C.
 difficile* 630 spores (Fig.
 2A).
-Clindamycin was chosen because we have previously demonstrated mice are rendered susceptible, but consistently cleared *C.
-difficile* within 9 days [@Jenior2018; @Tomkovich2019], clindamycin is frequently implicated with human CDIs [@Guh2018], and is also part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008].
 The day after infection, *C.
@@
 2B).
-Interestingly, variation in *C.
-difficile* CFU levels across sources of mice emerged from days 3-7 post-infection (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig.
-2B and Table S4), suggesting mouse source is associated with *C.
+However, between 3 and 7 days post-infection, we observed variation in *C.
+difficile* levels across sources of mice (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig.
+2B and Table S6).
+This suggested the mouse source was associated with *C.
 difficile* clearance.
-We conducted two experiments and while the colonization dynamics were similar across most sources of mice, there was some variation between the 2 experiments, particularly for the Schloss and Envigo mice (Fig.
+While the colonization dynamics were similar between the two experiments, the Schloss mice took longer to clear in the 1st experiment and the Envigo mice took longer to clear in the 2nd experiment (Fig.
 S2A-B).
-Although *C.
-difficile* 630 causes mild symptoms in mice comparied to other *C.
-difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig.
-2C and Table S5).
-Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment.
-Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to lose more weight, have higher *C.
-difficile* CFU levels and take longer to clear the infection compared to the other sources of mice (although there was variation between experiments with Schloss and Envigo mice), which was particularly evident 7 days post-infection (Fig.
+The change in the mice's weight significantly varied across sources of mice with the most weight lost two days post-infection (Fig.
+2C and Table S7).
+There was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment.
+Mice obtained from Jackson, Taconic, and Envigo tended to lose more weight, have higher *C.
+difficile* CFU levels and take longer to clear the infection compared to the other sources of mice (although there was variation between experiments with Schloss and Envigo mice).
+This was particularly evident 7 days post-infection (Fig.
 2B-C, Fig.
@@
 ```
-**Clindamycin treatment alters bacteria in all sources, but a subset of bacterial differences across sources persists.** Given the well known variation in mouse microbiomes across breeding colonies [@Alegre2019], we hypothesized that variation in *C.
+**Clindamycin treatment alters bacteria in all sources, but a subset of bacterial differences across sources persists.** Given the variation in mouse microbiotas that we observed across breeding colonies, we hypothesized that variation in *C.
 difficile* clearance would be explained by microbiota variation across the 6 sources of mice.
-We used 16S rRNA sequencing to characterize microbiota variation across sources after clindamycin treatment and throughout the rest of the experiment.
-PERMANOVA analysis revealed source was the dominant factor that explained the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage (mice from the same sources were housed together, primarily at a density of 2 mice per cage) and day of the experiment (Movie S1 and Table S2).
-Since the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the communities after clindamycin treatment (day 0) and post-infection (day 1).
+As expected, clindamycin treatment decreased the richness and Shannon diversity across all sources of mice (Fig.
+3A-B).
+Interestingly, significant differences in diversity metrics between sources (*P*~FDR~ < 0.05) emerged after clindamycin treatment, with Charles River mice having higher richness and Shannon diversity than most of the other sources (Fig 3A-B and Tables S1-2).
+The clindamycin treatment decreased the variation in community structures between sources of mice.
+Source and cage effects explained almost all of the observed variation between communities (combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig.
+3C and Table S3).
+However, there were only `r source_otus_d0` OTUs (Fig.
+3D and Table S8) with relative abundances that significantly varied between sources.
+Next we identified the bacteria that shifted after clindamycin treatment, regardless of source by analyzing paired fecal samples from mice that were collected at baseline and after clindamycin treatment.
+We identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment in most mice (Fig.
+3E and Table S9).
+When we compared the list of significant clindamycin impacted bacteria with the bacteria that varied between sources post-clindamycin, we found 4 OTUs (*Enterobacteriaceae (OTU 1)*, *Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) overlapped (Fig.
+3D-E and Tables S8-9).
+Importantly, some of the OTUs that varied between sources also shifted with clindamycin treatment.
+For example, *Proteus* increased after clindamycin treatment (Fig.
+3D), but only in Taconic mice.
+*Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment (Fig.
+3D).
+These findings demonstrate that clindamycin had a consistent impact on the fecal bacterial communities of mice from all sources and only a subset of the OTUs continued to vary between sources.
 
-As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice (Fig.
-3A-B).
-Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after clindamycin treatment, with Charles River mice having higher richness and Shannon diversity than most of the other sources (Fig 3A-B and Table S1).
-The variation in community structures was also diminished by clindamycin treatment (Fig.
-3C and Table S2).
-Next we identified the bacteria that shifted after clindamycin treatment, regardless of source by analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment.
-We identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment (Fig.
-3D and Table S6) in most mice.
-Despite the significant impact of clindamycin, there were still `r source_otus_d0` OTUs (Fig.
-3E and Table S3) with relative abundances that significantly varied across sources.
-Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that varied across sources post-clindamycin, we found 4 OTUs (*Enterobacteriaceae (OTU 1)*, *Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) overlapped (Fig.
-3D-E and Tables S3, S6).
-These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all sources and only a subset of the OTUs continued to vary across sources.
+**Microbiota variation between sources is maintained after *C.
+difficile* challenge.** One day post-infection, significant differences in diversity metrics remained across sources (*P*~FDR~ < 0.05, Fig 4A-B and Tables S1-2).
+Although the Charles River mice had more diverse microbiotas and were also able to clear *C.
+difficile* faster than some of the other sources.
+Microbiota diversity did not explain the observed variation in *C.
+difficile* colonization across sources considering the Young and Schloss mice had the lowest diversity after clindamycin treatment and were able to clear *C.
+difficile* earlier than Jackson, Taconic and Envigo mice.
+Source and the interactions between source and cage effects continued to explain most of the observed community variation (combined R^2^ = `r d1_rsq_sum`; *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig.
+4D and Table S3).
+One day after *C.
+difficile* challenge, there were `r source_otus_d1` OTUs (Fig.
+4D and Table S10) with significantly different relative abundances across sources.
 
-**Microbiota variation across sources is maintained after *C.
-difficile* challenge.** Post-infection, significant differences in diversity metrics  were maintained across sources (*P*~FDR~ < 0.05, Fig 4A-B and Table S1).
-Although the Charles River mice had more diverse microbiotas, the Young and Schloss lab mice were also able to clear *C.
-difficile* faster than some of the other sources, suggesting microbiota diversity alone does not explain the observed variation in *C.
-difficile* colonization across sources.
-Source and cage interaction continued to explain most of the observed community variation (combined R^2^ = `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig.
-4D and Table S2).
-After *C.
-difficile* challenge, there were `r source_otus_d1` OTUs (Fig.
-4D and Table S3) with significantly different relative abundances across sources.
-In spite of the experimental perturbations that occurred, there were `r shared_source_otus_dn1to1` OTUs with relative abundances that consistently varied across sources at baseline, post-clindamycin, and post-infection.
-Importantly, some of the OTUs that consistently varied across sources also shifted with clindamycin treatment.
-For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice.
-*Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment.
-In summary, mouse bacterial communities varied significantly according to source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C.
+Throughout the experiment the source of mice continued to be the dominant factor that explained the observed variation across fecal communities (PERMANOVA R^2^ = `r rsq_adonis_source`, *P* `r format.pval(p_adonis_source, eps = 0.001)`) followed by interactions between cage effects  and the day of the experiment (Movie S1 and Table S11).
+Mice fecal samples from the same source of mice continued to cluster closely to each other throughout the experiment.
+By 7 days post-infection, when approximately `r format(100-percent_colonized_d7, digits = 1)`% mice had cleared *C.
+difficile*, most of the mice still had not recovered to their baseline community structure (Fig.
+4E).
+Distance from the baseline community between sources did not explain the variation in *C.
+difficile* clearance since Schloss and Young mice cleared *C.
+difficile* faster, but their communities were a greater distance from baseline 7 days post-infection.
+In summary, mouse bacterial communities varied significantly between sources throughout the course of the experiment and a consistent subset of bacterial taxa remained different between sources regardless of clindamycin and *C.
 difficile* challenge.
@@
 interp_combined <- c(interp_otus_dn1, interp_otus_d0, interp_otus_d1)
+
+#List of OTUs that overlapped with day 0 models:
+#Overlap between day 0 and day -1 model:
+d0_vs_dn1 <- interp_otus %>% filter(model_input_day == 0 | model_input_day == -1) %>% filter(duplicated(OTU)) %>% pull(OTU)
+#Overlap between day 0 and day 1 model:
+d0_vs_d1 <- interp_otus %>% filter(model_input_day == 0 | model_input_day == 1) %>% filter(duplicated(OTU)) %>% pull(OTU)
+#Overlap between day 0 model and the day -1 or 1 models:
+overlapping_across_models <- c(d0_vs_dn1, d0_vs_d1)
+overlapping_across_models <- str_replace(overlapping_across_models, " \\(.*\\)", "") #remove (OTU) from string
+list_overlapping_across_models <- paste(overlapping_across_models, collapse= ", ") #Format to output in paper
 
@@
 key_otus <- unique(c(key_source_otus, key_clind_otus))
+key_otus <- str_replace(key_otus, " \\(.*\\)", "") #remove (OTU) from string
 list_key_otus <- paste(key_otus, collapse= ", ") #Format to output in paper
@@
 **Baseline, post-clindamycin, and post-infection community data can predict mice that will clear *C.
-difficile* by 7 days post-infection.** After identifying taxa that varied by source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C.
+difficile* by 7 days post-infection.** After identifying taxa that varied between sources, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C.
 difficile* colonization at day 7 (Fig.
-2D, Fig.
+2B, Fig.
 S2C).
-We trained three L2-regularized logistic regression models with input bacterial community data from the baseline, post-clindamycin, and post-infection timepoints of the experiment to predict *C.
+We trained three L2-regularized logistic regression models with input bacterial community data from the baseline (day = -1), post-clindamycin (day = 0), and post-infection (day = 1) timepoints of the experiment to predict *C.
 difficile* colonization status on day 7 (Fig.
@@
 All models were better at predicting *C.
-difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S7).
-Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than the other models with an area under the receiving operator characteristic curve (AUROC) of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S7).
-Thus, we were able to use bacterial relative abundance data alone to differentiate mice that had cleared *C.
+difficile* colonization status on day 7 than random chance (all *P* `r format.pval(max_p_vs_random, eps = 0.001)`, Table S12).
+The model based on the post-clindamycin (day 0) community OTU data performed significantly better than the other models with an area under the receiving operator characteristic curve (AUROC) of `r median_otu_day0_AUROC` (*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for pairwise comparisons, Table S13).
+Thus, we were able to use bacterial relative abundance data from the time of *C.
+difficile* challenge to differentiate mice that had cleared *C.
 difficile* before day 7 from the mice still colonized with *C.
 difficile* at that timepoint.
-Interestingly, the model built with OTU relative abundance data post-clindamycin treatment had the best performance, suggesting how the bacterial community responds to clindamycin treatment has the greatest influence on subsequent *C.
+This result suggests the bacterial community's response to clindamycin treatment had the greatest influence on subsequent *C.
 difficile* colonization dynamics.
 
-Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 OTUs that had the highest absolute feature weights in each of the 6 models (Table S8).
-First, we looked at OTUs from the model with the best performance that was based on the post-clindamycin treatment bacterial community data.
-While most of the 20 OTUs had low relative abundances on day 0, *Enterobacteriace*, *Bacteroides* and *Proteus* had high relative abundances in at least one source of mice and significantly varied across sources (Fig.
+To examine the bacteria that were driving each model’s performance, we selected the 20 OTUs that had the highest absolute feature weights in each of the 3 models (Table S14).
+First, we looked at OTUs from the model with the best performance, which was based on the post-clindamycin treatment (day 0) bacterial community data.
+Out of the 10 highest ranked OTUs, 7 OTUs (*Bacteroides*, *Escherichia/Shigella*, 2 *Lachnospiraceae*,  *Lactobacillus*, *Porphyromonadaceae*, and *Ruminococcaceae*) were associated with *C.
+difficile* colonization 7 days post-infection, while 3 OTUs (*Enterobacteriaceae*, *Lachnospiraceae*, *Porphyromonadaceae*) were associated with clearance (Fig.
 5A).
-Next, the top 20 taxa from each model were compared to the list of taxa that varied across source at the same timepoint (Fig.
-1D, 3E, 4D and Table S3) and the taxa that were altered by clindamycin treatment (Fig.
-3D and Table S6).
-We found a subset of OTUs that were important to the model and overlapped with bacteria that varied by either source, clindamycin treatment, or both (Fig.
+Next, we examined whether any of the top 20 ranked OTUs from the baseline (day 0) model were also important in the other 2 classification models based on baseline (day -1) and 1 day post-infection community data.
+We identified 6 OTUs (`r list_overlapping_across_models`) that were important to the day 0 model and either the baseline or 1 day post-infection models (Table S14).
+Thus, a subset of bacterial OTUs were important for determining *C.
+difficile* colonization dynamics across multiple timepoints.
+
+To determine whether the OTUs driving the classification models also varied between sources, were altered by clindamycin treatment, or both, we examined whether the top 20 ranked OTUs from each model overlapped with OTUs that varied between sources (Fig.
+1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin treatment (Fig.
+3E and Table S9).
+We found a subset of OTUs that were important to the baseline (day -1), post-clindamycin (day 0), and 1 day post-infection models and overlapped with bacteria that varied between sources, were altered by clindamycin treatment, or both (Fig.
 S4).
-Combining the overall results for the 3 models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig.
+Combining the overall comparison results for the 3 models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig.
 5B).
-Several OTUs (*`r list_key_otus`*) appeared across at least 2 models, so we examined how the relative abundances of these OTUs varied over the course of the experiment (Fig.
+Several OTUs (*`r list_key_otus`*) that overlapped with our previous analyses appeared across at least 2 models, so we examined how the relative abundances of these OTUs varied over the course of the experiment (Fig.
 6).
-Throughout the experiment, there was at least 1 timepoint where relative abundances of these OTUs significantly varied across sources (Table S9).
-Interestingly, there were no OTUs that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection with *C.
-difficile* 630, suggesting multiple bacteria influence the time needed to clear the infection.
+Throughout the experiment, there was at least 1 timepoint where relative abundances of these OTUs significantly varied between sources (Table S15).
+Interestingly, there were no OTUs that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection, suggesting multiple bacteria influence *C.
+difficile* colonization dynamics.
 Together, these results suggest the initial bacterial communities and their responses to clindamycin influence the clearance of *C.
@@
 By running our CDI model with mice from 6 different sources, we were able to identify bacterial taxa that were unique to sources throughout the experiment as well as taxa that were universally impacted by clindamycin.
-We trained L2 logistic regression models with baseline, post-clindamycin treatment, and post-infection fecal community data that could predict whether mice cleared *C.
+We trained L2 logistic regression models with baseline (day -1), post-clindamycin treatment (day 0), and 1-day post-infection fecal community data that could predict whether mice cleared *C.
 difficile* by 7 days post-infection better than random chance.
@@
 difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by source, were impacted by clindamycin treatment, or both.
-Overall, our results demonstrate clindamycin is sufficient to render mice from multiple sources susceptible to CDI and only a subset of the interindividual microbiota variation across mice from different sources was associated with the time needed to clear *C.
+Overall, our results demonstrated clindamycin was sufficient to render mice from multiple sources susceptible to CDI and only a subset of the interindividual microbiota variation across mice from different sources was associated with the time needed to clear *C.
 difficile*.
@@
 
-Although all mice were susceptible to *C.
-difficile* colonization, mice from Jackson, Taconic, and Envigo mice tended to remain colonized through at least 7 days post-infection.
-We identified a subset of bacteria that were important in predicting whether a mouse was still colonized with *C.
-difficile* 7 days post-infection.
-These results suggest a subset of the bacterial community is responsible for determining the length of time needed to clear *C.
-difficile* colonization.
-
-Differences in CDI mouse model studies have been attributed to different intestinal microbiotas of mice from different sources.
+Differences in CDI mouse model studies have been attributed to intestinal microbiota variation across sources.
 For example, groups using the same clindamycin treatment and C57BL/6 mice had different *C.
@@
 For example, a group found differential colonization outcomes after clindamycin treatment, with *C.
-difficile* 630 and M68 eventually becoming undectable while BI-7 remained detectable up to 70 days post-treatment [@Lawley2012].
+difficile* 630 and M68 infections eventually becoming undectable while strain BI-7 remained detectable up to 70 days post-treatment [@Lawley2012].
 We found the time needed to naturally clear *C.
@@
 Since all but 3 mice eventually cleared *C.
-difficile* 630 by 9 days post-infection and the model built with the post-clindamycin OTU relative abundance data had the best performance, our results suggest clindamycin treatment had a large role in determining *C.
+difficile* 630 by 9 days post-infection and the model built with the post-clindamycin (day 0) OTU relative abundance data had the best performance, our results suggest clindamycin treatment had a larger role in determining *C.
 difficile* susceptibility and clearance than the source of the mice.
@@
 Importantly, our study using mice from 6 different sources increased the variation between groups of mice compared to using 1 source alone, to better reflect the inter-individual microbiota variation observed in humans.
-Encouragingly, decreased *Bifidobacterium*, *Porphyromonas*, *Ruminococcaceae* and *Lachnospiraceae* and increased *Enterobacteriaceae*, *Enterococcus*, *Lactobacillus*, and *Proteus* have all been associated with human CDIs [@Mancabelli2017, @Duvallet2017] and were well represented in our study, suggesting most of the mouse sources are suitable for gaining insights into microbiota associated factors influencing *C.
+Encouragingly, decreased *Bifidobacterium*, *Porphyromonas*, *Ruminococcaceae* and *Lachnospiraceae* and increased *Enterobacteriaceae*, *Enterococcus*, *Lactobacillus*, and *Proteus* have all been associated with human CDIs [@Mancabelli2017, @Duvallet2017] and were well represented in our study, suggesting most of the mouse sources are suitable for gaining insights into the bacteria influencing *C.
 difficile* colonization and infections in humans.
-An important exception was *Enterococcus*, which was primarily absent from the mice from University of Michigan colonies and *Proteus*, which was only found in Taconic mice.
+An important exception was *Enterococcus*, which was primarily absent from University of Michigan colonies and *Proteus*, which was only found in Taconic mice.
 Importantly, the fact that some CDI-associated bacteria were only found in a subset of mice has important implications for future CDI mouse model studies.
@@
 There are several limitations to our work.
-The microbiome is composed of viruses, fungi, and parasites in addition to bacteria, and these non-bacterial members can also vary across mouse sources [@Rasmussen2019; @Mims2020].
+The microbiota is composed of viruses, fungi, and parasites in addition to bacteria, and these non-bacterial members can also vary across sources of mice [@Rasmussen2019; @Mims2020].
 While our study focused solely on the bacterial portion, viruses and fungi have also begun to be implicated in the context of CDIs or FMT treatments for recurrent CDIs [@Markey2018; @Stewart2019; @Ott2017; @Zuo2017; @Zuo2018].
@@
 difficile* colonization status and metagenomes from these bacteria have been shown to contain the bile acid-inducible gene cluster necessary for secondary bile acid formation and ability to produce butyrate [@Vital2019; @Collins2015].
-Interestingly, butyrate has previously been shown to vary across vendors and mediated resistance to *Citrobacter rodentium* infection, a model of enterohemorrhagic and enteropathogenic *Escherichia coli* infections [@Osbelt2020].
-Evidence for immunological toning differences in IgA and Th17 cells across mice from different vendors have also been documented and [@Fransen2015; @Ivanov2009] may also influence response to CDI, particularly in the context of severe CDIs [@Azrad2018; @Saleh2019].
+Interestingly, butyrate has previously been shown to vary across mouse vendors and mediated resistance to *Citrobacter rodentium* infection, a model of enterohemorrhagic and enteropathogenic *Escherichia coli* infections [@Osbelt2020].
+Evidence for immunological toning differences in IgA and Th17 cells across mice from different vendors have also been documented and [@Fransen2015; @Ivanov2009] could influence the host response to CDI [@Azrad2018; @Saleh2019].
 The outcome after *C.
@@
 difficile* 630.
-For those interested in dissecting the contribution of the microbiome to *C.
+For those interested in dissecting the contribution of the microbiota to *C.
 difficile* pathogenesis and treatments, using multiple sources of mice may yield more insights than a single model alone.
-Furthermore, for studies wanting to examine the interplay between a particular bacterial taxon such as *Enterococcus* and *C.
+Furthermore, for studies wanting to examine the interplay between a particular bactera such as *Enterococcus* and *C.
 difficile*, these results could serve as a resource for selecting which mice to order to address the question.
@@
 ST was supported by the Michigan Institute for Clincial and Health Research Postdoctoral Translation Scholars Program (UL1TR002240).
-We thank members of the Schloss lab for feedback on planning the experiments and data presentation, as well as code tutorials and feedback through Code club.
+We thank members of the Schloss lab for feedback on planning the experiments and data presentation, as well as code tutorials and feedback through Code Club.
 In particular, we want to thank Begüm Topçuoğlu for help with implementing L2 logistic regression models using her [pipeline](https://github.com/SchlossLab/ML_pipeline_microbiome), Ana Taylor for help with media preparation and sample collection, and Nicholas Lesniak for his critical feedback on the manuscript.
@@
 Female C57BL/7 mice were obtained from 6 different sources: The Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, Envigo, and two colonies at the University of Michigan (the Schloss lab colony and the Young lab colony).
-The Young lab colony was originally established with mice purchased from Jackson, and the Schloss lab colony was established 10 years ago with mice donated from the Young lab.
+The Young lab colony was originally established with mice purchased from Jackson, and the Schloss lab colony was established in 2010 with mice donated from the Young lab.
 The 4 groups of mice purchased from vendors were allowed to acclimate to the University of Michigan mouse facility for 13 days prior to starting the experiment.
@@
 To examine whether there was an overall difference in predictive performance across the 6 classification models we used the Kruskal-Wallis test followed by pairwise Wilcoxan comparisons with Benjamini-Hochberg correction for multiple hypothesis testing.
-The tidyverse package was used to wrangle and graph data (v `r paste(packageVersion("tidyverse")[1])`) [@Tidyverse2019].
+The tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) was used to wrangle and graph data  [@Tidyverse2019].
 
@@
 Microbiota variation is high between mice from different sources.** A-B.
-Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice at baseline (day -1) of the experiment.
-Differences across sources were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 for panel A (Table S1).
-None of the *P* values from pairwise Wilcoxon comparisons between sources were significant after Benjamini-Hochberg correction (Table S1).
+Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice at baseline (day -1 of the experiment).
+Differences between sources were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 for panel A (Table S1).
+None of the *P* values from pairwise Wilcoxon comparisons between sources were significant after Benjamini-Hochberg correction (Table S2).
 Gray lines represent the median values for each source of mice.
@@
 Principal Coordinates Analysis (PCoA) of $\theta_{YC}$ distances of baseline stool samples.
-Permutational multivariate analysis of variance (PERMANOVA) analysis demonstrated that source and the interaction between source and cage explained most of the variation (combined R^2^ = `r dn1_rsq_sum`, *P* = `r p_adonis_intial`,  see Table S2).
-For A-C:  each symbol representing the value for a stool sample from an individual mouse, circles represent experiment 1 mice and triangles represent experiment 2 mice.
+Source and the interaction between source and cage effects explained most of the variation (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, see Table S3).
+For A-C:  each symbol represents the value for a stool sample from an individual mouse, circles represent experiment 1 mice and triangles represent experiment 2 mice.
 D.
-Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the top 20 bacteria out of the `r source_otus_dn1` OTUs that varied across sources at baseline (Table S3).
+Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the top 20 bacteria out of the `r source_otus_dn1` OTUs that varied across sources at baseline (Table S5).
 
@@
 Clindamycin is sufficient to promote *C.
-difficile* colonization in all mice, but clearance time varies across sources of C57BL/6 mice.** A.
+difficile* colonization in all mice, but clearance time varies across sources.** A.
 Setup of the experimental timeline.
@@
 Mouse weight change measured in grams over time (N = `r min_weight_n`-`r max_weight_n` mice per timepoint), all mice were normalized to the weight recorded 1 day before infection.
-For B-C: timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S4 and Table S5) are reflected by the asterisk(s) above each timepoint (\*\, *P* < 0.05).
+For B-C: timepoints where differences between sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S6 and Table S7) are reflected by the asterisk above each timepoint (\*\, *P* < 0.05).
 Lines represent the median for each source and circles represent individual mice from experiment 1 while triangles represent mice from experiment 2.
@@
 Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice after clindamycin treatment (day 0).
-Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05.
-Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S1).
+Differences between sources were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 (Table S1).
+Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S2).
 C.
 PCoA of $\theta_{YC}$ distances from stools collected post-clindamycin.
-PERMANOVA analysis demonstrated that source and the interaction between source and cage explained most of the variation observed post-clindamycin (combined R^2^ = `r d0_rsq_sum`, *P* = `r p_adonis_intial`, see Table S2).For A-C, each symbol represents a stool sample from an individual mouse, with circles representing experiment 1 mice and triangles representing experiment 2 mice.
+Source and the interaction between source and cage effects explained most of the variation observed post-clindamycin (PERMANOVA combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, see Table S3).
+For A-C, each symbol represents a stool sample from an individual mouse, with circles representing experiment 1 mice and triangles representing experiment 2 mice.
 D.
+Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the `r source_otus_d0` OTUs (Table S8)that varied between sources after clindamycin treatment (day 0).
+E.
 Plots highlighting the median (point) and interquantile range (colored lines) of the top 10 OTUs out of `r clindamycin_otus_sig` with relative abundances that changed after clindamycin treatment (adjusted *P* value < 0.05).
-Data were analyzed by Wilcoxon signed rank test limited to mice that had paired sequence data for day -1 and 0 (N = `r n_paired_for_clind`), with Benjamini-Hochberg correction for testing all identified OTUs (Table S6).
+Data were analyzed by Wilcoxon signed rank test of mice that had paired sequence data for baseline (day -1) and post-clindamycin (day 0) timepoints (N = `r n_paired_for_clind`), with Benjamini-Hochberg correction for testing all identified OTUs (Table S9).
 The gray vertical line indicates the limit of detection.
-E.
-Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the `r source_otus_d0` OTUs that varied across sources after clindamycin treatment (Table S3).
 
@@
 difficile* challenge.** A-B.
-Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice post-infection (day 1).
-Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05.
-Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S1).
-PCoA of $\theta_{YC}$ distances of post-infection stool samples.
-PERMANOVA analysis demonstrated that source and the interaction between source and cage explained most of the variation (combined R^2^ = `r d1_rsq_sum`, *P* = `r p_adonis_intial`, Table S2).
-For A-C:  each symbol representing the value for a stool sample from an individual mouse, circles represent experiment 1 mice and triangles represent experiment 2 mice.
+Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice 1-day post-infection.
+Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 (Table S1).
+Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S2).
+PCoA of $\theta_{YC}$ distances of 1-day post-infection stool samples.
+Source and the interaction between source and cage effects explained most of the variation between fecal communities (PERMANOVA combined R^2^ = `r d1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, Table S3).
+For A-C:  each symbol represents the value for a stool sample from an individual mouse, circles represent experiment 1 mice and triangles represent experiment 2 mice.
 D.
-Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the top 20 bacteria out of the `r source_otus_d1` OTUs that varied across sources post-infection.
+Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the top 20 bacteria out of the `r source_otus_d1` OTUs that varied between sources 1-day post-infection.
 The gray vertical line indicates the limit of detection.
-For each timepoint OTUs with differential relative abundances across sources of mice were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs (Table S3).
+For each timepoint OTUs with differential relative abundances across sources of mice were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs (Table S10).
+E.
+$\theta_{YC}$ distances of fecal samples collected 7-days post-infection relative to the baseline (day -1) sample for each mouse.
+Each symbol represents an individual mouse.
+Gray lines represent the median for each source.
 
@@
 difficile* by day 7.** A.
-Baseline relative abundance data for 3 of the OTUs from the classification model based on day 0 OTU relative abundances that significantly varied across sources of mice and had high relative abundances in the community.
-Symbols represent the relative abundance data for an individual mouse, circles represent mice that cleared *C.
-difficile* by day 7, X-shapes represent mice that were still colonized with *C.
-difficile*, and open circles represent mice that did not have *C.
-difficile* CFU counts for day 7 post-infection.
+Baseline (day 0) relative abundance data for the 10 OTUs with the highest rankings based on feature weights in the baseline (day 0) classification model.
+Red font represents OTUs that correlated with *C.
+difficile* colonization and blue font represents OTUs that correlated with clearance.
+Symbols represent the relative abundance data for an individual mouse.
 Gray lines indicate the median relative abundances for each source.
-Asterisks are shown for pairwise Wilcoxon comparisons with Benjamini-Hochberg correction where *P* < 0.05.
 B.
 Venn diagram that combines Fig.
-S4 summaries of OTUs that were important to the day -1, 0, and 1 classification models (Table S8) and either overlapped with taxa that varied across sources at the same timepoint, were impacted by clindamycin treatment, or both.
-See Fig.
-S4 for separate comparisons of taxa from the day -1, 0, and 1 classification models.
+S4 summaries of OTUs that were important to the day -1, 0, and 1 classification models (Table S14) and either overlapped with taxa that varied across sources at the same timepoint, were impacted by clindamycin treatment, or both.
 Red OTUs were important to more than 1 classification model.
@@
 \includegraphics{figure_6.pdf}
-**Figure 6: Key OTUs vary across sources throughout the experiment.** A-D.
+**Figure 6: OTUs associated with *C.
+difficile* colonization dynamics vary across sources throughout the experiment.** A-D.
 Relative abundances of red OTUs from Fig.
@@
 D.
-*Porphyromonadaceae* (OTU 7), which was significantly impacted by clindamycin treatment and examining relative abundance dynamics over the course of the experiment, revealed timepoints where relative abundances also significantly varied across sources of mice.
-Symbols represent the relative abundance data for an individual mouse, circles represent mice that cleared *C.
-difficile* by day 7, X-shapes represent mice that were still colonized with *C.
-difficile*, and open circles represent mice that did not have *C.
-difficile* CFU counts for day 7 post-infection.
+*Porphyromonadaceae* (OTU 7), which was significantly impacted by clindamycin treatment and after examining relative abundance dynamics over the course of the experiment was found to also significantly vary between sources of mice on days -1, 5, 6, 7, and 9 of the experiment.
+Symbols represent the relative abundance data for an individual mouse.
 Colored lines indicate the median relative abundances for each source.
 The gray horizontal line represents the limit of detection.
-Timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S9) are identified by the asterisk(s) above each timepoint (\*\, P < 0.05).
+Timepoints where differences between sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S15) are identified by the asterisk above each timepoint (\*\, P < 0.05).
 
@@
 Each symbol represents a stool sample from an individual mouse with color corresponding to experiment and shape representing cage mates.
-PERMANOVA was performed within each group to examine the contributions of experiment and cage to observed variation.
-Experiment number and cage only significantly explained observed variation for mice from the Schloss (combined R^2^ = `r adonis_schloss_rsq_sum`; *P* $\le$ `r p_adonis_schloss_max`) and Young (combined R^2^ = `r adonis_young_rsq_sum`; *P* $\le$ `r p_adonis_young_max`) lab colonies (Table S2).
-G-H: Boxplots of the $\theta_{YC}$ distances of the 6 sources of mice relative to mice within the same source and experiment (G) or mice within the same source and between experiments (H) at baseline.
+Experiment number and cage effects explained most of the observed variation for samples from the Schloss (PERMANOVA combined R^2^ = `r adonis_schloss_rsq_sum`; *P* $\le$ `r p_adonis_schloss_max`) and Young (combined R^2^ = `r adonis_young_rsq_sum`; *P* $\le$ `r p_adonis_young_max`) mice (Table S4).
+G-H: Boxplots of the $\theta_{YC}$ distances of the 6 sources of mice relative to mice within the same source and experiment (G) or mice within the same source and between experiments (H) at baseline (day -1).
 Symbols represent individual mouse samples: circles for experiment 1 and triangles for experiment 2.
@@
 *C.
-difficile* CFU/gram stool on day 7 post-infection across sources of mice with asterisks for pairwise Wilcoxon comparisons with Benjamini-Hochberg correction where *P* < 0.05.
+difficile* CFU/gram stool 7-days post-infection across sources of mice with an asterisk for pairwise Wilcoxon comparisons with Benjamini-Hochberg correction where *P* < 0.05.
 D.
-Mouse weight change 2 days post-infection across sources of mice, no pairwise Wilcoxon comparisons were significant after Benjamini-Hochberg correction.
+Mouse weight change 2-days post-infection across sources of mice, no pairwise Wilcoxon comparisons were significant after Benjamini-Hochberg correction.
 For C-D: circles represent experiment 1 mice, triangles represent experiment 2 mice and gray lines indicate the median values for each group.
@@
 difficile* colonization status 7 days post-challenge.** A.
-Bar graph visualizations of overall day 7 C.
-difficile colonization status that were used as classification outcomes to build L2-regularized logistic regression models.
+Bar graph visualizations of overall 7-days post-infection *C.
+difficile* colonization status that were used as classification outcomes to build L2-regularized logistic regression models.
 Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU) based on CFU g/stool data from 7 days post-infection.
@@
 L2-regularized logistic regression classification model area under the receiving operator characteristic curve (AUROCs) to predict *C.
-difficile* CFU on day 7 post-infectoin (Fig.
-1D, Fig.
-S3) based on the OTU community relative abundances at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1).
-All models performed better than random chance (AUROC = 0.5; all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S7) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for all pairwise comparisons; Table S7).
-For list of the 20 OTUs that were ranked as most important to each model, see Table S8.
+difficile* CFU on day 7 post-infection (Fig.
+2B, Fig.
+S2C) based on the OTU community relative abundances at baseline (day -1), post-clindamycin (day 0), and 1-day post-infection.
+All models performed better than random chance (AUROC = 0.5, all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`, Table S12) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for all pairwise comparisons, Table S13).
+For list of the 20 OTUs that were ranked as most important to each model, see Table S14.
 
@@
 OTUs from classification models based on baseline, post-clindamycin treatment, or post-infection community data vary by source, clindamycin treatment, or both.** A-C.
-Venn diagrams of top 20 important OTUs from baseline (A), post-clindamycin treatment (B), and post-infection (C) classification models (Table S8) that overlapped with OTUs that varied across sources at the corresponding timepoint, were impacted by clindamycin treatment, or both.
-Bold OTUs signify OTUs that were important to more than 1 classification model.
+Venn diagrams of OTUs from the top 20 OTUs from the baseline (A), post-clindamycin treatment (B), and post-infection (C) classification models (Table S14) that overlapped with OTUs that varied across sources at the corresponding timepoint (Tables S5, 8, 10), were impacted by clindamycin treatment (Table S9), or both.
+Red OTUs signify OTUs that were important to more than 1 classification model.
 
@@
 ## Supplementary Tables and Movie
-All supplemental material is available at: https://github.com/SchlossLab/Tomkovich_Vendor_XXXX_2020/submission.
 
@@
 Large shifts in bacterial community structure occurred after clindamycin and *C.
-difficile* infection.** PCoA of $\theta_{YC}$ distances animated from 0 through 9 days post-infection.
-PERMANOVA analysis indicated source was the variable that explained the most observed variation across fecal communities (source R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment.
-Transparency of the circle corresponds to the day of the experiment, each circle represents a sample from an individual mouse at a specific timepoint.
-See Table S2 for PERMANOVA results).
+difficile* infection.** PCoA of $\theta_{YC}$ distances animated from days -1 through 9 of the experiment.
+Source was the variable that explained the most observed variation across fecal communities (PERMANOVA source R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`, Table S11) followed by interactions between cage and day of the experiment.
+Transparency of the symbol corresponds to the day of the experiment, each symbol represents a sample from an individual mouse at a specific timepoint.
 Circles represent mice from experiment 1 and triangles represent mice from expeirment 2.
 
-**Table S1. Diversity metrics Kruskal-Wallis and pairwise Wilcoxon statistical results.**
+**Tables S1-S15. Excel workbook of Tables S1-S15.**
 
-**Table S2. PERMANOVA results for mice at specific timepoints and across all timepoints.**
+**Table S1. Alpha diversity metrics Kruskal-Wallis statistical results.**
 
-**Table S3. OTUs with relative abudances that significantly vary across sources at baseline, post-clindamycin, or post-infection timepoints.**
+**Table S2. Alpha diversity metrics pairwise Wilcoxon statistical results.**
 
-**Table S4. *C. difficile* CFU statistical results.**
+**Table S3. PERMANOVA results for mice at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1).**
 
-**Table S5. Mouse weight change statistical results.**
+**Table S4. PERMANOVA results for each source of mice at baseline (day -1).**
 
-**Table S6. OTUs with relative abudances that significantly changed after clindamycin treatment.**
+**Table S5. OTUs with relative abudances that significantly vary between sources at baseline (day -1).**
 
-**Table S7. Statistical results of L2-regularized logistic regression model performances.**
+**Table S6. *C. difficile* CFU statistical results.**
 
-**Table S8. Top 20 most important OTUs for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
+**Table S7. Mouse weight change statistical results.**
 
-**Table S9. OTUs with relative abudances that significantly varied across sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**
+**Table S8. OTUs with relative abudances that significantly vary between sources post-clindamycin (day 0).**
+
+**Table S9. OTUs with relative abudances that significantly changed after clindamycin treatment.**
+
+**Table S10. OTUs with relative abudances that significantly vary between sources post-infection (day 1).**
+
+**Table S11. PERMANOVA results for mice across all timepoints.**
+
+**Table S12. Statistical results of L2-regularized logistic regression model performances compared to random chance.**
+
+**Table S13. Pairwise comparisons of L2-regularized logistic regression model performances.**
+
+**Table S14. Top 20 most important OTUs for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
+
+**Table S15. OTUs with relative abundances that significantly varied across sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**

### 2020-08-03 submission/manuscript.tex
Incorporate Pat's and my own edits

@@
-\documentclass[11pt,]{article}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+  11pt,
+]{article}
 \usepackage{lmodern}
@@
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
   \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1.0in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
 \usepackage{graphicx,grffile}
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{}
-    \pretitle{\vspace{\droptitle}}
-  \posttitle{}
-    \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
-
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{helvet} % Helvetica font
@@
 
+\author{}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 
+\hypertarget{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}{%
 \section{\texorpdfstring{The initial gut microbiota and response to
@@
 colonization in
-mice}{The initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice}}\label{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}
+mice}{The initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice}}\label{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}}
 
@@
 \newpage
-
 \linenumbers
 
-\subsection{Abstract}\label{abstract}
+\hypertarget{abstract}{%
+\subsection{Abstract}\label{abstract}}
 
@@
 day later and then measured \emph{C. difficile} colonization levels
-through 9 days post-infection. The microbiota was profiled via 16S rRNA
+through 9 days post-infection. The microbiota were profiled via 16S rRNA
 gene sequencing to examine the variation across sources and alterations
@@
 \emph{C. difficile} by 7 days post-infection using baseline,
-post-clindamycin, and post-infection community composition data. With
-these models, we identified 4 bacteria that varied across sources
+post-clindamycin, and 1-day post-infection community composition data.
+With these models, we identified 4 bacteria that varied across sources
 (\emph{Bacteroides}), were altered by clindamycin
@@
 \emph{Enterococcus}). Microbiota variation across sources better
-emulates human interindividual variation and can help identify bacterial
-drivers of phenotypic variation in the context of CDIs.
-
-\subsection{Importance}\label{importance}
+emulates human inter-individual variation and can help identify
+bacterial drivers of phenotypic variation in the context of CDIs.
+
+\hypertarget{importance}{%
+\subsection{Importance}\label{importance}}
 
 \emph{Clostridioides difficile} is a leading nosocomial infection.
-Although perturbation to the gut microbiota has been established as a
-key risk factor, there is variation in who becomes asymptomatically
-colonized, develops an infection, or has an infection with adverse
-outcomes. \emph{C. difficile} infection (CDI) mouse models are widely
-used to answer a variety of \emph{C. difficile} pathogenesis questions.
-However, the inter-individual variation between mice is less than what
-is observed in humans, particularly if just one source of mice is used.
-In this study, we administered clindamycin to mice from 6 different
-breeding colonies and challenged them with \emph{C. difficile}.
-Interestingly, only a subset of the bacteria that vary across sources
-were associated with how long \emph{C. difficile} was able to colonize.
-Future studies examining the interplay between the microbiota and
-\emph{C. difficile} should consider using mice from multiple sources to
-better reflect human interindividual variation.
-
-\newpage
-
-\subsection{Introduction}\label{introduction}
+Although perturbation to the gut microbiota is an established risk,
+there is variation in who becomes asymptomatically colonized, develops
+an infection, or has an infection with adverse outcomes. \emph{C.
+difficile} infection (CDI) mouse models are widely used to answer a
+variety of \emph{C. difficile} pathogenesis questions. However, the
+inter-individual variation between mice from the same breeding facility
+is less than what is observed in humans. Therefore, we challenged mice
+from 6 different breeding colonies with \emph{C. difficile}. We found
+that the starting microbial community structures and \emph{C. difficile}
+persistence varied by the source of mice. Interestingly, a subset of the
+bacteria that varied across sources were associated with how long
+\emph{C. difficile} was able to colonize. By increasing the
+inter-individual diversity of the starting communities, we were able to
+better model human diversity. This provided a more nuanced perspective
+of \emph{C. difficile} pathogenesis.
+
+\newpage
+
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
 
 Antibiotics are a common risk factor for \emph{Clostridioides difficile}
-infections (CDIs), but there is variation in who goes on to develop
-severe or recurrent CDIs after exposure (1, 2). Additionally,
-asymptomatic colonization, where \emph{C. difficile} is detectable, but
-symptoms are absent has been documented in infants and adults (3, 4).
-The intestinal microbiome has been implicated in asymptomatic
-colonization (5, 6), susceptibility to CDIs (7), and adverse CDI
-outcomes (9--12).
+infections (CDIs) due to there effect on the intestinal microbiota, but
+there is variation in who goes on to develop severe or recurrent CDIs
+after exposure (1, 2). Additionally, asymptomatic colonization, where
+\emph{C. difficile} is detectable, but symptoms are absent has been
+documented in infants and adults (3, 4). The intestinal microbiota has
+been implicated in asymptomatic colonization (5, 6), susceptibility to
+CDIs (7), and adverse CDI outcomes (9--12). However, it is not clear how
+much inter-individual microbiota variation contributes to the range of
+outcomes observed after \emph{C. difficile} exposure relative to other
+risk factors.
 
@@
 difficile} pathogenesis (13). The number of CDI mouse model studies has
-grown substantially since Chen et al. published their C57BL/6 model in
+grown substantially since Chen et al.~published their C57BL/6 model in
 2008, which disrupted the gut microbiota with antibiotics to enable
@@
 therapeutics for treating CDIs (15). However, variation in the
-microbiome between mice from the same breeding colony is much less than
-the variation observed between humans (16, 17). Additionally, studying
-the contribution of the microbiota to a particular disease phenotype in
-one set of lab mice after the same perturbation could yield a number of
-findings of which only a fraction may be driving the phenotype.
-
-In the past, our group has attempted to introduce more microbiome
-variation into the CDI mouse model by using a variety of antibiotic
-treatments (18--21). An alternative approach to maximize microbiome
-variation is to use mice from multiple sources (22, 23). Microbiome
-differences between different mouse vendors have been well documented
-and shown to influence susceptibility to a variety of diseases (24, 25),
-including enteric infections (22, 23, 26--30). Additionally, research
-groups have observed different CDI outcomes in mice despite using
-similar models and the microbiome has been proposed as one factor
-potentially mediating CDI susceptibility and outcomes (13, 18, 21,
-31--33). Here we examined how variations in the baseline microbiome and
-responses to clindamycin treatment in C57BL/6 mice from six different
-sources influenced susceptibility to \emph{C. difficile} colonization
-and the time needed to clear the infection.
-
-\subsection{Results}\label{results}
-
-\textbf{Microbiota variation is high between mice from different
-sources.} To test how the microbiotas of mice from different sources
-impact colonization dynamics after clindamycin exposure, we utilized
-C57BL/6 mice from 6 different sources: two colonies from the University
-of Michigan that were split from each other 10 years ago (the Young and
-Schloss lab colonies) and four commercial source: the Jackson
-Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo
-(which was formerly Harlan). These 4 vendors were chosen because they
-represent commonly used vendors for CDI studies in mice (26, 34--40). We
-used 16S rRNA gene sequencing to characterize the fecal bacterial
-communities across sources at baseline.
-
-Since antibiotics and other risk factors of CDIs are associated with
-decreased microbiota diversity (41), we first examined alpha diversity
-across the 6 sources of mice. Intially, there was a significant
-difference in the number of observed operational taxonomic units (OTUs),
-but not Shannon diversity index (\emph{P}\textsubscript{FDR} = 0.03 and
-\emph{P}\textsubscript{FDR} = 0.052, respectively) across sources of
-mice (Fig. 1A-B and Table S1). Next, we compared the community structure
-of mice from the 6 sources over the course of the experiment using
-principal coordinate analysis (PCoA) of the \(\theta_{YC}\) distances.
-Permutational multivariate analysis of variance (PERMANOVA) analysis
-revealed source and cage interactions explained most of the observed
-variation across fecal communities explained most of the observed
-community variation (combined R\textsuperscript{2} = 0.90, \emph{P} =
-0.0001, Fig. 3 and Table S2). Mice that are co-housed have been shown to
-have similar gut microbiotas due to coprophagy (42) and since mice
+microbiota between mice from the same breeding colony is much less than
+the inter-individual variation observed between humans (16, 17).
+Studying CDIs in mice with a homogenous microbiota is likely to
+overstate the importance of individual mechanisms. Using mice that have
+a more heterogenous microbiota would allow researchers to identify and
+validate more generalizable mechanisms responsible for CDI.
+
+In the past, our group has attempted to introduce more variation into
+the mouse microbiota by using a variety of antibiotic treatments
+(18--21). An alternative approach to maximize microbiota variation is to
+use mice from multiple sources (22, 23). The differences between the
+microbiota of mice from vendors have been well documented and shown to
+influence susceptibility to a variety of diseases (24, 25), including
+enteric infections (22, 23, 26--30). Different research groups have also
+observed variation in CDI outcomes despite using similar murine models
+(13, 18, 21, 31--33). Here we examined how variations in the baseline
+microbiota and responses to clindamycin treatment in C57BL/6 mice from
+six different sources influenced susceptibility to \emph{C. difficile}
+colonization and the time needed to clear the infection.
+
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
+
+\textbf{The variation in the microbiota is high between mice from
+different sources.} We obtained C57BL/6 mice from 6 different sources:
+two colonies from the University of Michigan that were split from each
+other in 2010 (the Young and Schloss lab colonies) and four commercial
+sources: the Jackson Laboratory, Charles River Laboratories, Taconic
+Biosciences, and Envigo (which was formerly Harlan). These 4 vendors
+were chosen because they are commonly used for murine CDI studies (26,
+34--40). Two experiments were conducted, approximately 3 months apart.
+
+We sequenced the 16S rRNA gene from fecal samples collected from these
+mice after they acclimated to the University of Michigan animal housing
+environment. We first examined the alpha diversity across the 6 sources
+of mice. There was a significant difference in the richness (i.e.~number
+of observed operational taxonomic units (OTUs)), but not Shannon
+diversity index (\emph{P}\textsubscript{FDR} = 0.03 and
+\emph{P}\textsubscript{FDR} = 0.052, respectively) across the sources of
+mice (Fig. 1A-B and Tables S1-2). Next, we compared the community
+structure of mice (Fig. 1C). Interactions between the source and cage
+effects explained most of the observed variation between fecal
+communities (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
+\textless{} 0.001, Fig. 1C and Table S3). Mice that are co-housed tend
+to have similar gut microbiotas due to coprophagy (41). Since mice
 within the same source were housed together, it is not surprising that
-cage also contributed to the observed microbiota variation. Since we
-conducted two experiments approximately 3 months apart, we also compared
-baseline communities across the 2 experiments. Experiment and cage
-significantly explained the observed variation only for the Schloss and
-Young lab mouse colonies (Fig. S2A-B and Table S2). However, most of the
-vendors also clustered by experiment (Fig. S2C-D, F), suggesting there
-was some community variation between the 2 experiments within each
-source, particularly for Schloss, Young, and Envigo mice (Fig. S2G-H).
-After finding differences at the community level, we next identified the
-bacteria that initially varied across sources of mice. There were 268
-OTUs with relative abundances that varied across sources (Fig. 1D and
-Table S3). By using mice from six sources we were able to increase the
-number of microbiota communities to test with the clindamycin \emph{C.
-difficile} colonization mouse model.
+the cage effect also contributed to the observed microbiota variation.
+There were some differences between the 2 experiments we conducted, as
+the experiment and cage effects significantly explained the observed
+community variation for the Schloss and Young lab mouse colonies (Fig.
+S2A-B and Table S4). However, most of the vendors also clustered by
+experiment (Fig. S2C-D, F), suggesting there was some community
+variation between the 2 experiments within each source, particularly for
+Schloss, Young, and Envigo mice (Fig. S2G-H). After finding differences
+at the community level, we next identified the bacteria that varied
+between sources of mice. There were 268 OTUs with relative abundances
+that were significantly different between the sources (Fig. 1D and Table
+S5). Though we saw differences between experiments at the community
+level, there were no OTUs that were significantly different between
+experiments within Schloss, Young, and Envigo mice (Data not shown). By
+using mice from six sources we were able to increase the number of
+microbiota communities to treat with clindamycin and challenge with
+\emph{C. difficile}.
 
@@
 difficile} 630 colonization, but clearance time varies across sources.}
-After the acclimation period, all mice were treated with 10 mg/kg
-clindamycin via intraperitoneal injection and one day later challenged
-with 10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 2A).
-Clindamycin was chosen because we have previously demonstrated mice are
-rendered susceptible, but consistently cleared \emph{C. difficile}
-within 9 days (21, 43), clindamycin is frequently implicated with human
-CDIs (44), and is also part of the antibiotic treatment for the
-frequently cited 2008 CDI mouse model (14). The day after infection,
-\emph{C. difficile} was detectable in all mice at a similar level
-(median CFU range: 2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15),
-indicating clindamycin rendered all mice susceptible regardless of
-source (Fig. 2B). Interestingly, variation in \emph{C. difficile} CFU
-levels across sources of mice emerged from days 3-7 post-infection (all
-\emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 2B and Table S4),
-suggesting mouse source is associated with \emph{C. difficile}
-clearance. We conducted two experiments and while the colonization
-dynamics were similar across most sources of mice, there was some
-variation between the 2 experiments, particularly for the Schloss and
-Envigo mice (Fig. S2A-B). Although \emph{C. difficile} 630 causes mild
-symptoms in mice comparied to other \emph{C. difficile} strains (45), we
-also saw that weight change significantly varied across sources of mice
-with the most weight lost two days post-infection (Fig. 2C and Table
-S5). Importantly, there was also one Jackson and one Envigo mouse that
+Clindamycin is frequently implicated with human CDIs (42) and was part
+of the antibiotic treatment for the frequently cited 2008 CDI mouse
+model (14). We have previously demonstrated mice are rendered
+susceptible to \emph{C. difficile}, but cleared the pathogen within 9
+days when treated with clindamycin alone (21, 43). All mice were treated
+with 10 mg/kg clindamycin via intraperitoneal injection and one day
+later challenged with 10\textsuperscript{3} \emph{C. difficile} 630
+spores (Fig. 2A). The day after infection, \emph{C. difficile} was
+detectable in all mice at a similar level (median CFU range:
+2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15), indicating
+clindamycin rendered all mice susceptible regardless of source (Fig.
+2B). However, between 3 and 7 days post-infection, we observed variation
+in \emph{C. difficile} levels across sources of mice (all
+\emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 2B and Table S6). This
+suggested the mouse source was associated with \emph{C. difficile}
+clearance. While the colonization dynamics were similar between the two
+experiments, the Schloss mice took longer to clear in the 1st experiment
+and the Envigo mice took longer to clear in the 2nd experiment (Fig.
+S2A-B). The change in the mice's weight significantly varied across
+sources of mice with the most weight lost two days post-infection (Fig.
+2C and Table S7). There was also one Jackson and one Envigo mouse that
 died between 1- and 3-days post-infection during the second experiment.
-Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to
-lose more weight, have higher \emph{C. difficile} CFU levels and take
-longer to clear the infection compared to the other sources of mice
-(although there was variation between experiments with Schloss and
-Envigo mice), which was particularly evident 7 days post-infection (Fig.
-2B-C, Fig. S2C-D), when 57\% of the mice were still colonized with
-\emph{C. difficile} (Fig. S2E). By 9 days post-infection the majority of
-the mice from all sources had cleared \emph{C. difficile} (Fig. 2C) with
-the exception of 1 Taconic mouse from the first experiment and 2 Envigo
-mice from the second experiment. Thus, clindamycin rendered all mice
+Mice obtained from Jackson, Taconic, and Envigo tended to lose more
+weight, have higher \emph{C. difficile} CFU levels and take longer to
+clear the infection compared to the other sources of mice (although
+there was variation between experiments with Schloss and Envigo mice).
+This was particularly evident 7 days post-infection (Fig. 2B-C, Fig.
+S2C-D), when 57\% of the mice were still colonized with \emph{C.
+difficile} (Fig. S2E). By 9 days post-infection the majority of the mice
+from all sources had cleared \emph{C. difficile} (Fig. 2C) with the
+exception of 1 Taconic mouse from the first experiment and 2 Envigo mice
+from the second experiment. Thus, clindamycin rendered all mice
 susceptible to \emph{C. difficile} 630 colonization, regardless of
@@
 \textbf{Clindamycin treatment alters bacteria in all sources, but a
-subset of bacterial differences across sources persists.} Given the well
-known variation in mouse microbiomes across breeding colonies (25), we
-hypothesized that variation in \emph{C. difficile} clearance would be
-explained by microbiota variation across the 6 sources of mice. We used
-16S rRNA sequencing to characterize microbiota variation across sources
-after clindamycin treatment and throughout the rest of the experiment.
-PERMANOVA analysis revealed source was the dominant factor that
-explained the observed variation across fecal communities
-(R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
-interactions between cage (mice from the same sources were housed
-together, primarily at a density of 2 mice per cage) and day of the
-experiment (Movie S1 and Table S2). Since the majority of the
-perturbations happened over the initial days of the experiment, we
-decided to focus on the communities after clindamycin treatment (day 0)
-and post-infection (day 1).
-
-As expected, clindamycin treatment decreased richness and Shannon
-diversity across all sources of mice (Fig. 3A-B). Interestingly,
-significant differences in diversity metrics across sources
-(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after clindamycin
-treatment, with Charles River mice having higher richness and Shannon
-diversity than most of the other sources (Fig 3A-B and Table S1). The
-variation in community structures was also diminished by clindamycin
-treatment (Fig. 3C and Table S2). Next we identified the bacteria that
-shifted after clindamycin treatment, regardless of source by analyzing
-all mice that had sequence data from fecal samples collected at baseline
-and after clindamycin treatment. We identified 153 OTUs that were
-altered after clindamycin treatment (Fig. 3D and Table S6) in most mice.
-Despite the significant impact of clindamycin, there were still 18 OTUs
-(Fig. 3E and Table S3) with relative abundances that significantly
-varied across sources. Interestingly, when we compared the list of
+subset of bacterial differences across sources persists.} Given the
+variation in mouse microbiotas that we observed across breeding
+colonies, we hypothesized that variation in \emph{C. difficile}
+clearance would be explained by microbiota variation across the 6
+sources of mice. As expected, clindamycin treatment decreased the
+richness and Shannon diversity across all sources of mice (Fig. 3A-B).
+Interestingly, significant differences in diversity metrics between
+sources (\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after
+clindamycin treatment, with Charles River mice having higher richness
+and Shannon diversity than most of the other sources (Fig 3A-B and
+Tables S1-2). The clindamycin treatment decreased the variation in
+community structures between sources of mice. Source and cage effects
+explained almost all of the observed variation between communities
+(combined R\textsuperscript{2} = 0.99, \emph{P} \textless{} 0.001; Fig.
+3C and Table S3). However, there were only 18 OTUs (Fig. 3D and Table
+S8) with relative abundances that significantly varied between sources.
+Next we identified the bacteria that shifted after clindamycin
+treatment, regardless of source by analyzing paired fecal samples from
+mice that were collected at baseline and after clindamycin treatment. We
+identified 153 OTUs that were altered after clindamycin treatment in
+most mice (Fig. 3E and Table S9). When we compared the list of
 significant clindamycin impacted bacteria with the bacteria that varied
-across sources post-clindamycin, we found 4 OTUs
+between sources post-clindamycin, we found 4 OTUs
 (\emph{Enterobacteriaceae (OTU 1)}, \emph{Lachnospiraceae} (OTU 130),
 \emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) overlapped
-(Fig. 3D-E and Tables S3, S6). These findings demonstrate that
-clindamycin has a consistent impact on the fecal bacterial communities
-of mice from all sources and only a subset of the OTUs continued to vary
-across sources.
-
-\textbf{Microbiota variation across sources is maintained after \emph{C.
-difficile} challenge.} Post-infection, significant differences in
-diversity metrics were maintained across sources
-(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Table S1).
-Although the Charles River mice had more diverse microbiotas, the Young
-and Schloss lab mice were also able to clear \emph{C. difficile} faster
-than some of the other sources, suggesting microbiota diversity alone
-does not explain the observed variation in \emph{C. difficile}
-colonization across sources. Source and cage interaction continued to
-explain most of the observed community variation (combined
-R\textsuperscript{2} = 0.88, respectively; \emph{P} = 0.0001; Fig. 4D
-and Table S2). After \emph{C. difficile} challenge, there were 44 OTUs
-(Fig. 4D and Table S3) with significantly different relative abundances
-across sources. In spite of the experimental perturbations that
-occurred, there were 12 OTUs with relative abundances that consistently
-varied across sources at baseline, post-clindamycin, and post-infection.
-Importantly, some of the OTUs that consistently varied across sources
-also shifted with clindamycin treatment. For example, \emph{Proteus}
-increased after clindamycin treatment, but only in Taconic mice.
-\emph{Enterococcus} was primarily found only in mice purchased from
-commercial vendors and also increased after clindamycin treatment. In
-summary, mouse bacterial communities varied significantly according to
-source throughout the course of the experiment and a consistent subset
-of bacterial taxa remained different across sources regardless of
-clindamycin and \emph{C. difficile} challenge.
+(Fig. 3D-E and Tables S8-9). Importantly, some of the OTUs that varied
+between sources also shifted with clindamycin treatment. For example,
+\emph{Proteus} increased after clindamycin treatment (Fig. 3D), but only
+in Taconic mice. \emph{Enterococcus} was primarily found only in mice
+purchased from commercial vendors and also increased after clindamycin
+treatment (Fig. 3D). These findings demonstrate that clindamycin had a
+consistent impact on the fecal bacterial communities of mice from all
+sources and only a subset of the OTUs continued to vary between sources.
+
+\textbf{Microbiota variation between sources is maintained after
+\emph{C. difficile} challenge.} One day post-infection, significant
+differences in diversity metrics remained across sources
+(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Tables
+S1-2). Although the Charles River mice had more diverse microbiotas and
+were also able to clear \emph{C. difficile} faster than some of the
+other sources. Microbiota diversity did not explain the observed
+variation in \emph{C. difficile} colonization across sources considering
+the Young and Schloss mice had the lowest diversity after clindamycin
+treatment and were able to clear \emph{C. difficile} earlier than
+Jackson, Taconic and Envigo mice. Source and the interactions between
+source and cage effects continued to explain most of the observed
+community variation (combined R\textsuperscript{2} = 0.88; \emph{P}
+\textless{} 0.001; Fig. 4D and Table S3). One day after \emph{C.
+difficile} challenge, there were 44 OTUs (Fig. 4D and Table S10) with
+significantly different relative abundances across sources.
+
+Throughout the experiment the source of mice continued to be the
+dominant factor that explained the observed variation across fecal
+communities (PERMANOVA R\textsuperscript{2} = 0.35, \emph{P} \textless{}
+0.001) followed by interactions between cage effects and the day of the
+experiment (Movie S1 and Table S11). Mice fecal samples from the same
+source of mice continued to cluster closely to each other throughout the
+experiment. By 7 days post-infection, when approximately 43\% mice had
+cleared \emph{C. difficile}, most of the mice still had not recovered to
+their baseline community structure (Fig. 4E). Distance from the baseline
+community between sources did not explain the variation in \emph{C.
+difficile} clearance since Schloss and Young mice cleared \emph{C.
+difficile} faster, but their communities were a greater distance from
+baseline 7 days post-infection. In summary, mouse bacterial communities
+varied significantly between sources throughout the course of the
+experiment and a consistent subset of bacterial taxa remained different
+between sources regardless of clindamycin and \emph{C. difficile}
+challenge.
 
@@
 can predict mice that will clear \emph{C. difficile} by 7 days
-post-infection.} After identifying taxa that varied by source, changed
-after clindamycin treatment, or both, we next wanted to determine which
-taxa were influencing the variation in \emph{C. difficile} colonization
-at day 7 (Fig. 2D, Fig. S2C). We trained three L2-regularized logistic
-regression models with input bacterial community data from the baseline,
-post-clindamycin, and post-infection timepoints of the experiment to
-predict \emph{C. difficile} colonization status on day 7 (Fig. S3A-B).
-All models were better at predicting \emph{C. difficile} colonization
-status on day 7 than random chance (all \emph{P} \(\le\) 1.6e-17; Table
-S7). Interestingly, the model based on the post-clindamycin (day 0)
-community OTU data performed significantly better than the other models
-with an area under the receiving operator characteristic curve (AUROC)
-of 0.78 (\emph{P}\textsubscript{FDR} \(\le\) 3.3e-06 for pairwise
-comparisons; Table S7). Thus, we were able to use bacterial relative
-abundance data alone to differentiate mice that had cleared \emph{C.
-difficile} before day 7 from the mice still colonized with \emph{C.
-difficile} at that timepoint. Interestingly, the model built with OTU
-relative abundance data post-clindamycin treatment had the best
-performance, suggesting how the bacterial community responds to
-clindamycin treatment has the greatest influence on subsequent \emph{C.
-difficile} colonization dynamics.
-
-Next, to examine the bacteria that were driving each model's
-performance, we pulled out the top 20 OTUs that had the highest absolute
-feature weights in each of the 6 models (Table S8). First, we looked at
-OTUs from the model with the best performance that was based on the
-post-clindamycin treatment bacterial community data. While most of the
-20 OTUs had low relative abundances on day 0, \emph{Enterobacteriace},
-\emph{Bacteroides} and \emph{Proteus} had high relative abundances in at
-least one source of mice and significantly varied across sources (Fig.
-5A). Next, the top 20 taxa from each model were compared to the list of
-taxa that varied across source at the same timepoint (Fig. 1D, 3E, 4D
-and Table S3) and the taxa that were altered by clindamycin treatment
-(Fig. 3D and Table S6). We found a subset of OTUs that were important to
-the model and overlapped with bacteria that varied by either source,
-clindamycin treatment, or both (Fig. S4). Combining the overall results
-for the 3 models identified 14 OTUs associated with source, 21 OTUs
-associated with clindamycin treatment, and 6 OTUs associated with both
-(Fig. 5B). Several OTUs (\emph{Bacteroides (OTU 2), Enterococcus (OTU
-23), Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)}) appeared
-across at least 2 models, so we examined how the relative abundances of
-these OTUs varied over the course of the experiment (Fig. 6). Throughout
-the experiment, there was at least 1 timepoint where relative abundances
-of these OTUs significantly varied across sources (Table S9).
-Interestingly, there were no OTUs that emerged as consistently enriched
-or depleted in mice that were colonized past 7 days post-infection with
-\emph{C. difficile} 630, suggesting multiple bacteria influence the time
-needed to clear the infection. Together, these results suggest the
-initial bacterial communities and their responses to clindamycin
-influence the clearance of \emph{C. difficile}.
-
-\subsection{Discussion}\label{discussion}
+post-infection.} After identifying taxa that varied between sources,
+changed after clindamycin treatment, or both, we next wanted to
+determine which taxa were influencing the variation in \emph{C.
+difficile} colonization at day 7 (Fig. 2B, Fig. S2C). We trained three
+L2-regularized logistic regression models with input bacterial community
+data from the baseline (day = -1), post-clindamycin (day = 0), and
+post-infection (day = 1) timepoints of the experiment to predict
+\emph{C. difficile} colonization status on day 7 (Fig. S3A-B). All
+models were better at predicting \emph{C. difficile} colonization status
+on day 7 than random chance (all \emph{P} \textless{} 0.001, Table S12).
+The model based on the post-clindamycin (day 0) community OTU data
+performed significantly better than the other models with an area under
+the receiving operator characteristic curve (AUROC) of 0.78
+(\emph{P}\textsubscript{FDR} \textless{} 0.001 for pairwise comparisons,
+Table S13). Thus, we were able to use bacterial relative abundance data
+from the time of \emph{C. difficile} challenge to differentiate mice
+that had cleared \emph{C. difficile} before day 7 from the mice still
+colonized with \emph{C. difficile} at that timepoint. This result
+suggests the bacterial community's response to clindamycin treatment had
+the greatest influence on subsequent \emph{C. difficile} colonization
+dynamics.
+
+To examine the bacteria that were driving each model's performance, we
+selected the 20 OTUs that had the highest absolute feature weights in
+each of the 3 models (Table S14). First, we looked at OTUs from the
+model with the best performance, which was based on the post-clindamycin
+treatment (day 0) bacterial community data. Out of the 10 highest ranked
+OTUs, 7 OTUs (\emph{Bacteroides}, \emph{Escherichia/Shigella}, 2
+\emph{Lachnospiraceae}, \emph{Lactobacillus}, \emph{Porphyromonadaceae},
+and \emph{Ruminococcaceae}) were associated with \emph{C. difficile}
+colonization 7 days post-infection, while 3 OTUs
+(\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
+\emph{Porphyromonadaceae}) were associated with clearance (Fig. 5A).
+Next, we examined whether any of the top 20 ranked OTUs from the
+baseline (day 0) model were also important in the other 2 classification
+models based on baseline (day -1) and 1 day post-infection community
+data. We identified 6 OTUs (Enterobacteriaceae, Ruminococcaceae,
+Lactobacillus, Bacteroides, Porphyromonadaceae, Erysipelotrichaceae)
+that were important to the day 0 model and either the baseline or 1 day
+post-infection models (Table S14). Thus, a subset of bacterial OTUs were
+important for determining \emph{C. difficile} colonization dynamics
+across multiple timepoints.
+
+To determine whether the OTUs driving the classification models also
+varied between sources, were altered by clindamycin treatment, or both,
+we examined whether the top 20 ranked OTUs from each model overlapped
+with OTUs that varied between sources (Fig. 1D, 3D, 4D and Tables S5,
+S8, S10) or were impacted by clindamycin treatment (Fig. 3E and Table
+S9). We found a subset of OTUs that were important to the baseline (day
+-1), post-clindamycin (day 0), and 1 day post-infection models and
+overlapped with bacteria that varied between sources, were altered by
+clindamycin treatment, or both (Fig. S4). Combining the overall
+comparison results for the 3 models identified 14 OTUs associated with
+source, 21 OTUs associated with clindamycin treatment, and 6 OTUs
+associated with both (Fig. 5B). Several OTUs (\emph{Bacteroides,
+Enterococcus, Enterobacteriaceae, Porphyromonadaceae}) that overlapped
+with our previous analyses appeared across at least 2 models, so we
+examined how the relative abundances of these OTUs varied over the
+course of the experiment (Fig. 6). Throughout the experiment, there was
+at least 1 timepoint where relative abundances of these OTUs
+significantly varied between sources (Table S15). Interestingly, there
+were no OTUs that emerged as consistently enriched or depleted in mice
+that were colonized past 7 days post-infection, suggesting multiple
+bacteria influence \emph{C. difficile} colonization dynamics. Together,
+these results suggest the initial bacterial communities and their
+responses to clindamycin influence the clearance of \emph{C. difficile}.
+
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
 
@@
 the experiment as well as taxa that were universally impacted by
-clindamycin. We trained L2 logistic regression models with baseline,
-post-clindamycin treatment, and post-infection fecal community data that
-could predict whether mice cleared \emph{C. difficile} by 7 days
-post-infection better than random chance. We identified
-\emph{Bacteroides (OTU 2), Enterococcus (OTU 23), Enterobacteriaceae
-(OTU 1), Porphyromonadaceae (OTU 7)} (Fig. 6) as candidate bacteria
-within these communities that were influencing variation in \emph{C.
-difficile} colonization dynamics since these bacteria were all important
-in the logistic regression models and varied by source, were impacted by
-clindamycin treatment, or both. Overall, our results demonstrate
-clindamycin is sufficient to render mice from multiple sources
+clindamycin. We trained L2 logistic regression models with baseline (day
+-1), post-clindamycin treatment (day 0), and 1-day post-infection fecal
+community data that could predict whether mice cleared \emph{C.
+difficile} by 7 days post-infection better than random chance. We
+identified \emph{Bacteroides, Enterococcus, Enterobacteriaceae,
+Porphyromonadaceae} (Fig. 6) as candidate bacteria within these
+communities that were influencing variation in \emph{C. difficile}
+colonization dynamics since these bacteria were all important in the
+logistic regression models and varied by source, were impacted by
+clindamycin treatment, or both. Overall, our results demonstrated
+clindamycin was sufficient to render mice from multiple sources
 susceptible to CDI and only a subset of the interindividual microbiota
@@
 
-Although all mice were susceptible to \emph{C. difficile} colonization,
-mice from Jackson, Taconic, and Envigo mice tended to remain colonized
-through at least 7 days post-infection. We identified a subset of
-bacteria that were important in predicting whether a mouse was still
-colonized with \emph{C. difficile} 7 days post-infection. These results
-suggest a subset of the bacterial community is responsible for
-determining the length of time needed to clear \emph{C. difficile}
-colonization.
-
-Differences in CDI mouse model studies have been attributed to different
-intestinal microbiotas of mice from different sources. For example,
-groups using the same clindamycin treatment and C57BL/6 mice had
-different \emph{C. difficile} outcomes, one having sustained
-colonization (32), while the other had transient (18), despite both
-using \emph{C. difficile} VPI 10643. Baseline differences in the
-microbiota composition have been hypothesized to partially explain the
-differences in colonization outcomes and overall susceptibility to
-\emph{C. difficile} after treatment with the same antibiotic (13, 31).
-The bacterial perturbations induced by clindamycin treatment have been
-well characterized and our findings agree with previous CDI mouse model
-work demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae}
-were associated with \emph{C. difficile} susceptibility and
+Differences in CDI mouse model studies have been attributed to
+intestinal microbiota variation across sources. For example, groups
+using the same clindamycin treatment and C57BL/6 mice had different
+\emph{C. difficile} outcomes, one having sustained colonization (32),
+while the other had transient (18), despite both using \emph{C.
+difficile} VPI 10643. Baseline differences in the microbiota composition
+have been hypothesized to partially explain the differences in
+colonization outcomes and overall susceptibility to \emph{C. difficile}
+after treatment with the same antibiotic (13, 31). The bacterial
+perturbations induced by clindamycin treatment have been well
+characterized and our findings agree with previous CDI mouse model work
+demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae} were
+associated with \emph{C. difficile} susceptibility and
 \emph{Porpyhromonadaceae}, \emph{Lachnospiraceae},
 \emph{Ruminococcaceae}, and \emph{Turicibacter} were associated with
-resistance (19, 21, 32, 33, 43, 46--48). While we have demonstrated that
+resistance (19, 21, 32, 33, 43--46). While we have demonstrated that
 susceptibility is uniform across sources of mice after clindamycin
@@
 example, a group found differential colonization outcomes after
-clindamycin treatment, with \emph{C. difficile} 630 and M68 eventually
-becoming undectable while BI-7 remained detectable up to 70 days
-post-treatment (47). We found the time needed to naturally clear
-\emph{C. difficile} varied across sources of mice implying that at least
-in the context of the same perturbation, microbiota differences seemed
-to influence infection outcome more than susceptibility. More
+clindamycin treatment, with \emph{C. difficile} 630 and M68 infections
+eventually becoming undectable while strain BI-7 remained detectable up
+to 70 days post-treatment (45). We found the time needed to naturally
+clear \emph{C. difficile} varied across sources of mice implying that at
+least in the context of the same perturbation, microbiota differences
+seemed to influence infection outcome more than susceptibility. More
 importantly, we were able to reduce the variation observed across
@@
 difficile} 630 by 9 days post-infection and the model built with the
-post-clindamycin OTU relative abundance data had the best performance,
-our results suggest clindamycin treatment had a large role in
-determining \emph{C. difficile} susceptibility and clearance than the
+post-clindamycin (day 0) OTU relative abundance data had the best
+performance, our results suggest clindamycin treatment had a larger role
+in determining \emph{C. difficile} susceptibility and clearance than the
 source of the mice.
@@
 communities tested in our clindamycin \emph{C. difficile} model. One
-alternative approach that has been used in some CDI studies (49--54) is
+alternative approach that has been used in some CDI studies (47--52) is
 to associate mice with human microbiotas. However, a major caveat to
 this method is the substantial loss of human microbiota community
-members upon transfer to mice (55, 56). Additionally with the exception
-of 2 recent studies (49, 50), most of the CDI mouse model studies to
+members upon transfer to mice (53, 54). Additionally with the exception
+of 2 recent studies (47, 48), most of the CDI mouse model studies to
 date associated mice with just 1 types of human microbiota either from a
-single donor or a single pool from multiple donors (51--54), which does
+single donor or a single pool from multiple donors (49--52), which does
 not aid in the goal of modeling the interpersonal variation seen in
@@
 study, suggesting most of the mouse sources are suitable for gaining
-insights into microbiota associated factors influencing \emph{C.
-difficile} colonization and infections in humans. An important exception
-was \emph{Enterococcus}, which was primarily absent from the mice from
-University of Michigan colonies and \emph{Proteus}, which was only found
-in Taconic mice. Importantly, the fact that some CDI-associated bacteria
-were only found in a subset of mice has important implications for
-future CDI mouse model studies.
-
-There are several limitations to our work. The microbiome is composed of
+insights into the bacteria influencing \emph{C. difficile} colonization
+and infections in humans. An important exception was
+\emph{Enterococcus}, which was primarily absent from University of
+Michigan colonies and \emph{Proteus}, which was only found in Taconic
+mice. Importantly, the fact that some CDI-associated bacteria were only
+found in a subset of mice has important implications for future CDI
+mouse model studies.
+
+There are several limitations to our work. The microbiota is composed of
 viruses, fungi, and parasites in addition to bacteria, and these
-non-bacterial members can also vary across mouse sources (57, 58). While
-our study focused solely on the bacterial portion, viruses and fungi
-have also begun to be implicated in the context of CDIs or FMT
-treatments for recurrent CDIs (35, 59--62). Beyond community
+non-bacterial members can also vary across sources of mice (55, 56).
+While our study focused solely on the bacterial portion, viruses and
+fungi have also begun to be implicated in the context of CDIs or FMT
+treatments for recurrent CDIs (35, 57--60). Beyond community
 composition, the metabolic function of the microbiota also has a CDI
-signature (20, 48, 63, 64) and can vary across mice from different
-sources (65). For example, microbial metabolites, particularly secondary
+signature (20, 46, 61, 62) and can vary across mice from different
+sources (63). For example, microbial metabolites, particularly secondary
 bile acids and butyrate production, have been implicated as important
-contributors to \emph{C. difficile} resistance (33, 47). Although, we
+contributors to \emph{C. difficile} resistance (33, 45). Although, we
 only looked at composition, \emph{Ruminococcaceae} and
@@
 necessary for secondary bile acid formation and ability to produce
-butyrate (52, 66). Interestingly, butyrate has previously been shown to
-vary across vendors and mediated resistance to \emph{Citrobacter
+butyrate (50, 64). Interestingly, butyrate has previously been shown to
+vary across mouse vendors and mediated resistance to \emph{Citrobacter
 rodentium} infection, a model of enterohemorrhagic and enteropathogenic
@@
 toning differences in IgA and Th17 cells across mice from different
-vendors have also been documented and (67, 68) may also influence
-response to CDI, particularly in the context of severe CDIs (69, 70).
-The outcome after \emph{C. difficile} exposure depends on a multitude of
-factors, including age, diet, and immunity; all of which are also
-influenced by the microbiota.
+vendors have also been documented and (65, 66) could influence the host
+response to CDI (67, 68). The outcome after \emph{C. difficile} exposure
+depends on a multitude of factors, including age, diet, and immunity;
+all of which are also influenced by the microbiota.
 
@@
 time mice remained colonized with \emph{C. difficile} 630. For those
-interested in dissecting the contribution of the microbiome to \emph{C.
+interested in dissecting the contribution of the microbiota to \emph{C.
 difficile} pathogenesis and treatments, using multiple sources of mice
 may yield more insights than a single model alone. Furthermore, for
-studies wanting to examine the interplay between a particular bacterial
-taxon such as \emph{Enterococcus} and \emph{C. difficile}, these results
-could serve as a resource for selecting which mice to order to address
-the question. Using mice from multiple sources helps model the
-interpersonal microbiota variation among humans to aid our understanding
-of how the gut microbiota contributes to CDIs.
-
-\newpage
-
-\subsection{Acknowledgements}\label{acknowledgements}
+studies wanting to examine the interplay between a particular bactera
+such as \emph{Enterococcus} and \emph{C. difficile}, these results could
+serve as a resource for selecting which mice to order to address the
+question. Using mice from multiple sources helps model the interpersonal
+microbiota variation among humans to aid our understanding of how the
+gut microbiota contributes to CDIs.
+
+\newpage
+
+\hypertarget{acknowledgements}{%
+\subsection{Acknowledgements}\label{acknowledgements}}
 
@@
 planning the experiments and data presentation, as well as code
-tutorials and feedback through Code club. In particular, we want to
+tutorials and feedback through Code Club. In particular, we want to
 thank Begüm Topçuoğlu for help with implementing L2 logistic regression
@@
 
-\subsection{Materials and Methods}\label{materials-and-methods}
+\hypertarget{materials-and-methods}{%
+\subsection{Materials and Methods}\label{materials-and-methods}}
 
@@
 established with mice purchased from Jackson, and the Schloss lab colony
-was established 10 years ago with mice donated from the Young lab. The 4
+was established in 2010 with mice donated from the Young lab. The 4
 groups of mice purchased from vendors were allowed to acclimate to the
@@
 Fisher Scientific) using custom barcoded primers, as previously
-described (71). The ZymoBIOMICS microbial community DNA standards was
-used as a mock community control (72) and water was used as a negative
+described (69). The ZymoBIOMICS microbial community DNA standards was
+used as a mock community control (70) and water was used as a negative
 control per 96-well extraction plate. The PCR amplicons were cleaned up
@@
 \textbf{(v) 16S rRNA gene sequence analysis.} mothur (v. 1.43) was used
-to process all sequences (73) with a previously published protocol (71).
-Reads were combined and aligned with the SILVA reference database (74).
+to process all sequences (71) with a previously published protocol (69).
+Reads were combined and aligned with the SILVA reference database (72).
 Chimeras were removed with the VSEARCH algorithm and taxonomic
 assignment was completed with a modified version (v16) of the Ribosomal
-Database Project reference database (v11.5) (75) with an 80\% cutoff.
+Database Project reference database (v11.5) (73) with an 80\% cutoff.
 Operational taxonomic units (OTUs) were assigned with a 97\% similarity
-threshold using the opticlust algorithm (76). To account for uneven
+threshold using the opticlust algorithm (74). To account for uneven
 sequencing across samples, samples were rarefied to 5,437 sequences
@@
 \(\theta_{YC}\) distance matrices with the adonis function in the vegan
-package (77) in R (78).
+package (75) in R (76).
 
@@
 post-infection. The L2-regularized logistic regression models were
-trained and tested using the caret package (79) in R as previously
-described (80) with the exception that we used 60\% training and 40\%
+trained and tested using the caret package (77) in R as previously
+described (78) with the exception that we used 60\% training and 40\%
 testing data splits for the cross-validation of the training data to
@@
 \textbf{(vii) Statistical analysis.} All statistical tests were
-performed in R (v 3.5.2) (78). The Kruskal-Wallis test was used to
+performed in R (v 4.0.2) (76). The Kruskal-Wallis test was used to
 analyze differences in \emph{C. difficile} CFU, mouse weight change, and
@@
 comparisons with Benjamini-Hochberg correction for multiple hypothesis
-testing. The tidyverse package was used to wrangle and graph data (v
-1.3.0) (81).
+testing. The tidyverse package (v 1.3.0) was used to wrangle and graph
+data (79).
 
@@
 
-\subsection{References}\label{references}
+\hypertarget{references}{%
+\subsection{References}\label{references}}
 
 \hypertarget{refs}{}
-\hypertarget{ref-Teng2019}{}
+\leavevmode\hypertarget{ref-Teng2019}{}%
 1. Teng C, Reveles KR, Obodozie-Ofoegbu OO, Frei CR. 2019.
@@
 
-\hypertarget{ref-Kelly2012}{}
-2. Kelly C. 2012. Can we identify patients at high risk of recurrent
+\leavevmode\hypertarget{ref-Kelly2012}{}%
+2. Kelly CP. 2012. Can we identify patients at high risk of recurrent
 \emph{Clostridium difficile} infection? Clinical Microbiology and
@@
 
-\hypertarget{ref-Zacharioudakis2015}{}
+\leavevmode\hypertarget{ref-Zacharioudakis2015}{}%
 3. Zacharioudakis IM, Zervou FN, Pliakos EE, Ziakas PD, Mylonakis E.
-2015. Colonization with toxinogenic \emph{C. difficile} upon hospital
+2015. Colonization with toxinogenic \emph{C. Difficile} upon hospital
 admission, and risk of infection: A systematic review and meta-analysis.
@@
 
-\hypertarget{ref-Crobach2018}{}
+\leavevmode\hypertarget{ref-Crobach2018}{}%
 4. Crobach MJT, Vernon JJ, Loo VG, Kong LY, Péchiné S, Wilcox MH,
@@
 
-\hypertarget{ref-Zhang2015}{}
+\leavevmode\hypertarget{ref-Zhang2015}{}%
 5. Zhang L, Dong D, Jiang C, Li Z, Wang X, Peng Y. 2015. Insight into
 alteration of gut microbiota in \emph{Clostridium difficile} infection
-and asymptomatic c. difficile colonization. Anaerobe 34:1--7.
-
-\hypertarget{ref-VanInsberghe2020}{}
+and asymptomatic c. Difficile colonization. Anaerobe 34:1--7.
+
+\leavevmode\hypertarget{ref-VanInsberghe2020}{}%
 6. VanInsberghe D, Elsherbini JA, Varian B, Poutahidis T, Erdman S, Polz
@@
 
-\hypertarget{ref-Mancabelli2017}{}
+\leavevmode\hypertarget{ref-Mancabelli2017}{}%
 7. Mancabelli L, Milani C, Lugli GA, Turroni F, Cocconi D, Sinderen D
@@
 
-\hypertarget{ref-Duvallet2017}{}
+\leavevmode\hypertarget{ref-Duvallet2017}{}%
 8. Duvallet C, Gibbons SM, Gurry T, Irizarry RA, Alm EJ. 2017.
@@
 
-\hypertarget{ref-Seekatz2016}{}
+\leavevmode\hypertarget{ref-Seekatz2016}{}%
 9. Seekatz AM, Rao K, Santhosh K, Young VB. 2016. Dynamics of the fecal
@@
 
-\hypertarget{ref-Khanna2016}{}
+\leavevmode\hypertarget{ref-Khanna2016}{}%
 10. Khanna S, Montassier E, Schmidt B, Patel R, Knights D, Pardi DS,
@@
 
-\hypertarget{ref-Pakpour2017}{}
+\leavevmode\hypertarget{ref-Pakpour2017}{}%
 11. Pakpour S, Bhanvadia A, Zhu R, Amarnani A, Gibbons SM, Gurry T, Alm
@@
 
-\hypertarget{ref-Lee2020}{}
+\leavevmode\hypertarget{ref-Lee2020}{}%
 12. Lee AA, Rao K, Limsrivilai J, Gillilland M, Malamet B, Briggs E,
@@
 recurrent \emph{Clostridioides difficile} infection in patients with and
-without ulcerative colitis. Inflammatory Bowel Diseases.
-
-\hypertarget{ref-Hutton2014}{}
+without ulcerative colitis. Inflammatory Bowel Diseases
+\url{https://doi.org/10.1093/ibd/izz335}.
+
+\leavevmode\hypertarget{ref-Hutton2014}{}%
 13. Hutton ML, Mackin KE, Chakravorty A, Lyras D. 2014. Small animal
@@
 
-\hypertarget{ref-Chen2008}{}
+\leavevmode\hypertarget{ref-Chen2008}{}%
 14. Chen X, Katchar K, Goldsmith JD, Nanthakumar N, Cheknis A, Gerding
@@
 
-\hypertarget{ref-Best2012}{}
+\leavevmode\hypertarget{ref-Best2012}{}%
 15. Best EL, Freeman J, Wilcox MH. 2012. Models for the study of
@@
 
-\hypertarget{ref-Baxter2014}{}
+\leavevmode\hypertarget{ref-Baxter2014}{}%
 16. Baxter NT, Wan JJ, Schubert AM, Jenior ML, Myers P, Schloss PD.
@@
 
-\hypertarget{ref-Nagpal2018}{}
+\leavevmode\hypertarget{ref-Nagpal2018}{}%
 17. Nagpal R, Wang S, Woods LCS, Seshie O, Chung ST, Shively CA,
@@
 
-\hypertarget{ref-Reeves2011}{}
+\leavevmode\hypertarget{ref-Reeves2011}{}%
 18. Reeves AE, Theriot CM, Bergin IL, Huffnagle GB, Schloss PD, Young
@@
 
-\hypertarget{ref-Schubert2015}{}
+\leavevmode\hypertarget{ref-Schubert2015}{}%
 19. Schubert AM, Sinani H, Schloss PD. 2015. Antibiotic-induced
@@
 
-\hypertarget{ref-Jenior2017}{}
+\leavevmode\hypertarget{ref-Jenior2017}{}%
 20. Jenior ML, Leslie JL, Young VB, Schloss PD. 2017. \emph{Clostridium
@@
 
-\hypertarget{ref-Jenior2018}{}
+\leavevmode\hypertarget{ref-Jenior2018}{}%
 21. Jenior ML, Leslie JL, Young VB, Schloss PD. 2018. \emph{Clostridium
@@
 
-\hypertarget{ref-Velazquez2019}{}
+\leavevmode\hypertarget{ref-Velazquez2019}{}%
 22. Velazquez EM, Nguyen H, Heasley KT, Saechao CH, Gil LM, Rogers AWL,
@@
 
-\hypertarget{ref-Osbelt2020}{}
+\leavevmode\hypertarget{ref-Osbelt2020}{}%
 23. Osbelt L, Thiemann S, Smit N, Lesker TR, Schröter M, Gálvez EJC,
@@
 
-\hypertarget{ref-Stough2016}{}
+\leavevmode\hypertarget{ref-Stough2016}{}%
 24. Stough JMA, Dearth SP, Denny JE, LeCleir GR, Schmidt NW, Campagna
@@
 
-\hypertarget{ref-Alegre2019}{}
+\leavevmode\hypertarget{ref-Alegre2019}{}%
 25. Alegre M-L. 2019. Mouse microbiomes: Overlooked culprits of
@@
 
-\hypertarget{ref-EtienneMesmin2017}{}
+\leavevmode\hypertarget{ref-EtienneMesmin2017}{}%
 26. Etienne-Mesmin L, Chassaing B, Adekunle O, Mattei LM, Bushman FD,
@@
 infect mouse colonies and protect against highly pathogenic \emph{C.
-difficile}. Gut 67:860--871.
-
-\hypertarget{ref-Lai2020}{}
+Difficile}. Gut 67:860--871.
+
+\leavevmode\hypertarget{ref-Lai2020}{}%
 27. Lai NY, Musser MA, Pinho-Ribeiro FA, Baral P, Jacobson A, Ma P,
@@
 
-\hypertarget{ref-Thiemann2017}{}
-28. Thiemann S, Smit N, Roy U, Lesker TR, Gálvez EJ, Helmecke J, Basic
+\leavevmode\hypertarget{ref-Thiemann2017}{}%
+28. Thiemann S, Smit N, Roy U, Lesker TR, Gálvez EJC, Helmecke J, Basic
 M, Bleich A, Goodman AL, Kalinke U, Flavell RA, Erhardt M, Strowig T.
@@
 
-\hypertarget{ref-Rolig2013}{}
+\leavevmode\hypertarget{ref-Rolig2013}{}%
 29. Rolig AS, Cech C, Ahler E, Carter JE, Ottemann KM. 2013. The degree
@@
 
-\hypertarget{ref-Ge2018}{}
+\leavevmode\hypertarget{ref-Ge2018}{}%
 30. Ge Z, Sheh A, Feng Y, Muthupalani S, Ge L, Wang C, Kurnick S,
@@
 
-\hypertarget{ref-Lawley2013}{}
+\leavevmode\hypertarget{ref-Lawley2013}{}%
 31. Lawley TD, Young VB. 2013. Murine models to study \emph{Clostridium
@@
 
-\hypertarget{ref-Buffie2011}{}
+\leavevmode\hypertarget{ref-Buffie2011}{}%
 32. Buffie CG, Jarchum I, Equinda M, Lipuma L, Gobourne A, Viale A,
@@
 
-\hypertarget{ref-Buffie2014}{}
+\leavevmode\hypertarget{ref-Buffie2014}{}%
 33. Buffie CG, Bucci V, Stein RR, McKenney PT, Ling L, Gobourne A, No D,
@@
 
-\hypertarget{ref-Spinler2016}{}
+\leavevmode\hypertarget{ref-Spinler2016}{}%
 34. Spinler JK, Brown A, Ross CL, Boonma P, Conner ME, Savidge TC. 2016.
@@
 
-\hypertarget{ref-Markey2018}{}
+\leavevmode\hypertarget{ref-Markey2018}{}%
 35. Markey L, Shaban L, Green ER, Lemon KP, Mecsas J, Kumamoto CA. 2018.
@@
 
-\hypertarget{ref-McKee2018}{}
+\leavevmode\hypertarget{ref-McKee2018}{}%
 36. McKee RW, Aleksanyan N, Garrett EM, Tamayo R. 2018. Type IV pili
@@
 
-\hypertarget{ref-Yamaguchi2020}{}
+\leavevmode\hypertarget{ref-Yamaguchi2020}{}%
 37. Yamaguchi T, Konishi H, Aoki K, Ishii Y, Chono K, Tateda K. 2020.
@@
 
-\hypertarget{ref-Stroke2018}{}
+\leavevmode\hypertarget{ref-Stroke2018}{}%
 38. Stroke IL, Letourneau JJ, Miller TE, Xu Y, Pechik I, Savoly DR, Ma
@@
 
-\hypertarget{ref-Quigley2019}{}
+\leavevmode\hypertarget{ref-Quigley2019}{}%
 39. Quigley L, Coakley M, Alemayehu D, Rea MC, Casey PG, O'Sullivan,
@@
 
-\hypertarget{ref-Mullish2019}{}
+\leavevmode\hypertarget{ref-Mullish2019}{}%
 40. Mullish BH, McDonald JAK, Pechlivanis A, Allegretti JR, Kao D,
@@
 
-\hypertarget{ref-Ross2016}{}
-41. Ross CL, Spinler JK, Savidge TC. 2016. Structural and functional
-changes within the gut microbiota and susceptibility to
-\emph{Clostridium difficile} infection. Anaerobe 41:37--43.
-
-\hypertarget{ref-Nguyen2015}{}
-42. Nguyen TLA, Vieira-Silva S, Liston A, Raes J. 2015. How informative
+\leavevmode\hypertarget{ref-Nguyen2015}{}%
+41. Nguyen TLA, Vieira-Silva S, Liston A, Raes J. 2015. How informative
 is the mouse for human gut microbiota research? Disease Models \&
@@
 
-\hypertarget{ref-Tomkovich2019}{}
+\leavevmode\hypertarget{ref-Guh2018}{}%
+42. Guh AY, Kutty PK. 2018. \emph{Clostridioides difficile} infection
+169:ITC49.
+
+\leavevmode\hypertarget{ref-Tomkovich2019}{}%
 43. Tomkovich S, Lesniak NA, Li Y, Bishop L, Fitzgerald MJ, Schloss PD.
@@
 
-\hypertarget{ref-Guh2018}{}
-44. Guh AY, Kutty PK. 2018. \emph{Clostridioides difficile} infection
-169:ITC49.
-
-\hypertarget{ref-Theriot2011}{}
-45. Theriot CM, Koumpouras CC, Carlson PE, Bergin II, Aronoff DM, Young
-VB. 2011. Cefoperazone-treated mice as an experimental platform to
-assess differential virulence of \emph{Clostridium difficile} strains.
-Gut Microbes 2:326--334.
-
-\hypertarget{ref-Lawley2009}{}
-46. Lawley TD, Clare S, Walker AW, Goulding D, Stabler RA, Croucher N,
+\leavevmode\hypertarget{ref-Lawley2009}{}%
+44. Lawley TD, Clare S, Walker AW, Goulding D, Stabler RA, Croucher N,
 Mastroeni P, Scott P, Raisen C, Mottram L, Fairweather NF, Wren BW,
@@
 
-\hypertarget{ref-Lawley2012}{}
-47. Lawley TD, Clare S, Walker AW, Stares MD, Connor TR, Raisen C,
+\leavevmode\hypertarget{ref-Lawley2012}{}%
+45. Lawley TD, Clare S, Walker AW, Stares MD, Connor TR, Raisen C,
 Goulding D, Rad R, Schreiber F, Brandt C, Deakin LJ, Pickard DJ, Duncan
@@
 
-\hypertarget{ref-Jump2014}{}
-48. Jump RLP, Polinkovsky A, Hurless K, Sitzlar B, Eckart K, Tomas M,
+\leavevmode\hypertarget{ref-Jump2014}{}%
+46. Jump RLP, Polinkovsky A, Hurless K, Sitzlar B, Eckart K, Tomas M,
 Deshpande A, Nerandzic MM, Donskey CJ. 2014. Metabolomics analysis
@@
 
-\hypertarget{ref-NagaoKitamoto2020}{}
-49. Nagao-Kitamoto H, Leslie JL, Kitamoto S, Jin C, Thomsson KA,
+\leavevmode\hypertarget{ref-NagaoKitamoto2020}{}%
+47. Nagao-Kitamoto H, Leslie JL, Kitamoto S, Jin C, Thomsson KA,
 Gillilland MG, Kuffa P, Goto Y, Jenq RR, Ishii C, Hirayama A, Seekatz
@@
 
-\hypertarget{ref-Battaglioli2018}{}
-50. Battaglioli EJ, Hale VL, Chen J, Jeraldo P, Ruiz-Mojica C, Schmidt
+\leavevmode\hypertarget{ref-Battaglioli2018}{}%
+48. Battaglioli EJ, Hale VL, Chen J, Jeraldo P, Ruiz-Mojica C, Schmidt
 BA, Rekdal VM, Till LM, Huq L, Smits SA, Moor WJ, Jones-Hall Y, Smyrk T,
@@
 
-\hypertarget{ref-Robinson2014}{}
-51. Robinson CD, Auchtung JM, Collins J, Britton RA. 2014. Epidemic
+\leavevmode\hypertarget{ref-Robinson2014}{}%
+49. Robinson CD, Auchtung JM, Collins J, Britton RA. 2014. Epidemic
 \emph{Clostridium difficile} strains demonstrate increased competitive
@@
 
-\hypertarget{ref-Collins2015}{}
-52. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. 2015.
+\leavevmode\hypertarget{ref-Collins2015}{}%
+50. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. 2015.
 Humanized microbiota mice as a model of recurrent \emph{Clostridium
@@
 
-\hypertarget{ref-Collins2018}{}
-53. Collins J, Robinson C, Danhof H, Knetsch CW, Leeuwen HC van, Lawley
+\leavevmode\hypertarget{ref-Collins2018}{}%
+51. Collins J, Robinson C, Danhof H, Knetsch CW, Leeuwen HC van, Lawley
 TD, Auchtung JM, Britton RA. 2018. Dietary trehalose enhances virulence
@@
 
-\hypertarget{ref-Hryckowian2018}{}
-54. Hryckowian AJ, Treuren WV, Smits SA, Davis NM, Gardner JO, Bouley
+\leavevmode\hypertarget{ref-Hryckowian2018}{}%
+52. Hryckowian AJ, Treuren WV, Smits SA, Davis NM, Gardner JO, Bouley
 DM, Sonnenburg JL. 2018. Microbiota-accessible carbohydrates suppress
@@
 
-\hypertarget{ref-Fouladi2020}{}
-55. Fouladi F, Glenny EM, Bulik-Sullivan EC, Tsilimigras MCB, Sioda M,
+\leavevmode\hypertarget{ref-Fouladi2020}{}%
+53. Fouladi F, Glenny EM, Bulik-Sullivan EC, Tsilimigras MCB, Sioda M,
 Thomas SA, Wang Y, Djukic Z, Tang Q, Tarantino LM, Bulik CM, Fodor AA,
@@
 microbial taxonomic abundance between humans and mice after gnotobiotic
-transfer. The ISME Journal.
-
-\hypertarget{ref-Walter2020}{}
-56. Walter J, Armet AM, Finlay BB, Shanahan F. 2020. Establishing or
+transfer. The ISME Journal
+\url{https://doi.org/10.1038/s41396-020-0645-z}.
+
+\leavevmode\hypertarget{ref-Walter2020}{}%
+54. Walter J, Armet AM, Finlay BB, Shanahan F. 2020. Establishing or
 exaggerating causality for the gut microbiome: Lessons from human
@@
 
-\hypertarget{ref-Rasmussen2019}{}
-57. Rasmussen TS, Vries L de, Kot W, Hansen LH, Castro-Mejía JL,
+\leavevmode\hypertarget{ref-Rasmussen2019}{}%
+55. Rasmussen TS, Vries L de, Kot W, Hansen LH, Castro-Mejía JL,
 Vogensen FK, Hansen AK, Nielsen DS. 2019. Mouse vendor influence on the
@@
 
-\hypertarget{ref-Mims2020}{}
-58. Mims TS, Abdallah QA, Watts S, White C, Han J, Willis KA, Pierre JF.
+\leavevmode\hypertarget{ref-Mims2020}{}%
+56. Mims TS, Abdallah QA, Watts S, White C, Han J, Willis KA, Pierre JF.
 2020. Variability in interkingdom gut microbiomes between different
@@
 
-\hypertarget{ref-Stewart2019}{}
-59. Stewart DB, Wright JR, Fowler M, McLimans CJ, Tokarev V, Amaniera I,
+\leavevmode\hypertarget{ref-Stewart2019}{}%
+57. Stewart DB, Wright JR, Fowler M, McLimans CJ, Tokarev V, Amaniera I,
 Baker O, Wong H-T, Brabec J, Drucker R, Lamendella R. 2019. Integrated
@@
 
-\hypertarget{ref-Ott2017}{}
-60. Ott SJ, Waetzig GH, Rehman A, Moltzau-Anderson J, Bharti R, Grasis
+\leavevmode\hypertarget{ref-Ott2017}{}%
+58. Ott SJ, Waetzig GH, Rehman A, Moltzau-Anderson J, Bharti R, Grasis
 JA, Cassidy L, Tholey A, Fickenscher H, Seegert D, Rosenstiel P,
@@
 
-\hypertarget{ref-Zuo2017}{}
-61. Zuo T, Wong SH, Lam K, Lui R, Cheung K, Tang W, Ching JYL, Chan PKS,
+\leavevmode\hypertarget{ref-Zuo2017}{}%
+59. Zuo T, Wong SH, Lam K, Lui R, Cheung K, Tang W, Ching JYL, Chan PKS,
 Chan MCW, Wu JCY, Chan FKL, Yu J, Sung JJY, Ng SC. 2017. Bacteriophage
@@
 
-\hypertarget{ref-Zuo2018}{}
-62. Zuo T, Wong SH, Cheung CP, Lam K, Lui R, Cheung K, Zhang F, Tang W,
+\leavevmode\hypertarget{ref-Zuo2018}{}%
+60. Zuo T, Wong SH, Cheung CP, Lam K, Lui R, Cheung K, Zhang F, Tang W,
 Ching JYL, Wu JCY, Chan PKS, Sung JJY, Yu J, Chan FKL, Ng SC. 2018. Gut
@@
 
-\hypertarget{ref-Robinson2019}{}
-63. Robinson JI, Weir WH, Crowley JR, Hink T, Reske KA, Kwon JH, Burnham
+\leavevmode\hypertarget{ref-Robinson2019}{}%
+61. Robinson JI, Weir WH, Crowley JR, Hink T, Reske KA, Kwon JH, Burnham
 C-AD, Dubberke ER, Mucha PJ, Henderson JP. 2019. Metabolomic networks
@@
 
-\hypertarget{ref-Fletcher2018}{}
-64. Fletcher JR, Erwin S, Lanzas C, Theriot CM. 2018. Shifts in the gut
+\leavevmode\hypertarget{ref-Fletcher2018}{}%
+62. Fletcher JR, Erwin S, Lanzas C, Theriot CM. 2018. Shifts in the gut
 metabolome and \emph{Clostridium difficile} transcriptome throughout
@@
 
-\hypertarget{ref-Xiao2015}{}
-65. Xiao L, Feng Q, Liang S, Sonne SB, Xia Z, Qiu X, Li X, Long H, Zhang
+\leavevmode\hypertarget{ref-Xiao2015}{}%
+63. Xiao L, Feng Q, Liang S, Sonne SB, Xia Z, Qiu X, Li X, Long H, Zhang
 J, Zhang D, Liu C, Fang Z, Chou J, Glanville J, Hao Q, Kotowska D,
@@
 
-\hypertarget{ref-Vital2019}{}
-66. Vital M, Rud T, Rath S, Pieper DH, Schlüter D. 2019. Diversity of
+\leavevmode\hypertarget{ref-Vital2019}{}%
+64. Vital M, Rud T, Rath S, Pieper DH, Schlüter D. 2019. Diversity of
 bacteria exhibiting bile acid-inducible 7alpha-dehydroxylation genes in
@@
 
-\hypertarget{ref-Fransen2015}{}
-67. Fransen F, Zagato E, Mazzini E, Fosso B, Manzari C, Aidy SE,
+\leavevmode\hypertarget{ref-Fransen2015}{}%
+65. Fransen F, Zagato E, Mazzini E, Fosso B, Manzari C, Aidy SE,
 Chiavelli A, D'Erchia AM, Sethi MK, Pabst O, Marzano M, Moretti S,
@@
 
-\hypertarget{ref-Ivanov2009}{}
-68. Ivanov II, Atarashi K, Manel N, Brodie EL, Shima T, Karaoz U, Wei D,
+\leavevmode\hypertarget{ref-Ivanov2009}{}%
+66. Ivanov II, Atarashi K, Manel N, Brodie EL, Shima T, Karaoz U, Wei D,
 Goldfarb KC, Santee CA, Lynch SV, Tanoue T, Imaoka A, Itoh K, Takeda K,
@@
 
-\hypertarget{ref-Azrad2018}{}
-69. Azrad M, Hamo Z, Tkhawkho L, Peretz A. 2018. Elevated serum
+\leavevmode\hypertarget{ref-Azrad2018}{}%
+67. Azrad M, Hamo Z, Tkhawkho L, Peretz A. 2018. Elevated serum
 immunoglobulin a levels in patients with \emph{Clostridium difficile}
@@
 
-\hypertarget{ref-Saleh2019}{}
-70. Saleh MM, Frisbee AL, Leslie JL, Buonomo EL, Cowardin CA, Ma JZ,
+\leavevmode\hypertarget{ref-Saleh2019}{}%
+68. Saleh MM, Frisbee AL, Leslie JL, Buonomo EL, Cowardin CA, Ma JZ,
 Simpson ME, Scully KW, Abhyankar MM, Petri WA. 2019. Colitis-induced
@@
 
-\hypertarget{ref-Kozich2013}{}
-71. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
+\leavevmode\hypertarget{ref-Kozich2013}{}%
+69. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 
-\hypertarget{ref-Sze2019}{}
-72. Sze MA, Schloss PD. 2019. The impact of DNA polymerase and number of
+\leavevmode\hypertarget{ref-Sze2019}{}%
+70. Sze MA, Schloss PD. 2019. The impact of DNA polymerase and number of
 rounds of amplification in PCR on 16S rRNA gene sequence data. mSphere
@@
 
-\hypertarget{ref-Schloss2009}{}
-73. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+\leavevmode\hypertarget{ref-Schloss2009}{}%
+71. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
@@
 
-\hypertarget{ref-Quast2012}{}
-74. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
+\leavevmode\hypertarget{ref-Quast2012}{}%
+72. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
 J, Glöckner FO. 2012. The SILVA ribosomal RNA gene database project:
@@
 
-\hypertarget{ref-Cole2013}{}
-75. Cole JR, Wang Q, Fish JA, Chai B, McGarrell DM, Sun Y, Brown CT,
+\leavevmode\hypertarget{ref-Cole2013}{}%
+73. Cole JR, Wang Q, Fish JA, Chai B, McGarrell DM, Sun Y, Brown CT,
 Porras-Alfaro A, Kuske CR, Tiedje JM. 2013. Ribosomal database project:
@@
 
-\hypertarget{ref-Westcott2017}{}
-76. Westcott SL, Schloss PD. 2017. OptiClust, an improved method for
+\leavevmode\hypertarget{ref-Westcott2017}{}%
+74. Westcott SL, Schloss PD. 2017. OptiClust, an improved method for
 assigning amplicon-based sequence data to operational taxonomic units.
@@
 
-\hypertarget{ref-Vegan2018}{}
-77. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+\leavevmode\hypertarget{ref-Vegan2018}{}%
+75. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 Minchin PR, O'Hara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
@@
 
-\hypertarget{ref-r_citation_2018}{}
-78. R Core Team. 2018. R: A language and environment for statistical
+\leavevmode\hypertarget{ref-r_citation_2018}{}%
+76. R Core Team. 2018. R: A language and environment for statistical
 computing. R Foundation for Statistical Computing, Vienna, Austria.
 
-\hypertarget{ref-Kuhn2008}{}
-79. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
+\leavevmode\hypertarget{ref-Kuhn2008}{}%
+77. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
 Journal of Statistical Software 28.
 
-\hypertarget{ref-Topcuoglu2020}{}
-80. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
+\leavevmode\hypertarget{ref-Topcuoglu2020}{}%
+78. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
 framework for effective application of machine learning to
@@
 
-\hypertarget{ref-Tidyverse2019}{}
-81. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
+\leavevmode\hypertarget{ref-Tidyverse2019}{}%
+79. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
 Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
@@
 
-\subsection{Figures}\label{figures}
+\hypertarget{figures}{%
+\subsection{Figures}\label{figures}}
 
@@
 (A) and Shannon diversity index values (B) across sources of mice at
-baseline (day -1) of the experiment. Differences across sources were
+baseline (day -1 of the experiment). Differences between sources were
 analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for
@@
 from pairwise Wilcoxon comparisons between sources were significant
-after Benjamini-Hochberg correction (Table S1). Gray lines represent the
+after Benjamini-Hochberg correction (Table S2). Gray lines represent the
 median values for each source of mice. C. Principal Coordinates Analysis
-(PCoA) of \(\theta_{YC}\) distances of baseline stool samples.
-Permutational multivariate analysis of variance (PERMANOVA) analysis
-demonstrated that source and the interaction between source and cage
-explained most of the variation (combined R\textsuperscript{2} = 0.90,
-\emph{P} = 0.0001, see Table S2). For A-C: each symbol representing the
+(PCoA) of \(\theta_{YC}\) distances of baseline stool samples. Source
+and the interaction between source and cage effects explained most of
+the variation (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
+\textless{} 0.001, see Table S3). For A-C: each symbol represents the
 value for a stool sample from an individual mouse, circles represent
@@
 of the relative abundances for the top 20 bacteria out of the 268 OTUs
-that varied across sources at baseline (Table S3).
+that varied across sources at baseline (Table S5).
 
@@
 difficile} colonization in all mice, but clearance time varies across
-sources of C57BL/6 mice.} A. Setup of the experimental timeline. Mice
-for the experiments were obtained from 6 different sources: the Schloss
-(N = 8) and Young lab (N = 9) colonies at the University of Michigan,
-the Jackson Laboratory (N = 8), Charles River Laboratory (N = 8),
-Taconic Biosciences (N = 8), and Envigo (N = 8). All mice were
-administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before
-challenge with \emph{C. difficile} 630 spores on day 0. Mice were
-weighed and feces was collected daily through the end of the experiment
-(9 days post-infection). Note: 3 mice died during course of experiment.
-1 Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse
+sources.} A. Setup of the experimental timeline. Mice for the
+experiments were obtained from 6 different sources: the Schloss (N = 8)
+and Young lab (N = 9) colonies at the University of Michigan, the
+Jackson Laboratory (N = 8), Charles River Laboratory (N = 8), Taconic
+Biosciences (N = 8), and Envigo (N = 8). All mice were administered 10
+mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with
+\emph{C. difficile} 630 spores on day 0. Mice were weighed and feces was
+collected daily through the end of the experiment (9 days
+post-infection). Note: 3 mice died during course of experiment. 1
+Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse
 between 1- and 3-days post-infection. B. \emph{C. difficile} CFU/gram
@@
 to the weight recorded 1 day before infection. For B-C: timepoints where
-differences across sources of mice were statistically significant by
+differences between sources of mice were statistically significant by
 Kruskal-Wallis test with Benjamini-Hochberg correction for testing
-across multiple days (Table S4 and Table S5) are reflected by the
-asterisk(s) above each timepoint (*, \emph{P} \textless{} 0.05). Lines
+across multiple days (Table S6 and Table S7) are reflected by the
+asterisk above each timepoint (*, \emph{P} \textless{} 0.05). Lines
 represent the median for each source and circles represent individual
@@
 diversity index values (B) across sources of mice after clindamycin
-treatment (day 0). Data were analyzed by Kruskal-Wallis test with
-Benjamini-Hochberg correction for testing each day of the experiment and
-the adjusted \emph{P} value was \textless{} 0.05. Significant \emph{P}
-values from the pairwise Wilcoxon comparisons between sources with
-Benjamini-Hochberg correction are shown (Table S1). C. PCoA of
-\(\theta_{YC}\) distances from stools collected post-clindamycin.
-PERMANOVA analysis demonstrated that source and the interaction between
-source and cage explained most of the variation observed
-post-clindamycin (combined R\textsuperscript{2} = 0.99, \emph{P} =
-0.0001, see Table S2).For A-C, each symbol represents a stool sample
+treatment (day 0). Differences between sources were analyzed by
+Kruskal-Wallis test with Benjamini-Hochberg correction for testing each
+day of the experiment and the adjusted \emph{P} value was \textless{}
+0.05 (Table S1). Significant \emph{P} values from the pairwise Wilcoxon
+comparisons between sources with Benjamini-Hochberg correction are shown
+(Table S2). C. PCoA of \(\theta_{YC}\) distances from stools collected
+post-clindamycin. Source and the interaction between source and cage
+effects explained most of the variation observed post-clindamycin
+(PERMANOVA combined R\textsuperscript{2} = 0.99, \emph{P} \textless{}
+0.001, see Table S3). For A-C, each symbol represents a stool sample
 from an individual mouse, with circles representing experiment 1 mice
 and triangles representing experiment 2 mice. D. Plots highlighting the
-median (point) and interquantile range (colored lines) of the top 10
-OTUs out of 153 with relative abundances that changed after clindamycin
-treatment (adjusted \emph{P} value \textless{} 0.05). Data were analyzed
-by Wilcoxon signed rank test limited to mice that had paired sequence
-data for day -1 and 0 (N = 31), with Benjamini-Hochberg correction for
-testing all identified OTUs (Table S6). The gray vertical line indicates
-the limit of detection. E. Plots highlighting the median (point) and
-interquantile range (colored lines) of the relative abundances for the
-18 OTUs that varied across sources after clindamycin treatment (Table
-S3).
+median (point) and interquantile range (colored lines) of the relative
+abundances for the 18 OTUs (Table S8)that varied between sources after
+clindamycin treatment (day 0). E. Plots highlighting the median (point)
+and interquantile range (colored lines) of the top 10 OTUs out of 153
+with relative abundances that changed after clindamycin treatment
+(adjusted \emph{P} value \textless{} 0.05). Data were analyzed by
+Wilcoxon signed rank test of mice that had paired sequence data for
+baseline (day -1) and post-clindamycin (day 0) timepoints (N = 31), with
+Benjamini-Hochberg correction for testing all identified OTUs (Table
+S9). The gray vertical line indicates the limit of detection.
 
@@
 Number of observed OTUs (A) and Shannon diversity index values (B)
-across sources of mice post-infection (day 1). Data were analyzed by
+across sources of mice 1-day post-infection. Data were analyzed by
 Kruskal-Wallis test with Benjamini-Hochberg correction for testing each
 day of the experiment and the adjusted \emph{P} value was \textless{}
-0.05. Significant \emph{P} values from the pairwise Wilcoxon comparisons
-between sources with Benjamini-Hochberg correction are shown (Table S1).
-PCoA of \(\theta_{YC}\) distances of post-infection stool samples.
-PERMANOVA analysis demonstrated that source and the interaction between
-source and cage explained most of the variation (combined
-R\textsuperscript{2} = 0.88, \emph{P} = 0.0001, Table S2). For A-C: each
-symbol representing the value for a stool sample from an individual
-mouse, circles represent experiment 1 mice and triangles represent
-experiment 2 mice. D. Plots highlighting the median (point) and
-interquantile range (colored lines) of the relative abundances for the
-top 20 bacteria out of the 44 OTUs that varied across sources
-post-infection. The gray vertical line indicates the limit of detection.
-For each timepoint OTUs with differential relative abundances across
-sources of mice were identified by Kruskal-Wallis test with
-Benjamini-Hochberg correction for testing all identified OTUs (Table
-S3).
+0.05 (Table S1). Significant \emph{P} values from the pairwise Wilcoxon
+comparisons between sources with Benjamini-Hochberg correction are shown
+(Table S2). PCoA of \(\theta_{YC}\) distances of 1-day post-infection
+stool samples. Source and the interaction between source and cage
+effects explained most of the variation between fecal communities
+(PERMANOVA combined R\textsuperscript{2} = 0.88, \emph{P} \textless{}
+0.001, Table S3). For A-C: each symbol represents the value for a stool
+sample from an individual mouse, circles represent experiment 1 mice and
+triangles represent experiment 2 mice. D. Plots highlighting the median
+(point) and interquantile range (colored lines) of the relative
+abundances for the top 20 bacteria out of the 44 OTUs that varied
+between sources 1-day post-infection. The gray vertical line indicates
+the limit of detection. For each timepoint OTUs with differential
+relative abundances across sources of mice were identified by
+Kruskal-Wallis test with Benjamini-Hochberg correction for testing all
+identified OTUs (Table S10). E. \(\theta_{YC}\) distances of fecal
+samples collected 7-days post-infection relative to the baseline (day
+-1) sample for each mouse. Each symbol represents an individual mouse.
+Gray lines represent the median for each source.
 
@@
 \includegraphics{figure_5.pdf} \textbf{Figure 5. Bacteria that influence
-whether mice cleared \emph{C. difficile} by day 7.} A. Baseline relative
-abundance data for 3 of the OTUs from the classification model based on
-day 0 OTU relative abundances that significantly varied across sources
-of mice and had high relative abundances in the community. Symbols
-represent the relative abundance data for an individual mouse, circles
-represent mice that cleared \emph{C. difficile} by day 7, X-shapes
-represent mice that were still colonized with \emph{C. difficile}, and
-open circles represent mice that did not have \emph{C. difficile} CFU
-counts for day 7 post-infection. Gray lines indicate the median relative
-abundances for each source. Asterisks are shown for pairwise Wilcoxon
-comparisons with Benjamini-Hochberg correction where \emph{P}
-\textless{} 0.05. B. Venn diagram that combines Fig. S4 summaries of
-OTUs that were important to the day -1, 0, and 1 classification models
-(Table S8) and either overlapped with taxa that varied across sources at
-the same timepoint, were impacted by clindamycin treatment, or both. See
-Fig. S4 for separate comparisons of taxa from the day -1, 0, and 1
-classification models. Red OTUs were important to more than 1
-classification model.
-
-\newpage
-
-\includegraphics{figure_6.pdf} \textbf{Figure 6: Key OTUs vary across
-sources throughout the experiment.} A-D. Relative abundances of red OTUs
-from Fig. 5A that were important for at least two classification models
-are shown over time. A. \emph{Bacteroides} (OTU 2), which varied across
-sources throughout the experiment. B-C. \emph{Enterobacteriaceae} (B)
-and \emph{Enterococcus} (C), which significantly varied across sources
-and were impacted by clindamycin treatment. D. \emph{Porphyromonadaceae}
+whether mice cleared \emph{C. difficile} by day 7.} A. Baseline (day 0)
+relative abundance data for the 10 OTUs with the highest rankings based
+on feature weights in the baseline (day 0) classification model. Red
+font represents OTUs that correlated with \emph{C. difficile}
+colonization and blue font represents OTUs that correlated with
+clearance. Symbols represent the relative abundance data for an
+individual mouse. Gray lines indicate the median relative abundances for
+each source. B. Venn diagram that combines Fig. S4 summaries of OTUs
+that were important to the day -1, 0, and 1 classification models (Table
+S14) and either overlapped with taxa that varied across sources at the
+same timepoint, were impacted by clindamycin treatment, or both. Red
+OTUs were important to more than 1 classification model.
+
+\newpage
+
+\includegraphics{figure_6.pdf} \textbf{Figure 6: OTUs associated with
+\emph{C. difficile} colonization dynamics vary across sources throughout
+the experiment.} A-D. Relative abundances of red OTUs from Fig. 5A that
+were important for at least two classification models are shown over
+time. A. \emph{Bacteroides} (OTU 2), which varied across sources
+throughout the experiment. B-C. \emph{Enterobacteriaceae} (B) and
+\emph{Enterococcus} (C), which significantly varied across sources and
+were impacted by clindamycin treatment. D. \emph{Porphyromonadaceae}
 (OTU 7), which was significantly impacted by clindamycin treatment and
-examining relative abundance dynamics over the course of the experiment,
-revealed timepoints where relative abundances also significantly varied
-across sources of mice. Symbols represent the relative abundance data
-for an individual mouse, circles represent mice that cleared \emph{C.
-difficile} by day 7, X-shapes represent mice that were still colonized
-with \emph{C. difficile}, and open circles represent mice that did not
-have \emph{C. difficile} CFU counts for day 7 post-infection. Colored
-lines indicate the median relative abundances for each source. The gray
-horizontal line represents the limit of detection. Timepoints where
-differences across sources of mice were statistically significant by
-Kruskal-Wallis test with Benjamini-Hochberg correction for testing
-across multiple days (Table S9) are identified by the asterisk(s) above
-each timepoint (*, P \textless{} 0.05).
+after examining relative abundance dynamics over the course of the
+experiment was found to also significantly vary between sources of mice
+on days -1, 5, 6, 7, and 9 of the experiment. Symbols represent the
+relative abundance data for an individual mouse. Colored lines indicate
+the median relative abundances for each source. The gray horizontal line
+represents the limit of detection. Timepoints where differences between
+sources of mice were statistically significant by Kruskal-Wallis test
+with Benjamini-Hochberg correction for testing across multiple days
+(Table S15) are identified by the asterisk above each timepoint (*, P
+\textless{} 0.05).
 
@@
 mouse with color corresponding to experiment and shape representing cage
-mates. PERMANOVA was performed within each group to examine the
-contributions of experiment and cage to observed variation. Experiment
-number and cage only significantly explained observed variation for mice
-from the Schloss (combined R\textsuperscript{2} = 0.99; \emph{P} \(\le\)
-0.033) and Young (combined R\textsuperscript{2} = 0.95; \emph{P} \(\le\)
-0.029) lab colonies (Table S2). G-H: Boxplots of the \(\theta_{YC}\)
-distances of the 6 sources of mice relative to mice within the same
-source and experiment (G) or mice within the same source and between
-experiments (H) at baseline. Symbols represent individual mouse samples:
-circles for experiment 1 and triangles for experiment 2.
+mates. Experiment number and cage effects explained most of the observed
+variation for samples from the Schloss (PERMANOVA combined
+R\textsuperscript{2} = 0.99; \emph{P} \(\le\) 0.033) and Young (combined
+R\textsuperscript{2} = 0.95; \emph{P} \(\le\) 0.03) mice (Table S4).
+G-H: Boxplots of the \(\theta_{YC}\) distances of the 6 sources of mice
+relative to mice within the same source and experiment (G) or mice
+within the same source and between experiments (H) at baseline (day -1).
+Symbols represent individual mouse samples: circles for experiment 1 and
+triangles for experiment 2.
 
@@
 represent individual mice and the black line represents the limit of
-detection. C. \emph{C. difficile} CFU/gram stool on day 7 post-infection
-across sources of mice with asterisks for pairwise Wilcoxon comparisons
-with Benjamini-Hochberg correction where \emph{P} \textless{} 0.05. D.
-Mouse weight change 2 days post-infection across sources of mice, no
-pairwise Wilcoxon comparisons were significant after Benjamini-Hochberg
-correction. For C-D: circles represent experiment 1 mice, triangles
-represent experiment 2 mice and gray lines indicate the median values
-for each group. E. Percent of mice that were colonized with \emph{C.
-difficile} over the course of the experiment. Each day the percent is
-calculated based on the mice where \emph{C. difficile} CFU was
-quantified for that particular day. Total N for each day: day 1 (N =
+detection. C. \emph{C. difficile} CFU/gram stool 7-days post-infection
+across sources of mice with an asterisk for pairwise Wilcoxon
+comparisons with Benjamini-Hochberg correction where \emph{P}
+\textless{} 0.05. D. Mouse weight change 2-days post-infection across
+sources of mice, no pairwise Wilcoxon comparisons were significant after
+Benjamini-Hochberg correction. For C-D: circles represent experiment 1
+mice, triangles represent experiment 2 mice and gray lines indicate the
+median values for each group. E. Percent of mice that were colonized
+with \emph{C. difficile} over the course of the experiment. Each day the
+percent is calculated based on the mice where \emph{C. difficile} CFU
+was quantified for that particular day. Total N for each day: day 1 (N =
 42), day 2 (N = 20), day 3 (N = 39), day 4 (N = 29), day 5 (N = 43), day
@@
 can predict \emph{C. difficile} colonization status 7 days
-post-challenge.} A. Bar graph visualizations of overall day 7 C.
-difficile colonization status that were used as classification outcomes
-to build L2-regularized logistic regression models. Mice were classified
-as colonized or cleared (not detectable at the limit of detection of 100
-CFU) based on CFU g/stool data from 7 days post-infection. B. \emph{C.
-difficile} CFU status on Day 7 within each mouse source. N = 5-9 mice
-per group. C. L2-regularized logistic regression classification model
-area under the receiving operator characteristic curve (AUROCs) to
-predict \emph{C. difficile} CFU on day 7 post-infectoin (Fig. 1D, Fig.
-S3) based on the OTU community relative abundances at baseline (day -1),
-post-clindamycin (day 0), and post-infection (day 1). All models
-performed better than random chance (AUROC = 0.5; all \emph{P} \(\le\)
-1.6e-17; Table S7) and the model built with post-clindamycin treated
-bacterial OTU relative abundances had the best performance
-((\emph{P}\textsubscript{FDR} \(\le\) 3.3e-06 for all pairwise
-comparisons; Table S7). For list of the 20 OTUs that were ranked as most
-important to each model, see Table S8.
+post-challenge.} A. Bar graph visualizations of overall 7-days
+post-infection \emph{C. difficile} colonization status that were used as
+classification outcomes to build L2-regularized logistic regression
+models. Mice were classified as colonized or cleared (not detectable at
+the limit of detection of 100 CFU) based on CFU g/stool data from 7 days
+post-infection. B. \emph{C. difficile} CFU status on Day 7 within each
+mouse source. N = 8-9 mice per group. C. L2-regularized logistic
+regression classification model area under the receiving operator
+characteristic curve (AUROCs) to predict \emph{C. difficile} CFU on day
+7 post-infection (Fig. 2B, Fig. S2C) based on the OTU community relative
+abundances at baseline (day -1), post-clindamycin (day 0), and 1-day
+post-infection. All models performed better than random chance (AUROC =
+0.5, all \emph{P} \(\le\) 1.6e-17, Table S12) and the model built with
+post-clindamycin treated bacterial OTU relative abundances had the best
+performance ((\emph{P}\textsubscript{FDR} \(\le\) 3.3e-06 for all
+pairwise comparisons, Table S13). For list of the 20 OTUs that were
+ranked as most important to each model, see Table S14.
 
@@
 post-infection community data vary by source, clindamycin treatment, or
-both.} A-C. Venn diagrams of top 20 important OTUs from baseline (A),
-post-clindamycin treatment (B), and post-infection (C) classification
-models (Table S8) that overlapped with OTUs that varied across sources
-at the corresponding timepoint, were impacted by clindamycin treatment,
-or both. Bold OTUs signify OTUs that were important to more than 1
-classification model.
-
-\newpage
-
+both.} A-C. Venn diagrams of OTUs from the top 20 OTUs from the baseline
+(A), post-clindamycin treatment (B), and post-infection (C)
+classification models (Table S14) that overlapped with OTUs that varied
+across sources at the corresponding timepoint (Tables S5, 8, 10), were
+impacted by clindamycin treatment (Table S9), or both. Red OTUs signify
+OTUs that were important to more than 1 classification model.
+
+\newpage
+
+\hypertarget{supplementary-tables-and-movie}{%
 \subsection{Supplementary Tables and
-Movie}\label{supplementary-tables-and-movie}
-
-All supplemental material is available at:
-\url{https://github.com/SchlossLab/Tomkovich_Vendor_XXXX_2020/submission}.
+Movie}\label{supplementary-tables-and-movie}}
 
@@
 after clindamycin and \emph{C. difficile} infection.} PCoA of
-\(\theta_{YC}\) distances animated from 0 through 9 days post-infection.
-PERMANOVA analysis indicated source was the variable that explained the
-most observed variation across fecal communities (source
-R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by interactions
-between cage and day of the experiment. Transparency of the circle
-corresponds to the day of the experiment, each circle represents a
-sample from an individual mouse at a specific timepoint. See Table S2
-for PERMANOVA results). Circles represent mice from experiment 1 and
-triangles represent mice from expeirment 2.
-
-\textbf{Table S1. Diversity metrics Kruskal-Wallis and pairwise Wilcoxon
-statistical results.}
-
-\textbf{Table S2. PERMANOVA results for mice at specific timepoints and
-across all timepoints.}
-
-\textbf{Table S3. OTUs with relative abudances that significantly vary
-across sources at baseline, post-clindamycin, or post-infection
-timepoints.}
-
-\textbf{Table S4. \emph{C. difficile} CFU statistical results.}
-
-\textbf{Table S5. Mouse weight change statistical results.}
-
-\textbf{Table S6. OTUs with relative abudances that significantly
+\(\theta_{YC}\) distances animated from days -1 through 9 of the
+experiment. Source was the variable that explained the most observed
+variation across fecal communities (PERMANOVA source
+R\textsuperscript{2} = 0.35, \emph{P} = 0.0001, Table S11) followed by
+interactions between cage and day of the experiment. Transparency of the
+symbol corresponds to the day of the experiment, each symbol represents
+a sample from an individual mouse at a specific timepoint. Circles
+represent mice from experiment 1 and triangles represent mice from
+expeirment 2.
+
+\textbf{Tables S1-S15. Excel workbook of Tables S1-S15.}
+
+\textbf{Table S1. Alpha diversity metrics Kruskal-Wallis statistical
+results.}
+
+\textbf{Table S2. Alpha diversity metrics pairwise Wilcoxon statistical
+results.}
+
+\textbf{Table S3. PERMANOVA results for mice at baseline (day -1),
+post-clindamycin (day 0), and post-infection (day 1).}
+
+\textbf{Table S4. PERMANOVA results for each source of mice at baseline
+(day -1).}
+
+\textbf{Table S5. OTUs with relative abudances that significantly vary
+between sources at baseline (day -1).}
+
+\textbf{Table S6. \emph{C. difficile} CFU statistical results.}
+
+\textbf{Table S7. Mouse weight change statistical results.}
+
+\textbf{Table S8. OTUs with relative abudances that significantly vary
+between sources post-clindamycin (day 0).}
+
+\textbf{Table S9. OTUs with relative abudances that significantly
 changed after clindamycin treatment.}
 
-\textbf{Table S7. Statistical results of L2-regularized logistic
+\textbf{Table S10. OTUs with relative abudances that significantly vary
+between sources post-infection (day 1).}
+
+\textbf{Table S11. PERMANOVA results for mice across all timepoints.}
+
+\textbf{Table S12. Statistical results of L2-regularized logistic
+regression model performances compared to random chance.}
+
+\textbf{Table S13. Pairwise comparisons of L2-regularized logistic
 regression model performances.}
 
-\textbf{Table S8. Top 20 most important OTUs for each of the 3
+\textbf{Table S14. Top 20 most important OTUs for each of the 3
 L2-regularized logistic regression models based on OTU relative
@@
 
-\textbf{Table S9. OTUs with relative abudances that significantly varied
-across sources of mice on at least 1 day of the experiment by
+\textbf{Table S15. OTUs with relative abundances that significantly
+varied across sources of mice on at least 1 day of the experiment by
 Kruskal-Wallis test.}

### 2020-08-11 submission/manuscript.Rmd
Incorporate Pat's edits

@@
 The gut microbiota has a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs).
-However, much of the mechanistic work examining CDIs in mouse models use animals obtained from a single university colony or vendor.
-We treated mice from 6 different sources (2 University of Michigan colonies and 4 vendors) with a single clindamycin dose, followed by a *C.
+However, much of the mechanistic work examining CDIs in mouse models use animals obtained from a single source.
+We treated mice from 6 sources (2 University of Michigan colonies and 4 vendors) with 1 clindamycin dose, followed by *C.
 difficile* challenge 1 day later and then measured *C.
@@
 difficile* challenge.
-While all sources of mice were colonized 1-day post-infection, variation emerged from days 3-7 post-infection with animals from some sources colonized with *C.
+While all mice were colonized 1-day post-infection, variation emerged from days 3-7 post-infection with animals from some sources colonized with *C.
 difficile* for longer and at higher levels.
@@
 difficile* by 7 days post-infection using baseline, post-clindamycin, and 1-day post-infection community composition data.
-With these models, we identified 4 bacteria that varied across sources (*Bacteroides*), were altered by clindamycin (*Porphyromonadaceae*), or both (*Enterobacteriaceae* and *Enterococcus*).
+With these models, we identified 4 bacteria that were predictive of whether *C.
+difficile* cleared.
+They varied across sources (*Bacteroides*), were altered by clindamycin (*Porphyromonadaceae*), or both (*Enterobacteriaceae* and *Enterococcus*).
 Microbiota variation across sources better emulates human inter-individual variation and can help identify bacterial drivers of phenotypic variation in the context of CDIs.
@@
 *Clostridioides difficile* is a leading nosocomial infection.
-Although perturbation to the gut microbiota is an established risk, there is variation in who becomes asymptomatically colonized, develops an infection, or has an infection with adverse outcomes.
-*C.
-difficile* infection (CDI) mouse models are widely used to answer a variety of *C.
+Although perturbation to the gut microbiota is an established risk, there is variation in who becomes asymptomatically colonized, develops an infection, or has adverse infection outcomes.
+Mouse models of *C.
+difficile* infection (CDI) are widely used to answer a variety of *C.
 difficile* pathogenesis questions.
@@
 Additionally, asymptomatic colonization, where *C.
-difficile* is detectable, but symptoms are absent has been documented in infants and adults [@Zacharioudakis2015; @Crobach2018].
+difficile* is detectable, but symptoms are absent, has been documented in infants and adults [@Zacharioudakis2015; @Crobach2018].
 The intestinal microbiota has been implicated in asymptomatic colonization [@Zhang2015; @VanInsberghe2020], susceptibility to CDIs [@Mancabelli2017, @Duvallet2017], and adverse CDI outcomes [@Seekatz2016; @Khanna2016; @Pakpour2017; @Lee2020].
@@
 1C).
-Interactions between the source and cage effects explained most of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, Fig.
+The source of mice and the interactions between the source and cage effects explained most of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, Fig.
 1C and Table S3).
 Mice that are co-housed tend to have similar gut microbiotas due to coprophagy [@Nguyen2015].
-Since mice within the same source were housed together, it is not surprising that the cage effect also contributed to the observed microbiota variation.
+Since mice within the same source were housed together, it was not surprising that the cage effect also contributed to the observed community variation.
 There were some differences between the 2 experiments we conducted, as the experiment and cage effects significantly explained the observed community variation for the Schloss and Young lab mouse colonies (Fig.
@@
 1D and Table S5).
-Though we saw differences between experiments at the community level, there were no OTUs that were significantly different between experiments within Schloss, Young, and Envigo mice (Data not shown).
-By using mice from six sources we were able to increase the number of microbiota communities to treat with clindamycin and challenge with *C.
-difficile*.
+Though we saw differences between experiments at the community level, there were no OTUs that were significantly different between experiments within Schloss, Young, and Envigo mice (all *P* > 0.05).
+By using mice from six sources we were able to increase the variation in the starting communities to evaluate in a clindamycin-based CDI model.
 
@@
 We have previously demonstrated mice are rendered susceptible to *C.
-difficile*, but cleared the pathogen within 9 days when treated with clindamycin alone [@Jenior2018; @Tomkovich2019].
+difficile*, but clear the pathogen within 9 days when treated with clindamycin alone [@Jenior2018; @Tomkovich2019].
 All mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C.
@@
 difficile* clearance.
-While the colonization dynamics were similar between the two experiments, the Schloss mice took longer to clear in the 1st experiment and the Envigo mice took longer to clear in the 2nd experiment (Fig.
+While the colonization dynamics were similar between the two experiments, the Schloss mice took longer to clear in the 1st experiment compared to the 2nd and the Envigo mice took longer to clear in the 2nd experiment compared to the 1st (Fig.
 S2A-B).
-The change in the mice's weight significantly varied across sources of mice with the most weight lost two days post-infection (Fig.
+The change in the mice's weight significantly varied across sources of mice with the most weight loss occurring two days post-infection (Fig.
 2C and Table S7).
@@
 difficile* (Fig.
-2C) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment.
+2B) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment.
 Thus, clindamycin rendered all mice susceptible to *C.
@@
 ```
-**Clindamycin treatment alters bacteria in all sources, but a subset of bacterial differences across sources persists.** Given the variation in mouse microbiotas that we observed across breeding colonies, we hypothesized that variation in *C.
-difficile* clearance would be explained by microbiota variation across the 6 sources of mice.
+**Clindamycin treatment alters bacteria in all sources, but a subset of bacterial differences across sources persists.** Given the variation in fecal communities that we observed across breeding colonies, we hypothesized that variation in *C.
+difficile* clearance would be explained by community variation across the 6 sources of mice.
 As expected, clindamycin treatment decreased the richness and Shannon diversity across all sources of mice (Fig.
@@
 The clindamycin treatment decreased the variation in community structures between sources of mice.
-Source and cage effects explained almost all of the observed variation between communities (combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig.
+The source of mice and the interactions between source and cage effects explained almost all of the observed variation between communities (combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig.
 3C and Table S3).
@@
 3D and Table S8) with relative abundances that significantly varied between sources.
-Next we identified the bacteria that shifted after clindamycin treatment, regardless of source by analyzing paired fecal samples from mice that were collected at baseline and after clindamycin treatment.
+Next, we identified the bacteria that shifted after clindamycin treatment, regardless of source by analyzing paired fecal samples from mice that were collected at baseline and after clindamycin treatment.
 We identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment in most mice (Fig.
@@
 3D), but only in Taconic mice.
-*Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment (Fig.
+*Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased in relative abundance after clindamycin treatment (Fig.
 3D).
@@
 difficile* challenge.** One day post-infection, significant differences in diversity metrics remained across sources (*P*~FDR~ < 0.05, Fig 4A-B and Tables S1-2).
-Although the Charles River mice had more diverse microbiotas and were also able to clear *C.
-difficile* faster than some of the other sources.
-Microbiota diversity did not explain the observed variation in *C.
-difficile* colonization across sources considering the Young and Schloss mice had the lowest diversity after clindamycin treatment and were able to clear *C.
+Although the Charles River mice had more diverse communities and were also able to clear *C.
+difficile* faster than the other sources, diversity did not explain the observed variation in *C.
+difficile* colonization across sources.
+The Young and Schloss mice had the lowest diversity 1 day post-infection and were able to clear *C.
 difficile* earlier than Jackson, Taconic and Envigo mice.
-Source and the interactions between source and cage effects continued to explain most of the observed community variation (combined R^2^ = `r d1_rsq_sum`; *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig.
-4D and Table S3).
+The source of mice and the interactions between source and cage effects continued to explain most of the observed community variation (combined R^2^ = `r d1_rsq_sum`; *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig.
+4C and Table S3).
 One day after *C.
@@
 
-Throughout the experiment the source of mice continued to be the dominant factor that explained the observed variation across fecal communities (PERMANOVA R^2^ = `r rsq_adonis_source`, *P* `r format.pval(p_adonis_source, eps = 0.001)`) followed by interactions between cage effects  and the day of the experiment (Movie S1 and Table S11).
+Throughout the experiment, the source of mice continued to be the dominant factor that explained the observed variation across fecal communities (PERMANOVA R^2^ = `r rsq_adonis_source`, *P* `r format.pval(p_adonis_source, eps = 0.001)`) followed by interactions between cage effects  and the day of the experiment (Movie S1 and Table S11).
 Mice fecal samples from the same source of mice continued to cluster closely to each other throughout the experiment.
 By 7 days post-infection, when approximately `r format(100-percent_colonized_d7, digits = 1)`% mice had cleared *C.
-difficile*, most of the mice still had not recovered to their baseline community structure (Fig.
+difficile*, most of the mice had not recovered to their baseline community structure (Fig.
 4E).
-Distance from the baseline community between sources did not explain the variation in *C.
-difficile* clearance since Schloss and Young mice cleared *C.
-difficile* faster, but their communities were a greater distance from baseline 7 days post-infection.
-In summary, mouse bacterial communities varied significantly between sources throughout the course of the experiment and a consistent subset of bacterial taxa remained different between sources regardless of clindamycin and *C.
+The distance to the baseline community did not explain the variation in *C.
+difficile* clearance as the Schloss and Young mice had mostly cleared *C.
+difficile*, but their communities were a greater distance from baseline 7 days post-infection compared to the Jackson and Taconic mice that were still colonized.
+In summary, mouse bacterial communities varied significantly between sources throughout the course of the experiment and a consistent subset of bacteria remained different between sources regardless of clindamycin and *C.
 difficile* challenge.
@@
 median_otu_day0_AUROC <- models_vs_random %>% filter(model_name == "otu_day0") %>% pull(median)
+median_otu_dayn1_AUROC <- models_vs_random %>% filter(model_name == "otu_dayn1") %>% pull(median)
+median_otu_day1_AUROC <- models_vs_random %>% filter(model_name == "otu_day1") %>% pull(median)
 
@@
 **Baseline, post-clindamycin, and post-infection community data can predict mice that will clear *C.
-difficile* by 7 days post-infection.** After identifying taxa that varied between sources, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C.
+difficile* by 7 days post-infection.** After identifying taxa that varied between sources, changed after clindamycin treatment, or both, we determined which taxa were influencing the variation in *C.
 difficile* colonization at day 7 (Fig.
@@
 difficile* colonization status on day 7 than random chance (all *P* `r format.pval(max_p_vs_random, eps = 0.001)`, Table S12).
-The model based on the post-clindamycin (day 0) community OTU data performed significantly better than the other models with an area under the receiving operator characteristic curve (AUROC) of `r median_otu_day0_AUROC` (*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for pairwise comparisons, Table S13).
+The model based on the post-clindamycin (AUROC = `r median_otu_day0_AUROC`) community OTU data performed significantly better than the baseline (AUROC = `r median_otu_dayn1_AUROC`) or the post-infection (AUROC = `r median_otu_day1_AUROC`) models (*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for pairwise comparisons, Fig.
+S3C and Table S13).
 Thus, we were able to use bacterial relative abundance data from the time of *C.
@@
 5A).
-Next, we examined whether any of the top 20 ranked OTUs from the baseline (day 0) model were also important in the other 2 classification models based on baseline (day -1) and 1 day post-infection community data.
-We identified 6 OTUs (`r list_overlapping_across_models`) that were important to the day 0 model and either the baseline or 1 day post-infection models (Table S14).
+Next, we examined whether any of the top 20 ranked OTUs from the post-clindamycin (day 0) model were also important in the other 2 classification models based on baseline (day -1) and 1 day post-infection community data.
+We identified 6 OTUs (*`r list_overlapping_across_models`*) that were important to the post-clindamycin model and either the baseline or 1 day post-infection models (Table S14).
 Thus, a subset of bacterial OTUs were important for determining *C.
@@
 
-To determine whether the OTUs driving the classification models also varied between sources, were altered by clindamycin treatment, or both, we examined whether the top 20 ranked OTUs from each model overlapped with OTUs that varied between sources (Fig.
+To determine whether the OTUs driving the classification models also varied between sources, were altered by clindamycin treatment, or both, we identified the OTUs from each model that varied between sources (Fig.
 1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin treatment (Fig.
-3E and Table S9).
-We found a subset of OTUs that were important to the baseline (day -1), post-clindamycin (day 0), and 1 day post-infection models and overlapped with bacteria that varied between sources, were altered by clindamycin treatment, or both (Fig.
+3E and Table S9; Fig.
 S4).
-Combining the overall comparison results for the 3 models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig.
+Comparing the features important to the 3 models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig.
 5B).
+Together, these results suggest that the initial bacterial communities and their responses to clindamycin influence the clearance of *C.
+difficile*.
+
 Several OTUs (*`r list_key_otus`*) that overlapped with our previous analyses appeared across at least 2 models, so we examined how the relative abundances of these OTUs varied over the course of the experiment (Fig.
 6).
-Throughout the experiment, there was at least 1 timepoint where relative abundances of these OTUs significantly varied between sources (Table S15).
-Interestingly, there were no OTUs that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection, suggesting multiple bacteria influence *C.
+Across the 9 days post-infection, there was at least 1 timepoint when the relative abundances of these OTUs significantly varied between sources (Table S15).
+Interestingly, there were no OTUs that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection, suggesting that multiple bacteria influence *C.
 difficile* colonization dynamics.
-Together, these results suggest the initial bacterial communities and their responses to clindamycin influence the clearance of *C.
-difficile*.
 
@@
 
-By running our CDI model with mice from 6 different sources, we were able to identify bacterial taxa that were unique to sources throughout the experiment as well as taxa that were universally impacted by clindamycin.
+Applying our CDI model with mice from 6 different sources, allowed us to identify bacterial taxa that were unique to different sources as well as taxa that were universally impacted by clindamycin.
 We trained L2 logistic regression models with baseline (day -1), post-clindamycin treatment (day 0), and 1-day post-infection fecal community data that could predict whether mice cleared *C.
@@
 difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by source, were impacted by clindamycin treatment, or both.
-Overall, our results demonstrated clindamycin was sufficient to render mice from multiple sources susceptible to CDI and only a subset of the interindividual microbiota variation across mice from different sources was associated with the time needed to clear *C.
+Overall, our results demonstrated clindamycin was sufficient to render mice from multiple sources susceptible to CDI and only a subset of the inter-individual microbiota variation across mice from different sources was associated with the time needed to clear *C.
 difficile*.
@@
 Other studies have taken similar approaches by using mice from multiple sources to identify bacteria that either promote colonization resistance or increase susceptibility to enteric infections [@Velazquez2019; @EtienneMesmin2017; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018].
-For example, in the context of *Salmonella* infections, *Enterobacteriaceae* and segmented filamentous bacteria have emerged as protective [@Velazquez2019; @Lai2020].
+For example, against *Salmonella* infections, *Enterobacteriaceae* and segmented filamentous bacteria have emerged as protective [@Velazquez2019; @Lai2020].
+We found *Enterobacteriaceae* increased in all sources of mice after clindamycin treatment, facilitating *C.
+difficile* colonization.
+However, there was also variation in *Enterobacteriaceae* relative abundance levels between sources that was associated with the variation in *C.
+difficile* colonization dynamics across sources.
+Thus, bacteria may have differential roles in determining susceptibility depending on the type of bacterial infection.
+
+Differences in CDI mouse model studies have been attributed to intestinal microbiota variation across sources.
+For example, researchers using the same clindamycin treatment and C57BL/6 mice had different *C.
+difficile* outcomes, one having sustained colonization [@Buffie2011], while the other had transient [@Reeves2011], despite both using *C.
+difficile* VPI 10643.
+Baseline differences in the microbiota composition have been hypothesized to partially explain the differences in colonization outcomes and overall susceptibility to *C.
+difficile* after treatment with the same antibiotic [@Lawley2013; @Hutton2014].
 A previous study with *C.
@@
 However, we did not identify any endogenous *C.
-difficile* strains prior to challenge, suggesting there were no endogenous protective strains in the mice we received and other bacterial taxa mediated the variation in *C.
+difficile* strains prior to challenge, suggesting there were no endogenous protective strains in the mice we received and other bacteria mediated the variation in *C.
 difficile* colonization across sources.
-
-Differences in CDI mouse model studies have been attributed to intestinal microbiota variation across sources.
-For example, groups using the same clindamycin treatment and C57BL/6 mice had different *C.
-difficile* outcomes, one having sustained colonization [@Buffie2011], while the other had transient [@Reeves2011], despite both using *C.
-difficile* VPI 10643.
-Baseline differences in the microbiota composition have been hypothesized to partially explain the differences in colonization outcomes and overall susceptibility to *C.
-difficile* after treatment with the same antibiotic [@Lawley2013; @Hutton2014].
-The bacterial perturbations induced by clindamycin treatment have been well characterized and our findings agree with previous CDI mouse model work demonstrating *Enterococcus* and *Enterobacteriaceae* were associated with *C.
-difficile* susceptibility and *Porpyhromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, and *Turicibacter* were associated with resistance [@Schubert2015; @Jenior2018; @Tomkovich2019; @Buffie2011; @Buffie2014; @Lawley2009; @Lawley2012; @Jump2014].
-While we have demonstrated that susceptibility is uniform across sources of mice after clindamycin treatment, there could be different outcomes for either susceptibility or clearance in the case of other antibiotic treatments.
 The *C.
@@
 For example, a group found differential colonization outcomes after clindamycin treatment, with *C.
-difficile* 630 and M68 infections eventually becoming undectable while strain BI-7 remained detectable up to 70 days post-treatment [@Lawley2012].
+difficile* 630 and M68 infections eventually becoming undetectable while strain BI-7 remained detectable up to 70 days post-treatment [@Lawley2012].
+
+The bacterial perturbations induced by clindamycin treatment have been well characterized and our findings agree with previous CDI mouse model work demonstrating *Enterococcus* and *Enterobacteriaceae* were associated with *C.
+difficile* susceptibility and *Porpyhromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, and *Turicibacter* were associated with resistance [@Schubert2015; @Jenior2018; @Tomkovich2019; @Buffie2011; @Buffie2014; @Lawley2009; @Lawley2012; @Jump2014].
+While we have demonstrated that susceptibility is uniform across sources of mice after clindamycin treatment, there could be different outcomes for either susceptibility or clearance in the case of other antibiotic treatments.
+
 We found the time needed to naturally clear *C.
@@
 
-Our approach successfully increased the diversity of murine bacterial communities tested in our clindamycin *C.
-difficile* model.
+Our approach successfully increased the inter-animal variation tested with our clindamycin model of CDI.
 One alternative approach that has been used in some CDI studies [@NagaoKitamoto2020; @Battaglioli2018; @Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018] is to associate mice with human microbiotas.
 However, a major caveat to this method is the substantial loss of human microbiota community members upon transfer to mice [@Fouladi2020; @Walter2020].
-Additionally with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of the CDI mouse model studies to date associated mice with just 1 types of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018], which does not aid in the goal of modeling the interpersonal variation seen in humans to understand how the microbiota influences susceptiblity to CDIs and adverse outcomes.
+Additionally with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of the CDI mouse model studies to date associated mice with just 1 types of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018], which does not aid in the goal of modeling the interpersonal variation seen in humans to understand how the microbiota influences susceptibility to CDIs and adverse outcomes.
 Importantly, our study using mice from 6 different sources increased the variation between groups of mice compared to using 1 source alone, to better reflect the inter-individual microbiota variation observed in humans.
+
+Another motivation for associating mice with human microbiotas,is to study the bacteria associated with the disease in humans.
 Encouragingly, decreased *Bifidobacterium*, *Porphyromonas*, *Ruminococcaceae* and *Lachnospiraceae* and increased *Enterobacteriaceae*, *Enterococcus*, *Lactobacillus*, and *Proteus* have all been associated with human CDIs [@Mancabelli2017, @Duvallet2017] and were well represented in our study, suggesting most of the mouse sources are suitable for gaining insights into the bacteria influencing *C.
@@
 An important exception was *Enterococcus*, which was primarily absent from University of Michigan colonies and *Proteus*, which was only found in Taconic mice.
-Importantly, the fact that some CDI-associated bacteria were only found in a subset of mice has important implications for future CDI mouse model studies.
+The fact that some CDI-associated bacteria were only found in a subset of mice has important implications for future CDI mouse model studies.
 
-There are several limitations to our work.
+Other microbiota and host factors that were outside the scope of our current study may also contribute to the differences in *C.
+difficile* colonization dynamics between sources of mice.
 The microbiota is composed of viruses, fungi, and parasites in addition to bacteria, and these non-bacterial members can also vary across sources of mice [@Rasmussen2019; @Mims2020].
@@
 difficile* resistance [@Buffie2014; @Lawley2012].
-Although, we only looked at composition, *Ruminococcaceae* and *Lachnospiraceae* both emerged as important taxa for classifying day 7 *C.
-difficile* colonization status and metagenomes from these bacteria have been shown to contain the bile acid-inducible gene cluster necessary for secondary bile acid formation and ability to produce butyrate [@Vital2019; @Collins2015].
 Interestingly, butyrate has previously been shown to vary across mouse vendors and mediated resistance to *Citrobacter rodentium* infection, a model of enterohemorrhagic and enteropathogenic *Escherichia coli* infections [@Osbelt2020].
@@
 
-We have demonstrated that the ways baseline microbiotas from different mouse sources respond to clindamycin treatment influences the length of time mice remained colonized with *C.
+We have demonstrated that the ways baseline microbiotas from different mouse sources respond to clindamycin treatment influence the length of time mice remained colonized with *C.
 difficile* 630.
-For those interested in dissecting the contribution of the microbiota to *C.
-difficile* pathogenesis and treatments, using multiple sources of mice may yield more insights than a single model alone.
-Furthermore, for studies wanting to examine the interplay between a particular bactera such as *Enterococcus* and *C.
-difficile*, these results could serve as a resource for selecting which mice to order to address the question.
-Using mice from multiple sources helps model the interpersonal microbiota variation among humans to aid our understanding of how the gut microbiota contributes to CDIs.
+To better understand the contribution of the microbiota to *C.
+difficile* pathogenesis and treatments, using multiple sources of mice may yield more insights than a single source.
+Furthermore, for studies wanting to examine the interplay between particular bacteria such as *Enterococcus* and *C.
+difficile*, these results could serve as a resource for selecting mice to address their question.
+Using mice from multiple sources helps model the interpersonal microbiota variation among humans to aid our understanding of how the gut microbiota provides colonization resistance to CDIs.
 
@@
 ST was supported by the Michigan Institute for Clincial and Health Research Postdoctoral Translation Scholars Program (UL1TR002240).
-We thank members of the Schloss lab for feedback on planning the experiments and data presentation, as well as code tutorials and feedback through Code Club.
-In particular, we want to thank Begüm Topçuoğlu for help with implementing L2 logistic regression models using her [pipeline](https://github.com/SchlossLab/ML_pipeline_microbiome), Ana Taylor for help with media preparation and sample collection, and Nicholas Lesniak for his critical feedback on the manuscript.
+We thank members of the Schloss lab for feedback on planning the experiments and data presentation.
+In particular, we thank Begüm Topçuoğlu for help with implementing L2 logistic regression models, Ana Taylor for help with media preparation and sample collection, and Nicholas Lesniak for his critical feedback on the manuscript.
 We also thank members of Vincent Young’s lab, particularly  Kimberly Vendrov, for guidance with the *C.
 difficile* infection mouse model and donating the mice.
-We also want to thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the institutional support for our mouse experiments.
+We also thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the institutional support for our mouse experiments.
 Finally, we thank Kwi Kim, Austin Campbell, and Kimberly Vendrov for their help in maintaining the Schloss lab's anaerobic chamber.
@@
 Reads were combined and aligned with the SILVA reference database [@Quast2012].
-Chimeras were removed with the VSEARCH algorithm and taxonomic assignment was completed with a modified version (v16) of the Ribosomal Database Project reference database (v11.5) [@Cole2013] with an 80% cutoff.
+Chimeras were removed with the VSEARCH algorithm and taxonomic assignment was completed with a modified version (v16) of the Ribosomal Database Project reference database (v11.5) [@Cole2013] with an 80% confidence cutoff.
 Operational taxonomic units (OTUs) were assigned with a 97% similarity threshold using the opticlust algorithm [@Westcott2017].
-To account for uneven sequencing across samples, samples were rarefied to 5,437 sequences 1,000 times for alpha and beta diversity analyses.
+To account for uneven sequencing across samples, samples were rarefied to 5,437 sequences 1,000 times for alpha and beta diversity analyses, and a single time to generate relative abundances for model training.
 PCoAs were generated based on $\theta_{YC}$ distances.
@@
 Bacterial relative abundances that varied across sources at the OTU level were identified with the Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
-OTUs impacted by clindamycin treatment were identified using the Wilcoxon signed rank test with matched pairs of mice samples for day -1 and day 0.
+The Wilcoxon rank sum test was used to test for OTUs that differed between experiments within the Schloss, Young, and Envigo sources with Benjamini-Hochberg correction for testing all identified OTUs.
+OTUs impacted by clindamycin treatment were identified using the paired Wilcoxon signed rank test with matched pairs of mice samples for day -1 and day 0.
 To determine whether classification models had better performance (test AUROCs) than random chance (0.5), we used the one-sample Wilcoxon signed rank test.
@@
 D.
-Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the top 20 bacteria out of the `r source_otus_dn1` OTUs that varied across sources at baseline (Table S5).
+The median (point) and interquantile range (colored lines) of the relative abundances for the 20 most significant OTUs out of the `r source_otus_dn1` OTUs that varied across sources at baseline (Table S5).
 
@@
 Differences between sources were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 (Table S1).
-Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S2).
+Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are displayed as the first initial of each group compared to the group that they are listed above (Table S2).
 C.
@@
 D.
-Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the `r source_otus_d0` OTUs (Table S8)that varied between sources after clindamycin treatment (day 0).
+The median (point) and interquantile range (colored lines) of the relative abundances for the `r source_otus_d0` OTUs (Table S8) that varied between sources after clindamycin treatment (day 0).
 E.
-Plots highlighting the median (point) and interquantile range (colored lines) of the top 10 OTUs out of `r clindamycin_otus_sig` with relative abundances that changed after clindamycin treatment (adjusted *P* value < 0.05).
-Data were analyzed by Wilcoxon signed rank test of mice that had paired sequence data for baseline (day -1) and post-clindamycin (day 0) timepoints (N = `r n_paired_for_clind`), with Benjamini-Hochberg correction for testing all identified OTUs (Table S9).
+The median (point) and interquantile range (colored lines) of the top 10 most significant OTUs out of `r clindamycin_otus_sig` with relative abundances that changed because of the clindamycin treatment (adjusted *P* value < 0.05).
+Data were analyzed by paired Wilcoxon signed rank test of mice that had paired sequence data for baseline (day -1) and post-clindamycin (day 0) timepoints (N = `r n_paired_for_clind`), with Benjamini-Hochberg correction for testing all identified OTUs (Table S9).
 The gray vertical line indicates the limit of detection.
@@
 Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 (Table S1).
-Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are shown (Table S2).
+Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are displayed as the first initial of each group compared to the group that they are listed above (Table S2).
 PCoA of $\theta_{YC}$ distances of 1-day post-infection stool samples.
@@
 D.
-Plots highlighting the median (point) and interquantile range (colored lines) of the relative abundances for the top 20 bacteria out of the `r source_otus_d1` OTUs that varied between sources 1-day post-infection.
+The median (point) and interquantile range (colored lines) of the relative abundances for the top 20 most significant OTUs out of the `r source_otus_d1` OTUs that varied between sources 1-day post-infection.
 The gray vertical line indicates the limit of detection.
@@
 difficile* by day 7.** A.
-Baseline (day 0) relative abundance data for the 10 OTUs with the highest rankings based on feature weights in the baseline (day 0) classification model.
+Post-clindamycin (day 0) relative abundance data for the 10 OTUs with the highest rankings based on feature weights in the post-clindamycin (day 0) classification model.
 Red font represents OTUs that correlated with *C.
@@
 Symbols represent the relative abundance data for an individual mouse.
-Gray lines indicate the median relative abundances for each source.
+Gray bars indicate the median relative abundances for each source.
 B.
-Venn diagram that combines Fig.
-S4 summaries of OTUs that were important to the day -1, 0, and 1 classification models (Table S14) and either overlapped with taxa that varied across sources at the same timepoint, were impacted by clindamycin treatment, or both.
-Red OTUs were important to more than 1 classification model.
+Venn diagram that combines  OTUs that were important to the day -1, 0, and 1 classification models ( Fig.
+S4, Table S14) and either overlapped with taxa that varied across sources at the same timepoint, were impacted by clindamycin treatment, or both.
+Bold OTUs were important to more than 1 classification model.
 
@@
 difficile* colonization dynamics vary across sources throughout the experiment.** A-D.
-Relative abundances of red OTUs from Fig.
-5A that were important for at least two classification models are shown over time.
+Relative abundances of bold OTUs from Fig.
+5B that were important in at least two classification models are shown over time.
 A.
@@
 S2C) based on the OTU community relative abundances at baseline (day -1), post-clindamycin (day 0), and 1-day post-infection.
-All models performed better than random chance (AUROC = 0.5, all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`, Table S12) and the model built with post-clindamycin treated bacterial OTU relative abundances had the best performance ((*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for all pairwise comparisons, Table S13).
-For list of the 20 OTUs that were ranked as most important to each model, see Table S14.
+All models performed better than random chance (AUROC = 0.5, all *P* `r format.pval(max_p_vs_random, eps = 0.001)`, Table S12) and the model built with post-clindamycin bacterial OTU relative abundances had the best performance ((*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for all pairwise comparisons, Table S13).
+See Table S14 for list of the 20 OTUs that were ranked as most important to each model.
 
@@
 **Figure S4.
-OTUs from classification models based on baseline, post-clindamycin treatment, or post-infection community data vary by source, clindamycin treatment, or both.** A-C.
-Venn diagrams of OTUs from the top 20 OTUs from the baseline (A), post-clindamycin treatment (B), and post-infection (C) classification models (Table S14) that overlapped with OTUs that varied across sources at the corresponding timepoint (Tables S5, 8, 10), were impacted by clindamycin treatment (Table S9), or both.
-Red OTUs signify OTUs that were important to more than 1 classification model.
+OTUs from classification models based on baseline, post-clindamycin treatment, or 1-day post-infection community data vary by source, clindamycin treatment, or both.** A-C.
+Venn diagrams of OTUs from the top 20 OTUs from the baseline (A), post-clindamycin treatment (B), and 1-day post-infection (C) classification models (Table S14) that overlapped with OTUs that varied across sources at the corresponding timepoint (Tables S5, 8, 10), were impacted by clindamycin treatment (Table S9), or both.
+Bold OTUs were important to more than 1 classification model.
 
@@
 **Movie S1.
-Large shifts in bacterial community structure occurred after clindamycin and *C.
+Large shifts in bacterial community structures occurred after clindamycin and *C.
 difficile* infection.** PCoA of $\theta_{YC}$ distances animated from days -1 through 9 of the experiment.
-Source was the variable that explained the most observed variation across fecal communities (PERMANOVA source R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`, Table S11) followed by interactions between cage and day of the experiment.
+Source was the variable that explained the most observed variation across fecal communities (PERMANOVA source R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`, Table S11) followed by interactions between cage effects and day of the experiment.
 Transparency of the symbol corresponds to the day of the experiment, each symbol represents a sample from an individual mouse at a specific timepoint.
@@
 
-**Table S5. OTUs with relative abudances that significantly vary between sources at baseline (day -1).**
+**Table S5. OTUs with relative abumdances that significantly vary between sources at baseline (day -1).**
 
@@
 
-**Table S8. OTUs with relative abudances that significantly vary between sources post-clindamycin (day 0).**
+**Table S8. OTUs with relative abundances that significantly vary between sources post-clindamycin (day 0).**
 
-**Table S9. OTUs with relative abudances that significantly changed after clindamycin treatment.**
+**Table S9. OTUs with relative abundances that significantly changed after clindamycin treatment.**
 
-**Table S10. OTUs with relative abudances that significantly vary between sources post-infection (day 1).**
+**Table S10. OTUs with relative abundances that significantly vary between sources 1-day post-infection.**
 
@@
 
-**Table S15. OTUs with relative abundances that significantly varied across sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**
+**Table S15. OTUs with relative abundances that significantly varied between sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**

### 2020-08-11 submission/manuscript.tex
Incorporate Pat's edits

@@
 mechanistic work examining CDIs in mouse models use animals obtained
-from a single university colony or vendor. We treated mice from 6
-different sources (2 University of Michigan colonies and 4 vendors) with
-a single clindamycin dose, followed by a \emph{C. difficile} challenge 1
-day later and then measured \emph{C. difficile} colonization levels
-through 9 days post-infection. The microbiota were profiled via 16S rRNA
-gene sequencing to examine the variation across sources and alterations
-due to clindamycin treatment and \emph{C. difficile} challenge. While
-all sources of mice were colonized 1-day post-infection, variation
-emerged from days 3-7 post-infection with animals from some sources
-colonized with \emph{C. difficile} for longer and at higher levels. We
-identified bacteria that varied in relative abundance across sources and
-throughout the experiment. Some bacteria were consistently impacted by
-clindamycin treatment in all sources of mice including
-\emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and
+from a single source. We treated mice from 6 sources (2 University of
+Michigan colonies and 4 vendors) with 1 clindamycin dose, followed by
+\emph{C. difficile} challenge 1 day later and then measured \emph{C.
+difficile} colonization levels through 9 days post-infection. The
+microbiota were profiled via 16S rRNA gene sequencing to examine the
+variation across sources and alterations due to clindamycin treatment
+and \emph{C. difficile} challenge. While all mice were colonized 1-day
+post-infection, variation emerged from days 3-7 post-infection with
+animals from some sources colonized with \emph{C. difficile} for longer
+and at higher levels. We identified bacteria that varied in relative
+abundance across sources and throughout the experiment. Some bacteria
+were consistently impacted by clindamycin treatment in all sources of
+mice including \emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and
 \emph{Enterobacteriaceae}. To identify bacteria that were most important
@@
 post-clindamycin, and 1-day post-infection community composition data.
-With these models, we identified 4 bacteria that varied across sources
+With these models, we identified 4 bacteria that were predictive of
+whether \emph{C. difficile} cleared. They varied across sources
 (\emph{Bacteroides}), were altered by clindamycin
@@
 there is variation in who becomes asymptomatically colonized, develops
-an infection, or has an infection with adverse outcomes. \emph{C.
-difficile} infection (CDI) mouse models are widely used to answer a
-variety of \emph{C. difficile} pathogenesis questions. However, the
+an infection, or has adverse infection outcomes. Mouse models of
+\emph{C. difficile} infection (CDI) are widely used to answer a variety
+of \emph{C. difficile} pathogenesis questions. However, the
 inter-individual variation between mice from the same breeding facility
@@
 after exposure (1, 2). Additionally, asymptomatic colonization, where
-\emph{C. difficile} is detectable, but symptoms are absent has been
+\emph{C. difficile} is detectable, but symptoms are absent, has been
 documented in infants and adults (3, 4). The intestinal microbiota has
@@
 mice (Fig. 1A-B and Tables S1-2). Next, we compared the community
-structure of mice (Fig. 1C). Interactions between the source and cage
-effects explained most of the observed variation between fecal
-communities (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
-\textless{} 0.001, Fig. 1C and Table S3). Mice that are co-housed tend
-to have similar gut microbiotas due to coprophagy (41). Since mice
-within the same source were housed together, it is not surprising that
-the cage effect also contributed to the observed microbiota variation.
-There were some differences between the 2 experiments we conducted, as
-the experiment and cage effects significantly explained the observed
-community variation for the Schloss and Young lab mouse colonies (Fig.
-S2A-B and Table S4). However, most of the vendors also clustered by
-experiment (Fig. S2C-D, F), suggesting there was some community
-variation between the 2 experiments within each source, particularly for
-Schloss, Young, and Envigo mice (Fig. S2G-H). After finding differences
-at the community level, we next identified the bacteria that varied
-between sources of mice. There were 268 OTUs with relative abundances
-that were significantly different between the sources (Fig. 1D and Table
-S5). Though we saw differences between experiments at the community
-level, there were no OTUs that were significantly different between
-experiments within Schloss, Young, and Envigo mice (Data not shown). By
-using mice from six sources we were able to increase the number of
-microbiota communities to treat with clindamycin and challenge with
-\emph{C. difficile}.
+structure of mice (Fig. 1C). The source of mice and the interactions
+between the source and cage effects explained most of the observed
+variation between fecal communities (PERMANOVA combined
+R\textsuperscript{2} = 0.90, \emph{P} \textless{} 0.001, Fig. 1C and
+Table S3). Mice that are co-housed tend to have similar gut microbiotas
+due to coprophagy (41). Since mice within the same source were housed
+together, it was not surprising that the cage effect also contributed to
+the observed community variation. There were some differences between
+the 2 experiments we conducted, as the experiment and cage effects
+significantly explained the observed community variation for the Schloss
+and Young lab mouse colonies (Fig. S2A-B and Table S4). However, most of
+the vendors also clustered by experiment (Fig. S2C-D, F), suggesting
+there was some community variation between the 2 experiments within each
+source, particularly for Schloss, Young, and Envigo mice (Fig. S2G-H).
+After finding differences at the community level, we next identified the
+bacteria that varied between sources of mice. There were 268 OTUs with
+relative abundances that were significantly different between the
+sources (Fig. 1D and Table S5). Though we saw differences between
+experiments at the community level, there were no OTUs that were
+significantly different between experiments within Schloss, Young, and
+Envigo mice (all \emph{P} \textgreater{} 0.05). By using mice from six
+sources we were able to increase the variation in the starting
+communities to evaluate in a clindamycin-based CDI model.
 
@@
 model (14). We have previously demonstrated mice are rendered
-susceptible to \emph{C. difficile}, but cleared the pathogen within 9
-days when treated with clindamycin alone (21, 43). All mice were treated
-with 10 mg/kg clindamycin via intraperitoneal injection and one day
-later challenged with 10\textsuperscript{3} \emph{C. difficile} 630
-spores (Fig. 2A). The day after infection, \emph{C. difficile} was
-detectable in all mice at a similar level (median CFU range:
-2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15), indicating
-clindamycin rendered all mice susceptible regardless of source (Fig.
-2B). However, between 3 and 7 days post-infection, we observed variation
-in \emph{C. difficile} levels across sources of mice (all
-\emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 2B and Table S6). This
-suggested the mouse source was associated with \emph{C. difficile}
-clearance. While the colonization dynamics were similar between the two
-experiments, the Schloss mice took longer to clear in the 1st experiment
-and the Envigo mice took longer to clear in the 2nd experiment (Fig.
-S2A-B). The change in the mice's weight significantly varied across
-sources of mice with the most weight lost two days post-infection (Fig.
-2C and Table S7). There was also one Jackson and one Envigo mouse that
-died between 1- and 3-days post-infection during the second experiment.
-Mice obtained from Jackson, Taconic, and Envigo tended to lose more
-weight, have higher \emph{C. difficile} CFU levels and take longer to
-clear the infection compared to the other sources of mice (although
-there was variation between experiments with Schloss and Envigo mice).
-This was particularly evident 7 days post-infection (Fig. 2B-C, Fig.
-S2C-D), when 57\% of the mice were still colonized with \emph{C.
-difficile} (Fig. S2E). By 9 days post-infection the majority of the mice
-from all sources had cleared \emph{C. difficile} (Fig. 2C) with the
-exception of 1 Taconic mouse from the first experiment and 2 Envigo mice
-from the second experiment. Thus, clindamycin rendered all mice
+susceptible to \emph{C. difficile}, but clear the pathogen within 9 days
+when treated with clindamycin alone (21, 43). All mice were treated with
+10 mg/kg clindamycin via intraperitoneal injection and one day later
+challenged with 10\textsuperscript{3} \emph{C. difficile} 630 spores
+(Fig. 2A). The day after infection, \emph{C. difficile} was detectable
+in all mice at a similar level (median CFU range: 2.2e+07-1.3e+08;
+\emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
+mice susceptible regardless of source (Fig. 2B). However, between 3 and
+7 days post-infection, we observed variation in \emph{C. difficile}
+levels across sources of mice (all \emph{P}\textsubscript{FDR} \(\le\)
+0.019; Fig. 2B and Table S6). This suggested the mouse source was
+associated with \emph{C. difficile} clearance. While the colonization
+dynamics were similar between the two experiments, the Schloss mice took
+longer to clear in the 1st experiment compared to the 2nd and the Envigo
+mice took longer to clear in the 2nd experiment compared to the 1st
+(Fig. S2A-B). The change in the mice's weight significantly varied
+across sources of mice with the most weight loss occurring two days
+post-infection (Fig. 2C and Table S7). There was also one Jackson and
+one Envigo mouse that died between 1- and 3-days post-infection during
+the second experiment. Mice obtained from Jackson, Taconic, and Envigo
+tended to lose more weight, have higher \emph{C. difficile} CFU levels
+and take longer to clear the infection compared to the other sources of
+mice (although there was variation between experiments with Schloss and
+Envigo mice). This was particularly evident 7 days post-infection (Fig.
+2B-C, Fig. S2C-D), when 57\% of the mice were still colonized with
+\emph{C. difficile} (Fig. S2E). By 9 days post-infection the majority of
+the mice from all sources had cleared \emph{C. difficile} (Fig. 2B) with
+the exception of 1 Taconic mouse from the first experiment and 2 Envigo
+mice from the second experiment. Thus, clindamycin rendered all mice
 susceptible to \emph{C. difficile} 630 colonization, regardless of
@@
 subset of bacterial differences across sources persists.} Given the
-variation in mouse microbiotas that we observed across breeding
+variation in fecal communities that we observed across breeding
 colonies, we hypothesized that variation in \emph{C. difficile}
-clearance would be explained by microbiota variation across the 6
-sources of mice. As expected, clindamycin treatment decreased the
-richness and Shannon diversity across all sources of mice (Fig. 3A-B).
-Interestingly, significant differences in diversity metrics between
-sources (\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after
-clindamycin treatment, with Charles River mice having higher richness
-and Shannon diversity than most of the other sources (Fig 3A-B and
-Tables S1-2). The clindamycin treatment decreased the variation in
-community structures between sources of mice. Source and cage effects
-explained almost all of the observed variation between communities
-(combined R\textsuperscript{2} = 0.99, \emph{P} \textless{} 0.001; Fig.
-3C and Table S3). However, there were only 18 OTUs (Fig. 3D and Table
-S8) with relative abundances that significantly varied between sources.
-Next we identified the bacteria that shifted after clindamycin
-treatment, regardless of source by analyzing paired fecal samples from
-mice that were collected at baseline and after clindamycin treatment. We
-identified 153 OTUs that were altered after clindamycin treatment in
-most mice (Fig. 3E and Table S9). When we compared the list of
-significant clindamycin impacted bacteria with the bacteria that varied
-between sources post-clindamycin, we found 4 OTUs
+clearance would be explained by community variation across the 6 sources
+of mice. As expected, clindamycin treatment decreased the richness and
+Shannon diversity across all sources of mice (Fig. 3A-B). Interestingly,
+significant differences in diversity metrics between sources
+(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after clindamycin
+treatment, with Charles River mice having higher richness and Shannon
+diversity than most of the other sources (Fig 3A-B and Tables S1-2). The
+clindamycin treatment decreased the variation in community structures
+between sources of mice. The source of mice and the interactions between
+source and cage effects explained almost all of the observed variation
+between communities (combined R\textsuperscript{2} = 0.99, \emph{P}
+\textless{} 0.001; Fig. 3C and Table S3). However, there were only 18
+OTUs (Fig. 3D and Table S8) with relative abundances that significantly
+varied between sources. Next, we identified the bacteria that shifted
+after clindamycin treatment, regardless of source by analyzing paired
+fecal samples from mice that were collected at baseline and after
+clindamycin treatment. We identified 153 OTUs that were altered after
+clindamycin treatment in most mice (Fig. 3E and Table S9). When we
+compared the list of significant clindamycin impacted bacteria with the
+bacteria that varied between sources post-clindamycin, we found 4 OTUs
 (\emph{Enterobacteriaceae (OTU 1)}, \emph{Lachnospiraceae} (OTU 130),
@@
 in Taconic mice. \emph{Enterococcus} was primarily found only in mice
-purchased from commercial vendors and also increased after clindamycin
-treatment (Fig. 3D). These findings demonstrate that clindamycin had a
-consistent impact on the fecal bacterial communities of mice from all
-sources and only a subset of the OTUs continued to vary between sources.
+purchased from commercial vendors and also increased in relative
+abundance after clindamycin treatment (Fig. 3D). These findings
+demonstrate that clindamycin had a consistent impact on the fecal
+bacterial communities of mice from all sources and only a subset of the
+OTUs continued to vary between sources.
 
@@
 (\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Tables
-S1-2). Although the Charles River mice had more diverse microbiotas and
-were also able to clear \emph{C. difficile} faster than some of the
-other sources. Microbiota diversity did not explain the observed
-variation in \emph{C. difficile} colonization across sources considering
-the Young and Schloss mice had the lowest diversity after clindamycin
-treatment and were able to clear \emph{C. difficile} earlier than
-Jackson, Taconic and Envigo mice. Source and the interactions between
-source and cage effects continued to explain most of the observed
-community variation (combined R\textsuperscript{2} = 0.88; \emph{P}
-\textless{} 0.001; Fig. 4D and Table S3). One day after \emph{C.
-difficile} challenge, there were 44 OTUs (Fig. 4D and Table S10) with
-significantly different relative abundances across sources.
-
-Throughout the experiment the source of mice continued to be the
+S1-2). Although the Charles River mice had more diverse communities and
+were also able to clear \emph{C. difficile} faster than the other
+sources, diversity did not explain the observed variation in \emph{C.
+difficile} colonization across sources. The Young and Schloss mice had
+the lowest diversity 1 day post-infection and were able to clear
+\emph{C. difficile} earlier than Jackson, Taconic and Envigo mice. The
+source of mice and the interactions between source and cage effects
+continued to explain most of the observed community variation (combined
+R\textsuperscript{2} = 0.88; \emph{P} \textless{} 0.001; Fig. 4C and
+Table S3). One day after \emph{C. difficile} challenge, there were 44
+OTUs (Fig. 4D and Table S10) with significantly different relative
+abundances across sources.
+
+Throughout the experiment, the source of mice continued to be the
 dominant factor that explained the observed variation across fecal
@@
 experiment. By 7 days post-infection, when approximately 43\% mice had
-cleared \emph{C. difficile}, most of the mice still had not recovered to
-their baseline community structure (Fig. 4E). Distance from the baseline
-community between sources did not explain the variation in \emph{C.
-difficile} clearance since Schloss and Young mice cleared \emph{C.
-difficile} faster, but their communities were a greater distance from
-baseline 7 days post-infection. In summary, mouse bacterial communities
-varied significantly between sources throughout the course of the
-experiment and a consistent subset of bacterial taxa remained different
-between sources regardless of clindamycin and \emph{C. difficile}
-challenge.
+cleared \emph{C. difficile}, most of the mice had not recovered to their
+baseline community structure (Fig. 4E). The distance to the baseline
+community did not explain the variation in \emph{C. difficile} clearance
+as the Schloss and Young mice had mostly cleared \emph{C. difficile},
+but their communities were a greater distance from baseline 7 days
+post-infection compared to the Jackson and Taconic mice that were still
+colonized. In summary, mouse bacterial communities varied significantly
+between sources throughout the course of the experiment and a consistent
+subset of bacteria remained different between sources regardless of
+clindamycin and \emph{C. difficile} challenge.
 
@@
 post-infection.} After identifying taxa that varied between sources,
-changed after clindamycin treatment, or both, we next wanted to
-determine which taxa were influencing the variation in \emph{C.
-difficile} colonization at day 7 (Fig. 2B, Fig. S2C). We trained three
-L2-regularized logistic regression models with input bacterial community
-data from the baseline (day = -1), post-clindamycin (day = 0), and
-post-infection (day = 1) timepoints of the experiment to predict
-\emph{C. difficile} colonization status on day 7 (Fig. S3A-B). All
-models were better at predicting \emph{C. difficile} colonization status
-on day 7 than random chance (all \emph{P} \textless{} 0.001, Table S12).
-The model based on the post-clindamycin (day 0) community OTU data
-performed significantly better than the other models with an area under
-the receiving operator characteristic curve (AUROC) of 0.78
-(\emph{P}\textsubscript{FDR} \textless{} 0.001 for pairwise comparisons,
-Table S13). Thus, we were able to use bacterial relative abundance data
-from the time of \emph{C. difficile} challenge to differentiate mice
-that had cleared \emph{C. difficile} before day 7 from the mice still
-colonized with \emph{C. difficile} at that timepoint. This result
-suggests the bacterial community's response to clindamycin treatment had
-the greatest influence on subsequent \emph{C. difficile} colonization
-dynamics.
+changed after clindamycin treatment, or both, we determined which taxa
+were influencing the variation in \emph{C. difficile} colonization at
+day 7 (Fig. 2B, Fig. S2C). We trained three L2-regularized logistic
+regression models with input bacterial community data from the baseline
+(day = -1), post-clindamycin (day = 0), and post-infection (day = 1)
+timepoints of the experiment to predict \emph{C. difficile} colonization
+status on day 7 (Fig. S3A-B). All models were better at predicting
+\emph{C. difficile} colonization status on day 7 than random chance (all
+\emph{P} \textless{} 0.001, Table S12). The model based on the
+post-clindamycin (AUROC = 0.78) community OTU data performed
+significantly better than the baseline (AUROC = 0.72) or the
+post-infection (AUROC = 0.67) models (\emph{P}\textsubscript{FDR}
+\textless{} 0.001 for pairwise comparisons, Fig. S3C and Table S13).
+Thus, we were able to use bacterial relative abundance data from the
+time of \emph{C. difficile} challenge to differentiate mice that had
+cleared \emph{C. difficile} before day 7 from the mice still colonized
+with \emph{C. difficile} at that timepoint. This result suggests the
+bacterial community's response to clindamycin treatment had the greatest
+influence on subsequent \emph{C. difficile} colonization dynamics.
 
@@
 Next, we examined whether any of the top 20 ranked OTUs from the
-baseline (day 0) model were also important in the other 2 classification
-models based on baseline (day -1) and 1 day post-infection community
-data. We identified 6 OTUs (Enterobacteriaceae, Ruminococcaceae,
-Lactobacillus, Bacteroides, Porphyromonadaceae, Erysipelotrichaceae)
-that were important to the day 0 model and either the baseline or 1 day
-post-infection models (Table S14). Thus, a subset of bacterial OTUs were
-important for determining \emph{C. difficile} colonization dynamics
-across multiple timepoints.
+post-clindamycin (day 0) model were also important in the other 2
+classification models based on baseline (day -1) and 1 day
+post-infection community data. We identified 6 OTUs
+(\emph{Enterobacteriaceae, Ruminococcaceae, Lactobacillus, Bacteroides,
+Porphyromonadaceae, Erysipelotrichaceae}) that were important to the
+post-clindamycin model and either the baseline or 1 day post-infection
+models (Table S14). Thus, a subset of bacterial OTUs were important for
+determining \emph{C. difficile} colonization dynamics across multiple
+timepoints.
 
@@
 varied between sources, were altered by clindamycin treatment, or both,
-we examined whether the top 20 ranked OTUs from each model overlapped
-with OTUs that varied between sources (Fig. 1D, 3D, 4D and Tables S5,
-S8, S10) or were impacted by clindamycin treatment (Fig. 3E and Table
-S9). We found a subset of OTUs that were important to the baseline (day
--1), post-clindamycin (day 0), and 1 day post-infection models and
-overlapped with bacteria that varied between sources, were altered by
-clindamycin treatment, or both (Fig. S4). Combining the overall
-comparison results for the 3 models identified 14 OTUs associated with
-source, 21 OTUs associated with clindamycin treatment, and 6 OTUs
-associated with both (Fig. 5B). Several OTUs (\emph{Bacteroides,
-Enterococcus, Enterobacteriaceae, Porphyromonadaceae}) that overlapped
-with our previous analyses appeared across at least 2 models, so we
-examined how the relative abundances of these OTUs varied over the
-course of the experiment (Fig. 6). Throughout the experiment, there was
-at least 1 timepoint where relative abundances of these OTUs
-significantly varied between sources (Table S15). Interestingly, there
-were no OTUs that emerged as consistently enriched or depleted in mice
-that were colonized past 7 days post-infection, suggesting multiple
-bacteria influence \emph{C. difficile} colonization dynamics. Together,
-these results suggest the initial bacterial communities and their
-responses to clindamycin influence the clearance of \emph{C. difficile}.
+we identified the OTUs from each model that varied between sources (Fig.
+1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin
+treatment (Fig. 3E and Table S9; Fig. S4). Comparing the features
+important to the 3 models identified 14 OTUs associated with source, 21
+OTUs associated with clindamycin treatment, and 6 OTUs associated with
+both (Fig. 5B). Together, these results suggest that the initial
+bacterial communities and their responses to clindamycin influence the
+clearance of \emph{C. difficile}.
+
+Several OTUs (\emph{Bacteroides, Enterococcus, Enterobacteriaceae,
+Porphyromonadaceae}) that overlapped with our previous analyses appeared
+across at least 2 models, so we examined how the relative abundances of
+these OTUs varied over the course of the experiment (Fig. 6). Across the
+9 days post-infection, there was at least 1 timepoint when the relative
+abundances of these OTUs significantly varied between sources (Table
+S15). Interestingly, there were no OTUs that emerged as consistently
+enriched or depleted in mice that were colonized past 7 days
+post-infection, suggesting that multiple bacteria influence \emph{C.
+difficile} colonization dynamics.
 
@@
 
-By running our CDI model with mice from 6 different sources, we were
-able to identify bacterial taxa that were unique to sources throughout
-the experiment as well as taxa that were universally impacted by
-clindamycin. We trained L2 logistic regression models with baseline (day
--1), post-clindamycin treatment (day 0), and 1-day post-infection fecal
-community data that could predict whether mice cleared \emph{C.
-difficile} by 7 days post-infection better than random chance. We
-identified \emph{Bacteroides, Enterococcus, Enterobacteriaceae,
-Porphyromonadaceae} (Fig. 6) as candidate bacteria within these
-communities that were influencing variation in \emph{C. difficile}
-colonization dynamics since these bacteria were all important in the
-logistic regression models and varied by source, were impacted by
-clindamycin treatment, or both. Overall, our results demonstrated
-clindamycin was sufficient to render mice from multiple sources
-susceptible to CDI and only a subset of the interindividual microbiota
-variation across mice from different sources was associated with the
-time needed to clear \emph{C. difficile}.
+Applying our CDI model with mice from 6 different sources, allowed us to
+identify bacterial taxa that were unique to different sources as well as
+taxa that were universally impacted by clindamycin. We trained L2
+logistic regression models with baseline (day -1), post-clindamycin
+treatment (day 0), and 1-day post-infection fecal community data that
+could predict whether mice cleared \emph{C. difficile} by 7 days
+post-infection better than random chance. We identified
+\emph{Bacteroides, Enterococcus, Enterobacteriaceae, Porphyromonadaceae}
+(Fig. 6) as candidate bacteria within these communities that were
+influencing variation in \emph{C. difficile} colonization dynamics since
+these bacteria were all important in the logistic regression models and
+varied by source, were impacted by clindamycin treatment, or both.
+Overall, our results demonstrated clindamycin was sufficient to render
+mice from multiple sources susceptible to CDI and only a subset of the
+inter-individual microbiota variation across mice from different sources
+was associated with the time needed to clear \emph{C. difficile}.
 
@@
 or increase susceptibility to enteric infections (22, 23, 26--30). For
-example, in the context of \emph{Salmonella} infections,
-\emph{Enterobacteriaceae} and segmented filamentous bacteria have
-emerged as protective (22, 27). A previous study with \emph{C.
-difficile} identified an endogenous protective \emph{C. difficile}
-strain LEM1 that bloomed after antibiotic treatment in mice from Jackson
-or Charles River Laboratories, but not Taconic that protected mice
-against the more toxigenic \emph{C. difficile} VPI10463 (26). Given that
-we obtained mice from the same vendors, we checked all mice for
-endogenous \emph{C. difficile} by plating stool samples that were
-collected after clindamycin treatment. However, we did not identify any
-endogenous \emph{C. difficile} strains prior to challenge, suggesting
-there were no endogenous protective strains in the mice we received and
-other bacterial taxa mediated the variation in \emph{C. difficile}
-colonization across sources.
+example, against \emph{Salmonella} infections, \emph{Enterobacteriaceae}
+and segmented filamentous bacteria have emerged as protective (22, 27).
+We found \emph{Enterobacteriaceae} increased in all sources of mice
+after clindamycin treatment, facilitating \emph{C. difficile}
+colonization. However, there was also variation in
+\emph{Enterobacteriaceae} relative abundance levels between sources that
+was associated with the variation in \emph{C. difficile} colonization
+dynamics across sources. Thus, bacteria may have differential roles in
+determining susceptibility depending on the type of bacterial infection.
 
 Differences in CDI mouse model studies have been attributed to
-intestinal microbiota variation across sources. For example, groups
+intestinal microbiota variation across sources. For example, researchers
 using the same clindamycin treatment and C57BL/6 mice had different
@@
 colonization outcomes and overall susceptibility to \emph{C. difficile}
-after treatment with the same antibiotic (13, 31). The bacterial
-perturbations induced by clindamycin treatment have been well
-characterized and our findings agree with previous CDI mouse model work
-demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae} were
-associated with \emph{C. difficile} susceptibility and
+after treatment with the same antibiotic (13, 31). A previous study with
+\emph{C. difficile} identified an endogenous protective \emph{C.
+difficile} strain LEM1 that bloomed after antibiotic treatment in mice
+from Jackson or Charles River Laboratories, but not Taconic that
+protected mice against the more toxigenic \emph{C. difficile} VPI10463
+(26). Given that we obtained mice from the same vendors, we checked all
+mice for endogenous \emph{C. difficile} by plating stool samples that
+were collected after clindamycin treatment. However, we did not identify
+any endogenous \emph{C. difficile} strains prior to challenge,
+suggesting there were no endogenous protective strains in the mice we
+received and other bacteria mediated the variation in \emph{C.
+difficile} colonization across sources. The \emph{C. difficile} strain
+used could also be contributing to the variation in \emph{C. difficile}
+outcomes seen across different research groups. For example, a group
+found differential colonization outcomes after clindamycin treatment,
+with \emph{C. difficile} 630 and M68 infections eventually becoming
+undetectable while strain BI-7 remained detectable up to 70 days
+post-treatment (44).
+
+The bacterial perturbations induced by clindamycin treatment have been
+well characterized and our findings agree with previous CDI mouse model
+work demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae}
+were associated with \emph{C. difficile} susceptibility and
 \emph{Porpyhromonadaceae}, \emph{Lachnospiraceae},
@@
 treatment, there could be different outcomes for either susceptibility
-or clearance in the case of other antibiotic treatments. The \emph{C.
-difficile} strain used could also be contributing to the variation in
-\emph{C. difficile} outcomes seen across different research groups. For
-example, a group found differential colonization outcomes after
-clindamycin treatment, with \emph{C. difficile} 630 and M68 infections
-eventually becoming undectable while strain BI-7 remained detectable up
-to 70 days post-treatment (45). We found the time needed to naturally
-clear \emph{C. difficile} varied across sources of mice implying that at
-least in the context of the same perturbation, microbiota differences
-seemed to influence infection outcome more than susceptibility. More
-importantly, we were able to reduce the variation observed across
-sources to identify a subset of OTUs that were also important for
-predicting \emph{C. difficile} colonization status 7 days
-post-infection. Since all but 3 mice eventually cleared \emph{C.
-difficile} 630 by 9 days post-infection and the model built with the
-post-clindamycin (day 0) OTU relative abundance data had the best
-performance, our results suggest clindamycin treatment had a larger role
-in determining \emph{C. difficile} susceptibility and clearance than the
-source of the mice.
-
-Our approach successfully increased the diversity of murine bacterial
-communities tested in our clindamycin \emph{C. difficile} model. One
-alternative approach that has been used in some CDI studies (47--52) is
-to associate mice with human microbiotas. However, a major caveat to
-this method is the substantial loss of human microbiota community
-members upon transfer to mice (53, 54). Additionally with the exception
-of 2 recent studies (47, 48), most of the CDI mouse model studies to
-date associated mice with just 1 types of human microbiota either from a
-single donor or a single pool from multiple donors (49--52), which does
-not aid in the goal of modeling the interpersonal variation seen in
-humans to understand how the microbiota influences susceptiblity to CDIs
-and adverse outcomes. Importantly, our study using mice from 6 different
-sources increased the variation between groups of mice compared to using
-1 source alone, to better reflect the inter-individual microbiota
-variation observed in humans. Encouragingly, decreased
-\emph{Bifidobacterium}, \emph{Porphyromonas}, \emph{Ruminococcaceae} and
-\emph{Lachnospiraceae} and increased \emph{Enterobacteriaceae},
-\emph{Enterococcus}, \emph{Lactobacillus}, and \emph{Proteus} have all
-been associated with human CDIs (7) and were well represented in our
-study, suggesting most of the mouse sources are suitable for gaining
-insights into the bacteria influencing \emph{C. difficile} colonization
-and infections in humans. An important exception was
-\emph{Enterococcus}, which was primarily absent from University of
+or clearance in the case of other antibiotic treatments.
+
+We found the time needed to naturally clear \emph{C. difficile} varied
+across sources of mice implying that at least in the context of the same
+perturbation, microbiota differences seemed to influence infection
+outcome more than susceptibility. More importantly, we were able to
+reduce the variation observed across sources to identify a subset of
+OTUs that were also important for predicting \emph{C. difficile}
+colonization status 7 days post-infection. Since all but 3 mice
+eventually cleared \emph{C. difficile} 630 by 9 days post-infection and
+the model built with the post-clindamycin (day 0) OTU relative abundance
+data had the best performance, our results suggest clindamycin treatment
+had a larger role in determining \emph{C. difficile} susceptibility and
+clearance than the source of the mice.
+
+Our approach successfully increased the inter-animal variation tested
+with our clindamycin model of CDI. One alternative approach that has
+been used in some CDI studies (47--52) is to associate mice with human
+microbiotas. However, a major caveat to this method is the substantial
+loss of human microbiota community members upon transfer to mice (53,
+54). Additionally with the exception of 2 recent studies (47, 48), most
+of the CDI mouse model studies to date associated mice with just 1 types
+of human microbiota either from a single donor or a single pool from
+multiple donors (49--52), which does not aid in the goal of modeling the
+interpersonal variation seen in humans to understand how the microbiota
+influences susceptibility to CDIs and adverse outcomes. Importantly, our
+study using mice from 6 different sources increased the variation
+between groups of mice compared to using 1 source alone, to better
+reflect the inter-individual microbiota variation observed in humans.
+
+Another motivation for associating mice with human microbiotas,is to
+study the bacteria associated with the disease in humans. Encouragingly,
+decreased \emph{Bifidobacterium}, \emph{Porphyromonas},
+\emph{Ruminococcaceae} and \emph{Lachnospiraceae} and increased
+\emph{Enterobacteriaceae}, \emph{Enterococcus}, \emph{Lactobacillus},
+and \emph{Proteus} have all been associated with human CDIs (7) and were
+well represented in our study, suggesting most of the mouse sources are
+suitable for gaining insights into the bacteria influencing \emph{C.
+difficile} colonization and infections in humans. An important exception
+was \emph{Enterococcus}, which was primarily absent from University of
 Michigan colonies and \emph{Proteus}, which was only found in Taconic
-mice. Importantly, the fact that some CDI-associated bacteria were only
-found in a subset of mice has important implications for future CDI
-mouse model studies.
-
-There are several limitations to our work. The microbiota is composed of
-viruses, fungi, and parasites in addition to bacteria, and these
-non-bacterial members can also vary across sources of mice (55, 56).
-While our study focused solely on the bacterial portion, viruses and
-fungi have also begun to be implicated in the context of CDIs or FMT
-treatments for recurrent CDIs (35, 57--60). Beyond community
+mice. The fact that some CDI-associated bacteria were only found in a
+subset of mice has important implications for future CDI mouse model
+studies.
+
+Other microbiota and host factors that were outside the scope of our
+current study may also contribute to the differences in \emph{C.
+difficile} colonization dynamics between sources of mice. The microbiota
+is composed of viruses, fungi, and parasites in addition to bacteria,
+and these non-bacterial members can also vary across sources of mice
+(55, 56). While our study focused solely on the bacterial portion,
+viruses and fungi have also begun to be implicated in the context of
+CDIs or FMT treatments for recurrent CDIs (35, 57--60). Beyond community
 composition, the metabolic function of the microbiota also has a CDI
@@
 bile acids and butyrate production, have been implicated as important
-contributors to \emph{C. difficile} resistance (33, 45). Although, we
-only looked at composition, \emph{Ruminococcaceae} and
-\emph{Lachnospiraceae} both emerged as important taxa for classifying
-day 7 \emph{C. difficile} colonization status and metagenomes from these
-bacteria have been shown to contain the bile acid-inducible gene cluster
-necessary for secondary bile acid formation and ability to produce
-butyrate (50, 64). Interestingly, butyrate has previously been shown to
-vary across mouse vendors and mediated resistance to \emph{Citrobacter
-rodentium} infection, a model of enterohemorrhagic and enteropathogenic
-\emph{Escherichia coli} infections (23). Evidence for immunological
-toning differences in IgA and Th17 cells across mice from different
-vendors have also been documented and (65, 66) could influence the host
-response to CDI (67, 68). The outcome after \emph{C. difficile} exposure
-depends on a multitude of factors, including age, diet, and immunity;
-all of which are also influenced by the microbiota.
+contributors to \emph{C. difficile} resistance (33, 44). Interestingly,
+butyrate has previously been shown to vary across mouse vendors and
+mediated resistance to \emph{Citrobacter rodentium} infection, a model
+of enterohemorrhagic and enteropathogenic \emph{Escherichia coli}
+infections (23). Evidence for immunological toning differences in IgA
+and Th17 cells across mice from different vendors have also been
+documented and (64, 65) could influence the host response to CDI (66,
+67). The outcome after \emph{C. difficile} exposure depends on a
+multitude of factors, including age, diet, and immunity; all of which
+are also influenced by the microbiota.
 
 We have demonstrated that the ways baseline microbiotas from different
-mouse sources respond to clindamycin treatment influences the length of
-time mice remained colonized with \emph{C. difficile} 630. For those
-interested in dissecting the contribution of the microbiota to \emph{C.
-difficile} pathogenesis and treatments, using multiple sources of mice
-may yield more insights than a single model alone. Furthermore, for
-studies wanting to examine the interplay between a particular bactera
-such as \emph{Enterococcus} and \emph{C. difficile}, these results could
-serve as a resource for selecting which mice to order to address the
-question. Using mice from multiple sources helps model the interpersonal
-microbiota variation among humans to aid our understanding of how the
-gut microbiota contributes to CDIs.
+mouse sources respond to clindamycin treatment influence the length of
+time mice remained colonized with \emph{C. difficile} 630. To better
+understand the contribution of the microbiota to \emph{C. difficile}
+pathogenesis and treatments, using multiple sources of mice may yield
+more insights than a single source. Furthermore, for studies wanting to
+examine the interplay between particular bacteria such as
+\emph{Enterococcus} and \emph{C. difficile}, these results could serve
+as a resource for selecting mice to address their question. Using mice
+from multiple sources helps model the interpersonal microbiota variation
+among humans to aid our understanding of how the gut microbiota provides
+colonization resistance to CDIs.
 
@@
 (UL1TR002240). We thank members of the Schloss lab for feedback on
-planning the experiments and data presentation, as well as code
-tutorials and feedback through Code Club. In particular, we want to
-thank Begüm Topçuoğlu for help with implementing L2 logistic regression
-models using her
-\href{https://github.com/SchlossLab/ML_pipeline_microbiome}{pipeline},
-Ana Taylor for help with media preparation and sample collection, and
-Nicholas Lesniak for his critical feedback on the manuscript. We also
-thank members of Vincent Young's lab, particularly Kimberly Vendrov, for
-guidance with the \emph{C. difficile} infection mouse model and donating
-the mice. We also want to thank the Unit for Laboratory Animal Medicine
-at the University of Michigan for maintaining our mouse colony and
-providing the institutional support for our mouse experiments. Finally,
-we thank Kwi Kim, Austin Campbell, and Kimberly Vendrov for their help
-in maintaining the Schloss lab's anaerobic chamber.
+planning the experiments and data presentation. In particular, we thank
+Begüm Topçuoğlu for help with implementing L2 logistic regression
+models, Ana Taylor for help with media preparation and sample
+collection, and Nicholas Lesniak for his critical feedback on the
+manuscript. We also thank members of Vincent Young's lab, particularly
+Kimberly Vendrov, for guidance with the \emph{C. difficile} infection
+mouse model and donating the mice. We also thank the Unit for Laboratory
+Animal Medicine at the University of Michigan for maintaining our mouse
+colony and providing the institutional support for our mouse
+experiments. Finally, we thank Kwi Kim, Austin Campbell, and Kimberly
+Vendrov for their help in maintaining the Schloss lab's anaerobic
+chamber.
 
@@
 Fisher Scientific) using custom barcoded primers, as previously
-described (69). The ZymoBIOMICS microbial community DNA standards was
-used as a mock community control (70) and water was used as a negative
+described (68). The ZymoBIOMICS microbial community DNA standards was
+used as a mock community control (69) and water was used as a negative
 control per 96-well extraction plate. The PCR amplicons were cleaned up
@@
 \textbf{(v) 16S rRNA gene sequence analysis.} mothur (v. 1.43) was used
-to process all sequences (71) with a previously published protocol (69).
-Reads were combined and aligned with the SILVA reference database (72).
+to process all sequences (70) with a previously published protocol (68).
+Reads were combined and aligned with the SILVA reference database (71).
 Chimeras were removed with the VSEARCH algorithm and taxonomic
 assignment was completed with a modified version (v16) of the Ribosomal
-Database Project reference database (v11.5) (73) with an 80\% cutoff.
-Operational taxonomic units (OTUs) were assigned with a 97\% similarity
-threshold using the opticlust algorithm (74). To account for uneven
-sequencing across samples, samples were rarefied to 5,437 sequences
-1,000 times for alpha and beta diversity analyses. PCoAs were generated
-based on \(\theta_{YC}\) distances. Permutational multivariate analysis
-of variance (PERMANOVA) was performed on mothur-generated
-\(\theta_{YC}\) distance matrices with the adonis function in the vegan
-package (75) in R (76).
+Database Project reference database (v11.5) (72) with an 80\% confidence
+cutoff. Operational taxonomic units (OTUs) were assigned with a 97\%
+similarity threshold using the opticlust algorithm (73). To account for
+uneven sequencing across samples, samples were rarefied to 5,437
+sequences 1,000 times for alpha and beta diversity analyses, and a
+single time to generate relative abundances for model training. PCoAs
+were generated based on \(\theta_{YC}\) distances. Permutational
+multivariate analysis of variance (PERMANOVA) was performed on
+mothur-generated \(\theta_{YC}\) distance matrices with the adonis
+function in the vegan package (74) in R (75).
 
@@
 post-infection. The L2-regularized logistic regression models were
-trained and tested using the caret package (77) in R as previously
-described (78) with the exception that we used 60\% training and 40\%
+trained and tested using the caret package (76) in R as previously
+described (77) with the exception that we used 60\% training and 40\%
 testing data splits for the cross-validation of the training data to
@@
 \textbf{(vii) Statistical analysis.} All statistical tests were
-performed in R (v 4.0.2) (76). The Kruskal-Wallis test was used to
+performed in R (v 4.0.2) (75). The Kruskal-Wallis test was used to
 analyze differences in \emph{C. difficile} CFU, mouse weight change, and
@@
 followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg
-correction. OTUs impacted by clindamycin treatment were identified using
-the Wilcoxon signed rank test with matched pairs of mice samples for day
--1 and day 0. To determine whether classification models had better
+correction. The Wilcoxon rank sum test was used to test for OTUs that
+differed between experiments within the Schloss, Young, and Envigo
+sources with Benjamini-Hochberg correction for testing all identified
+OTUs. OTUs impacted by clindamycin treatment were identified using the
+paired Wilcoxon signed rank test with matched pairs of mice samples for
+day -1 and day 0. To determine whether classification models had better
 performance (test AUROCs) than random chance (0.5), we used the
@@
 testing. The tidyverse package (v 1.3.0) was used to wrangle and graph
-data (79).
+data (78).
 
@@
 
+\leavevmode\hypertarget{ref-Lawley2012}{}%
+44. Lawley TD, Clare S, Walker AW, Stares MD, Connor TR, Raisen C,
+Goulding D, Rad R, Schreiber F, Brandt C, Deakin LJ, Pickard DJ, Duncan
+SH, Flint HJ, Clark TG, Parkhill J, Dougan G. 2012. Targeted restoration
+of the intestinal microbiota with a simple, defined bacteriotherapy
+resolves relapsing \emph{Clostridium difficile} disease in mice. PLoS
+Pathogens 8:e1002995.
+
 \leavevmode\hypertarget{ref-Lawley2009}{}%
-44. Lawley TD, Clare S, Walker AW, Goulding D, Stabler RA, Croucher N,
+45. Lawley TD, Clare S, Walker AW, Goulding D, Stabler RA, Croucher N,
 Mastroeni P, Scott P, Raisen C, Mottram L, Fairweather NF, Wren BW,
@@
 and Immunity 77:3661--3669.
-
-\leavevmode\hypertarget{ref-Lawley2012}{}%
-45. Lawley TD, Clare S, Walker AW, Stares MD, Connor TR, Raisen C,
-Goulding D, Rad R, Schreiber F, Brandt C, Deakin LJ, Pickard DJ, Duncan
-SH, Flint HJ, Clark TG, Parkhill J, Dougan G. 2012. Targeted restoration
-of the intestinal microbiota with a simple, defined bacteriotherapy
-resolves relapsing \emph{Clostridium difficile} disease in mice. PLoS
-Pathogens 8:e1002995.
 
@@
 
-\leavevmode\hypertarget{ref-Vital2019}{}%
-64. Vital M, Rud T, Rath S, Pieper DH, Schlüter D. 2019. Diversity of
-bacteria exhibiting bile acid-inducible 7alpha-dehydroxylation genes in
-the human gut. Computational and Structural Biotechnology Journal
-17:1016--1019.
-
 \leavevmode\hypertarget{ref-Fransen2015}{}%
-65. Fransen F, Zagato E, Mazzini E, Fosso B, Manzari C, Aidy SE,
+64. Fransen F, Zagato E, Mazzini E, Fosso B, Manzari C, Aidy SE,
 Chiavelli A, D'Erchia AM, Sethi MK, Pabst O, Marzano M, Moretti S,
@@
 \leavevmode\hypertarget{ref-Ivanov2009}{}%
-66. Ivanov II, Atarashi K, Manel N, Brodie EL, Shima T, Karaoz U, Wei D,
+65. Ivanov II, Atarashi K, Manel N, Brodie EL, Shima T, Karaoz U, Wei D,
 Goldfarb KC, Santee CA, Lynch SV, Tanoue T, Imaoka A, Itoh K, Takeda K,
@@
 \leavevmode\hypertarget{ref-Azrad2018}{}%
-67. Azrad M, Hamo Z, Tkhawkho L, Peretz A. 2018. Elevated serum
+66. Azrad M, Hamo Z, Tkhawkho L, Peretz A. 2018. Elevated serum
 immunoglobulin a levels in patients with \emph{Clostridium difficile}
@@
 \leavevmode\hypertarget{ref-Saleh2019}{}%
-68. Saleh MM, Frisbee AL, Leslie JL, Buonomo EL, Cowardin CA, Ma JZ,
+67. Saleh MM, Frisbee AL, Leslie JL, Buonomo EL, Cowardin CA, Ma JZ,
 Simpson ME, Scully KW, Abhyankar MM, Petri WA. 2019. Colitis-induced
@@
 \leavevmode\hypertarget{ref-Kozich2013}{}%
-69. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
+68. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \leavevmode\hypertarget{ref-Sze2019}{}%
-70. Sze MA, Schloss PD. 2019. The impact of DNA polymerase and number of
+69. Sze MA, Schloss PD. 2019. The impact of DNA polymerase and number of
 rounds of amplification in PCR on 16S rRNA gene sequence data. mSphere
@@
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-71. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+70. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
@@
 \leavevmode\hypertarget{ref-Quast2012}{}%
-72. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
+71. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
 J, Glöckner FO. 2012. The SILVA ribosomal RNA gene database project:
@@
 \leavevmode\hypertarget{ref-Cole2013}{}%
-73. Cole JR, Wang Q, Fish JA, Chai B, McGarrell DM, Sun Y, Brown CT,
+72. Cole JR, Wang Q, Fish JA, Chai B, McGarrell DM, Sun Y, Brown CT,
 Porras-Alfaro A, Kuske CR, Tiedje JM. 2013. Ribosomal database project:
@@
 \leavevmode\hypertarget{ref-Westcott2017}{}%
-74. Westcott SL, Schloss PD. 2017. OptiClust, an improved method for
+73. Westcott SL, Schloss PD. 2017. OptiClust, an improved method for
 assigning amplicon-based sequence data to operational taxonomic units.
@@
 \leavevmode\hypertarget{ref-Vegan2018}{}%
-75. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+74. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 Minchin PR, O'Hara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
@@
 \leavevmode\hypertarget{ref-r_citation_2018}{}%
-76. R Core Team. 2018. R: A language and environment for statistical
+75. R Core Team. 2018. R: A language and environment for statistical
 computing. R Foundation for Statistical Computing, Vienna, Austria.
@@
 \leavevmode\hypertarget{ref-Kuhn2008}{}%
-77. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
+76. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
 Journal of Statistical Software 28.
@@
 \leavevmode\hypertarget{ref-Topcuoglu2020}{}%
-78. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
+77. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
 framework for effective application of machine learning to
@@
 \leavevmode\hypertarget{ref-Tidyverse2019}{}%
-79. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
+78. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
 Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
@@
 value for a stool sample from an individual mouse, circles represent
-experiment 1 mice and triangles represent experiment 2 mice. D. Plots
-highlighting the median (point) and interquantile range (colored lines)
-of the relative abundances for the top 20 bacteria out of the 268 OTUs
-that varied across sources at baseline (Table S5).
+experiment 1 mice and triangles represent experiment 2 mice. D. The
+median (point) and interquantile range (colored lines) of the relative
+abundances for the 20 most significant OTUs out of the 268 OTUs that
+varied across sources at baseline (Table S5).
 
@@
 0.05 (Table S1). Significant \emph{P} values from the pairwise Wilcoxon
-comparisons between sources with Benjamini-Hochberg correction are shown
-(Table S2). C. PCoA of \(\theta_{YC}\) distances from stools collected
-post-clindamycin. Source and the interaction between source and cage
-effects explained most of the variation observed post-clindamycin
-(PERMANOVA combined R\textsuperscript{2} = 0.99, \emph{P} \textless{}
-0.001, see Table S3). For A-C, each symbol represents a stool sample
-from an individual mouse, with circles representing experiment 1 mice
-and triangles representing experiment 2 mice. D. Plots highlighting the
-median (point) and interquantile range (colored lines) of the relative
-abundances for the 18 OTUs (Table S8)that varied between sources after
-clindamycin treatment (day 0). E. Plots highlighting the median (point)
-and interquantile range (colored lines) of the top 10 OTUs out of 153
-with relative abundances that changed after clindamycin treatment
-(adjusted \emph{P} value \textless{} 0.05). Data were analyzed by
-Wilcoxon signed rank test of mice that had paired sequence data for
-baseline (day -1) and post-clindamycin (day 0) timepoints (N = 31), with
-Benjamini-Hochberg correction for testing all identified OTUs (Table
-S9). The gray vertical line indicates the limit of detection.
+comparisons between sources with Benjamini-Hochberg correction are
+displayed as the first initial of each group compared to the group that
+they are listed above (Table S2). C. PCoA of \(\theta_{YC}\) distances
+from stools collected post-clindamycin. Source and the interaction
+between source and cage effects explained most of the variation observed
+post-clindamycin (PERMANOVA combined R\textsuperscript{2} = 0.99,
+\emph{P} \textless{} 0.001, see Table S3). For A-C, each symbol
+represents a stool sample from an individual mouse, with circles
+representing experiment 1 mice and triangles representing experiment 2
+mice. D. The median (point) and interquantile range (colored lines) of
+the relative abundances for the 18 OTUs (Table S8) that varied between
+sources after clindamycin treatment (day 0). E. The median (point) and
+interquantile range (colored lines) of the top 10 most significant OTUs
+out of 153 with relative abundances that changed because of the
+clindamycin treatment (adjusted \emph{P} value \textless{} 0.05). Data
+were analyzed by paired Wilcoxon signed rank test of mice that had
+paired sequence data for baseline (day -1) and post-clindamycin (day 0)
+timepoints (N = 31), with Benjamini-Hochberg correction for testing all
+identified OTUs (Table S9). The gray vertical line indicates the limit
+of detection.
 
@@
 0.05 (Table S1). Significant \emph{P} values from the pairwise Wilcoxon
-comparisons between sources with Benjamini-Hochberg correction are shown
-(Table S2). PCoA of \(\theta_{YC}\) distances of 1-day post-infection
-stool samples. Source and the interaction between source and cage
-effects explained most of the variation between fecal communities
-(PERMANOVA combined R\textsuperscript{2} = 0.88, \emph{P} \textless{}
-0.001, Table S3). For A-C: each symbol represents the value for a stool
-sample from an individual mouse, circles represent experiment 1 mice and
-triangles represent experiment 2 mice. D. Plots highlighting the median
-(point) and interquantile range (colored lines) of the relative
-abundances for the top 20 bacteria out of the 44 OTUs that varied
-between sources 1-day post-infection. The gray vertical line indicates
-the limit of detection. For each timepoint OTUs with differential
-relative abundances across sources of mice were identified by
-Kruskal-Wallis test with Benjamini-Hochberg correction for testing all
-identified OTUs (Table S10). E. \(\theta_{YC}\) distances of fecal
+comparisons between sources with Benjamini-Hochberg correction are
+displayed as the first initial of each group compared to the group that
+they are listed above (Table S2). PCoA of \(\theta_{YC}\) distances of
+1-day post-infection stool samples. Source and the interaction between
+source and cage effects explained most of the variation between fecal
+communities (PERMANOVA combined R\textsuperscript{2} = 0.88, \emph{P}
+\textless{} 0.001, Table S3). For A-C: each symbol represents the value
+for a stool sample from an individual mouse, circles represent
+experiment 1 mice and triangles represent experiment 2 mice. D. The
+median (point) and interquantile range (colored lines) of the relative
+abundances for the top 20 most significant OTUs out of the 44 OTUs that
+varied between sources 1-day post-infection. The gray vertical line
+indicates the limit of detection. For each timepoint OTUs with
+differential relative abundances across sources of mice were identified
+by Kruskal-Wallis test with Benjamini-Hochberg correction for testing
+all identified OTUs (Table S10). E. \(\theta_{YC}\) distances of fecal
 samples collected 7-days post-infection relative to the baseline (day
@@
 \includegraphics{figure_5.pdf} \textbf{Figure 5. Bacteria that influence
-whether mice cleared \emph{C. difficile} by day 7.} A. Baseline (day 0)
-relative abundance data for the 10 OTUs with the highest rankings based
-on feature weights in the baseline (day 0) classification model. Red
-font represents OTUs that correlated with \emph{C. difficile}
-colonization and blue font represents OTUs that correlated with
-clearance. Symbols represent the relative abundance data for an
-individual mouse. Gray lines indicate the median relative abundances for
-each source. B. Venn diagram that combines Fig. S4 summaries of OTUs
-that were important to the day -1, 0, and 1 classification models (Table
+whether mice cleared \emph{C. difficile} by day 7.} A. Post-clindamycin
+(day 0) relative abundance data for the 10 OTUs with the highest
+rankings based on feature weights in the post-clindamycin (day 0)
+classification model. Red font represents OTUs that correlated with
+\emph{C. difficile} colonization and blue font represents OTUs that
+correlated with clearance. Symbols represent the relative abundance data
+for an individual mouse. Gray bars indicate the median relative
+abundances for each source. B. Venn diagram that combines OTUs that were
+important to the day -1, 0, and 1 classification models ( Fig. S4, Table
 S14) and either overlapped with taxa that varied across sources at the
-same timepoint, were impacted by clindamycin treatment, or both. Red
+same timepoint, were impacted by clindamycin treatment, or both. Bold
 OTUs were important to more than 1 classification model.
@@
 \emph{C. difficile} colonization dynamics vary across sources throughout
-the experiment.} A-D. Relative abundances of red OTUs from Fig. 5A that
-were important for at least two classification models are shown over
+the experiment.} A-D. Relative abundances of bold OTUs from Fig. 5B that
+were important in at least two classification models are shown over
 time. A. \emph{Bacteroides} (OTU 2), which varied across sources
@@
 post-infection. All models performed better than random chance (AUROC =
-0.5, all \emph{P} \(\le\) 1.6e-17, Table S12) and the model built with
-post-clindamycin treated bacterial OTU relative abundances had the best
-performance ((\emph{P}\textsubscript{FDR} \(\le\) 3.3e-06 for all
-pairwise comparisons, Table S13). For list of the 20 OTUs that were
-ranked as most important to each model, see Table S14.
+0.5, all \emph{P} \textless{} 0.001, Table S12) and the model built with
+post-clindamycin bacterial OTU relative abundances had the best
+performance ((\emph{P}\textsubscript{FDR} \textless{} 0.001 for all
+pairwise comparisons, Table S13). See Table S14 for list of the 20 OTUs
+that were ranked as most important to each model.
 
@@
 classification models based on baseline, post-clindamycin treatment, or
-post-infection community data vary by source, clindamycin treatment, or
-both.} A-C. Venn diagrams of OTUs from the top 20 OTUs from the baseline
-(A), post-clindamycin treatment (B), and post-infection (C)
-classification models (Table S14) that overlapped with OTUs that varied
-across sources at the corresponding timepoint (Tables S5, 8, 10), were
-impacted by clindamycin treatment (Table S9), or both. Red OTUs signify
-OTUs that were important to more than 1 classification model.
+1-day post-infection community data vary by source, clindamycin
+treatment, or both.} A-C. Venn diagrams of OTUs from the top 20 OTUs
+from the baseline (A), post-clindamycin treatment (B), and 1-day
+post-infection (C) classification models (Table S14) that overlapped
+with OTUs that varied across sources at the corresponding timepoint
+(Tables S5, 8, 10), were impacted by clindamycin treatment (Table S9),
+or both. Bold OTUs were important to more than 1 classification model.
 
@@
 
-\textbf{Movie S1. Large shifts in bacterial community structure occurred
-after clindamycin and \emph{C. difficile} infection.} PCoA of
+\textbf{Movie S1. Large shifts in bacterial community structures
+occurred after clindamycin and \emph{C. difficile} infection.} PCoA of
 \(\theta_{YC}\) distances animated from days -1 through 9 of the
@@
 R\textsuperscript{2} = 0.35, \emph{P} = 0.0001, Table S11) followed by
-interactions between cage and day of the experiment. Transparency of the
-symbol corresponds to the day of the experiment, each symbol represents
-a sample from an individual mouse at a specific timepoint. Circles
-represent mice from experiment 1 and triangles represent mice from
-expeirment 2.
+interactions between cage effects and day of the experiment.
+Transparency of the symbol corresponds to the day of the experiment,
+each symbol represents a sample from an individual mouse at a specific
+timepoint. Circles represent mice from experiment 1 and triangles
+represent mice from expeirment 2.
 
@@
 
-\textbf{Table S5. OTUs with relative abudances that significantly vary
+\textbf{Table S5. OTUs with relative abumdances that significantly vary
 between sources at baseline (day -1).}
@@
 
-\textbf{Table S8. OTUs with relative abudances that significantly vary
+\textbf{Table S8. OTUs with relative abundances that significantly vary
 between sources post-clindamycin (day 0).}
 
-\textbf{Table S9. OTUs with relative abudances that significantly
+\textbf{Table S9. OTUs with relative abundances that significantly
 changed after clindamycin treatment.}
 
-\textbf{Table S10. OTUs with relative abudances that significantly vary
-between sources post-infection (day 1).}
+\textbf{Table S10. OTUs with relative abundances that significantly vary
+between sources 1-day post-infection.}
 
@@
 \textbf{Table S15. OTUs with relative abundances that significantly
-varied across sources of mice on at least 1 day of the experiment by
+varied between sources of mice on at least 1 day of the experiment by
 Kruskal-Wallis test.}

### 2020-08-20 code/figure_3.R
Additional figure edits

@@ -6,9 +6,9 @@ c <- ggdraw() + draw_image("results/figures/pcoa_day0.png")
 left_panel <- plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=1)
 d <- ggdraw() + draw_image("results/figures/D0top18_otus.png")
 e <- ggdraw() + draw_image("results/figures/clind_impacted_otus_plot.png")
+right_panel <- plot_grid(d, e, labels = c("D", "E"), label_size = 12, ncol=2, rel_widths = c(1, 1)) 
+right_panel <- plot_grid(right_panel, NULL, labels = NULL, label_size = 12, nrow=2, rel_heights = c(1, 0.5)) #Add NULL plot to fix margin spacing in final figure
 
-plot_grid(left_panel, d, e, labels = c("", "D", "E"), label_size = 12, ncol=3)+
-  ggsave("results/figures/figure_3.pdf", width=8, height=4.25)+
-  ggsave("submission/figure_3.pdf", width=8, height=4.25)
-
-
+plot_grid(left_panel, right_panel, labels = c("", "D", "E"), label_size = 12, ncol=2, axis = "t", align = "v", rel_heights = c(1, 3), rel_widths = c(1, 2))+
+  ggsave("results/figures/figure_3.pdf", width=8, height=6)+
+  ggsave("submission/figure_3.pdf", width=8, height=6)

### 2020-08-20 code/taxa_stats.R

@@ -404,6 +404,7 @@ otu_over_time <- function(otu_plot, x_annotation, y_position, label){
     annotate("text", y = y_position, x = x_annotation, label = label, size =8)+
     theme_classic()+
     theme(plot.title=element_markdown(hjust = 0.5),
+          panel.grid.minor.x = element_line(size = 0.4, color = "grey"),  # Add gray lines to clearly separate symbols by days)
           text = element_text(size = 14)) # Change font size for entire plot
 }
 

### 2020-08-20 code/venn_diagram_comparisons.R

@@ -105,7 +105,7 @@ venn_otus <- function(source_comp, clind_comp, title){
   #Make data frame to annotate number of unique and overlapping OTUs onto venn diagram plot
   df_venn_otus <- as.data.frame(otus) %>%
     mutate(x = c(-15, 0, 15),
-           y = c(5, 5, 5))
+           y = c(4, 4, 4))
   #List of OTUs that are unique to each group or overlap to add to the plot
   source_unique_taxa <- sort(setdiff(source_comp, clind_comp))  
   # intersection
@@ -147,7 +147,7 @@ venn_otus <- function(source_comp, clind_comp, title){
     scale_colour_manual(values = c('cornflowerblue', 'firebrick', 'gold'), guide = FALSE) +
     labs(fill = NULL) +
     annotate("text", x = df_venn_otus$x, y = df_venn_otus$y, label = df_venn_otus[,1], size = 5)+
-    annotate("text", x = c(-10, 10), y = c(10, 10), label = c("Source", "Clindamycin"), size = 5)+
+    annotate("text", x = c(-10, 10), y = c(10.5, 10.5), label = c("Source", "Clindamycin"), size = 5)+
     annotate(geom='richtext', label = source_overlap, x = -16.5, y = 0.5, size = 2.7, fill = NA, label.color = NA, color = "black")+
     annotate(geom='richtext', label = source_unique, x = -19, y = -13, size = 2.7, fill = NA, label.color = NA, color = "grey27")+
     annotate(geom='richtext', label = clind_overlap, x = 16, y = 0.5, size = 2.7, fill = NA, label.color = NA, color = "black")+

### 2020-08-20 code/tables_supplemental.R
Fix typo and regenerate tables

@@ -41,7 +41,7 @@ weight_kruskal_stats_adjust <- read_tsv("data/process/weight_stats_all_days.tsv"
 kw_w_sig_d0 <- otu_stats_dn1to1_combined %>% filter(day == 0)
 
 #Table S9: OTUs that changed after clindamycin treatment----
-o_dn1to0_pairs_stats_adjust <- read_tsv(path = "data/process/otu_stats_dn1to0.tsv") 
+o_dn1to0_pairs_stats_adjust <- read_tsv("data/process/otu_stats_dn1to0.tsv") 
 
 #Table S10: OTUs that vary between sources on day 1----
 kw_w_sig_d1 <- otu_stats_dn1to1_combined %>% filter(day == 1)

### 2020-08-20 submission/manuscript.Rmd
Additional edits

@@
 However, much of the mechanistic work examining CDIs in mouse models use animals obtained from a single source.
-We treated mice from 6 sources (2 University of Michigan colonies and 4 vendors) with 1 clindamycin dose, followed by *C.
-difficile* challenge 1 day later and then measured *C.
-difficile* colonization levels through 9 days post-infection.
+We treated mice from 6 sources (2 University of Michigan colonies and 4 commercial vendors) with clindamycin, followed by a *C.
+difficile* challenge and then measured *C.
+difficile* colonization levels throughout the infection.
 The microbiota were profiled via 16S rRNA gene sequencing to examine the variation across sources and alterations due to clindamycin treatment and *C.
@@
 To identify bacteria that were most important to colonization regardless of the source, we created logistic regression models that successfully classified mice based on whether they cleared *C.
-difficile* by 7 days post-infection using baseline, post-clindamycin, and 1-day post-infection community composition data.
+difficile* by 7 days post-infection using community composition data at baseline, post-clindamycin, and 1-day post-infection.
 With these models, we identified 4 bacteria that were predictive of whether *C.
@@
 They varied across sources (*Bacteroides*), were altered by clindamycin (*Porphyromonadaceae*), or both (*Enterobacteriaceae* and *Enterococcus*).
-Microbiota variation across sources better emulates human inter-individual variation and can help identify bacterial drivers of phenotypic variation in the context of CDIs.
+Allowing for microbiota variation across sources better emulates human inter-individual variation and can help identify bacterial drivers of phenotypic variation in the context of CDIs.
 
@@
 
-Antibiotics are a common risk factor for *Clostridioides difficile* infections (CDIs) due to there effect on the intestinal microbiota, but there is variation in who goes on to develop severe or recurrent CDIs after exposure [@Teng2019; @Kelly2012].
+Antibiotics are a common risk factor for *Clostridioides difficile* infections (CDIs) due to their effect on the intestinal microbiota, but there is variation in who goes on to develop severe or recurrent CDIs after exposure [@Teng2019; @Kelly2012].
 Additionally, asymptomatic colonization, where *C.
@@
 However, variation in the microbiota between mice from the same breeding colony is much less than the inter-individual variation observed between humans [@Baxter2014; @Nagpal2018].
-Studying CDIs in mice with a homogenous microbiota is likely to overstate the importance of individual mechanisms.
-Using mice that have a more heterogenous microbiota would allow researchers to identify and validate more generalizable mechanisms responsible for CDI.
+Studying CDIs in mice with a homogeneous microbiota is likely to overstate the importance of individual mechanisms.
+Using mice that have a more heterogeneous microbiota would allow researchers to identify and validate more generalizable mechanisms responsible for CDI.
 
@@
 The differences between the microbiota of mice from vendors have been well documented and shown to influence susceptibility to a variety of diseases [@Stough2016; @Alegre2019], including enteric infections [@EtienneMesmin2017; @Velazquez2019; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018].
-Different research groups have also observed variation in CDI outcomes despite using similar murine models [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014].
-Here we examined how variations in the baseline microbiota and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C.
+Different research groups have also observed different CDI outcomes despite using similar murine models [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014].
+Here we examined how variation in the baseline microbiota and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C.
 difficile* colonization and the time needed to clear the infection.
@@
 
-**The variation in the microbiota is high between mice from different sources.** We obtained C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other in 2010 (the Young and Schloss lab colonies) and four commercial sources: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan).
+**The variation in the microbiota is high between mice from different sources.** We obtained C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other in 2010 (the Young and Schloss lab colonies) and four commercial vendors: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan).
 These 4 vendors were chosen because they are commonly used for murine CDI studies [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019].
@@
 There was a significant difference in the richness (i.e.
-number of observed operational taxonomic units (OTUs)), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_richness` and *P*~FDR~ = `r padj_dn1_shannon`, respectively) across the sources of mice (Fig.
+number of observed operational taxonomic units (OTUs)), but not Shannon diversity index across the sources of mice (*P*~FDR~ = `r padj_dn1_richness` and *P*~FDR~ = `r padj_dn1_shannon`, respectively; Fig.
 1A-B and Tables S1-2).
@@
 1C).
-The source of mice and the interactions between the source and cage effects explained most of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, Fig.
+The source of mice and the interactions between the source and cage effects explained most of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig.
 1C and Table S3).
@@
 2B and Table S6).
-This suggested the mouse source was associated with *C.
+This suggested the source of mice was associated with *C.
 difficile* clearance.
-While the colonization dynamics were similar between the two experiments, the Schloss mice took longer to clear in the 1st experiment compared to the 2nd and the Envigo mice took longer to clear in the 2nd experiment compared to the 1st (Fig.
+While the colonization dynamics were similar between the two experiments, the Schloss mice took longer to clear in the first experiment compared to the second and the Envigo mice took longer to clear in the second experiment compared to the first (Fig.
 S2A-B).
@@
 By 9 days post-infection the majority of the mice from all sources had cleared *C.
-difficile* (Fig.
-2B) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment.
+difficile* with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment  (Fig.
+2B).
 Thus, clindamycin rendered all mice susceptible to *C.
@@
 3A-B).
-Interestingly, significant differences in diversity metrics between sources (*P*~FDR~ < 0.05) emerged after clindamycin treatment, with Charles River mice having higher richness and Shannon diversity than most of the other sources (Fig 3A-B and Tables S1-2).
+Interestingly, significant differences in diversity metrics between sources emerged after clindamycin treatment, with Charles River mice having higher richness and Shannon diversity than most of the other sources (*P*~FDR~ < 0.05; Fig 3A-B and Tables S1-2).
 The clindamycin treatment decreased the variation in community structures between sources of mice.
@@
 3C and Table S3).
-However, there were only `r source_otus_d0` OTUs (Fig.
-3D and Table S8) with relative abundances that significantly varied between sources.
+However, there were only `r source_otus_d0` OTUs with relative abundances that significantly varied between sources (Fig.
+3D and Table S8).
 Next, we identified the bacteria that shifted after clindamycin treatment, regardless of source by analyzing paired fecal samples from mice that were collected at baseline and after clindamycin treatment.
@@
 3E and Table S9).
-When we compared the list of significant clindamycin impacted bacteria with the bacteria that varied between sources post-clindamycin, we found 4 OTUs (*Enterobacteriaceae (OTU 1)*, *Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) overlapped (Fig.
+When we compared the list of significant clindamycin impacted bacteria with the bacteria that varied between sources post-clindamycin, we found 4 OTUs that were shared between the lists (*Enterobacteriaceae (OTU 1)*, *Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23); Fig.
 3D-E and Tables S8-9).
@@
 One day after *C.
-difficile* challenge, there were `r source_otus_d1` OTUs (Fig.
-4D and Table S10) with significantly different relative abundances across sources.
+difficile* challenge, there were `r source_otus_d1` OTUs with significantly different relative abundances across sources (Fig.
+4D and Table S10).
 
 Throughout the experiment, the source of mice continued to be the dominant factor that explained the observed variation across fecal communities (PERMANOVA R^2^ = `r rsq_adonis_source`, *P* `r format.pval(p_adonis_source, eps = 0.001)`) followed by interactions between cage effects  and the day of the experiment (Movie S1 and Table S11).
-Mice fecal samples from the same source of mice continued to cluster closely to each other throughout the experiment.
+Fecal samples from the same source of mice continued to cluster closely to each other throughout the experiment.
 By 7 days post-infection, when approximately `r format(100-percent_colonized_d7, digits = 1)`% mice had cleared *C.
@@
 S2C).
-We trained three L2-regularized logistic regression models with input bacterial community data from the baseline (day = -1), post-clindamycin (day = 0), and post-infection (day = 1) timepoints of the experiment to predict *C.
+We trained three L2-regularized logistic regression models with either input bacterial community data from the baseline (day = -1), post-clindamycin (day = 0), or post-infection (day = 1) timepoints of the experiment to predict *C.
 difficile* colonization status on day 7 (Fig.
@@
 difficile* colonization status on day 7 than random chance (all *P* `r format.pval(max_p_vs_random, eps = 0.001)`, Table S12).
-The model based on the post-clindamycin (AUROC = `r median_otu_day0_AUROC`) community OTU data performed significantly better than the baseline (AUROC = `r median_otu_dayn1_AUROC`) or the post-infection (AUROC = `r median_otu_day1_AUROC`) models (*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for pairwise comparisons, Fig.
+The model based on the post-clindamycin (AUROC = `r median_otu_day0_AUROC`) community OTU data performed significantly better than the baseline (AUROC = `r median_otu_dayn1_AUROC`) or the post-infection (AUROC = `r median_otu_day1_AUROC`) models (*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for pairwise comparisons; Fig.
 S3C and Table S13).
@@
 difficile* at that timepoint.
-This result suggests the bacterial community's response to clindamycin treatment had the greatest influence on subsequent *C.
+This result suggests that the bacterial community's response to clindamycin treatment had the greatest influence on subsequent *C.
 difficile* colonization dynamics.
@@
 First, we looked at OTUs from the model with the best performance, which was based on the post-clindamycin treatment (day 0) bacterial community data.
-Out of the 10 highest ranked OTUs, 7 OTUs (*Bacteroides*, *Escherichia/Shigella*, 2 *Lachnospiraceae*,  *Lactobacillus*, *Porphyromonadaceae*, and *Ruminococcaceae*) were associated with *C.
-difficile* colonization 7 days post-infection, while 3 OTUs (*Enterobacteriaceae*, *Lachnospiraceae*, *Porphyromonadaceae*) were associated with clearance (Fig.
+Out of the 10 highest ranked OTUs, 7 OTUs were associated with *C.
+difficile* colonization 7 days post-infection (*Bacteroides*, *Escherichia/Shigella*, 2 *Lachnospiraceae*,  *Lactobacillus*, *Porphyromonadaceae*, and *Ruminococcaceae*), while 3 OTUs were associated with clearance (*Enterobacteriaceae*, *Lachnospiraceae*, *Porphyromonadaceae*; Fig.
 5A).
 Next, we examined whether any of the top 20 ranked OTUs from the post-clindamycin (day 0) model were also important in the other 2 classification models based on baseline (day -1) and 1 day post-infection community data.
-We identified 6 OTUs (*`r list_overlapping_across_models`*) that were important to the post-clindamycin model and either the baseline or 1 day post-infection models (Table S14).
+We identified 6 OTUs that were important to the post-clindamycin model and either the baseline or 1 day post-infection models (*`r list_overlapping_across_models`*; Table S14).
 Thus, a subset of bacterial OTUs were important for determining *C.
@@
 5B).
-Together, these results suggest that the initial bacterial communities and their responses to clindamycin influence the clearance of *C.
+Together, these results suggest that the initial bacterial communities and their responses to clindamycin influenced the clearance of *C.
 difficile*.
 
-Several OTUs (*`r list_key_otus`*) that overlapped with our previous analyses appeared across at least 2 models, so we examined how the relative abundances of these OTUs varied over the course of the experiment (Fig.
+Several OTUs that overlapped with our previous analyses appeared across at least 2 models (*`r list_key_otus`*), so we examined how the relative abundances of these OTUs varied over the course of the experiment (Fig.
 6).
@@
 
-Applying our CDI model with mice from 6 different sources, allowed us to identify bacterial taxa that were unique to different sources as well as taxa that were universally impacted by clindamycin.
-We trained L2 logistic regression models with baseline (day -1), post-clindamycin treatment (day 0), and 1-day post-infection fecal community data that could predict whether mice cleared *C.
+Applying our CDI model to 6 different sources of mice, allowed us to identify bacterial taxa that were unique to different sources as well as taxa that were universally impacted by clindamycin.
+We trained logistic regression models with baseline (day -1), post-clindamycin treatment (day 0), and 1-day post-infection fecal community data that could predict whether mice cleared *C.
 difficile* by 7 days post-infection better than random chance.
 We identified *`r list_key_otus`* (Fig.
-6) as candidate bacteria within these communities that were influencing variation in *C.
+6) as candidate bacteria within these communities that influenced variation in *C.
 difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by source, were impacted by clindamycin treatment, or both.
-Overall, our results demonstrated clindamycin was sufficient to render mice from multiple sources susceptible to CDI and only a subset of the inter-individual microbiota variation across mice from different sources was associated with the time needed to clear *C.
+Overall, our results demonstrated clindamycin was sufficient to render mice from multiple sources susceptible to CDI and only a subset of the inter-individual microbiota variation across mice from different sources was needed to predict which mice could clear *C.
 difficile*.
 
-Other studies have taken similar approaches by using mice from multiple sources to identify bacteria that either promote colonization resistance or increase susceptibility to enteric infections [@Velazquez2019; @EtienneMesmin2017; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018].
+Other studies have used mice from multiple sources to identify bacteria that either promote colonization resistance or increase susceptibility to enteric infections [@Velazquez2019; @EtienneMesmin2017; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018].
 For example, against *Salmonella* infections, *Enterobacteriaceae* and segmented filamentous bacteria have emerged as protective [@Velazquez2019; @Lai2020].
@@
 For example, researchers using the same clindamycin treatment and C57BL/6 mice had different *C.
-difficile* outcomes, one having sustained colonization [@Buffie2011], while the other had transient [@Reeves2011], despite both using *C.
+difficile* outcomes, one having sustained colonization [@Buffie2011], while the other had transient colonization [@Reeves2011], despite both using *C.
 difficile* VPI 10643.
@@
 difficile* after treatment with the same antibiotic [@Lawley2013; @Hutton2014].
+When we treated mice from 6 different sources with clindamycin and challenged them with *C.
+difficile* 630, we found microbiota variation across sources impacted colonization outcomes, but not susceptibility.
 A previous study with *C.
@@
 difficile* 630 and M68 infections eventually becoming undetectable while strain BI-7 remained detectable up to 70 days post-treatment [@Lawley2012].
-
 The bacterial perturbations induced by clindamycin treatment have been well characterized and our findings agree with previous CDI mouse model work demonstrating *Enterococcus* and *Enterobacteriaceae* were associated with *C.
@@
 We found the time needed to naturally clear *C.
-difficile* varied across sources of mice implying that at least in the context of the same perturbation, microbiota differences seemed to influence infection outcome more than susceptibility.
-More importantly, we were able to reduce the variation observed across sources to identify a subset of OTUs that were also important for predicting *C.
+difficile* varied across sources of mice implying that at least in the context of the same perturbation, microbiota differences influence infection outcome.
+More importantly, we were able to explain the variation observed across sources with a subset of OTUs that were also important for predicting *C.
 difficile* colonization status 7 days post-infection.
@@
 
-Our approach successfully increased the inter-animal variation tested with our clindamycin model of CDI.
-One alternative approach that has been used in some CDI studies [@NagaoKitamoto2020; @Battaglioli2018; @Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018] is to associate mice with human microbiotas.
+Using mice from multiple sources successfully increased the inter-animal variation.
+One alternative approach that has been used in some CDI studies is to associate mice with human microbiotas [@NagaoKitamoto2020; @Battaglioli2018; @Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018].
 However, a major caveat to this method is the substantial loss of human microbiota community members upon transfer to mice [@Fouladi2020; @Walter2020].
-Additionally with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of the CDI mouse model studies to date associated mice with just 1 types of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018], which does not aid in the goal of modeling the interpersonal variation seen in humans to understand how the microbiota influences susceptibility to CDIs and adverse outcomes.
+Additionally with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of these studies associated mice with just 1 types of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018].
+This approach does not aid in the goal of modeling the interpersonal variation seen in humans to understand how the microbiota influences susceptibility to CDIs and adverse outcomes.
 Importantly, our study using mice from 6 different sources increased the variation between groups of mice compared to using 1 source alone, to better reflect the inter-individual microbiota variation observed in humans.
 
-Another motivation for associating mice with human microbiotas,is to study the bacteria associated with the disease in humans.
-Encouragingly, decreased *Bifidobacterium*, *Porphyromonas*, *Ruminococcaceae* and *Lachnospiraceae* and increased *Enterobacteriaceae*, *Enterococcus*, *Lactobacillus*, and *Proteus* have all been associated with human CDIs [@Mancabelli2017, @Duvallet2017] and were well represented in our study, suggesting most of the mouse sources are suitable for gaining insights into the bacteria influencing *C.
+Another motivation for associating mice with human microbiotas is to study the bacteria associated with the disease in humans.
+Decreased *Bifidobacterium*, *Porphyromonas*, *Ruminococcaceae* and *Lachnospiraceae* and increased *Enterobacteriaceae*, *Enterococcus*, *Lactobacillus*, and *Proteus* have all been associated with human CDIs [@Mancabelli2017, @Duvallet2017].
+Encouragingly, these populations were well represented in our study, suggesting most of the mouse sources are suitable for gaining insights into the bacteria influencing *C.
 difficile* colonization and infections in humans.
 An important exception was *Enterococcus*, which was primarily absent from University of Michigan colonies and *Proteus*, which was only found in Taconic mice.
-The fact that some CDI-associated bacteria were only found in a subset of mice has important implications for future CDI mouse model studies.
+The fact that some CDI-associated bacteria were only found in a subset of mice has important implications for future CDI mouse model studies, but also models the natural patchiness of microbial populations in humans.
 
@@
 Furthermore, for studies wanting to examine the interplay between particular bacteria such as *Enterococcus* and *C.
-difficile*, these results could serve as a resource for selecting mice to address their question.
+difficile*, these results could serve as a resource for selecting mice to address the question.
 Using mice from multiple sources helps model the interpersonal microbiota variation among humans to aid our understanding of how the gut microbiota provides colonization resistance to CDIs.
@@
 This work was supported by the National Institutes of Health (U01AI124255).
-ST was supported by the Michigan Institute for Clincial and Health Research Postdoctoral Translation Scholars Program (UL1TR002240).
+ST was supported by the Michigan Institute for Clincial and Health Research Postdoctoral Translation Scholars Program (UL1TR002240 from the National Center for Advancing Translational Sciences).
 We thank members of the Schloss lab for feedback on planning the experiments and data presentation.
-In particular, we thank Begüm Topçuoğlu for help with implementing L2 logistic regression models, Ana Taylor for help with media preparation and sample collection, and Nicholas Lesniak for his critical feedback on the manuscript.
+In particular, we thank Begüm Topçuoğlu for help with implementing logistic regression models, Ana Taylor for help with media preparation and sample collection, and Nicholas Lesniak for his critical feedback on the manuscript.
 We also thank members of Vincent Young’s lab, particularly  Kimberly Vendrov, for guidance with the *C.
@@
 Female C57BL/7 mice were obtained from 6 different sources: The Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, Envigo, and two colonies at the University of Michigan (the Schloss lab colony and the Young lab colony).
-The Young lab colony was originally established with mice purchased from Jackson, and the Schloss lab colony was established in 2010 with mice donated from the Young lab.
+The Young lab colony was originally established with mice purchased from Jackson in 2002, and the Schloss lab colony was established in 2010 with mice donated from the Young lab.
 The 4 groups of mice purchased from vendors were allowed to acclimate to the University of Michigan mouse facility for 13 days prior to starting the experiment.
@@
 
-**(ii) Antibiotic treatment.** After the 13-day acclimation period and 1 day prior to challenge (Fig.
-1A), all mice received 10 mg/kg clindamycin (filter sterilized through a 0.22 micron syringe filter prior to administration) via intraperitoneal injection.
+**(ii) Antibiotic treatment.** After the 13-day acclimation period, all mice received 10 mg/kg clindamycin (filter sterilized through a 0.22 micron syringe filter prior to administration) via intraperitoneal injection (Fig.
+1A).
 
@@
 difficile* strain 630 via oral gavage post-infection 1 day after clindamycin treatment as described previously [@Jenior2018].
-Mice weights and stool samples were taken daily through 9 days post-challenge.
+Mice weights and stool samples were taken daily through 9 days post-challenge (Fig.
+1A).
 Collected stool was split for *C.
-difficile* CFU quantification and 16S rRNA sequencing analysis.
-*C.
-difficile* quantification stool samples were transferred to the anaerobic chamber, serially diluted in PBS, plated on taurocholate-cycloserine-cefoxitin-fructose agar (TCCFA) plates, and counted after 24 hours of incubation at 37&deg;C under anaerobic conditions.
+difficile* quantification and 16S rRNA sequencing analysis.
+For *C.
+difficile* quantification, stool samples were transferred to the anaerobic chamber, serially diluted in PBS, plated on taurocholate-cycloserine-cefoxitin-fructose agar (TCCFA) plates, and counted after 24 hours of incubation at 37&deg;C under anaerobic conditions.
 A sample from the day 0 timepoint (post-clindamycin and prior to *C.
@@
 **(vi) Classification model training and evaluation.** Models were generated based on mice that were categorized as either cleared or colonized 7 days post-infection and had sequencing data from the baseline (day -1), post-clindamycin (day 0), and post-infection (day 1) timepoints of the experiment.
-Input bacterial community relative abundance data at the OTU level from the baseline, post-clindamycin, and post-infection timepoints was used to generate 6 classification models that predicted *C.
+Input bacterial community relative abundance data at the OTU level from the baseline, post-clindamycin, and 1-day post-infection timepoints was used to generate 3 classification models that predicted *C.
 difficile* colonization status 7 days post-infection.
@@
 Code was modified from https://github.com/SchlossLab/ML_pipeline_microbiome to update the classification outcomes and change the data split ratios.
-The modified repository to regenerate this analysis is available at https://github.com/tomkoset/ML_pipeline_microbiome.
+The modified repository to regenerate our modeling analysis is available at https://github.com/tomkoset/ML_pipeline_microbiome.
 
@@
 The Wilcoxon rank sum test was used to test for OTUs that differed between experiments within the Schloss, Young, and Envigo sources with Benjamini-Hochberg correction for testing all identified OTUs.
-OTUs impacted by clindamycin treatment were identified using the paired Wilcoxon signed rank test with matched pairs of mice samples for day -1 and day 0.
+OTUs impacted by clindamycin treatment were identified using the paired Wilcoxon signed rank test with matched pairs of mice samples from day -1 and day 0.
 To determine whether classification models had better performance (test AUROCs) than random chance (0.5), we used the one-sample Wilcoxon signed rank test.
-To examine whether there was an overall difference in predictive performance across the 6 classification models we used the Kruskal-Wallis test followed by pairwise Wilcoxan comparisons with Benjamini-Hochberg correction for multiple hypothesis testing.
+To examine whether there was an overall difference in predictive performance across the 3 classification models we used the Kruskal-Wallis test followed by pairwise Wilcoxan comparisons with Benjamini-Hochberg correction for multiple hypothesis testing.
 The tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) was used to wrangle and graph data  [@Tidyverse2019].
@@
 Principal Coordinates Analysis (PCoA) of $\theta_{YC}$ distances of baseline stool samples.
-Source and the interaction between source and cage effects explained most of the variation (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, see Table S3).
+Source and the interaction between source and cage effects explained most of the variation (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Table S3).
 For A-C:  each symbol represents the value for a stool sample from an individual mouse, circles represent experiment 1 mice and triangles represent experiment 2 mice.
@@
 PCoA of $\theta_{YC}$ distances from stools collected post-clindamycin.
-Source and the interaction between source and cage effects explained most of the variation observed post-clindamycin (PERMANOVA combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, see Table S3).
+Source and the interaction between source and cage effects explained most of the variation observed post-clindamycin (PERMANOVA combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Table S3).
 For A-C, each symbol represents a stool sample from an individual mouse, with circles representing experiment 1 mice and triangles representing experiment 2 mice.
@@
 PCoA of $\theta_{YC}$ distances of 1-day post-infection stool samples.
-Source and the interaction between source and cage effects explained most of the variation between fecal communities (PERMANOVA combined R^2^ = `r d1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, Table S3).
+Source and the interaction between source and cage effects explained most of the variation between fecal communities (PERMANOVA combined R^2^ = `r d1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Table S3).
 For A-C:  each symbol represents the value for a stool sample from an individual mouse, circles represent experiment 1 mice and triangles represent experiment 2 mice.
@@
 **Figure 5.
-Bacteria that influence whether mice cleared *C.
+Bacteria that influenced whether mice cleared *C.
 difficile* by day 7.** A.
@@
 \includegraphics{figure_S1.pdf}
+
 **Figure S1.
@@
 \includegraphics{figure_S4.pdf}
+
 **Figure S4.

### 2020-08-20 submission/manuscript.tex
Additional edits

@@
 from a single source. We treated mice from 6 sources (2 University of
-Michigan colonies and 4 vendors) with 1 clindamycin dose, followed by
-\emph{C. difficile} challenge 1 day later and then measured \emph{C.
-difficile} colonization levels through 9 days post-infection. The
-microbiota were profiled via 16S rRNA gene sequencing to examine the
-variation across sources and alterations due to clindamycin treatment
-and \emph{C. difficile} challenge. While all mice were colonized 1-day
+Michigan colonies and 4 commercial vendors) with clindamycin, followed
+by a \emph{C. difficile} challenge and then measured \emph{C. difficile}
+colonization levels throughout the infection. The microbiota were
+profiled via 16S rRNA gene sequencing to examine the variation across
+sources and alterations due to clindamycin treatment and \emph{C.
+difficile} challenge. While all mice were colonized 1-day
 post-infection, variation emerged from days 3-7 post-infection with
@@
 models that successfully classified mice based on whether they cleared
-\emph{C. difficile} by 7 days post-infection using baseline,
-post-clindamycin, and 1-day post-infection community composition data.
-With these models, we identified 4 bacteria that were predictive of
-whether \emph{C. difficile} cleared. They varied across sources
+\emph{C. difficile} by 7 days post-infection using community composition
+data at baseline, post-clindamycin, and 1-day post-infection. With these
+models, we identified 4 bacteria that were predictive of whether
+\emph{C. difficile} cleared. They varied across sources
 (\emph{Bacteroides}), were altered by clindamycin
 (\emph{Porphyromonadaceae}), or both (\emph{Enterobacteriaceae} and
-\emph{Enterococcus}). Microbiota variation across sources better
-emulates human inter-individual variation and can help identify
+\emph{Enterococcus}). Allowing for microbiota variation across sources
+better emulates human inter-individual variation and can help identify
 bacterial drivers of phenotypic variation in the context of CDIs.
@@
 Antibiotics are a common risk factor for \emph{Clostridioides difficile}
-infections (CDIs) due to there effect on the intestinal microbiota, but
+infections (CDIs) due to their effect on the intestinal microbiota, but
 there is variation in who goes on to develop severe or recurrent CDIs
@@
 the inter-individual variation observed between humans (16, 17).
-Studying CDIs in mice with a homogenous microbiota is likely to
+Studying CDIs in mice with a homogeneous microbiota is likely to
 overstate the importance of individual mechanisms. Using mice that have
-a more heterogenous microbiota would allow researchers to identify and
+a more heterogeneous microbiota would allow researchers to identify and
 validate more generalizable mechanisms responsible for CDI.
@@
 enteric infections (22, 23, 26--30). Different research groups have also
-observed variation in CDI outcomes despite using similar murine models
-(13, 18, 21, 31--33). Here we examined how variations in the baseline
+observed different CDI outcomes despite using similar murine models (13,
+18, 21, 31--33). Here we examined how variation in the baseline
 microbiota and responses to clindamycin treatment in C57BL/6 mice from
@@
 other in 2010 (the Young and Schloss lab colonies) and four commercial
-sources: the Jackson Laboratory, Charles River Laboratories, Taconic
+vendors: the Jackson Laboratory, Charles River Laboratories, Taconic
 Biosciences, and Envigo (which was formerly Harlan). These 4 vendors
@@
 of observed operational taxonomic units (OTUs)), but not Shannon
-diversity index (\emph{P}\textsubscript{FDR} = 0.03 and
-\emph{P}\textsubscript{FDR} = 0.052, respectively) across the sources of
-mice (Fig. 1A-B and Tables S1-2). Next, we compared the community
-structure of mice (Fig. 1C). The source of mice and the interactions
-between the source and cage effects explained most of the observed
-variation between fecal communities (PERMANOVA combined
-R\textsuperscript{2} = 0.90, \emph{P} \textless{} 0.001, Fig. 1C and
-Table S3). Mice that are co-housed tend to have similar gut microbiotas
-due to coprophagy (41). Since mice within the same source were housed
-together, it was not surprising that the cage effect also contributed to
-the observed community variation. There were some differences between
-the 2 experiments we conducted, as the experiment and cage effects
-significantly explained the observed community variation for the Schloss
-and Young lab mouse colonies (Fig. S2A-B and Table S4). However, most of
-the vendors also clustered by experiment (Fig. S2C-D, F), suggesting
-there was some community variation between the 2 experiments within each
-source, particularly for Schloss, Young, and Envigo mice (Fig. S2G-H).
-After finding differences at the community level, we next identified the
-bacteria that varied between sources of mice. There were 268 OTUs with
-relative abundances that were significantly different between the
-sources (Fig. 1D and Table S5). Though we saw differences between
-experiments at the community level, there were no OTUs that were
-significantly different between experiments within Schloss, Young, and
-Envigo mice (all \emph{P} \textgreater{} 0.05). By using mice from six
-sources we were able to increase the variation in the starting
-communities to evaluate in a clindamycin-based CDI model.
+diversity index across the sources of mice (\emph{P}\textsubscript{FDR}
+= 0.03 and \emph{P}\textsubscript{FDR} = 0.052, respectively; Fig. 1A-B
+and Tables S1-2). Next, we compared the community structure of mice
+(Fig. 1C). The source of mice and the interactions between the source
+and cage effects explained most of the observed variation between fecal
+communities (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
+\textless{} 0.001; Fig. 1C and Table S3). Mice that are co-housed tend
+to have similar gut microbiotas due to coprophagy (41). Since mice
+within the same source were housed together, it was not surprising that
+the cage effect also contributed to the observed community variation.
+There were some differences between the 2 experiments we conducted, as
+the experiment and cage effects significantly explained the observed
+community variation for the Schloss and Young lab mouse colonies (Fig.
+S2A-B and Table S4). However, most of the vendors also clustered by
+experiment (Fig. S2C-D, F), suggesting there was some community
+variation between the 2 experiments within each source, particularly for
+Schloss, Young, and Envigo mice (Fig. S2G-H). After finding differences
+at the community level, we next identified the bacteria that varied
+between sources of mice. There were 268 OTUs with relative abundances
+that were significantly different between the sources (Fig. 1D and Table
+S5). Though we saw differences between experiments at the community
+level, there were no OTUs that were significantly different between
+experiments within Schloss, Young, and Envigo mice (all \emph{P}
+\textgreater{} 0.05). By using mice from six sources we were able to
+increase the variation in the starting communities to evaluate in a
+clindamycin-based CDI model.
 
@@
 levels across sources of mice (all \emph{P}\textsubscript{FDR} \(\le\)
-0.019; Fig. 2B and Table S6). This suggested the mouse source was
+0.019; Fig. 2B and Table S6). This suggested the source of mice was
 associated with \emph{C. difficile} clearance. While the colonization
 dynamics were similar between the two experiments, the Schloss mice took
-longer to clear in the 1st experiment compared to the 2nd and the Envigo
-mice took longer to clear in the 2nd experiment compared to the 1st
-(Fig. S2A-B). The change in the mice's weight significantly varied
-across sources of mice with the most weight loss occurring two days
-post-infection (Fig. 2C and Table S7). There was also one Jackson and
-one Envigo mouse that died between 1- and 3-days post-infection during
-the second experiment. Mice obtained from Jackson, Taconic, and Envigo
-tended to lose more weight, have higher \emph{C. difficile} CFU levels
-and take longer to clear the infection compared to the other sources of
-mice (although there was variation between experiments with Schloss and
-Envigo mice). This was particularly evident 7 days post-infection (Fig.
-2B-C, Fig. S2C-D), when 57\% of the mice were still colonized with
-\emph{C. difficile} (Fig. S2E). By 9 days post-infection the majority of
-the mice from all sources had cleared \emph{C. difficile} (Fig. 2B) with
-the exception of 1 Taconic mouse from the first experiment and 2 Envigo
-mice from the second experiment. Thus, clindamycin rendered all mice
-susceptible to \emph{C. difficile} 630 colonization, regardless of
-source, but there was significant variation in disease phenotype across
-the sources of mice.
+longer to clear in the first experiment compared to the second and the
+Envigo mice took longer to clear in the second experiment compared to
+the first (Fig. S2A-B). The change in the mice's weight significantly
+varied across sources of mice with the most weight loss occurring two
+days post-infection (Fig. 2C and Table S7). There was also one Jackson
+and one Envigo mouse that died between 1- and 3-days post-infection
+during the second experiment. Mice obtained from Jackson, Taconic, and
+Envigo tended to lose more weight, have higher \emph{C. difficile} CFU
+levels and take longer to clear the infection compared to the other
+sources of mice (although there was variation between experiments with
+Schloss and Envigo mice). This was particularly evident 7 days
+post-infection (Fig. 2B-C, Fig. S2C-D), when 57\% of the mice were still
+colonized with \emph{C. difficile} (Fig. S2E). By 9 days post-infection
+the majority of the mice from all sources had cleared \emph{C.
+difficile} with the exception of 1 Taconic mouse from the first
+experiment and 2 Envigo mice from the second experiment (Fig. 2B). Thus,
+clindamycin rendered all mice susceptible to \emph{C. difficile} 630
+colonization, regardless of source, but there was significant variation
+in disease phenotype across the sources of mice.
 
@@
 Shannon diversity across all sources of mice (Fig. 3A-B). Interestingly,
-significant differences in diversity metrics between sources
-(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after clindamycin
-treatment, with Charles River mice having higher richness and Shannon
-diversity than most of the other sources (Fig 3A-B and Tables S1-2). The
-clindamycin treatment decreased the variation in community structures
-between sources of mice. The source of mice and the interactions between
-source and cage effects explained almost all of the observed variation
-between communities (combined R\textsuperscript{2} = 0.99, \emph{P}
-\textless{} 0.001; Fig. 3C and Table S3). However, there were only 18
-OTUs (Fig. 3D and Table S8) with relative abundances that significantly
-varied between sources. Next, we identified the bacteria that shifted
-after clindamycin treatment, regardless of source by analyzing paired
-fecal samples from mice that were collected at baseline and after
-clindamycin treatment. We identified 153 OTUs that were altered after
-clindamycin treatment in most mice (Fig. 3E and Table S9). When we
-compared the list of significant clindamycin impacted bacteria with the
-bacteria that varied between sources post-clindamycin, we found 4 OTUs
-(\emph{Enterobacteriaceae (OTU 1)}, \emph{Lachnospiraceae} (OTU 130),
-\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) overlapped
-(Fig. 3D-E and Tables S8-9). Importantly, some of the OTUs that varied
-between sources also shifted with clindamycin treatment. For example,
-\emph{Proteus} increased after clindamycin treatment (Fig. 3D), but only
-in Taconic mice. \emph{Enterococcus} was primarily found only in mice
-purchased from commercial vendors and also increased in relative
-abundance after clindamycin treatment (Fig. 3D). These findings
-demonstrate that clindamycin had a consistent impact on the fecal
-bacterial communities of mice from all sources and only a subset of the
-OTUs continued to vary between sources.
+significant differences in diversity metrics between sources emerged
+after clindamycin treatment, with Charles River mice having higher
+richness and Shannon diversity than most of the other sources
+(\emph{P}\textsubscript{FDR} \textless{} 0.05; Fig 3A-B and Tables
+S1-2). The clindamycin treatment decreased the variation in community
+structures between sources of mice. The source of mice and the
+interactions between source and cage effects explained almost all of the
+observed variation between communities (combined R\textsuperscript{2} =
+0.99, \emph{P} \textless{} 0.001; Fig. 3C and Table S3). However, there
+were only 18 OTUs with relative abundances that significantly varied
+between sources (Fig. 3D and Table S8). Next, we identified the bacteria
+that shifted after clindamycin treatment, regardless of source by
+analyzing paired fecal samples from mice that were collected at baseline
+and after clindamycin treatment. We identified 153 OTUs that were
+altered after clindamycin treatment in most mice (Fig. 3E and Table S9).
+When we compared the list of significant clindamycin impacted bacteria
+with the bacteria that varied between sources post-clindamycin, we found
+4 OTUs that were shared between the lists (\emph{Enterobacteriaceae (OTU
+1)}, \emph{Lachnospiraceae} (OTU 130), \emph{Lactobacillus} (OTU 6),
+\emph{Enterococcus} (OTU 23); Fig. 3D-E and Tables S8-9). Importantly,
+some of the OTUs that varied between sources also shifted with
+clindamycin treatment. For example, \emph{Proteus} increased after
+clindamycin treatment (Fig. 3D), but only in Taconic mice.
+\emph{Enterococcus} was primarily found only in mice purchased from
+commercial vendors and also increased in relative abundance after
+clindamycin treatment (Fig. 3D). These findings demonstrate that
+clindamycin had a consistent impact on the fecal bacterial communities
+of mice from all sources and only a subset of the OTUs continued to vary
+between sources.
 
@@
 Table S3). One day after \emph{C. difficile} challenge, there were 44
-OTUs (Fig. 4D and Table S10) with significantly different relative
-abundances across sources.
+OTUs with significantly different relative abundances across sources
+(Fig. 4D and Table S10).
 
@@
 0.001) followed by interactions between cage effects and the day of the
-experiment (Movie S1 and Table S11). Mice fecal samples from the same
-source of mice continued to cluster closely to each other throughout the
+experiment (Movie S1 and Table S11). Fecal samples from the same source
+of mice continued to cluster closely to each other throughout the
 experiment. By 7 days post-infection, when approximately 43\% mice had
@@
 day 7 (Fig. 2B, Fig. S2C). We trained three L2-regularized logistic
-regression models with input bacterial community data from the baseline
-(day = -1), post-clindamycin (day = 0), and post-infection (day = 1)
-timepoints of the experiment to predict \emph{C. difficile} colonization
-status on day 7 (Fig. S3A-B). All models were better at predicting
-\emph{C. difficile} colonization status on day 7 than random chance (all
-\emph{P} \textless{} 0.001, Table S12). The model based on the
-post-clindamycin (AUROC = 0.78) community OTU data performed
+regression models with either input bacterial community data from the
+baseline (day = -1), post-clindamycin (day = 0), or post-infection (day
+= 1) timepoints of the experiment to predict \emph{C. difficile}
+colonization status on day 7 (Fig. S3A-B). All models were better at
+predicting \emph{C. difficile} colonization status on day 7 than random
+chance (all \emph{P} \textless{} 0.001, Table S12). The model based on
+the post-clindamycin (AUROC = 0.78) community OTU data performed
 significantly better than the baseline (AUROC = 0.72) or the
 post-infection (AUROC = 0.67) models (\emph{P}\textsubscript{FDR}
-\textless{} 0.001 for pairwise comparisons, Fig. S3C and Table S13).
+\textless{} 0.001 for pairwise comparisons; Fig. S3C and Table S13).
 Thus, we were able to use bacterial relative abundance data from the
@@
 cleared \emph{C. difficile} before day 7 from the mice still colonized
-with \emph{C. difficile} at that timepoint. This result suggests the
-bacterial community's response to clindamycin treatment had the greatest
-influence on subsequent \emph{C. difficile} colonization dynamics.
+with \emph{C. difficile} at that timepoint. This result suggests that
+the bacterial community's response to clindamycin treatment had the
+greatest influence on subsequent \emph{C. difficile} colonization
+dynamics.
 
@@
 treatment (day 0) bacterial community data. Out of the 10 highest ranked
-OTUs, 7 OTUs (\emph{Bacteroides}, \emph{Escherichia/Shigella}, 2
+OTUs, 7 OTUs were associated with \emph{C. difficile} colonization 7
+days post-infection (\emph{Bacteroides}, \emph{Escherichia/Shigella}, 2
 \emph{Lachnospiraceae}, \emph{Lactobacillus}, \emph{Porphyromonadaceae},
-and \emph{Ruminococcaceae}) were associated with \emph{C. difficile}
-colonization 7 days post-infection, while 3 OTUs
+and \emph{Ruminococcaceae}), while 3 OTUs were associated with clearance
 (\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
-\emph{Porphyromonadaceae}) were associated with clearance (Fig. 5A).
-Next, we examined whether any of the top 20 ranked OTUs from the
-post-clindamycin (day 0) model were also important in the other 2
-classification models based on baseline (day -1) and 1 day
-post-infection community data. We identified 6 OTUs
-(\emph{Enterobacteriaceae, Ruminococcaceae, Lactobacillus, Bacteroides,
-Porphyromonadaceae, Erysipelotrichaceae}) that were important to the
-post-clindamycin model and either the baseline or 1 day post-infection
-models (Table S14). Thus, a subset of bacterial OTUs were important for
+\emph{Porphyromonadaceae}; Fig. 5A). Next, we examined whether any of
+the top 20 ranked OTUs from the post-clindamycin (day 0) model were also
+important in the other 2 classification models based on baseline (day
+-1) and 1 day post-infection community data. We identified 6 OTUs that
+were important to the post-clindamycin model and either the baseline or
+1 day post-infection models (\emph{Enterobacteriaceae, Ruminococcaceae,
+Lactobacillus, Bacteroides, Porphyromonadaceae, Erysipelotrichaceae};
+Table S14). Thus, a subset of bacterial OTUs were important for
 determining \emph{C. difficile} colonization dynamics across multiple
@@
 both (Fig. 5B). Together, these results suggest that the initial
-bacterial communities and their responses to clindamycin influence the
+bacterial communities and their responses to clindamycin influenced the
 clearance of \emph{C. difficile}.
 
-Several OTUs (\emph{Bacteroides, Enterococcus, Enterobacteriaceae,
-Porphyromonadaceae}) that overlapped with our previous analyses appeared
-across at least 2 models, so we examined how the relative abundances of
+Several OTUs that overlapped with our previous analyses appeared across
+at least 2 models (\emph{Bacteroides, Enterococcus, Enterobacteriaceae,
+Porphyromonadaceae}), so we examined how the relative abundances of
 these OTUs varied over the course of the experiment (Fig. 6). Across the
@@
 
-Applying our CDI model with mice from 6 different sources, allowed us to
+Applying our CDI model to 6 different sources of mice, allowed us to
 identify bacterial taxa that were unique to different sources as well as
-taxa that were universally impacted by clindamycin. We trained L2
-logistic regression models with baseline (day -1), post-clindamycin
-treatment (day 0), and 1-day post-infection fecal community data that
-could predict whether mice cleared \emph{C. difficile} by 7 days
+taxa that were universally impacted by clindamycin. We trained logistic
+regression models with baseline (day -1), post-clindamycin treatment
+(day 0), and 1-day post-infection fecal community data that could
+predict whether mice cleared \emph{C. difficile} by 7 days
 post-infection better than random chance. We identified
 \emph{Bacteroides, Enterococcus, Enterobacteriaceae, Porphyromonadaceae}
-(Fig. 6) as candidate bacteria within these communities that were
-influencing variation in \emph{C. difficile} colonization dynamics since
-these bacteria were all important in the logistic regression models and
-varied by source, were impacted by clindamycin treatment, or both.
-Overall, our results demonstrated clindamycin was sufficient to render
-mice from multiple sources susceptible to CDI and only a subset of the
+(Fig. 6) as candidate bacteria within these communities that influenced
+variation in \emph{C. difficile} colonization dynamics since these
+bacteria were all important in the logistic regression models and varied
+by source, were impacted by clindamycin treatment, or both. Overall, our
+results demonstrated clindamycin was sufficient to render mice from
+multiple sources susceptible to CDI and only a subset of the
 inter-individual microbiota variation across mice from different sources
-was associated with the time needed to clear \emph{C. difficile}.
-
-Other studies have taken similar approaches by using mice from multiple
-sources to identify bacteria that either promote colonization resistance
-or increase susceptibility to enteric infections (22, 23, 26--30). For
-example, against \emph{Salmonella} infections, \emph{Enterobacteriaceae}
-and segmented filamentous bacteria have emerged as protective (22, 27).
-We found \emph{Enterobacteriaceae} increased in all sources of mice
-after clindamycin treatment, facilitating \emph{C. difficile}
-colonization. However, there was also variation in
-\emph{Enterobacteriaceae} relative abundance levels between sources that
-was associated with the variation in \emph{C. difficile} colonization
-dynamics across sources. Thus, bacteria may have differential roles in
-determining susceptibility depending on the type of bacterial infection.
+was needed to predict which mice could clear \emph{C. difficile}.
+
+Other studies have used mice from multiple sources to identify bacteria
+that either promote colonization resistance or increase susceptibility
+to enteric infections (22, 23, 26--30). For example, against
+\emph{Salmonella} infections, \emph{Enterobacteriaceae} and segmented
+filamentous bacteria have emerged as protective (22, 27). We found
+\emph{Enterobacteriaceae} increased in all sources of mice after
+clindamycin treatment, facilitating \emph{C. difficile} colonization.
+However, there was also variation in \emph{Enterobacteriaceae} relative
+abundance levels between sources that was associated with the variation
+in \emph{C. difficile} colonization dynamics across sources. Thus,
+bacteria may have differential roles in determining susceptibility
+depending on the type of bacterial infection.
 
@@
 \emph{C. difficile} outcomes, one having sustained colonization (32),
-while the other had transient (18), despite both using \emph{C.
-difficile} VPI 10643. Baseline differences in the microbiota composition
-have been hypothesized to partially explain the differences in
-colonization outcomes and overall susceptibility to \emph{C. difficile}
-after treatment with the same antibiotic (13, 31). A previous study with
-\emph{C. difficile} identified an endogenous protective \emph{C.
-difficile} strain LEM1 that bloomed after antibiotic treatment in mice
-from Jackson or Charles River Laboratories, but not Taconic that
-protected mice against the more toxigenic \emph{C. difficile} VPI10463
-(26). Given that we obtained mice from the same vendors, we checked all
-mice for endogenous \emph{C. difficile} by plating stool samples that
-were collected after clindamycin treatment. However, we did not identify
-any endogenous \emph{C. difficile} strains prior to challenge,
-suggesting there were no endogenous protective strains in the mice we
-received and other bacteria mediated the variation in \emph{C.
-difficile} colonization across sources. The \emph{C. difficile} strain
-used could also be contributing to the variation in \emph{C. difficile}
-outcomes seen across different research groups. For example, a group
-found differential colonization outcomes after clindamycin treatment,
-with \emph{C. difficile} 630 and M68 infections eventually becoming
-undetectable while strain BI-7 remained detectable up to 70 days
-post-treatment (44).
-
-The bacterial perturbations induced by clindamycin treatment have been
-well characterized and our findings agree with previous CDI mouse model
-work demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae}
-were associated with \emph{C. difficile} susceptibility and
+while the other had transient colonization (18), despite both using
+\emph{C. difficile} VPI 10643. Baseline differences in the microbiota
+composition have been hypothesized to partially explain the differences
+in colonization outcomes and overall susceptibility to \emph{C.
+difficile} after treatment with the same antibiotic (13, 31). When we
+treated mice from 6 different sources with clindamycin and challenged
+them with \emph{C. difficile} 630, we found microbiota variation across
+sources impacted colonization outcomes, but not susceptibility. A
+previous study with \emph{C. difficile} identified an endogenous
+protective \emph{C. difficile} strain LEM1 that bloomed after antibiotic
+treatment in mice from Jackson or Charles River Laboratories, but not
+Taconic that protected mice against the more toxigenic \emph{C.
+difficile} VPI10463 (26). Given that we obtained mice from the same
+vendors, we checked all mice for endogenous \emph{C. difficile} by
+plating stool samples that were collected after clindamycin treatment.
+However, we did not identify any endogenous \emph{C. difficile} strains
+prior to challenge, suggesting there were no endogenous protective
+strains in the mice we received and other bacteria mediated the
+variation in \emph{C. difficile} colonization across sources. The
+\emph{C. difficile} strain used could also be contributing to the
+variation in \emph{C. difficile} outcomes seen across different research
+groups. For example, a group found differential colonization outcomes
+after clindamycin treatment, with \emph{C. difficile} 630 and M68
+infections eventually becoming undetectable while strain BI-7 remained
+detectable up to 70 days post-treatment (44). The bacterial
+perturbations induced by clindamycin treatment have been well
+characterized and our findings agree with previous CDI mouse model work
+demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae} were
+associated with \emph{C. difficile} susceptibility and
 \emph{Porpyhromonadaceae}, \emph{Lachnospiraceae},
@@
 across sources of mice implying that at least in the context of the same
-perturbation, microbiota differences seemed to influence infection
-outcome more than susceptibility. More importantly, we were able to
-reduce the variation observed across sources to identify a subset of
-OTUs that were also important for predicting \emph{C. difficile}
-colonization status 7 days post-infection. Since all but 3 mice
-eventually cleared \emph{C. difficile} 630 by 9 days post-infection and
-the model built with the post-clindamycin (day 0) OTU relative abundance
-data had the best performance, our results suggest clindamycin treatment
-had a larger role in determining \emph{C. difficile} susceptibility and
-clearance than the source of the mice.
-
-Our approach successfully increased the inter-animal variation tested
-with our clindamycin model of CDI. One alternative approach that has
-been used in some CDI studies (47--52) is to associate mice with human
-microbiotas. However, a major caveat to this method is the substantial
-loss of human microbiota community members upon transfer to mice (53,
-54). Additionally with the exception of 2 recent studies (47, 48), most
-of the CDI mouse model studies to date associated mice with just 1 types
-of human microbiota either from a single donor or a single pool from
-multiple donors (49--52), which does not aid in the goal of modeling the
-interpersonal variation seen in humans to understand how the microbiota
-influences susceptibility to CDIs and adverse outcomes. Importantly, our
-study using mice from 6 different sources increased the variation
-between groups of mice compared to using 1 source alone, to better
-reflect the inter-individual microbiota variation observed in humans.
-
-Another motivation for associating mice with human microbiotas,is to
-study the bacteria associated with the disease in humans. Encouragingly,
-decreased \emph{Bifidobacterium}, \emph{Porphyromonas},
-\emph{Ruminococcaceae} and \emph{Lachnospiraceae} and increased
-\emph{Enterobacteriaceae}, \emph{Enterococcus}, \emph{Lactobacillus},
-and \emph{Proteus} have all been associated with human CDIs (7) and were
-well represented in our study, suggesting most of the mouse sources are
-suitable for gaining insights into the bacteria influencing \emph{C.
+perturbation, microbiota differences influence infection outcome. More
+importantly, we were able to explain the variation observed across
+sources with a subset of OTUs that were also important for predicting
+\emph{C. difficile} colonization status 7 days post-infection. Since all
+but 3 mice eventually cleared \emph{C. difficile} 630 by 9 days
+post-infection and the model built with the post-clindamycin (day 0) OTU
+relative abundance data had the best performance, our results suggest
+clindamycin treatment had a larger role in determining \emph{C.
+difficile} susceptibility and clearance than the source of the mice.
+
+Using mice from multiple sources successfully increased the inter-animal
+variation. One alternative approach that has been used in some CDI
+studies is to associate mice with human microbiotas (47--52). However, a
+major caveat to this method is the substantial loss of human microbiota
+community members upon transfer to mice (53, 54). Additionally with the
+exception of 2 recent studies (47, 48), most of these studies associated
+mice with just 1 types of human microbiota either from a single donor or
+a single pool from multiple donors (49--52). This approach does not aid
+in the goal of modeling the interpersonal variation seen in humans to
+understand how the microbiota influences susceptibility to CDIs and
+adverse outcomes. Importantly, our study using mice from 6 different
+sources increased the variation between groups of mice compared to using
+1 source alone, to better reflect the inter-individual microbiota
+variation observed in humans.
+
+Another motivation for associating mice with human microbiotas is to
+study the bacteria associated with the disease in humans. Decreased
+\emph{Bifidobacterium}, \emph{Porphyromonas}, \emph{Ruminococcaceae} and
+\emph{Lachnospiraceae} and increased \emph{Enterobacteriaceae},
+\emph{Enterococcus}, \emph{Lactobacillus}, and \emph{Proteus} have all
+been associated with human CDIs (7). Encouragingly, these populations
+were well represented in our study, suggesting most of the mouse sources
+are suitable for gaining insights into the bacteria influencing \emph{C.
 difficile} colonization and infections in humans. An important exception
@@
 subset of mice has important implications for future CDI mouse model
-studies.
+studies, but also models the natural patchiness of microbial populations
+in humans.
 
@@
 \emph{Enterococcus} and \emph{C. difficile}, these results could serve
-as a resource for selecting mice to address their question. Using mice
+as a resource for selecting mice to address the question. Using mice
 from multiple sources helps model the interpersonal microbiota variation
@@
 and Health Research Postdoctoral Translation Scholars Program
-(UL1TR002240). We thank members of the Schloss lab for feedback on
-planning the experiments and data presentation. In particular, we thank
-Begüm Topçuoğlu for help with implementing L2 logistic regression
-models, Ana Taylor for help with media preparation and sample
-collection, and Nicholas Lesniak for his critical feedback on the
-manuscript. We also thank members of Vincent Young's lab, particularly
-Kimberly Vendrov, for guidance with the \emph{C. difficile} infection
-mouse model and donating the mice. We also thank the Unit for Laboratory
-Animal Medicine at the University of Michigan for maintaining our mouse
-colony and providing the institutional support for our mouse
-experiments. Finally, we thank Kwi Kim, Austin Campbell, and Kimberly
-Vendrov for their help in maintaining the Schloss lab's anaerobic
-chamber.
+(UL1TR002240 from the National Center for Advancing Translational
+Sciences). We thank members of the Schloss lab for feedback on planning
+the experiments and data presentation. In particular, we thank Begüm
+Topçuoğlu for help with implementing logistic regression models, Ana
+Taylor for help with media preparation and sample collection, and
+Nicholas Lesniak for his critical feedback on the manuscript. We also
+thank members of Vincent Young's lab, particularly Kimberly Vendrov, for
+guidance with the \emph{C. difficile} infection mouse model and donating
+the mice. We also thank the Unit for Laboratory Animal Medicine at the
+University of Michigan for maintaining our mouse colony and providing
+the institutional support for our mouse experiments. Finally, we thank
+Kwi Kim, Austin Campbell, and Kimberly Vendrov for their help in
+maintaining the Schloss lab's anaerobic chamber.
 
@@
 colony and the Young lab colony). The Young lab colony was originally
-established with mice purchased from Jackson, and the Schloss lab colony
-was established in 2010 with mice donated from the Young lab. The 4
-groups of mice purchased from vendors were allowed to acclimate to the
-University of Michigan mouse facility for 13 days prior to starting the
-experiment. At least 4 female mice (age 5-10 weeks) were obtained per
-source and mice from the same source were primarily housed at a density
-of 2 mice per cage. The experiment was repeated once, approximately 3
-months after the start of the first experiment.
-
-\textbf{(ii) Antibiotic treatment.} After the 13-day acclimation period
-and 1 day prior to challenge (Fig. 1A), all mice received 10 mg/kg
-clindamycin (filter sterilized through a 0.22 micron syringe filter
-prior to administration) via intraperitoneal injection.
+established with mice purchased from Jackson in 2002, and the Schloss
+lab colony was established in 2010 with mice donated from the Young lab.
+The 4 groups of mice purchased from vendors were allowed to acclimate to
+the University of Michigan mouse facility for 13 days prior to starting
+the experiment. At least 4 female mice (age 5-10 weeks) were obtained
+per source and mice from the same source were primarily housed at a
+density of 2 mice per cage. The experiment was repeated once,
+approximately 3 months after the start of the first experiment.
+
+\textbf{(ii) Antibiotic treatment.} After the 13-day acclimation period,
+all mice received 10 mg/kg clindamycin (filter sterilized through a 0.22
+micron syringe filter prior to administration) via intraperitoneal
+injection (Fig. 1A).
 
@@
 described previously (21). Mice weights and stool samples were taken
-daily through 9 days post-challenge. Collected stool was split for
-\emph{C. difficile} CFU quantification and 16S rRNA sequencing analysis.
-\emph{C. difficile} quantification stool samples were transferred to the
-anaerobic chamber, serially diluted in PBS, plated on
+daily through 9 days post-challenge (Fig. 1A). Collected stool was split
+for \emph{C. difficile} quantification and 16S rRNA sequencing analysis.
+For \emph{C. difficile} quantification, stool samples were transferred
+to the anaerobic chamber, serially diluted in PBS, plated on
 taurocholate-cycloserine-cefoxitin-fructose agar (TCCFA) plates, and
@@
 abundance data at the OTU level from the baseline, post-clindamycin, and
-post-infection timepoints was used to generate 6 classification models
-that predicted \emph{C. difficile} colonization status 7 days
+1-day post-infection timepoints was used to generate 3 classification
+models that predicted \emph{C. difficile} colonization status 7 days
 post-infection. The L2-regularized logistic regression models were
@@
 classification outcomes and change the data split ratios. The modified
-repository to regenerate this analysis is available at
+repository to regenerate our modeling analysis is available at
 \url{https://github.com/tomkoset/ML_pipeline_microbiome}.
@@
 OTUs. OTUs impacted by clindamycin treatment were identified using the
-paired Wilcoxon signed rank test with matched pairs of mice samples for
+paired Wilcoxon signed rank test with matched pairs of mice samples from
 day -1 and day 0. To determine whether classification models had better
@@
 one-sample Wilcoxon signed rank test. To examine whether there was an
-overall difference in predictive performance across the 6 classification
+overall difference in predictive performance across the 3 classification
 models we used the Kruskal-Wallis test followed by pairwise Wilcoxan
@@
 the variation (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
-\textless{} 0.001, see Table S3). For A-C: each symbol represents the
-value for a stool sample from an individual mouse, circles represent
+\textless{} 0.001; Table S3). For A-C: each symbol represents the value
+for a stool sample from an individual mouse, circles represent
 experiment 1 mice and triangles represent experiment 2 mice. D. The
@@
 post-clindamycin (PERMANOVA combined R\textsuperscript{2} = 0.99,
-\emph{P} \textless{} 0.001, see Table S3). For A-C, each symbol
-represents a stool sample from an individual mouse, with circles
-representing experiment 1 mice and triangles representing experiment 2
-mice. D. The median (point) and interquantile range (colored lines) of
-the relative abundances for the 18 OTUs (Table S8) that varied between
-sources after clindamycin treatment (day 0). E. The median (point) and
-interquantile range (colored lines) of the top 10 most significant OTUs
-out of 153 with relative abundances that changed because of the
-clindamycin treatment (adjusted \emph{P} value \textless{} 0.05). Data
-were analyzed by paired Wilcoxon signed rank test of mice that had
-paired sequence data for baseline (day -1) and post-clindamycin (day 0)
-timepoints (N = 31), with Benjamini-Hochberg correction for testing all
-identified OTUs (Table S9). The gray vertical line indicates the limit
-of detection.
+\emph{P} \textless{} 0.001; Table S3). For A-C, each symbol represents a
+stool sample from an individual mouse, with circles representing
+experiment 1 mice and triangles representing experiment 2 mice. D. The
+median (point) and interquantile range (colored lines) of the relative
+abundances for the 18 OTUs (Table S8) that varied between sources after
+clindamycin treatment (day 0). E. The median (point) and interquantile
+range (colored lines) of the top 10 most significant OTUs out of 153
+with relative abundances that changed because of the clindamycin
+treatment (adjusted \emph{P} value \textless{} 0.05). Data were analyzed
+by paired Wilcoxon signed rank test of mice that had paired sequence
+data for baseline (day -1) and post-clindamycin (day 0) timepoints (N =
+31), with Benjamini-Hochberg correction for testing all identified OTUs
+(Table S9). The gray vertical line indicates the limit of detection.
 
@@
 communities (PERMANOVA combined R\textsuperscript{2} = 0.88, \emph{P}
-\textless{} 0.001, Table S3). For A-C: each symbol represents the value
+\textless{} 0.001; Table S3). For A-C: each symbol represents the value
 for a stool sample from an individual mouse, circles represent
@@
 
-\includegraphics{figure_5.pdf} \textbf{Figure 5. Bacteria that influence
-whether mice cleared \emph{C. difficile} by day 7.} A. Post-clindamycin
-(day 0) relative abundance data for the 10 OTUs with the highest
-rankings based on feature weights in the post-clindamycin (day 0)
-classification model. Red font represents OTUs that correlated with
-\emph{C. difficile} colonization and blue font represents OTUs that
+\includegraphics{figure_5.pdf} \textbf{Figure 5. Bacteria that
+influenced whether mice cleared \emph{C. difficile} by day 7.} A.
+Post-clindamycin (day 0) relative abundance data for the 10 OTUs with
+the highest rankings based on feature weights in the post-clindamycin
+(day 0) classification model. Red font represents OTUs that correlated
+with \emph{C. difficile} colonization and blue font represents OTUs that
 correlated with clearance. Symbols represent the relative abundance data
@@
 
-\includegraphics{figure_S1.pdf} \textbf{Figure S1. Bacterial communities
-vary between experiments for some sources.} A-F. PCoA of \(\theta_{YC}\)
-distances for the baseline fecal bacterial communities within each
-source of mice. Each symbol represents a stool sample from an individual
-mouse with color corresponding to experiment and shape representing cage
-mates. Experiment number and cage effects explained most of the observed
+\includegraphics{figure_S1.pdf}
+
+\textbf{Figure S1. Bacterial communities vary between experiments for
+some sources.} A-F. PCoA of \(\theta_{YC}\) distances for the baseline
+fecal bacterial communities within each source of mice. Each symbol
+represents a stool sample from an individual mouse with color
+corresponding to experiment and shape representing cage mates.
+Experiment number and cage effects explained most of the observed
 variation for samples from the Schloss (PERMANOVA combined
@@
 
-\includegraphics{figure_S4.pdf} \textbf{Figure S4. OTUs from
-classification models based on baseline, post-clindamycin treatment, or
-1-day post-infection community data vary by source, clindamycin
-treatment, or both.} A-C. Venn diagrams of OTUs from the top 20 OTUs
-from the baseline (A), post-clindamycin treatment (B), and 1-day
-post-infection (C) classification models (Table S14) that overlapped
-with OTUs that varied across sources at the corresponding timepoint
-(Tables S5, 8, 10), were impacted by clindamycin treatment (Table S9),
-or both. Bold OTUs were important to more than 1 classification model.
+\includegraphics{figure_S4.pdf}
+
+\textbf{Figure S4. OTUs from classification models based on baseline,
+post-clindamycin treatment, or 1-day post-infection community data vary
+by source, clindamycin treatment, or both.} A-C. Venn diagrams of OTUs
+from the top 20 OTUs from the baseline (A), post-clindamycin treatment
+(B), and 1-day post-infection (C) classification models (Table S14) that
+overlapped with OTUs that varied across sources at the corresponding
+timepoint (Tables S5, 8, 10), were impacted by clindamycin treatment
+(Table S9), or both. Bold OTUs were important to more than 1
+classification model.
 

### 2020-09-20 code/cfu_plot.R
Rearrange Figure S2

@@ -81,6 +81,15 @@ summarize_plot <- function(df){
           panel.grid.minor.x = element_line(size = 0.4, color = "grey"))  # Add gray lines to clearly separate symbols by days)
 }
 
+#Plots of CFU data by experiment----
+facet_labels <- c("1st experiment", "2nd experiment")
+names(facet_labels) <- c("1", "2")
+exp1_2_cfu <- summarize_plot(cfu_data_final)+
+  facet_wrap( ~ experiment, labeller = labeller(experiment = facet_labels), scales = "fixed")+
+  theme(plot.title = element_text(hjust = 0.5),#Center plot title
+        strip.background = element_blank()) #get rid of box around facet_wrap labels
+save_plot(filename = paste0("results/figures/exp1_2_cfu_time.png"), plot = exp1_2_cfu, base_aspect_ratio = 2)
+  
 #CFU plot for the 1st experiment----
 exp1_cfu <- summarize_plot(cfu_data_final %>% filter(experiment == 1))+
   ggtitle("1st experiment")+ #Title plot

### 2020-09-20 code/figure_S2.R

@@ -1,15 +1,21 @@
 source("code/functions.R")
 
-a <- ggdraw() + draw_image("results/figures/exp1_cfu_time.png")
-b <- ggdraw() + draw_image("results/figures/exp2_cfu_time.png")
+#Since A and B were faceted, need to add letter labels separately
+letter_annotation <- data.frame(x_pos = c(.02, .55),
+                                y_pos = c(.97, .97), 
+                                lab = c("A", "B")) #To label the different panels of the plot
+
+a <- ggdraw() + draw_image("results/figures/exp1_2_cfu_time.png")+
+  geom_text(data = letter_annotation, aes( x = x_pos, y = y_pos, label = lab), size = 4.5, fontface = "bold")
 c <- ggdraw() + draw_image("results/figures/C.diff_CFU_D7_stats.png")
 d <- ggdraw() + draw_image("results/figures/weight_D2.png")
 e <- ggdraw() + draw_image("results/figures/C.diff_percent_colonized.png")
-right_panel <- plot_grid(c, d, e, labels = c("C", "D", "E"), label_size = 12, nrow = 1)
+bottom_panel <- plot_grid(c, d, e, labels = c("C", "D", "E"), label_size = 12, nrow = 1)
 
 
-plot_grid(a, right_panel, b,  labels = c("A", "", "B"), label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_S2.pdf", width=8, height=4.5)+
-  ggsave("submission/figure_S2.pdf", width=8, height=4.5)
+
+plot_grid(a, bottom_panel, labels = c("", ""), label_size = 12, ncol=1, rel_heights = c(1.4, 1))+
+  ggsave("results/figures/figure_S2.pdf", width=8, height=6)+
+  ggsave("submission/figure_S2.pdf", width=8, height=6)
 
 

### 2020-09-21 submission/manuscript.Rmd
Edits for revision

@@
 
-# The initial gut microbiota and response to antibiotic perturbation influence *Clostridioides difficile* colonization in mice
+# The initial gut microbiota and response to antibiotic perturbation influence *Clostridioides difficile* clearance in mice
 
@@
 
-We sequenced the 16S rRNA gene from fecal samples collected from these mice after they acclimated to the University of Michigan animal housing environment.
+We sequenced the V4 region of the 16S rRNA gene from fecal samples collected from these mice after they acclimated to the University of Michigan animal housing environment.
 We first examined the alpha diversity across the 6 sources of mice.
@@
 There were some differences between the 2 experiments we conducted, as the experiment and cage effects significantly explained the observed community variation for the Schloss and Young lab mouse colonies (Fig.
-S2A-B and Table S4).
+S1A-B and Table S4).
 However, most of the vendors also clustered by experiment (Fig.
-S2C-D, F), suggesting there was some community variation between the 2 experiments within each source, particularly for Schloss, Young, and Envigo mice (Fig.
-S2G-H).
+S1C-D, F), suggesting there was some community variation between the 2 experiments within each source, particularly for Schloss, Young, and Envigo mice (Fig.
+S1G-H).
 After finding differences at the community level, we next identified the bacteria that varied between sources of mice.
-There were `r source_otus_dn1` OTUs with relative abundances that were significantly different between the sources (Fig.
+There were `r source_otus_dn1` OTUs with relative abundances that were significantly different between the sources at baseline (Fig.
 1D and Table S5).
-Though we saw differences between experiments at the community level, there were no OTUs that were significantly different between experiments within Schloss, Young, and Envigo mice (all *P* > 0.05).
+Though we saw differences between experiments at the community level, there were no OTUs that were significantly different between experiments within Schloss, Young, and Envigo mice at baseline (all *P* > 0.05).
 By using mice from six sources we were able to increase the variation in the starting communities to evaluate in a clindamycin-based CDI model.
@@
 We have previously demonstrated mice are rendered susceptible to *C.
-difficile*, but clear the pathogen within 9 days when treated with clindamycin alone [@Jenior2018; @Tomkovich2019].
+difficile*, but clear the pathogen within 9 days, thus colonization is transient when treated with clindamycin alone [@Jenior2018; @Tomkovich2019].
 All mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C.
@@
 difficile* clearance.
-While the colonization dynamics were similar between the two experiments, the Schloss mice took longer to clear in the first experiment compared to the second and the Envigo mice took longer to clear in the second experiment compared to the first (Fig.
+While the colonization dynamics were similar between the two experiments, the Schloss mice took longer to clear *C.
+difficile* in the first experiment compared to the second and the Envigo mice took longer to clear *C.
+difficile* in the second experiment compared to the first (Fig.
 S2A-B).
@@
 3C and Table S3).
-However, there were only `r source_otus_d0` OTUs with relative abundances that significantly varied between sources (Fig.
+However, there were only `r source_otus_d0` OTUs with relative abundances that significantly varied between sources after clindamycin treatment (Fig.
 3D and Table S8).
@@
 3D), but only in Taconic mice.
-*Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased in relative abundance after clindamycin treatment (Fig.
+*Enterococcus* was primarily found in mice purchased from commercial vendors and also increased in relative abundance after clindamycin treatment (Fig.
 3D).
@@
 S2C).
-We trained three L2-regularized logistic regression models with either input bacterial community data from the baseline (day = -1), post-clindamycin (day = 0), or post-infection (day = 1) timepoints of the experiment to predict *C.
+We trained three L2-regularized logistic regression models with either input bacterial community data from the 6 sources of mice at the baseline (day = -1), post-clindamycin (day = 0), or post-infection (day = 1) timepoints of the experiment to predict *C.
 difficile* colonization status on day 7 (Fig.
@@
 difficile* colonization 7 days post-infection (*Bacteroides*, *Escherichia/Shigella*, 2 *Lachnospiraceae*,  *Lactobacillus*, *Porphyromonadaceae*, and *Ruminococcaceae*), while 3 OTUs were associated with clearance (*Enterobacteriaceae*, *Lachnospiraceae*, *Porphyromonadaceae*; Fig.
+5A).
+On day 0, the majority of these OTUs were impacted by clindamycin and had relative abundances that were close to the limit of detection (Fig.
 5A).
@@
 Applying our CDI model to 6 different sources of mice, allowed us to identify bacterial taxa that were unique to different sources as well as taxa that were universally impacted by clindamycin.
-We trained logistic regression models with baseline (day -1), post-clindamycin treatment (day 0), and 1-day post-infection fecal community data that could predict whether mice cleared *C.
+We trained L2-regularized logistic regression models with baseline (day -1), post-clindamycin treatment (day 0), and 1-day post-infection fecal community data that could predict whether mice cleared *C.
 difficile* by 7 days post-infection better than random chance.
@@
 For example, against *Salmonella* infections, *Enterobacteriaceae* and segmented filamentous bacteria have emerged as protective [@Velazquez2019; @Lai2020].
-We found *Enterobacteriaceae* increased in all sources of mice after clindamycin treatment, facilitating *C.
+We found *Enterobacteriaceae* increased in all sources of mice after clindamycin treatment, positively correlating with *C.
 difficile* colonization.
@@
 difficile* 630 and M68 infections eventually becoming undetectable while strain BI-7 remained detectable up to 70 days post-treatment [@Lawley2012].
+One study limitation is that we only used female mice.
+Sex has been shown to influence microbiota variation in mice [@Wang2019], so we used female mice to reduce this confounding variable and also match the sex used in previous CDI studies that administered clindamycin to mice [@Buffie2011; @Lawley2009; @Lawley2012; @Buffie2014].
 The bacterial perturbations induced by clindamycin treatment have been well characterized and our findings agree with previous CDI mouse model work demonstrating *Enterococcus* and *Enterobacteriaceae* were associated with *C.
@@
 However, a major caveat to this method is the substantial loss of human microbiota community members upon transfer to mice [@Fouladi2020; @Walter2020].
-Additionally with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of these studies associated mice with just 1 types of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018].
+Additionally, with the exception of 2 recent studies [@NagaoKitamoto2020; @Battaglioli2018], most of these studies associated mice with just 1 type of human microbiota either from a single donor or a single pool from multiple donors [@Robinson2014; @Collins2015; @Collins2018; @Hryckowian2018].
 This approach does not aid in the goal of modeling the interpersonal variation seen in humans to understand how the microbiota influences susceptibility to CDIs and adverse outcomes.
@@
 Interestingly, butyrate has previously been shown to vary across mouse vendors and mediated resistance to *Citrobacter rodentium* infection, a model of enterohemorrhagic and enteropathogenic *Escherichia coli* infections [@Osbelt2020].
-Evidence for immunological toning differences in IgA and Th17 cells across mice from different vendors have also been documented and [@Fransen2015; @Ivanov2009] could influence the host response to CDI [@Azrad2018; @Saleh2019].
+Evidence for immunological toning differences in IgA and Th17 cells across mice from different vendors have also been documented [@Fransen2015; @Ivanov2009] and could influence the host response to CDI [@Azrad2018; @Saleh2019], particularly relevant for *C.
+difficile* strains that induce more severe disease than *C.
+difficile* 630.
 The outcome after *C.
-difficile* exposure depends on a multitude of factors, including age, diet, and immunity; all of which are also influenced by the microbiota.
+difficile* exposure depends on a multitude of factors, including genetics, age, diet, and immunity; all of which also influence the microbiota.
 
@@
 difficile* strain 630 via oral gavage post-infection 1 day after clindamycin treatment as described previously [@Jenior2018].
-Mice weights and stool samples were taken daily through 9 days post-challenge (Fig.
+Mice weights and stool samples were taken daily through 9 days post-infection (Fig.
 1A).
@@
 Operational taxonomic units (OTUs) were assigned with a 97% similarity threshold using the opticlust algorithm [@Westcott2017].
+Based on the mock communities, our overall sequencing error rate was 0.0112% and all  water controls had less than 1000 sequences (range: 18-875).
 To account for uneven sequencing across samples, samples were rarefied to 5,437 sequences 1,000 times for alpha and beta diversity analyses, and a single time to generate relative abundances for model training.
-PCoAs were generated based on $\theta_{YC}$ distances.
+PCoAs were generated based on the Yue and Clayton measure of dissimilarity ($\theta_{YC}$) distances.
 Permutational multivariate analysis of variance (PERMANOVA) was performed on mothur-generated $\theta_{YC}$ distance matrices with the adonis function in the vegan package [@Vegan2018] in R [@r_citation_2018].
@@
 difficile* colonization status 7 days post-infection.
-The L2-regularized logistic regression models were trained and tested using the caret package [@Kuhn2008] in R as previously described [@Topcuoglu2020] with the exception that we used 60% training and 40% testing data splits for the cross-validation of the training data to select the best cost hyperparameter and the testing of the held out test data to measure model performance.
+The L2-regularized logistic regression models were trained and tested using the caret package [@Kuhn2008] in R as previously described [@Topcuoglu2020] with the exception that we used 60% training and 40% testing data splits for testing of the held out test data to measure model performance and repeated k-fold cross-validation of the training data to select the best cost hyperparameter.
 The modified training to testing ratio was selected to accommodate the small number of samples in the dataset.
@@
 
-**(viii) Code availability.** Code for all data analysis and generating this manuscript is available at https://github.com/SchlossLab/Tomkovich_Vendor_XXXX_2020.
+**(viii) Code availability.** Code for all data analysis and generating this manuscript is available at https://github.com/SchlossLab/Tomkovich_Vendor_mSphere_2020.
 
@@
 D.
-The median (point) and interquantile range (colored lines) of the relative abundances for the 20 most significant OTUs out of the `r source_otus_dn1` OTUs that varied across sources at baseline (Table S5).
+The median (point) and interquantile range (colored lines) of the relative abundances for the 20 most significant OTUs out of the `r source_otus_dn1` OTUs that varied across sources at baseline by Kruskal-Wallis test with Benjamini-Hochberg correction (Table S5).
 
@@
 Mice for the experiments were obtained from 6 different sources: the Schloss (N = `r schloss_i`) and Young lab (N = `r young_i`) colonies at the University of Michigan, the Jackson Laboratory (N = `r jackson_i`), Charles River Laboratory (N = `r cr_i`), Taconic Biosciences (N = `r taconic_i`), and Envigo (N = `r envigo_i`).
+Mice that were ordered from commercial vendors acclimated to the University of Michigan mouse facility for 13 days prior to antibiotic administration.
 All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with *C.
@@
 Gray bars indicate the median relative abundances for each source.
+The gray horizontal lines indicates the limit of detection.
 B.
@@
 Bacterial community composition before, after clindamycin perturbation, and post-infection can predict *C.
-difficile* colonization status 7 days post-challenge.** A.
+difficile* colonization status 7 days post-infection.** A.
 Bar graph visualizations of overall 7-days post-infection *C.

### 2020-09-21 submission/manuscript.tex
Edits for revision

@@
 
-\hypertarget{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}{%
+\hypertarget{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-clearance-in-mice}{%
 \section{\texorpdfstring{The initial gut microbiota and response to
 antibiotic perturbation influence \emph{Clostridioides difficile}
-colonization in
-mice}{The initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice}}\label{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}}
+clearance in
+mice}{The initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile clearance in mice}}\label{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-clearance-in-mice}}
 
@@
 
-We sequenced the 16S rRNA gene from fecal samples collected from these
-mice after they acclimated to the University of Michigan animal housing
-environment. We first examined the alpha diversity across the 6 sources
-of mice. There was a significant difference in the richness (i.e.~number
-of observed operational taxonomic units (OTUs)), but not Shannon
-diversity index across the sources of mice (\emph{P}\textsubscript{FDR}
-= 0.03 and \emph{P}\textsubscript{FDR} = 0.052, respectively; Fig. 1A-B
-and Tables S1-2). Next, we compared the community structure of mice
-(Fig. 1C). The source of mice and the interactions between the source
-and cage effects explained most of the observed variation between fecal
-communities (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
-\textless{} 0.001; Fig. 1C and Table S3). Mice that are co-housed tend
-to have similar gut microbiotas due to coprophagy (41). Since mice
-within the same source were housed together, it was not surprising that
-the cage effect also contributed to the observed community variation.
-There were some differences between the 2 experiments we conducted, as
-the experiment and cage effects significantly explained the observed
+We sequenced the V4 region of the 16S rRNA gene from fecal samples
+collected from these mice after they acclimated to the University of
+Michigan animal housing environment. We first examined the alpha
+diversity across the 6 sources of mice. There was a significant
+difference in the richness (i.e.~number of observed operational
+taxonomic units (OTUs)), but not Shannon diversity index across the
+sources of mice (\emph{P}\textsubscript{FDR} = 0.03 and
+\emph{P}\textsubscript{FDR} = 0.052, respectively; Fig. 1A-B and Tables
+S1-2). Next, we compared the community structure of mice (Fig. 1C). The
+source of mice and the interactions between the source and cage effects
+explained most of the observed variation between fecal communities
+(PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P} \textless{}
+0.001; Fig. 1C and Table S3). Mice that are co-housed tend to have
+similar gut microbiotas due to coprophagy (41). Since mice within the
+same source were housed together, it was not surprising that the cage
+effect also contributed to the observed community variation. There were
+some differences between the 2 experiments we conducted, as the
+experiment and cage effects significantly explained the observed
 community variation for the Schloss and Young lab mouse colonies (Fig.
-S2A-B and Table S4). However, most of the vendors also clustered by
-experiment (Fig. S2C-D, F), suggesting there was some community
+S1A-B and Table S4). However, most of the vendors also clustered by
+experiment (Fig. S1C-D, F), suggesting there was some community
 variation between the 2 experiments within each source, particularly for
-Schloss, Young, and Envigo mice (Fig. S2G-H). After finding differences
+Schloss, Young, and Envigo mice (Fig. S1G-H). After finding differences
 at the community level, we next identified the bacteria that varied
 between sources of mice. There were 268 OTUs with relative abundances
-that were significantly different between the sources (Fig. 1D and Table
-S5). Though we saw differences between experiments at the community
-level, there were no OTUs that were significantly different between
-experiments within Schloss, Young, and Envigo mice (all \emph{P}
-\textgreater{} 0.05). By using mice from six sources we were able to
-increase the variation in the starting communities to evaluate in a
-clindamycin-based CDI model.
+that were significantly different between the sources at baseline (Fig.
+1D and Table S5). Though we saw differences between experiments at the
+community level, there were no OTUs that were significantly different
+between experiments within Schloss, Young, and Envigo mice at baseline
+(all \emph{P} \textgreater{} 0.05). By using mice from six sources we
+were able to increase the variation in the starting communities to
+evaluate in a clindamycin-based CDI model.
 
@@
 model (14). We have previously demonstrated mice are rendered
-susceptible to \emph{C. difficile}, but clear the pathogen within 9 days
-when treated with clindamycin alone (21, 43). All mice were treated with
-10 mg/kg clindamycin via intraperitoneal injection and one day later
-challenged with 10\textsuperscript{3} \emph{C. difficile} 630 spores
-(Fig. 2A). The day after infection, \emph{C. difficile} was detectable
-in all mice at a similar level (median CFU range: 2.2e+07-1.3e+08;
+susceptible to \emph{C. difficile}, but clear the pathogen within 9
+days, thus colonization is transient when treated with clindamycin alone
+(21, 43). All mice were treated with 10 mg/kg clindamycin via
+intraperitoneal injection and one day later challenged with
+10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 2A). The day
+after infection, \emph{C. difficile} was detectable in all mice at a
+similar level (median CFU range: 2.2e+07-1.3e+08;
 \emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
@@
 dynamics were similar between the two experiments, the Schloss mice took
-longer to clear in the first experiment compared to the second and the
-Envigo mice took longer to clear in the second experiment compared to
-the first (Fig. S2A-B). The change in the mice's weight significantly
-varied across sources of mice with the most weight loss occurring two
-days post-infection (Fig. 2C and Table S7). There was also one Jackson
-and one Envigo mouse that died between 1- and 3-days post-infection
-during the second experiment. Mice obtained from Jackson, Taconic, and
-Envigo tended to lose more weight, have higher \emph{C. difficile} CFU
-levels and take longer to clear the infection compared to the other
-sources of mice (although there was variation between experiments with
-Schloss and Envigo mice). This was particularly evident 7 days
-post-infection (Fig. 2B-C, Fig. S2C-D), when 57\% of the mice were still
-colonized with \emph{C. difficile} (Fig. S2E). By 9 days post-infection
-the majority of the mice from all sources had cleared \emph{C.
-difficile} with the exception of 1 Taconic mouse from the first
-experiment and 2 Envigo mice from the second experiment (Fig. 2B). Thus,
-clindamycin rendered all mice susceptible to \emph{C. difficile} 630
-colonization, regardless of source, but there was significant variation
-in disease phenotype across the sources of mice.
+longer to clear \emph{C. difficile} in the first experiment compared to
+the second and the Envigo mice took longer to clear \emph{C. difficile}
+in the second experiment compared to the first (Fig. S2A-B). The change
+in the mice's weight significantly varied across sources of mice with
+the most weight loss occurring two days post-infection (Fig. 2C and
+Table S7). There was also one Jackson and one Envigo mouse that died
+between 1- and 3-days post-infection during the second experiment. Mice
+obtained from Jackson, Taconic, and Envigo tended to lose more weight,
+have higher \emph{C. difficile} CFU levels and take longer to clear the
+infection compared to the other sources of mice (although there was
+variation between experiments with Schloss and Envigo mice). This was
+particularly evident 7 days post-infection (Fig. 2B-C, Fig. S2C-D), when
+57\% of the mice were still colonized with \emph{C. difficile} (Fig.
+S2E). By 9 days post-infection the majority of the mice from all sources
+had cleared \emph{C. difficile} with the exception of 1 Taconic mouse
+from the first experiment and 2 Envigo mice from the second experiment
+(Fig. 2B). Thus, clindamycin rendered all mice susceptible to \emph{C.
+difficile} 630 colonization, regardless of source, but there was
+significant variation in disease phenotype across the sources of mice.
 
@@
 were only 18 OTUs with relative abundances that significantly varied
-between sources (Fig. 3D and Table S8). Next, we identified the bacteria
-that shifted after clindamycin treatment, regardless of source by
-analyzing paired fecal samples from mice that were collected at baseline
-and after clindamycin treatment. We identified 153 OTUs that were
-altered after clindamycin treatment in most mice (Fig. 3E and Table S9).
-When we compared the list of significant clindamycin impacted bacteria
-with the bacteria that varied between sources post-clindamycin, we found
-4 OTUs that were shared between the lists (\emph{Enterobacteriaceae (OTU
-1)}, \emph{Lachnospiraceae} (OTU 130), \emph{Lactobacillus} (OTU 6),
+between sources after clindamycin treatment (Fig. 3D and Table S8).
+Next, we identified the bacteria that shifted after clindamycin
+treatment, regardless of source by analyzing paired fecal samples from
+mice that were collected at baseline and after clindamycin treatment. We
+identified 153 OTUs that were altered after clindamycin treatment in
+most mice (Fig. 3E and Table S9). When we compared the list of
+significant clindamycin impacted bacteria with the bacteria that varied
+between sources post-clindamycin, we found 4 OTUs that were shared
+between the lists (\emph{Enterobacteriaceae (OTU 1)},
+\emph{Lachnospiraceae} (OTU 130), \emph{Lactobacillus} (OTU 6),
 \emph{Enterococcus} (OTU 23); Fig. 3D-E and Tables S8-9). Importantly,
@@
 clindamycin treatment (Fig. 3D), but only in Taconic mice.
-\emph{Enterococcus} was primarily found only in mice purchased from
+\emph{Enterococcus} was primarily found in mice purchased from
 commercial vendors and also increased in relative abundance after
@@
 day 7 (Fig. 2B, Fig. S2C). We trained three L2-regularized logistic
-regression models with either input bacterial community data from the
-baseline (day = -1), post-clindamycin (day = 0), or post-infection (day
-= 1) timepoints of the experiment to predict \emph{C. difficile}
-colonization status on day 7 (Fig. S3A-B). All models were better at
-predicting \emph{C. difficile} colonization status on day 7 than random
-chance (all \emph{P} \textless{} 0.001, Table S12). The model based on
-the post-clindamycin (AUROC = 0.78) community OTU data performed
-significantly better than the baseline (AUROC = 0.72) or the
-post-infection (AUROC = 0.67) models (\emph{P}\textsubscript{FDR}
+regression models with either input bacterial community data from the 6
+sources of mice at the baseline (day = -1), post-clindamycin (day = 0),
+or post-infection (day = 1) timepoints of the experiment to predict
+\emph{C. difficile} colonization status on day 7 (Fig. S3A-B). All
+models were better at predicting \emph{C. difficile} colonization status
+on day 7 than random chance (all \emph{P} \textless{} 0.001, Table S12).
+The model based on the post-clindamycin (AUROC = 0.78) community OTU
+data performed significantly better than the baseline (AUROC = 0.72) or
+the post-infection (AUROC = 0.67) models (\emph{P}\textsubscript{FDR}
 \textless{} 0.001 for pairwise comparisons; Fig. S3C and Table S13).
@@
 (\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
-\emph{Porphyromonadaceae}; Fig. 5A). Next, we examined whether any of
-the top 20 ranked OTUs from the post-clindamycin (day 0) model were also
-important in the other 2 classification models based on baseline (day
--1) and 1 day post-infection community data. We identified 6 OTUs that
-were important to the post-clindamycin model and either the baseline or
-1 day post-infection models (\emph{Enterobacteriaceae, Ruminococcaceae,
-Lactobacillus, Bacteroides, Porphyromonadaceae, Erysipelotrichaceae};
-Table S14). Thus, a subset of bacterial OTUs were important for
-determining \emph{C. difficile} colonization dynamics across multiple
-timepoints.
+\emph{Porphyromonadaceae}; Fig. 5A). On day 0, the majority of these
+OTUs were impacted by clindamycin and had relative abundances that were
+close to the limit of detection (Fig. 5A). Next, we examined whether any
+of the top 20 ranked OTUs from the post-clindamycin (day 0) model were
+also important in the other 2 classification models based on baseline
+(day -1) and 1 day post-infection community data. We identified 6 OTUs
+that were important to the post-clindamycin model and either the
+baseline or 1 day post-infection models (\emph{Enterobacteriaceae,
+Ruminococcaceae, Lactobacillus, Bacteroides, Porphyromonadaceae,
+Erysipelotrichaceae}; Table S14). Thus, a subset of bacterial OTUs were
+important for determining \emph{C. difficile} colonization dynamics
+across multiple timepoints.
 
@@
 identify bacterial taxa that were unique to different sources as well as
-taxa that were universally impacted by clindamycin. We trained logistic
-regression models with baseline (day -1), post-clindamycin treatment
-(day 0), and 1-day post-infection fecal community data that could
-predict whether mice cleared \emph{C. difficile} by 7 days
-post-infection better than random chance. We identified
-\emph{Bacteroides, Enterococcus, Enterobacteriaceae, Porphyromonadaceae}
-(Fig. 6) as candidate bacteria within these communities that influenced
-variation in \emph{C. difficile} colonization dynamics since these
-bacteria were all important in the logistic regression models and varied
-by source, were impacted by clindamycin treatment, or both. Overall, our
-results demonstrated clindamycin was sufficient to render mice from
-multiple sources susceptible to CDI and only a subset of the
-inter-individual microbiota variation across mice from different sources
-was needed to predict which mice could clear \emph{C. difficile}.
+taxa that were universally impacted by clindamycin. We trained
+L2-regularized logistic regression models with baseline (day -1),
+post-clindamycin treatment (day 0), and 1-day post-infection fecal
+community data that could predict whether mice cleared \emph{C.
+difficile} by 7 days post-infection better than random chance. We
+identified \emph{Bacteroides, Enterococcus, Enterobacteriaceae,
+Porphyromonadaceae} (Fig. 6) as candidate bacteria within these
+communities that influenced variation in \emph{C. difficile}
+colonization dynamics since these bacteria were all important in the
+logistic regression models and varied by source, were impacted by
+clindamycin treatment, or both. Overall, our results demonstrated
+clindamycin was sufficient to render mice from multiple sources
+susceptible to CDI and only a subset of the inter-individual microbiota
+variation across mice from different sources was needed to predict which
+mice could clear \emph{C. difficile}.
 
@@
 \emph{Enterobacteriaceae} increased in all sources of mice after
-clindamycin treatment, facilitating \emph{C. difficile} colonization.
-However, there was also variation in \emph{Enterobacteriaceae} relative
-abundance levels between sources that was associated with the variation
-in \emph{C. difficile} colonization dynamics across sources. Thus,
-bacteria may have differential roles in determining susceptibility
-depending on the type of bacterial infection.
+clindamycin treatment, positively correlating with \emph{C. difficile}
+colonization. However, there was also variation in
+\emph{Enterobacteriaceae} relative abundance levels between sources that
+was associated with the variation in \emph{C. difficile} colonization
+dynamics across sources. Thus, bacteria may have differential roles in
+determining susceptibility depending on the type of bacterial infection.
 
@@
 infections eventually becoming undetectable while strain BI-7 remained
-detectable up to 70 days post-treatment (44). The bacterial
+detectable up to 70 days post-treatment (44). One study limitation is
+that we only used female mice. Sex has been shown to influence
+microbiota variation in mice (45), so we used female mice to reduce this
+confounding variable and also match the sex used in previous CDI studies
+that administered clindamycin to mice (32, 32, 44, 46). The bacterial
 perturbations induced by clindamycin treatment have been well
@@
 \emph{Ruminococcaceae}, and \emph{Turicibacter} were associated with
-resistance (19, 21, 32, 33, 43--46). While we have demonstrated that
-susceptibility is uniform across sources of mice after clindamycin
+resistance (19, 21, 32, 33, 43, 44, 46, 47). While we have demonstrated
+that susceptibility is uniform across sources of mice after clindamycin
 treatment, there could be different outcomes for either susceptibility
@@
 variation. One alternative approach that has been used in some CDI
-studies is to associate mice with human microbiotas (47--52). However, a
+studies is to associate mice with human microbiotas (48--53). However, a
 major caveat to this method is the substantial loss of human microbiota
-community members upon transfer to mice (53, 54). Additionally with the
-exception of 2 recent studies (47, 48), most of these studies associated
-mice with just 1 types of human microbiota either from a single donor or
-a single pool from multiple donors (49--52). This approach does not aid
+community members upon transfer to mice (54, 55). Additionally, with the
+exception of 2 recent studies (48, 49), most of these studies associated
+mice with just 1 type of human microbiota either from a single donor or
+a single pool from multiple donors (50--53). This approach does not aid
 in the goal of modeling the interpersonal variation seen in humans to
@@
 and these non-bacterial members can also vary across sources of mice
-(55, 56). While our study focused solely on the bacterial portion,
+(56, 57). While our study focused solely on the bacterial portion,
 viruses and fungi have also begun to be implicated in the context of
-CDIs or FMT treatments for recurrent CDIs (35, 57--60). Beyond community
+CDIs or FMT treatments for recurrent CDIs (35, 58--61). Beyond community
 composition, the metabolic function of the microbiota also has a CDI
-signature (20, 46, 61, 62) and can vary across mice from different
-sources (63). For example, microbial metabolites, particularly secondary
+signature (20, 47, 62, 63) and can vary across mice from different
+sources (64). For example, microbial metabolites, particularly secondary
 bile acids and butyrate production, have been implicated as important
@@
 and Th17 cells across mice from different vendors have also been
-documented and (64, 65) could influence the host response to CDI (66,
-67). The outcome after \emph{C. difficile} exposure depends on a
-multitude of factors, including age, diet, and immunity; all of which
-are also influenced by the microbiota.
+documented (65, 66) and could influence the host response to CDI (67,
+68), particularly relevant for \emph{C. difficile} strains that induce
+more severe disease than \emph{C. difficile} 630. The outcome after
+\emph{C. difficile} exposure depends on a multitude of factors,
+including genetics, age, diet, and immunity; all of which also influence
+the microbiota.
 
@@
 described previously (21). Mice weights and stool samples were taken
-daily through 9 days post-challenge (Fig. 1A). Collected stool was split
+daily through 9 days post-infection (Fig. 1A). Collected stool was split
 for \emph{C. difficile} quantification and 16S rRNA sequencing analysis.
@@
 Fisher Scientific) using custom barcoded primers, as previously
-described (68). The ZymoBIOMICS microbial community DNA standards was
-used as a mock community control (69) and water was used as a negative
+described (69). The ZymoBIOMICS microbial community DNA standards was
+used as a mock community control (70) and water was used as a negative
 control per 96-well extraction plate. The PCR amplicons were cleaned up
@@
 \textbf{(v) 16S rRNA gene sequence analysis.} mothur (v. 1.43) was used
-to process all sequences (70) with a previously published protocol (68).
-Reads were combined and aligned with the SILVA reference database (71).
+to process all sequences (71) with a previously published protocol (69).
+Reads were combined and aligned with the SILVA reference database (72).
 Chimeras were removed with the VSEARCH algorithm and taxonomic
 assignment was completed with a modified version (v16) of the Ribosomal
-Database Project reference database (v11.5) (72) with an 80\% confidence
+Database Project reference database (v11.5) (73) with an 80\% confidence
 cutoff. Operational taxonomic units (OTUs) were assigned with a 97\%
-similarity threshold using the opticlust algorithm (73). To account for
-uneven sequencing across samples, samples were rarefied to 5,437
+similarity threshold using the opticlust algorithm (74). Based on the
+mock communities, our overall sequencing error rate was 0.0112\% and all
+water controls had less than 1000 sequences (range: 18-875). To account
+for uneven sequencing across samples, samples were rarefied to 5,437
 sequences 1,000 times for alpha and beta diversity analyses, and a
 single time to generate relative abundances for model training. PCoAs
-were generated based on \(\theta_{YC}\) distances. Permutational
-multivariate analysis of variance (PERMANOVA) was performed on
-mothur-generated \(\theta_{YC}\) distance matrices with the adonis
-function in the vegan package (74) in R (75).
+were generated based on the Yue and Clayton measure of dissimilarity
+(\(\theta_{YC}\)) distances. Permutational multivariate analysis of
+variance (PERMANOVA) was performed on mothur-generated \(\theta_{YC}\)
+distance matrices with the adonis function in the vegan package (75) in
+R (76).
 
@@
 post-infection. The L2-regularized logistic regression models were
-trained and tested using the caret package (76) in R as previously
-described (77) with the exception that we used 60\% training and 40\%
-testing data splits for the cross-validation of the training data to
-select the best cost hyperparameter and the testing of the held out test
-data to measure model performance. The modified training to testing
-ratio was selected to accommodate the small number of samples in the
-dataset. Code was modified from
+trained and tested using the caret package (77) in R as previously
+described (78) with the exception that we used 60\% training and 40\%
+testing data splits for testing of the held out test data to measure
+model performance and repeated k-fold cross-validation of the training
+data to select the best cost hyperparameter. The modified training to
+testing ratio was selected to accommodate the small number of samples in
+the dataset. Code was modified from
 \url{https://github.com/SchlossLab/ML_pipeline_microbiome} to update the
@@
 \textbf{(vii) Statistical analysis.} All statistical tests were
-performed in R (v 4.0.2) (75). The Kruskal-Wallis test was used to
+performed in R (v 4.0.2) (76). The Kruskal-Wallis test was used to
 analyze differences in \emph{C. difficile} CFU, mouse weight change, and
@@
 testing. The tidyverse package (v 1.3.0) was used to wrangle and graph
-data (78).
+data (79).
 
@@
 generating this manuscript is available at
-\url{https://github.com/SchlossLab/Tomkovich_Vendor_XXXX_2020}.
+\url{https://github.com/SchlossLab/Tomkovich_Vendor_mSphere_2020}.
 
@@
 
+\leavevmode\hypertarget{ref-Wang2019}{}%
+45. Wang J, Lang T, Shen J, Dai J, Tian L, Wang X. 2019. Core gut
+bacteria analysis of healthy mice. Frontiers in Microbiology 10.
+
 \leavevmode\hypertarget{ref-Lawley2009}{}%
-45. Lawley TD, Clare S, Walker AW, Goulding D, Stabler RA, Croucher N,
+46. Lawley TD, Clare S, Walker AW, Goulding D, Stabler RA, Croucher N,
 Mastroeni P, Scott P, Raisen C, Mottram L, Fairweather NF, Wren BW,
@@
 \leavevmode\hypertarget{ref-Jump2014}{}%
-46. Jump RLP, Polinkovsky A, Hurless K, Sitzlar B, Eckart K, Tomas M,
+47. Jump RLP, Polinkovsky A, Hurless K, Sitzlar B, Eckart K, Tomas M,
 Deshpande A, Nerandzic MM, Donskey CJ. 2014. Metabolomics analysis
@@
 \leavevmode\hypertarget{ref-NagaoKitamoto2020}{}%
-47. Nagao-Kitamoto H, Leslie JL, Kitamoto S, Jin C, Thomsson KA,
+48. Nagao-Kitamoto H, Leslie JL, Kitamoto S, Jin C, Thomsson KA,
 Gillilland MG, Kuffa P, Goto Y, Jenq RR, Ishii C, Hirayama A, Seekatz
@@
 \leavevmode\hypertarget{ref-Battaglioli2018}{}%
-48. Battaglioli EJ, Hale VL, Chen J, Jeraldo P, Ruiz-Mojica C, Schmidt
+49. Battaglioli EJ, Hale VL, Chen J, Jeraldo P, Ruiz-Mojica C, Schmidt
 BA, Rekdal VM, Till LM, Huq L, Smits SA, Moor WJ, Jones-Hall Y, Smyrk T,
@@
 \leavevmode\hypertarget{ref-Robinson2014}{}%
-49. Robinson CD, Auchtung JM, Collins J, Britton RA. 2014. Epidemic
+50. Robinson CD, Auchtung JM, Collins J, Britton RA. 2014. Epidemic
 \emph{Clostridium difficile} strains demonstrate increased competitive
@@
 \leavevmode\hypertarget{ref-Collins2015}{}%
-50. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. 2015.
+51. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. 2015.
 Humanized microbiota mice as a model of recurrent \emph{Clostridium
@@
 \leavevmode\hypertarget{ref-Collins2018}{}%
-51. Collins J, Robinson C, Danhof H, Knetsch CW, Leeuwen HC van, Lawley
+52. Collins J, Robinson C, Danhof H, Knetsch CW, Leeuwen HC van, Lawley
 TD, Auchtung JM, Britton RA. 2018. Dietary trehalose enhances virulence
@@
 \leavevmode\hypertarget{ref-Hryckowian2018}{}%
-52. Hryckowian AJ, Treuren WV, Smits SA, Davis NM, Gardner JO, Bouley
+53. Hryckowian AJ, Treuren WV, Smits SA, Davis NM, Gardner JO, Bouley
 DM, Sonnenburg JL. 2018. Microbiota-accessible carbohydrates suppress
@@
 \leavevmode\hypertarget{ref-Fouladi2020}{}%
-53. Fouladi F, Glenny EM, Bulik-Sullivan EC, Tsilimigras MCB, Sioda M,
+54. Fouladi F, Glenny EM, Bulik-Sullivan EC, Tsilimigras MCB, Sioda M,
 Thomas SA, Wang Y, Djukic Z, Tang Q, Tarantino LM, Bulik CM, Fodor AA,
@@
 \leavevmode\hypertarget{ref-Walter2020}{}%
-54. Walter J, Armet AM, Finlay BB, Shanahan F. 2020. Establishing or
+55. Walter J, Armet AM, Finlay BB, Shanahan F. 2020. Establishing or
 exaggerating causality for the gut microbiome: Lessons from human
@@
 \leavevmode\hypertarget{ref-Rasmussen2019}{}%
-55. Rasmussen TS, Vries L de, Kot W, Hansen LH, Castro-Mejía JL,
+56. Rasmussen TS, Vries L de, Kot W, Hansen LH, Castro-Mejía JL,
 Vogensen FK, Hansen AK, Nielsen DS. 2019. Mouse vendor influence on the
@@
 \leavevmode\hypertarget{ref-Mims2020}{}%
-56. Mims TS, Abdallah QA, Watts S, White C, Han J, Willis KA, Pierre JF.
+57. Mims TS, Abdallah QA, Watts S, White C, Han J, Willis KA, Pierre JF.
 2020. Variability in interkingdom gut microbiomes between different
@@
 \leavevmode\hypertarget{ref-Stewart2019}{}%
-57. Stewart DB, Wright JR, Fowler M, McLimans CJ, Tokarev V, Amaniera I,
+58. Stewart DB, Wright JR, Fowler M, McLimans CJ, Tokarev V, Amaniera I,
 Baker O, Wong H-T, Brabec J, Drucker R, Lamendella R. 2019. Integrated
@@
 \leavevmode\hypertarget{ref-Ott2017}{}%
-58. Ott SJ, Waetzig GH, Rehman A, Moltzau-Anderson J, Bharti R, Grasis
+59. Ott SJ, Waetzig GH, Rehman A, Moltzau-Anderson J, Bharti R, Grasis
 JA, Cassidy L, Tholey A, Fickenscher H, Seegert D, Rosenstiel P,
@@
 \leavevmode\hypertarget{ref-Zuo2017}{}%
-59. Zuo T, Wong SH, Lam K, Lui R, Cheung K, Tang W, Ching JYL, Chan PKS,
+60. Zuo T, Wong SH, Lam K, Lui R, Cheung K, Tang W, Ching JYL, Chan PKS,
 Chan MCW, Wu JCY, Chan FKL, Yu J, Sung JJY, Ng SC. 2017. Bacteriophage
@@
 \leavevmode\hypertarget{ref-Zuo2018}{}%
-60. Zuo T, Wong SH, Cheung CP, Lam K, Lui R, Cheung K, Zhang F, Tang W,
+61. Zuo T, Wong SH, Cheung CP, Lam K, Lui R, Cheung K, Zhang F, Tang W,
 Ching JYL, Wu JCY, Chan PKS, Sung JJY, Yu J, Chan FKL, Ng SC. 2018. Gut
@@
 \leavevmode\hypertarget{ref-Robinson2019}{}%
-61. Robinson JI, Weir WH, Crowley JR, Hink T, Reske KA, Kwon JH, Burnham
+62. Robinson JI, Weir WH, Crowley JR, Hink T, Reske KA, Kwon JH, Burnham
 C-AD, Dubberke ER, Mucha PJ, Henderson JP. 2019. Metabolomic networks
@@
 \leavevmode\hypertarget{ref-Fletcher2018}{}%
-62. Fletcher JR, Erwin S, Lanzas C, Theriot CM. 2018. Shifts in the gut
+63. Fletcher JR, Erwin S, Lanzas C, Theriot CM. 2018. Shifts in the gut
 metabolome and \emph{Clostridium difficile} transcriptome throughout
@@
 \leavevmode\hypertarget{ref-Xiao2015}{}%
-63. Xiao L, Feng Q, Liang S, Sonne SB, Xia Z, Qiu X, Li X, Long H, Zhang
+64. Xiao L, Feng Q, Liang S, Sonne SB, Xia Z, Qiu X, Li X, Long H, Zhang
 J, Zhang D, Liu C, Fang Z, Chou J, Glanville J, Hao Q, Kotowska D,
@@
 \leavevmode\hypertarget{ref-Fransen2015}{}%
-64. Fransen F, Zagato E, Mazzini E, Fosso B, Manzari C, Aidy SE,
+65. Fransen F, Zagato E, Mazzini E, Fosso B, Manzari C, Aidy SE,
 Chiavelli A, D'Erchia AM, Sethi MK, Pabst O, Marzano M, Moretti S,
@@
 \leavevmode\hypertarget{ref-Ivanov2009}{}%
-65. Ivanov II, Atarashi K, Manel N, Brodie EL, Shima T, Karaoz U, Wei D,
+66. Ivanov II, Atarashi K, Manel N, Brodie EL, Shima T, Karaoz U, Wei D,
 Goldfarb KC, Santee CA, Lynch SV, Tanoue T, Imaoka A, Itoh K, Takeda K,
@@
 \leavevmode\hypertarget{ref-Azrad2018}{}%
-66. Azrad M, Hamo Z, Tkhawkho L, Peretz A. 2018. Elevated serum
+67. Azrad M, Hamo Z, Tkhawkho L, Peretz A. 2018. Elevated serum
 immunoglobulin a levels in patients with \emph{Clostridium difficile}
@@
 \leavevmode\hypertarget{ref-Saleh2019}{}%
-67. Saleh MM, Frisbee AL, Leslie JL, Buonomo EL, Cowardin CA, Ma JZ,
+68. Saleh MM, Frisbee AL, Leslie JL, Buonomo EL, Cowardin CA, Ma JZ,
 Simpson ME, Scully KW, Abhyankar MM, Petri WA. 2019. Colitis-induced
@@
 \leavevmode\hypertarget{ref-Kozich2013}{}%
-68. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
+69. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \leavevmode\hypertarget{ref-Sze2019}{}%
-69. Sze MA, Schloss PD. 2019. The impact of DNA polymerase and number of
+70. Sze MA, Schloss PD. 2019. The impact of DNA polymerase and number of
 rounds of amplification in PCR on 16S rRNA gene sequence data. mSphere
@@
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-70. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+71. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
@@
 \leavevmode\hypertarget{ref-Quast2012}{}%
-71. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
+72. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
 J, Glöckner FO. 2012. The SILVA ribosomal RNA gene database project:
@@
 \leavevmode\hypertarget{ref-Cole2013}{}%
-72. Cole JR, Wang Q, Fish JA, Chai B, McGarrell DM, Sun Y, Brown CT,
+73. Cole JR, Wang Q, Fish JA, Chai B, McGarrell DM, Sun Y, Brown CT,
 Porras-Alfaro A, Kuske CR, Tiedje JM. 2013. Ribosomal database project:
@@
 \leavevmode\hypertarget{ref-Westcott2017}{}%
-73. Westcott SL, Schloss PD. 2017. OptiClust, an improved method for
+74. Westcott SL, Schloss PD. 2017. OptiClust, an improved method for
 assigning amplicon-based sequence data to operational taxonomic units.
@@
 \leavevmode\hypertarget{ref-Vegan2018}{}%
-74. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+75. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 Minchin PR, O'Hara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
@@
 \leavevmode\hypertarget{ref-r_citation_2018}{}%
-75. R Core Team. 2018. R: A language and environment for statistical
+76. R Core Team. 2018. R: A language and environment for statistical
 computing. R Foundation for Statistical Computing, Vienna, Austria.
@@
 \leavevmode\hypertarget{ref-Kuhn2008}{}%
-76. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
+77. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
 Journal of Statistical Software 28.
@@
 \leavevmode\hypertarget{ref-Topcuoglu2020}{}%
-77. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
+78. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
 framework for effective application of machine learning to
@@
 \leavevmode\hypertarget{ref-Tidyverse2019}{}%
-78. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
+79. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
 Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
@@
 abundances for the 20 most significant OTUs out of the 268 OTUs that
-varied across sources at baseline (Table S5).
+varied across sources at baseline by Kruskal-Wallis test with
+Benjamini-Hochberg correction (Table S5).
 
@@
 Jackson Laboratory (N = 8), Charles River Laboratory (N = 8), Taconic
-Biosciences (N = 8), and Envigo (N = 8). All mice were administered 10
-mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with
-\emph{C. difficile} 630 spores on day 0. Mice were weighed and feces was
-collected daily through the end of the experiment (9 days
-post-infection). Note: 3 mice died during course of experiment. 1
-Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse
+Biosciences (N = 8), and Envigo (N = 8). Mice that were ordered from
+commercial vendors acclimated to the University of Michigan mouse
+facility for 13 days prior to antibiotic administration. All mice were
+administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before
+challenge with \emph{C. difficile} 630 spores on day 0. Mice were
+weighed and feces was collected daily through the end of the experiment
+(9 days post-infection). Note: 3 mice died during course of experiment.
+1 Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse
 between 1- and 3-days post-infection. B. \emph{C. difficile} CFU/gram
@@
 for an individual mouse. Gray bars indicate the median relative
-abundances for each source. B. Venn diagram that combines OTUs that were
+abundances for each source. The gray horizontal lines indicates the
+limit of detection. B. Venn diagram that combines OTUs that were
 important to the day -1, 0, and 1 classification models ( Fig. S4, Table
@@
 can predict \emph{C. difficile} colonization status 7 days
-post-challenge.} A. Bar graph visualizations of overall 7-days
+post-infection.} A. Bar graph visualizations of overall 7-days
 post-infection \emph{C. difficile} colonization status that were used as

### 2020-09-21 submission/response_to_reviewers.md
Add response to reviewers for revision

---
output:
  pdf_document: default
  html_document: default
---
### Reviewer #1 (Comments for the Author):

**In this manuscript the authors investigate the generally ignored contribution of the source of animals in microbiota analyses, by determining the microbiota profiles of mice from 6 sources during antibiotic treatment (clindamycin) and subsequent challenge with a toxigenic C. difficile strain. This allowed them to determine genera that are consistently impacted, irrespective of source, and developed models that were predictive of C. difficile clearance based on key species.**

**The study has implications for the way microbiota studies are reported and raises a number of interesting points regarding variability between subjects in microbiome analysis; the transparency with respect to code and data is commendable. I also appreciate the fact that the authors took care to formulate a title that does justice to the limitations of the study ("in mice"). Nevertheless, I would suggest that the use of "C. difficile clearance" more accurately reflects the content of the manuscript (as authors also state in L232-233 that susceptibility, possibly associated with colonization, is not affected).**

> We thank the reviewer for this suggestion and have incorporated clearance into the manuscript title:
"The initial gut microbiota and response to antibiotic perturbation influence *Clostridioides difficile* clearance in mice"

**Generally, experiments are performed well, conclusions justified, and the manuscript is well written.**

**There are some issues that I would like to suggest to improve the manuscript further and to facilitate this I provide detailed comments below.**

**The authors do not define colonization. E.g. in the Abstract it is written that: "all mice were colonized 1-day post-infection" (L8-10). Are the authors sure this is colonization, not passage? The same can be said at multiple points throughout the manuscript (e.g. L101 "the colonization dynamics were siumilar"). I think the authors should either define their definition of colonization specifically, or rephrase (in other places this is done more accurately, e.g. "C. difficile clearance", "C. difficile was detectable"). In the reviewer's opinion, initial CFU likely represent transient presence of C. difficile upon outgrowth, but do not reflect (stable) colonization.**

> Our mouse model challenges mice with a low dose of *C. difficile* spores instead of vegetative cells to ensure the *C. difficile* we detect post-challenge germinated and expanded in the intestine. We have modified the following sentence to indicate colonization is transient in the clindamycin mouse model:
"We have previously demonstrated mice are rendered susceptible to *C. difficile*, but clear the pathogen within 9 days, thus colonization is transient when treated with clindamycin alone (21, 43)."

**There is limited discussion on the fact that the mouse model used allows recovery of nearly all mice (630 infections in mice are rarely lethal). It would be appropriate to present in the Discussion a paragraph on the possible implications of their findings for strains that could cause more severe disease.**

> We acknowlege that *C. difficile* 630 does produce mild infections in mice compared to other strains. It is interesting that we did see 2 deaths after *C. difficile* challenge with *C. difficile* 630. We expect that differences across sources of mice could yield to variation in disease severity and have modified the following sentence in the discussion to address this point:
"Evidence for immunological toning differences in IgA and Th17 cells across mice from different vendors have also been documented (65, 66) and could influence the host response to CDI (67, 68), particularly relevant for strains that induce more severe disease than *C. difficile* 630."

**There appears to be an issue with Supplemental Figure 2, as text refers to panels up to H, but Figure only contains panels up to E. Possibly, some panels have been lost during conversion. Please check. Also, the figure has poorly visible panels C-E. I suggest placing A and B next to each other, and increase the size of panels C-E and place them underneath panels A and B.**

> Thank you for calling attention to this discrepancy, we have corrected the text to refer to Fig. S1, which does have the correct number of panels:
"There were some differences between the 2 experiments we conducted, as the experiment and cage effects significantly explained the observed community variation for the Schloss and Young lab mouse colonies (Fig. S1A-B and Table S4). However, most of the vendors also clustered by experiment (Fig. S1C-D, F), suggesting there was some community variation between the 2 experiments within each source, particularly for Schloss, Young, and Envigo mice (Fig. S1G-H)."

>We have also rearranged Fig. S2 as suggested by the reviewer.

**Authors have longitudinal data for all experiments (daily fecal samples), but appear not to leverage this potential as only pairwise comparisons appear to have been made. Methods designed for repeated measurements instead of cross-sectional based methods like Metalonda or MetaSplines could be used. Statistics on the longitudinal dataset could give insight into the consistency of the significantly different abundance of OTUs between colonized/cleared mice at day 7, and could strengthen the findings (also relevant for Figure 6).**

> Thank you for the suggestion, we opted for comparisons at specific timepoints because of the variation in the mice we had samples and sequence data for each day. We did test whether any taxa  significantly varied between the colonized/cleared mice on day 7, but none were significant after correcting for multiple hypothesis testing. We suspect that the majority of the fluctuations occured during the first 3 days of the experiment.

**Minor comments**
**1. L6: please include the region used (V4)**

> Due to abstract word count constraints we specify that we sequenced the V4 region when we first mention 16S rRNA sequencing in the results section (L68, see point 3 below).

**2. L64: were the Michigan mice originally obtained from a commercial vendor? If so, which one?**

> Yes, the Michigan mice were originally obtained from Jackson in 2002, which is stated in the Material and Methods section.

**3. L68: "sequenced the 16S rRNA genes" or "performed 16S rRNA sequencing"**

> We have modified to specify the region sequenced, per the reviewer's suggestion with point 1 above:
"sequenced the V4 region of the 16S rRNA gene""

**4. L70 and on: the section states that "after they acclimatized". It is unclear how this relates to the study setup, in Figure 2 for instance. I would welcome a more detailed description on this.**

> In the Materials and Methods section we define the acclimatization period as the 13 days between when the mice ordered from commercial vendors arrived at the University of Michigan and the start of the experiment. We have added the following sentection to the Figure 2A legend to define the acclimatization period:
"Mice that were ordered from commercial vendors acclimated to the University of Michigan mouse facility for 13 days prior to antibiotic administration."

**5. L83-87: in the reviewer's view, it is not entirely clear what comparisons are made (which groups, timepoints). Can they authors clarify?**

> We are comparing OTUs that vary between the 6 sources and OTUs within a single source that vary between experiments. Both comparisons are referring to the baseline (day -1 timepoint). We have modified the sentences to clarify:
"After finding differences at the community level, we next identified the bacteria that varied between the 6 sources of mice. There were 268 OTUs with relative abundances that were significantly different between the sources at baseline (Fig. 1D and Table S5). Though we saw differences between experiments at the community level, there were no OTUs that were significantly different between experiments within Schloss, Young, and Envigo mice at baseline (all *P* > 0.05)."

**6. L84: please place the number in perspective to the total number of OTUs detected (is 268 a minor or a major fraction of what you detect?).**

> After samples were rarefied to 5,437 sequences, we detected 1,935 OTUs in our dataset. We display the number of OTUs (richness) detected in each mouse in Fig. 1A, 3A, and 4A to place the number of OTUs that vary between sources of mice in context.

**7. L101-102: to clear - better "to clear C. difficile"**

> We have added *C. difficile* to the sentence as suggested by the reviewer:
"While the colonization dynamics were similar between the two experiments, the Schloss mice took longer to clear *C. difficile* in the first experiment compared to the second and the Envigo mice took longer to clear *C. difficile* in the second experiment compared to the first (Fig. S2A-B)."

**8. L125-126: as earlier (see point 5), I would appreciate a clarification with regard to for instance timepoint.**

> We have modified the sentence to indicate the timepoint:
"However, there were only 18 OTUs with relative abundances that significantly varied between sources after clindamycin treatment (Fig. 3D and Table S8)."

**9. L135: "was primarily found only", suggest removing only or primarily**

> We have removed only as suggested:
"*Enterococcus* was primarily found in mice purchased from commercial vendors and also increased in relative abundance after clindamycin treatment (Fig. 3D)."

**10. L165-167: samples were grouped per timepoint (different sources of mice were put in 3 groups depending on timepoint), correct? Please clarify.**

> That's correct, we took the OTU relative abundances from all sources of mice that had sequence data at the 3 timepoints: baseline (day -1), post-clindamycin (day 0), and post-infection (day 1). We have modified the sentence to clarify:
"We trained three L2-regularized logistic regression models with either input bacterial community data from the 6 sources of mice at the baseline (day = -1), post-clindamycin (day = 0), or post-infection (day = 1) timepoints of the experiment to predict *C. difficile* colonization status on day 7 (Fig. S3A-B)."

**11. L208: Perhaps more accurate to state 'ridge regression model' than logistic regression. A standard logistic regression model does normally not include any penalization.**

> For consistency with how we refer to the logistic regression models throughout the rest of the manuscript, we have added L2-regularized to specify the type of model:
"We trained L2-regularized logistic regression models with baseline (day -1), post-clindamycin treatment (day 0), and 1-day post-infection fecal community data that could predict whether mice cleared *C. difficile* by 7 days post-infection better than random chance."

**12. L221: facilitating is overstating. The data shows that it is associated with colonization, but facilitating would require experimental evidence to this effect.**

> We have modified facilitating to positively correlating:
"We found *Enterobacteriaceae* increased in all sources of mice after clindamycin treatment, positively correlating with *C. difficile* colonization"

**13. L257; certain mouse models use antibiotics other than clindamycin. I would appreciate a statement to this effect (how should the authors' findings be seen in this respect), considering that clindamycin treatment appears to be the dominant factor in determining susceptibility.**

> We agree that different antibiotic treatments will likely result in different *C. difficle* infection outcomes in mice. We refer the reviewer to L248-250 in the discussion section of the manuscript: "While we have demonstrated that susceptibility is uniform across sources of mice after clindamycin treatment, there could be different outcomes for either susceptibility or clearance in the case of other antibiotic treatments."

**14. L262: comma missing after Additionally**

> We have added the comma after Additionally.

**15. L263: types should be type**

> We have corrected to type.

**16. L293 and 294: I think the authors mean to state that all these factors influence the microbiota, not the other way around? (at least for diet and age)**

> That is what we meant and we have corrected to the following:
"The outcome after *C. difficile* exposure depends on a multitude of factors, including age, diet, and immunity; all of which also influence the microbiota."

**17. L346-348: please include a single sentence on the results of the controls, to verify quality of processing.**

> We thank the reviewer for suggesting we add in the control results. We have added the following sentence to the Materials and Methods section:
"Based on the mock communities, our overall sequencing error rate was 0.0112% and all  water controls had less than 1000 sequences."

**18. Line 360-362: Could the authors perhaps give some explanation on the used distance metrics? I am not familiar with this specific metric. More importantly, it is unclear to this reviewer whether the assumption of homogeneity of dispersion between the comparison groups is tested (can be done using the betadisper function). It is known that even if centroids of comparison groups are very close together in e.g. a PCoA plot, adonis can still yield a very significant value due to different dispersions between the groups. You would then not be looking at actual different compositions between groups, but at (large) differences in composition within a group. If this is the case for the current analyses, it would be useful to report this, or an alternative (recent) test could be used which is robust to this issue (https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-019-0659-9).**

> $\theta_{YC}$ distances are based on the Yue & Clayton measure of dissimilarity. We have modified the Materials and Methods Section to clarify:
"PCoAs were generated based on the Yue and Clayton measure of dissimilarity ($\theta_{YC}$) distances."

> Regarding the PERMANOVA tests used to analyze overall differences between the community structures of the 6 sources of mice, we agree that there are caveats to this approach. We also determined the specific OTUs with different relative abundances between sources of mice to determine which OTUs were contributing to the overall differences between the community structures.

**19. L371: please provide some additional detail about the cross-validation (e.g. k-fold cross validation, repetition of cross-validations). Why did the authors opt for a ridge regression and not a lasso or elastic net regression?**

> We have modified the sentence to include additional detail on the cross-validation:
"The L2-regularized logistic regression models were trained and tested using the caret package (77) in R as previously described (78) with the exception that we used 60% training and 40% testing data splits for testing of the held out test data to measure model performance and repeated k-fold cross-validation of the training data to select the best cost hyperparameter."

> We chose ridge (L2-regularized) logistic regression because our group has previously implemented this model with microbiome data (Topcuoglu et al mBio 2020) and found the L2-regularized logistic regression model performed almost as well as a random forest model when classifying colorectal cancer cases using fecal microbiota data.

**20. L377-378: have the authors considerd using a linear mixed model to investigate the dynamics of C. difficile clearance over time? (as I understand it now per-timepoint comparisons were performed).**

> The reviewer is correct in that we chose per timepoint comparisons. We chose to analyze at specific timepoints because we did not have datapoints for *C. difficile* CFU and sequencing data for each mouse across every timepoint.

**21. In general, is anything known about the genetic make-up of the mice from the different sources? Are there distinct genetic differences that could underlie the differences in microbiota?**

> Yes, there is evidence of genetic differences across different C57BL/6 substrains that could influence the microbiota. Teasing out the degree to which environmental factors and genetic differences contribute to the microbiota differences we observed between the sources mice was outside the scope of our current study. We have modified the following sentence to acknowledge genetics as an additional factor influencing the microbiota.
"The outcome after *C. difficile* exposure depends on a multitude of factors, including genetics, age, diet, and immunity; all of which also influence the microbiota. "

**22. Fig. 1D: please indicate which test was used.**

> We have added that OTUs were identified by the Kruskal-Wallis test with Benjamini-Hochberg correction to the Fig. 1D legend.

**23. Fig. 3: what is the reason for using high values for the y-axis, when no datapoints are shown in the higher regions? Would it be better to use a narrower scale so smaller differences between groups can be better appreciated?**

> The high values on the y-axis of Fig. 3A-B allow comparison with Fig. 1A-B and 4A-B plots, which have the same y-axis scale and highlight the overall impact of clindamycin on community diversity across all sources of mice.

**24. Fig 5A: the horizontal light grey bars at the same position in each panel are not defined. What does this indicate?**

>This line indicates the limit of detection, we have added the following to the Fig. 5A legend to clarify:
"The gray horizontal lines indicates the limit of detection."

### Reviewer #2 (Comments for the Author):

**The initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice**

**The study by Tomkovich et al. shows how the starting microbiota from mice purchased from six different vendors is different. They also define how the microbiota is altered with the addition of clindamycin followed by C. difficile (CD) challenge and monitor colonization over a period of 9 days. The main take a ways from this study are that mice from different vendors, including in house colonies, have different starting microbiotas. When they are treated with clindamycin to render them susceptible to CD colonization, they again are altered differently. From here the authors assessed CD load and saw that colonization patterns were affected differently. They went on to create a model to determine which OTUs were associated with CD clearance and or persistence over this time period. Again, adding to the list of bacteria that are potentially protective and or antagonistic to CD colonization.**

**This is a well written and well powered study. I also appreciate the transparency of the study by including many supplemental files that include all the data. I also think it will be received very well from the field as it is showing how we could potentially get different results based on vendor and to consider this when working with CD mouse models. The study does fall short on mechanism and does not dig into how these bacteria are able to contribute to clearance, BUT that was not the point of the study. I do have a few minor comments that I think it would be nice for the authors to address below. Overall, very nice study!**

**Minor comments:**

**1.The authors use post infection vs. post challenge in the paper. Since they are not looking at disease I would suggest they go with post challenge throughout the paper. Also, did they look at disease at all throughout as this would have also been really interesting? I know they looked at weight loss, but this does not always correlate with disease. I also see that there were three mice that died during the experiments. Can the authors comment on why? Was this from CDI?**

>  For consistency with our group's previously published studies with *C. difficile* 630, we have changed all instances of post-challenge to post-infection. Our hypothesis going into this study was the microbiota variation across different sources of mice would impact *C. difficile* colonization resistance. Previous studies have characterized *C. difficile* 630, and found it to be mild in terms of the amount of histological colitis and weight loss that was observed in contrast to mice infected with VPI 10463 (Theriot et al. Gut Microbes 2011). Thus, we opted to focus on clearance and not disease severity. We suspect that the 2 deaths we observed in our study that occured post-infection were from the CDI (the other death happened prior to infection and was unrelated). Reviewer #1 had a similar comment related to the strain we were using and we have modified a sentence in the discussion to address disease severity variation between *C. difficile* strains:
"Evidence for immunological toning differences in IgA and Th17 cells across mice from different vendors have also been documented and could influence the host response to CDI, particularly relevant for strains that induce more severe disease than *C. difficile* 630."

**2.I see that only female mice were used in this study. Can the author's note this as a limitation in the discussion and or justify why they did this?**

> Since sex has been shown to influence microbiota composition in mice (Wang et al. Frontiers in Microbiology 2019 "Core gut bacteria analysis of healthy mice"), we opted to eliminate this confounding factor by using only female mice. Additionally, we were interested in placing our findings in context to studies that have observed different *C. difficile* infection outcomes despite using similar mouse models and several of these studies also used only female mice. We have added the following to the discussion to highlight this limitation:
"One study limitation is that we only used female mice. Sex has been shown to influence microbiota variation in mice (45), so we used female mice to reduce this confounding variable and also match the sex used in previous CDI studies that administered clindamycin to mice (32, 33, 44, 46)."

**3.I found Figure 5A to confusing. I would suggest trying to make this clearer for the reader.**

> Figure 5A displays the 10 highest ranked OTUs in our model with the best performance. We have added the following sentence to the text to clarify:
"On day 0, the majority of these OTUs were impacted by clindamycin and had relative abundances that were close to the limit of detection (Fig. 5A)."

### 2020-09-25 submission/response_to_reviewers.md
Make double spaced
Add edits from Pat

@@
   html_document: default
+header-includes:
+    - \usepackage{setspace}\doublespacing  
 ---
@@
 > $\theta_{YC}$ distances are based on the Yue & Clayton measure of dissimilarity. We have modified the Materials and Methods Section to clarify:
-"PCoAs were generated based on the Yue and Clayton measure of dissimilarity ($\theta_{YC}$) distances."
+"PCoAs were generated based on the Yue and Clayton measure of dissimilarity ($\theta_{YC}$) distances (75)."
 
-> Regarding the PERMANOVA tests used to analyze overall differences between the community structures of the 6 sources of mice, we agree that there are caveats to this approach.
-We also determined the specific OTUs with different relative abundances between sources of mice to determine which OTUs were contributing to the overall differences between the community structures.
+> Regarding the PERMANOVA tests used to analyze overall differences between the community structures of the 6 sources of mice, we agree that there are numerous approaches to differentiate microbial community structures.
+In our previous work we found that heteroscedacticity did not affect the Type I error of the PERMANOVA test (see AMOVA data in first column of Table 1 from DOI: 10.1038/ismej.2008.5).
+Regardless, the data for which PERMANOVA was used in the current study are pretty obviously different when looking at the ordinations (e.g.
+see Fig 1C).
+Furthermore, when we investigated the suggested test, there was not much documentation in the package to show us how to use the tool and it was not clear that it was possible to employ more sophisticated experimental designs than a single classification ANOVA.
+In contrast, we were able to use vegan to build a statistical model to account for the vendor and experiment.
+Based on these points, we are confident in the results of the statistical tests using PERMANOVA as implemented in the vegan R package.
 
@@
 > We have modified the sentence to include additional detail on the cross-validation:
-"The L2-regularized logistic regression models were trained and tested using the caret package (77) in R as previously described (78) with the exception that we used 60% training and 40% testing data splits for testing of the held out test data to measure model performance and repeated k-fold cross-validation of the training data to select the best cost hyperparameter."
+"The L2-regularized logistic regression models were trained and tested using the caret package (78) in R as previously described (79) with the exception that we used 60% training and 40% testing data splits for testing of the held out test data to measure model performance and repeated 2.5-fold cross-validation of the training data to select the best cost hyperparameter."
 
-> We chose ridge (L2-regularized) logistic regression because our group has previously implemented this model with microbiome data (Topcuoglu et al mBio 2020) and found the L2-regularized logistic regression model performed almost as well as a random forest model when classifying colorectal cancer cases using fecal microbiota data.
+> We chose ridge (L2-regularized) logistic regression because our group has previously implemented this model with microbiome data (Topcuoglu et al mBio 2020) and found the L2-regularized logistic regression model performed almost as well as a random forest model when classifying colorectal cancer cases using fecal microbiota data and did not require as many samples as random forest to avoid over fitting the data.
 
@@
 
->This line indicates the limit of detection, we have added the following to the Fig. 5A legend to clarify:
-"The gray horizontal lines indicates the limit of detection."
+>These lines indicatee the limit of detection, we have added the following to the Fig. 5A legend to clarify:
+"The gray horizontal lines indicate the limit of detection."
 

### 2020-09-25 submission/manuscript.Rmd
Add edits from Pat

@@
 To account for uneven sequencing across samples, samples were rarefied to 5,437 sequences 1,000 times for alpha and beta diversity analyses, and a single time to generate relative abundances for model training.
-PCoAs were generated based on the Yue and Clayton measure of dissimilarity ($\theta_{YC}$) distances.
+PCoAs were generated based on the Yue and Clayton measure of dissimilarity ($\theta_{YC}$) distances [@Yue2005].
 Permutational multivariate analysis of variance (PERMANOVA) was performed on mothur-generated $\theta_{YC}$ distance matrices with the adonis function in the vegan package [@Vegan2018] in R [@r_citation_2018].
@@
 difficile* colonization status 7 days post-infection.
-The L2-regularized logistic regression models were trained and tested using the caret package [@Kuhn2008] in R as previously described [@Topcuoglu2020] with the exception that we used 60% training and 40% testing data splits for testing of the held out test data to measure model performance and repeated k-fold cross-validation of the training data to select the best cost hyperparameter.
+The L2-regularized logistic regression models were trained and tested using the caret package [@Kuhn2008] in R as previously described [@Topcuoglu2020] with the exception that we used 60% training and 40% testing data splits for testing of the held out test data to measure model performance and repeated 2.5-fold cross-validation of the training data to select the best cost hyperparameter.
 The modified training to testing ratio was selected to accommodate the small number of samples in the dataset.
@@
 Gray bars indicate the median relative abundances for each source.
-The gray horizontal lines indicates the limit of detection.
+The gray horizontal lines indicate the limit of detection.
 B.

### 2020-09-25 submission/manuscript.tex
Add edits from Pat

@@
 confounding variable and also match the sex used in previous CDI studies
-that administered clindamycin to mice (32, 32, 44, 46). The bacterial
+that administered clindamycin to mice (32, 33, 44, 46). The bacterial
 perturbations induced by clindamycin treatment have been well
@@
 were generated based on the Yue and Clayton measure of dissimilarity
-(\(\theta_{YC}\)) distances. Permutational multivariate analysis of
+(\(\theta_{YC}\)) distances (75). Permutational multivariate analysis of
 variance (PERMANOVA) was performed on mothur-generated \(\theta_{YC}\)
-distance matrices with the adonis function in the vegan package (75) in
-R (76).
+distance matrices with the adonis function in the vegan package (76) in
+R (77).
 
@@
 post-infection. The L2-regularized logistic regression models were
-trained and tested using the caret package (77) in R as previously
-described (78) with the exception that we used 60\% training and 40\%
+trained and tested using the caret package (78) in R as previously
+described (79) with the exception that we used 60\% training and 40\%
 testing data splits for testing of the held out test data to measure
-model performance and repeated k-fold cross-validation of the training
+model performance and repeated 2.5-fold cross-validation of the training
 data to select the best cost hyperparameter. The modified training to
@@
 \textbf{(vii) Statistical analysis.} All statistical tests were
-performed in R (v 4.0.2) (76). The Kruskal-Wallis test was used to
+performed in R (v 4.0.2) (77). The Kruskal-Wallis test was used to
 analyze differences in \emph{C. difficile} CFU, mouse weight change, and
@@
 testing. The tidyverse package (v 1.3.0) was used to wrangle and graph
-data (79).
+data (80).
 
@@
 
+\leavevmode\hypertarget{ref-Yue2005}{}%
+75. Yue JC, Clayton MK. 2005. A similarity measure based on species
+proportions. Communications in Statistics - Theory and Methods
+34:2123--2131.
+
 \leavevmode\hypertarget{ref-Vegan2018}{}%
-75. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+76. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 Minchin PR, O'Hara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
@@
 \leavevmode\hypertarget{ref-r_citation_2018}{}%
-76. R Core Team. 2018. R: A language and environment for statistical
+77. R Core Team. 2018. R: A language and environment for statistical
 computing. R Foundation for Statistical Computing, Vienna, Austria.
@@
 \leavevmode\hypertarget{ref-Kuhn2008}{}%
-77. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
+78. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
 Journal of Statistical Software 28.
@@
 \leavevmode\hypertarget{ref-Topcuoglu2020}{}%
-78. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
+79. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
 framework for effective application of machine learning to
@@
 \leavevmode\hypertarget{ref-Tidyverse2019}{}%
-79. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
+80. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
 Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
@@
 for an individual mouse. Gray bars indicate the median relative
-abundances for each source. The gray horizontal lines indicates the
-limit of detection. B. Venn diagram that combines OTUs that were
-important to the day -1, 0, and 1 classification models ( Fig. S4, Table
-S14) and either overlapped with taxa that varied across sources at the
-same timepoint, were impacted by clindamycin treatment, or both. Bold
-OTUs were important to more than 1 classification model.
+abundances for each source. The gray horizontal lines indicate the limit
+of detection. B. Venn diagram that combines OTUs that were important to
+the day -1, 0, and 1 classification models ( Fig. S4, Table S14) and
+either overlapped with taxa that varied across sources at the same
+timepoint, were impacted by clindamycin treatment, or both. Bold OTUs
+were important to more than 1 classification model.
 

### 2020-09-25 submission/manuscript.Rmd
Remove figures from manuscript, prepare files for submitting revision

@@
 ```
-\includegraphics{figure_1.pdf}
 **Figure 1.
@@
 \newpage
-\includegraphics{figure_2.pdf}
-
 **Figure 2.
@@
 \newpage
-\includegraphics{figure_3.pdf}
 **Figure 3.
@@
 \newpage
-\includegraphics{figure_4.pdf}
 **Figure 4.
@@
 \newpage
-\includegraphics{figure_5.pdf}
 **Figure 5.
@@
 \newpage
-\includegraphics{figure_6.pdf}
 **Figure 6: OTUs associated with *C.
@@
 \newpage
-\includegraphics{figure_S1.pdf}
-
 **Figure S1.
@@
 \newpage
-\includegraphics{figure_S2.pdf}
 **Figure S2.
@@
 \newpage
-\includegraphics{figure_S3.pdf}
 **Figure S3.
@@
 \newpage
-\includegraphics{figure_S4.pdf}
-
 **Figure S4.

### 2020-09-25 submission/manuscript.tex
Remove figures from manuscript, prepare files for submitting revision

@@
 
-\includegraphics{figure_1.pdf} \textbf{Figure 1. Microbiota variation is
-high between mice from different sources.} A-B. Number of observed OTUs
-(A) and Shannon diversity index values (B) across sources of mice at
-baseline (day -1 of the experiment). Differences between sources were
-analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for
-testing each day of the experiment and the adjusted \emph{P} value was
-\textless{} 0.05 for panel A (Table S1). None of the \emph{P} values
-from pairwise Wilcoxon comparisons between sources were significant
-after Benjamini-Hochberg correction (Table S2). Gray lines represent the
+\textbf{Figure 1. Microbiota variation is high between mice from
+different sources.} A-B. Number of observed OTUs (A) and Shannon
+diversity index values (B) across sources of mice at baseline (day -1 of
+the experiment). Differences between sources were analyzed by
+Kruskal-Wallis test with Benjamini-Hochberg correction for testing each
+day of the experiment and the adjusted \emph{P} value was \textless{}
+0.05 for panel A (Table S1). None of the \emph{P} values from pairwise
+Wilcoxon comparisons between sources were significant after
+Benjamini-Hochberg correction (Table S2). Gray lines represent the
 median values for each source of mice. C. Principal Coordinates Analysis
@@
 \newpage
-
-\includegraphics{figure_2.pdf}
 
@@
 
-\includegraphics{figure_3.pdf} \textbf{Figure 3. Clindamycin treatment
-alters bacteria in all sources, but a subset of bacterial differences
-across sources persists.} A-B. Number of observed OTUs (A) and Shannon
-diversity index values (B) across sources of mice after clindamycin
-treatment (day 0). Differences between sources were analyzed by
-Kruskal-Wallis test with Benjamini-Hochberg correction for testing each
-day of the experiment and the adjusted \emph{P} value was \textless{}
-0.05 (Table S1). Significant \emph{P} values from the pairwise Wilcoxon
-comparisons between sources with Benjamini-Hochberg correction are
-displayed as the first initial of each group compared to the group that
-they are listed above (Table S2). C. PCoA of \(\theta_{YC}\) distances
-from stools collected post-clindamycin. Source and the interaction
-between source and cage effects explained most of the variation observed
-post-clindamycin (PERMANOVA combined R\textsuperscript{2} = 0.99,
-\emph{P} \textless{} 0.001; Table S3). For A-C, each symbol represents a
-stool sample from an individual mouse, with circles representing
-experiment 1 mice and triangles representing experiment 2 mice. D. The
-median (point) and interquantile range (colored lines) of the relative
-abundances for the 18 OTUs (Table S8) that varied between sources after
-clindamycin treatment (day 0). E. The median (point) and interquantile
-range (colored lines) of the top 10 most significant OTUs out of 153
-with relative abundances that changed because of the clindamycin
-treatment (adjusted \emph{P} value \textless{} 0.05). Data were analyzed
-by paired Wilcoxon signed rank test of mice that had paired sequence
-data for baseline (day -1) and post-clindamycin (day 0) timepoints (N =
-31), with Benjamini-Hochberg correction for testing all identified OTUs
-(Table S9). The gray vertical line indicates the limit of detection.
+\textbf{Figure 3. Clindamycin treatment alters bacteria in all sources,
+but a subset of bacterial differences across sources persists.} A-B.
+Number of observed OTUs (A) and Shannon diversity index values (B)
+across sources of mice after clindamycin treatment (day 0). Differences
+between sources were analyzed by Kruskal-Wallis test with
+Benjamini-Hochberg correction for testing each day of the experiment and
+the adjusted \emph{P} value was \textless{} 0.05 (Table S1). Significant
+\emph{P} values from the pairwise Wilcoxon comparisons between sources
+with Benjamini-Hochberg correction are displayed as the first initial of
+each group compared to the group that they are listed above (Table S2).
+C. PCoA of \(\theta_{YC}\) distances from stools collected
+post-clindamycin. Source and the interaction between source and cage
+effects explained most of the variation observed post-clindamycin
+(PERMANOVA combined R\textsuperscript{2} = 0.99, \emph{P} \textless{}
+0.001; Table S3). For A-C, each symbol represents a stool sample from an
+individual mouse, with circles representing experiment 1 mice and
+triangles representing experiment 2 mice. D. The median (point) and
+interquantile range (colored lines) of the relative abundances for the
+18 OTUs (Table S8) that varied between sources after clindamycin
+treatment (day 0). E. The median (point) and interquantile range
+(colored lines) of the top 10 most significant OTUs out of 153 with
+relative abundances that changed because of the clindamycin treatment
+(adjusted \emph{P} value \textless{} 0.05). Data were analyzed by paired
+Wilcoxon signed rank test of mice that had paired sequence data for
+baseline (day -1) and post-clindamycin (day 0) timepoints (N = 31), with
+Benjamini-Hochberg correction for testing all identified OTUs (Table
+S9). The gray vertical line indicates the limit of detection.
 
@@
 
-\includegraphics{figure_4.pdf} \textbf{Figure 4. Microbiota variation
-across sources is maintained after \emph{C. difficile} challenge.} A-B.
-Number of observed OTUs (A) and Shannon diversity index values (B)
-across sources of mice 1-day post-infection. Data were analyzed by
-Kruskal-Wallis test with Benjamini-Hochberg correction for testing each
-day of the experiment and the adjusted \emph{P} value was \textless{}
-0.05 (Table S1). Significant \emph{P} values from the pairwise Wilcoxon
-comparisons between sources with Benjamini-Hochberg correction are
-displayed as the first initial of each group compared to the group that
-they are listed above (Table S2). PCoA of \(\theta_{YC}\) distances of
-1-day post-infection stool samples. Source and the interaction between
-source and cage effects explained most of the variation between fecal
-communities (PERMANOVA combined R\textsuperscript{2} = 0.88, \emph{P}
-\textless{} 0.001; Table S3). For A-C: each symbol represents the value
-for a stool sample from an individual mouse, circles represent
-experiment 1 mice and triangles represent experiment 2 mice. D. The
-median (point) and interquantile range (colored lines) of the relative
-abundances for the top 20 most significant OTUs out of the 44 OTUs that
-varied between sources 1-day post-infection. The gray vertical line
-indicates the limit of detection. For each timepoint OTUs with
-differential relative abundances across sources of mice were identified
-by Kruskal-Wallis test with Benjamini-Hochberg correction for testing
-all identified OTUs (Table S10). E. \(\theta_{YC}\) distances of fecal
-samples collected 7-days post-infection relative to the baseline (day
--1) sample for each mouse. Each symbol represents an individual mouse.
-Gray lines represent the median for each source.
+\textbf{Figure 4. Microbiota variation across sources is maintained
+after \emph{C. difficile} challenge.} A-B. Number of observed OTUs (A)
+and Shannon diversity index values (B) across sources of mice 1-day
+post-infection. Data were analyzed by Kruskal-Wallis test with
+Benjamini-Hochberg correction for testing each day of the experiment and
+the adjusted \emph{P} value was \textless{} 0.05 (Table S1). Significant
+\emph{P} values from the pairwise Wilcoxon comparisons between sources
+with Benjamini-Hochberg correction are displayed as the first initial of
+each group compared to the group that they are listed above (Table S2).
+PCoA of \(\theta_{YC}\) distances of 1-day post-infection stool samples.
+Source and the interaction between source and cage effects explained
+most of the variation between fecal communities (PERMANOVA combined
+R\textsuperscript{2} = 0.88, \emph{P} \textless{} 0.001; Table S3). For
+A-C: each symbol represents the value for a stool sample from an
+individual mouse, circles represent experiment 1 mice and triangles
+represent experiment 2 mice. D. The median (point) and interquantile
+range (colored lines) of the relative abundances for the top 20 most
+significant OTUs out of the 44 OTUs that varied between sources 1-day
+post-infection. The gray vertical line indicates the limit of detection.
+For each timepoint OTUs with differential relative abundances across
+sources of mice were identified by Kruskal-Wallis test with
+Benjamini-Hochberg correction for testing all identified OTUs (Table
+S10). E. \(\theta_{YC}\) distances of fecal samples collected 7-days
+post-infection relative to the baseline (day -1) sample for each mouse.
+Each symbol represents an individual mouse. Gray lines represent the
+median for each source.
 
@@
 
-\includegraphics{figure_5.pdf} \textbf{Figure 5. Bacteria that
-influenced whether mice cleared \emph{C. difficile} by day 7.} A.
-Post-clindamycin (day 0) relative abundance data for the 10 OTUs with
-the highest rankings based on feature weights in the post-clindamycin
-(day 0) classification model. Red font represents OTUs that correlated
-with \emph{C. difficile} colonization and blue font represents OTUs that
-correlated with clearance. Symbols represent the relative abundance data
-for an individual mouse. Gray bars indicate the median relative
-abundances for each source. The gray horizontal lines indicate the limit
-of detection. B. Venn diagram that combines OTUs that were important to
-the day -1, 0, and 1 classification models ( Fig. S4, Table S14) and
-either overlapped with taxa that varied across sources at the same
-timepoint, were impacted by clindamycin treatment, or both. Bold OTUs
-were important to more than 1 classification model.
+\textbf{Figure 5. Bacteria that influenced whether mice cleared \emph{C.
+difficile} by day 7.} A. Post-clindamycin (day 0) relative abundance
+data for the 10 OTUs with the highest rankings based on feature weights
+in the post-clindamycin (day 0) classification model. Red font
+represents OTUs that correlated with \emph{C. difficile} colonization
+and blue font represents OTUs that correlated with clearance. Symbols
+represent the relative abundance data for an individual mouse. Gray bars
+indicate the median relative abundances for each source. The gray
+horizontal lines indicate the limit of detection. B. Venn diagram that
+combines OTUs that were important to the day -1, 0, and 1 classification
+models ( Fig. S4, Table S14) and either overlapped with taxa that varied
+across sources at the same timepoint, were impacted by clindamycin
+treatment, or both. Bold OTUs were important to more than 1
+classification model.
 
@@
 
-\includegraphics{figure_6.pdf} \textbf{Figure 6: OTUs associated with
-\emph{C. difficile} colonization dynamics vary across sources throughout
-the experiment.} A-D. Relative abundances of bold OTUs from Fig. 5B that
-were important in at least two classification models are shown over
-time. A. \emph{Bacteroides} (OTU 2), which varied across sources
-throughout the experiment. B-C. \emph{Enterobacteriaceae} (B) and
-\emph{Enterococcus} (C), which significantly varied across sources and
-were impacted by clindamycin treatment. D. \emph{Porphyromonadaceae}
-(OTU 7), which was significantly impacted by clindamycin treatment and
-after examining relative abundance dynamics over the course of the
-experiment was found to also significantly vary between sources of mice
-on days -1, 5, 6, 7, and 9 of the experiment. Symbols represent the
-relative abundance data for an individual mouse. Colored lines indicate
-the median relative abundances for each source. The gray horizontal line
-represents the limit of detection. Timepoints where differences between
-sources of mice were statistically significant by Kruskal-Wallis test
-with Benjamini-Hochberg correction for testing across multiple days
-(Table S15) are identified by the asterisk above each timepoint (*, P
-\textless{} 0.05).
+\textbf{Figure 6: OTUs associated with \emph{C. difficile} colonization
+dynamics vary across sources throughout the experiment.} A-D. Relative
+abundances of bold OTUs from Fig. 5B that were important in at least two
+classification models are shown over time. A. \emph{Bacteroides} (OTU
+2), which varied across sources throughout the experiment. B-C.
+\emph{Enterobacteriaceae} (B) and \emph{Enterococcus} (C), which
+significantly varied across sources and were impacted by clindamycin
+treatment. D. \emph{Porphyromonadaceae} (OTU 7), which was significantly
+impacted by clindamycin treatment and after examining relative abundance
+dynamics over the course of the experiment was found to also
+significantly vary between sources of mice on days -1, 5, 6, 7, and 9 of
+the experiment. Symbols represent the relative abundance data for an
+individual mouse. Colored lines indicate the median relative abundances
+for each source. The gray horizontal line represents the limit of
+detection. Timepoints where differences between sources of mice were
+statistically significant by Kruskal-Wallis test with Benjamini-Hochberg
+correction for testing across multiple days (Table S15) are identified
+by the asterisk above each timepoint (*, P \textless{} 0.05).
 
 \newpage
-
-\includegraphics{figure_S1.pdf}
 
@@
 
-\includegraphics{figure_S2.pdf} \textbf{Figure S2. \emph{C. difficile}
-CFU variation across sources varies slightly between the 2 experiments.}
-A-B. \emph{C. difficile} CFU/gram of stool quantification over time for
-experiment 1 (A) and 2 (B). Experiments were conducted approximately 3
-months apart. Lines represent the median CFU for each source, symbols
-represent individual mice and the black line represents the limit of
-detection. C. \emph{C. difficile} CFU/gram stool 7-days post-infection
-across sources of mice with an asterisk for pairwise Wilcoxon
-comparisons with Benjamini-Hochberg correction where \emph{P}
-\textless{} 0.05. D. Mouse weight change 2-days post-infection across
-sources of mice, no pairwise Wilcoxon comparisons were significant after
-Benjamini-Hochberg correction. For C-D: circles represent experiment 1
-mice, triangles represent experiment 2 mice and gray lines indicate the
-median values for each group. E. Percent of mice that were colonized
-with \emph{C. difficile} over the course of the experiment. Each day the
-percent is calculated based on the mice where \emph{C. difficile} CFU
-was quantified for that particular day. Total N for each day: day 1 (N =
+\textbf{Figure S2. \emph{C. difficile} CFU variation across sources
+varies slightly between the 2 experiments.} A-B. \emph{C. difficile}
+CFU/gram of stool quantification over time for experiment 1 (A) and 2
+(B). Experiments were conducted approximately 3 months apart. Lines
+represent the median CFU for each source, symbols represent individual
+mice and the black line represents the limit of detection. C. \emph{C.
+difficile} CFU/gram stool 7-days post-infection across sources of mice
+with an asterisk for pairwise Wilcoxon comparisons with
+Benjamini-Hochberg correction where \emph{P} \textless{} 0.05. D. Mouse
+weight change 2-days post-infection across sources of mice, no pairwise
+Wilcoxon comparisons were significant after Benjamini-Hochberg
+correction. For C-D: circles represent experiment 1 mice, triangles
+represent experiment 2 mice and gray lines indicate the median values
+for each group. E. Percent of mice that were colonized with \emph{C.
+difficile} over the course of the experiment. Each day the percent is
+calculated based on the mice where \emph{C. difficile} CFU was
+quantified for that particular day. Total N for each day: day 1 (N =
 42), day 2 (N = 20), day 3 (N = 39), day 4 (N = 29), day 5 (N = 43), day
@@
 
-\includegraphics{figure_S3.pdf} \textbf{Figure S3. Bacterial community
-composition before, after clindamycin perturbation, and post-infection
-can predict \emph{C. difficile} colonization status 7 days
-post-infection.} A. Bar graph visualizations of overall 7-days
-post-infection \emph{C. difficile} colonization status that were used as
-classification outcomes to build L2-regularized logistic regression
-models. Mice were classified as colonized or cleared (not detectable at
-the limit of detection of 100 CFU) based on CFU g/stool data from 7 days
-post-infection. B. \emph{C. difficile} CFU status on Day 7 within each
-mouse source. N = 8-9 mice per group. C. L2-regularized logistic
-regression classification model area under the receiving operator
-characteristic curve (AUROCs) to predict \emph{C. difficile} CFU on day
-7 post-infection (Fig. 2B, Fig. S2C) based on the OTU community relative
-abundances at baseline (day -1), post-clindamycin (day 0), and 1-day
-post-infection. All models performed better than random chance (AUROC =
-0.5, all \emph{P} \textless{} 0.001, Table S12) and the model built with
-post-clindamycin bacterial OTU relative abundances had the best
-performance ((\emph{P}\textsubscript{FDR} \textless{} 0.001 for all
-pairwise comparisons, Table S13). See Table S14 for list of the 20 OTUs
-that were ranked as most important to each model.
+\textbf{Figure S3. Bacterial community composition before, after
+clindamycin perturbation, and post-infection can predict \emph{C.
+difficile} colonization status 7 days post-infection.} A. Bar graph
+visualizations of overall 7-days post-infection \emph{C. difficile}
+colonization status that were used as classification outcomes to build
+L2-regularized logistic regression models. Mice were classified as
+colonized or cleared (not detectable at the limit of detection of 100
+CFU) based on CFU g/stool data from 7 days post-infection. B. \emph{C.
+difficile} CFU status on Day 7 within each mouse source. N = 8-9 mice
+per group. C. L2-regularized logistic regression classification model
+area under the receiving operator characteristic curve (AUROCs) to
+predict \emph{C. difficile} CFU on day 7 post-infection (Fig. 2B, Fig.
+S2C) based on the OTU community relative abundances at baseline (day
+-1), post-clindamycin (day 0), and 1-day post-infection. All models
+performed better than random chance (AUROC = 0.5, all \emph{P}
+\textless{} 0.001, Table S12) and the model built with post-clindamycin
+bacterial OTU relative abundances had the best performance
+((\emph{P}\textsubscript{FDR} \textless{} 0.001 for all pairwise
+comparisons, Table S13). See Table S14 for list of the 20 OTUs that were
+ranked as most important to each model.
 
 \newpage
-
-\includegraphics{figure_S4.pdf}
 

### 2020-09-25 code/figure_1.R
Format figures for revision submission

@@ -7,7 +7,7 @@ left_panel <- plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, nco
 d <- ggdraw() + draw_image("results/figures/Dn1top20_otus.png")
 
 plot_grid(left_panel, d, labels = c("", "D"), label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_1.pdf", width=8, height=5.5)+
-  ggsave("submission/figure_1.pdf", width=8, height=5.5)
+  ggsave("results/figures/figure_1.pdf", width=6.875, height=4.73)+
+  ggsave("submission/figure_1.tiff", width=6.875, height=4.73, dpi = 300, device = "tiff", compression = "lzw", units = "in")
 
 

### 2020-09-25 code/figure_2.R

@@ -7,7 +7,7 @@ c <- ggdraw() + draw_image("results/figures/weight_over_time.png")
 
 plot_grid(a, b, c, labels = c("A", "B", "C", ""), label_size = 12, ncol=1)+
   ggsave("results/figures/figure_2.pdf", width=5, height=7.5)+
-  ggsave("submission/figure_2.pdf", width=5, height=7.5)
+  ggsave("submission/figure_2.tiff", width=5, height=7.5, dpi = 300, device = "tiff", compression = "lzw", units = "in")
 
 
 

### 2020-09-25 code/figure_3.R

@@ -10,5 +10,5 @@ right_panel <- plot_grid(d, e, labels = c("D", "E"), label_size = 12, ncol=2, re
 right_panel <- plot_grid(right_panel, NULL, labels = NULL, label_size = 12, nrow=2, rel_heights = c(1, 0.5)) #Add NULL plot to fix margin spacing in final figure
 
 plot_grid(left_panel, right_panel, labels = c("", "D", "E"), label_size = 12, ncol=2, axis = "t", align = "v", rel_heights = c(1, 3), rel_widths = c(1, 2))+
-  ggsave("results/figures/figure_3.pdf", width=8, height=6)+
-  ggsave("submission/figure_3.pdf", width=8, height=6)
+  ggsave("results/figures/figure_3.pdf", width=6.875, height=5.16)+
+  ggsave("submission/figure_3.tiff", width=6.875, height=5.16, dpi = 300, device = "tiff", compression = "lzw", units = "in")

### 2020-09-25 code/figure_4.R

@@ -8,5 +8,5 @@ left_panel <- plot_grid(a, b, c, e, labels = c("A", "B", "C", "E"), label_size =
 d <- ggdraw() + draw_image("results/figures/D1top20_otus.png")
 
 plot_grid(left_panel, d, labels = c("", "D"), label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_4.pdf", width=8, height=6)+
-  ggsave("submission/figure_4.pdf", width=8, height=6)
+  ggsave("results/figures/figure_4.pdf", width=6.875, height=5.16)+
+  ggsave("submission/figure_4.tiff", width=6.875, height=5.16, dpi = 300, device = "tiff", compression = "lzw", units = "in")

### 2020-09-25 code/figure_5.R

@@ -4,6 +4,6 @@ a <- ggdraw() + draw_image("results/figures/d0_top10_otus.png")
 b <- ggdraw() + draw_image("results/figures/venn_overall_otus.png")
 
 plot_grid(a, b, labels = "AUTO", label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_5.pdf", width=7, height=9)+
-  ggsave("submission/figure_5.pdf", width=7, height=9)
+  ggsave("results/figures/figure_5.pdf", width=6.875, height=8.84)+
+  ggsave("submission/figure_5.tiff", width=6.875, height=8.84, dpi = 300, device = "tiff", compression = "lzw", units = "in")
 

### 2020-09-25 code/figure_6.R

@@ -9,5 +9,5 @@ legend <- ggdraw()+ draw_image("results/figures/overlap_OTUs_legend.png")
 
 
 plot_grid(figure, legend, labels = NULL, label_size = 12, ncol=1, rel_heights = c(1, 0.2))+
-  ggsave("results/figures/figure_6.pdf", width=8, height=5)+
-  ggsave("submission/figure_6.pdf", width=8, height=5)
+  ggsave("results/figures/figure_6.pdf", width=6.875, height=4.3)+
+  ggsave("submission/figure_6.tiff", width=6.875, height=4.3, dpi = 300, device = "tiff", compression = "lzw", units = "in")

### 2020-09-25 code/figure_S1.R

@@ -10,8 +10,8 @@ g <- ggdraw() + draw_image("results/figures/within_exp_dn1.png")
 h <- ggdraw() + draw_image("results/figures/between_exp_dn1.png")
 
 plot_grid(a, b, c, d, e, f, g, h, labels = "AUTO", label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_S1.pdf", width=6, height=10)+
-  ggsave("submission/figure_S1.pdf", width=6, height=10)
+  ggsave("results/figures/figure_S1.pdf", width=5.4, height=9.0625)+
+  ggsave("submission/figure_S1.tiff", width=5.4, height=9.0625, dpi = 300, device = "tiff", compression = "lzw", units = "in")
 
 
 

### 2020-09-25 code/figure_S2.R

@@ -15,7 +15,7 @@ bottom_panel <- plot_grid(c, d, e, labels = c("C", "D", "E"), label_size = 12, n
 
 
 plot_grid(a, bottom_panel, labels = c("", ""), label_size = 12, ncol=1, rel_heights = c(1.4, 1))+
-  ggsave("results/figures/figure_S2.pdf", width=8, height=6)+
-  ggsave("submission/figure_S2.pdf", width=8, height=6)
+  ggsave("results/figures/figure_S2.pdf", width=6.875, height=5.16)+
+  ggsave("submission/figure_S2.tiff", width=6.875, height=5.16, dpi = 300, device = "tiff", compression = "lzw", units = "in")
 
 

### 2020-09-25 code/figure_S3.R

@@ -5,6 +5,6 @@ b <- ggdraw() + draw_image("results/figures/c.diff_status_d7.png")
 c <- ggdraw() + draw_image("results/figures/class._logistic_regression_60.png")
 
 plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=2)+
-  ggsave("results/figures/figure_S3.pdf", width=7.5, height=5.5)+
-  ggsave("submission/figure_S3.pdf", width=7.5, height=5.5)
+  ggsave("results/figures/figure_S3.pdf", width=6.875, height=5.04)+
+  ggsave("submission/figure_S3.tiff", width=6.875, height=5.04, dpi = 300, device = "tiff", compression = "lzw", units = "in")
  

### 2020-09-25 code/figure_S4.R

@@ -7,4 +7,4 @@ c <- ggdraw() + draw_image("results/figures/venn_d1_otus.png")
 
 plot_grid(a, b, c, labels = "AUTO", label_size = 12, ncol=1)+
   ggsave("results/figures/figure_S4.pdf", width=6, height=8)+
-  ggsave("submission/figure_S4.pdf", width=6, height=8)
+  ggsave("submission/figure_S4.tiff", width=6, height=8, dpi = 300, device = "tiff", compression = "lzw", units = "in")

### 2020-09-30 code/supplemental_data_set_S1.R
Transform Tables S1-S15 into Data Set S1

@@ -1,0 +1,82 @@
+source("code/functions.R")
+
+#Generate single excel workbook containing sheets for Data Set S1
+
+#Sheets 1 and 2: Alpha diversity Kruskal-Wallis and pairwise wilcoxon results----
+#Combine Kruskal-Wallis test results for shannon and richness diversity metrics
+shannon_results <- read_tsv("data/process/shannon_stats_all_days.tsv") %>% 
+  mutate(diversity_metric = "Shannon Diversity")
+richness_results <- read_tsv("data/process/richness_stats_all_days.tsv") %>% 
+  mutate(diversity_metric = "Richness")
+
+#Sheet 1: Diversity kruskal-Wallis test results with adjusted P values
+diversity_kruskal_stats_adjust <- rbind(shannon_results, richness_results)
+
+#Combine pairwise wilcoxon test results for shannon and richness diversity metrics
+shannon_pairwise_results <- read_tsv("data/process/shannon_stats_sig_days.tsv") %>% 
+  mutate(diversity_metric = "Shannon Diversity")
+richness_pairwise_results <- read_tsv("data/process/richness_stats_sig_days.tsv") %>% 
+  mutate(diversity_metric = "Richness")
+#Sheet 2: Diversity pairwise wilcoxon results
+diversity_pairwise_results <- rbind(shannon_pairwise_results, richness_pairwise_results)
+
+#Write PERMANOVA results to data set excel file.
+#Sheet 3: PERMANOVA results for day -1, day 0, and day 1----
+adonis_dn1_to_0 <- read_tsv("data/process/adonis_dn1-1.tsv")
+#Sheet 4: PERMANOVA results within sources at day -1----
+dn1_source <- read_tsv("data/process/adonis_dn1_source.tsv")
+
+#OTU statistics for differences across sources of mice on day -1, 0 and 1----
+otu_stats_dn1to1_combined <- read_tsv("data/process/otu_stats_dn1to1_combined.tsv")
+#Sheet 5: OTUs that vary between sources on day -1----
+kw_w_sig_dn1 <- otu_stats_dn1to1_combined %>% filter(day == -1)
+
+#Sheet 6: C. difficile CFU statistical results----
+cfu_kruskal_stats_adjust <- read_tsv("data/process/cfu_stats_all_days.tsv")
+
+#Sheet 7: Mouse weight change statistical results----
+weight_kruskal_stats_adjust <- read_tsv("data/process/weight_stats_all_days.tsv")
+
[…]

### 2020-09-30 code/tables_supplemental.R

@@ -1,77 +1,0 @@
-source("code/functions.R")
-
-#Generate single excel workbook containing sheets of all suppplemental tables
-
-#Table S1 and S2: Alpha diversity Kruskal-Wallis and pairwise wilcoxon results----
-#Combine Kruskal-Wallis test results for shannon and richness diversity metrics
-shannon_results <- read_tsv("data/process/shannon_stats_all_days.tsv") %>% 
-  mutate(diversity_metric = "Shannon Diversity")
-richness_results <- read_tsv("data/process/richness_stats_all_days.tsv") %>% 
-  mutate(diversity_metric = "Richness")
-
-#Table S1: Diversity kruskal-Wallis test results with adjusted P values
-diversity_kruskal_stats_adjust <- rbind(shannon_results, richness_results)
-
-#Combine pairwise wilcoxon test results for shannon and richness diversity metrics
-shannon_pairwise_results <- read_tsv("data/process/shannon_stats_sig_days.tsv") %>% 
-  mutate(diversity_metric = "Shannon Diversity")
-richness_pairwise_results <- read_tsv("data/process/richness_stats_sig_days.tsv") %>% 
-  mutate(diversity_metric = "Richness")
-#Table S2: Diversity pairwise wilcoxon results
-diversity_pairwise_results <- rbind(shannon_pairwise_results, richness_pairwise_results)
-
-#Write PERMANOVA results to supplemental table excel file.
-#Table S3: PERMANOVA results for day -1, day 0, and day 1----
-adonis_dn1_to_0 <- read_tsv("data/process/adonis_dn1-1.tsv")
-#Table S4: PERMANOVA results within sources at day -1----
-dn1_source <- read_tsv("data/process/adonis_dn1_source.tsv")
-
-#OTU statistics for differences across sources of mice on day -1, 0 and 1----
-otu_stats_dn1to1_combined <- read_tsv("data/process/otu_stats_dn1to1_combined.tsv")
-#Table S5: OTUs that vary between sources on day -1----
-kw_w_sig_dn1 <- otu_stats_dn1to1_combined %>% filter(day == -1)
-
-#Table S6: C. difficile CFU statistical results----
-cfu_kruskal_stats_adjust <- read_tsv("data/process/cfu_stats_all_days.tsv")
-
-#Table S7: Mouse weight change statistical results----
-weight_kruskal_stats_adjust <- read_tsv("data/process/weight_stats_all_days.tsv")
-
[…]

### 2020-09-30 submission/manuscript.Rmd
Update all references from Table S# Data Set S1, Sheet# in manuscript

@@
 number of observed operational taxonomic units (OTUs)), but not Shannon diversity index across the sources of mice (*P*~FDR~ = `r padj_dn1_richness` and *P*~FDR~ = `r padj_dn1_shannon`, respectively; Fig.
-1A-B and Tables S1-2).
+1A-B and Data Set S1, Sheets 1-2).
 Next, we compared the community structure of mice (Fig.
@@
 The source of mice and the interactions between the source and cage effects explained most of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig.
-1C and Table S3).
+1C and Data Set S1, Sheet 3).
 Mice that are co-housed tend to have similar gut microbiotas due to coprophagy [@Nguyen2015].
@@
 There were some differences between the 2 experiments we conducted, as the experiment and cage effects significantly explained the observed community variation for the Schloss and Young lab mouse colonies (Fig.
-S1A-B and Table S4).
+S1A-B and Data Set S1, Sheet 4).
 However, most of the vendors also clustered by experiment (Fig.
@@
 There were `r source_otus_dn1` OTUs with relative abundances that were significantly different between the sources at baseline (Fig.
-1D and Table S5).
+1D and Data Set S1, Sheet 5).
 Though we saw differences between experiments at the community level, there were no OTUs that were significantly different between experiments within Schloss, Young, and Envigo mice at baseline (all *P* > 0.05).
@@
 difficile* levels across sources of mice (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig.
-2B and Table S6).
+2B and Data Set S1, Sheet 6).
 This suggested the source of mice was associated with *C.
@@
 The change in the mice's weight significantly varied across sources of mice with the most weight loss occurring two days post-infection (Fig.
-2C and Table S7).
+2C and Data Set S1, Sheet 7).
 There was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment.
@@
 3A-B).
-Interestingly, significant differences in diversity metrics between sources emerged after clindamycin treatment, with Charles River mice having higher richness and Shannon diversity than most of the other sources (*P*~FDR~ < 0.05; Fig 3A-B and Tables S1-2).
+Interestingly, significant differences in diversity metrics between sources emerged after clindamycin treatment, with Charles River mice having higher richness and Shannon diversity than most of the other sources (*P*~FDR~ < 0.05; Fig 3A-B and Data Set S1, Sheets 1-2).
 The clindamycin treatment decreased the variation in community structures between sources of mice.
 The source of mice and the interactions between source and cage effects explained almost all of the observed variation between communities (combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig.
-3C and Table S3).
+3C and Data Set S1, Sheet 3).
 However, there were only `r source_otus_d0` OTUs with relative abundances that significantly varied between sources after clindamycin treatment (Fig.
-3D and Table S8).
+3D and Data Set S1, Sheet 8).
 Next, we identified the bacteria that shifted after clindamycin treatment, regardless of source by analyzing paired fecal samples from mice that were collected at baseline and after clindamycin treatment.
 We identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment in most mice (Fig.
-3E and Table S9).
-When we compared the list of significant clindamycin impacted bacteria with the bacteria that varied between sources post-clindamycin, we found 4 OTUs that were shared between the lists (*Enterobacteriaceae (OTU 1)*, *Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23); Fig.
-3D-E and Tables S8-9).
+3E and Data Set S1, Sheet 9).
+When we compared the list of significant clindamycin impacted bacteria with the bacteria that varied between sources post-clindamycin, we found 4 OTUs that were shared between the lists (*Enterobacteriaceae* (OTU 1), *Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23); Fig.
+3D-E and Data Set S1, Sheets 8-9).
 Importantly, some of the OTUs that varied between sources also shifted with clindamycin treatment.
@@
 **Microbiota variation between sources is maintained after *C.
-difficile* challenge.** One day post-infection, significant differences in diversity metrics remained across sources (*P*~FDR~ < 0.05, Fig 4A-B and Tables S1-2).
+difficile* challenge.** One day post-infection, significant differences in diversity metrics remained across sources (*P*~FDR~ < 0.05, Fig 4A-B and Data Set S1, Sheets 1-2).
 Although the Charles River mice had more diverse communities and were also able to clear *C.
@@
 The source of mice and the interactions between source and cage effects continued to explain most of the observed community variation (combined R^2^ = `r d1_rsq_sum`; *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig.
-4C and Table S3).
+4C and Data Set S1, Sheet 3).
 One day after *C.
 difficile* challenge, there were `r source_otus_d1` OTUs with significantly different relative abundances across sources (Fig.
-4D and Table S10).
+4D and Data Set S1, Sheet 10).
 
-Throughout the experiment, the source of mice continued to be the dominant factor that explained the observed variation across fecal communities (PERMANOVA R^2^ = `r rsq_adonis_source`, *P* `r format.pval(p_adonis_source, eps = 0.001)`) followed by interactions between cage effects  and the day of the experiment (Movie S1 and Table S11).
+Throughout the experiment, the source of mice continued to be the dominant factor that explained the observed variation across fecal communities (PERMANOVA R^2^ = `r rsq_adonis_source`, *P* `r format.pval(p_adonis_source, eps = 0.001)`) followed by interactions between cage effects  and the day of the experiment (Movie S1 and Data Set S1, Sheet 11).
 Fecal samples from the same source of mice continued to cluster closely to each other throughout the experiment.
@@
 All models were better at predicting *C.
-difficile* colonization status on day 7 than random chance (all *P* `r format.pval(max_p_vs_random, eps = 0.001)`, Table S12).
+difficile* colonization status on day 7 than random chance (all *P* `r format.pval(max_p_vs_random, eps = 0.001)`, Data Set S1, Sheet 12).
 The model based on the post-clindamycin (AUROC = `r median_otu_day0_AUROC`) community OTU data performed significantly better than the baseline (AUROC = `r median_otu_dayn1_AUROC`) or the post-infection (AUROC = `r median_otu_day1_AUROC`) models (*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for pairwise comparisons; Fig.
-S3C and Table S13).
+S3C and Data Set S1, Sheet 13).
 Thus, we were able to use bacterial relative abundance data from the time of *C.
@@
 
-To examine the bacteria that were driving each model’s performance, we selected the 20 OTUs that had the highest absolute feature weights in each of the 3 models (Table S14).
+To examine the bacteria that were driving each model’s performance, we selected the 20 OTUs that had the highest absolute feature weights in each of the 3 models (Data Set S1, Sheet 14).
 First, we looked at OTUs from the model with the best performance, which was based on the post-clindamycin treatment (day 0) bacterial community data.
@@
 Next, we examined whether any of the top 20 ranked OTUs from the post-clindamycin (day 0) model were also important in the other 2 classification models based on baseline (day -1) and 1 day post-infection community data.
-We identified 6 OTUs that were important to the post-clindamycin model and either the baseline or 1 day post-infection models (*`r list_overlapping_across_models`*; Table S14).
+We identified 6 OTUs that were important to the post-clindamycin model and either the baseline or 1 day post-infection models (*`r list_overlapping_across_models`*; Data Set S1, Sheet 14).
 Thus, a subset of bacterial OTUs were important for determining *C.
@@
 To determine whether the OTUs driving the classification models also varied between sources, were altered by clindamycin treatment, or both, we identified the OTUs from each model that varied between sources (Fig.
-1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin treatment (Fig.
-3E and Table S9; Fig.
+1D, 3D, 4D and Data Set S1, Sheets 5, 8, and 10) or were impacted by clindamycin treatment (Fig.
+3E and Data Set S1, Sheet 9; Fig.
 S4).
@@
 6).
-Across the 9 days post-infection, there was at least 1 timepoint when the relative abundances of these OTUs significantly varied between sources (Table S15).
+Across the 9 days post-infection, there was at least 1 timepoint when the relative abundances of these OTUs significantly varied between sources (Data Set S1, Sheet 15).
 Interestingly, there were no OTUs that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection, suggesting that multiple bacteria influence *C.
@@
 Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice at baseline (day -1 of the experiment).
-Differences between sources were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 for panel A (Table S1).
-None of the *P* values from pairwise Wilcoxon comparisons between sources were significant after Benjamini-Hochberg correction (Table S2).
+Differences between sources were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 for panel A (Data Set S1, Sheet 1).
+None of the *P* values from pairwise Wilcoxon comparisons between sources were significant after Benjamini-Hochberg correction (Data Set S1, Sheet 2).
 Gray lines represent the median values for each source of mice.
@@
 Principal Coordinates Analysis (PCoA) of $\theta_{YC}$ distances of baseline stool samples.
-Source and the interaction between source and cage effects explained most of the variation (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Table S3).
+Source and the interaction between source and cage effects explained most of the variation (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Data Set S1, Sheet 3).
 For A-C:  each symbol represents the value for a stool sample from an individual mouse, circles represent experiment 1 mice and triangles represent experiment 2 mice.
 D.
-The median (point) and interquantile range (colored lines) of the relative abundances for the 20 most significant OTUs out of the `r source_otus_dn1` OTUs that varied across sources at baseline by Kruskal-Wallis test with Benjamini-Hochberg correction (Table S5).
+The median (point) and interquantile range (colored lines) of the relative abundances for the 20 most significant OTUs out of the `r source_otus_dn1` OTUs that varied across sources at baseline by Kruskal-Wallis test with Benjamini-Hochberg correction (Data Set S1, Sheet 5).
 
@@
 Mouse weight change measured in grams over time (N = `r min_weight_n`-`r max_weight_n` mice per timepoint), all mice were normalized to the weight recorded 1 day before infection.
-For B-C: timepoints where differences between sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S6 and Table S7) are reflected by the asterisk above each timepoint (\*\, *P* < 0.05).
+For B-C: timepoints where differences between sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Data Set S1, Sheets 6-7) are reflected by the asterisk above each timepoint (\*\, *P* < 0.05).
 Lines represent the median for each source and circles represent individual mice from experiment 1 while triangles represent mice from experiment 2.
@@
 Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice after clindamycin treatment (day 0).
-Differences between sources were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 (Table S1).
-Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are displayed as the first initial of each group compared to the group that they are listed above (Table S2).
+Differences between sources were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 (Data Set S1, Sheet 1).
+Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are displayed as the first initial of each group compared to the group that they are listed above (Data Set S1, Sheet 2).
 C.
 PCoA of $\theta_{YC}$ distances from stools collected post-clindamycin.
-Source and the interaction between source and cage effects explained most of the variation observed post-clindamycin (PERMANOVA combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Table S3).
+Source and the interaction between source and cage effects explained most of the variation observed post-clindamycin (PERMANOVA combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Data Set S1, Sheet 3).
 For A-C, each symbol represents a stool sample from an individual mouse, with circles representing experiment 1 mice and triangles representing experiment 2 mice.
 D.
-The median (point) and interquantile range (colored lines) of the relative abundances for the `r source_otus_d0` OTUs (Table S8) that varied between sources after clindamycin treatment (day 0).
+The median (point) and interquantile range (colored lines) of the relative abundances for the `r source_otus_d0` OTUs (Data Set S1, Sheet 8) that varied between sources after clindamycin treatment (day 0).
 E.
 The median (point) and interquantile range (colored lines) of the top 10 most significant OTUs out of `r clindamycin_otus_sig` with relative abundances that changed because of the clindamycin treatment (adjusted *P* value < 0.05).
-Data were analyzed by paired Wilcoxon signed rank test of mice that had paired sequence data for baseline (day -1) and post-clindamycin (day 0) timepoints (N = `r n_paired_for_clind`), with Benjamini-Hochberg correction for testing all identified OTUs (Table S9).
+Data were analyzed by paired Wilcoxon signed rank test of mice that had paired sequence data for baseline (day -1) and post-clindamycin (day 0) timepoints (N = `r n_paired_for_clind`), with Benjamini-Hochberg correction for testing all identified OTUs (Data Set S1, Sheet 9).
 The gray vertical line indicates the limit of detection.
@@
 Number of observed OTUs (A) and Shannon diversity index values (B) across sources of mice 1-day post-infection.
-Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 (Table S1).
-Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are displayed as the first initial of each group compared to the group that they are listed above (Table S2).
+Data were analyzed by Kruskal-Wallis test with Benjamini-Hochberg correction for testing each day of the experiment and the adjusted *P* value was < 0.05 (Data Set S1, Sheet 1).
+Significant *P* values from the pairwise Wilcoxon comparisons between sources with Benjamini-Hochberg correction are displayed as the first initial of each group compared to the group that they are listed above (Data Set S1, Sheet 2).
 PCoA of $\theta_{YC}$ distances of 1-day post-infection stool samples.
-Source and the interaction between source and cage effects explained most of the variation between fecal communities (PERMANOVA combined R^2^ = `r d1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Table S3).
+Source and the interaction between source and cage effects explained most of the variation between fecal communities (PERMANOVA combined R^2^ = `r d1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Data Set S1, Sheet 3).
 For A-C:  each symbol represents the value for a stool sample from an individual mouse, circles represent experiment 1 mice and triangles represent experiment 2 mice.
@@
 The gray vertical line indicates the limit of detection.
-For each timepoint OTUs with differential relative abundances across sources of mice were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs (Table S10).
+For each timepoint OTUs with differential relative abundances across sources of mice were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs (Data Set S1, Sheet 10).
 E.
@@
 Venn diagram that combines  OTUs that were important to the day -1, 0, and 1 classification models ( Fig.
-S4, Table S14) and either overlapped with taxa that varied across sources at the same timepoint, were impacted by clindamycin treatment, or both.
+S4, Data Set S1, Sheet 14) and either overlapped with taxa that varied across sources at the same timepoint, were impacted by clindamycin treatment, or both.
 Bold OTUs were important to more than 1 classification model.
@@
 The gray horizontal line represents the limit of detection.
-Timepoints where differences between sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S15) are identified by the asterisk above each timepoint (\*\, P < 0.05).
+Timepoints where differences between sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Data Set S1, Sheet 15) are identified by the asterisk above each timepoint (\*\, P < 0.05).
 
@@
 Each symbol represents a stool sample from an individual mouse with color corresponding to experiment and shape representing cage mates.
-Experiment number and cage effects explained most of the observed variation for samples from the Schloss (PERMANOVA combined R^2^ = `r adonis_schloss_rsq_sum`; *P* $\le$ `r p_adonis_schloss_max`) and Young (combined R^2^ = `r adonis_young_rsq_sum`; *P* $\le$ `r p_adonis_young_max`) mice (Table S4).
+Experiment number and cage effects explained most of the observed variation for samples from the Schloss (PERMANOVA combined R^2^ = `r adonis_schloss_rsq_sum`; *P* $\le$ `r p_adonis_schloss_max`) and Young (combined R^2^ = `r adonis_young_rsq_sum`; *P* $\le$ `r p_adonis_young_max`) mice (Data Set S1, Sheet 4).
 G-H: Boxplots of the $\theta_{YC}$ distances of the 6 sources of mice relative to mice within the same source and experiment (G) or mice within the same source and between experiments (H) at baseline (day -1).
@@
 S2C) based on the OTU community relative abundances at baseline (day -1), post-clindamycin (day 0), and 1-day post-infection.
-All models performed better than random chance (AUROC = 0.5, all *P* `r format.pval(max_p_vs_random, eps = 0.001)`, Table S12) and the model built with post-clindamycin bacterial OTU relative abundances had the best performance ((*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for all pairwise comparisons, Table S13).
-See Table S14 for list of the 20 OTUs that were ranked as most important to each model.
+All models performed better than random chance (AUROC = 0.5, all *P* `r format.pval(max_p_vs_random, eps = 0.001)`, Data Set S1, Sheet 12) and the model built with post-clindamycin bacterial OTU relative abundances had the best performance ((*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for all pairwise comparisons, Data Set S1, Sheet 13).
+See Data Set S1, Sheet 14 for list of the 20 OTUs that were ranked as most important to each model.
 
@@
 OTUs from classification models based on baseline, post-clindamycin treatment, or 1-day post-infection community data vary by source, clindamycin treatment, or both.** A-C.
-Venn diagrams of OTUs from the top 20 OTUs from the baseline (A), post-clindamycin treatment (B), and 1-day post-infection (C) classification models (Table S14) that overlapped with OTUs that varied across sources at the corresponding timepoint (Tables S5, 8, 10), were impacted by clindamycin treatment (Table S9), or both.
+Venn diagrams of OTUs from the top 20 OTUs from the baseline (A), post-clindamycin treatment (B), and 1-day post-infection (C) classification models (Data Set S1, Sheet 14) that overlapped with OTUs that varied across sources at the corresponding timepoint (Data Set S1, Sheets 5, 8, and 10), were impacted by clindamycin treatment (Data Set S1, Sheet 9), or both.
 Bold OTUs were important to more than 1 classification model.
@@
 \newpage
-## Supplementary Tables and Movie
+## Supplementary Movie and Data Set S1
 
@@
 difficile* infection.** PCoA of $\theta_{YC}$ distances animated from days -1 through 9 of the experiment.
-Source was the variable that explained the most observed variation across fecal communities (PERMANOVA source R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`, Table S11) followed by interactions between cage effects and day of the experiment.
+Source was the variable that explained the most observed variation across fecal communities (PERMANOVA source R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`, Data Set S1, Sheet 11) followed by interactions between cage effects and day of the experiment.
 Transparency of the symbol corresponds to the day of the experiment, each symbol represents a sample from an individual mouse at a specific timepoint.
@@
 
-**Tables S1-S15. Excel workbook of Tables S1-S15.**
+**Data Set S1, Sheets 1-15. Excel workbook with 15 sheets.**
 
-**Table S1. Alpha diversity metrics Kruskal-Wallis statistical results.**
+**Data Set S1, Sheet 1. Alpha diversity metrics Kruskal-Wallis statistical results.**
 
-**Table S2. Alpha diversity metrics pairwise Wilcoxon statistical results.**
+**Data Set S1, Sheet 2. Alpha diversity metrics pairwise Wilcoxon statistical results.**
 
-**Table S3. PERMANOVA results for mice at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1).**
+**Data Set S1, Sheet 3. PERMANOVA results for mice at baseline (day -1), post-clindamycin (day 0), and post-infection (day 1).**
 
-**Table S4. PERMANOVA results for each source of mice at baseline (day -1).**
+**Data Set S1, Sheet 4. PERMANOVA results for each source of mice at baseline (day -1).**
 
-**Table S5. OTUs with relative abumdances that significantly vary between sources at baseline (day -1).**
+**Data Set S1, Sheet 5. OTUs with relative abumdances that significantly vary between sources at baseline (day -1).**
 
-**Table S6. *C. difficile* CFU statistical results.**
+**Data Set S1, Sheet 6. *C. difficile* CFU statistical results.**
 
-**Table S7. Mouse weight change statistical results.**
+**Data Set S1, Sheet 7. Mouse weight change statistical results.**
 
-**Table S8. OTUs with relative abundances that significantly vary between sources post-clindamycin (day 0).**
+**Data Set S1, Sheet 8. OTUs with relative abundances that significantly vary between sources post-clindamycin (day 0).**
 
-**Table S9. OTUs with relative abundances that significantly changed after clindamycin treatment.**
+**Data Set S1, Sheet 9. OTUs with relative abundances that significantly changed after clindamycin treatment.**
 
-**Table S10. OTUs with relative abundances that significantly vary between sources 1-day post-infection.**
+**Data Set S1, Sheet 10. OTUs with relative abundances that significantly vary between sources 1-day post-infection.**
 
-**Table S11. PERMANOVA results for mice across all timepoints.**
+**Data Set S1, Sheet 11. PERMANOVA results for mice across all timepoints.**
 
-**Table S12. Statistical results of L2-regularized logistic regression model performances compared to random chance.**
+**Data Set S1, Sheet 12. Statistical results of L2-regularized logistic regression model performances compared to random chance.**
 
-**Table S13. Pairwise comparisons of L2-regularized logistic regression model performances.**
+**Data Set S1, Sheet 13. Pairwise comparisons of L2-regularized logistic regression model performances.**
 
-**Table S14. Top 20 most important OTUs for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
+**Data Set S1, Sheet 14. Top 20 most important OTUs for each of the 3 L2-regularized logistic regression models based on OTU relative abundance data.**
 
-**Table S15. OTUs with relative abundances that significantly varied between sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**
+**Data Set S1, Sheet 15. OTUs with relative abundances that significantly varied between sources of mice on at least 1 day of the experiment by Kruskal-Wallis test.**

### 2020-09-30 submission/manuscript.tex
Update all references from Table S# Data Set S1, Sheet# in manuscript

@@
 sources of mice (\emph{P}\textsubscript{FDR} = 0.03 and
-\emph{P}\textsubscript{FDR} = 0.052, respectively; Fig. 1A-B and Tables
-S1-2). Next, we compared the community structure of mice (Fig. 1C). The
-source of mice and the interactions between the source and cage effects
-explained most of the observed variation between fecal communities
-(PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P} \textless{}
-0.001; Fig. 1C and Table S3). Mice that are co-housed tend to have
-similar gut microbiotas due to coprophagy (41). Since mice within the
-same source were housed together, it was not surprising that the cage
-effect also contributed to the observed community variation. There were
-some differences between the 2 experiments we conducted, as the
-experiment and cage effects significantly explained the observed
-community variation for the Schloss and Young lab mouse colonies (Fig.
-S1A-B and Table S4). However, most of the vendors also clustered by
-experiment (Fig. S1C-D, F), suggesting there was some community
-variation between the 2 experiments within each source, particularly for
-Schloss, Young, and Envigo mice (Fig. S1G-H). After finding differences
-at the community level, we next identified the bacteria that varied
-between sources of mice. There were 268 OTUs with relative abundances
-that were significantly different between the sources at baseline (Fig.
-1D and Table S5). Though we saw differences between experiments at the
-community level, there were no OTUs that were significantly different
-between experiments within Schloss, Young, and Envigo mice at baseline
-(all \emph{P} \textgreater{} 0.05). By using mice from six sources we
-were able to increase the variation in the starting communities to
-evaluate in a clindamycin-based CDI model.
+\emph{P}\textsubscript{FDR} = 0.052, respectively; Fig. 1A-B and Data
+Set S1, Sheets 1-2). Next, we compared the community structure of mice
+(Fig. 1C). The source of mice and the interactions between the source
+and cage effects explained most of the observed variation between fecal
+communities (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
+\textless{} 0.001; Fig. 1C and Data Set S1, Sheet 3). Mice that are
+co-housed tend to have similar gut microbiotas due to coprophagy (41).
+Since mice within the same source were housed together, it was not
+surprising that the cage effect also contributed to the observed
+community variation. There were some differences between the 2
+experiments we conducted, as the experiment and cage effects
+significantly explained the observed community variation for the Schloss
+and Young lab mouse colonies (Fig. S1A-B and Data Set S1, Sheet 4).
+However, most of the vendors also clustered by experiment (Fig. S1C-D,
+F), suggesting there was some community variation between the 2
+experiments within each source, particularly for Schloss, Young, and
+Envigo mice (Fig. S1G-H). After finding differences at the community
+level, we next identified the bacteria that varied between sources of
+mice. There were 268 OTUs with relative abundances that were
+significantly different between the sources at baseline (Fig. 1D and
+Data Set S1, Sheet 5). Though we saw differences between experiments at
+the community level, there were no OTUs that were significantly
+different between experiments within Schloss, Young, and Envigo mice at
+baseline (all \emph{P} \textgreater{} 0.05). By using mice from six
+sources we were able to increase the variation in the starting
+communities to evaluate in a clindamycin-based CDI model.
 
@@
 levels across sources of mice (all \emph{P}\textsubscript{FDR} \(\le\)
-0.019; Fig. 2B and Table S6). This suggested the source of mice was
-associated with \emph{C. difficile} clearance. While the colonization
-dynamics were similar between the two experiments, the Schloss mice took
-longer to clear \emph{C. difficile} in the first experiment compared to
-the second and the Envigo mice took longer to clear \emph{C. difficile}
-in the second experiment compared to the first (Fig. S2A-B). The change
-in the mice's weight significantly varied across sources of mice with
-the most weight loss occurring two days post-infection (Fig. 2C and
-Table S7). There was also one Jackson and one Envigo mouse that died
-between 1- and 3-days post-infection during the second experiment. Mice
-obtained from Jackson, Taconic, and Envigo tended to lose more weight,
-have higher \emph{C. difficile} CFU levels and take longer to clear the
-infection compared to the other sources of mice (although there was
-variation between experiments with Schloss and Envigo mice). This was
-particularly evident 7 days post-infection (Fig. 2B-C, Fig. S2C-D), when
-57\% of the mice were still colonized with \emph{C. difficile} (Fig.
-S2E). By 9 days post-infection the majority of the mice from all sources
-had cleared \emph{C. difficile} with the exception of 1 Taconic mouse
-from the first experiment and 2 Envigo mice from the second experiment
-(Fig. 2B). Thus, clindamycin rendered all mice susceptible to \emph{C.
-difficile} 630 colonization, regardless of source, but there was
-significant variation in disease phenotype across the sources of mice.
+0.019; Fig. 2B and Data Set S1, Sheet 6). This suggested the source of
+mice was associated with \emph{C. difficile} clearance. While the
+colonization dynamics were similar between the two experiments, the
+Schloss mice took longer to clear \emph{C. difficile} in the first
+experiment compared to the second and the Envigo mice took longer to
+clear \emph{C. difficile} in the second experiment compared to the first
+(Fig. S2A-B). The change in the mice's weight significantly varied
+across sources of mice with the most weight loss occurring two days
+post-infection (Fig. 2C and Data Set S1, Sheet 7). There was also one
+Jackson and one Envigo mouse that died between 1- and 3-days
+post-infection during the second experiment. Mice obtained from Jackson,
+Taconic, and Envigo tended to lose more weight, have higher \emph{C.
+difficile} CFU levels and take longer to clear the infection compared to
+the other sources of mice (although there was variation between
+experiments with Schloss and Envigo mice). This was particularly evident
+7 days post-infection (Fig. 2B-C, Fig. S2C-D), when 57\% of the mice
+were still colonized with \emph{C. difficile} (Fig. S2E). By 9 days
+post-infection the majority of the mice from all sources had cleared
+\emph{C. difficile} with the exception of 1 Taconic mouse from the first
+experiment and 2 Envigo mice from the second experiment (Fig. 2B). Thus,
+clindamycin rendered all mice susceptible to \emph{C. difficile} 630
+colonization, regardless of source, but there was significant variation
+in disease phenotype across the sources of mice.
 
@@
 richness and Shannon diversity than most of the other sources
-(\emph{P}\textsubscript{FDR} \textless{} 0.05; Fig 3A-B and Tables
-S1-2). The clindamycin treatment decreased the variation in community
-structures between sources of mice. The source of mice and the
+(\emph{P}\textsubscript{FDR} \textless{} 0.05; Fig 3A-B and Data Set S1,
+Sheets 1-2). The clindamycin treatment decreased the variation in
+community structures between sources of mice. The source of mice and the
 interactions between source and cage effects explained almost all of the
 observed variation between communities (combined R\textsuperscript{2} =
-0.99, \emph{P} \textless{} 0.001; Fig. 3C and Table S3). However, there
-were only 18 OTUs with relative abundances that significantly varied
-between sources after clindamycin treatment (Fig. 3D and Table S8).
-Next, we identified the bacteria that shifted after clindamycin
-treatment, regardless of source by analyzing paired fecal samples from
-mice that were collected at baseline and after clindamycin treatment. We
-identified 153 OTUs that were altered after clindamycin treatment in
-most mice (Fig. 3E and Table S9). When we compared the list of
-significant clindamycin impacted bacteria with the bacteria that varied
-between sources post-clindamycin, we found 4 OTUs that were shared
-between the lists (\emph{Enterobacteriaceae (OTU 1)},
-\emph{Lachnospiraceae} (OTU 130), \emph{Lactobacillus} (OTU 6),
-\emph{Enterococcus} (OTU 23); Fig. 3D-E and Tables S8-9). Importantly,
-some of the OTUs that varied between sources also shifted with
-clindamycin treatment. For example, \emph{Proteus} increased after
+0.99, \emph{P} \textless{} 0.001; Fig. 3C and Data Set S1, Sheet 3).
+However, there were only 18 OTUs with relative abundances that
+significantly varied between sources after clindamycin treatment (Fig.
+3D and Data Set S1, Sheet 8). Next, we identified the bacteria that
+shifted after clindamycin treatment, regardless of source by analyzing
+paired fecal samples from mice that were collected at baseline and after
+clindamycin treatment. We identified 153 OTUs that were altered after
+clindamycin treatment in most mice (Fig. 3E and Data Set S1, Sheet 9).
+When we compared the list of significant clindamycin impacted bacteria
+with the bacteria that varied between sources post-clindamycin, we found
+4 OTUs that were shared between the lists (\emph{Enterobacteriaceae}
+(OTU 1), \emph{Lachnospiraceae} (OTU 130), \emph{Lactobacillus} (OTU 6),
+\emph{Enterococcus} (OTU 23); Fig. 3D-E and Data Set S1, Sheets 8-9).
+Importantly, some of the OTUs that varied between sources also shifted
+with clindamycin treatment. For example, \emph{Proteus} increased after
 clindamycin treatment (Fig. 3D), but only in Taconic mice.
@@
 differences in diversity metrics remained across sources
-(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Tables
-S1-2). Although the Charles River mice had more diverse communities and
-were also able to clear \emph{C. difficile} faster than the other
-sources, diversity did not explain the observed variation in \emph{C.
-difficile} colonization across sources. The Young and Schloss mice had
-the lowest diversity 1 day post-infection and were able to clear
-\emph{C. difficile} earlier than Jackson, Taconic and Envigo mice. The
-source of mice and the interactions between source and cage effects
+(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Data Set S1,
+Sheets 1-2). Although the Charles River mice had more diverse
+communities and were also able to clear \emph{C. difficile} faster than
+the other sources, diversity did not explain the observed variation in
+\emph{C. difficile} colonization across sources. The Young and Schloss
+mice had the lowest diversity 1 day post-infection and were able to
+clear \emph{C. difficile} earlier than Jackson, Taconic and Envigo mice.
+The source of mice and the interactions between source and cage effects
 continued to explain most of the observed community variation (combined
 R\textsuperscript{2} = 0.88; \emph{P} \textless{} 0.001; Fig. 4C and
-Table S3). One day after \emph{C. difficile} challenge, there were 44
-OTUs with significantly different relative abundances across sources
-(Fig. 4D and Table S10).
+Data Set S1, Sheet 3). One day after \emph{C. difficile} challenge,
+there were 44 OTUs with significantly different relative abundances
+across sources (Fig. 4D and Data Set S1, Sheet 10).
 
@@
 0.001) followed by interactions between cage effects and the day of the
-experiment (Movie S1 and Table S11). Fecal samples from the same source
-of mice continued to cluster closely to each other throughout the
-experiment. By 7 days post-infection, when approximately 43\% mice had
-cleared \emph{C. difficile}, most of the mice had not recovered to their
-baseline community structure (Fig. 4E). The distance to the baseline
-community did not explain the variation in \emph{C. difficile} clearance
-as the Schloss and Young mice had mostly cleared \emph{C. difficile},
-but their communities were a greater distance from baseline 7 days
-post-infection compared to the Jackson and Taconic mice that were still
-colonized. In summary, mouse bacterial communities varied significantly
-between sources throughout the course of the experiment and a consistent
-subset of bacteria remained different between sources regardless of
-clindamycin and \emph{C. difficile} challenge.
+experiment (Movie S1 and Data Set S1, Sheet 11). Fecal samples from the
+same source of mice continued to cluster closely to each other
+throughout the experiment. By 7 days post-infection, when approximately
+43\% mice had cleared \emph{C. difficile}, most of the mice had not
+recovered to their baseline community structure (Fig. 4E). The distance
+to the baseline community did not explain the variation in \emph{C.
+difficile} clearance as the Schloss and Young mice had mostly cleared
+\emph{C. difficile}, but their communities were a greater distance from
+baseline 7 days post-infection compared to the Jackson and Taconic mice
+that were still colonized. In summary, mouse bacterial communities
+varied significantly between sources throughout the course of the
+experiment and a consistent subset of bacteria remained different
+between sources regardless of clindamycin and \emph{C. difficile}
+challenge.
 
@@
 models were better at predicting \emph{C. difficile} colonization status
-on day 7 than random chance (all \emph{P} \textless{} 0.001, Table S12).
-The model based on the post-clindamycin (AUROC = 0.78) community OTU
-data performed significantly better than the baseline (AUROC = 0.72) or
-the post-infection (AUROC = 0.67) models (\emph{P}\textsubscript{FDR}
-\textless{} 0.001 for pairwise comparisons; Fig. S3C and Table S13).
-Thus, we were able to use bacterial relative abundance data from the
-time of \emph{C. difficile} challenge to differentiate mice that had
-cleared \emph{C. difficile} before day 7 from the mice still colonized
-with \emph{C. difficile} at that timepoint. This result suggests that
-the bacterial community's response to clindamycin treatment had the
-greatest influence on subsequent \emph{C. difficile} colonization
-dynamics.
+on day 7 than random chance (all \emph{P} \textless{} 0.001, Data Set
+S1, Sheet 12). The model based on the post-clindamycin (AUROC = 0.78)
+community OTU data performed significantly better than the baseline
+(AUROC = 0.72) or the post-infection (AUROC = 0.67) models
+(\emph{P}\textsubscript{FDR} \textless{} 0.001 for pairwise comparisons;
+Fig. S3C and Data Set S1, Sheet 13). Thus, we were able to use bacterial
+relative abundance data from the time of \emph{C. difficile} challenge
+to differentiate mice that had cleared \emph{C. difficile} before day 7
+from the mice still colonized with \emph{C. difficile} at that
+timepoint. This result suggests that the bacterial community's response
+to clindamycin treatment had the greatest influence on subsequent
+\emph{C. difficile} colonization dynamics.
 
@@
 selected the 20 OTUs that had the highest absolute feature weights in
-each of the 3 models (Table S14). First, we looked at OTUs from the
-model with the best performance, which was based on the post-clindamycin
-treatment (day 0) bacterial community data. Out of the 10 highest ranked
-OTUs, 7 OTUs were associated with \emph{C. difficile} colonization 7
-days post-infection (\emph{Bacteroides}, \emph{Escherichia/Shigella}, 2
-\emph{Lachnospiraceae}, \emph{Lactobacillus}, \emph{Porphyromonadaceae},
-and \emph{Ruminococcaceae}), while 3 OTUs were associated with clearance
+each of the 3 models (Data Set S1, Sheet 14). First, we looked at OTUs
+from the model with the best performance, which was based on the
+post-clindamycin treatment (day 0) bacterial community data. Out of the
+10 highest ranked OTUs, 7 OTUs were associated with \emph{C. difficile}
+colonization 7 days post-infection (\emph{Bacteroides},
+\emph{Escherichia/Shigella}, 2 \emph{Lachnospiraceae},
+\emph{Lactobacillus}, \emph{Porphyromonadaceae}, and
+\emph{Ruminococcaceae}), while 3 OTUs were associated with clearance
 (\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
@@
 Ruminococcaceae, Lactobacillus, Bacteroides, Porphyromonadaceae,
-Erysipelotrichaceae}; Table S14). Thus, a subset of bacterial OTUs were
-important for determining \emph{C. difficile} colonization dynamics
-across multiple timepoints.
+Erysipelotrichaceae}; Data Set S1, Sheet 14). Thus, a subset of
+bacterial OTUs were important for determining \emph{C. difficile}
+colonization dynamics across multiple timepoints.
 
@@
 we identified the OTUs from each model that varied between sources (Fig.
-1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin
-treatment (Fig. 3E and Table S9; Fig. S4). Comparing the features
-important to the 3 models identified 14 OTUs associated with source, 21
-OTUs associated with clindamycin treatment, and 6 OTUs associated with
-both (Fig. 5B). Together, these results suggest that the initial
-bacterial communities and their responses to clindamycin influenced the
-clearance of \emph{C. difficile}.
+1D, 3D, 4D and Data Set S1, Sheets 5, 8, and 10) or were impacted by
+clindamycin treatment (Fig. 3E and Data Set S1, Sheet 9; Fig. S4).
+Comparing the features important to the 3 models identified 14 OTUs
+associated with source, 21 OTUs associated with clindamycin treatment,
+and 6 OTUs associated with both (Fig. 5B). Together, these results
+suggest that the initial bacterial communities and their responses to
+clindamycin influenced the clearance of \emph{C. difficile}.
 
@@
 9 days post-infection, there was at least 1 timepoint when the relative
-abundances of these OTUs significantly varied between sources (Table
-S15). Interestingly, there were no OTUs that emerged as consistently
-enriched or depleted in mice that were colonized past 7 days
-post-infection, suggesting that multiple bacteria influence \emph{C.
-difficile} colonization dynamics.
+abundances of these OTUs significantly varied between sources (Data Set
+S1, Sheet 15). Interestingly, there were no OTUs that emerged as
+consistently enriched or depleted in mice that were colonized past 7
+days post-infection, suggesting that multiple bacteria influence
+\emph{C. difficile} colonization dynamics.
 
@@
 day of the experiment and the adjusted \emph{P} value was \textless{}
-0.05 for panel A (Table S1). None of the \emph{P} values from pairwise
-Wilcoxon comparisons between sources were significant after
-Benjamini-Hochberg correction (Table S2). Gray lines represent the
-median values for each source of mice. C. Principal Coordinates Analysis
-(PCoA) of \(\theta_{YC}\) distances of baseline stool samples. Source
-and the interaction between source and cage effects explained most of
-the variation (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
-\textless{} 0.001; Table S3). For A-C: each symbol represents the value
-for a stool sample from an individual mouse, circles represent
-experiment 1 mice and triangles represent experiment 2 mice. D. The
-median (point) and interquantile range (colored lines) of the relative
-abundances for the 20 most significant OTUs out of the 268 OTUs that
-varied across sources at baseline by Kruskal-Wallis test with
-Benjamini-Hochberg correction (Table S5).
+0.05 for panel A (Data Set S1, Sheet 1). None of the \emph{P} values
+from pairwise Wilcoxon comparisons between sources were significant
+after Benjamini-Hochberg correction (Data Set S1, Sheet 2). Gray lines
+represent the median values for each source of mice. C. Principal
+Coordinates Analysis (PCoA) of \(\theta_{YC}\) distances of baseline
+stool samples. Source and the interaction between source and cage
+effects explained most of the variation (PERMANOVA combined
+R\textsuperscript{2} = 0.90, \emph{P} \textless{} 0.001; Data Set S1,
+Sheet 3). For A-C: each symbol represents the value for a stool sample
+from an individual mouse, circles represent experiment 1 mice and
+triangles represent experiment 2 mice. D. The median (point) and
+interquantile range (colored lines) of the relative abundances for the
+20 most significant OTUs out of the 268 OTUs that varied across sources
+at baseline by Kruskal-Wallis test with Benjamini-Hochberg correction
+(Data Set S1, Sheet 5).
 
@@
 Kruskal-Wallis test with Benjamini-Hochberg correction for testing
-across multiple days (Table S6 and Table S7) are reflected by the
+across multiple days (Data Set S1, Sheets 6-7) are reflected by the
 asterisk above each timepoint (*, \emph{P} \textless{} 0.05). Lines
@@
 Benjamini-Hochberg correction for testing each day of the experiment and
-the adjusted \emph{P} value was \textless{} 0.05 (Table S1). Significant
-\emph{P} values from the pairwise Wilcoxon comparisons between sources
-with Benjamini-Hochberg correction are displayed as the first initial of
-each group compared to the group that they are listed above (Table S2).
-C. PCoA of \(\theta_{YC}\) distances from stools collected
-post-clindamycin. Source and the interaction between source and cage
-effects explained most of the variation observed post-clindamycin
-(PERMANOVA combined R\textsuperscript{2} = 0.99, \emph{P} \textless{}
-0.001; Table S3). For A-C, each symbol represents a stool sample from an
-individual mouse, with circles representing experiment 1 mice and
-triangles representing experiment 2 mice. D. The median (point) and
-interquantile range (colored lines) of the relative abundances for the
-18 OTUs (Table S8) that varied between sources after clindamycin
-treatment (day 0). E. The median (point) and interquantile range
-(colored lines) of the top 10 most significant OTUs out of 153 with
-relative abundances that changed because of the clindamycin treatment
-(adjusted \emph{P} value \textless{} 0.05). Data were analyzed by paired
-Wilcoxon signed rank test of mice that had paired sequence data for
-baseline (day -1) and post-clindamycin (day 0) timepoints (N = 31), with
-Benjamini-Hochberg correction for testing all identified OTUs (Table
-S9). The gray vertical line indicates the limit of detection.
+the adjusted \emph{P} value was \textless{} 0.05 (Data Set S1, Sheet 1).
+Significant \emph{P} values from the pairwise Wilcoxon comparisons
+between sources with Benjamini-Hochberg correction are displayed as the
+first initial of each group compared to the group that they are listed
+above (Data Set S1, Sheet 2). C. PCoA of \(\theta_{YC}\) distances from
+stools collected post-clindamycin. Source and the interaction between
+source and cage effects explained most of the variation observed
+post-clindamycin (PERMANOVA combined R\textsuperscript{2} = 0.99,
+\emph{P} \textless{} 0.001; Data Set S1, Sheet 3). For A-C, each symbol
+represents a stool sample from an individual mouse, with circles
+representing experiment 1 mice and triangles representing experiment 2
+mice. D. The median (point) and interquantile range (colored lines) of
+the relative abundances for the 18 OTUs (Data Set S1, Sheet 8) that
+varied between sources after clindamycin treatment (day 0). E. The
+median (point) and interquantile range (colored lines) of the top 10
+most significant OTUs out of 153 with relative abundances that changed
+because of the clindamycin treatment (adjusted \emph{P} value
+\textless{} 0.05). Data were analyzed by paired Wilcoxon signed rank
+test of mice that had paired sequence data for baseline (day -1) and
+post-clindamycin (day 0) timepoints (N = 31), with Benjamini-Hochberg
+correction for testing all identified OTUs (Data Set S1, Sheet 9). The
+gray vertical line indicates the limit of detection.
 
@@
 Benjamini-Hochberg correction for testing each day of the experiment and
-the adjusted \emph{P} value was \textless{} 0.05 (Table S1). Significant
-\emph{P} values from the pairwise Wilcoxon comparisons between sources
-with Benjamini-Hochberg correction are displayed as the first initial of
-each group compared to the group that they are listed above (Table S2).
-PCoA of \(\theta_{YC}\) distances of 1-day post-infection stool samples.
-Source and the interaction between source and cage effects explained
-most of the variation between fecal communities (PERMANOVA combined
-R\textsuperscript{2} = 0.88, \emph{P} \textless{} 0.001; Table S3). For
-A-C: each symbol represents the value for a stool sample from an
-individual mouse, circles represent experiment 1 mice and triangles
-represent experiment 2 mice. D. The median (point) and interquantile
-range (colored lines) of the relative abundances for the top 20 most
-significant OTUs out of the 44 OTUs that varied between sources 1-day
-post-infection. The gray vertical line indicates the limit of detection.
-For each timepoint OTUs with differential relative abundances across
-sources of mice were identified by Kruskal-Wallis test with
-Benjamini-Hochberg correction for testing all identified OTUs (Table
-S10). E. \(\theta_{YC}\) distances of fecal samples collected 7-days
-post-infection relative to the baseline (day -1) sample for each mouse.
-Each symbol represents an individual mouse. Gray lines represent the
-median for each source.
+the adjusted \emph{P} value was \textless{} 0.05 (Data Set S1, Sheet 1).
+Significant \emph{P} values from the pairwise Wilcoxon comparisons
+between sources with Benjamini-Hochberg correction are displayed as the
+first initial of each group compared to the group that they are listed
+above (Data Set S1, Sheet 2). PCoA of \(\theta_{YC}\) distances of 1-day
+post-infection stool samples. Source and the interaction between source
+and cage effects explained most of the variation between fecal
+communities (PERMANOVA combined R\textsuperscript{2} = 0.88, \emph{P}
+\textless{} 0.001; Data Set S1, Sheet 3). For A-C: each symbol
+represents the value for a stool sample from an individual mouse,
+circles represent experiment 1 mice and triangles represent experiment 2
+mice. D. The median (point) and interquantile range (colored lines) of
+the relative abundances for the top 20 most significant OTUs out of the
+44 OTUs that varied between sources 1-day post-infection. The gray
+vertical line indicates the limit of detection. For each timepoint OTUs
+with differential relative abundances across sources of mice were
+identified by Kruskal-Wallis test with Benjamini-Hochberg correction for
+testing all identified OTUs (Data Set S1, Sheet 10). E. \(\theta_{YC}\)
+distances of fecal samples collected 7-days post-infection relative to
+the baseline (day -1) sample for each mouse. Each symbol represents an
+individual mouse. Gray lines represent the median for each source.
 
@@
 combines OTUs that were important to the day -1, 0, and 1 classification
-models ( Fig. S4, Table S14) and either overlapped with taxa that varied
-across sources at the same timepoint, were impacted by clindamycin
-treatment, or both. Bold OTUs were important to more than 1
+models ( Fig. S4, Data Set S1, Sheet 14) and either overlapped with taxa
+that varied across sources at the same timepoint, were impacted by
+clindamycin treatment, or both. Bold OTUs were important to more than 1
 classification model.
@@
 statistically significant by Kruskal-Wallis test with Benjamini-Hochberg
-correction for testing across multiple days (Table S15) are identified
-by the asterisk above each timepoint (*, P \textless{} 0.05).
+correction for testing across multiple days (Data Set S1, Sheet 15) are
+identified by the asterisk above each timepoint (*, P \textless{} 0.05).
 
@@
 R\textsuperscript{2} = 0.99; \emph{P} \(\le\) 0.033) and Young (combined
-R\textsuperscript{2} = 0.95; \emph{P} \(\le\) 0.03) mice (Table S4).
-G-H: Boxplots of the \(\theta_{YC}\) distances of the 6 sources of mice
-relative to mice within the same source and experiment (G) or mice
-within the same source and between experiments (H) at baseline (day -1).
-Symbols represent individual mouse samples: circles for experiment 1 and
-triangles for experiment 2.
+R\textsuperscript{2} = 0.95; \emph{P} \(\le\) 0.03) mice (Data Set S1,
+Sheet 4). G-H: Boxplots of the \(\theta_{YC}\) distances of the 6
+sources of mice relative to mice within the same source and experiment
+(G) or mice within the same source and between experiments (H) at
+baseline (day -1). Symbols represent individual mouse samples: circles
+for experiment 1 and triangles for experiment 2.
 
@@
 performed better than random chance (AUROC = 0.5, all \emph{P}
-\textless{} 0.001, Table S12) and the model built with post-clindamycin
-bacterial OTU relative abundances had the best performance
-((\emph{P}\textsubscript{FDR} \textless{} 0.001 for all pairwise
-comparisons, Table S13). See Table S14 for list of the 20 OTUs that were
-ranked as most important to each model.
+\textless{} 0.001, Data Set S1, Sheet 12) and the model built with
+post-clindamycin bacterial OTU relative abundances had the best
+performance ((\emph{P}\textsubscript{FDR} \textless{} 0.001 for all
+pairwise comparisons, Data Set S1, Sheet 13). See Data Set S1, Sheet 14
+for list of the 20 OTUs that were ranked as most important to each
+model.
 
@@
 from the top 20 OTUs from the baseline (A), post-clindamycin treatment
-(B), and 1-day post-infection (C) classification models (Table S14) that
-overlapped with OTUs that varied across sources at the corresponding
-timepoint (Tables S5, 8, 10), were impacted by clindamycin treatment
-(Table S9), or both. Bold OTUs were important to more than 1
-classification model.
+(B), and 1-day post-infection (C) classification models (Data Set S1,
+Sheet 14) that overlapped with OTUs that varied across sources at the
+corresponding timepoint (Data Set S1, Sheets 5, 8, and 10), were
+impacted by clindamycin treatment (Data Set S1, Sheet 9), or both. Bold
+OTUs were important to more than 1 classification model.
 
@@
 
-\hypertarget{supplementary-tables-and-movie}{%
-\subsection{Supplementary Tables and
-Movie}\label{supplementary-tables-and-movie}}
+\hypertarget{supplementary-movie-and-data-set-s1}{%
+\subsection{Supplementary Movie and Data Set
+S1}\label{supplementary-movie-and-data-set-s1}}
 
@@
 variation across fecal communities (PERMANOVA source
-R\textsuperscript{2} = 0.35, \emph{P} = 0.0001, Table S11) followed by
-interactions between cage effects and day of the experiment.
+R\textsuperscript{2} = 0.35, \emph{P} = 0.0001, Data Set S1, Sheet 11)
+followed by interactions between cage effects and day of the experiment.
 Transparency of the symbol corresponds to the day of the experiment,
@@
 
-\textbf{Tables S1-S15. Excel workbook of Tables S1-S15.}
-
-\textbf{Table S1. Alpha diversity metrics Kruskal-Wallis statistical
+\textbf{Data Set S1, Sheets 1-15. Excel workbook with 15 sheets.}
+
+\textbf{Data Set S1, Sheet 1. Alpha diversity metrics Kruskal-Wallis
+statistical results.}
+
+\textbf{Data Set S1, Sheet 2. Alpha diversity metrics pairwise Wilcoxon
+statistical results.}
+
+\textbf{Data Set S1, Sheet 3. PERMANOVA results for mice at baseline
+(day -1), post-clindamycin (day 0), and post-infection (day 1).}
+
+\textbf{Data Set S1, Sheet 4. PERMANOVA results for each source of mice
+at baseline (day -1).}
+
+\textbf{Data Set S1, Sheet 5. OTUs with relative abumdances that
+significantly vary between sources at baseline (day -1).}
+
+\textbf{Data Set S1, Sheet 6. \emph{C. difficile} CFU statistical
 results.}
 
-\textbf{Table S2. Alpha diversity metrics pairwise Wilcoxon statistical
-results.}
-
-\textbf{Table S3. PERMANOVA results for mice at baseline (day -1),
-post-clindamycin (day 0), and post-infection (day 1).}
-
-\textbf{Table S4. PERMANOVA results for each source of mice at baseline
-(day -1).}
-
-\textbf{Table S5. OTUs with relative abumdances that significantly vary
-between sources at baseline (day -1).}
-
-\textbf{Table S6. \emph{C. difficile} CFU statistical results.}
-
-\textbf{Table S7. Mouse weight change statistical results.}
-
-\textbf{Table S8. OTUs with relative abundances that significantly vary
-between sources post-clindamycin (day 0).}
-
-\textbf{Table S9. OTUs with relative abundances that significantly
-changed after clindamycin treatment.}
-
-\textbf{Table S10. OTUs with relative abundances that significantly vary
-between sources 1-day post-infection.}
-
-\textbf{Table S11. PERMANOVA results for mice across all timepoints.}
-
-\textbf{Table S12. Statistical results of L2-regularized logistic
-regression model performances compared to random chance.}
-
-\textbf{Table S13. Pairwise comparisons of L2-regularized logistic
-regression model performances.}
-
-\textbf{Table S14. Top 20 most important OTUs for each of the 3
-L2-regularized logistic regression models based on OTU relative
+\textbf{Data Set S1, Sheet 7. Mouse weight change statistical results.}
+
+\textbf{Data Set S1, Sheet 8. OTUs with relative abundances that
+significantly vary between sources post-clindamycin (day 0).}
+
+\textbf{Data Set S1, Sheet 9. OTUs with relative abundances that
+significantly changed after clindamycin treatment.}
+
+\textbf{Data Set S1, Sheet 10. OTUs with relative abundances that
+significantly vary between sources 1-day post-infection.}
+
+\textbf{Data Set S1, Sheet 11. PERMANOVA results for mice across all
+timepoints.}
+
+\textbf{Data Set S1, Sheet 12. Statistical results of L2-regularized
+logistic regression model performances compared to random chance.}
+
+\textbf{Data Set S1, Sheet 13. Pairwise comparisons of L2-regularized
+logistic regression model performances.}
+
+\textbf{Data Set S1, Sheet 14. Top 20 most important OTUs for each of
+the 3 L2-regularized logistic regression models based on OTU relative
 abundance data.}
 
-\textbf{Table S15. OTUs with relative abundances that significantly
-varied between sources of mice on at least 1 day of the experiment by
-Kruskal-Wallis test.}
+\textbf{Data Set S1, Sheet 15. OTUs with relative abundances that
+significantly varied between sources of mice on at least 1 day of the
+experiment by Kruskal-Wallis test.}
 

### 2020-10-14 submission/manuscript.Rmd
Edits from proofs

@@
 difficile* is detectable, but symptoms are absent, has been documented in infants and adults [@Zacharioudakis2015; @Crobach2018].
-The intestinal microbiota has been implicated in asymptomatic colonization [@Zhang2015; @VanInsberghe2020], susceptibility to CDIs [@Mancabelli2017, @Duvallet2017], and adverse CDI outcomes [@Seekatz2016; @Khanna2016; @Pakpour2017; @Lee2020].
+The intestinal microbiota has been implicated in asymptomatic colonization [@Zhang2015; @VanInsberghe2020], susceptibility to CDIs [@Mancabelli2017; @Duvallet2017], and adverse CDI outcomes [@Seekatz2016; @Khanna2016; @Pakpour2017; @Lee2020].
 However, it is not clear how much inter-individual microbiota variation contributes to the range of outcomes observed after *C.
@@
 The microbiota is composed of viruses, fungi, and parasites in addition to bacteria, and these non-bacterial members can also vary across sources of mice [@Rasmussen2019; @Mims2020].
-While our study focused solely on the bacterial portion, viruses and fungi have also begun to be implicated in the context of CDIs or FMT treatments for recurrent CDIs [@Markey2018; @Stewart2019; @Ott2017; @Zuo2017; @Zuo2018].
+While our study focused solely on the bacterial portion, viruses and fungi have also begun to be implicated in the context of CDIs or fecal microbiota transplant (FMT) treatments for recurrent CDIs [@Markey2018; @Stewart2019; @Ott2017; @Zuo2017; @Zuo2018].
 Beyond community composition, the metabolic function of the microbiota also has a CDI signature [@Robinson2019; @Jenior2017; @Fletcher2018; @Jump2014] and can vary across mice from different sources [@Xiao2015].
@@
 **(i) Animals.** All experiments were approved by the University of Michigan Animal Care and Use Committee (IACUC) under protocol number PRO00006983.
-Female C57BL/7 mice were obtained from 6 different sources: The Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, Envigo, and two colonies at the University of Michigan (the Schloss lab colony and the Young lab colony).
+Female C57BL/6 mice were obtained from 6 different sources: The Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, Envigo, and two colonies at the University of Michigan (the Schloss lab colony and the Young lab colony).
 The Young lab colony was originally established with mice purchased from Jackson in 2002, and the Schloss lab colony was established in 2010 with mice donated from the Young lab.

### 2020-10-14 submission/manuscript.tex
Edits from proofs

@@
 been implicated in asymptomatic colonization (5, 6), susceptibility to
-CDIs (7), and adverse CDI outcomes (9--12). However, it is not clear how
-much inter-individual microbiota variation contributes to the range of
-outcomes observed after \emph{C. difficile} exposure relative to other
-risk factors.
+CDIs (7, 8), and adverse CDI outcomes (9--12). However, it is not clear
+how much inter-individual microbiota variation contributes to the range
+of outcomes observed after \emph{C. difficile} exposure relative to
+other risk factors.
 
@@
 viruses and fungi have also begun to be implicated in the context of
-CDIs or FMT treatments for recurrent CDIs (35, 58--61). Beyond community
-composition, the metabolic function of the microbiota also has a CDI
-signature (20, 47, 62, 63) and can vary across mice from different
-sources (64). For example, microbial metabolites, particularly secondary
-bile acids and butyrate production, have been implicated as important
-contributors to \emph{C. difficile} resistance (33, 44). Interestingly,
-butyrate has previously been shown to vary across mouse vendors and
-mediated resistance to \emph{Citrobacter rodentium} infection, a model
-of enterohemorrhagic and enteropathogenic \emph{Escherichia coli}
-infections (23). Evidence for immunological toning differences in IgA
-and Th17 cells across mice from different vendors have also been
-documented (65, 66) and could influence the host response to CDI (67,
-68), particularly relevant for \emph{C. difficile} strains that induce
-more severe disease than \emph{C. difficile} 630. The outcome after
-\emph{C. difficile} exposure depends on a multitude of factors,
-including genetics, age, diet, and immunity; all of which also influence
-the microbiota.
+CDIs or fecal microbiota transplant (FMT) treatments for recurrent CDIs
+(35, 58--61). Beyond community composition, the metabolic function of
+the microbiota also has a CDI signature (20, 47, 62, 63) and can vary
+across mice from different sources (64). For example, microbial
+metabolites, particularly secondary bile acids and butyrate production,
+have been implicated as important contributors to \emph{C. difficile}
+resistance (33, 44). Interestingly, butyrate has previously been shown
+to vary across mouse vendors and mediated resistance to
+\emph{Citrobacter rodentium} infection, a model of enterohemorrhagic and
+enteropathogenic \emph{Escherichia coli} infections (23). Evidence for
+immunological toning differences in IgA and Th17 cells across mice from
+different vendors have also been documented (65, 66) and could influence
+the host response to CDI (67, 68), particularly relevant for \emph{C.
+difficile} strains that induce more severe disease than \emph{C.
+difficile} 630. The outcome after \emph{C. difficile} exposure depends
+on a multitude of factors, including genetics, age, diet, and immunity;
+all of which also influence the microbiota.
 
@@
 Michigan Animal Care and Use Committee (IACUC) under protocol number
-PRO00006983. Female C57BL/7 mice were obtained from 6 different sources:
+PRO00006983. Female C57BL/6 mice were obtained from 6 different sources:
 The Jackson Laboratory, Charles River Laboratories, Taconic Biosciences,
